### 2021-05-29 manuscript/main.tex
added the whole body draft

\documentclass{bioinfo}
\copyrightyear{2021} \pubyear{2021}

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Manuscript Category}

\begin{document}
    \firstpage{1}

    \subtitle{Genome Analysis}

    \title[short Title]{ODGI: scalable tools for working with pangenome graphs}
    \author[Sample \textit{et~al}.]{
        Andrea Guarracino\,$^{\text{\sfb 1,}*}$,
        Simon Heumos\,$^{\text{\sfb 2}}$,
        XXXXX,
        Pjotr Prins\,$^{\text{\sfb 3}}$ and
        Erik Garrison\,$^{\text{\sfb 3,}*}$
    }
    \address{$^{\text{\sf 1}}$Department, Institution, City, Post Code, Country \\
        $^{\text{\sf 2}}$Department, Institution, City, Post Code, Country \\
        $^{\text{\sf 3}}$Department, Institution, City, Post Code, Country
    }

    \corresp{$^\ast$To whom correspondence should be addressed.}

    \history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

    \editor{Associate Editor: XXXXXXX}

    \abstract{
        \textbf{Motivation:}
        Pangenome graphs provide a complete representation of the mutual alignment of
        collections of genomes. These models, therefore, offer the opportunity to study the entire genetic diversity of
        a population, bringing new challenges to how to analyze hundreds of genomes on a gigabase scale. In addition,
        vertebrate genomes present highly repetitive regions which increase the complexity of the analysis.
        Efficient and flexible tools are therefore required to work on pangenome graphs of any complexity and scale.\\
        \textbf{Results:} Here we present ODGI, a new suite of tools to analyze and manipulate pangenome graphs
        structured in the variation graph model. ODGI supports graphs in the Graphical Fragment Assembly format version
        1 (GFAv1). The tools scale efficiently to large collections of eukaryotic genomes, allowing users to work easily
        with pangenome graphs of tens to thousands of large genomes while maintaining fast execution runtime and low
        memory overhead.\\
        \textbf{Availability:} ODGI is written in C++. Source code and user manual are freely available at
        \href{https://github.com/pangenome/odgi}{https://github.com/pangenome/odgi}.\\
        \textbf{Contact:}
        \href{andreaguarracino@outlook.com}{andreaguarracino@outlook.com},
        \href{erik.garrison@gmail.com}{erik.garrison@gmail.com}\\
        \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.}
    \maketitle

    \section{Introduction}

    A pangenome models the full set of genomic elements in a given species or clade  \citep{32453966}. Indeed, these
    data structures encode the mutual relationships between all the genomes represented, in contrast to reference-based
    approaches which relate sequences only to a particular genome, chosen as reference. A class of methods to represent
    pangenomes involves the \textit{sequence graphs} \citep{2488477}, where similar orthologous regions between genomes
    collapse into a single representative sequence. In \textit{node-labeled} sequence graphs, nodes indicate DNA
    sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph. A
    \textit{bidirected} sequence graph represents both strands of DNA. On this model, \textit{variation graphs} add the
    concept of \textit{path} to embed linear sequences (e.g., genomes or haplotypes) into the graph \citep{30125266}.
    Paths provide a stable coordinate system, allowing graph annotations and interoperability between different graphs.

    Thanks to advances in sequencing technologies, new genome assemblies are produced at a high rate. This amount of
    data offers the opportunity to study genomic variation as never before, but it also brings challenges in how to
    represent and manipulate hundreds of genomes at a gigabase scale. Variation graphs compactly represent the full
    genetic variation across a population, but graph-based data structures present computational overheads: in addition
    to the sequences, they must also represent the graph topology. Several genome data structures were developed to
    address these challenges \citep{33040146}. Memory-efficient representations do not allow dynamic updates to the
    graph, such as adding new genomes or simplifying the graph topology. Furthermore, dynamic updates must perform
    efficiently on the complex graphs generated from whole-genome alignment of vertebrates. Highly repetitive regions,
    such as centromeres, segmental duplications, and acrocentric chromosomes lead to graphs which present paths with
    very high-depth nodes. The node depth is defined as the number of times in which the node is crossed by all the
    paths embedded in the graph. High depth increases the complexity of the operations performed on the paths and the
    nodes they cross.

    To overcome this limitation, we have developed an Optimized Dynamic Genome/Graph Implementation (ODGI). Each tool
    loads graphs into the succinct dynamic HandleGraph model \citep{33040146} and applies methods based on the
    HandleGraph API. Several algorithms are shared with the VG toolkit (VG) \citep{30125266}, the state-of-the-art
    toolkit for working with pangenome graphs. However, although similar in scope to VG, ODGI importantly differs in
    its design to support highly complex pangenome graphs. The majority of the ODGI tools are implemented in an
    index-free manner, avoiding the burning task needed to create an index structure. Indeed, index-constrained
    implementations would require building indexes also for single or little operations on the graphs. Moreover,
    thanks to its efficient path representation, ODGI tools can build the internal graph representation (the ODGI
    format) in parallel, which is one of the major bottlenecks when working with very large and deep variation graphs,
    efficiently scaling up to hundreds of large vertebrate genomes.

%    Text Text Text Text Text Text Text Text Text Text Text Text Text
%    Text Text Text Text Text Text Text Text. Figure~\ref{fig:01}
%    shows that the above method Text Text Text Text Text Text Text
%    Text Text Text Text Text.  \citep{Bag01} wants to know about
%        {\ldots}{\ldots} text follows.
%    \begin{equation}
%        \sum \text{\it x}+ \text{\it y} =\text{\it Z}\label{eq:01}\vspace*{-10pt}
%    \end{equation}

%\enlargethispage{12pt}

%    \section{Approach}
%
%    Equation~(\ref{eq:01}) Text Text Text Text Text Text Text Text
%    Text Text Text Text Text Text Text Text Text Text Text Text Text.
%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%    Text Text Text Text Text Text Text Text Text Text Text.
%    \citealp{Boffelli03} might want to know about text text text text
%    .....

    \begin{methods}
        \section{Methods}

        ODGI is designed to build and modify paths in parallel, keeping its representation in memory efficient. It
        applies concepts first introduced in the dynamic version of the Graph BWT citep{31406990}, a scalable
        implementation of the graph extension of the positional Burrows-Wheeler transform \citep{28702075}. Specifically,
        we build a \textit{node-local} model of the graph, where the graph is stored in a vector of node structures.
        The identifier of each node is its offset in this vector plus 1. Each structure contains the sequence of the
        node, its edges in both directions (forward and backward), and a vector of \textit{path steps} that describes
        how the embedded paths walk across the node. Each step records the path identifier and the previous and next
        path steps on the same path. To reduce the encoding, we encode path steps using a local alphabet that maps the
        \textit{N} neighbors of the node into the range \textit{1..N}. To further save space, \textit{node deltas},
        rather than IDs, are stored in this alphabet. The delta between two nodes is defined as their distance in the
        graph vector (i.e., the difference between the node offsets). The path step vector, of length \textit{n}, is
        stored using a dynamic succinct integer vector that requires \textit{O(5nw)} bits, where \textit{w} is
        \textit{\textasciitilde log\textsubscript{2}(N)}. Finally, to allow operating on path steps in parallel, each node structure includes a
        byte-width mutex lock.

%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text
%        text\vspace*{1pt}

%        \begin{itemize}
%            \item for bulleted list, use itemize
%            \item for bulleted list, use itemize
%            \item for bulleted list, use itemize\vspace*{1pt}
%        \end{itemize}

%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text\vadjust{\newpage}.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%
%
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%
%
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%
%        \subsection{This is subheading}
%
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text
%        text. Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%
%        \subsubsection{This is subsubheading}
%
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%
%
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text\break
%        text text
%
%
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text
%        textText Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%
%        \enlargethispage{6pt}
%
%
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%
%
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%
%
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%
%
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%        Text Text Text Text Text Text Text Text Text Text Text Text Text
%        Text Text Text Text Text Text Text Text.
%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%        Text Text Text Text Text Text Text Text Text Text Text.
%        \citealp{Boffelli03} might want to know about text text text text
%
%        \begin{table}[!t]
%            \processtable{This is table caption\label{Tab:01}} {\begin{tabular}{@{}llll@{}}
%                                                                    \toprule head1 &
%                                                                    head2 & head3 & head4 \\\midrule
%                                                                    row1 & row1 & row1 & row1 \\
%                                                                    row2 & row2 & row2 & row2 \\
%                                                                    row3 & row3 & row3 & row3 \\
%                                                                    row4 & row4 & row4 & row4 \\\botrule
%            \end{tabular}}{This is a footnote}
%        \end{table}

    \end{methods}

    \section{Results}

    ODGI is entirely focused on “first-order” operations on the graph (Table~\ref{tab:01}), providing, unlike VG, no
    method to construct graphs from raw sequences, map reads, or call variants. It supports interrogation of the graph
    for basic dimensional properties, positional and path-related queries, and visualization in one-dimension (1D) and
    two-dimensions (2D). ODGI also supports operations on the graph such as subset, subdivide, break, combine,
    normalize, or order its components, nodes, and paths. Most of the tools are designed to be applied together,
    piping the output of one tool into the next, thereby preventing the creation of intermediate files, and reducing
    the number of IO operations.

    \subsection{Functionalities}

    Variation graphs are commonly represented by a subset of version 1 of the Graphical Fragment Assembly format
    (\citep{GFAv1}): \texttt{odgi build} and \texttt{odgi view} allow to convert graphs from GFA to ODGI format and
    vice versa, respectively.

    In the variation graph model, paths have to respect the graph’s topology: this can be verified with
    \texttt{odgi validate}, to ensure no errors in the input or edited graphs.

    By visualizing pangenome graphs, we can gain insight into the relationship between genomes. \texttt{odgi viz}
    applies binding and direct rendering to a raster image to generate a 1D representation able to scale up to gigabase
    pangenome graphs (Figure~1.A\vphantom{\ref{fig:01}}). However, pangenome graphs may present complex structures to
    visualize: \texttt{odgi sort} finds the best node order to simplify 1D graph visualization by applying several
    cutting-edge sorting algorithms. \texttt{odgi drawn} extends the visualization in 2D, taking in input the layouts
    built by \texttt{odgi layout}.

    Applying \texttt{odgi stats}, users can retrieve metrics describing the graph properties, such as the number of
    nodes, edges, paths, and graph length. ODGI also offers more advanced tools for interrogating the graphs.
    \texttt{odgi bin} summarizes the path information into bins of specified size, enabling a summarized view of
    gigabase scale graphs. \texttt{odgi depth} returns the node depth as defined by query criteria, allowing users to
    detect the complex regions in the graph due to sequences with highly identical repeats. Complex motifs can be also
    detected with \texttt{odgi degree}, which returns the node degree as defined by query criteria. High degree nodes
    are the mirror of misassemblies or problems in the pangenome building, making the tool useful for debugging the
    process as well.

    \begin{table}[!t]
        \processtable{A selection of the tools available in ODGI
        \label{tab:01}} {
            \begin{tabular}{@{}ll@{}}
                \toprule Name & Description \\
                \midrule
                row1          & row1        \\
                row2          & row2        \\
                row3          & row3        \\
                row4          & row4        \\
                \botrule
            \end{tabular}}

    \end{table}

    ODGI allows edit operations on the graphs. Cycles in the graph complicate downstream analyses: \texttt{odgi break}
    removes the cycles, reducing the complexity of the graph topology. \texttt{odgi groom} removes spurious inverting
    links by exploring the graph from the orientation supported by most paths. To enable efficient sequence alignment
    against the graph, long nodes can be divided into shorter nodes at a maximum requested size using
    \texttt{odgi chop}. Partial order alignment, which consists of aligning sequences against a directed acyclic graph
    (DAG), is frequently used in pangenome building pipelines, but the current implementations return DAGs with 1-bp
    long nodes; \texttt{odgi unchop} allows joining nodes that can be merged without changing the graph topology, nor
    the embedded sequences, obtaining an equivalent, but more compact, representation of the graph.

    Pangenome graphs can embed multiple chromosomes as separated connected components (inter-chromosomal structural
    variants would join the components into bigger ones). \texttt{odgi explode} separates the connected components in
    different ODGI format files, while \texttt{odgi squeeze} allows merging multiple graphs into the same ODGI format
    file, preventing node ID collisions. \texttt{odgi extract} allows extracting specific subregions of the graph as
    defined by query criteria, thereby simplifying the downstream analyses and reducing the resources to work only
    with the extracted region.

    In variation graphs the coordinates are provided by the embedded path sequences. Indeed, the node IDs are not
    meant to be stable. \texttt{odgi position} finds, translates, and liftovers graph and path positions between
    different graphs by exploiting their shared path sequences (Figure~1.B\vphantom{\ref{fig:01}}).

    \begin{figure}[!tpb]%figure1
        \fboxsep=0pt\colorbox{gray}{
            \begin{minipage}[t]{235pt}
                \vbox to 100pt{
                    \vfill\hbox to
                    235pt{\hfill\fontsize{24pt}{24pt}\selectfont FPO\hfill}
                    \vfill
                }
            \end{minipage}}
%\centerline{\includegraphics{fig01.eps}}
        \caption{
            Visualizing the Major Histocompatibility Complex (MHC) \textit{locus} pangenome. A) odgi viz 1D
            visualization of 8 haploid, phased human genome assemblies plus the chm13 and GRCh38 reference genomes.
            Graph nodes’ are arranged from left to right. The colored bars represent the linearized renderings of the
            embedded paths versus the nodes in a binary matrix. The black lines under the paths are the links, which
            represent the graph topology. B) Bandage \citep{26099265} representation of the graph topology of a
            consensus MHC pangenome graph representing only variations bigger than 100 bps. Gene labels were applied
            exploiting \texttt{odgi position} to liftover gene coordinates from the full graph into the consensus
            graph.
        }\label{fig:01}
    \end{figure}

    \subsection{Performance on human data}

    Table~\ref{tab:02} shows a comparison of the operations necessary to extract a subgraph using ODGI and VG. The
    example consists in extracting the centromeres from a pangenome graph made from 88 haploid, phased human genome
    assemblies from 44 individuals from the Human Pangenome Reference Consortium year 1 assembly (see Data
    Availability), plus the chm13 and GRCh38 reference genomes, for a total of 90 haplotypes. All commands are provided
    in the Supplementary Material.

    \begin{table}[!t]
        \processtable{Centromeres extraction comparison using ODGI and VG
        \label{tab:02}} {
            \begin{tabular}{@{}lllll@{}}
                \toprule Operation & Tool & Runtime (mm:ss) & Memory (GB) & Output size (GB) \\
                \midrule
                row1               & row1 & row1            & row1        & row1             \\
                row2               & row2 & row2            & row2        & row2             \\
                row3               & row3 & row3            & row3        & row3             \\
                row4               & row4 & row4            & row4        & row4             \\
                \botrule
            \end{tabular}}

    \end{table}

    To work with ODGI and VG it is first necessary to convert the graph from GFA to ODGI and VG format, respectively.
    Thanks to the ODGI format's path implementation, \texttt{odgi build} embeds the paths in parallel, outperforming
    \texttt{vg convert} in the same task. For the same reason, \texttt{odgi extract} results 20 times faster than
    \texttt{vg chunk} in extracting a portion from the full graph.

    \section{Discussion}
    We implemented a tool suite, ODGI, which is simple to work with to analyse and manipulate pangenome graphs at a
    population scale. ODGI's path implementation allows working with graphs of any complexity, without particular
    limitations. We have demonstrated that ODGI tools efficiently perform with pangenome graphs embedding full human
    chromosomes. ODGI is not only a tool suite but also a library, already used in pangenome graphs building
    (\citep{pggb}) and normalization (\citep{smoothxg}). Furthermore, users can add new tools, exploiting ODGI as a
    framework where to implement algorithms working on variation graphs. Therefore, ODGI offers an efficient and
    extendable suite of tools ready for the upcoming challenges in the analyses of hundreds of genomes at a gigabase
    scale that characterize the pangenomic era.

%\begin{figure}[!tpb]%figure2
%%\centerline{\includegraphics{fig02.eps}}
%\caption{Caption, caption.}\label{fig:02}
%\end{figure}

%    Text Text Text Text Text Text Text Text Text Text Text Text Text
%    Text Text Text Text Text Text Text Text.
%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%    Text Text Text Text Text Text Text Text Text Text Text.
%    \citealp{Boffelli03} might want to know about text text text text
%    Text Text Text Text Text Text Text Text Text Text Text Text Text
%    Text Text Text Text Text Text Text Text.
%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%    Text Text Text Text Text Text Text Text Text Text Text.
%    \citealp{Boffelli03} might want to know about text text text text
%    Text Text Text Text Text Text Text Text Text Text Text Text Text
%    Text Text Text Text Text Text Text Text.
%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%    Text Text Text Text Text Text Text Text Text Text Text.
%    \citealp{Boffelli03} might want to know about text text text text
%
%    \subsection{Test1}
%
%    Text Text Text Text Text Text Text Text Text Text Text Text Text
%    Text Text Text Text Text Text Text Text.
%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%    Text Text Text Text Text Text Text Text Text Text Text.
%    \citealp{Boffelli03} might want to know about text text text text
%    Text Text Text Text Text Text Text Text Text Text Text Text Text
%    Text Text Text Text Text Text Text Text.
%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%    Text Text Text Text Text Text Text Text Text Text Text.
%    \citealp{Boffelli03} might want to know about text text text text
%    Text Text Text Text Text Text Text Text Text Text Text Text Text
%    Text Text Text Text Text Text Text Text.
%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%    Text Text Text Text Text Text Text Text Text Text Text.
%    \citealp{Boffelli03} might want to know about text text text text
%
%
%    \section{Discussion}
%
%    Text Text Text Text Text Text Text Text Text Text Text Text Text
%    Text Text Text Text Text Text Text Text.
%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%    Text Text Text Text Text Text Text Text Text Text Text.
%    \citealp{Boffelli03} might want to know about text text text text
%    Text Text Text Text Text Text Text Text Text Text Text Text Text
%    Text Text Text Text Text Text Text Text.
%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%    Text Text Text Text Text Text Text Text Text Text Text.
%    \citealp{Boffelli03} might want to know about text text text text
%    Text Text Text Text Text Text Text Text Text Text.
%
%
%    Table~\ref{tab:01} shows that Text Text Text Text Text Text Text
%    Text Text Text Text. Figure~2\vphantom{\ref{fig:02}} shows that
%    the above method Text Text. Text Text Text Text Text Text Text
%    Text Text. Figure~2\vphantom{\ref{fig:02}} shows that the above
%    method Text Text. Text Text Text Text Text Text Text Text Text.
%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%    Text.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%     please remove the " % " symbol from \centerline{\includegraphics{fig01.eps}}
%     as it may ignore the figures.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%    \section{Conclusion}
%
%    (Table~\ref{tab:01}) Text Text Text Text Text Text Text Text Text
%    Text Text Text Text Text Text Text Text Text Text Text Text.
%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%    Text Text Text Text Text Text Text Text Text Text Text.
%    \citealp{Boffelli03} might want to know about text text text text
%    Text Text Text Text Text Text Text Text Text Text Text Text Text
%    Text Text Text Text Text Text Text Text.
%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%    Text Text Text Text Text Text Text Text Text Text Text.
%    \citealp{Boffelli03} might want to know about text text text text
%    Text Text Text Text Text Text Text Text Text Text Text Text Text
%    Text Text Text Text Text Text Text Text.
%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%    Text Text Text Text Text Text Text Text Text Text Text.
%
%
%    Text Text Text Text Text Text Text Text Text Text Text Text Text
%    Text Text Text Text Text Text Text Text.
%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%    Text Text Text Text Text Text Text Text Text Text Text.
%    \citealp{Boffelli03} might want to know about text text text text
%
%    \begin{enumerate}
%        \item this is item, use enumerate
%        \item this is item, use enumerate
%        \item this is item, use enumerate
%    \end{enumerate}
%
%    Text Text Text Text Text Text Text Text Text Text Text Text Text
%    Text Text Text Text Text Text Text Text.
%    Figure~2\vphantom{\ref{fig:02}} shows\vadjust{\pagebreak} that the
%    above method Text Text Text Text Text Text Text Text Text Text
%    Text Text.  \citealp{Boffelli03} might want to know about text
%    text text text Text Text Text Text Text Text Text Text Text Text
%    Text Text Text Text Text Text Text Text Text Text Text.
%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%    Text Text Text Text Text Text Text Text Text Text Text.
%    \citealp{Boffelli03} might want to know about text text text text
%    Text Text Text Text Text Text Text Text Text Text Text Text Text
%    Text Text Text Text Text Text Text\break Text.
%
%
%    Text Text Text Text Text Text Text Text Text Text Text Text Text
%    Text Text Text Text Text Text Text Text.
%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
%    Text Text Text\vspace*{-10pt}

    \vspace*{+12pt}
    \textit{Financial Support}: none declared.

    \textit{Conflict of Interest}: none declared.
    \vspace*{+12pt}

    \section*{Data availability}

    Data used to build Human pangenome graphs is available at \href{https://github.com/human-pangenomics/HPP_Year1_Data_Freeze_v1.0}{https://github.com/human-pangenomics/HPP\_Year1\_Data\_Freeze\_v1.0}.

    \section*{Acknowledgements}
    %ToDo
    XXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX.

    \section*{Funding}
    %ToDo
    This work has been supported by the XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX.

%    \vspace*{-12pt}

%\bibliographystyle{natbib}
%\bibliographystyle{achemnat}
%\bibliographystyle{plainnat}
%\bibliographystyle{abbrv}
%\bibliographystyle{bioinformatics}
%
%\bibliographystyle{plain}
%
%\bibliography{Document}

    %ToDo
    \begin{thebibliography}{}

        \bibitem[Bofelli {\it et~al}., 2000]{Boffelli03}
        Bofelli,F., Name2, Name3 (2003) Article title, {\it Journal Name}, {\bf 199}, 133-154.

        \bibitem[Bag {\it et~al}., 2001]{Bag01}
        Bag,M., Name2, Name3 (2001) Article title, {\it Journal Name}, {\bf 99}, 33-54.

        \bibitem[Yoo \textit{et~al}., 2003]{Yoo03}
        Yoo,M.S. \textit{et~al}. (2003) Oxidative stress regulated genes
        in nigral dopaminergic neurnol cell: correlation with the known
        pathology in Parkinson's disease. \textit{Brain Res. Mol. Brain
        Res.}, \textbf{110}(Suppl. 1), 76--84.

        \bibitem[Lehmann, 1986]{Leh86}
        Lehmann,E.L. (1986) Chapter title. \textit{Book Title}. Vol.~1, 2nd edn. Springer-Verlag, New York.

        \bibitem[Crenshaw and Jones, 2003]{Cre03}
        Crenshaw, B.,III, and Jones, W.B.,Jr (2003) The future of clinical
        cancer management: one tumor, one chip. \textit{Bioinformatics},
        doi:10.1093/bioinformatics/btn000.

        \bibitem[Auhtor \textit{et~al}. (2000)]{Aut00}
        Auhtor,A.B. \textit{et~al}. (2000) Chapter title. In Smith, A.C.
        (ed.), \textit{Book Title}, 2nd edn. Publisher, Location, Vol. 1, pp.
        ???--???.

        \bibitem[Bardet, 1920]{Bar20}
        Bardet, G. (1920) Sur un syndrome d'obesite infantile avec
        polydactylie et retinite pigmentaire (contribution a l'etude des
        formes cliniques de l'obesite hypophysaire). PhD Thesis, name of
        institution, Paris, France.

    \end{thebibliography}
\end{document}

### 2021-05-29 manuscript/main.tex
reduced the comments

@@
 
-%    Text Text Text Text Text Text Text Text Text Text Text Text Text
-%    Text Text Text Text Text Text Text Text. Figure~\ref{fig:01}
-%    shows that the above method Text Text Text Text Text Text Text
-%    Text Text Text Text Text.  \citep{Bag01} wants to know about
-%        {\ldots}{\ldots} text follows.
-%    \begin{equation}
-%        \sum \text{\it x}+ \text{\it y} =\text{\it Z}\label{eq:01}\vspace*{-10pt}
-%    \end{equation}
-
 %\enlargethispage{12pt}
-
-%    \section{Approach}
-%
-%    Equation~(\ref{eq:01}) Text Text Text Text Text Text Text Text
-%    Text Text Text Text Text Text Text Text Text Text Text Text Text.
-%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%    Text Text Text Text Text Text Text Text Text Text Text.
-%    \citealp{Boffelli03} might want to know about text text text text
-%    .....
 
@@
         ODGI is designed to build and modify paths in parallel, keeping its representation in memory efficient. It
-        applies concepts first introduced in the dynamic version of the Graph BWT citep{31406990}, a scalable
+        applies concepts first introduced in the dynamic version of the Graph BWT \citep{31406990}, a scalable
         implementation of the graph extension of the positional Burrows-Wheeler transform \citep{28702075}. Specifically,
@@
 
+%        Text Text Text Text Text Text Text Text Text Text Text Text Text
+%        Text Text Text Text Text Text Text Text. Figure~\ref{fig:01}
+%        shows that the above method Text Text Text Text Text Text Text
+%        Text Text Text Text Text.  \citep{Bag01} wants to know about
+%            {\ldots}{\ldots} text follows.
+%        \begin{equation}
+%            \sum \text{\it x}+ \text{\it y} =\text{\it Z}\label{eq:01}\vspace*{-10pt}
+%        \end{equation}
+%
 %        Text Text Text Text Text Text Text Text Text Text Text Text Text
@@
 %        text\vspace*{1pt}
-
+%
 %        \begin{itemize}
@@
 %        \end{itemize}
-
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
 %
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text\vadjust{\newpage}.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%
-%
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%
-%
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
+%        \begin{enumerate}
+%            \item this is item, use enumerate
+%            \item this is item, use enumerate
+%            \item this is item, use enumerate
+%        \end{enumerate}
 %
@@
 %        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text
-%        text. Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%
-%        \subsubsection{This is subsubheading}
-%
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
 %
 %
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text\break
-%        text text
-%
-%
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text
-%        textText Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%
 %        \enlargethispage{6pt}
-%
-%
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%
-%
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%
-%
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%
-%
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
 %
@@
 
-%    Text Text Text Text Text Text Text Text Text Text Text Text Text
-%    Text Text Text Text Text Text Text Text.
-%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%    Text Text Text Text Text Text Text Text Text Text Text.
-%    \citealp{Boffelli03} might want to know about text text text text
-%    Text Text Text Text Text Text Text Text Text Text Text Text Text
-%    Text Text Text Text Text Text Text Text.
-%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%    Text Text Text Text Text Text Text Text Text Text Text.
-%    \citealp{Boffelli03} might want to know about text text text text
-%    Text Text Text Text Text Text Text Text Text Text Text Text Text
-%    Text Text Text Text Text Text Text Text.
-%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%    Text Text Text Text Text Text Text Text Text Text Text.
-%    \citealp{Boffelli03} might want to know about text text text text
-%
-%    \subsection{Test1}
-%
-%    Text Text Text Text Text Text Text Text Text Text Text Text Text
-%    Text Text Text Text Text Text Text Text.
-%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%    Text Text Text Text Text Text Text Text Text Text Text.
-%    \citealp{Boffelli03} might want to know about text text text text
-%    Text Text Text Text Text Text Text Text Text Text Text Text Text
-%    Text Text Text Text Text Text Text Text.
-%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%    Text Text Text Text Text Text Text Text Text Text Text.
-%    \citealp{Boffelli03} might want to know about text text text text
-%    Text Text Text Text Text Text Text Text Text Text Text Text Text
-%    Text Text Text Text Text Text Text Text.
-%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%    Text Text Text Text Text Text Text Text Text Text Text.
-%    \citealp{Boffelli03} might want to know about text text text text
-%
-%
-%    \section{Discussion}
-%
-%    Text Text Text Text Text Text Text Text Text Text Text Text Text
-%    Text Text Text Text Text Text Text Text.
-%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%    Text Text Text Text Text Text Text Text Text Text Text.
-%    \citealp{Boffelli03} might want to know about text text text text
-%    Text Text Text Text Text Text Text Text Text Text Text Text Text
-%    Text Text Text Text Text Text Text Text.
-%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%    Text Text Text Text Text Text Text Text Text Text Text.
-%    \citealp{Boffelli03} might want to know about text text text text
-%    Text Text Text Text Text Text Text Text Text Text.
-%
-%
-%    Table~\ref{tab:01} shows that Text Text Text Text Text Text Text
-%    Text Text Text Text. Figure~2\vphantom{\ref{fig:02}} shows that
-%    the above method Text Text. Text Text Text Text Text Text Text
-%    Text Text. Figure~2\vphantom{\ref{fig:02}} shows that the above
-%    method Text Text. Text Text Text Text Text Text Text Text Text.
-%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%    Text.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%    \section{Conclusion}
-%
-%    (Table~\ref{tab:01}) Text Text Text Text Text Text Text Text Text
-%    Text Text Text Text Text Text Text Text Text Text Text Text.
-%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%    Text Text Text Text Text Text Text Text Text Text Text.
-%    \citealp{Boffelli03} might want to know about text text text text
-%    Text Text Text Text Text Text Text Text Text Text Text Text Text
-%    Text Text Text Text Text Text Text Text.
-%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%    Text Text Text Text Text Text Text Text Text Text Text.
-%    \citealp{Boffelli03} might want to know about text text text text
-%    Text Text Text Text Text Text Text Text Text Text Text Text Text
-%    Text Text Text Text Text Text Text Text.
-%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%    Text Text Text Text Text Text Text Text Text Text Text.
-%
-%
-%    Text Text Text Text Text Text Text Text Text Text Text Text Text
-%    Text Text Text Text Text Text Text Text.
-%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%    Text Text Text Text Text Text Text Text Text Text Text.
-%    \citealp{Boffelli03} might want to know about text text text text
-%
-%    \begin{enumerate}
-%        \item this is item, use enumerate
-%        \item this is item, use enumerate
-%        \item this is item, use enumerate
-%    \end{enumerate}
-%
-%    Text Text Text Text Text Text Text Text Text Text Text Text Text
-%    Text Text Text Text Text Text Text Text.
-%    Figure~2\vphantom{\ref{fig:02}} shows\vadjust{\pagebreak} that the
-%    above method Text Text Text Text Text Text Text Text Text Text
-%    Text Text.  \citealp{Boffelli03} might want to know about text
-%    text text text Text Text Text Text Text Text Text Text Text Text
-%    Text Text Text Text Text Text Text Text Text Text Text.
-%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%    Text Text Text Text Text Text Text Text Text Text Text.
-%    \citealp{Boffelli03} might want to know about text text text text
-%    Text Text Text Text Text Text Text Text Text Text Text Text Text
-%    Text Text Text Text Text Text Text\break Text.
-%
-%
-%    Text Text Text Text Text Text Text Text Text Text Text Text Text
-%    Text Text Text Text Text Text Text Text.
-%    Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%    Text Text Text\vspace*{-10pt}
 

### 2021-05-29 manuscript/main.tex
fixed non-ASCII character

@@
 
-    ODGI is entirely focused on “first-order” operations on the graph (Table~\ref{tab:01}), providing, unlike VG, no
+    ODGI is entirely focused on "first-order" operations on the graph (Table~\ref{tab:01}), providing, unlike VG, no
     method to construct graphs from raw sequences, map reads, or call variants. It supports interrogation of the graph
@@
 
-    In the variation graph model, paths have to respect the graph’s topology: this can be verified with
+    In the variation graph model, paths have to respect the graph's topology: this can be verified with
     \texttt{odgi validate}, to ensure no errors in the input or edited graphs.

### 2021-05-29 manuscript/main.tex
added non-breaking space before the references

@@
 
-    A pangenome models the full set of genomic elements in a given species or clade \citep{32453966}. Indeed, these
+    A pangenome models the full set of genomic elements in a given species or clade~\citep{32453966}. Indeed, these
     data structures encode the mutual relationships between all the genomes represented, in contrast to reference-based
     approaches which relate sequences only to a particular genome, chosen as reference. A class of methods to represent
-    pangenomes involves the \textit{sequence graphs} \citep{2488477}, where similar orthologous regions between genomes
+    pangenomes involves the \textit{sequence graphs}~\citep{2488477}, where similar orthologous regions between genomes
     collapse into a single representative sequence. In \textit{node-labeled} sequence graphs, nodes indicate DNA
@@
     \textit{bidirected} sequence graph represents both strands of DNA. On this model, \textit{variation graphs} add the
-    concept of \textit{path} to embed linear sequences (e.g., genomes or haplotypes) into the graph \citep{30125266}.
+    concept of \textit{path} to embed linear sequences (e.g., genomes or haplotypes) into the graph~\citep{30125266}.
     Paths provide a stable coordinate system, allowing graph annotations and interoperability between different graphs.
@@
     to the sequences, they must also represent the graph topology. Several genome data structures were developed to
-    address these challenges \citep{33040146}. Memory-efficient representations do not allow dynamic updates to the
+    address these challenges~\citep{33040146}. Memory-efficient representations do not allow dynamic updates to the
     graph, such as adding new genomes or simplifying the graph topology. Furthermore, dynamic updates must perform
@@
     To overcome this limitation, we have developed an Optimized Dynamic Genome/Graph Implementation (ODGI). Each tool
-    loads graphs into the succinct dynamic HandleGraph model \citep{33040146} and applies methods based on the
-    HandleGraph API. Several algorithms are shared with the VG toolkit (VG) \citep{30125266}, the state-of-the-art
+    loads graphs into the succinct dynamic HandleGraph model~\citep{33040146} and applies methods based on the
+    HandleGraph API. Several algorithms are shared with the VG toolkit (VG)~\citep{30125266}, the state-of-the-art
     toolkit for working with pangenome graphs. However, although similar in scope to VG, ODGI importantly differs in
@@
         ODGI is designed to build and modify paths in parallel, keeping its representation in memory efficient. It
-        applies concepts first introduced in the dynamic version of the Graph BWT \citep{31406990}, a scalable
-        implementation of the graph extension of the positional Burrows-Wheeler transform \citep{28702075}. Specifically,
+        applies concepts first introduced in the dynamic version of the Graph BWT~\citep{31406990}, a scalable
+        implementation of the graph extension of the positional Burrows-Wheeler transform~\citep{28702075}. Specifically,
         we build a \textit{node-local} model of the graph, where the graph is stored in a vector of node structures.
@@
             embedded paths versus the nodes in a binary matrix. The black lines under the paths are the links, which
-            represent the graph topology. B) Bandage \citep{26099265} representation of the graph topology of a
+            represent the graph topology. B) Bandage~\citep{26099265} representation of the graph topology of a
             consensus MHC pangenome graph representing only variations bigger than 100 bps. Gene labels were applied

### 2021-05-31 manuscript/main.tex
Edit title and abstract

@@
 \documentclass{bioinfo}
+
+\usepackage{relsize} % for Cpp
+\usepackage{xspace} % for Cpp
+
+\newcommand{\odgi}{ODGI}
+\newcommand{\cmd}[1]{{\texttt{#1}}}
+\newcommand{\Rplus}{\protect\hspace{-.1em}\protect\raisebox{.35ex}{\smaller{\smaller\textbf{+}}}}
+\newcommand{\pp}{\Rplus\Rplus}
+\newcommand{\Cpp}{\mbox{C\Rplus\Rplus}\xspace}
+
 \copyrightyear{2021} \pubyear{2021}
@@
 
-    \title[short Title]{ODGI: scalable tools for working with pangenome graphs}
+    \title[short Title]{ODGI: scalable tools for pangenome graphs}
     \author[Sample \textit{et~al}.]{
@@
         Simon Heumos\,$^{\text{\sfb 2}}$,
-        XXXXX,
-        Pjotr Prins\,$^{\text{\sfb 3}}$ and
-        Erik Garrison\,$^{\text{\sfb 3,}*}$
+        % XXXXX,
+        Pjotr~Prins\,$^{\text{\sfb 3}}$ and
+        Erik~Garrison\,$^{\text{\sfb 3,}*}$
     }
@@
         $^{\text{\sf 2}}$Department, Institution, City, Post Code, Country \\
-        $^{\text{\sf 3}}$Department, Institution, City, Post Code, Country
+        $^{\text{\sf 3}}$Genetics Genomics \& Informatics, UTHSC, Memphis, TN, USA
     }
@@
 
-    \abstract{
-        \textbf{Motivation:}
-        Pangenome graphs provide a complete representation of the mutual alignment of
-        collections of genomes. These models, therefore, offer the opportunity to study the entire genetic diversity of
-        a population, bringing new challenges to how to analyze hundreds of genomes on a gigabase scale. In addition,
-        vertebrate genomes present highly repetitive regions which increase the complexity of the analysis.
-        Efficient and flexible tools are therefore required to work on pangenome graphs of any complexity and scale.\\
-        \textbf{Results:} Here we present ODGI, a new suite of tools to analyze and manipulate pangenome graphs
-        structured in the variation graph model. ODGI supports graphs in the Graphical Fragment Assembly format version
-        1 (GFAv1). The tools scale efficiently to large collections of eukaryotic genomes, allowing users to work easily
-        with pangenome graphs of tens to thousands of large genomes while maintaining fast execution runtime and low
-        memory overhead.\\
-        \textbf{Availability:} ODGI is written in C++. Source code and user manual are freely available at
+    \abstract{ \textbf{Motivation:} Pangenome graphs provide a
+      complete and `lossless' representation of genomic information
+      that allows for complex population-sized analysis of genomes
+      without introducing reference genome bias. Here we present a
+      range of novel tools that makes use of the \odgi\ efficient
+      in-memory representation of DNA/RNA graphs. \odgi\ can handle
+      thousands of large genomes while maintaining fast execution
+      runtime and low memory overhead. \odgi\ includes tools for
+      detecting complex regions, extracting loci, exploratory
+      analysis, sorting, removing artifacts and visualisation.
+      %
+      \\
+      \textbf{Availability:} \odgi\ is software written in the \Cpp\
+      programming language and published under the libre MIT license. Source
+      code and user manual can be downloaded from
       \href{https://github.com/pangenome/odgi}{https://github.com/pangenome/odgi}.\\
+      %
         \textbf{Contact:}
         \href{andreaguarracino@outlook.com}{andreaguarracino@outlook.com},
-        \href{erik.garrison@gmail.com}{erik.garrison@gmail.com}\\
-        \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.}
+        \href{erik.garrison@gmail.com}{erik.garrison@gmail.com}\\ \textbf{Supplementary
+          information:} Supplementary data are available at
+        \textit{Bioinformatics} online.}
+
     \maketitle

### 2021-05-31 manuscript/main.tex
Remove table - we'll do this in text

@@
 
-    \begin{table}[!t]
-        \processtable{A selection of the tools available in ODGI
-        \label{tab:01}} {
-            \begin{tabular}{@{}ll@{}}
-                \toprule Name & Description \\
-                \midrule
-                row1          & row1        \\
-                row2          & row2        \\
-                row3          & row3        \\
-                row4          & row4        \\
-                \botrule
-            \end{tabular}}
-
-    \end{table}
-
     ODGI allows edit operations on the graphs. Cycles in the graph complicate downstream analyses: \texttt{odgi break}
@@
 
-    \vspace*{+12pt}
-    \textit{Financial Support}: none declared.
+    % \vspace*{+12pt}
+    % \textit{Financial Support}: none declared.
 
@@
 
-    \section*{Data availability}
+    % \section*{Data availability}
 
-    Data used to build Human pangenome graphs is available at \href{https://github.com/human-pangenomics/HPP_Year1_Data_Freeze_v1.0}{https://github.com/human-pangenomics/HPP\_Year1\_Data\_Freeze\_v1.0}.
+    % Data used to build Human pangenome graphs is available at \href{https://github.com/human-pangenomics/HPP_Year1_Data_Freeze_v1.0}{https://github.com/human-pangenomics/HPP\_Year1\_Data\_Freeze\_v1.0}.
 
-    \section*{Acknowledgements}
+    % \section*{Acknowledgements}
     %ToDo
-    XXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX.
+    % XXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX.
 
-    \section*{Funding}
+    % \section*{Funding}
     %ToDo
-    This work has been supported by the XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX.
+    % This work has been supported by the XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX.
 

### 2021-05-31 manuscript/main.tex
Remove table - we'll do this in text
Proposed feature layout

@@
 \newcommand{\cmd}[1]{{\texttt{#1}}}
+\newcommand{\cmdbf}[1]{{\textbf{#1}}}
+\newcommand{\topic}[1]{{\cmdbf{#1}}:}
 \newcommand{\Rplus}{\protect\hspace{-.1em}\protect\raisebox{.35ex}{\smaller{\smaller\textbf{+}}}}
@@
 
-    \subsection{Functionalities}
+    \subsection{Features}
 
-    Variation graphs are commonly represented by a subset of version 1 of the Graphical Fragment Assembly format
-    (\citep{GFAv1}): \texttt{odgi build} and \texttt{odgi view} allow to convert graphs from GFA to ODGI format and
-    vice versa, respectively.
+    \odgi\ has a wide range of commands. Important features are
 
-    In the variation graph model, paths have to respect the graph's topology: this can be verified with
-    \texttt{odgi validate}, to ensure no errors in the input or edited graphs.
+    \topic{odgi build, view \& validate} pangenome graphs are commonly
+    represented by a subset of Graphical Fragment Assembly (GFAv1)
+    format\citep{GFAv1}.  \cmd{odgi build} and \cmd{odgi view} allow
+    to convert graphs from GFA to ODGI format and vice versa.  In the
+    variation graph model, paths have to respect the graph's topology:
+    this can be verified with \texttt{odgi validate}, to ensure no
+    errors in the input or edited graphs.
 
-    By visualizing pangenome graphs, we can gain insight into the relationship between genomes. \texttt{odgi viz}
-    applies binding and direct rendering to a raster image to generate a 1D representation able to scale up to gigabase
-    pangenome graphs (Figure~1.A\vphantom{\ref{fig:01}}). However, pangenome graphs may present complex structures to
-    visualize: \texttt{odgi sort} finds the best node order to simplify 1D graph visualization by applying several
-    cutting-edge sorting algorithms. \texttt{odgi drawn} extends the visualization in 2D, taking in input the layouts
-    built by \texttt{odgi layout}.
+    \topic{odgi viz, sort, drawn \& layout} by visualizing pangenome
+    graphs, we can gain insight into the relationship between
+    genomes. \cmd{odgi viz} applies binding and direct rendering to
+    a raster image to generate a 1D representation able to scale up to
+    gigabase pangenome graphs
+    (Figure~1.A\vphantom{\ref{fig:01}}). However, pangenome graphs may
+    present complex structures to visualize: \cmd{odgi sort} finds
+    the best node order to simplify 1D graph visualization by applying
+    several cutting-edge sorting algorithms. \cmd{odgi drawn}
+    extends the visualization in 2D, taking in input the layouts built
+    by \cmd{odgi layout}.
 
-    Applying \texttt{odgi stats}, users can retrieve metrics describing the graph properties, such as the number of
+    \topic{odgi stats} users can retrieve metrics describing the graph properties, such as the number of
     nodes, edges, paths, and graph length. ODGI also offers more advanced tools for interrogating the graphs.

### 2021-05-31 manuscript/main.tex
Merge introduction stuff where it belongs
Stacked figure
reworked the Abstract a little, explicitly including the Results

@@
 
-    \abstract{ \textbf{Motivation:} Pangenome graphs provide a
-      complete and `lossless' representation of genomic information
-      that allows for complex population-sized analysis of genomes
-      without introducing reference genome bias. Here we present a
-      range of novel tools that makes use of the \odgi\ efficient
-      in-memory representation of DNA/RNA graphs. \odgi\ can handle
-      thousands of large genomes while maintaining fast execution
-      runtime and low memory overhead. \odgi\ includes tools for
-      detecting complex regions, extracting loci, exploratory
-      analysis, sorting, removing artifacts and visualisation.
+    \abstract{
+        \textbf{Motivation:} Pangenome graphs provide a complete representation of
+        genomic information that allows for complex population-sized analysis of genomes
+        without introducing reference genome bias.
         %
         \\
-      \textbf{Availability:} \odgi\ is software written in the \Cpp\
-      programming language and published under the libre MIT license. Source
-      code and user manual can be downloaded from
+        \textbf{Results:} Here we present \odgi\, a novel range of tools that makes use of an efficient
+        in-memory representation of DNA graphs structured in the variation graph model. \odgi\ can
+        handle hundreds of large genomes while maintaining fast execution runtime and low memory overhead.
+        \odgi\ includes tools for detecting complex regions, extracting \textit{loci}, exploratory analysis,
+        sorting, removing artifacts, and graphs manipulation and visualisation.
+        %
+        \\
+        \textbf{Availability:} \odgi\ is software written in the \Cpp\ programming language and
+        published under the libre MIT license. Source code and user manual can be downloaded from
         \href{https://github.com/pangenome/odgi}{https://github.com/pangenome/odgi}.\\
@@
         \href{andreaguarracino@outlook.com}{andreaguarracino@outlook.com},
-        \href{erik.garrison@gmail.com}{erik.garrison@gmail.com}\\ \textbf{Supplementary
-          information:} Supplementary data are available at
-        \textit{Bioinformatics} online.}
+        \href{erik.garrison@gmail.com}{erik.garrison@gmail.com}\\
+        \\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.}
 
@@
 
-%\enlargethispage{12pt}
-
-    \begin{methods}
-        \section{Methods}
+    ODGI is entirely focused on "first-order" operations on the graph
+    (Table~\ref{tab:01}), providing, unlike VG, no method to construct
+    graphs from raw sequences, map reads, or call variants. It
+    supports interrogation of the graph for basic dimensional
+    properties, positional and path-related queries, and visualization
+    in one-dimension (1D) and two-dimensions (2D). ODGI also supports
+    operations on the graph such as subset, subdivide, break, combine,
+    normalize, or order its components, nodes, and paths. Most of the
+    tools are designed to be applied together, piping the output of
+    one tool into the next, thereby preventing the creation of
+    intermediate files, and reducing the number of IO operations.
 
@@
 
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text. Figure~\ref{fig:01}
-%        shows that the above method Text Text Text Text Text Text Text
-%        Text Text Text Text Text.  \citep{Bag01} wants to know about
-%            {\ldots}{\ldots} text follows.
-%        \begin{equation}
-%            \sum \text{\it x}+ \text{\it y} =\text{\it Z}\label{eq:01}\vspace*{-10pt}
-%        \end{equation}
-%
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text
-%        text\vspace*{1pt}
-%
-%        \begin{itemize}
-%            \item for bulleted list, use itemize
-%            \item for bulleted list, use itemize
-%            \item for bulleted list, use itemize\vspace*{1pt}
-%        \end{itemize}
-%
-%        \begin{enumerate}
-%            \item this is item, use enumerate
-%            \item this is item, use enumerate
-%            \item this is item, use enumerate
-%        \end{enumerate}
-%
-%        \subsection{This is subheading}
-%
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%        Figure~2\vphantom{\ref{fig:02}} shows that the above method Text
-%        Text Text Text Text Text Text Text Text Text Text Text.
-%        \citealp{Boffelli03} might want to know about text text text text
-%        Text Text Text Text Text Text Text Text Text Text Text Text Text
-%        Text Text Text Text Text Text Text Text.
-%
-%
-%        \enlargethispage{6pt}
-%
-%        \begin{table}[!t]
-%            \processtable{This is table caption\label{Tab:01}} {\begin{tabular}{@{}llll@{}}
-%                                                                    \toprule head1 &
-%                                                                    head2 & head3 & head4 \\\midrule
-%                                                                    row1 & row1 & row1 & row1 \\
-%                                                                    row2 & row2 & row2 & row2 \\
-%                                                                    row3 & row3 & row3 & row3 \\
-%                                                                    row4 & row4 & row4 & row4 \\\botrule
-%            \end{tabular}}{This is a footnote}
-%        \end{table}
-
-    \end{methods}
-
     \section{Results}
-
-    ODGI is entirely focused on "first-order" operations on the graph (Table~\ref{tab:01}), providing, unlike VG, no
-    method to construct graphs from raw sequences, map reads, or call variants. It supports interrogation of the graph
-    for basic dimensional properties, positional and path-related queries, and visualization in one-dimension (1D) and
-    two-dimensions (2D). ODGI also supports operations on the graph such as subset, subdivide, break, combine,
-    normalize, or order its components, nodes, and paths. Most of the tools are designed to be applied together,
-    piping the output of one tool into the next, thereby preventing the creation of intermediate files, and reducing
-    the number of IO operations.
-
-    \subsection{Features}
 
@@
 
-    \begin{figure}[!tpb]%figure1
-        \fboxsep=0pt\colorbox{gray}{
-            \begin{minipage}[t]{235pt}
-                \vbox to 100pt{
-                    \vfill\hbox to
-                    235pt{\hfill\fontsize{24pt}{24pt}\selectfont FPO\hfill}
-                    \vfill
-                }
-            \end{minipage}}
-%\centerline{\includegraphics{fig01.eps}}
-        \caption{
-            Visualizing the Major Histocompatibility Complex (MHC) \textit{locus} pangenome. A) odgi viz 1D
-            visualization of 8 haploid, phased human genome assemblies plus the chm13 and GRCh38 reference genomes.
-            Graph nodes’ are arranged from left to right. The colored bars represent the linearized renderings of the
-            embedded paths versus the nodes in a binary matrix. The black lines under the paths are the links, which
-            represent the graph topology. B) Bandage~\citep{26099265} representation of the graph topology of a
-            consensus MHC pangenome graph representing only variations bigger than 100 bps. Gene labels were applied
-            exploiting \texttt{odgi position} to liftover gene coordinates from the full graph into the consensus
-            graph.
-        }\label{fig:01}
+    \begin{figure}
+        \begin{minipage}[b]{0.48\textwidth}
+            \includegraphics[width=\linewidth]{fig/mhc-align.eps}
+            \caption{(a) alignment}
+            \label{fig:1}
+        \end{minipage}
+
+        \begin{minipage}[b]{0.48\textwidth}
+            \includegraphics[width=\linewidth]{fig/mhc-pangenome.eps}
+            \caption{(b) Visualizing the Major Histocompatibility Complex
+                (MHC) \textit{locus} pangenome. A) odgi viz 1D visualization
+                of 8 haploid, phased human genome assemblies plus the chm13
+                and GRCh38 reference genomes. Graph nodes’ are arranged
+                from left to right. The colored bars represent the
+                linearized renderings of the embedded paths versus the nodes
+                in a binary matrix. The black lines under the paths are the
+                links, which represent the graph topology. B)
+                Bandage~\citep{26099265} representation of the graph
+                topology of a consensus MHC pangenome graph representing
+                only variations bigger than 100 bps. Gene labels were
+                applied exploiting \texttt{odgi position} to liftover gene
+                coordinates from the full graph into the consensus graph.  }
+
+            \label{fig:2}
+        \end{minipage}
     \end{figure}

### 2021-05-31 manuscript/main.tex
use \odgi\

@@
     HandleGraph API. Several algorithms are shared with the VG toolkit (VG)~\citep{30125266}, the state-of-the-art
-    toolkit for working with pangenome graphs. However, although similar in scope to VG, ODGI importantly differs in
-    its design to support highly complex pangenome graphs. The majority of the ODGI tools are implemented in an
+    toolkit for working with pangenome graphs. However, although similar in scope to VG, \odgi\ importantly differs in
+    its design to support highly complex pangenome graphs. The majority of the \odgi\ tools are implemented in an
     index-free manner, avoiding the burning task needed to create an index structure. Indeed, index-constrained
     implementations would require building indexes also for single or little operations on the graphs. Moreover,
-    thanks to its efficient path representation, ODGI tools can build the internal graph representation (the ODGI
+    thanks to its efficient path representation, \odgi\ tools can build the internal graph representation (the \odgi\
     format) in parallel, which is one of the major bottlenecks when working with very large and deep variation graphs,
@@
 
-    ODGI is entirely focused on "first-order" operations on the graph
+    \odgi\ is entirely focused on "first-order" operations on the graph
     (Table~\ref{tab:01}), providing, unlike VG, no method to construct
@@
     properties, positional and path-related queries, and visualization
-    in one-dimension (1D) and two-dimensions (2D). ODGI also supports
+    in one-dimension (1D) and two-dimensions (2D). \odgi\ also supports
     operations on the graph such as subset, subdivide, break, combine,
@@
 
-    ODGI is designed to build and modify paths in parallel, keeping its representation in memory efficient. It
+    \odgi\ is designed to build and modify paths in parallel, keeping its representation in memory efficient. It
     applies concepts first introduced in the dynamic version of the Graph BWT~\citep{31406990}, a scalable
@@
     \topic{odgi stats} users can retrieve metrics describing the graph properties, such as the number of
-    nodes, edges, paths, and graph length. ODGI also offers more advanced tools for interrogating the graphs.
+    nodes, edges, paths, and graph length. \odgi\ also offers more advanced tools for interrogating the graphs.
     \texttt{odgi bin} summarizes the path information into bins of specified size, enabling a summarized view of
@@
 
-    ODGI allows edit operations on the graphs. Cycles in the graph complicate downstream analyses: \texttt{odgi break}
+    \odgi\ allows edit operations on the graphs. Cycles in the graph complicate downstream analyses: \texttt{odgi break}
     removes the cycles, reducing the complexity of the graph topology. \texttt{odgi groom} removes spurious inverting
@@
     variants would join the components into bigger ones). \texttt{odgi explode} separates the connected components in
-    different ODGI format files, while \texttt{odgi squeeze} allows merging multiple graphs into the same ODGI format
+    different \odgi\ format files, while \texttt{odgi squeeze} allows merging multiple graphs into the same \odgi\ format
     file, preventing node ID collisions. \texttt{odgi extract} allows extracting specific subregions of the graph as
@@
             \caption{(b) Visualizing the Major Histocompatibility Complex
-                (MHC) \textit{locus} pangenome. A) odgi viz 1D visualization
+                (MHC) \textit{locus} pangenome. A) \texttt{odgi viz} 1D visualization
                 of 8 haploid, phased human genome assemblies plus the chm13
@@
 
-    Table~\ref{tab:02} shows a comparison of the operations necessary to extract a subgraph using ODGI and VG. The
+    Table~\ref{tab:02} shows a comparison of the operations necessary to extract a subgraph using \odgi\ and VG. The
     example consists in extracting the centromeres from a pangenome graph made from 88 haploid, phased human genome
@@
     \begin{table}[!t]
-        \processtable{Centromeres extraction comparison using ODGI and VG
+        \processtable{Centromeres extraction comparison using \odgi\ and VG
         \label{tab:02}} {
@@
 
-    To work with ODGI and VG it is first necessary to convert the graph from GFA to ODGI and VG format, respectively.
-    Thanks to the ODGI format's path implementation, \texttt{odgi build} embeds the paths in parallel, outperforming
+    To work with \odgi\ and VG it is first necessary to convert the graph from GFA to \odgi\ and VG format, respectively.
+    Thanks to the \odgi\ format's path implementation, \texttt{odgi build} embeds the paths in parallel, outperforming
     \texttt{vg convert} in the same task. For the same reason, \texttt{odgi extract} results 20 times faster than
@@
     \section{Discussion}
-    We implemented a tool suite, ODGI, which is simple to work with to analyse and manipulate pangenome graphs at a
-    population scale. ODGI's path implementation allows working with graphs of any complexity, without particular
-    limitations. We have demonstrated that ODGI tools efficiently perform with pangenome graphs embedding full human
-    chromosomes. ODGI is not only a tool suite but also a library, already used in pangenome graphs building
+    We implemented a tool suite, \odgi\, which is simple to work with to analyse and manipulate pangenome graphs at a
+    population scale. \odgi\'s path implementation allows working with graphs of any complexity, without particular
+    limitations. We have demonstrated that \odgi\ tools efficiently perform with pangenome graphs embedding full human
+    chromosomes. \odgi\ is not only a tool suite but also a library, already used in pangenome graphs building
     (\citep{pggb}) and normalization (\citep{smoothxg}). Furthermore, users can add new tools, exploiting ODGI as a
-    framework where to implement algorithms working on variation graphs. Therefore, ODGI offers an efficient and
+    framework where to implement algorithms working on variation graphs. Therefore, \odgi\ offers an efficient and
     extendable suite of tools ready for the upcoming challenges in the analyses of hundreds of genomes at a gigabase

### 2021-05-31 manuscript/main.tex
let's try to fit with a Methods section
extended the proposed layout

@@
 
+    \section{Methods}
+
     \odgi\ is designed to build and modify paths in parallel, keeping its representation in memory efficient. It
@@
     stored using a dynamic succinct integer vector that requires \textit{O(5nw)} bits, where \textit{w} is
-    \textit{\textasciitilde log\textsubscript{2}(N)}. Finally, to allow operating on path steps in parallel, each node structure includes a
-    byte-width mutex lock.
+    \textit{\textasciitilde log\textsubscript{2}(N)}. Finally, to allow operating on path steps in parallel, each node
+    structure includes a byte-width mutex lock.
 
@@
 
-    \odgi\ has a wide range of commands. Important features are
+    \odgi\ has a wide range of commands.
 
@@
     represented by a subset of Graphical Fragment Assembly (GFAv1)
-    format\citep{GFAv1}.  \cmd{odgi build} and \cmd{odgi view} allow
-    to convert graphs from GFA to ODGI format and vice versa. In the
+    format \citep{GFAv1}; \cmd{odgi build} and \cmd{odgi view} allow
+    to convert graphs from GFA to \odgi\ format and vice versa. In the
     variation graph model, paths have to respect the graph's topology:
-    this can be verified with \texttt{odgi validate}, to ensure no
+    this can be verified with \cmd{odgi validate}, to ensure no
     errors in the input or edited graphs.
@@
     a raster image to generate a 1D representation able to scale up to
-    gigabase pangenome graphs
-    (Figure~1.A\vphantom{\ref{fig:01}}). However, pangenome graphs may
-    present complex structures to visualize: \cmd{odgi sort} finds
-    the best node order to simplify 1D graph visualization by applying
-    several cutting-edge sorting algorithms. \cmd{odgi drawn}
-    extends the visualization in 2D, taking in input the layouts built
-    by \cmd{odgi layout}.
+    gigabase pangenome graphs (Figure~1\vphantom{\ref{fig:1}}).
+    However, pangenome graphs may present complex structures to
+    visualize: \cmd{odgi sort} finds the best node order to simplify
+    1D graph visualization by applying several cutting-edge sorting
+    algorithms. \cmd{odgi drawn} extends the visualization in 2D,
+    taking in input the layouts built by \cmd{odgi layout}.
 
-    \topic{odgi stats} users can retrieve metrics describing the graph properties, such as the number of
+    \topic{odgi stats, bin, depth \& degree} users can retrieve metrics describing the graph properties, such as the number of
     nodes, edges, paths, and graph length. \odgi\ also offers more advanced tools for interrogating the graphs.
-    \texttt{odgi bin} summarizes the path information into bins of specified size, enabling a summarized view of
-    gigabase scale graphs. \texttt{odgi depth} returns the node depth as defined by query criteria, allowing users to
+    \cmd{odgi bin} summarizes the path information into bins of specified size, enabling a summarized view of
+    gigabase scale graphs. \cmd{odgi depth} returns the node depth as defined by query criteria, allowing users to
     detect the complex regions in the graph due to sequences with highly identical repeats. Complex motifs can be also
-    detected with \texttt{odgi degree}, which returns the node degree as defined by query criteria. High degree nodes
+    detected with \cmd{odgi degree}, which returns the node degree as defined by query criteria. High degree nodes
     are the mirror of misassemblies or problems in the pangenome building, making the tool useful for debugging the
@@
 
-    \odgi\ allows edit operations on the graphs. Cycles in the graph complicate downstream analyses: \texttt{odgi break}
-    removes the cycles, reducing the complexity of the graph topology. \texttt{odgi groom} removes spurious inverting
-    links by exploring the graph from the orientation supported by most paths. To enable efficient sequence alignment
-    against the graph, long nodes can be divided into shorter nodes at a maximum requested size using
-    \texttt{odgi chop}. Partial order alignment, which consists of aligning sequences against a directed acyclic graph
-    (DAG), is frequently used in pangenome building pipelines, but the current implementations return DAGs with 1-bp
-    long nodes; \texttt{odgi unchop} allows joining nodes that can be merged without changing the graph topology, nor
-    the embedded sequences, obtaining an equivalent, but more compact, representation of the graph.
+    \topic{odgi break, groom, chop \& unchop} \odgi\ allows edit operations on the graphs. Cycles in the graph
+    complicate downstream analyses: \cmd{odgi break} removes the cycles, reducing the complexity of the graph topology.
+    \cmd{odgi groom} removes spurious inverting links by exploring the graph from the orientation supported by most
+    paths. To enable efficient sequence alignment against the graph, long nodes can be divided into shorter nodes
+    at a maximum requested size using \cmd{odgi chop}. Partial order alignment, which consists of aligning sequences
+    against a directed acyclic graph (DAG), is frequently used in pangenome building pipelines, but the current
+    implementations return DAGs with 1-bp long nodes; \cmd{odgi unchop} allows joining nodes that can be merged
+    without changing the graph topology, nor the embedded sequences, obtaining an equivalent, but more compact,
+    representation of the graph.
 
-    Pangenome graphs can embed multiple chromosomes as separated connected components (inter-chromosomal structural
-    variants would join the components into bigger ones). \texttt{odgi explode} separates the connected components in
-    different \odgi\ format files, while \texttt{odgi squeeze} allows merging multiple graphs into the same \odgi\ format
-    file, preventing node ID collisions. \texttt{odgi extract} allows extracting specific subregions of the graph as
-    defined by query criteria, thereby simplifying the downstream analyses and reducing the resources to work only
-    with the extracted region.
+    \topic{odgi explode, squeeze \& extract} pangenome graphs can embed multiple chromosomes as separated connected
+    components (inter-chromosomal structural variants would join the components into bigger ones).
+    \cmd{odgi explode} separates the connected components in different \odgi\ format files, while \cmd{odgi squeeze}
+    allows merging multiple graphs into the same \odgi\ format file, preventing node ID collisions. \cmd{odgi extract}
+    allows extracting specific subregions of the graph as defined by query criteria, thereby simplifying the downstream
+    analyses and reducing the resources to work only with the extracted region.
 
-    In variation graphs the coordinates are provided by the embedded path sequences. Indeed, the node IDs are not
-    meant to be stable. \texttt{odgi position} finds, translates, and liftovers graph and path positions between
-    different graphs by exploiting their shared path sequences (Figure~1.B\vphantom{\ref{fig:01}}).
+    \topic{odgi position} in variation graphs the coordinates are provided by the embedded path sequences. Indeed, the node
+    IDs are not meant to be stable. \cmd{odgi position} finds, translates, and liftovers graph and path positions between
+    different graphs by exploiting their shared path sequences (Figure~2\vphantom{\ref{fig:2}}).
 
@@
         \begin{minipage}[b]{0.48\textwidth}
-            \includegraphics[width=\linewidth]{fig/mhc-align.eps}
+            \includegraphics[width=\linewidth]{fig/mhc-align}
             \caption{(a) alignment}
@@
         \begin{minipage}[b]{0.48\textwidth}
-            \includegraphics[width=\linewidth]{fig/mhc-pangenome.eps}
+            \includegraphics[width=\linewidth]{fig/mhc-pangenome}
             \caption{(b) Visualizing the Major Histocompatibility Complex
-                (MHC) \textit{locus} pangenome. A) \texttt{odgi viz} 1D visualization
+                (MHC) \textit{locus} pangenome. A) \cmd{odgi viz} 1D visualization
                 of 8 haploid, phased human genome assemblies plus the chm13
@@
                 only variations bigger than 100 bps. Gene labels were
-                applied exploiting \texttt{odgi position} to liftover gene
+                applied exploiting \cmd{odgi position} to liftover gene
                 coordinates from the full graph into the consensus graph.  }
@@
     To work with \odgi\ and VG it is first necessary to convert the graph from GFA to \odgi\ and VG format, respectively.
-    Thanks to the \odgi\ format's path implementation, \texttt{odgi build} embeds the paths in parallel, outperforming
-    \texttt{vg convert} in the same task. For the same reason, \texttt{odgi extract} results 20 times faster than
-    \texttt{vg chunk} in extracting a portion from the full graph.
+    Thanks to the \odgi\ format's path implementation, \cmd{odgi build} embeds the paths in parallel, outperforming
+    \cmd{vg convert} in the same task. For the same reason, \cmd{odgi extract} results 20 times faster than
+    \cmd{vg chunk} in extracting a portion from the full graph.
 

### 2021-05-31 manuscript/main.tex
filled the table

@@
 
+    %TODO to repeat on a bigger machine to avoid vg crashing
     \begin{table}[!t]
@@
                 \midrule
-                row1               & row1 & row1            & row1        & row1             \\
-                row2               & row2 & row2            & row2        & row2             \\
-                row3               & row3 & row3            & row3        & row3             \\
-                row4               & row4 & row4            & row4        & row4             \\
+                Format conversion                           &       &       &                   \\
+                                    & \cmd{odgi build}      & 1:35  & 10.39 & 5.4               \\
+                                    & \cmd{vg convert}      & 8:14  & 28.43 & 6.1               \\
+                Subgraph extraction                         &       &       &                   \\
+                                    & \cmd{odgi extract}    & 1:15  & 9.43  & 2.7               \\
+                                    & \cmd{vg chunk}        & 21:09 & 59.33 & 1.7               \\
                 \botrule

### 2021-05-31 manuscript/main.tex
replaced short title
fix character

@@
 
-    \title[short Title]{ODGI: scalable tools for pangenome graphs}
+    \title[ODGI: scalable tools for pangenome graphs]{ODGI: scalable tools for pangenome graphs}
     \author[Sample \textit{et~al}.]{
@@
     We implemented a tool suite, \odgi\, which is simple to work with to analyse and manipulate pangenome graphs at a
-    population scale. \odgi\'s path implementation allows working with graphs of any complexity, without particular
+    population scale. ODGI's path implementation allows working with graphs of any complexity, without particular
     limitations. We have demonstrated that \odgi\ tools efficiently perform with pangenome graphs embedding full human

### 2021-05-31 manuscript/figure_p.tex
Figure

--- 2021-05-31 manuscript/main.tex
@@
-\documentclass{bioinfo}
+\begin{figure*}[ht]
+  % \vspace{-8pt}
+   \begin{minipage}[b]{0.48\textwidth}
+  % \begin{center}
 
-\usepackage{relsize} % for Cpp
-\usepackage{xspace} % for Cpp
-
-\newcommand{\odgi}{ODGI}
-\newcommand{\cmd}[1]{{\texttt{#1}}}
-\newcommand{\cmdbf}[1]{{\textbf{#1}}}
-\newcommand{\topic}[1]{{\cmdbf{#1}}:}
-\newcommand{\Rplus}{\protect\hspace{-.1em}\protect\raisebox{.35ex}{\smaller{\smaller\textbf{+}}}}
-\newcommand{\pp}{\Rplus\Rplus}
-\newcommand{\Cpp}{\mbox{C\Rplus\Rplus}\xspace}
-
-\copyrightyear{2021} \pubyear{2021}
-
-\access{Advance Access Publication Date: Day Month Year}
-\appnotes{Manuscript Category}
-
-\begin{document}
-    \firstpage{1}
-
-    \subtitle{Genome Analysis}
-
-    \title[short Title]{ODGI: scalable tools for pangenome graphs}
-    \author[Sample \textit{et~al}.]{
-        Andrea Guarracino\,$^{\text{\sfb 1,}*}$,
-        Simon Heumos\,$^{\text{\sfb 2}}$,
-    % XXXXX,
-        Pjotr~Prins\,$^{\text{\sfb 3}}$ and
-        Erik~Garrison\,$^{\text{\sfb 3,}*}$
+    \subfigure[Aligned MHC pangenome]{
+      \includegraphics[width=\textwidth]{fig/mhc-align.eps}
+      \label{subfig1}
     }
-    \address{$^{\text{\sf 1}}$Department, Institution, City, Post Code, Country \\
-        $^{\text{\sf 2}}$Department, Institution, City, Post Code, Country \\
-        $^{\text{\sf 3}}$Genetics Genomics \& Informatics, UTHSC, Memphis, TN, USA
+    \vspace{-11pt}
+    \newline
+    \subfigure[Consensus MHC pangenome]{
+      \includegraphics[width=\textwidth]{fig/mhc-pangenome.eps}
+      \label{subfig2}
     }
-
-    \corresp{$^\ast$To whom correspondence should be addressed.}
-
-    \history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
-
-    \editor{Associate Editor: XXXXXXX}
-
-    \abstract{
-        \textbf{Motivation:} Pangenome graphs provide a complete representation of
-        genomic information that allows for complex population-sized analysis of genomes
-        without introducing reference genome bias.
-        %
-        \\
-        \textbf{Results:} Here we present \odgi\, a novel range of tools that makes use of an efficient
-        in-memory representation of DNA graphs structured in the variation graph model. \odgi\ can
-        handle hundreds of large genomes while maintaining fast execution runtime and low memory overhead.
-        \odgi\ includes tools for detecting complex regions, extracting \textit{loci}, exploratory analysis,
-        sorting, removing artifacts, and graphs manipulation and visualisation.
-        %
-        \\
-        \textbf{Availability:} \odgi\ is software written in the \Cpp\ programming language and
-        published under the libre MIT license. Source code and user manual can be downloaded from
-        \href{https://github.com/pangenome/odgi}{https://github.com/pangenome/odgi}.\\
-        %
-        \textbf{Contact:}
-        \href{andreaguarracino@outlook.com}{andreaguarracino@outlook.com},
-        \href{erik.garrison@gmail.com}{erik.garrison@gmail.com}\\
-        \\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.}
-
-    \maketitle
-
-    \section{Introduction}
-
-    A pangenome models the full set of genomic elements in a given species or clade~\citep{32453966}. Indeed, these
-    data structures encode the mutual relationships between all the genomes represented, in contrast to reference-based
-    approaches which relate sequences only to a particular genome, chosen as reference. A class of methods to represent
-    pangenomes involves the \textit{sequence graphs}~\citep{2488477}, where similar orthologous regions between genomes
-    collapse into a single representative sequence. In \textit{node-labeled} sequence graphs, nodes indicate DNA
-    sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph. A
-    \textit{bidirected} sequence graph represents both strands of DNA. On this model, \textit{variation graphs} add the
-    concept of \textit{path} to embed linear sequences (e.g., genomes or haplotypes) into the graph~\citep{30125266}.
-    Paths provide a stable coordinate system, allowing graph annotations and interoperability between different graphs.
-
-    Thanks to advances in sequencing technologies, new genome assemblies are produced at a high rate. This amount of
-    data offers the opportunity to study genomic variation as never before, but it also brings challenges in how to
-    represent and manipulate hundreds of genomes at a gigabase scale. Variation graphs compactly represent the full
-    genetic variation across a population, but graph-based data structures present computational overheads: in addition
-    to the sequences, they must also represent the graph topology. Several genome data structures were developed to
-    address these challenges~\citep{33040146}. Memory-efficient representations do not allow dynamic updates to the
-    graph, such as adding new genomes or simplifying the graph topology. Furthermore, dynamic updates must perform
-    efficiently on the complex graphs generated from whole-genome alignment of vertebrates. Highly repetitive regions,
-    such as centromeres, segmental duplications, and acrocentric chromosomes lead to graphs which present paths with
-    very high-depth nodes. The node depth is defined as the number of times in which the node is crossed by all the
-    paths embedded in the graph. High depth increases the complexity of the operations performed on the paths and the
-    nodes they cross.
-
-    To overcome this limitation, we have developed an Optimized Dynamic Genome/Graph Implementation (ODGI). Each tool
-    loads graphs into the succinct dynamic HandleGraph model~\citep{33040146} and applies methods based on the
-    HandleGraph API. Several algorithms are shared with the VG toolkit (VG)~\citep{30125266}, the state-of-the-art
-    toolkit for working with pangenome graphs. However, although similar in scope to VG, \odgi\ importantly differs in
-    its design to support highly complex pangenome graphs. The majority of the \odgi\ tools are implemented in an
-    index-free manner, avoiding the burning task needed to create an index structure. Indeed, index-constrained
-    implementations would require building indexes also for single or little operations on the graphs. Moreover,
-    thanks to its efficient path representation, \odgi\ tools can build the internal graph representation (the \odgi\
-    format) in parallel, which is one of the major bottlenecks when working with very large and deep variation graphs,
-    efficiently scaling up to hundreds of large vertebrate genomes.
-
-    \odgi\ is entirely focused on "first-order" operations on the graph
-    (Table~\ref{tab:01}), providing, unlike VG, no method to construct
-    graphs from raw sequences, map reads, or call variants. It
-    supports interrogation of the graph for basic dimensional
-    properties, positional and path-related queries, and visualization
-    in one-dimension (1D) and two-dimensions (2D). \odgi\ also supports
-    operations on the graph such as subset, subdivide, break, combine,
-    normalize, or order its components, nodes, and paths. Most of the
-    tools are designed to be applied together, piping the output of
-    one tool into the next, thereby preventing the creation of
-    intermediate files, and reducing the number of IO operations.
-
-    \section{Methods}
-
-    \odgi\ is designed to build and modify paths in parallel, keeping its representation in memory efficient. It
-    applies concepts first introduced in the dynamic version of the Graph BWT~\citep{31406990}, a scalable
-    implementation of the graph extension of the positional Burrows-Wheeler transform~\citep{28702075}. Specifically,
-    we build a \textit{node-local} model of the graph, where the graph is stored in a vector of node structures.
-    The identifier of each node is its offset in this vector plus 1. Each structure contains the sequence of the
-    node, its edges in both directions (forward and backward), and a vector of \textit{path steps} that describes
-    how the embedded paths walk across the node. Each step records the path identifier and the previous and next
-    path steps on the same path. To reduce the encoding, we encode path steps using a local alphabet that maps the
-    \textit{N} neighbors of the node into the range \textit{1..N}. To further save space, \textit{node deltas},
-    rather than IDs, are stored in this alphabet. The delta between two nodes is defined as their distance in the
-    graph vector (i.e., the difference between the node offsets). The path step vector, of length \textit{n}, is
-    stored using a dynamic succinct integer vector that requires \textit{O(5nw)} bits, where \textit{w} is
-    \textit{\textasciitilde log\textsubscript{2}(N)}. Finally, to allow operating on path steps in parallel, each node
-    structure includes a byte-width mutex lock.
-
-    \section{Results}
-
-    \odgi\ has a wide range of commands.
-
-    \topic{odgi build, view \& validate} pangenome graphs are commonly
-    represented by a subset of Graphical Fragment Assembly (GFAv1)
-    format \citep{GFAv1}; \cmd{odgi build} and \cmd{odgi view} allow
-    to convert graphs from GFA to \odgi\ format and vice versa. In the
-    variation graph model, paths have to respect the graph's topology:
-    this can be verified with \cmd{odgi validate}, to ensure no
-    errors in the input or edited graphs.
-
-    \topic{odgi viz, sort, drawn \& layout} by visualizing pangenome
-    graphs, we can gain insight into the relationship between
-    genomes. \cmd{odgi viz} applies binding and direct rendering to
-    a raster image to generate a 1D representation able to scale up to
-    gigabase pangenome graphs (Figure~1\vphantom{\ref{fig:1}}).
-    However, pangenome graphs may present complex structures to
-    visualize: \cmd{odgi sort} finds the best node order to simplify
-    1D graph visualization by applying several cutting-edge sorting
-    algorithms. \cmd{odgi drawn} extends the visualization in 2D,
-    taking in input the layouts built by \cmd{odgi layout}.
-
-    \topic{odgi stats, bin, depth \& degree} users can retrieve metrics describing the graph properties, such as the number of
-    nodes, edges, paths, and graph length. \odgi\ also offers more advanced tools for interrogating the graphs.
-    \cmd{odgi bin} summarizes the path information into bins of specified size, enabling a summarized view of
-    gigabase scale graphs. \cmd{odgi depth} returns the node depth as defined by query criteria, allowing users to
-    detect the complex regions in the graph due to sequences with highly identical repeats. Complex motifs can be also
-    detected with \cmd{odgi degree}, which returns the node degree as defined by query criteria. High degree nodes
-    are the mirror of misassemblies or problems in the pangenome building, making the tool useful for debugging the
-    process as well.
-
-    \topic{odgi break, groom, chop \& unchop} \odgi\ allows edit operations on the graphs. Cycles in the graph
-    complicate downstream analyses: \cmd{odgi break} removes the cycles, reducing the complexity of the graph topology.
-    \cmd{odgi groom} removes spurious inverting links by exploring the graph from the orientation supported by most
-    paths. To enable efficient sequence alignment against the graph, long nodes can be divided into shorter nodes
-    at a maximum requested size using \cmd{odgi chop}. Partial order alignment, which consists of aligning sequences
-    against a directed acyclic graph (DAG), is frequently used in pangenome building pipelines, but the current
-    implementations return DAGs with 1-bp long nodes; \cmd{odgi unchop} allows joining nodes that can be merged
-    without changing the graph topology, nor the embedded sequences, obtaining an equivalent, but more compact,
-    representation of the graph.
-
-    \topic{odgi explode, squeeze \& extract} pangenome graphs can embed multiple chromosomes as separated connected
-    components (inter-chromosomal structural variants would join the components into bigger ones).
-    \cmd{odgi explode} separates the connected components in different \odgi\ format files, while \cmd{odgi squeeze}
-    allows merging multiple graphs into the same \odgi\ format file, preventing node ID collisions. \cmd{odgi extract}
-    allows extracting specific subregions of the graph as defined by query criteria, thereby simplifying the downstream
-    analyses and reducing the resources to work only with the extracted region.
-
-    \topic{odgi position} in variation graphs the coordinates are provided by the embedded path sequences. Indeed, the node
-    IDs are not meant to be stable. \cmd{odgi position} finds, translates, and liftovers graph and path positions between
-    different graphs by exploiting their shared path sequences (Figure~2\vphantom{\ref{fig:2}}).
-
-    \begin{figure}
-        \begin{minipage}[b]{0.48\textwidth}
-            \includegraphics[width=\linewidth]{fig/mhc-align}
-            \caption{(a) alignment}
-            \label{fig:1}
-        \end{minipage}
-
-        \begin{minipage}[b]{0.48\textwidth}
-            \includegraphics[width=\linewidth]{fig/mhc-pangenome}
             \caption{(b) Visualizing the Major Histocompatibility Complex
@@
 
-            \label{fig:2}
+  % \end{center}
   \end{minipage}
-    \end{figure}
-
-    \subsection{Performance on human data}
-
-    Table~\ref{tab:02} shows a comparison of the operations necessary to extract a subgraph using \odgi\ and VG. The
-    example consists in extracting the centromeres from a pangenome graph made from 88 haploid, phased human genome
-    assemblies from 44 individuals from the Human Pangenome Reference Consortium year 1 assembly (see Data
-    Availability), plus the chm13 and GRCh38 reference genomes, for a total of 90 haplotypes. All commands are provided
-    in the Supplementary Material.
-
-    \begin{table}[!t]
-        \processtable{Centromeres extraction comparison using \odgi\ and VG
-        \label{tab:02}} {
-            \begin{tabular}{@{}lllll@{}}
-                \toprule Operation & Tool & Runtime (mm:ss) & Memory (GB) & Output size (GB) \\
-                \midrule
-                row1               & row1 & row1            & row1        & row1             \\
-                row2               & row2 & row2            & row2        & row2             \\
-                row3               & row3 & row3            & row3        & row3             \\
-                row4               & row4 & row4            & row4        & row4             \\
-                \botrule
-            \end{tabular}}
-
-    \end{table}
-
-    To work with \odgi\ and VG it is first necessary to convert the graph from GFA to \odgi\ and VG format, respectively.
-    Thanks to the \odgi\ format's path implementation, \cmd{odgi build} embeds the paths in parallel, outperforming
-    \cmd{vg convert} in the same task. For the same reason, \cmd{odgi extract} results 20 times faster than
-    \cmd{vg chunk} in extracting a portion from the full graph.
-
-    \section{Discussion}
-    We implemented a tool suite, \odgi\, which is simple to work with to analyse and manipulate pangenome graphs at a
-    population scale. \odgi\'s path implementation allows working with graphs of any complexity, without particular
-    limitations. We have demonstrated that \odgi\ tools efficiently perform with pangenome graphs embedding full human
-    chromosomes. \odgi\ is not only a tool suite but also a library, already used in pangenome graphs building
-    (\citep{pggb}) and normalization (\citep{smoothxg}). Furthermore, users can add new tools, exploiting ODGI as a
-    framework where to implement algorithms working on variation graphs. Therefore, \odgi\ offers an efficient and
-    extendable suite of tools ready for the upcoming challenges in the analyses of hundreds of genomes at a gigabase
-    scale that characterize the pangenomic era.
-
-%\begin{figure}[!tpb]%figure2
-%%\centerline{\includegraphics{fig02.eps}}
-%\caption{Caption, caption.}\label{fig:02}
-%\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%     please remove the " % " symbol from \centerline{\includegraphics{fig01.eps}}
-%     as it may ignore the figures.
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-    % \vspace*{+12pt}
-    % \textit{Financial Support}: none declared.
-
-    \textit{Conflict of Interest}: none declared.
-    \vspace*{+12pt}
-
-    % \section*{Data availability}
-
-    % Data used to build Human pangenome graphs is available at \href{https://github.com/human-pangenomics/HPP_Year1_Data_Freeze_v1.0}{https://github.com/human-pangenomics/HPP\_Year1\_Data\_Freeze\_v1.0}.
-
-    % \section*{Acknowledgements}
-    %ToDo
-    % XXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX.
-
-    % \section*{Funding}
-    %ToDo
-    % This work has been supported by the XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX.
-
-%    \vspace*{-12pt}
-
-%\bibliographystyle{natbib}
-%\bibliographystyle{achemnat}
-%\bibliographystyle{plainnat}
-%\bibliographystyle{abbrv}
-%\bibliographystyle{bioinformatics}
-%
-%\bibliographystyle{plain}
-%
-%\bibliography{Document}
-
-    %ToDo
-    \begin{thebibliography}{}
-
-        \bibitem[Bofelli {\it et~al}., 2000]{Boffelli03}
-        Bofelli,F., Name2, Name3 (2003) Article title, {\it Journal Name}, {\bf 199}, 133-154.
-
-        \bibitem[Bag {\it et~al}., 2001]{Bag01}
-        Bag,M., Name2, Name3 (2001) Article title, {\it Journal Name}, {\bf 99}, 33-54.
-
-        \bibitem[Yoo \textit{et~al}., 2003]{Yoo03}
-        Yoo,M.S. \textit{et~al}. (2003) Oxidative stress regulated genes
-        in nigral dopaminergic neurnol cell: correlation with the known
-        pathology in Parkinson's disease. \textit{Brain Res. Mol. Brain
-        Res.}, \textbf{110}(Suppl. 1), 76--84.
-
-        \bibitem[Lehmann, 1986]{Leh86}
-        Lehmann,E.L. (1986) Chapter title. \textit{Book Title}. Vol.~1, 2nd edn. Springer-Verlag, New York.
-
-        \bibitem[Crenshaw and Jones, 2003]{Cre03}
-        Crenshaw, B.,III, and Jones, W.B.,Jr (2003) The future of clinical
-        cancer management: one tumor, one chip. \textit{Bioinformatics},
-        doi:10.1093/bioinformatics/btn000.
-
-        \bibitem[Auhtor \textit{et~al}. (2000)]{Aut00}
-        Auhtor,A.B. \textit{et~al}. (2000) Chapter title. In Smith, A.C.
-        (ed.), \textit{Book Title}, 2nd edn. Publisher, Location, Vol. 1, pp.
-        ???--???.
-
-        \bibitem[Bardet, 1920]{Bar20}
-        Bardet, G. (1920) Sur un syndrome d'obesite infantile avec
-        polydactylie et retinite pigmentaire (contribution a l'etude des
-        formes cliniques de l'obesite hypophysaire). PhD Thesis, name of
-        institution, Paris, France.
-
-    \end{thebibliography}
-\end{document}
+\end{figure*}

### 2021-05-31 manuscript/main.tex
Figure

@@
 \usepackage{xspace} % for Cpp
+\usepackage{subfigure}
 
@@
 
-    \begin{figure}
-        \begin{minipage}[b]{0.48\textwidth}
-            \includegraphics[width=\linewidth]{fig/mhc-align}
-            \caption{(a) alignment}
-            \label{fig:1}
-        \end{minipage}
-
-        \begin{minipage}[b]{0.48\textwidth}
-            \includegraphics[width=\linewidth]{fig/mhc-pangenome}
-            \caption{(b) Visualizing the Major Histocompatibility Complex
-                (MHC) \textit{locus} pangenome. A) \cmd{odgi viz} 1D visualization
-                of 8 haploid, phased human genome assemblies plus the chm13
-                and GRCh38 reference genomes. Graph nodes’ are arranged
-                from left to right. The colored bars represent the
-                linearized renderings of the embedded paths versus the nodes
-                in a binary matrix. The black lines under the paths are the
-                links, which represent the graph topology. B)
-                Bandage~\citep{26099265} representation of the graph
-                topology of a consensus MHC pangenome graph representing
-                only variations bigger than 100 bps. Gene labels were
-                applied exploiting \cmd{odgi position} to liftover gene
-                coordinates from the full graph into the consensus graph.  }
-
-            \label{fig:2}
-        \end{minipage}
-    \end{figure}
+    \input{figure_p}
 

### 2021-05-31 manuscript/figure_p.tex
Reorganise figures
Caption

@@
 \begin{figure*}[ht]
-  % \vspace{-8pt}
-   \begin{minipage}[b]{0.48\textwidth}
+  \vspace{-10pt}
+   \begin{minipage}[b]{\textwidth}
   % \begin{center}
 
-    \subfigure[Aligned MHC pangenome]{
-      \includegraphics[width=\textwidth]{fig/mhc-align.eps}
+    \subfigure[aligned MHC pangenome]{
+      \includegraphics[width=.48\textwidth]{fig/mhc-align.eps}
       \label{subfig1}
@@
     \vspace{-11pt}
-    \newline
-    \subfigure[Consensus MHC pangenome]{
-      \includegraphics[width=\textwidth]{fig/mhc-pangenome.eps}
+    % \newline
+    \subfigure[consensus MHC pangenome]{
+      \includegraphics[width=.48\textwidth]{fig/mhc-pangenome.eps}
       \label{subfig2}
     }
-            \caption{(b) Visualizing the Major Histocompatibility Complex
-                (MHC) \textit{locus} pangenome. A) \cmd{odgi viz} 1D visualization
-                of 8 haploid, phased human genome assemblies plus the chm13
-                and GRCh38 reference genomes. Graph nodes’ are arranged
-                from left to right. The colored bars represent the
-                linearized renderings of the embedded paths versus the nodes
-                in a binary matrix. The black lines under the paths are the
-                links, which represent the graph topology. B)
-                Bandage~\citep{26099265} representation of the graph
-                topology of a consensus MHC pangenome graph representing
-                only variations bigger than 100 bps. Gene labels were
-                applied exploiting \cmd{odgi position} to liftover gene
-                coordinates from the full graph into the consensus graph.  }
+    \caption{Examples of output generated by \odgi\ visualizing the
+      Major Histocompatibility Complex (MHC) \textit{locus}. (a)
+      \cmd{odgi viz} projection of full MHC of eight haploid phased
+      human genome assemblies, plus the chm13 cell line and GRCh38
+      reference genomes as supplied by the human reference pangenome
+      consortium\citep{RPC}. The coloured bars represent the
+      linearized paths --- representing individuals --- as a zoomed
+      out multi-sequence alignment. The black lines at the bottom
+      represent the graph topology. (b) consensus representation of
+      the graph of the same MHC assemblies showing variations larger
+      than $100$ base pairs. The gene label coordinates were super
+      imposed by \cmd{odgi position} from GRCh38. Figure genrated by
+      the Bandage tool\citep{26099265} }
 

### 2021-05-31 manuscript/main.tex
Reorganise figures
fit table
1st_topic

@@
     \maketitle
+
+    \input{figure_p}
 
@@
 
-    \topic{odgi build, view \& validate} pangenome graphs are commonly
-    represented by a subset of Graphical Fragment Assembly (GFAv1)
-    format \citep{GFAv1}; \cmd{odgi build} and \cmd{odgi view} allow
-    to convert graphs from GFA to \odgi\ format and vice versa. In the
-    variation graph model, paths have to respect the graph's topology:
-    this can be verified with \cmd{odgi validate}, to ensure no
-    errors in the input or edited graphs.
+    \topic{odgi build, view \& validate} pangenome graphs are
+    represented by a subset of the textual Graphical Fragment Assembly
+    (GFAv1) format\citep{GFAv1}; \cmd{odgi build} and \cmd{odgi view}
+    convert graphs from GFA to the \odgi\ format and vice versa.
+    \cmd{odgi validate} ascertains the graph data is correct, such as
+    validating that paths respect the graph's topology.
 
@@
 
-    \input{figure_p}
-
     \subsection{Performance on human data}
@@
           \label{tab:02}} {
+          \resizebox{.99\columnwidth}{!}{
             \begin{tabular}{@{}lllll@{}}
@@
                 \botrule
-            \end{tabular}}
+            \end{tabular}}}
 

### 2021-05-31 manuscript/main.tex
headers
2nd topic

@@
 
-    \subtitle{Genome Analysis}
+    \subtitle{Pangenome analysis}
 
     \title[ODGI: scalable tools for pangenome graphs]{ODGI: scalable tools for pangenome graphs}
-    \author[Sample \textit{et~al}.]{
+    \author[Guarracino, Heumos \textit{et~al}.]{
         Andrea Guarracino\,$^{\text{\sfb 1,}*}$,
@@
 
-    \topic{odgi viz, sort, drawn \& layout} by visualizing pangenome
-    graphs, we can gain insight into the relationship between
-    genomes. \cmd{odgi viz} applies binding and direct rendering to
-    a raster image to generate a 1D representation able to scale up to
-    gigabase pangenome graphs (Figure~1\vphantom{\ref{fig:1}}).
-    However, pangenome graphs may present complex structures to
-    visualize: \cmd{odgi sort} finds the best node order to simplify
-    1D graph visualization by applying several cutting-edge sorting
-    algorithms. \cmd{odgi drawn} extends the visualization in 2D,
+    \topic{odgi viz, sort, drawn \& layout} pangenome visualization is important to gain insight
+    into the relationship between the embedded genomes. \cmd{odgi viz} generates a 1D representation
+    able to scale up to gigabase pangenome graphs (Figure~1\vphantom{\ref{fig:1}}).
+    \cmd{odgi sort} finds the best node order to simplify the graph visualization by applying several
+    cutting-edge sorting algorithms. \cmd{odgi drawn} extends the visualization in 2D,
     taking in input the layouts built by \cmd{odgi layout}.

### 2021-05-31 manuscript/main.tex
topic_2

@@
 
-    \topic{odgi viz, sort, drawn \& layout} pangenome visualization is important to gain insight
-    into the relationship between the embedded genomes. \cmd{odgi viz} generates a 1D representation
-    able to scale up to gigabase pangenome graphs (Figure~1\vphantom{\ref{fig:1}}).
-    \cmd{odgi sort} finds the best node order to simplify the graph visualization by applying several
-    cutting-edge sorting algorithms. \cmd{odgi drawn} extends the visualization in 2D,
-    taking in input the layouts built by \cmd{odgi layout}.
+    \topic{odgi viz, sort, drawn \& layout} pangenome visualization
+    provides convenient insight into genomic variation (see
+    Fig.~\ref{fig:1}). \cmd{odgi viz} generates a linearized
+    representation of the pangenome and is capable of handling full
+    length human chromosomes (Fig.~\ref{fig:1}a).  \cmd{odgi sort}
+    reorganises node order and simplifies the graph by applying
+    several algorithms, including the novel `NAME the
+    algorithm'. \cmd{odgi layout \& drawn} bring out the consensus
+    graph (Fig.~\ref{fig:1}b).
 

### 2021-05-31 manuscript/main.tex
draw in the present

@@
 
-    \topic{odgi viz, sort, drawn \& layout} pangenome visualization
+    \topic{odgi viz, sort, draw \& layout} pangenome visualization
     provides convenient insight into genomic variation (see
@@
     several algorithms, including the novel `NAME the
-    algorithm'. \cmd{odgi layout \& drawn} bring out the consensus
+    algorithm'. \cmd{odgi layout \& draw} bring out the consensus
     graph (Fig.~\ref{fig:1}b).

### 2021-05-31 manuscript/main.tex
fixed odgi layout and draw
added hot sorting algorithm name

@@
     reorganises node order and simplifies the graph by applying
-    several algorithms, including the novel `NAME the
-    algorithm'. \cmd{odgi layout \& draw} bring out the consensus
-    graph (Fig.~\ref{fig:1}b).
+    several algorithms, including the novel `Path-Guided Stochastic Gradient Descent`
+    algorithm. \cmd{odgi layout} and \cmd{odgi draw} bring the
+    visualization in 2D. (Fig.~\ref{fig:1}b).
 

### 2021-05-31 manuscript/main.tex
topic 3rd

@@
 
-    \topic{odgi stats, bin, depth \& degree} users can retrieve metrics describing the graph properties, such as the number of
-    nodes, edges, paths, and graph length. \odgi\ also offers more advanced tools for interrogating the graphs.
+    \topic{odgi stats, bin, depth \& degree} pangenome statistics allow exploring genome complexity at different
+    levels of abstraction. \cmd{odgi stats} returns the number of nodes, edges, paths, and graph length, while
     \cmd{odgi bin} summarizes the path information into bins of specified size, enabling a summarized view of
-    gigabase scale graphs. \cmd{odgi depth} returns the node depth as defined by query criteria, allowing users to
-    detect the complex regions in the graph due to sequences with highly identical repeats. Complex motifs can be also
-    detected with \cmd{odgi degree}, which returns the node degree as defined by query criteria. High degree nodes
-    are the mirror of misassemblies or problems in the pangenome building, making the tool useful for debugging the
-    process as well.
+    gigabase scale graphs. \cmd{odgi depth} and \cmd{odgi degree} return the node depth and degree, respectively, as
+    defined by query criteria, allowing users to detect complex regions due to sequences with highly identical
+    repeats, or problems in the pangenome graph.
 

### 2021-05-31 manuscript/main.tex
Edits

@@
     \topic{odgi viz, sort, draw \& layout} pangenome visualization
-    provides convenient insight into genomic variation (see
-    Fig.~\ref{fig:1}). \cmd{odgi viz} generates a linearized
-    representation of the pangenome and is capable of handling full
-    length human chromosomes (Fig.~\ref{fig:1}a). \cmd{odgi sort}
-    reorganises node order and simplifies the graph by applying
-    several algorithms, including the novel `Path-Guided Stochastic Gradient Descent`
-    algorithm. \cmd{odgi layout} and \cmd{odgi draw} bring the
-    visualization in 2D. (Fig.~\ref{fig:1}b).
+    provides convenient insight into genomic variation. \cmd{odgi viz}
+    generates a linearized representation of the pangenome and is
+    capable of handling full length human chromosomes
+    (Fig.~\ref{fig:1}a). \cmd{odgi sort} reorganises node order and
+    simplifies the graph by applying several algorithms, including the
+    novel `Path-Guided Stochastic Gradient Descent' algorithm.
+    \cmd{odgi layout} generates a file with the X-Y coordinates for
+    all nodes and \cmd{odgi draw} draws the actual layout of the graph
+    (Fig.~\ref{fig:1}b).
 
-    \topic{odgi stats, bin, depth \& degree} pangenome statistics allow exploring genome complexity at different
-    levels of abstraction. \cmd{odgi stats} returns the number of nodes, edges, paths, and graph length, while
-    \cmd{odgi bin} summarizes the path information into bins of specified size, enabling a summarized view of
-    gigabase scale graphs. \cmd{odgi depth} and \cmd{odgi degree} return the node depth and degree, respectively, as
-    defined by query criteria, allowing users to detect complex regions due to sequences with highly identical
-    repeats, or problems in the pangenome graph.
+    \topic{odgi stats, bin, depth \& degree} pangenome numerical
+    counts provide insight into genome complexity in other
+    ways. E.g., \cmd{odgi stats} returns the number of nodes,
+    edges, paths, and graph length. \cmd{odgi bin} summarizes path
+    information into bins of a specified size, thereby enabling a
+    summarized view of gigabase scale graphs. \cmd{odgi depth} and
+    \cmd{odgi degree} return the node depth and degree as defined by
+    user provided criteria. These methods allow detection of complex
+    genomic regions.
 

### 2021-05-31 manuscript/figure_p.tex
topic_2
caption includes tools

@@
       \includegraphics[width=.48\textwidth]{fig/mhc-align.eps}
-      \label{subfig1}
+      \label{fig:1a}
     }
@@
       \includegraphics[width=.48\textwidth]{fig/mhc-pangenome.eps}
-      \label{subfig2}
+      \label{fig:1b}
     }
@@
       Major Histocompatibility Complex (MHC) \textit{locus}. (a)
-      \cmd{odgi viz} projection of full MHC of eight haploid phased
-      human genome assemblies, plus the chm13 cell line and GRCh38
-      reference genomes as supplied by the human reference pangenome
-      consortium\citep{RPC}. The coloured bars represent the
-      linearized paths --- representing individuals --- as a zoomed
-      out multi-sequence alignment. The black lines at the bottom
-      represent the graph topology. (b) consensus representation of
-      the graph of the same MHC assemblies showing variations larger
-      than $100$ base pairs. The gene label coordinates were super
-      imposed by \cmd{odgi position} from GRCh38. Figure genrated by
-      the Bandage tool\citep{26099265} }
-
+      projection of full MHC of eight haploid phased human genome
+      assemblies, plus the chm13 cell line and GRCh38 reference
+      genomes as supplied by the human reference pangenome
+      consortium\citep{RPC} by \cmd{odgi build+sort+viz}. The coloured
+      bars represent the linearized paths --- representing individuals
+      --- as a zoomed out multi-sequence alignment. The black lines at
+      the bottom represent the graph topology. (b) consensus
+      representation of the graph of the same MHC assemblies showing
+      variations larger than $100$ base pairs created by \cmd{odgi
+        build}. The gene label coordinates were super imposed by
+      \cmd{odgi position} from GRCh38. Figure genrated by the Bandage
+      tool\citep{26099265} }
+    \label{fig:1}
   % \end{center}

### 2021-05-31 manuscript/main.tex
4th topic

@@
 
-    \topic{odgi break, groom, chop \& unchop} \odgi\ allows edit operations on the graphs. Cycles in the graph
-    complicate downstream analyses: \cmd{odgi break} removes the cycles, reducing the complexity of the graph topology.
-    \cmd{odgi groom} removes spurious inverting links by exploring the graph from the orientation supported by most
-    paths. To enable efficient sequence alignment against the graph, long nodes can be divided into shorter nodes
-    at a maximum requested size using \cmd{odgi chop}. Partial order alignment, which consists of aligning sequences
-    against a directed acyclic graph (DAG), is frequently used in pangenome building pipelines, but the current
-    implementations return DAGs with 1-bp long nodes; \cmd{odgi unchop} allows joining nodes that can be merged
-    without changing the graph topology, nor the embedded sequences, obtaining an equivalent, but more compact,
-    representation of the graph.
+    \topic{odgi break, groom, chop \& unchop} \odgi\ pangenome graphs are dynamic structures in \odgi\.
+    \cmd{odgi break} removes cycles in the graph, reducing its topology complexity. \cmd{odgi groom}
+    removes spurious inverting links by exploring the graph from the orientation supported by most paths. \cmd{odgi chop}
+    divides long nodes into shorter ones at a maximum requested size, simplifying downstream analysis.
+    \cmd{odgi unchop} allows joining nodes that can be merged without changing the graph topology, nor the
+    embedded sequences, obtaining an equivalent, but more compact, representation of the graph.
 

### 2021-05-31 manuscript/main.tex
5th topic

@@
     \topic{odgi explode, squeeze \& extract} pangenome graphs can embed multiple chromosomes as separated connected
-    components (inter-chromosomal structural variants would join the components into bigger ones).
-    \cmd{odgi explode} separates the connected components in different \odgi\ format files, while \cmd{odgi squeeze}
-    allows merging multiple graphs into the same \odgi\ format file, preventing node ID collisions. \cmd{odgi extract}
-    allows extracting specific subregions of the graph as defined by query criteria, thereby simplifying the downstream
-    analyses and reducing the resources to work only with the extracted region.
+    components. \cmd{odgi explode} separates them in different files, while \cmd{odgi squeeze} merges
+    multiple graphs into the same file, preventing node ID collisions. \cmd{odgi extract} extracts regions of the graph
+    as defined by query criteria, thereby reducing the resources to work only with the subgraph.
 

### 2021-05-31 manuscript/main.tex
6th topic
little fix to compile

@@
 
-    \topic{odgi break, groom, chop \& unchop} \odgi\ pangenome graphs are dynamic structures in \odgi\.
+    \topic{odgi break, groom, chop \& unchop} \odgi\ pangenome graphs are dynamic structures in ODGI.
     \cmd{odgi break} removes cycles in the graph, reducing its topology complexity. \cmd{odgi groom}
@@
 
-    \topic{odgi position} in variation graphs the coordinates are provided by the embedded path sequences. Indeed, the node
-    IDs are not meant to be stable. \cmd{odgi position} finds, translates, and liftovers graph and path positions between
-    different graphs by exploiting their shared path sequences (Figure~2\vphantom{\ref{fig:2}}).
+    \topic{odgi position} embedded paths provide stable coordinates regardless of the graph in which they are.
+    \cmd{odgi position} finds, translates, and liftovers graph and path positions between different graphs by
+    exploiting their shared path sequences (Figure~2\vphantom{\ref{fig:2}}).
 

### 2021-05-31 manuscript/figure_p.tex
reworked figure caption

@@
         }
-    \caption{Examples of output generated by \odgi\ visualizing the
-      Major Histocompatibility Complex (MHC) \textit{locus}. (a)
-      projection of full MHC of eight haploid phased human genome
-      assemblies, plus the chm13 cell line and GRCh38 reference
-      genomes as supplied by the human reference pangenome
-      consortium\citep{RPC} by \cmd{odgi build+sort+viz}. The coloured
-      bars represent the linearized paths --- representing individuals
-      --- as a zoomed out multi-sequence alignment. The black lines at
-      the bottom represent the graph topology. (b) consensus
-      representation of the graph of the same MHC assemblies showing
-      variations larger than $100$ base pairs created by \cmd{odgi
-        build}. The gene label coordinates were super imposed by
-      \cmd{odgi position} from GRCh38. Figure genrated by the Bandage
-      tool\citep{26099265} }
+        \caption{Visualizing the Major Histocompatibility Complex (MHC) \textit{locus}.
+            (a) projection of full MHC \textit{locus} of eight haploid phased human genome
+            assemblies, plus the chm13 cell line and GRCh38 reference genomes, as supplied by the
+            Human Reference Pangenome Consortium\citep{HRPC}, created by \cmd{odgi build+sort+viz}. The coloured
+            bars represent the linearized paths, representing contigs, as a zoomed out multi-sequence
+            alignment. The black lines at the bottom represent the graph topology.
+            (b) consensus graph representation of the same assemblies showing variations larger than $100$ base pairs.
+            The gene label coordinates were super imposed by using \cmd{odgi position} from GRCh38.
+            Figure generated by the Bandage tool\citep{26099265}.
+        }
         \label{fig:1}

### 2021-05-31 manuscript/main.tex
authors things

@@
     \author[Guarracino, Heumos \textit{et~al}.]{
-        Andrea Guarracino\,$^{\text{\sfb 1,}*}$,
-        Simon Heumos\,$^{\text{\sfb 2}}$,
+        Andrea Guarracino\,$^{\text{\sfb 1,}\dagger,*}$,
+        Simon Heumos\,$^{\text{\sfb 2,}\dagger}$,
     % XXXXX,
@@
     }
-    \address{$^{\text{\sf 1}}$Department, Institution, City, Post Code, Country \\
-        $^{\text{\sf 2}}$Department, Institution, City, Post Code, Country \\
+    \address{$^{\text{\sf 1}}$Biology, University of Tor Vergata, Rome, 00133, Italy \\
+        $^{\text{\sf 2}}$QBiC, University of Tübingen, Tübingen, XXXXXXXXXX, Germany \\
         $^{\text{\sf 3}}$Genetics Genomics \& Informatics, UTHSC, Memphis, TN, USA
@@
     \corresp{$^\ast$To whom correspondence should be addressed.}
+    \corresp{$^\dagger$Shared first authorship.}
 
@@
         \href{erik.garrison@gmail.com}{erik.garrison@gmail.com}\\
-        \\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.}
+        \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.}
 

### 2021-05-31 manuscript/main.tex
shorter intro

@@
 
-    A pangenome models the full set of genomic elements in a given species or clade~\citep{32453966}. Indeed, these
-    data structures encode the mutual relationships between all the genomes represented, in contrast to reference-based
-    approaches which relate sequences only to a particular genome, chosen as reference. A class of methods to represent
+    A pangenome models the full set of genomic elements in a population~\citep{32453966}. Indeed, these data structures
+    encode the mutual relationships between all the genomes represented, in contrast to reference-based approaches
+    which relate sequences only to a particular genome, chosen as reference. A class of methods to represent
     pangenomes involves the \textit{sequence graphs}~\citep{2488477}, where similar orthologous regions between genomes
@@
     address these challenges~\citep{33040146}. Memory-efficient representations do not allow dynamic updates to the
-    graph, such as adding new genomes or simplifying the graph topology. Furthermore, dynamic updates must perform
-    efficiently on the complex graphs generated from whole-genome alignment of vertebrates. Highly repetitive regions,
-    such as centromeres, segmental duplications, and acrocentric chromosomes lead to graphs which present paths with
-    very high-depth nodes. The node depth is defined as the number of times in which the node is crossed by all the
-    paths embedded in the graph. High depth increases the complexity of the operations performed on the paths and the
-    nodes they cross.
+    graph. Furthermore, the changes must perform efficiently on the complex graphs generated from whole-genome
+    alignment of vertebrates. Highly repetitive regions, such as centromeres, segmental duplications, and acrocentric
+    chromosomes lead to graphs which present very high-depth nodes. The node depth is defined as the
+    number of times in which the node is crossed by all the paths embedded in the graph. High depth increases the
+    complexity of the operations performed on the paths and the nodes they cross.
 
@@
     its design to support highly complex pangenome graphs. The majority of the \odgi\ tools are implemented in an
-    index-free manner, avoiding the burning task needed to create an index structure. Indeed, index-constrained
-    implementations would require building indexes also for single or little operations on the graphs. Moreover,
+    index-free manner, avoiding the burning task to create an index structure. Indeed, index-constrained
+    implementations require building indexes also for single or little operations on the graphs. Moreover,
     thanks to its efficient path representation, \odgi\ tools can build the internal graph representation (the \odgi\
-    format) in parallel, which is one of the major bottlenecks when working with very large and deep variation graphs,
-    efficiently scaling up to hundreds of large vertebrate genomes.
+    format) in parallel, which is one of the major bottlenecks when working with very large and deep graphs,
+    scaling up to hundreds of large vertebrate genomes.
 

### 2021-06-01 manuscript/figure_p.tex
Caption edited

@@
         }
-        \caption{Visualizing the Major Histocompatibility Complex (MHC) \textit{locus}.
-            (a) projection of full MHC \textit{locus} of eight haploid phased human genome
-            assemblies, plus the chm13 cell line and GRCh38 reference genomes, as supplied by the
-            Human Reference Pangenome Consortium\citep{HRPC}, created by \cmd{odgi build+sort+viz}. The coloured
-            bars represent the linearized paths, representing contigs, as a zoomed out multi-sequence
-            alignment. The black lines at the bottom represent the graph topology.
-            (b) consensus graph representation of the same assemblies showing variations larger than $100$ base pairs.
-            The gene label coordinates were super imposed by using \cmd{odgi position} from GRCh38.
-            Figure generated by the Bandage tool\citep{26099265}.
+        \caption{Pangenome visualisations of the Major Histocompatibility Complex (MHC) \textit{locus} by \odgi.
+            (a) projection of full MHC \textit{locus} of haploid phased human genome
+          assemblies, plus the chm13 cell line and GRCh38 reference genome (note the missing patch in the reference
+          genome) as supplied by the
+            Human Reference Pangenome Consortium\citep{HRPC} and created by \cmd{odgi build+sort+viz}. The coloured
+            bars represent the linearised paths --- representing contigs as a zoomed out multi-sequence
+            alignment. The black lines at the bottom show the graph topology.
+            (b) consensus graph representation by \cmd{odgi build} of the same assemblies showing variations larger than $100$ base pairs.
+            Loops or `bubbles' display contigs uniquely diverging from the consensus, caused by variation in repeats.
+            The gene labels are super imposed by \cmd{odgi position} from a GRCh38 BED file and visualised by the Bandage tool\citep{26099265}.
         }

### 2021-06-01 manuscript/main.tex
First authors - only one correspondent
In an app note we say features
Changed font for cmd

@@
 \newcommand{\odgi}{ODGI}
-\newcommand{\cmd}[1]{{\texttt{#1}}}
+\newcommand{\cmd}[1]{{\scriptsize\textrm{#1}}}
 \newcommand{\cmdbf}[1]{{\textbf{#1}}}
@@
     \corresp{$^\ast$To whom correspondence should be addressed.}
-    \corresp{$^\dagger$Shared first authorship.}
+    \corresp{$^\dagger$The authors wish it to be known that the first two authors should be regarded as joint First Authors.}
 
@@
         \textbf{Contact:}
-        \href{andreaguarracino@outlook.com}{andreaguarracino@outlook.com},
+        % \href{andreaguarracino@outlook.com}{andreaguarracino@outlook.com},
         \href{erik.garrison@gmail.com}{erik.garrison@gmail.com}\\
@@
 
-    \section{Results}
+    \section{Features}
 
@@
     (GFAv1) format\citep{GFAv1}; \cmd{odgi build} and \cmd{odgi view}
-    convert graphs from GFA to the \odgi\ format and vice versa.
+    convert graphs from GFA to the \odgi\ format and \textit{vice versa}.
     \cmd{odgi validate} ascertains the graph data is correct, such as

### 2021-06-01 manuscript/main.tex
Fixed topics

@@
     supports interrogation of the graph for basic dimensional
-    properties, positional and path-related queries, and visualization
+    properties, positional and path-related queries, and visualisation
     in one-dimension (1D) and two-dimensions (2D). \odgi\ also supports
@@
 
-    \topic{odgi build, view \& validate} pangenome graphs are
-    represented by a subset of the textual Graphical Fragment Assembly
+    \topic{odgi build, view \& validate} pangenome graphs can be
+    stored in the textual Graphical Fragment Assembly
     (GFAv1) format\citep{GFAv1}; \cmd{odgi build} and \cmd{odgi view}
@@
 
-    \topic{odgi viz, sort, draw \& layout} pangenome visualization
+    \topic{odgi viz, sort, draw \& layout} pangenome visualisation
     provides convenient insight into genomic variation. \cmd{odgi viz}
@@
     \cmd{odgi layout} generates a file with the X-Y coordinates for
-    all nodes and \cmd{odgi draw} draws the actual layout of the graph
+    all nodes and \cmd{odgi draw} draws a layout of the graph
     (Fig.~\ref{fig:1}b).
@@
     ways. E.g., \cmd{odgi stats} returns the number of nodes,
-    edges, paths, and graph length. \cmd{odgi bin} summarizes path
+    edges, paths, and graph length. \cmd{odgi bin} summarises path
     information into bins of a specified size, thereby enabling a
-    summarized view of gigabase scale graphs. \cmd{odgi depth} and
+    summarised view of large genomes. \cmd{odgi depth} and
     \cmd{odgi degree} return the node depth and degree as defined by
@@
 
-    \topic{odgi break, groom, chop \& unchop} \odgi\ pangenome graphs are dynamic structures in ODGI.
-    \cmd{odgi break} removes cycles in the graph, reducing its topology complexity. \cmd{odgi groom}
-    removes spurious inverting links by exploring the graph from the orientation supported by most paths. \cmd{odgi chop}
-    divides long nodes into shorter ones at a maximum requested size, simplifying downstream analysis.
-    \cmd{odgi unchop} allows joining nodes that can be merged without changing the graph topology, nor the
-    embedded sequences, obtaining an equivalent, but more compact, representation of the graph.
+    \topic{odgi break, groom, chop \& unchop} pangenome graphs can be
+    reorganised.  \cmd{odgi break} removes circular structures in the
+    graph, thereby reducing complexity. \cmd{odgi groom} removes
+    spurious inverted links by aligning the graph from the orientation
+    that is supported by most paths. \cmd{odgi chop} divides long
+    nodes into shorter ones at a maximum requested size, thereby
+    simplifying downstream analysis.  \cmd{odgi unchop} joins nodes
+    and embedded sequences that do not change the graph topology,
+    thereby obtaining a more compact representation of the graph.
 
-    \topic{odgi explode, squeeze \& extract} pangenome graphs can embed multiple chromosomes as separated connected
-    components. \cmd{odgi explode} separates them in different files, while \cmd{odgi squeeze} merges
-    multiple graphs into the same file, preventing node ID collisions. \cmd{odgi extract} extracts regions of the graph
-    as defined by query criteria, thereby reducing the resources to work only with the subgraph.
+    \topic{odgi explode, squeeze \& extract} pangenome are constructed
+    as a large graph. \cmd{odgi explode} separates units, such as
+    chromosomes, into different files.  \cmd{odgi squeeze} merges
+    multiple graphs into the same file whilst preventing node ID
+    collisions. \cmd{odgi extract} extracts regions of the graph as
+    defined by certain criteria, allowing downstream processing of
+    smaller subgraphs.
 
-    \topic{odgi position} embedded paths provide stable coordinates regardless of the graph in which they are.
-    \cmd{odgi position} finds, translates, and liftovers graph and path positions between different graphs by
-    exploiting their shared path sequences (Figure~2\vphantom{\ref{fig:2}}).
+    \topic{odgi position} pangenome graphs are flexible when it comes
+    to coordinate systems. \cmd{odgi position} can use the coordinate
+    system from a contained reference genome --- a dynamic liftover
+    --- to display coordiates and other localised features, as is shown in
+    Fig.~\ref{fig:1b}.
 

### 2021-06-01 manuscript/main.tex
Table

@@
 \newcommand{\odgi}{ODGI}
+\newcommand{\FIXME}{\textbf{!!FIXME!!}}
 \newcommand{\cmd}[1]{{\scriptsize\textrm{#1}}}
@@
 
-    Table~\ref{tab:02} shows a comparison of the operations necessary to extract a subgraph using \odgi\ and VG. The
-    example consists in extracting the centromeres from a pangenome graph made from 88 haploid, phased human genome
-    assemblies from 44 individuals from the Human Pangenome Reference Consortium year 1 assembly (see Data
-    Availability), plus the chm13 and GRCh38 reference genomes, for a total of 90 haplotypes. All commands are provided
-    in the Supplementary Material.
+    Table~\ref{tab:02} shows a comparison of the operations necessary to extract a subgraph using \odgi\ and VG.
 
@@
     \begin{table}[!t]
-        \processtable{Centromeres extraction comparison using \odgi\ and VG
+      \processtable{Example of performance improvement between
+        \cmd{odgi extract} and \cmd{VG something}. Centromere region
+        is extracted from CHR ?? of the HPRC year 1
+        assembly\citep{HPRC}, i.e., 88 haploid, phased human genome
+        assemblies from 44 individuals plus the chm13 and GRCh38
+        reference genomes stored in 100GB(?) GFA format. All timings
+        were performed on a server class machine on SSD MODEL with X
+        GB of RAM and 48 CPU cores (Y x 12 core AMD Opteron) Processor
+        MODEL @ 3.3 GHz with 8MB L2 Cache.\FIXME
           \label{tab:02}} {

### 2021-06-01 manuscript/main.tex
added few missing info in the table caption

@@
         \cmd{odgi extract} and \cmd{VG something}. Centromere region
-        is extracted from CHR ?? of the HPRC year 1
+        is extracted from CHR 6 of the HPRC year 1
         assembly\citep{HPRC}, i.e., 88 haploid, phased human genome
         assemblies from 44 individuals plus the chm13 and GRCh38
-        reference genomes stored in 100GB(?) GFA format. All timings
+        reference genomes stored in 4.3GB GFA format. All timings
         were performed on a server class machine on SSD MODEL with X

### 2021-06-01 manuscript/main.tex
edits

@@
 \access{Advance Access Publication Date: Day Month Year}
-\appnotes{Manuscript Category}
+\appnotes{Application Note}
 
@@
 
-    \abstract{
-        \textbf{Motivation:} Pangenome graphs provide a complete representation of
-        genomic information that allows for complex population-sized analysis of genomes
-        without introducing reference genome bias.
+    \abstract{ \textbf{Motivation:} Pangenomes address the
+      shortcomings of mainline genomics workflows, in particular with
+      regard to reference genome bias and the identifation of complex
+      structural variants.
         %
-        \\
-        \textbf{Results:} Here we present \odgi\, a novel range of tools that makes use of an efficient
-        in-memory representation of DNA graphs structured in the variation graph model. \odgi\ can
-        handle hundreds of large genomes while maintaining fast execution runtime and low memory overhead.
-        \odgi\ includes tools for detecting complex regions, extracting \textit{loci}, exploratory analysis,
-        sorting, removing artifacts, and graphs manipulation and visualisation.
+        \\ \textbf{Results:} Here we present \odgi\, a novel range of
+        scalable tools that makes use of an efficient in-memory
+        representation of DNA graphs. \odgi\ can handle hundreds of
+        large genomes at the same time while maintaining fast
+        execution runtime and low memory overheads.  \odgi\ includes
+        tools for detecting complex regions, extracting \textit{loci},
+        exploratory analysis, sorting, removing artifacts, and graphs
+        manipulation and visualisation.
         %
@@
         \textbf{Availability:} \odgi\ is software written in the \Cpp\ programming language and
-        published under the libre MIT license. Source code and user manual can be downloaded from
+        published under the permissive MIT license. Source code and user manual can be downloaded from
         \href{https://github.com/pangenome/odgi}{https://github.com/pangenome/odgi}.\\
@@
         % \href{andreaguarracino@outlook.com}{andreaguarracino@outlook.com},
-        \href{erik.garrison@gmail.com}{erik.garrison@gmail.com}\\
+        egarris5@uthcs.edu\\
         \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.}
@@
 
-    \subsection{Performance on human data}
-
     Table~\ref{tab:02} shows a comparison of the operations necessary to extract a subgraph using \odgi\ and VG.
@@
       \processtable{Example of performance improvement between
-        \cmd{odgi extract} and \cmd{VG something}. Centromere region
-        is extracted from CHR 6 of the HPRC year 1
-        assembly\citep{HPRC}, i.e., 88 haploid, phased human genome
-        assemblies from 44 individuals plus the chm13 and GRCh38
-        reference genomes stored in 4.3GB GFA format. All timings
-        were performed on a server class machine on SSD MODEL with X
-        GB of RAM and 48 CPU cores (Y x 12 core AMD Opteron) Processor
-        MODEL @ 3.3 GHz with 8MB L2 Cache.\FIXME
+        \cmd{odgi} and \cmd{VG} equivalent commands with output file
+        sizes. The centromere region was extracted from CHR 6 of the
+        HPRC year one assembly\citep{HPRC}, i.e., 88 haploid, phased
+        human genome assemblies from 44 individuals plus the chm13
+        cell line and GRCh38 reference genomes stored in 4.3GB GFA
+        format.
+        %
+        First GFS was converted from GFA to \odgi\ and \cmd{vg}
+        formats, respectively.  \odgi's path implementation can
+        process paths in parallel, outperforming \cmd{vg}. For the
+        same reason \cmd{odgi extract} is $20\times$ faster than
+        \cmd{vg chunk} extracting a portion from the full
+        graph. \FIXME: vg never finished.
+        %
+        All timings were performed on a server class machine
+        on SSD MODEL with X GB of RAM and 48 CPU cores (Y x 12 core
+        AMD Opteron) Processor MODEL @ 3.3 GHz with 8MB L2
+        Cache.\FIXME
           \label{tab:02}} {
@@
     \end{table}
-
-    To work with \odgi\ and VG it is first necessary to convert the graph from GFA to \odgi\ and VG format, respectively.
-    Thanks to the \odgi\ format's path implementation, \cmd{odgi build} embeds the paths in parallel, outperforming
-    \cmd{vg convert} in the same task. For the same reason, \cmd{odgi extract} results 20 times faster than
-    \cmd{vg chunk} in extracting a portion from the full graph.
 

### 2021-06-01 manuscript/main.tex
Bib works

@@
 \usepackage{subfigure}
+
+\usepackage{natbib}
 
@@
 
+    Table~\ref{tab:02} shows a comparison of the operations necessary to extract a subgraph using \odgi\ and VG.
+
+    %TODO to repeat on a bigger machine to avoid vg crashing
+    \begin{table}[!t]
+      \processtable{Example of performance improvement between
+        \cmd{odgi} and \cmd{VG} equivalent commands with output file
+        sizes. The centromere region was extracted from CHR 6 of the
+        HPRC year one assembly\citep{HPRC}, i.e., 88 haploid, phased
+        human genome assemblies from 44 individuals plus the chm13
+        cell line and GRCh38 reference genomes stored in 4.3GB GFA
+        format\citep{GFA}.
+        %
+        First GFS was converted from GFA to \odgi\ and \cmd{vg}
+        formats, respectively.  \odgi's path implementation can
+        process paths in parallel, outperforming \cmd{vg}. For the
+        same reason \cmd{odgi extract} is $20\times$ faster than
+        \cmd{vg chunk} extracting a portion from the full
+        graph. \FIXME: vg never finished.
+        %
+        All timings were performed on a server class machine
+        on SSD MODEL with X GB of RAM and 48 CPU cores (Y x 12 core
+        AMD Opteron) Processor MODEL @ 3.3 GHz with 8MB L2
+        Cache.\FIXME
+          \label{tab:02}} {
+          \resizebox{.99\columnwidth}{!}{
+            \begin{tabular}{@{}lllll@{}}
+                \toprule Operation & Tool & Runtime (mm:ss) & Memory (GB) & Output size (GB)    \\
+                \midrule
+                Format conversion                           &       &       &                   \\
+                                    & \cmd{odgi build}      & 1:35  & 10.39 & 5.4               \\
+                                    & \cmd{vg convert}      & 8:14  & 28.43 & 6.1               \\
+                Subgraph extraction                         &       &       &                   \\
+                                    & \cmd{odgi extract}    & 1:15  & 9.43  & 2.7               \\
+                                    & \cmd{vg chunk}        & 21:09 & 59.33 & 1.7               \\
+                \botrule
+            \end{tabular}}}
+
+    \end{table}
+
     \section{Features}
@@
     stored in the textual Graphical Fragment Assembly
-    (GFAv1) format\citep{GFAv1}; \cmd{odgi build} and \cmd{odgi view}
+    (GFAv1) format\citep{GFA}; \cmd{odgi build} and \cmd{odgi view}
     convert graphs from GFA to the \odgi\ format and \textit{vice versa}.
@@
 
-    Table~\ref{tab:02} shows a comparison of the operations necessary to extract a subgraph using \odgi\ and VG.
-
-    %TODO to repeat on a bigger machine to avoid vg crashing
-    \begin{table}[!t]
-      \processtable{Example of performance improvement between
-        \cmd{odgi} and \cmd{VG} equivalent commands with output file
-        sizes. The centromere region was extracted from CHR 6 of the
-        HPRC year one assembly\citep{HPRC}, i.e., 88 haploid, phased
-        human genome assemblies from 44 individuals plus the chm13
-        cell line and GRCh38 reference genomes stored in 4.3GB GFA
-        format.
-        %
-        First GFS was converted from GFA to \odgi\ and \cmd{vg}
-        formats, respectively.  \odgi's path implementation can
-        process paths in parallel, outperforming \cmd{vg}. For the
-        same reason \cmd{odgi extract} is $20\times$ faster than
-        \cmd{vg chunk} extracting a portion from the full
-        graph. \FIXME: vg never finished.
-        %
-        All timings were performed on a server class machine
-        on SSD MODEL with X GB of RAM and 48 CPU cores (Y x 12 core
-        AMD Opteron) Processor MODEL @ 3.3 GHz with 8MB L2
-        Cache.\FIXME
-          \label{tab:02}} {
-          \resizebox{.99\columnwidth}{!}{
-            \begin{tabular}{@{}lllll@{}}
-                \toprule Operation & Tool & Runtime (mm:ss) & Memory (GB) & Output size (GB)    \\
-                \midrule
-                Format conversion                           &       &       &                   \\
-                                    & \cmd{odgi build}      & 1:35  & 10.39 & 5.4               \\
-                                    & \cmd{vg convert}      & 8:14  & 28.43 & 6.1               \\
-                Subgraph extraction                         &       &       &                   \\
-                                    & \cmd{odgi extract}    & 1:15  & 9.43  & 2.7               \\
-                                    & \cmd{vg chunk}        & 21:09 & 59.33 & 1.7               \\
-                \botrule
-            \end{tabular}}}
-
-    \end{table}
-
     \section{Discussion}
@@
 
-%\bibliographystyle{natbib}
+\bibliographystyle{natbib}
 %\bibliographystyle{achemnat}
@@
 
-    %ToDo
-    \begin{thebibliography}{}
+\bibliography{bibliography}
 
-        \bibitem[Bofelli {\it et~al}., 2000]{Boffelli03}
-        Bofelli,F., Name2, Name3 (2003) Article title, {\it Journal Name}, {\bf 199}, 133-154.
-
-        \bibitem[Bag {\it et~al}., 2001]{Bag01}
-        Bag,M., Name2, Name3 (2001) Article title, {\it Journal Name}, {\bf 99}, 33-54.
-
-        \bibitem[Yoo \textit{et~al}., 2003]{Yoo03}
-        Yoo,M.S. \textit{et~al}. (2003) Oxidative stress regulated genes
-        in nigral dopaminergic neurnol cell: correlation with the known
-        pathology in Parkinson's disease. \textit{Brain Res. Mol. Brain
-        Res.}, \textbf{110}(Suppl. 1), 76--84.
-
-        \bibitem[Lehmann, 1986]{Leh86}
-        Lehmann,E.L. (1986) Chapter title. \textit{Book Title}. Vol.~1, 2nd edn. Springer-Verlag, New York.
-
-        \bibitem[Crenshaw and Jones, 2003]{Cre03}
-        Crenshaw, B.,III, and Jones, W.B.,Jr (2003) The future of clinical
-        cancer management: one tumor, one chip. \textit{Bioinformatics},
-        doi:10.1093/bioinformatics/btn000.
-
-        \bibitem[Auhtor \textit{et~al}. (2000)]{Aut00}
-        Auhtor,A.B. \textit{et~al}. (2000) Chapter title. In Smith, A.C.
-        (ed.), \textit{Book Title}, 2nd edn. Publisher, Location, Vol. 1, pp.
-        ???--???.
-
-        \bibitem[Bardet, 1920]{Bar20}
-        Bardet, G. (1920) Sur un syndrome d'obesite infantile avec
-        polydactylie et retinite pigmentaire (contribution a l'etude des
-        formes cliniques de l'obesite hypophysaire). PhD Thesis, name of
-        institution, Paris, France.
-
-    \end{thebibliography}
 \end{document}

### 2021-06-01 manuscript/main.tex
Shortened

@@
 \newcommand{\odgi}{ODGI}
+\newcommand{\vg}{VG}
+
 \newcommand{\FIXME}{\textbf{!!FIXME!!}}
@@
 
-    A pangenome models the full set of genomic elements in a population~\citep{32453966}. Indeed, these data structures
-    encode the mutual relationships between all the genomes represented, in contrast to reference-based approaches
-    which relate sequences only to a particular genome, chosen as reference. A class of methods to represent
-    pangenomes involves the \textit{sequence graphs}~\citep{2488477}, where similar orthologous regions between genomes
-    collapse into a single representative sequence. In \textit{node-labeled} sequence graphs, nodes indicate DNA
-    sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph. A
-    \textit{bidirected} sequence graph represents both strands of DNA. On this model, \textit{variation graphs} add the
-    concept of \textit{path} to embed linear sequences (e.g., genomes or haplotypes) into the graph~\citep{30125266}.
-    Paths provide a stable coordinate system, allowing graph annotations and interoperability between different graphs.
+    Thanks to advances in sequencing technologies, new genome
+    assemblies are produced at a high rate, including the human
+    pangenome reference consortium (HPRC) and telomere-to-telemere
+    efforts, e.g.~\citep{Miga:2020}.  A pangenome models the full set
+    of genomic elements in a given species or clade. Pangenomics thus
+    stands in contrast to reference-based genomic approaches which
+    relate sequences to a particular consensus model of the
+    genome~\citep{Eizenga:2020}. A pangenome can elegantly be
+    represented by a graph data structure incorporating sequences as
+    nodes and their connections as edges. These nodes can be shared
+    for homologs and orthologs.  A `bidirected' graph represents both
+    strands of DNA and so-called paths represent contigs that run
+    through the nodes. A path can represent a chromosome or part
+    thereof.
 
-    Thanks to advances in sequencing technologies, new genome assemblies are produced at a high rate. This amount of
-    data offers the opportunity to study genomic variation as never before, but it also brings challenges in how to
-    represent and manipulate hundreds of genomes at a gigabase scale. Variation graphs compactly represent the full
-    genetic variation across a population, but graph-based data structures present computational overheads: in addition
-    to the sequences, they must also represent the graph topology. Several genome data structures were developed to
-    address these challenges~\citep{33040146}. Memory-efficient representations do not allow dynamic updates to the
-    graph. Furthermore, the changes must perform efficiently on the complex graphs generated from whole-genome
-    alignment of vertebrates. Highly repetitive regions, such as centromeres, segmental duplications, and acrocentric
-    chromosomes lead to graphs which present very high-depth nodes. The node depth is defined as the
-    number of times in which the node is crossed by all the paths embedded in the graph. High depth increases the
-    complexity of the operations performed on the paths and the nodes they cross.
+    A major challenge is writing software that can deal with the size
+    of these graphs. vgtools pioneered graph
+    processing~\citep{vgtools,Eizenga:2020b}, but was not designed for
+    processing large data (see Table~\ref{tab:02}). With \odgi\ we
+    wrote a new toolset and API in \Cpp\ named Optimized Dynamic
+    Genome/Graph Implementation (ODGI).  that implements is a memory
+    efficient graph structure that can be dynamically updated using
+    multiple CPUs. This is particularly relevant for highly repetitive
+    regions. Segmental duplications, and acrocentric chromosomes lead
+    to graphs which present very high node depths which greatly
+    increases processing complexity.
 
-    To overcome this limitation, we have developed an Optimized Dynamic Genome/Graph Implementation (ODGI). Each tool
-    loads graphs into the succinct dynamic HandleGraph model~\citep{33040146} and applies methods based on the
-    HandleGraph API. Several algorithms are shared with the VG toolkit (VG)~\citep{30125266}, the state-of-the-art
-    toolkit for working with pangenome graphs. However, although similar in scope to VG, \odgi\ importantly differs in
-    its design to support highly complex pangenome graphs. The majority of the \odgi\ tools are implemented in an
-    index-free manner, avoiding the burning task to create an index structure. Indeed, index-constrained
-    implementations require building indexes also for single or little operations on the graphs. Moreover,
-    thanks to its efficient path representation, \odgi\ tools can build the internal graph representation (the \odgi\
-    format) in parallel, which is one of the major bottlenecks when working with very large and deep graphs,
-    scaling up to hundreds of large vertebrate genomes.
+    Note that \odgi\ is focused on `first-order' graph
+    operations.  Other tools are required to construct graphs from raw
+    sequences, map reads, or call variants.  \odgi\ provides
+    operations on an existing graph, such as subset, subdivide, break,
+    combine, normalize and order its components, nodes, and
+    paths. Most of the tools are designed to be applied together,
+    piping the output of one tool into the next.
 
-    \odgi\ is entirely focused on "first-order" operations on the graph
-    (Table~\ref{tab:01}), providing, unlike VG, no method to construct
-    graphs from raw sequences, map reads, or call variants. It
-    supports interrogation of the graph for basic dimensional
-    properties, positional and path-related queries, and visualisation
-    in one-dimension (1D) and two-dimensions (2D). \odgi\ also supports
-    operations on the graph such as subset, subdivide, break, combine,
-    normalize, or order its components, nodes, and paths. Most of the
-    tools are designed to be applied together, piping the output of
-    one tool into the next, thereby preventing the creation of
-    intermediate files, and reducing the number of IO operations.
+    \vspace{-0.1in}
 
-    \section{Methods}
+    \section{Results}
 
-    \odgi\ is designed to build and modify paths in parallel, keeping its representation in memory efficient. It
-    applies concepts first introduced in the dynamic version of the Graph BWT~\citep{31406990}, a scalable
-    implementation of the graph extension of the positional Burrows-Wheeler transform~\citep{28702075}. Specifically,
-    we build a \textit{node-local} model of the graph, where the graph is stored in a vector of node structures.
-    The identifier of each node is its offset in this vector plus 1. Each structure contains the sequence of the
-    node, its edges in both directions (forward and backward), and a vector of \textit{path steps} that describes
-    how the embedded paths walk across the node. Each step records the path identifier and the previous and next
-    path steps on the same path. To reduce the encoding, we encode path steps using a local alphabet that maps the
-    \textit{N} neighbors of the node into the range \textit{1..N}. To further save space, \textit{node deltas},
-    rather than IDs, are stored in this alphabet. The delta between two nodes is defined as their distance in the
-    graph vector (i.e., the difference between the node offsets). The path step vector, of length \textit{n}, is
-    stored using a dynamic succinct integer vector that requires \textit{O(5nw)} bits, where \textit{w} is
-    \textit{\textasciitilde log\textsubscript{2}(N)}. Finally, to allow operating on path steps in parallel, each node
-    structure includes a byte-width mutex lock.
-
-    Table~\ref{tab:02} shows a comparison of the operations necessary to extract a subgraph using \odgi\ and VG.
-
-    %TODO to repeat on a bigger machine to avoid vg crashing
-    \begin{table}[!t]
-      \processtable{Example of performance improvement between
-        \cmd{odgi} and \cmd{VG} equivalent commands with output file
-        sizes. The centromere region was extracted from CHR 6 of the
-        HPRC year one assembly\citep{HPRC}, i.e., 88 haploid, phased
-        human genome assemblies from 44 individuals plus the chm13
-        cell line and GRCh38 reference genomes stored in 4.3GB GFA
-        format\citep{GFA}.
-        %
-        First GFS was converted from GFA to \odgi\ and \cmd{vg}
-        formats, respectively.  \odgi's path implementation can
-        process paths in parallel, outperforming \cmd{vg}. For the
-        same reason \cmd{odgi extract} is $20\times$ faster than
-        \cmd{vg chunk} extracting a portion from the full
-        graph. \FIXME: vg never finished.
-        %
-        All timings were performed on a server class machine
-        on SSD MODEL with X GB of RAM and 48 CPU cores (Y x 12 core
-        AMD Opteron) Processor MODEL @ 3.3 GHz with 8MB L2
-        Cache.\FIXME
-          \label{tab:02}} {
-          \resizebox{.99\columnwidth}{!}{
-            \begin{tabular}{@{}lllll@{}}
-                \toprule Operation & Tool & Runtime (mm:ss) & Memory (GB) & Output size (GB)    \\
-                \midrule
-                Format conversion                           &       &       &                   \\
-                                    & \cmd{odgi build}      & 1:35  & 10.39 & 5.4               \\
-                                    & \cmd{vg convert}      & 8:14  & 28.43 & 6.1               \\
-                Subgraph extraction                         &       &       &                   \\
-                                    & \cmd{odgi extract}    & 1:15  & 9.43  & 2.7               \\
-                                    & \cmd{vg chunk}        & 21:09 & 59.33 & 1.7               \\
-                \botrule
-            \end{tabular}}}
-
-    \end{table}
-
-    \section{Features}
-
-    \odgi\ has a wide range of commands.
+    \odgi\ is designed to build and modify paths in parallel. It
+    applies concepts first introduced in the dynamic version of the
+    graph extension of the positional Burrows-Wheeler
+    transform~\citep{Siren:2020}. For implementation details see the
+    \odgi\ online documentation. \odgi\ includes the following tools
 
@@
     \section{Discussion}
-    We implemented a tool suite, \odgi\, which is simple to work with to analyse and manipulate pangenome graphs at a
-    population scale. ODGI's path implementation allows working with graphs of any complexity, without particular
-    limitations. We have demonstrated that \odgi\ tools efficiently perform with pangenome graphs embedding full human
-    chromosomes. \odgi\ is not only a tool suite but also a library, already used in pangenome graphs building
-    (\citep{pggb}) and normalization (\citep{smoothxg}). Furthermore, users can add new tools, exploiting ODGI as a
-    framework where to implement algorithms working on variation graphs. Therefore, \odgi\ offers an efficient and
-    extendable suite of tools ready for the upcoming challenges in the analyses of hundreds of genomes at a gigabase
-    scale that characterize the pangenomic era.
+
+    Pangenome graphs stand to become a ubiquitous tool in
+    genomics\citep{Eizenga:2020}.  With \odgi\ we implemented a
+    state-of-the-art tool suite, that can transform, analyse and
+    visualise pangenome graphs at scale.  \odgi\ also exposes an API
+    that is used by other tools for pangenome graph building and
+    normalization. Future work will add support for RNA and protein
+    sequences and expand on metadata capabilities of large pangenome
+    graphs.
+
+    %TODO to repeat on a bigger machine to avoid vg crashing
+    \begin{table}[th]
+      \processtable{\small
+        Example of performance improvement between
+        \cmd{odgi} and \cmd{vg}~\citep{vgtools} equivalent commands with output file
+        sizes. The centromere region was extracted from CHR 6 of the
+        HPRC year one assembly, i.e., 88~haploid, phased
+        human genome assemblies from 44~individuals plus the chm13
+        cell line and GRCh38 reference genomes stored in 4.3GB~GFA
+        format\citep{GFA}.
+        %
+        First GFS was converted from GFA to \odgi\ and \vg\
+        formats, respectively.  \odgi's path implementation can
+        process paths in parallel, outperforming \vg. For the
+        same reason \cmd{odgi extract} is $20\times$ faster than
+        \cmd{vg chunk} extracting a portion from the full
+        graph. \FIXME: vg never finished.
+        %
+        All timings were performed on a server class machine
+        on SSD MODEL with X GB of RAM and 48 CPU cores (Y x 12 core
+        AMD Opteron) Processor MODEL @ 3.3 GHz with 8MB L2
+        Cache.\FIXME
+        % \vspace{0.5in}
+        \label{tab:02}}
+                   {
+                     % \resizebox{.8\textwidth}{!}{
+            \begin{tabular*}{.5\textwidth}{@{}lllll@{}}
+                \toprule Operation & Tool & Runtime (mm:ss) & Memory (GB) & Output size (GB)    \\
+                \midrule
+                Format conversion                           &       &       &                   \\
+                                    & \cmd{odgi build}      & 1:35  & 10.39 & 5.4               \\
+                                    & \cmd{vg convert}      & 8:14  & 28.43 & 6.1               \\
+                Subgraph extraction                         &       &       &                   \\
+                                    & \cmd{odgi extract}    & 1:15  & 9.43  & 2.7               \\
+                                    & \cmd{vg chunk}        & 21:09 & 59.33 & 1.7               \\
+                \botrule
+            \end{tabular*}} %}
+
+    \end{table}
 
@@
     \textit{Conflict of Interest}: none declared.
-    \vspace*{+12pt}
+    % \vspace*{+12pt}
 

### 2021-06-01 manuscript/main.tex
edits

@@
     \abstract{ \textbf{Motivation:} Pangenomes address the
-      shortcomings of mainline genomics workflows, in particular with
-      regard to reference genome bias and the identifation of complex
+      shortcomings of mainline genomics workflows, in particular
+      concerning reference genome bias and identification of complex
       structural variants.
@@
         tools for detecting complex regions, extracting \textit{loci},
-        exploratory analysis, sorting, removing artifacts, and graphs
+        exploratory analysis, sorting, removing artifacts, and graph
         manipulation and visualisation.
@@
         egarris5@uthcs.edu\\
-        \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.}
+        \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online. \FIXME}
 
@@
     efforts, e.g.~\citep{Miga:2020}.  A pangenome models the full set
-    of genomic elements in a given species or clade. Pangenomics thus
-    stands in contrast to reference-based genomic approaches which
+    of genomic elements in a given species or clade. Pangenomics
+    therefore contrasts with reference genome based approaches which
     relate sequences to a particular consensus model of the
@@
     represented by a graph data structure incorporating sequences as
-    nodes and their connections as edges. These nodes can be shared
-    for homologs and orthologs.  A `bidirected' graph represents both
-    strands of DNA and so-called paths represent contigs that run
-    through the nodes. A path can represent a chromosome or part
-    thereof.
+    nodes and their connections as edges. These nodes are shared for
+    homologs, paralogs and orthologs.  A `bidirected' graph represents
+    both strands of DNA and so-called paths represent contigs that run
+    through the nodes. A path can represent an individual chromosome
+    or part thereof. Such a graph-based approach is particularly
+    relevant for highly repetitive regions. Segmental duplications,
+    and acrocentric chromosomes lead to graphs which present very high
+    node depths which greatly increases processing complexity.
 
-    A major challenge is writing software that can deal with the size
-    of these graphs. vgtools pioneered graph
+    A major challenge is writing software that can deal with the sheer
+    size of these graphs representing hundreds, and soon thousands, of
+    human genomes. The \vg\ toolkit pioneered graph
     processing~\citep{vgtools,Eizenga:2020b}, but was not designed for
-    processing large data (see Table~\ref{tab:02}). With \odgi\ we
+    large data (see Table~\ref{tab:02}). With \odgi\ we
     wrote a new toolset and API in \Cpp\ named Optimized Dynamic
-    Genome/Graph Implementation (ODGI).  that implements is a memory
-    efficient graph structure that can be dynamically updated using
-    multiple CPUs. This is particularly relevant for highly repetitive
-    regions. Segmental duplications, and acrocentric chromosomes lead
-    to graphs which present very high node depths which greatly
-    increases processing complexity.
+    Genome/Graph Implementation (ODGI). \odgi\ implements a efficient
+    graph structure in RAM that can be dynamically updated using
+    multiple CPUs.
 
@@
     transform~\citep{Siren:2020}. For implementation details see the
-    \odgi\ online documentation. \odgi\ includes the following tools
+    \odgi\ online documentation. \odgi\ includes the following tools:
 
-    \topic{odgi build, view \& validate} pangenome graphs can be
+    \indent \topic{odgi build, view \& validate} pangenome graphs can be
     stored in the textual Graphical Fragment Assembly

### 2021-06-02 manuscript/main.tex
Fix table

@@
 
-    \input{figure_p}
-
     \section{Introduction}
@@
     \vspace{-0.1in}
+
+    \input{figure_p}
 
@@
 
+        %TODO to repeat on a bigger machine to avoid vg crashing
+    \begin{table}[h!]
+        \caption{ Example of performance improvement between
+            \cmd{odgi} and \cmd{vg}~\citep{vgtools} equivalent
+            commands with output file sizes. The centromere region was
+            extracted from CHR 6 of the HPRC year one assembly, i.e.,
+            88~haploid, phased human genome assemblies from
+            44~individuals plus the chm13 cell line and GRCh38
+            reference genomes stored in 4.3GB~GFA format\citep{GFA}.
+            %
+            First GFS was converted from GFA to \odgi\ and
+            \vg\ formats, respectively.  \odgi's path implementation
+            can process paths in parallel, outperforming \vg. For the
+            same reason \cmd{odgi extract} is $20\times$ faster than
+            \cmd{vg chunk} extracting a portion from the full
+            graph. \FIXME: vg never finished.
+            %
+            All timings were performed on a server class machine
+            on SSD MODEL with X GB of RAM and 48 CPU cores (Y x 12 core
+        AMD Opteron) Processor MODEL @ 3.3 GHz with 8MB L2
+        Cache.\FIXME}
+        \begin{tabular}{{@{}lllll@{}}}
+          \toprule
+          Operation & Command & Runtime & Memory  & File size     \\
+                    &      & (mm:ss) & (GB) & (GB)           \\
+          \midrule
+          Covert GFA to native format                         &       &       &                   \\
+          & \cmd{odgi build}      & 1:35  & 10.39 & 5.4               \\
+          & \cmd{vg convert}      & 8:14  & 28.43 & 6.1               \\
+          Extract subgraph                         &       &       &                   \\
+          & \cmd{odgi extract}    & 1:15  & 9.43  & 2.7               \\
+          & \cmd{vg chunk}        & 21:09 & 59.33 & 1.7               \\
+          \botrule
+        \end{tabular}
+        \label{tab:02}
+
+    \end{table}
+
     \section{Discussion}
@@
     graphs.
-
-    %TODO to repeat on a bigger machine to avoid vg crashing
-    \begin{table}[th]
-      \processtable{\small
-        Example of performance improvement between
-        \cmd{odgi} and \cmd{vg}~\citep{vgtools} equivalent commands with output file
-        sizes. The centromere region was extracted from CHR 6 of the
-        HPRC year one assembly, i.e., 88~haploid, phased
-        human genome assemblies from 44~individuals plus the chm13
-        cell line and GRCh38 reference genomes stored in 4.3GB~GFA
-        format\citep{GFA}.
-        %
-        First GFS was converted from GFA to \odgi\ and \vg\
-        formats, respectively.  \odgi's path implementation can
-        process paths in parallel, outperforming \vg. For the
-        same reason \cmd{odgi extract} is $20\times$ faster than
-        \cmd{vg chunk} extracting a portion from the full
-        graph. \FIXME: vg never finished.
-        %
-        All timings were performed on a server class machine
-        on SSD MODEL with X GB of RAM and 48 CPU cores (Y x 12 core
-        AMD Opteron) Processor MODEL @ 3.3 GHz with 8MB L2
-        Cache.\FIXME
-        % \vspace{0.5in}
-        \label{tab:02}}
-                   {
-                     % \resizebox{.8\textwidth}{!}{
-            \begin{tabular*}{.5\textwidth}{@{}lllll@{}}
-                \toprule Operation & Tool & Runtime (mm:ss) & Memory (GB) & Output size (GB)    \\
-                \midrule
-                Format conversion                           &       &       &                   \\
-                                    & \cmd{odgi build}      & 1:35  & 10.39 & 5.4               \\
-                                    & \cmd{vg convert}      & 8:14  & 28.43 & 6.1               \\
-                Subgraph extraction                         &       &       &                   \\
-                                    & \cmd{odgi extract}    & 1:15  & 9.43  & 2.7               \\
-                                    & \cmd{vg chunk}        & 21:09 & 59.33 & 1.7               \\
-                \botrule
-            \end{tabular*}} %}
-
-    \end{table}
 
@@
     \textit{Conflict of Interest}: none declared.
-    % \vspace*{+12pt}
+    \vspace*{+24pt}
 

### 2021-06-02 manuscript/figure_p.tex
caption: remove line on topology because you can not see it really
figure
Shortened
Fix table
edits

@@
 \begin{figure*}[ht]
-    \vspace{-10pt}
+    % \vspace{-5pt}
     \begin{minipage}[b]{\textwidth}
@@
 
-        \subfigure[aligned MHC pangenome]{
-            \includegraphics[width=.48\textwidth]{fig/mhc-align.eps}
+        \subfigure[MHC aligned pangenome]{
+            \includegraphics[width=.6\textwidth]{fig/mhc-align.eps}
             \label{fig:1a}
@@
         % \newline
-        \subfigure[consensus MHC pangenome]{
-            \includegraphics[width=.48\textwidth]{fig/mhc-pangenome.eps}
+        \subfigure[MHC consensus pangenome]{
+            \includegraphics[width=.4\textwidth]{fig/mhc-pangenome.eps}
             \label{fig:1b}
         }
-        \caption{Pangenome visualisations of the Major Histocompatibility Complex (MHC) \textit{locus} by \odgi.
+        \caption{Pangenome visualisation of the Major Histocompatibility Complex (MHC) \textit{locus} by \odgi.
             (a) projection of full MHC \textit{locus} of haploid phased human genome
-          assemblies, plus the chm13 cell line and GRCh38 reference genome (note the missing patch in the reference
-          genome) as supplied by the
-            Human Reference Pangenome Consortium\citep{HRPC} and created by \cmd{odgi build+sort+viz}. The coloured
+          assemblies, plus the chm13 cell line and GRCh38 reference genome as supplied by the
+            HPRC and created by \cmd{odgi build+sort+viz}. The coloured
             bars represent the linearised paths --- representing contigs as a zoomed out multi-sequence
-            alignment. The black lines at the bottom show the graph topology.
+            alignment. The black lines at the bottom represent the graph topology.
             (b) consensus graph representation by \cmd{odgi build} of the same assemblies showing variations larger than $100$ base pairs.
-            Loops or `bubbles' display contigs uniquely diverging from the consensus, caused by variation in repeats.
-            The gene labels are super imposed by \cmd{odgi position} from a GRCh38 BED file and visualised by the Bandage tool\citep{26099265}.
+            Loops or `bubbles' display contigs uniquely diverging from the consensus, caused by variation in paralogs/repeats.
+            The gene labels are super imposed by \cmd{odgi position} from a GRCh38 BED file and visualised by the Bandage tool~\citep{Wick:2015}.
         }

### 2021-06-02 manuscript/main.tex
edits

@@
 \newcommand{\FIXME}{\textbf{!!FIXME!!}}
-\newcommand{\cmd}[1]{{\scriptsize\textrm{#1}}}
+\newcommand{\cmd}[1]{\textbf{\textrm{#1}}}
 \newcommand{\cmdbf}[1]{{\textbf{#1}}}
@@
     \author[Guarracino, Heumos \textit{et~al}.]{
-        Andrea Guarracino\,$^{\text{\sfb 1,}\dagger,*}$,
-        Simon Heumos\,$^{\text{\sfb 2,}\dagger}$,
+        Andrea Guarracino\,$^{\text{\sfb 1,}*}$,
+        Simon Heumos\,$^{\text{\sfb 2,}*}$,
     % XXXXX,
         Pjotr~Prins\,$^{\text{\sfb 3}}$ and
-        Erik~Garrison\,$^{\text{\sfb 3,}*}$
+        Erik~Garrison\,$^{\text{\sfb 3,}\dagger}$
+        \footnote{The authors wish it to be known that the first two authors should be regarded as joint first authors.}
     }
@@
         $^{\text{\sf 2}}$QBiC, University of Tübingen, Tübingen, XXXXXXXXXX, Germany \\
-        $^{\text{\sf 3}}$Genetics Genomics \& Informatics, UTHSC, Memphis, TN, USA
+        $^{\text{\sf 3}}$Dept. Genetics Genomics \& Informatics, UTHSC, Memphis, TN, USA
     }
@@
     \corresp{$^\ast$To whom correspondence should be addressed.}
-    \corresp{$^\dagger$The authors wish it to be known that the first two authors should be regarded as joint First Authors.}
 
@@
 
-    \abstract{ \textbf{Motivation:} Pangenomes address the
-      shortcomings of mainline genomics workflows, in particular
-      concerning reference genome bias and identification of complex
+    \abstract{ \textbf{Motivation:} Pangenomes graphs address some of the
+      shortcomings of mainline genomics, in particular
+      with regard to reference genome bias and identification of complex
       structural variants.
@@
         \\ \textbf{Results:} Here we present \odgi\, a novel range of
-        scalable tools that makes use of an efficient in-memory
+        scalable tools that implements new algorithms and has an efficient in-memory
         representation of DNA graphs. \odgi\ can handle hundreds of
-        large genomes at the same time while maintaining fast
-        execution runtime and low memory overheads.  \odgi\ includes
+        large genomes in memory with fast
+        parallel execution.  \odgi\ includes
         tools for detecting complex regions, extracting \textit{loci},
         exploratory analysis, sorting, removing artifacts, and graph
-        manipulation and visualisation.
+        manipulation and visualisation. \odgi\ also provides an API that allows
+        for writing new tool.
         %
@@
         \textbf{Availability:} \odgi\ is software written in the \Cpp\ programming language and
-        published under the permissive MIT license. Source code and user manual can be downloaded from
+        published under the permissive MIT license. Source code and documentation can be downloaded from
         \href{https://github.com/pangenome/odgi}{https://github.com/pangenome/odgi}.\\
@@
         % \href{andreaguarracino@outlook.com}{andreaguarracino@outlook.com},
-        egarris5@uthcs.edu\\
-        \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online. \FIXME}
-
+        egarris5@uthcs.edu
+        % \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online. \FIXME
+        }
     \maketitle
@@
 
-    Thanks to advances in sequencing technologies, new genome
-    assemblies are produced at a high rate, including the human
-    pangenome reference consortium (HPRC) and telomere-to-telemere
-    efforts, e.g.~\citep{Miga:2020}.  A pangenome models the full set
-    of genomic elements in a given species or clade. Pangenomics
-    therefore contrasts with reference genome based approaches which
-    relate sequences to a particular consensus model of the
-    genome~\citep{Eizenga:2020}. A pangenome can elegantly be
-    represented by a graph data structure incorporating sequences as
-    nodes and their connections as edges. These nodes are shared for
-    homologs, paralogs and orthologs.  A `bidirected' graph represents
-    both strands of DNA and so-called paths represent contigs that run
-    through the nodes. A path can represent an individual chromosome
-    or part thereof. Such a graph-based approach is particularly
-    relevant for highly repetitive regions. Segmental duplications,
-    and acrocentric chromosomes lead to graphs which present very high
-    node depths which greatly increases processing complexity.
+    Thanks to advances in sequencing technology, new full-length
+    genome assemblies are being produced at a high rate, including
+    efforts from the Human Pangenome Reference Consortium (HPRC) and
+    telomere-to-telemere projects, e.g.~\citep{Miga:2020}.  In
+    principle, a pangenome can model a full set of genomic elements of
+    individuals in a given species or clade. Pangenomics contrasts
+    with reference genome based approaches which relate sequences to a
+    single consensus model of the genome~\citep{Eizenga:2020}. A
+    pangenome can elegantly be represented by a graph data structure
+    incorporating sequences as nodes and their connections as
+    edges. These nodes are shared for homologs, paralogs and
+    orthologs. So-called paths represent contigs that run through the
+    nodes. A path can represent an individual chromosome or part
+    thereof.  A bi-directed graph can even contain both strands of DNA
+    as paths. A graph-based approach is particularly relevant for
+    exploring complex variation and highly repetitive regions.
 
@@
     processing~\citep{vgtools,Eizenga:2020b}, but was not designed for
-    large data (see Table~\ref{tab:02}). With \odgi\ we
+    large data (see Table~\ref{tab:02}). We
     wrote a new toolset and API in \Cpp\ named Optimized Dynamic
     Genome/Graph Implementation (ODGI). \odgi\ implements a efficient
-    graph structure in RAM that can be dynamically updated using
-    multiple CPUs.
+    graph structure in computer memory (RAM) that can be dynamically updated using
+    multiple CPU cores in parallel.
 
@@
     paths. Most of the tools are designed to be applied together,
-    piping the output of one tool into the next.
+    piping the output from one tool into the next.
 
-    \vspace{-0.1in}
+    % \vspace{-0.1in}
 
@@
     convert graphs from GFA to the \odgi\ format and \textit{vice versa}.
-    \cmd{odgi validate} ascertains the graph data is correct, such as
-    validating that paths respect the graph's topology.
+    \cmd{odgi validate} ascertains the graph data is correct, e.g.,
+    validate paths respect the graph topology.
 
@@
     capable of handling full length human chromosomes
-    (Fig.~\ref{fig:1}a). \cmd{odgi sort} reorganises node order and
+    (see Fig.~\ref{fig:1}a). \cmd{odgi sort} reorganises node order and
     simplifies the graph by applying several algorithms, including the
-    novel `Path-Guided Stochastic Gradient Descent' algorithm.
+    novel `path-guided stochastic gradient descent' algorithm.
     \cmd{odgi layout} generates a file with the X-Y coordinates for
@@
     system from a contained reference genome --- a dynamic liftover
-    --- to display coordiates and other localised features, as is shown in
-    Fig.~\ref{fig:1b}.
+    --- to display coordinates and other localised features, as is shown in
+    Fig.~\ref{fig:1b}. Note that multiple reference genomes can be contained
+    in the graph and any contained path can be a reference.
+
+    \section{Discussion}
+
+    Pangenome graphs stand to become a ubiquitous tool in
+    genomics~\citep{Eizenga:2020}.  With \odgi\ we implemented a
+    state-of-the-art tool suite, that can transform, analyse and
+    visualise pangenome graphs at scale.  \odgi\ also exposes an API
+    that can be used by other tools, e.g., for genotyping, pangenome
+    graph building and normalization. Future work will add support for
+    RNA and protein sequences and expand on metadata capabilities of
+    large pangenome graphs.
+
+%\begin{figure}[!tpb]%figure2
+%%\centerline{\includegraphics{fig02.eps}}
+%\caption{Caption, caption.}\label{fig:02}
+%\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%     please remove the " % " symbol from \centerline{\includegraphics{fig01.eps}}
+%     as it may ignore the figures.
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+    % \vspace*{+12pt}
+    % \textit{Financial Support}: none declared.
+
+    \textit{Conflict of Interest}: none declared.
+    \vspace*{+24pt}
 
@@
             44~individuals plus the chm13 cell line and GRCh38
-            reference genomes stored in 4.3GB~GFA format\citep{GFA}.
+            reference genomes stored in 4.3GB~GFA format~\citep{GFA}.
             %
@@
 
-    \section{Discussion}
-
-    Pangenome graphs stand to become a ubiquitous tool in
-    genomics\citep{Eizenga:2020}.  With \odgi\ we implemented a
-    state-of-the-art tool suite, that can transform, analyse and
-    visualise pangenome graphs at scale.  \odgi\ also exposes an API
-    that is used by other tools for pangenome graph building and
-    normalization. Future work will add support for RNA and protein
-    sequences and expand on metadata capabilities of large pangenome
-    graphs.
-
-%\begin{figure}[!tpb]%figure2
-%%\centerline{\includegraphics{fig02.eps}}
-%\caption{Caption, caption.}\label{fig:02}
-%\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%     please remove the " % " symbol from \centerline{\includegraphics{fig01.eps}}
-%     as it may ignore the figures.
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-    % \vspace*{+12pt}
-    % \textit{Financial Support}: none declared.
-
-    \textit{Conflict of Interest}: none declared.
-    \vspace*{+24pt}
-
     % \section*{Data availability}

### 2021-06-02 manuscript/main.tex
edits

@@
 \newcommand{\FIXME}{\textbf{!!FIXME!!}}
-\newcommand{\cmd}[1]{\textbf{\textrm{#1}}}
+\newcommand{\cmd}[1]{\textrm{#1}}
 \newcommand{\cmdbf}[1]{{\textbf{#1}}}
@@
         exploratory analysis, sorting, removing artifacts, and graph
-        manipulation and visualisation. \odgi\ also provides an API that allows
+        manipulation and visualisation. \odgi\ also provides an extensible API that allows
         for writing new tool.
@@
         egarris5@uthcs.edu
-        % \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online. \FIXME
+        \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online. \FIXME
         }
@@
     edges. These nodes are shared for homologs, paralogs and
-    orthologs. So-called paths represent contigs that run through the
-    nodes. A path can represent an individual chromosome or part
-    thereof.  A bi-directed graph can even contain both strands of DNA
+    orthologs. So-called paths represent reads, contigs, haplotypes, individual chromosomes or part thereof, that run through the
+    nodes.   A bi-directed graph can even contain both strands of DNA
     as paths. A graph-based approach is particularly relevant for
@@
     human genomes. The \vg\ toolkit pioneered graph
-    processing~\citep{vgtools,Eizenga:2020b}, but was not designed for
+    processing~\citep{vgtools,Eizenga:2020b}, but does not scale for
     large data (see Table~\ref{tab:02}). We
@@
     Genome/Graph Implementation (ODGI). \odgi\ implements a efficient
-    graph structure in computer memory (RAM) that can be dynamically updated using
+    graph structure in computer memory that can be dynamically updated using
     multiple CPU cores in parallel.
@@
     transform~\citep{Siren:2020}. For implementation details see the
-    \odgi\ online documentation. \odgi\ includes the following tools:
+    \odgi\ online documentation. \odgi\ currently comes with over 30
+    tools, including:
 
@@
     \cmd{odgi validate} ascertains the graph data is correct, e.g.,
-    validate paths respect the graph topology.
+    it validates paths with respect the graph topology.
 
@@
     state-of-the-art tool suite, that can transform, analyse and
-    visualise pangenome graphs at scale.  \odgi\ also exposes an API
+    visualise pangenome graphs at scale.  \odgi\ also exposes an extensible API
     that can be used by other tools, e.g., for genotyping, pangenome
@@
             %
-            First GFS was converted from GFA to \odgi\ and
-            \vg\ formats, respectively.  \odgi's path implementation
+            First GFA was to converted to xg and next to \odgi\ and
+            \vg\ formats respectively.  \odgi's path implementation
             can process paths in parallel, outperforming \vg. For the

### 2021-06-02 manuscript/main.tex
reduced the noise

@@
 
-    \corresp{$^\ast$To whom correspondence should be addressed.}
+    \corresp{$^\dagger$To whom correspondence should be addressed.}
 
@@
     stored in the textual Graphical Fragment Assembly
-    (GFAv1) format\citep{GFA}; \cmd{odgi build} and \cmd{odgi view}
+    (GFAv1) format\citep{GFA}. \cmd{odgi build} and \cmd{odgi view}
     convert graphs from GFA to the \odgi\ format and \textit{vice versa}.
@@
     novel `path-guided stochastic gradient descent' algorithm.
-    \cmd{odgi layout} generates a file with the X-Y coordinates for
-    all nodes and \cmd{odgi draw} draws a layout of the graph
-    (Fig.~\ref{fig:1}b).
+    \cmd{odgi layout} generates a layout file with the X-Y coordinates for
+    all nodes and \cmd{odgi draw} draws it, bringing the pangenome representation
+    in two dimensions.
 
@@
     counts provide insight into genome complexity in other
-    ways. E.g., \cmd{odgi stats} returns the number of nodes,
+    ways. \cmd{odgi stats} returns the number of nodes,
     edges, paths, and graph length. \cmd{odgi bin} summarises path
@@
 
-    \topic{odgi explode, squeeze \& extract} pangenome are constructed
-    as a large graph. \cmd{odgi explode} separates units, such as
+    \topic{odgi explode, squeeze \& extract} pangenomes are constructed
+    as large graphs. \cmd{odgi explode} separates units, such as
     chromosomes, into different files.  \cmd{odgi squeeze} merges

### 2021-06-02 manuscript/figure_p.tex
put odgi build in the image caption in a better place

@@
             alignment. The black lines at the bottom represent the graph topology.
-            (b) consensus graph representation by \cmd{odgi build} of the same assemblies showing variations larger than $100$ base pairs.
+            (b) consensus graph representation of the same assemblies showing variations larger than $100$ base pairs.
             Loops or `bubbles' display contigs uniquely diverging from the consensus, caused by variation in paralogs/repeats.
-            The gene labels are super imposed by \cmd{odgi position} from a GRCh38 BED file and visualised by the Bandage tool~\citep{Wick:2015}.
+            The gene labels are super imposed by \cmd{odgi build+position} from a GRCh38 BED file and visualised by the
+            Bandage tool~\citep{Wick:2015}.
         }

### 2021-06-02 manuscript/main.tex
edit abstract

@@
         representation of DNA graphs. \odgi\ can handle hundreds of
-        large genomes in memory with fast
-        parallel execution.  \odgi\ includes
-        tools for detecting complex regions, extracting \textit{loci},
-        exploratory analysis, sorting, removing artifacts, and graph
-        manipulation and visualisation. \odgi\ also provides an extensible API that allows
-        for writing new tool.
+        large genomes in memory with fast parallel execution. \odgi\ includes
+        tools for detecting complex regions, extracting \textit{loci}, removing artifacts,
+        exploratory analysis, and graph sorting, manipulation and visualisation.
+        \odgi\ also provides an extensible API that allows for writing new tool.
         %
@@
     Thanks to advances in sequencing technology, new full-length
-    genome assemblies are being produced at a high rate, including
+    genome assemblies are produced at a high rate, including
     efforts from the Human Pangenome Reference Consortium (HPRC) and

### 2021-06-02 manuscript/main.tex
the Fig 1 doesn't shot a full length chromosome; pills of biology; little edits

@@
         egarris5@uthcs.edu
+        \\
         \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online. \FIXME
@@
     edges. These nodes are shared for homologs, paralogs and
-    orthologs. So-called paths represent reads, contigs, haplotypes, individual chromosomes or part thereof, that run through the
+    orthologs. So-called paths can represent reads, contigs, haplotypes,
+    individual chromosomes or part thereof, that run through the
     nodes. A bi-directed graph can even contain both strands of DNA
     as paths. A graph-based approach is particularly relevant for
-    exploring complex variation and highly repetitive regions.
+    exploring complex variation and highly repetitive regions present
+    which characterises the vertebrate genomes.
 
@@
     large data (see Table~\ref{tab:02}). We
-    wrote a new toolset and API in \Cpp\ named Optimized Dynamic
-    Genome/Graph Implementation (ODGI). \odgi\ implements a efficient
-    graph structure in computer memory that can be dynamically updated using
-    multiple CPU cores in parallel.
+    wrote a new toolset and Application Programming Interface (API)
+    in \Cpp\ named Optimized Dynamic Genome/Graph Implementation (ODGI).
+    \odgi\ implements a efficient graph structure in computer memory
+    that can be dynamically updated using multiple CPU cores in parallel.
 
@@
     provides convenient insight into genomic variation. \cmd{odgi viz}
-    generates a linearized representation of the pangenome and is
-    capable of handling full length human chromosomes
-    (see Fig.~\ref{fig:1}a). \cmd{odgi sort} reorganises node order and
-    simplifies the graph by applying several algorithms, including the
-    novel `path-guided stochastic gradient descent' algorithm.
-    \cmd{odgi layout} generates a layout file with the X-Y coordinates for
-    all nodes and \cmd{odgi draw} draws it, bringing the pangenome representation
-    in two dimensions.
+    generates a linearized representation of the pangenome (see Fig.~\ref{fig:1}a)
+    and is capable of handling full length human chromosomes. \cmd{odgi sort}
+    reorganises node order and simplifies the graph by applying several
+    algorithms, including the novel 'path-guided stochastic gradient descent'
+    algorithm. \cmd{odgi layout} generates a layout file with the X-Y
+    coordinates for all nodes and \cmd{odgi draw} draws it, bringing
+    the pangenome representation in two dimensions.
 
@@
     \cmd{odgi degree} return the node depth and degree as defined by
-    user provided criteria. These methods allow detection of complex
-    genomic regions.
+    user provided criteria. These methods allow the detection of complex
+    genomic regions generated by highly repetitive sequences present in
+    centromeres, segmental duplications, and acrocentric chromosomes.
 
@@
     reorganised. \cmd{odgi break} removes circular structures in the
-    graph, thereby reducing complexity. \cmd{odgi groom} removes
+    graph, thereby reducing its complexity. \cmd{odgi groom} removes
     spurious inverted links by aligning the graph from the orientation
@@
     Fig.~\ref{fig:1b}. Note that multiple reference genomes can be contained
-    in the graph and any contained path can be a reference.
+    in the graph and any contained path can be used as a reference.
 
@@
     genomics~\citep{Eizenga:2020}. With \odgi\ we implemented a
-    state-of-the-art tool suite, that can transform, analyse and
-    visualise pangenome graphs at scale.  \odgi\ also exposes an extensible API
-    that can be used by other tools, e.g., for genotyping, pangenome
-    graph building and normalization. Future work will add support for
-    RNA and protein sequences and expand on metadata capabilities of
-    large pangenome graphs.
+    state-of-the-art tool suite that can transform, analyse and
+    visualise pangenome graphs at large scale. \odgi\ also exposes an
+    extensible API that can be used by other tools, e.g., for genotyping,
+    pangenome graph building \citep{pggb} and normalization \citep {smoothxg}.
+    Future work will add support for RNA and protein sequences and
+    expand on metadata capabilities of large pangenome graphs.
 

### 2021-06-02 manuscript/main.tex
Removed extra citations

@@
     visualise pangenome graphs at large scale. \odgi\ also exposes an
-    extensible API that can be used by other tools, e.g., for genotyping,
-    pangenome graph building \citep{pggb} and normalization \citep {smoothxg}.
-    Future work will add support for RNA and protein sequences and
-    expand on metadata capabilities of large pangenome graphs.
+    extensible API that can be used by other tools, e.g., for
+    genotyping, pangenome graph building and normalization.  Future
+    work will add support for RNA and protein sequences and expand on
+    metadata capabilities of large pangenome graphs.
 

### 2021-06-02 manuscript/main.tex
fixes

@@
     Thanks to advances in sequencing technology, new full-length
-    genome assemblies are produced at a high rate, including
-    efforts from the Human Pangenome Reference Consortium (HPRC) and
+    genome assemblies are produced at a high rate, including efforts
+    from the Human Pangenome Reference Consortium (HPRC) and
     telomere-to-telemere projects, e.g.~\citep{Miga:2020}. In
@@
     incorporating sequences as nodes and their connections as
-    edges. These nodes are shared for homologs, paralogs and
-    orthologs. So-called paths can represent reads, contigs, haplotypes,
-    individual chromosomes or part thereof, that run through the
-    nodes. A bi-directed graph can even contain both strands of DNA
-    as paths. A graph-based approach is particularly relevant for
-    exploring complex variation and highly repetitive regions present
-    which characterises the vertebrate genomes.
+    edges. These nodes are shared for identical sequences, such as
+    homologs, paralogs and orthologs. So-called paths can represent
+    reads, contigs, haplotypes, individual chromosomes or part
+    thereof, that run through the nodes. A bi-directed graph can even
+    contain both strands of DNA as paths. A graph-based approach is
+    particularly relevant for exploring complex variation and highly
+    repetitive regions present which characterises the vertebrate
+    genomes.
 
@@
     human genomes. The \vg\ toolkit pioneered graph
-    processing~\citep{vgtools,Eizenga:2020b}, but does not scale for
-    large data (see Table~\ref{tab:02}). We
-    wrote a new toolset and Application Programming Interface (API)
-    in \Cpp\ named Optimized Dynamic Genome/Graph Implementation (ODGI).
+    processing~\citep{vgtools,Eizenga:2020b}, but we found it does not
+    necessarily scale for large data (see Table~\ref{tab:02}). We
+    wrote a new toolset and Application Programming Interface (API) in
+    \Cpp\ named Optimized Dynamic Genome/Graph Implementation (ODGI).
     \odgi\ implements a efficient graph structure in computer memory
-    that can be dynamically updated using multiple CPU cores in parallel.
+    that can be dynamically updated using multiple CPU cores in
+    parallel.
 
@@
 
-    % \vspace{-0.1in}
+    \input{figure_p}
 
-    \input{figure_p}
+    \vspace{-0.1in}
 
@@
     transform~\citep{Siren:2020}. For implementation details see the
-    \odgi\ online documentation. \odgi\ currently comes with over 30
-    tools, including:
+    \odgi\ online documentation. \odgi\ currently comes with more than 30
+    tools and here we present a subset of functionality:
 
@@
     stored in the textual Graphical Fragment Assembly
-    (GFAv1) format\citep{GFA}. \cmd{odgi build} and \cmd{odgi view}
+    format~\citep{GFA}. \cmd{odgi build} and \cmd{odgi view}
     convert graphs from GFA to the \odgi\ format and \textit{vice versa}.
     \cmd{odgi validate} ascertains the graph data is correct, e.g.,
-    it validates paths with respect the graph topology.
+    it validates paths with respect to the graph topology.
 
@@
     reorganises node order and simplifies the graph by applying several
-    algorithms, including the novel 'path-guided stochastic gradient descent'
+    algorithms, including our novel 'path-guided stochastic gradient descent'
     algorithm. \cmd{odgi layout} generates a layout file with the X-Y
@@
     \topic{odgi stats, bin, depth \& degree} pangenome numerical
-    counts provide insight into genome complexity in other
+    counts provide insight into complexity in other
     ways. \cmd{odgi stats} returns the number of nodes,
@@
     user provided criteria. These methods allow the detection of complex
-    genomic regions generated by highly repetitive sequences present in
-    centromeres, segmental duplications, and acrocentric chromosomes.
+    genomic regions generated by highly repetitive sequences.
 

### 2021-06-02 manuscript/main.tex
edits

@@
     particularly relevant for exploring complex variation and highly
-    repetitive regions present which characterises the vertebrate
-    genomes.
+    repetitive regions, such as the Major Histocompatibility Complex
+    (MHC, Fig.~\ref{fig:1}).
 
@@
 
-    \vspace{-0.1in}
+    \vspace{-0.15in}
 
@@
     convert graphs from GFA to the \odgi\ format and \textit{vice versa}.
-    \cmd{odgi validate} ascertains the graph data is correct, e.g.,
-    it validates paths with respect to the graph topology.
+    \cmd{odgi validate} ascertains the graph data is correct.
 
@@
     algorithm. \cmd{odgi layout} generates a layout file with the X-Y
-    coordinates for all nodes and \cmd{odgi draw} draws it, bringing
-    the pangenome representation in two dimensions.
+    coordinates for all nodes and \cmd{odgi draw} creates the figure.
 
@@
     summarised view of large genomes. \cmd{odgi depth} and
-    \cmd{odgi degree} return the node depth and degree as defined by
+    \cmd{odgi degree} compute node depth and degree as defined by
     user provided criteria. These methods allow the detection of complex
@@
 
+    \vspace{-0.1in}
+
     \section{Discussion}
@@
     metadata capabilities of large pangenome graphs.
+
+    \input{table}
 
@@
 
+    \textit{Financial Support}: \FIXME\ Garrison \& Prins.
     % \vspace*{+12pt}
-    % \textit{Financial Support}: none declared.
 
     \textit{Conflict of Interest}: none declared.
-    \vspace*{+24pt}
 
-        %TODO to repeat on a bigger machine to avoid vg crashing
-    \begin{table}[h!]
-        \caption{ Example of performance improvement between
-            \cmd{odgi} and \cmd{vg}~\citep{vgtools} equivalent
-            commands with output file sizes. The centromere region was
-            extracted from CHR 6 of the HPRC year one assembly, i.e.,
-            88~haploid, phased human genome assemblies from
-            44~individuals plus the chm13 cell line and GRCh38
-            reference genomes stored in 4.3GB~GFA format~\citep{GFA}.
-            %
-            First GFA was to converted to xg and next to \odgi\ and
-            \vg\ formats respectively.  \odgi's path implementation
-            can process paths in parallel, outperforming \vg. For the
-            same reason \cmd{odgi extract} is $20\times$ faster than
-            \cmd{vg chunk} extracting a portion from the full
-            graph. \FIXME: vg never finished.
-            %
-            All timings were performed on a server class machine
-            on SSD MODEL with X GB of RAM and 48 CPU cores (Y x 12 core
-        AMD Opteron) Processor MODEL @ 3.3 GHz with 8MB L2
-        Cache.\FIXME}
-        \begin{tabular}{{@{}lllll@{}}}
-          \toprule
-          Operation & Command & Runtime & Memory  & File size     \\
-                    &      & (mm:ss) & (GB) & (GB)           \\
-          \midrule
-          Covert GFA to native format                         &       &       &                   \\
-          & \cmd{odgi build}      & 1:35  & 10.39 & 5.4               \\
-          & \cmd{vg convert}      & 8:14  & 28.43 & 6.1               \\
-          Extract subgraph                         &       &       &                   \\
-          & \cmd{odgi extract}    & 1:15  & 9.43  & 2.7               \\
-          & \cmd{vg chunk}        & 21:09 & 59.33 & 1.7               \\
-          \botrule
-        \end{tabular}
-        \label{tab:02}
-
-    \end{table}
+    % \vspace*{+12pt}
 
@@
 %    \vspace*{-12pt}
+    \vspace{-0.15in}
 

### 2021-06-02 manuscript/table.tex
edits

        %TODO to repeat on a bigger machine to avoid vg crashing
    \begin{table}[h!]
        \caption{ Example of performance improvement between
            \cmd{odgi} and \cmd{vg}~\citep{vgtools} equivalent
            commands with output file sizes. The centromere region was
            extracted from CHR 6 of the HPRC year one assembly, i.e.,
            88~haploid, phased human genome assemblies from
            44~individuals plus the chm13 cell line and GRCh38
            reference genomes stored in 4.3GB~GFA format~\citep{GFA}.
            %
            First GFA was to converted to xg and next to \odgi\ and
            \vg\ formats respectively.  \odgi's path implementation
            can process paths in parallel, outperforming \vg. For the
            same reason \cmd{odgi extract} is $20\times$ faster than
            \cmd{vg chunk} extracting a portion from the full
            graph. \FIXME: vg never finished.
            %
            All timings were performed on a server class machine
            on SSD MODEL with X GB of RAM and 48 CPU cores (Y x 12 core
        AMD Opteron) Processor MODEL @ 3.3 GHz with 8MB L2
        Cache.\FIXME}
        \begin{tabular}{{@{}lllll@{}}}
          \toprule
          Operation & Command & Runtime & Memory  & File size     \\
                    &      & (mm:ss) & (GB) & (GB)           \\
          \midrule
          Covert GFA to native format                         &       &       &                   \\
          & \cmd{odgi build}      & 1:35  & 10.39 & 5.4               \\
          & \cmd{vg convert}      & 8:14  & 28.43 & 6.1               \\
          Extract subgraph                         &       &       &                   \\
          & \cmd{odgi extract}    & 1:15  & 9.43  & 2.7               \\
          & \cmd{vg chunk}        & 21:09 & 59.33 & 1.7               \\
          \botrule
        \end{tabular}
        \label{tab:02}

    \end{table}

### 2021-06-04 manuscript/main.tex
complete affiliation of SH

@@
     \address{$^{\text{\sf 1}}$Biology, University of Tor Vergata, Rome, 00133, Italy \\
-        $^{\text{\sf 2}}$QBiC, University of Tübingen, Tübingen, XXXXXXXXXX, Germany \\
+        $^{\text{\sf 2}}$Quantitative Biology Center (QBiC), University of Tübingen, Tübingen, 72076, Germany \\
         $^{\text{\sf 3}}$Dept. Genetics Genomics \& Informatics, UTHSC, Memphis, TN, USA

### 2021-06-04 manuscript/main.tex
added financial support of SH
more motivation&biology, less bold statements, no API

@@
       with regard to reference genome bias and identification of complex
-      structural variants.
+      structural variants. Hence, fast and versatile software is required to ask intricate questions of such data in an efficient way.
         %
-        \\ \textbf{Results:} Here we present \odgi\, a novel range of
-        scalable tools that implements new algorithms and has an efficient in-memory
-        representation of DNA graphs. \odgi\ can handle hundreds of
-        large genomes in memory with fast parallel execution. \odgi\ includes
-        tools for detecting complex regions, extracting \textit{loci}, removing artifacts,
-        exploratory analysis, and graph sorting, manipulation and visualisation.
-        \odgi\ also provides an extensible API that allows for writing new tool.
+        \\ \textbf{Results:} Here we present \odgi\ a novel range of scalable tools that implements new algorithms and has an efficient in-memory representation of DNA graphs.
+\odgi\ includes tools for detecting complex regions, extracting \textit{loci}, removing artifacts, exploratory analysis, and graph sorting, manipulation and visualisation.
+Its fast parallel execution facilitates routine pangenomic tasks as well as pipelines that can quickly answer complex biological questions of very large pangenome graphs.
         %
         \\
-        \textbf{Availability:} \odgi\ is software written in the \Cpp\ programming language and
-        published under the permissive MIT license. Source code and documentation can be downloaded from
+        \textbf{Availability:} \odgi\ is \Cpp\ software
+        published under the permissive MIT license. Source code and documentation are available at
         \href{https://github.com/pangenome/odgi}{https://github.com/pangenome/odgi}.\\

### 2021-06-04 manuscript/main.tex
add crucial T2T reference, rework Miga:2020 reference

--- 2021-06-04 manuscript/main.tex
@@
     from the Human Pangenome Reference Consortium (HPRC) and
-    telomere-to-telemere projects, e.g.~\citep{Miga:2020}. In
+    telomere-to-telemere projects, e.g.~\citep{Miga:2020, Logsdon:2021}. In
     principle, a pangenome can model a full set of genomic elements of

### 2021-07-08 manuscript/table.tex
typo

@@
           \midrule
-          Covert GFA to native format                         &       &       &                   \\
+          Convert GFA to native format                         &       &       &                   \\
           & \cmd{odgi build}      & 1:35  & 10.39 & 5.4               \\

### 2021-07-11 manuscript/main.tex
typos

--- 2021-06-04 manuscript/main.tex
@@
     from the Human Pangenome Reference Consortium (HPRC) and
-    telomere-to-telemere projects, e.g.~\citep{Miga:2020}. In
+    telomere-to-telomere projects, e.g.~\citep{Miga:2020}. In
     principle, a pangenome can model a full set of genomic elements of

### 2021-08-09 manuscript/main.tex
add manuscript template

@@
+\PassOptionsToPackage{utf8}{inputenc}
 \documentclass{bioinfo}
 
-\usepackage{relsize} % for Cpp
-\usepackage{xspace} % for Cpp
-\usepackage{subfigure}
+\usepackage[draft]{hyperref}
+\usepackage{makecell}
 
-\usepackage{natbib}
+\newcommand{\vocab}{\textbf}
 
-\newcommand{\odgi}{ODGI}
-\newcommand{\vg}{VG}
-
-\newcommand{\FIXME}{\textbf{!!FIXME!!}}
-\newcommand{\cmd}[1]{\textrm{#1}}
-\newcommand{\cmdbf}[1]{{\textbf{#1}}}
-\newcommand{\topic}[1]{{\cmdbf{#1}}:}
-\newcommand{\Rplus}{\protect\hspace{-.1em}\protect\raisebox{.35ex}{\smaller{\smaller\textbf{+}}}}
-\newcommand{\pp}{\Rplus\Rplus}
-\newcommand{\Cpp}{\mbox{C\Rplus\Rplus}\xspace}
-
-\copyrightyear{2021} \pubyear{2021}
+\copyrightyear{XXXX} \pubyear{XXXX}
 
 \access{Advance Access Publication Date: Day Month Year}
-\appnotes{Application Note}
+\appnotes{Genome Analysis}
 
@@
 
-    \subtitle{Pangenome analysis}
+\subtitle{Genome Analysis}
 
-    \title[ODGI: scalable tools for pangenome graphs]{ODGI: scalable tools for pangenome graphs}
-    \author[Guarracino, Heumos \textit{et~al}.]{
-        Andrea Guarracino\,$^{\text{\sfb 1,}*}$,
-        Simon Heumos\,$^{\text{\sfb 2,}*}$,
-    % XXXXX,
-        Pjotr~Prins\,$^{\text{\sfb 3}}$ and
-        Erik~Garrison\,$^{\text{\sfb 3,}\dagger}$
-        \footnote{The authors wish it to be known that the first two authors should be regarded as joint first authors.}
-    }
-    \address{$^{\text{\sf 1}}$Biology, University of Tor Vergata, Rome, 00133, Italy \\
-        $^{\text{\sf 2}}$Quantitative Biology Center (QBiC), University of Tübingen, Tübingen, 72076, Germany \\
-        $^{\text{\sf 3}}$Dept. Genetics Genomics \& Informatics, UTHSC, Memphis, TN, USA
+\title[short Title]{ODGI: scalable tools for working with pangenome graphs}
+\author[Heumos, \textit{et~al}.]{
+Simon Heumos\,$^{\text{\sfb 1, \sfb 2} \dagger}$,
+Andrea Guarracino\,$^{\text{\sfb 2} \dagger}$,
+Pjotr Prins\,$^{\text{\sfb 3}}$,
+and Erik Garrison\,$^{\text{\sfb 3}*}$
 }
 
-    \corresp{$^\dagger$To whom correspondence should be addressed.}
+\address{
+$^{\text{\sf 1}}$Quantitative Biology Center (QBiC), University of T\"ubingen, T\"ubingen, Germany, 72076
+$^{\text{\sf 2}}$University of Tor Vergata, Rome, Italy and \\
+$^{\text{\sf 3}}$University of Tennessee Health Science Center, Memphis, TN, USA
+}
+
+\corresp{$^\ast$To whom correspondence should be addressed. \\
+$^\dagger$Contributed equally.}
 
@@
 
-    \abstract{ \textbf{Motivation:} Pangenomes graphs address some of the
-      shortcomings of mainline genomics, in particular
-      with regard to reference genome bias and identification of complex
-      structural variants.
-        %
-        \\ \textbf{Results:} Here we present \odgi\, a novel range of
-        scalable tools that implements new algorithms and has an efficient in-memory
-        representation of DNA graphs. \odgi\ can handle hundreds of
-        large genomes in memory with fast parallel execution. \odgi\ includes
-        tools for detecting complex regions, extracting \textit{loci}, removing artifacts,
-        exploratory analysis, and graph sorting, manipulation and visualisation.
-        \odgi\ also provides an extensible API that allows for writing new tool.
-        %
-        \\
-        \textbf{Availability:} \odgi\ is software written in the \Cpp\ programming language and
-        published under the permissive MIT license. Source code and documentation can be downloaded from
-        \href{https://github.com/pangenome/odgi}{https://github.com/pangenome/odgi}.\\
-        %
-        \textbf{Contact:}
-        % \href{andreaguarracino@outlook.com}{andreaguarracino@outlook.com},
-        egarris5@uthcs.edu
-        \\
-        \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online. \FIXME
+\abstract{
+\textbf{Motivation:} Pangenome graphs provide a complete representation of the mutual alignment of collections of genomes.
+These models, therefore, offer the opportunity to study the entire genetic diversity of a population, bringing new challenges to how to analyze hundreds of genomes on a gigabase scale.
+In addition, vertebrate genomes present highly repetitive regions which increase the complexity of the analysis.
+Efficient and flexible tools are therefore required to work on pangenome graphs of any complexity and scale. \\
+\textbf{Results:} Pangenome graphs provide a complete representation of the mutual alignment of collections of genomes.
+These models, therefore, offer the opportunity to study the entire genetic diversity of a population, bringing new challenges to how to analyze hundreds of genomes on a gigabase scale.
+In addition, vertebrate genomes present highly repetitive regions which increase the complexity of the analysis.
+Efficient and flexible tools are therefore required to work on pangenome graphs of any complexity and scale. \\
+\textbf{Availability:} ODGI is written in C++. Its source code is freely available at \url{https://github.com/pangenome/odgi} with documentation at \url{https://odgi.readthedocs.io}.
+ODGI can be installed via Bioconda \url{https://bioconda.github.io/recipes/odgi/README.html} or Guix https://github.com/ekg/guix-genomics/blob/master/odgi.scm. \\
+\textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
+\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
 }
+
 \maketitle
@@
 
-    Thanks to advances in sequencing technology, new full-length
-    genome assemblies are produced at a high rate, including efforts
-    from the Human Pangenome Reference Consortium (HPRC) and
-    telomere-to-telomere projects, e.g.~\citep{Miga:2020}. In
-    principle, a pangenome can model a full set of genomic elements of
-    individuals in a given species or clade. Pangenomics contrasts
-    with reference genome based approaches which relate sequences to a
-    single consensus model of the genome~\citep{Eizenga:2020}. A
-    pangenome can elegantly be represented by a graph data structure
-    incorporating sequences as nodes and their connections as
-    edges. These nodes are shared for identical sequences, such as
-    homologs, paralogs and orthologs. So-called paths can represent
-    reads, contigs, haplotypes, individual chromosomes or part
-    thereof, that run through the nodes. A bi-directed graph can even
-    contain both strands of DNA as paths. A graph-based approach is
-    particularly relevant for exploring complex variation and highly
-    repetitive regions, such as the Major Histocompatibility Complex
-    (MHC, Fig.~\ref{fig:1}).
+\cite{Eizenga_2020}
 
-    A major challenge is writing software that can deal with the sheer
-    size of these graphs representing hundreds, and soon thousands, of
-    human genomes. The \vg\ toolkit pioneered graph
-    processing~\citep{vgtools,Eizenga:2020b}, but we found it does not
-    necessarily scale for large data (see Table~\ref{tab:02}). We
-    wrote a new toolset and Application Programming Interface (API) in
-    \Cpp\ named Optimized Dynamic Genome/Graph Implementation (ODGI).
-    \odgi\ implements a efficient graph structure in computer memory
-    that can be dynamically updated using multiple CPU cores in
-    parallel.
+\begin{methods}
 
-    Note that \odgi\ is focused on `first-order' graph
-    operations. Other tools are required to construct graphs from raw
-    sequences, map reads, or call variants. \odgi\ provides
-    operations on an existing graph, such as subset, subdivide, break,
-    combine, normalize and order its components, nodes, and
-    paths. Most of the tools are designed to be applied together,
-    piping the output from one tool into the next.
+\section{Implementation}
 
-    \input{figure_p}
+\subsection{Data model}
 
-    \vspace{-0.15in}
+\subsection{The \textsc{HandleGraph} interface}
 
-    \section{Results}
+\subsection{Code availability}
 
-    \odgi\ is designed to build and modify paths in parallel. It
-    applies concepts first introduced in the dynamic version of the
-    graph extension of the positional Burrows-Wheeler
-    transform~\citep{Siren:2020}. For implementation details see the
-    \odgi\ online documentation. \odgi\ currently comes with more than 30
-    tools and here we present a subset of functionality:
+\end{methods}
 
-    \indent \topic{odgi build, view \& validate} pangenome graphs can be
-    stored in the textual Graphical Fragment Assembly
-    format~\citep{GFA}. \cmd{odgi build} and \cmd{odgi view}
-    convert graphs from GFA to the \odgi\ format and \textit{vice versa}.
-    \cmd{odgi validate} ascertains the graph data is correct.
+\section{Evaluation}
 
-    \topic{odgi viz, sort, draw \& layout} pangenome visualisation
-    provides convenient insight into genomic variation. \cmd{odgi viz}
-    generates a linearized representation of the pangenome (see Fig.~\ref{fig:1}a)
-    and is capable of handling full length human chromosomes. \cmd{odgi sort}
-    reorganises node order and simplifies the graph by applying several
-    algorithms, including our novel 'path-guided stochastic gradient descent'
-    algorithm. \cmd{odgi layout} generates a layout file with the X-Y
-    coordinates for all nodes and \cmd{odgi draw} creates the figure.
+\subsection{Human genome with structural variants}
 
-    \topic{odgi stats, bin, depth \& degree} pangenome numerical
-    counts provide insight into complexity in other
-    ways. \cmd{odgi stats} returns the number of nodes,
-    edges, paths, and graph length. \cmd{odgi bin} summarises path
-    information into bins of a specified size, thereby enabling a
-    summarised view of large genomes. \cmd{odgi depth} and
-    \cmd{odgi degree} compute node depth and degree as defined by
-    user provided criteria. These methods allow the detection of complex
-    genomic regions generated by highly repetitive sequences.
+\subsection{Genome graph collection}
 
-    \topic{odgi break, groom, chop \& unchop} pangenome graphs can be
-    reorganised. \cmd{odgi break} removes circular structures in the
-    graph, thereby reducing its complexity. \cmd{odgi groom} removes
-    spurious inverted links by aligning the graph from the orientation
-    that is supported by most paths. \cmd{odgi chop} divides long
-    nodes into shorter ones at a maximum requested size, thereby
-    simplifying downstream analysis. \cmd{odgi unchop} joins nodes
-    and embedded sequences that do not change the graph topology,
-    thereby obtaining a more compact representation of the graph.
-
-    \topic{odgi explode, squeeze \& extract} pangenomes are constructed
-    as large graphs. \cmd{odgi explode} separates units, such as
-    chromosomes, into different files. \cmd{odgi squeeze} merges
-    multiple graphs into the same file whilst preventing node ID
-    collisions. \cmd{odgi extract} extracts regions of the graph as
-    defined by certain criteria, allowing downstream processing of
-    smaller subgraphs.
-
-    \topic{odgi position} pangenome graphs are flexible when it comes
-    to coordinate systems. \cmd{odgi position} can use the coordinate
-    system from a contained reference genome --- a dynamic liftover
-    --- to display coordinates and other localised features, as is shown in
-    Fig.~\ref{fig:1b}. Note that multiple reference genomes can be contained
-    in the graph and any contained path can be used as a reference.
-
-    \vspace{-0.1in}
+\subsection{1000 Genome Project chromosome graphs}
 
@@
 
-    Pangenome graphs stand to become a ubiquitous tool in
-    genomics~\citep{Eizenga:2020}. With \odgi\ we implemented a
-    state-of-the-art tool suite that can transform, analyse and
-    visualise pangenome graphs at large scale. \odgi\ also exposes an
-    extensible API that can be used by other tools, e.g., for
-    genotyping, pangenome graph building and normalization.  Future
-    work will add support for RNA and protein sequences and expand on
-    metadata capabilities of large pangenome graphs.
+\section*{Funding}
 
-    \input{table}
-
-%\begin{figure}[!tpb]%figure2
-%%\centerline{\includegraphics{fig02.eps}}
-%\caption{Caption, caption.}\label{fig:02}
-%\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%
-%     please remove the " % " symbol from \centerline{\includegraphics{fig01.eps}}
-%     as it may ignore the figures.
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-    \textit{Financial Support}: \FIXME\ Garrison \& Prins.
-    % \vspace*{+12pt}
-
-    \textit{Conflict of Interest}: none declared.
-
-    % \vspace*{+12pt}
-
-    % \section*{Data availability}
-
-    % Data used to build Human pangenome graphs is available at \href{https://github.com/human-pangenomics/HPP_Year1_Data_Freeze_v1.0}{https://github.com/human-pangenomics/HPP\_Year1\_Data\_Freeze\_v1.0}.
-
-    % \section*{Acknowledgements}
-    %ToDo
-    % XXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX.
-
-    % \section*{Funding}
-    %ToDo
-    % This work has been supported by the XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX XXXXXXXX.
-
-%    \vspace*{-12pt}
-    \vspace{-0.15in}
+%This work was supported, in part, by the National Institutes of Health (award numbers U01HG010961, U41HG010972, R01HG010485, 2U41HG007234, 5U54HG007990, 5T32HG008345-04, U01HL137183 to B.P.) and the W.
+M.
+Keck Foundation (award number DT06172015 to B.P.).
+S.H. acknowledges funding from the Central Innovation Programme (ZIM) for SMEs of the Federal Ministry for Economic Affairs and Energy of Germany.
 
@@
 %
-%\bibliography{Document}
+\bibliography{document}
 
-\bibliography{bibliography}
+% \begin{thebibliography}{}
 
+% \bibitem[Bofelli {\it et~al}., 2000]{Boffelli03}
+% Bofelli,F., Name2, Name3 (2003) Article title, {\it Journal Name}, {\bf 199}, 133-154.
+
+% \bibitem[Bag {\it et~al}., 2001]{Bag01}
+% Bag,M., Name2, Name3 (2001) Article title, {\it Journal Name}, {\bf 99}, 33-54.
+
+% \bibitem[Yoo \textit{et~al}., 2003]{Yoo03}
+% Yoo,M.S. \textit{et~al}. (2003) Oxidative stress regulated genes
+% in nigral dopaminergic neurnol cell: correlation with the known
+% pathology in Parkinson's disease. \textit{Brain Res. Mol. Brain
+% Res.}, \textbf{110}(Suppl. 1), 76--84.
+
+% \bibitem[Lehmann, 1986]{Leh86}
+% Lehmann,E.L. (1986) Chapter title. \textit{Book Title}. Vol.~1, 2nd edn. Springer-Verlag, New York.
+
+% \bibitem[Crenshaw and Jones, 2003]{Cre03}
+% Crenshaw, B.,III, and Jones, W.B.,Jr (2003) The future of clinical
+% cancer management: one tumor, one chip. \textit{Bioinformatics},
+% doi:10.1093/bioinformatics/btn000.
+
+% \bibitem[Auhtor \textit{et~al}. (2000)]{Aut00}
+% Auhtor,A.B. \textit{et~al}. (2000) Chapter title. In Smith, A.C.
+% (ed.), \textit{Book Title}, 2nd edn. Publisher, Location, Vol. 1, pp.
+% ???--???.
+
+% \bibitem[Bardet, 1920]{Bar20}
+% Bardet, G. (1920) Sur un syndrome d'obesite infantile avec
+% polydactylie et retinite pigmentaire (contribution a l'etude des
+% formes cliniques de l'obesite hypophysaire). PhD Thesis, name of
+% institution, Paris, France.
+
+% \end{thebibliography}
 \end{document}

### 2021-08-09 manuscript/main.tex
fill the template

@@
 \usepackage{makecell}
+\usepackage{comment}
 
@@
 
+A pangenome models the full set of genomic elements in a given species or clade {32453966}.
+Indeed, these data structures encode the mutual relationships between all the genomes represented, in contrast to reference-based approaches which relate sequences only to a particular genome, chosen as reference.
+A class of methods to represent pangenomes involves the sequence graphs {2488477}, where similar orthologous regions between genomes collapse into a single representative sequence.
+In node-labeled sequence graphs, nodes indicate DNA sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph.
+A bidirected sequence graph represents both strands of DNA.
+On this model, variation graphs add the concept of path to embed linear sequences (e.g., genomes or haplotypes) into the graph {30125266}.
+Paths provide a stable coordinate system, allowing graph annotations and interoperability between different graphs.
+
+Thanks to advances in sequencing technologies, new genome assemblies are produced at a high rate.
+This amount of data offers the opportunity to study genomic variation as never before, but it also brings challenges in how to represent and manipulate hundreds of genomes at a gigabase scale.
+Variation graphs compactly represent the full genetic variation across a population, but graph-based data structures present computational overheads: in addition to the sequences, they must also represent the graph topology.
+Several genome data structures were developed to address these challenges {33040146}.
+Memory-efficient representations do not allow dynamic updates to the graph, such as adding new genomes or simplifying the graph topology.
+Furthermore, dynamic updates must perform efficiently on the complex graphs generated from whole-genome alignment of vertebrates.
+Highly repetitive regions, such as centromeres, segmental duplications, and acrocentric chromosomes lead to graphs that present paths with very high-depth nodes.
+The node depth is defined as the number of times in which the node is crossed by all the paths embedded in the graph.
+High depth increases the complexity of the operations performed on the paths and the nodes they cross.
+
+To overcome this limitation, we have developed an Optimized Dynamic Genome/Graph Implementation (ODGI).
+Each tool loads graphs into the succinct dynamic HandleGraph model {33040146} and applies methods based on the HandleGraph API.
+Several algorithms are shared with the VG toolkit (VG) {30125266}, the state-of-the-art toolkit for working with pangenome graphs.
+However, although similar in scope to VG, ODGI importantly differs in its design to support highly complex pangenome graphs.
+The majority of the ODGI tools are implemented in an index-free manner, avoiding the burning task needed to create an index structure.
+Indeed, index-constrained implementations would require building indexes also for single or little operations on the graphs.
+Moreover, thanks to its efficient path representation, ODGI tools can build and walk the internal graph representation (the ODGI format) in parallel, which is one of the major bottlenecks when working with very large and deep variation graphs, efficiently scaling up to hundreds of large vertebrate genomes.
+
 \cite{Eizenga_2020}
@@
 
-\subsection{Data model}
-
-\subsection{The \textsc{HandleGraph} interface}
-
-\subsection{Code availability}
+ODGI is designed to build and modify paths in parallel, keeping its representation in memory efficient.
+It applies concepts first introduced in the dynamic version of the Graph BWT {31406990} (gBWT), a scalable implementation of the graph extension of the positional Burrows-Wheeler transform {28702075}.
+Specifically, we build a node-local model of the graph, where the graph is stored in a vector of node structures.
+The identifier of each node is its offset in this vector plus 1.
+Each structure contains the sequence of the node, its edges in both directions (forward and backward), and a vector of path steps that describes how the embedded paths walk across the node.
+Each step records the path identifier and the previous and next path steps on the same path.
+To reduce the encoding, we encode path steps using a local alphabet that maps the N neighbors of the node into the range 1..N.
+To further save space, node deltas, rather than IDs, are stored in this alphabet.
+The delta between two nodes is defined as their distance in the graph vector (i.e., the difference between the node offsets).
+The path step vector, of length n, is stored using a dynamic succinct integer vector that requires O(5nw) bits, where w is ~log2(N).
+Finally, to allow operating on path steps in parallel, each node structure includes a byte-width mutex lock.
 
@@
 
-\section{Evaluation}
+\section{Results}
 
-\subsection{Human genome with structural variants}
+ODGI is entirely focused on “first-order” operations on the graph (Table 1), providing, unlike VG, no method to construct graphs from raw sequences or map reads.
+It supports interrogation of the graph for basic dimensional properties, positional and path-related queries, and visualization in one-dimension (1D) and two-dimensions (2D).
+ODGI also supports operations on the graph such as subset, subdivide, break, combine, normalize, or order its components, nodes, and paths.
+Most of the tools are designed to be applied together, piping the output of one tool into the next, thereby preventing the creation of intermediate files, and reducing the number of IO operations.
 
-\subsection{Genome graph collection}
+\subsection{Core functionality}
 
-\subsection{1000 Genome Project chromosome graphs}
+Variation graphs are commonly represented by a subset of version 1 of the Graphical Fragment Assembly format (GFAv1): odgi build and odgi view allow to convert graphs from GFA to ODGI format and vice versa, respectively.
+In the variation graph model, paths have to respect the graph’s topology: this can be verified with odgi validate, to ensure no errors in the input or edited graphs.
+
+By visualizing pangenome graphs, we can gain insight into the relationship between genomes.
+odgi viz applies binning and direct rendering to a raster image to generate a 1D representation able to scale up to gigabase pangenome graphs (Figure 1.A).
+However, pangenome graphs may present complex structures to visualize: odgi sort finds the best node order to simplify 1D graph visualization by applying several cutting-edge sorting algorithms.
+odgi draw extends the visualization in 2D, taking in input the layouts built by odgi layout.
+
+Applying odgi stats, users can retrieve metrics describing the graph properties, such as the number of nodes, edges, paths, and graph length.
+ODGI also offers more advanced tools for interrogating the graphs.
+odgi bin summarizes the path information into bins of specified size, enabling a summarized view of gigabase scale graphs.
+odgi depth returns the node depth as defined by query criteria, allowing users to detect the complex regions in the graph due to sequences with highly identical repeats.
+Complex motifs can be also detected with odgi degree, which returns the node degree as defined by query criteria.
+High degree nodes are the mirror of misassemblies or problems in the pangenome building, making the tool useful for debugging the process as well.
+
+\begin{comment}
+
+% the table
+
+Table 1 - A selection of the tools available in ODGI
+Name
+Description
+For conversion
+
+build
+Convert graphs from GFA to ODGI format.
+view
+Convert graphs from ODGI to GFA format.
+For validation
+
+validate
+Check if the graph is consistent.
+For interrogation
+
+stats
+Return graph metrics.
+bin
+Summarize path information in bins of fixed size.
+depth
+Return node depth as defined by query criteria.
+degree
+Return node degree as defined by query criteria.
+For manipulation
+
+break
+Break cycles in the graph, reducing its complexity.
+groom
+Resolve spurious inverting links.
+chop
+Chop long nodes into short ones while preserving graph topology and node order.
+unchop
+Join nodes that can be merged without changing the graph topology, nor the embedded sequences.
+explode
+Separate the graph into the connected components.
+squeeze
+Squeeze multiple graphs into the same file, preventing node ID collisions.
+extract
+Extract parts of the graph as defined by query criteria.
+For coordinate projections
+
+position
+Coordinate projections between nodes and paths in different graphs.
+For visualization
+
+viz
+Visualize the graph in one dimension.
+sort
+Sort the graph.
+layout
+Generate two-dimension graph layouts.
+draw
+Draw two-dimension layouts.
+
+\end{comment}
+
+ODGI allows edit operations on the graphs.
+Cycles in the graph complicate downstream analyses: odgi break removes the cycles, reducing the complexity of the graph topology.
+odgi groom removes spurious inverting links by exploring the graph from the orientation supported by most paths.
+To enable efficient sequence alignment against the graph, long nodes can be divided into shorter nodes at a maximum requested size using odgi chop.
+Partial order alignment, which consists of aligning sequences against a directed acyclic graph (DAG), is frequently used in pangenome building pipelines, but the current implementations return DAGs with 1-bp long nodes; odgi unchop allows joining nodes that can be merged without changing the graph topology, nor the embedded sequences, obtaining an equivalent, but more compact, representation of the graph.
+
+Pangenome graphs can embed multiple chromosomes as separated connected components (inter-chromosomal structural variants would join the components into bigger ones).
+odgi explode separates the connected components in different ODGI format files, while odgi squeeze allows merging multiple graphs into the same ODGI format file, preventing node ID collisions.
+odgi extract allows extracting specific subregions of the graph as defined by query criteria, thereby simplifying the downstream analyses and reducing the resources to work only with the extracted region.
+
+In variation graphs the coordinates are provided by the embedded path sequences.
+Indeed, the node IDs are not meant to be stable.
+odgi position finds, translates, and liftovers graph and path positions between different graphs by exploiting their shared path sequences (Figure 1.B).
+
+\subsection{Performance on human data}
+
+Table 2 shows a comparison of the operations necessary to extract a subgraph using ODGI and VG.
+The example consists in extracting the centromeres from a pangenome graph made from 88 haploid, phased human genome assemblies from 44 individuals from the Human Pangenome Reference Consortium year 1 assembly (see Data Availability), plus the chm13 and GRCh38 reference genomes, for a total of 90 haplotypes.
+All commands are provided in the Supplementary Material.
+
+To work with ODGI and VG it is first necessary to convert the graph from GFA to ODGI and VG format, respectively.
+Thanks to the ODGI format’s path implementation, odgi build embeds the paths in parallel, outperforming vg convert in the same task.
+For the same reason, odgi extract results 20 times faster than vg chunk in extracting a portion from the full graph.
+
+\begin{comment}
+
+% table on human perf
+
+Table 2 - Centromere extraction comparison using ODGI and VG
+Operation
+Tool
+Runtime
+(mm:ss)
+Memory
+(GB)
+Output size
+(GB)
+Format conversion
+
+odgi build
+1:35
+10,39
+5,4
+
+vg convert
+8:14
+28,43
+6,1
+Subgraph extraction
+
+odgi extract
+1:15
+9,43
+2,7
+
+vg chunk
+21:09
+59,33
+1,7
+
+\end{comment}
 
@@
 S.H. acknowledges funding from the Central Innovation Programme (ZIM) for SMEs of the Federal Ministry for Economic Affairs and Energy of Germany.
+
+\section*{Data availability}
+
+Data used to build Human pangenome graphs is available at \url{https://github.com/human-pangenomics/HPP_Year1_Data_Freeze_v1.0}.
 

### 2021-08-11 manuscript/main.tex
stuff

@@
 
-\title[short Title]{ODGI: scalable tools for working with pangenome graphs}
+\title[short Title]{ODGI: tools for working with pangenome graphs}
 \author[Heumos, \textit{et~al}.]{
@@
 
+% XXX key message of the paper is that we have collected a set of algorithms that enable easy use of pangenome graphs for investigating biology
+
 \abstract{
 \textbf{Motivation:} Pangenome graphs provide a complete representation of the mutual alignment of collections of genomes.
-These models, therefore, offer the opportunity to study the entire genetic diversity of a population, bringing new challenges to how to analyze hundreds of genomes on a gigabase scale.
-In addition, vertebrate genomes present highly repetitive regions which increase the complexity of the analysis.
+These models offer the opportunity to study the total genomic diversity of a population, including structurally complex regions.
+Analyzing hundreds of gigabase-scale genomes using this model is difficult and not well-supported by existing tools.
+%In addition, vertebrate genomes present highly repetitive regions which increase the complexity of the analysis.
 Efficient and flexible tools are therefore required to work on pangenome graphs of any complexity and scale. \\
-\textbf{Results:} Pangenome graphs provide a complete representation of the mutual alignment of collections of genomes.
-These models, therefore, offer the opportunity to study the entire genetic diversity of a population, bringing new challenges to how to analyze hundreds of genomes on a gigabase scale.
-In addition, vertebrate genomes present highly repetitive regions which increase the complexity of the analysis.
-Efficient and flexible tools are therefore required to work on pangenome graphs of any complexity and scale. \\
+\textbf{Results:} Here we present ODGI, a new suite of tools to analyze and manipulate pangenome graphs structured in the variation graph model.
+ODGI supports graphs in the Graphical Fragment Assembly format version 1 (GFAv1).
+The tools scale efficiently to large collections of eukaryotic genomes, allowing users to work easily with pangenome graphs of tens to thousands of large genomes while maintaining fast execution runtime and low memory overhead.
 \textbf{Availability:} ODGI is written in C++. Its source code is freely available at \url{https://github.com/pangenome/odgi} with documentation at \url{https://odgi.readthedocs.io}.
@@
 
-\begin{methods}
+%\begin{methods}
 
@@
 Each step records the path identifier and the previous and next path steps on the same path.
-To reduce the encoding, we encode path steps using a local alphabet that maps the N neighbors of the node into the range 1..N.
+To reduce the encoding, we encode path steps using a local alphabet that maps the N neighbors of the node into the range $1\ldots N$.
 To further save space, node deltas, rather than IDs, are stored in this alphabet.
@@
 
-\end{methods}
-
-\section{Results}
+%\end{methods}
 
@@
 
-\subsection{Core functionality}
+%\subsection{Core functionality}
 
@@
 Applying odgi stats, users can retrieve metrics describing the graph properties, such as the number of nodes, edges, paths, and graph length.
-ODGI also offers more advanced tools for interrogating the graphs.
-odgi bin summarizes the path information into bins of specified size, enabling a summarized view of gigabase scale graphs.
+ODGI also offers more advanced tools for interrogating the graphs. odgi bin summarizes the path information into bins of specified size, enabling a summarized view of gigabase scale graphs.
 odgi depth returns the node depth as defined by query criteria, allowing users to detect the complex regions in the graph due to sequences with highly identical repeats.
@@
 High degree nodes are the mirror of misassemblies or problems in the pangenome building, making the tool useful for debugging the process as well.
-
-\begin{comment}
-
-% the table
-
-Table 1 - A selection of the tools available in ODGI
-Name
-Description
-For conversion
-
-build
-Convert graphs from GFA to ODGI format.
-view
-Convert graphs from ODGI to GFA format.
-For validation
-
-validate
-Check if the graph is consistent.
-For interrogation
-
-stats
-Return graph metrics.
-bin
-Summarize path information in bins of fixed size.
-depth
-Return node depth as defined by query criteria.
-degree
-Return node degree as defined by query criteria.
-For manipulation
-
-break
-Break cycles in the graph, reducing its complexity.
-groom
-Resolve spurious inverting links.
-chop
-Chop long nodes into short ones while preserving graph topology and node order.
-unchop
-Join nodes that can be merged without changing the graph topology, nor the embedded sequences.
-explode
-Separate the graph into the connected components.
-squeeze
-Squeeze multiple graphs into the same file, preventing node ID collisions.
-extract
-Extract parts of the graph as defined by query criteria.
-For coordinate projections
-
-position
-Coordinate projections between nodes and paths in different graphs.
-For visualization
-
-viz
-Visualize the graph in one dimension.
-sort
-Sort the graph.
-layout
-Generate two-dimension graph layouts.
-draw
-Draw two-dimension layouts.
-
-\end{comment}
 
@@
 
-\subsection{Performance on human data}
-
-Table 2 shows a comparison of the operations necessary to extract a subgraph using ODGI and VG.
-The example consists in extracting the centromeres from a pangenome graph made from 88 haploid, phased human genome assemblies from 44 individuals from the Human Pangenome Reference Consortium year 1 assembly (see Data Availability), plus the chm13 and GRCh38 reference genomes, for a total of 90 haplotypes.
-All commands are provided in the Supplementary Material.
-
-To work with ODGI and VG it is first necessary to convert the graph from GFA to ODGI and VG format, respectively.
-Thanks to the ODGI format’s path implementation, odgi build embeds the paths in parallel, outperforming vg convert in the same task.
-For the same reason, odgi extract results 20 times faster than vg chunk in extracting a portion from the full graph.
-
 \begin{comment}
-
-% table on human perf
-
-Table 2 - Centromere extraction comparison using ODGI and VG
-Operation
-Tool
-Runtime
-(mm:ss)
-Memory
-(GB)
-Output size
-(GB)
-Format conversion
-
-odgi build
-1:35
-10,39
-5,4
-
-vg convert
-8:14
-28,43
-6,1
-Subgraph extraction
-
-odgi extract
-1:15
-9,43
-2,7
-
-vg chunk
-21:09
-59,33
-1,7
-
+key message of the paper is that we have collected a set of algorithms that enable easy use of pangenome graphs for investigating biology
+-> build model solves problem of working with big graphs in memory
+-> view (convert to GFA) & paths solve problem of exporting basic features of the graph (e.g. paths)
+-> stats (understand basic size / structure) & bin & degree & depth solves problem of understanding the overall structure and size of the graph
+-> sort (groom) & layout solves problem of finding latent structure in the pangenome
+-> viz & draw provides a human-viewable readout of the graph
+-> chop & unchop & squeeze & break & prune & explode lets us break apart or combine the graph nodes and topology
+-> position & tips & untangle (jaccard based coordinate conversion) provides a way to map coordinates between any genomes in the graph (e.g. liftover!)
+-> extract lets us pull out specific regions of the graph based on path ranges, nodes and positions
 \end{comment}
 
+\section{Results}
+
+\subsection{Building the \textsc{ODGI} model}
+
+% erik
+
+% build
+Solves problem of working with big and complex graphs in limited memory.
+
+Figure: encoding model as a cartoon --  HTT exon 1 as tiny example graph (maybe)
+
+\subsection{Converting to standard text formats}
+
+% pjotr
+
+% view / paths
+Allows us to export basic data of the graph (genomes and graph structure).
+
+Figure: .dot showing the relationship between all the tools and the data formats we use.
+
+\subsection{Obtaining metrics of the pangenome graph}
+
+% simon
+% chr8 longs
+
+% stats / bin / degree / depth / multiQC
+Solves the problem of understanding the overall structure, size, and features of the graph and its sequences.
+
+Figure: small graph with all the stats shown. HTT exon 1 as tiny example graph
+
+\subsection{Finding latent structure in the pangenome graph topology}
+
+% simon
+
+% sort / layout / tension??
+Allows us to understand the sparse structures typically found in pangenome graphs.
+
+\subsection{Visualizing pangenome graphs}
+
+% andrea (viz) & erik (draw)
+
+% viz / draw
+These provide scalable ways of geneating human-viewable pictures of the high-level structure of the graph.
+% interesting that we don't go base-level at all... we use vg for that
+
+Figure here with 1D + 2D vizs.
+
+\subsection{Editing the graph structure}
+
+% andrea (?)
+
+% chop / unchop / squeeze / break / prune / explode
+These are commonly-needed basic operations on the topology of the graph.
+
+\subsection{Untangling and navigating the pangenome with path-based coordinates}
+
+% pjotr
+
+% position / tips
+% untangle (this follows on the jaccard graph mapping concepts in the previous section}
+The graph is a model of an alignment of many genomes.
+We can use coordinates in any genome to refer to it, but this requires a few basic operations ...
+Obtaining unambiguous mappings between different genomes requires the use of a new kind of graph based mapping (path graph jaccard).
+
+Figure showing the untangling of C4 ?
+Figure showing tips over a single human chromosome from HPRC.
+
+\subsection{Extracting regions of interest}
+
+% andrea (?)
+
+% extract
+Pangenome graphs are very large, but we often only want to work with a small portion (e.g. a single gene).
+We can extract such regions using coordinates on the paths in the graph to guide us.
+
+Figure showing extraction process---either schematic or ``real''.
+
 \section{Discussion}
+
+% erik & pjotr & anyone
 

### 2021-08-11 manuscript/main.tex
authors

@@
 
-\title[short Title]{ODGI: tools for working with pangenome graphs}
+\title[short Title]{ODGI: understanding pangenome graphs}
 \author[Heumos, \textit{et~al}.]{
-Simon Heumos\,$^{\text{\sfb 1, \sfb 2} \dagger}$,
-Andrea Guarracino\,$^{\text{\sfb 2} \dagger}$,
+Andrea Guarracino\,$^{\text{\sfb 1} \dagger}$,
+Simon Heumos\,$^{\text{\sfb 2} \dagger}$,
 Pjotr Prins\,$^{\text{\sfb 3}}$,
@@
 \address{
-$^{\text{\sf 1}}$Quantitative Biology Center (QBiC), University of T\"ubingen, T\"ubingen, Germany, 72076
-$^{\text{\sf 2}}$University of Tor Vergata, Rome, Italy and \\
+$^{\text{\sf 1}}$University of Tor Vergata, Rome, Italy and \\
+$^{\text{\sf 2}}$Quantitative Biology Center (QBiC), University of T\"ubingen, T\"ubingen, Germany, 72076
 $^{\text{\sf 3}}$University of Tennessee Health Science Center, Memphis, TN, USA

### 2021-08-11 manuscript/main.tex
andrea for view
update authors section

@@
 Simon Heumos\,$^{\text{\sfb 2} \dagger}$,
+Sven Nahnsen\,$^{\text{\sfb 2}}$,
 Pjotr Prins\,$^{\text{\sfb 3}}$,
@@
 \address{
-$^{\text{\sf 1}}$University of Tor Vergata, Rome, Italy and \\
-$^{\text{\sf 2}}$Quantitative Biology Center (QBiC), University of T\"ubingen, T\"ubingen, Germany, 72076
+$^{\text{\sf 1}}$University of Tor Vergata, Rome, Italy \\
+$^{\text{\sf 2}}$Quantitative Biology Center (QBiC), University of T\"ubingen, T\"ubingen, Germany, 72076 \\
 $^{\text{\sf 3}}$University of Tennessee Health Science Center, Memphis, TN, USA
@@
 
-% pjotr
+% pjotr & andrea
 

### 2021-09-27 manuscript/main.tex
take over untangle
drafting metrics section

@@
 
-% stats / bin / degree / depth / multiQC
+% stats / multiQC / bin / degree / depth
+
+Applying odgi stats, users can retrieve metrics describing the graph properties, such as the number of nodes, edges, paths, and graph length.
+ODGI also offers more advanced tools for interrogating the graphs.
+odgi bin summarizes the path information into bins of specified size, enabling a summarized view of gigabase scale graphs.
+odgi depth returns the node depth as defined by query criteria, allowing users to detect the complex regions in the graph due to sequences with highly identical repeats.
+Complex motifs can be also detected with odgi degree, which returns the node degree as defined by query criteria.
+High degree nodes are the mirror of misassemblies or problems in the pangenome building, making the tool useful for debugging the process as well.
 Solves the problem of understanding the overall structure, size, and features of the graph and its sequences.
 
-Figure: small graph with all the stats shown. HTT exon 1 as tiny example graph
+Figure: small graph with all the stats shown. A screenshot of MultiQC, visualize summary per bin somehow, R visualizations of degree and depth HTT exon 1 as tiny example graph
 
@@
 
-% sort / layout / tension??
+% sort / layout / NO tension! We keep that for the sorting paper
 Allows us to understand the sparse structures typically found in pangenome graphs.
@@
 
-%This work was supported, in part, by the National Institutes of Health (award numbers U01HG010961, U41HG010972, R01HG010485, 2U41HG007234, 5U54HG007990, 5T32HG008345-04, U01HL137183 to B.P.) and the W.
-M.
-Keck Foundation (award number DT06172015 to B.P.).
 S.H. acknowledges funding from the Central Innovation Programme (ZIM) for SMEs of the Federal Ministry for Economic Affairs and Energy of Germany.
+This work was supported by the BMBF-funded de.NBI Cloud within the German Network for Bioinformatics Infrastructure (de.NBI) (031A537B, 031A533A, 031A538A, 031A533B, 031A535A, 031A537C, 031A534A, 031A532B).
 

### 2021-09-28 manuscript/main.tex
adding citations where possible

@@
 
-A pangenome models the full set of genomic elements in a given species or clade {32453966}.
+A pangenome models the full set of genomic elements in a given species or clade \cite{Eizenga_2020}.
 Indeed, these data structures encode the mutual relationships between all the genomes represented, in contrast to reference-based approaches which relate sequences only to a particular genome, chosen as reference.
-A class of methods to represent pangenomes involves the sequence graphs {2488477}, where similar orthologous regions between genomes collapse into a single representative sequence.
+A class of methods to represent pangenomes involves the sequence graphs \cite{Paten:2017}, where similar orthologous regions between genomes collapse into a single representative sequence.
 In node-labeled sequence graphs, nodes indicate DNA sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph.
 A bidirected sequence graph represents both strands of DNA.
-On this model, variation graphs add the concept of path to embed linear sequences (e.g., genomes or haplotypes) into the graph {30125266}.
+On this model, variation graphs add the concept of path to embed linear sequences (e.g., genomes or haplotypes) into the graph.
 Paths provide a stable coordinate system, allowing graph annotations and interoperability between different graphs.
 
-Thanks to advances in sequencing technologies, new genome assemblies are produced at a high rate.
+Thanks to advances in sequencing technologies, new full-length genome assemblies are produced at a high rate, including efforst from the Human Pangenome Reference Consortium \cite{HPRC} (HPRC) and Telomere-to-Telomere (T2T) projects \cite{Miga:2020}, \cite{Logsdon_2021}, \cite{Nurk_2021} from the T2T consortium \cite{T2T}.
 This amount of data offers the opportunity to study genomic variation as never before, but it also brings challenges in how to represent and manipulate hundreds of genomes at a gigabase scale.
 Variation graphs compactly represent the full genetic variation across a population, but graph-based data structures present computational overheads: in addition to the sequences, they must also represent the graph topology.
-Several genome data structures were developed to address these challenges {33040146}.
+Several genome data structures were developed to address these challenges \cite{Eizenga_2020_BX}.
 Memory-efficient representations do not allow dynamic updates to the graph, such as adding new genomes or simplifying the graph topology.
@@
 To overcome this limitation, we have developed an Optimized Dynamic Genome/Graph Implementation (ODGI).
-Each tool loads graphs into the succinct dynamic HandleGraph model {33040146} and applies methods based on the HandleGraph API.
-Several algorithms are shared with the VG toolkit (VG) {30125266}, the state-of-the-art toolkit for working with pangenome graphs.
+Each tool loads graphs into the succinct dynamic HandleGraph model \cite{Eizenga_2020_BX} and applies methods based on the HandleGraph API.
+Several algorithms are shared with the VG toolkit (VG) \cite{Garrison:2018}, the state-of-the-art toolkit for working with pangenome graphs.
 However, although similar in scope to VG, ODGI importantly differs in its design to support highly complex pangenome graphs.
@@
 Moreover, thanks to its efficient path representation, ODGI tools can build and walk the internal graph representation (the ODGI format) in parallel, which is one of the major bottlenecks when working with very large and deep variation graphs, efficiently scaling up to hundreds of large vertebrate genomes.
-
-\cite{Eizenga_2020}
 
@@
 ODGI is designed to build and modify paths in parallel, keeping its representation in memory efficient.
-It applies concepts first introduced in the dynamic version of the Graph BWT {31406990} (gBWT), a scalable implementation of the graph extension of the positional Burrows-Wheeler transform {28702075}.
+It applies concepts first introduced in the dynamic version of the Graph BWT \cite{Siren:2020} (gBWT), a scalable implementation of the graph extension of the positional Burrows-Wheeler transform \cite{Durbin_2014}.
 Specifically, we build a node-local model of the graph, where the graph is stored in a vector of node structures.
@@
 
-% pjotr
+% erik
 

### 2021-09-29 manuscript/main.tex
citation style matches BX rules, web pages must be footnotes

--- 2021-09-27 manuscript/main.tex
@@
 
-A pangenome models the full set of genomic elements in a given species or clade {32453966}.
+A pangenome models the full set of genomic elements in a given species or clade \citep{Eizenga_2020}.
 Indeed, these data structures encode the mutual relationships between all the genomes represented, in contrast to reference-based approaches which relate sequences only to a particular genome, chosen as reference.
-A class of methods to represent pangenomes involves the sequence graphs {2488477}, where similar orthologous regions between genomes collapse into a single representative sequence.
+A class of methods to represent pangenomes involves the sequence graphs \citep{Paten:2017}, where similar orthologous regions between genomes collapse into a single representative sequence.
 In node-labeled sequence graphs, nodes indicate DNA sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph.
 A bidirected sequence graph represents both strands of DNA.
-On this model, variation graphs add the concept of path to embed linear sequences (e.g., genomes or haplotypes) into the graph {30125266}.
+On this model, variation graphs add the concept of path to embed linear sequences (e.g., genomes or haplotypes) into the graph.
 Paths provide a stable coordinate system, allowing graph annotations and interoperability between different graphs.
 
-Thanks to advances in sequencing technologies, new genome assemblies are produced at a high rate.
+Thanks to advances in sequencing technologies, new full-length genome assemblies are produced at a high rate, including efforst from the Human Pangenome Reference Consortium\footnote{https://humanpangenome.org/ (accessed Oct 2021)} (HPRC) and Telomere-to-Telomere (T2T) projects \citep{Miga:2020, Logsdon_2021, Nurk_2021}, from the T2T consortium\footnote{https://sites.google.com/ucsc.edu/t2tworkinggroup/home (accessed Oct 2021)}.
 This amount of data offers the opportunity to study genomic variation as never before, but it also brings challenges in how to represent and manipulate hundreds of genomes at a gigabase scale.
 Variation graphs compactly represent the full genetic variation across a population, but graph-based data structures present computational overheads: in addition to the sequences, they must also represent the graph topology.
-Several genome data structures were developed to address these challenges {33040146}.
+Several genome data structures were developed to address these challenges \citep{Eizenga_2020_BX}.
 Memory-efficient representations do not allow dynamic updates to the graph, such as adding new genomes or simplifying the graph topology.
@@
 To overcome this limitation, we have developed an Optimized Dynamic Genome/Graph Implementation (ODGI).
-Each tool loads graphs into the succinct dynamic HandleGraph model {33040146} and applies methods based on the HandleGraph API.
-Several algorithms are shared with the VG toolkit (VG) {30125266}, the state-of-the-art toolkit for working with pangenome graphs.
+Each tool loads graphs into the succinct dynamic HandleGraph model \citep{Eizenga_2020_BX} and applies methods based on the HandleGraph API.
+Several algorithms are shared with the VG toolkit (VG) \citep{Garrison:2018}, the state-of-the-art toolkit for working with pangenome graphs.
 However, although similar in scope to VG, ODGI importantly differs in its design to support highly complex pangenome graphs.
@@
 ODGI is designed to build and modify paths in parallel, keeping its representation in memory efficient.
-It applies concepts first introduced in the dynamic version of the Graph BWT {31406990} (gBWT), a scalable implementation of the graph extension of the positional Burrows-Wheeler transform {28702075}.
+It applies concepts first introduced in the dynamic version of the Graph BWT \citep{Siren:2020} (gBWT), a scalable implementation of the graph extension of the positional Burrows-Wheeler transform \citep{Durbin_2014}.
 Specifically, we build a node-local model of the graph, where the graph is stored in a vector of node structures.
@@
 Each step records the path identifier and the previous and next path steps on the same path.
-To reduce the encoding, we encode path steps using a local alphabet that maps the N neighbors of the node into the range $1\ldots N$.
+To reduce the encoding, we encode path steps using a local alphabet that maps the N neighbours of the node into the range $1\ldots N$.
 To further save space, node deltas, rather than IDs, are stored in this alphabet.

### 2021-09-30 manuscript/main.tex
discussion draft, metrics text

@@
 
-Thanks to advances in sequencing technologies, new full-length genome assemblies are produced at a high rate, including efforst from the Human Pangenome Reference Consortium\footnote{https://humanpangenome.org/ (accessed Oct 2021)} (HPRC) and Telomere-to-Telomere (T2T) projects \citep{Miga:2020, Logsdon_2021, Nurk_2021}, from the T2T consortium\footnote{https://sites.google.com/ucsc.edu/t2tworkinggroup/home (accessed Oct 2021)}.
+Thanks to advances in sequencing technologies, new full-length genome assemblies are produced at a high rate, including efforst from the Human Pangenome Reference Consortium\footnote{\url{https://humanpangenome.org/} (accessed Oct 2021)} (HPRC) and Telomere-to-Telomere (T2T) projects \citep{Miga:2020, Logsdon_2021, Nurk_2021}, from the T2T consortium\footnote{\url{https://sites.google.com/ucsc.edu/t2tworkinggroup/home} (accessed Oct 2021)}.
 This amount of data offers the opportunity to study genomic variation as never before, but it also brings challenges in how to represent and manipulate hundreds of genomes at a gigabase scale.
@@
 
-\subsection{Obtaining metrics of the pangenome graph}
-
-% simon
-% chr8 longs
-
-% stats / multiQC / bin / degree / depth
-
-Applying odgi stats, users can retrieve metrics describing the graph properties, such as the number of nodes, edges, paths, and graph length.
-ODGI also offers more advanced tools for interrogating the graphs.
-odgi bin summarizes the path information into bins of specified size, enabling a summarized view of gigabase scale graphs.
-odgi depth returns the node depth as defined by query criteria, allowing users to detect the complex regions in the graph due to sequences with highly identical repeats.
-Complex motifs can be also detected with odgi degree, which returns the node degree as defined by query criteria.
-High degree nodes are the mirror of misassemblies or problems in the pangenome building, making the tool useful for debugging the process as well.
-Solves the problem of understanding the overall structure, size, and features of the graph and its sequences.
-
-Figure: small graph with all the stats shown. A screenshot of MultiQC, visualize summary per bin somehow, R visualizations of degree and depth HTT exon 1 as tiny example graph
-
-\subsection{Finding latent structure in the pangenome graph topology}
-
-% simon
-
-% sort / layout / NO tension! We keep that for the sorting paper
-Allows us to understand the sparse structures typically found in pangenome graphs.
-
 \subsection{Visualizing pangenome graphs}
@@
 Figure here with 1D + 2D vizs.
+
+\subsection{Finding latent structure in the pangenome graph topology}
+
+% simon
+
+% sort / layout / NO tension! We keep that for the sorting paper
+Allows us to understand the sparse structures typically found in pangenome graphs.
+
+\subsection{Obtaining metrics of the pangenome graph}
+
+% simon
+% HTT exon1 
+% Only chr8 longs if I can't produce someting nice
+
+% stats / multiQC / bin / degree / depth
+
+% Solves the problem of understanding the overall structure, size, and features of the graph and its sequences.
+
+Graphs statistics provide alternative ways to gain insight into pangenomes complexity revealing the overall structure, size, and features of a graph and its sequences.
+
+Applying \texttt{odgi stats}, users can retrieve metrics describing the graph properties, such as the number of nodes, edges, paths, and graph length.
+MultiQC's \citep{Ewels_2016} ODGI module\footnote{\url{https://multiqc.info/docs/\#odgi} (accessed Oct 2021)} provides an interactive way to comparatively explore the statistics of an arbitrary number of graphs.
+ODGI also offers more advanced tools for the interrogation of the graphs.
+\texttt{odgi bin} summarizes the path information into bins of a specified size, enabling a summarized view of gigabase scale graphs.
+\texttt{odgi depth} returns the node depth as defined by query criteria, allowing users to detect the complex regions in the graph due to collapsed sequences with highly identical repeats.
+Complex motifs can be also detected with \texttt{odgi degree}, which returns the node degree as defined by query criteria.
+High degree nodes are the mirror of misassemblies or problems in the pangenome building, making the tool useful for validation.
+
+% Figure: small graph with all the stats shown. A screenshot of MultiQC, visualize summary per bin somehow, R visualizations of degree and depth HTT exon 1 as tiny example graph
 
@@
 % erik & pjotr & anyone
+
+Pangenome graphs stand to become a ubiquitous tool in genomics \citep{Eizenga_2020}.
+With ODGI we implemented a state-of-the-art tool suite that can transform, analyse, simplify, validate, and visualise pangenome graphs at large scale.
+Lifting over annotations and linearizing nested graph structures place the suite as the bridge between traditional linear reference genome analysis and pangenome graphs.
+ODGI is a unique set of tools that enables scientists to explore and discover the underlying biology of pangenome graphs.
+Already, the tools are the backbone of pipelines such as the Pangenome Graph Builder\footnote{\url{https://github.com/pangenome/pggb} (accessed Oct 2021)} (PGGB) or nf-core/pangenome\footnote{\url{https://github.com/nf-core/pangenome} (accessed Oct 2021)}.
+Future work will add support for RNA and protein sequences and expand on metadata capabilities of large pangenome graphs.
 

### 2021-10-04 manuscript/fig_metrics.tex
metrics: figure, bib, script

\begin{figure*}[h!]
	\begin{subfigure}{\linewidth}
		\centering
		% include first image
		\includegraphics[width=1.0\linewidth]{fig/metrics/chr4.pan.HTTex1.gfa.multiqc_odgi_stats.png}  
		\caption{Put your sub-caption here}
		\label{fig:metrics-multiqc}
	\end{subfigure}

	\begin{subfigure}{.5\linewidth}
		\centering
		% include second image
		\includegraphics[width=1.0\linewidth]{fig/metrics/chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.png}  
		\caption{Put your sub-caption here}
		\label{fig:sub-second}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
	\centering
	% include second image
	\includegraphics[width=1.0\linewidth]{fig/metrics/chr4.HTT.chm13.depth.w1.bed.pdf}  
	\caption{Put your sub-caption here}
	\label{fig:sub-sixth}
	\end{subfigure}

	\begin{subfigure}{.5\linewidth}
		\centering
		% include third image
		\includegraphics[width=1.0\linewidth]{fig/metrics/chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.w2.convert.png}  
		\caption{Put your sub-caption here}
		\label{fig:sub-third}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\centering
		% include fourth image
		\includegraphics[width=1.0\linewidth]{fig/metrics/chr4.HTT.chm13.degree.w1.bed.pdf}  
		\caption{Put your sub-caption here}
		\label{fig:sub-fourth}
	\end{subfigure}
%	\includegraphics[width=\linewidth]{fig/metrics/chr4.pan.HTTex1.gfa.multiqc_odgi_stats.png}
	\caption{Various outputs of the ODGI metrics tools.}
	\label{fig:metrics}
\end{figure*}

### 2021-10-04 manuscript/main.tex
metrics: figure, bib, script

@@
 \usepackage{comment}
+\usepackage{subcaption}
 
@@
 \title[short Title]{ODGI: understanding pangenome graphs}
-\author[Heumos, \textit{et~al}.]{
+\author[Guarracino, Heumos \textit{et~al}.]{
 Andrea Guarracino\,$^{\text{\sfb 1} \dagger}$,
@@
 % sort / layout / NO tension! We keep that for the sorting paper
-Allows us to understand the sparse structures typically found in pangenome graphs.
+
+% Allows us to understand the sparse structures typically found in pangenome graphs.
+
+Pangenome graphs built from raw sets of alignments may have complex structures which can introduce difficulty in downstream analyses.
+Uncovering their latent topology allows us to understand the sparse structures typically found in pangenome graphs.
+
+\texttt{odgi sort} offers multiple sorting algorithms to find the best node order in one dimension.
+Nodes can be sorted randomly, topologically, by breaking cycles in the graph, by grooming, or by using the novel path-guided stochastic gradient descent (PG-SGD) algorithm.
+can be combined together to a series of sorts.
 
@@
 \texttt{odgi bin} summarizes the path information into bins of a specified size, enabling a summarized view of gigabase scale graphs.
-\texttt{odgi depth} returns the node depth as defined by query criteria, allowing users to detect the complex regions in the graph due to collapsed sequences with highly identical repeats.
+This concept found its way into the pangenome graph ontology \cite{Yokoyama2020}.
+\texttt{odgi depth} returns the node depth as defined by query criteria, allowing users to assess the complexity of regions in the graph.
+Collapsed sequences with highly identical repeats indicate intricate loci.
 Complex motifs can be also detected with \texttt{odgi degree}, which returns the node degree as defined by query criteria.
 High degree nodes are the mirror of misassemblies or problems in the pangenome building, making the tool useful for validation.
+In Fig.
+\ref{fig:metrics}, the characteristics of a human pangenome graph of the exon 1 huntingtin gene \citep{Sathasivam2013} (\textit{HTTexon1}), obtained with the ODGI metrics tools, are shown.
+
+\include{fig_metrics.tex}
 
@@
 
+rdf sparql side, bring in annotation for great biological interpretation
+
+some cancer stuff?!
+
+(recent) interactive visualization tools?!
+
 \section*{Funding}
@@
 Data used to build Human pangenome graphs is available at \url{https://github.com/human-pangenomics/HPP_Year1_Data_Freeze_v1.0}.
+TODO!
 

### 2021-10-04 scripts/metrics.sh

@@ -1,0 +1,35 @@
+#!/bin/bash
+
+wget https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/scratch/2021_07_30_pggb/chroms/chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.gz
+gunzip chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.gz
+wget https://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/genes/hg38.ncbiRefSeq.gtf.gz
+gunzip hg38.ncbiRefSeq.gtf.gz
+
+grep HTT hg38.ncbiRefSeq.gtf | grep "exon_number \"1\"" | grep "gene_name \"HTT\"" | grep -P "\texon"
+#chr4	ncbiRefSeq.2021-09-09	exon	3074681	3075088	.	+	.	gene_id "HTT"; transcript_id "NM_001388492.1"; exon_number "1"; exon_id "NM_001388492.1.1"; gene_name "HTT";
+#chr4	ncbiRefSeq.2021-09-09	exon	3074681	3075088	.	+	.	gene_id "HTT"; transcript_id "NM_002111.8"; exon_number "1"; exon_id "NM_002111.8.1"; gene_name "HTT";
+
+odgi build -g chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa -o chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og -t 16 -P
+odgi paths -L -i chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og > chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.paths
+
+odgi extract -i chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og -r grch38#chr4:3074681-3075088 -o chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og -P -t 16 -E
+odgi sort -i chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og -o chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og -O
[…]

### 2021-10-05 manuscript/fig_metrics.tex
finalized metrics figure caption

@@
 	\begin{subfigure}{\linewidth}
+		\caption{}
 		\centering
@@
 		\includegraphics[width=1.0\linewidth]{fig/metrics/chr4.pan.HTTex1.gfa.multiqc_odgi_stats.png}  
-		\caption{Put your sub-caption here}
 		\label{fig:metrics-multiqc}
@@
 	\begin{subfigure}{.5\linewidth}
+		\caption{}
 		\centering
@@
 		\includegraphics[width=1.0\linewidth]{fig/metrics/chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.png}  
-		\caption{Put your sub-caption here}
-		\label{fig:sub-second}
+		\label{fig:metrics-viz}
 	\end{subfigure}
 	\begin{subfigure}{.5\linewidth}
+		\caption{}
 		\centering
@@
 		\includegraphics[width=1.0\linewidth]{fig/metrics/chr4.HTT.chm13.depth.w1.bed.pdf}  
-	\caption{Put your sub-caption here}
-	\label{fig:sub-sixth}
+		\label{fig:metrics-depth}
 	\end{subfigure}
@@
 		\includegraphics[width=1.0\linewidth]{fig/metrics/chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.w2.convert.png}  
-		\caption{Put your sub-caption here}
-		\label{fig:sub-third}
+		\label{fig:metrics-bin}
 	\end{subfigure}
 	\begin{subfigure}{.5\linewidth}
+		\caption{}
 		\centering
@@
 		\includegraphics[width=1.0\linewidth]{fig/metrics/chr4.HTT.chm13.degree.w1.bed.pdf}  
-		\caption{Put your sub-caption here}
-		\label{fig:sub-fourth}
+		\label{fig:metrics-degree}
 	\end{subfigure}
 %	\includegraphics[width=\linewidth]{fig/metrics/chr4.pan.HTTex1.gfa.multiqc_odgi_stats.png}
-	\caption{Various outputs of the ODGI metrics tools.}
+	\caption{Features of a human pangenome graph of the exon 1 huntingtin gene (\textit{HTTexon1}): \textbf{(a)} Excerpt of vital statistics of the \textit{HTTexon1} graph displayed by MultiQC's ODGI module.
+The very high GC content of 73.0\% compared to a human genomic mean GC content of 40.9\% \cite{Piovesan2019} is in accordance with the literature (see for example \cite{Sathasivam2013, Neueder2017}).
+\textbf{(b)} 1D visualization of the \textit{HTTexon1} graph with a bin width of 1.
+\textbf{(c)} Per nucleotide node depth distribution of CHM13 in the \textit{HTTexon1} graph.
+The alternating depth around position 200 indicates polymorphic variation.
+\textbf{(d)} 1D visualization of the \textit{HTTexon1} graph with a bin width of 2.
+Compared to above, the graph's width is exactly one half, therefore small variation is not visible anymore and the resolution of graph topology decreases.
+\textbf{(e)} Per nucleotide node degree distribution of CHM13 in the \textit{HTTexon1} graph.
+The polymorphic region is around position 200 complementing the above node depth analysis.
+Figures \textbf{(b)-(e)} clearly highlight the variant region around position 200.}
 	\label{fig:metrics}

### 2021-10-05 manuscript/main.tex
finalized metrics figure caption

@@
 In Fig.
-\ref{fig:metrics}, the characteristics of a human pangenome graph of the exon 1 huntingtin gene \citep{Sathasivam2013} (\textit{HTTexon1}), obtained with the ODGI metrics tools, are shown.
+\ref{fig:metrics}, the characteristics of a human pangenome graph of the exon 1 huntingtin gene \citep{Sathasivam2013,Neueder2017} (\textit{HTTexon1}), obtained with the ODGI metrics tools, are shown.
 
@@
 
-rdf sparql side, bring in annotation for great biological interpretation
+rdf sparql side, bring in annotation for great biological interpretation \cite{Yokoyama2020}, federated queries, sharing, FAIR
+
+(recent) interactive visualization tools?! CITE PANACHE, also indicate that no published tool for interactive exploration of such graphs exist
 
 some cancer stuff?!
-
-(recent) interactive visualization tools?!
 

### 2021-10-06 manuscript/fig_metrics.tex
towards a nice metrics figure

@@
 
-	\begin{subfigure}{.5\linewidth}
+	\begin{subfigure}{\linewidth}
 		\caption{}
@@
 		% include second image
-		\includegraphics[width=1.0\linewidth]{fig/metrics/chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.png}  
+		\includegraphics[width=1.0\linewidth, trim=-1.5cm 0 -1cm 0 ]{fig/metrics/chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.a1.png.svg.pdf}  
 		\label{fig:metrics-viz}
 	\end{subfigure}
-	\begin{subfigure}{.5\linewidth}
+	\begin{subfigure}{\linewidth}
 		\caption{}
@@
 		% include second image
-		\includegraphics[width=1.0\linewidth]{fig/metrics/chr4.HTT.chm13.depth.w1.bed.pdf}  
+		\includegraphics[width=\linewidth,trim=0 0 0cm +2cm]{fig/metrics/chr4.HTT.chm13.depth.w1.bed.pdf}  
 		\label{fig:metrics-depth}
 	\end{subfigure}
-
-	\begin{subfigure}{.5\linewidth}
-		\caption{}
-		\centering
-		% include third image
-		\includegraphics[width=1.0\linewidth]{fig/metrics/chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.w2.convert.png}  
-		\label{fig:metrics-bin}
-	\end{subfigure}
-	\begin{subfigure}{.5\linewidth}
+	\begin{subfigure}{1\linewidth}
 		\caption{}
@@
 		% include fourth image
-		\includegraphics[width=1.0\linewidth]{fig/metrics/chr4.HTT.chm13.degree.w1.bed.pdf}  
+		\includegraphics[width=\linewidth,trim=0 0 0 +2cm]{fig/metrics/chr4.HTT.chm13.degree.w1.bed.pdf}  
 		\label{fig:metrics-degree}

### 2021-10-06 scripts/metrics.sh

@@ -14,9 +14,32 @@ odgi paths -L -i chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og > chr4.pan.fa
 
 odgi extract -i chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og -r grch38#chr4:3074681-3075088 -o chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og -P -t 16 -E
 odgi sort -i chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og -o chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og -O
-odgi viz -i chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og -o chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.png
-odgi viz -i chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og -w 2 -o chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.w2.png
-convert -size 852x940 xc:white chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.w2.png -gravity West -composite chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.w2.convert.png
+
+odgi paths -i chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og -L > chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.paths
+#### FIXME ####
+# we accidentally cut off stuff from chm13 and grch13 -> fix this first
[…]

### 2021-10-06 manuscript/main.tex
towards a nice metrics figure
metrics figure and section ready

@@
 % viz / draw
-These provide scalable ways of geneating human-viewable pictures of the high-level structure of the graph.
-% interesting that we don't go base-level at all... we use vg for that
+These provide scalable ways of generating human-viewable pictures of the high-level structure of the graph.
+% interesting that we don't go base-level at all... we use vg for that, or STM
 
@@
 Complex motifs can be also detected with \texttt{odgi degree}, which returns the node degree as defined by query criteria.
-High degree nodes are the mirror of misassemblies or problems in the pangenome building, making the tool useful for validation.
+High degree nodes can be the mirror of variation, misassemblies, or problems in the pangenome building, making the tool useful for validation.
 In Fig.

### 2021-10-06 scripts/metrics.sh

@@ -1,5 +1,7 @@
 #!/bin/bash
 
+SVG=$1
+
 wget https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/scratch/2021_07_30_pggb/chroms/chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.gz
 gunzip chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.gz
 wget https://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/genes/hg38.ncbiRefSeq.gtf.gz
@@ -15,32 +17,6 @@ odgi paths -L -i chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og > chr4.pan.fa
 odgi extract -i chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og -r grch38#chr4:3074681-3075088 -o chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og -P -t 16 -E
 odgi sort -i chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og -o chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og -O
 
-odgi paths -i chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og -L > chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.paths
-#### FIXME ####
-# we accidentally cut off stuff from chm13 and grch13 -> fix this first
-cat chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.paths | cut -f 1 -d':' | cut -f 1,2 -d'#' | cut -c3- > chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.paths_tiny
-odgi view -i chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og -g > chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.gfa
[…]

### 2021-10-06 manuscript/fig_metrics.tex

@@
 		% include second image
-		\includegraphics[width=1.0\linewidth, trim=-1.5cm 0 -1cm 0 ]{fig/metrics/chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.a1.png.svg.pdf}  
+		\includegraphics[width=1.0\linewidth, trim=0 0 -1cm 0 ]{fig/metrics/chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.tiny.og.png.svg.pdf}  
 		\label{fig:metrics-viz}
@@
 		% include second image
-		\includegraphics[width=\linewidth,trim=0 0 0cm +2cm]{fig/metrics/chr4.HTT.chm13.depth.w1.bed.pdf}  
+		\includegraphics[width=\linewidth,trim=+.225cm 0 +0.425cm +2cm]{fig/metrics/chr4.HTT.chm13.depth.w1.bed.pdf}  
 		\label{fig:metrics-depth}
@@
 		% include fourth image
-		\includegraphics[width=\linewidth,trim=0 0 0 +2cm]{fig/metrics/chr4.HTT.chm13.degree.w1.bed.pdf}  
+		\includegraphics[width=\linewidth,trim=+.225cm 0 +.425cm +2cm]{fig/metrics/chr4.HTT.chm13.degree.w1.bed.pdf}  
 		\label{fig:metrics-degree}
@@
 The very high GC content of 73.0\% compared to a human genomic mean GC content of 40.9\% \cite{Piovesan2019} is in accordance with the literature (see for example \cite{Sathasivam2013, Neueder2017}).
-\textbf{(b)} 1D visualization of the \textit{HTTexon1} graph with a bin width of 1.
+\textbf{(b)} \texttt{odgi viz} visualization of the \textit{HTTexon1} graph.
 \textbf{(c)} Per nucleotide node depth distribution of CHM13 in the \textit{HTTexon1} graph.
 The alternating depth around position 200 indicates polymorphic variation.
-\textbf{(d)} 1D visualization of the \textit{HTTexon1} graph with a bin width of 2.
-Compared to above, the graph's width is exactly one half, therefore small variation is not visible anymore and the resolution of graph topology decreases.
-\textbf{(e)} Per nucleotide node degree distribution of CHM13 in the \textit{HTTexon1} graph.
-The polymorphic region is around position 200 complementing the above node depth analysis.
-Figures \textbf{(b)-(e)} clearly highlight the variant region around position 200.}
+\textbf{(d)} Per nucleotide node degree distribution of CHM13 in the \textit{HTTexon1} graph.
+The polymorphic region around position 200 complements the above node depth analysis.
+Figures \textbf{(b)-(d)} clearly highlight the variant region around position 200 showing the variable number of glutamine residues of the different individuals as reported by \cite{Nance1999}.}
 	\label{fig:metrics}

### 2021-10-06 manuscript/main.tex
abstract overhaul

@@
 %In addition, vertebrate genomes present highly repetitive regions which increase the complexity of the analysis.
-Efficient and flexible tools are therefore required to work on pangenome graphs of any complexity and scale. \\
-\textbf{Results:} Here we present ODGI, a new suite of tools to analyze and manipulate pangenome graphs structured in the variation graph model.
-ODGI supports graphs in the Graphical Fragment Assembly format version 1 (GFAv1).
-The tools scale efficiently to large collections of eukaryotic genomes, allowing users to work easily with pangenome graphs of tens to thousands of large genomes while maintaining fast execution runtime and low memory overhead.
-\textbf{Availability:} ODGI is written in C++. Its source code is freely available at \url{https://github.com/pangenome/odgi} with documentation at \url{https://odgi.readthedocs.io}.
-ODGI can be installed via Bioconda \url{https://bioconda.github.io/recipes/odgi/README.html} or Guix https://github.com/ekg/guix-genomics/blob/master/odgi.scm. \\
+Hence, fast and versatile software is required to ask intricate questions of such data in an efficient way. \\
+\textbf{Results:} Here we present ODGI, a new suite of tools that implements scalable algorithms and has an efficient in-memory representation of DNA variation graphs.
+ODGI includes tools for detecting complex regions, extracting \textit{loci}, removing artifacts, exploratory analysis, manipulation, validation, and visualization.
+Its fast parallel execution facilitates routine pangenomic tasks as well as pipelines that can quickly answer complex biological questions of gigabase-scale pangenome graphs.
+\\
+%ODGI supports graphs in the Graphical Fragment Assembly format version 1 (GFAv1).
+%The tools scale efficiently to large collections of eukaryotic genomes, allowing users to work easily with pangenome graphs of tens to thousands of large genomes while maintaining fast execution runtime and low memory overhead.
+\textbf{Availability:} ODGI is ++ software published under the MIT license.
+Source code can be downloaded from \url{https://github.com/pangenome/odgi} while the documentation is available at \url{https://odgi.readthedocs.io}.
+ODGI can be installed via Bioconda \url{https://bioconda.github.io/recipes/odgi/README.html} or Guix \url{https://github.com/ekg/guix-genomics/blob/master/odgi.scm}. \\
 \textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\

### 2021-10-11 manuscript/main.tex
start working through text

@@
 \textbf{Results:} Here we present ODGI, a new suite of tools that implements scalable algorithms and has an efficient in-memory representation of DNA variation graphs.
-ODGI includes tools for detecting complex regions, extracting \textit{loci}, removing artifacts, exploratory analysis, manipulation, validation, and visualization.
+ODGI includes tools for detecting complex regions, extracting loci, removing artifacts, exploratory analysis, manipulation, validation, and visualization.
 Its fast parallel execution facilitates routine pangenomic tasks as well as pipelines that can quickly answer complex biological questions of gigabase-scale pangenome graphs.
@@
 A pangenome models the full set of genomic elements in a given species or clade \citep{Eizenga_2020}.
-Indeed, these data structures encode the mutual relationships between all the genomes represented, in contrast to reference-based approaches which relate sequences only to a particular genome, chosen as reference.
+In contrast to reference-based approaches which relate sequences only to a particular genome, these data structures encode the mutual relationships between all the genomes represented.
 A class of methods to represent pangenomes involves the sequence graphs \citep{Paten:2017}, where similar orthologous regions between genomes collapse into a single representative sequence.
 In node-labeled sequence graphs, nodes indicate DNA sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph.
-A bidirected sequence graph represents both strands of DNA.
-On this model, variation graphs add the concept of path to embed linear sequences (e.g., genomes or haplotypes) into the graph.
+A bidirected sequence graph represents both strands of DNA. On this model, variation graphs add the concept of path to embed linear sequences (e.g., genomes or haplotypes) into the graph.
 Paths provide a stable coordinate system, allowing graph annotations and interoperability between different graphs.
@@
 Several genome data structures were developed to address these challenges \citep{Eizenga_2020_BX}.
-Memory-efficient representations do not allow dynamic updates to the graph, such as adding new genomes or simplifying the graph topology.
-Furthermore, dynamic updates must perform efficiently on the complex graphs generated from whole-genome alignment of vertebrates.
-Highly repetitive regions, such as centromeres, segmental duplications, and acrocentric chromosomes lead to graphs that present paths with very high-depth nodes.
-The node depth is defined as the number of times in which the node is crossed by all the paths embedded in the graph.
-High depth increases the complexity of the operations performed on the paths and the nodes they cross.
+Here, we build on these approaches, developing essential functions for graph interrogation and manipulation, and refining our implementation to scale to the complexity found in T2T genome assemblies produced with modern long read sequencing.
 
-To overcome this limitation, we have developed an Optimized Dynamic Genome/Graph Implementation (ODGI).
-Each tool loads graphs into the succinct dynamic HandleGraph model \citep{Eizenga_2020_BX} and applies methods based on the HandleGraph API.
-Several algorithms are shared with the VG toolkit (VG) \citep{Garrison:2018}, the state-of-the-art toolkit for working with pangenome graphs.
-However, although similar in scope to VG, ODGI importantly differs in its design to support highly complex pangenome graphs.
-The majority of the ODGI tools are implemented in an index-free manner, avoiding the burning task needed to create an index structure.
-Indeed, index-constrained implementations would require building indexes also for single or little operations on the graphs.
-Moreover, thanks to its efficient path representation, ODGI tools can build and walk the internal graph representation (the ODGI format) in parallel, which is one of the major bottlenecks when working with very large and deep variation graphs, efficiently scaling up to hundreds of large vertebrate genomes.
+%that we have found to be essential during our work in the HPRC.
+%Most memory-efficient representations do not allow dynamic updates to the graph, such as adding new genomes or simplifying the graph topology.
+%Furthermore, dynamic updates must perform efficiently on the complex graphs generated from whole-genome alignment of vertebrates.
+%Highly repetitive regions, such as centromeres, segmental duplications, and acrocentric chromosomes lead to graphs that present paths with very high-depth nodes.
+%The node depth is defined as the number of times in which the node is crossed by all the paths embedded in the graph.
+%High depth increases the complexity of the operations performed on the paths and the nodes they cross.
+
+%To overcome this limitation, we have developed an 
+We implement these functions in the Optimized Dynamic Genome/Graph Implementation (ODGI).
+Each tool in ODGI loads graphs into a succinct dynamic HandleGraph model \citep{Eizenga_2020_BX} and applies methods based on the HandleGraph API.
+In addition to reusing and extending algorithms shared with the VG toolkit (VG) \citep{Garrison:2018}, we develop new methods to focused specifically on problems encountered when building pangenome graphs at the scale of vertebrate populations.
+%However, although similar in scope to VG, ODGI importantly differs in its design to support highly complex pangenome graphs.
+To ease interactive use, the majority of the ODGI tools are implemented in an index-free manner, avoiding the need to create index structures at each step of complex graph processing pipelines.
+%Indeed, index-constrained implementations would require building indexes also for single or little operations on the graphs.
+Thanks to ODGI's efficient path representation, its tools can work with variation graphs with highly complex regions.
+This eliminates one of the major bottlenecks when working with very large and deep variation graphs, and allows researchers to build and interrogate graphs of previously-inaccessible complexity and scale.
+%efficiently scaling up to hundreds of large vertebrate genomes.
 
@@
 
-ODGI is designed to build and modify paths in parallel, keeping its representation in memory efficient.
-It applies concepts first introduced in the dynamic version of the Graph BWT \citep{Siren:2020} (gBWT), a scalable implementation of the graph extension of the positional Burrows-Wheeler transform \citep{Durbin_2014}.
-Specifically, we build a node-local model of the graph, where the graph is stored in a vector of node structures.
-The identifier of each node is its offset in this vector plus 1.
-Each structure contains the sequence of the node, its edges in both directions (forward and backward), and a vector of path steps that describes how the embedded paths walk across the node.
-Each step records the path identifier and the previous and next path steps on the same path.
-To reduce the encoding, we encode path steps using a local alphabet that maps the N neighbours of the node into the range $1\ldots N$.
+%ODGI is designed to build and modify paths in parallel, keeping its representation in memory efficient.
+The ODGI toolkit builds on existing approaches to efficiently store and manipulate variation graphs \citep{Eizenga_2020_BX}. %, presenting a refined and well-tested implementation.
+We begin with several key properties which hold for most pangenome variation graphs that we have built: 1) The graphs are relatively sparse, with low node degree; 2) The graphs can be sorted, so that most edges go between nodes that are close together in the sort order (and identifier space); 3) There is high similarity between paths in the graph.
+%We incorporate concepts first introduced in the dynamic version of the Graph BWT \citep{Siren:2020} (GBWT), a scalable implementation of the graph extension of the positional Burrows-Wheeler transform \citep{Durbin_2014}.
+To exploit these properties, we follow a design first explored in the dynamic Graph BWT (GBWT) \citep{Siren:2020} and build a node-centric model of the graph, where the graph is stored in a vector of node structures, each of which presents a node-local view of the graph sequence, topology, and path layout.
+%The identifier of each node is its offset in this vector plus 1.
+Each structure contains the sequence of the node, its edges in both directions (forward and backward), and a vector of path steps that describes the previous and next steps in paths that walk across the node.
+%Each step records the path identifier and the previous and next path steps on the same path.
+As in the GBWT, to compress the encoding, we encode path steps using a local alphabet that maps the N neighbours of the node into the range $1\ldots N$.
 To further save space, node deltas, rather than IDs, are stored in this alphabet.

### 2021-10-11 manuscript/main.tex
text sort / layout section

@@
 Pangenome graphs built from raw sets of alignments may have complex structures which can introduce difficulty in downstream analyses.
-Uncovering their latent topology allows us to understand the sparse structures typically found in pangenome graphs.
+Uncovering their latent topology allows us to understand their typical sparse structures.
 
 \texttt{odgi sort} offers multiple sorting algorithms to find the best node order in one dimension.
-Nodes can be sorted randomly, topologically, by breaking cycles in the graph, by grooming, or by using the novel path-guided stochastic gradient descent (PG-SGD) algorithm.
-can be combined together to a series of sorts.
+Most notably, nodes can be sorted topologically, randomly, by breaking cycles in the graph, by grooming, or by using a novel path-guided stochastic gradient descent (PG-SGD) algorithm.
+The latter one exploits the biological information in the paths for sorting the nodes.
+Combining a series of sorts can mitigate the weaknesses of individually applied sorting algorithms. 
+In Figure \textbf{\textit{TODO MHC wgg.85}}, a comparison of an unsorted human pangenome MHC graph and a sorted one are displayed.
+
+For the establishment of a layout in two dimensions \texttt{odgi layout} applies a 2D adjusted PG-SGD algorithm.
+A calculated layout can be visualized statically with \texttt{odgi draw} (see Figure \textit{\textbf{TODO MHC wgg.85}}) or interactively explored with the in-development \texttt{gfaestus}\footnote{\url{https://github.com/chfi/gfaestus}}.
 
@@
 \texttt{odgi depth} returns the node depth as defined by query criteria, allowing users to assess the complexity of regions in the graph.
-Collapsed sequences with highly identical repeats indicate intricate loci.
+Collapsed sequences with highly identical repeats indicate intricate \textit{loci}. 
 Complex motifs can be also detected with \texttt{odgi degree}, which returns the node degree as defined by query criteria. 
@@
 In Fig.
-\ref{fig:metrics}, the characteristics of a human pangenome graph of the exon 1 huntingtin gene \citep{Sathasivam2013,Neueder2017} (\textit{HTTexon1}), obtained with the ODGI metrics tools, are shown.
+\ref{fig:metrics}, the characteristics of a 90 haplotypes human pangenome graph of the exon 1 huntingtin gene \citep{Sathasivam2013,Neueder2017} (\textit{HTTexon1}), obtained with the ODGI metrics tools, are shown.
 

### 2021-10-11 manuscript/main.tex
work it

--- 2021-10-11 manuscript/main.tex
@@
 %The tools scale efficiently to large collections of eukaryotic genomes, allowing users to work easily with pangenome graphs of tens to thousands of large genomes while maintaining fast execution runtime and low memory overhead.
-\textbf{Availability:} ODGI is ++ software published under the MIT license.
+\textbf{Availability:} ODGI is under the MIT license.
 Source code can be downloaded from \url{https://github.com/pangenome/odgi} while the documentation is available at \url{https://odgi.readthedocs.io}.
@@
 A pangenome models the full set of genomic elements in a given species or clade \citep{Eizenga_2020}.
-In contrast to reference-based approaches which relate sequences only to a particular genome, these data structures encode the mutual relationships between all the genomes represented.
-A class of methods to represent pangenomes involves the sequence graphs \citep{Paten:2017}, where similar orthologous regions between genomes collapse into a single representative sequence.
-In node-labeled sequence graphs, nodes indicate DNA sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph.
-A bidirected sequence graph represents both strands of DNA. On this model, variation graphs add the concept of path to embed linear sequences (e.g., genomes or haplotypes) into the graph.
-Paths provide a stable coordinate system, allowing graph annotations and interoperability between different graphs.
+In contrast to reference-based approaches which relate sequences to a single genome, these data structures encode the mutual relationships between all the genomes represented.
+In pangenome graphs \citep{Paten:2017}, homologous regions between genomes are compressed into a single representative model of all alleles present in the pangenome.
+These flexible models let us encode any kind of variation, allowing the generation of truly unbiased data systems on which to base analyses of genome evolution.
+Although these structures are of clear utility to researchers \citep{cpang2018}, and have been the focus of numerous applications in read mapping \cite{Garrison:2018,Baaijens_2019,Hickey:2020,Sibbesen_2021}, we observe that the community lacks a set of tools focused on graph manipulation and interrogation.
 
-Thanks to advances in sequencing technologies, new full-length genome assemblies are produced at a high rate, including efforst from the Human Pangenome Reference Consortium\footnote{\url{https://humanpangenome.org/} (accessed Oct 2021)} (HPRC) and Telomere-to-Telomere (T2T) projects \citep{Miga:2020, Logsdon_2021, Nurk_2021}, from the T2T consortium\footnote{\url{https://sites.google.com/ucsc.edu/t2tworkinggroup/home} (accessed Oct 2021)}.
+%In node-labeled sequence graphs, nodes indicate DNA sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph.
+
+%A bidirected sequence graph represents both strands of DNA.
+%On this model, variation graphs add the concept of path to embed linear sequences (e.g., genomes or haplotypes) into the graph.
+%Paths provide a stable coordinate system, and allow the graph to support annotations and  different graphs.
+
+%\footnote{\url{https://humanpangenome.org/} (accessed Oct 2021)}
+%\footnote{\url{https://sites.google.com/ucsc.edu/t2tworkinggroup/home} (accessed Oct 2021)}.
+
+Thanks to advances in sequencing technologies, new full-length genome assemblies are produced at a high rate, including efforts from the Human Pangenome Reference Consortium (HPRC) and Telomere-to-Telomere (T2T) projects \citep{Miga:2020, Logsdon_2021, Nurk_2021}, from the T2T consortium.
 This amount of data offers the opportunity to study genomic variation as never before, but it also brings challenges in how to represent and manipulate hundreds of genomes at a gigabase scale.

### 2021-10-11 manuscript/fig_metrics.tex

@@
 %	\includegraphics[width=\linewidth]{fig/metrics/chr4.pan.HTTex1.gfa.multiqc_odgi_stats.png}
-	\caption{Features of a human pangenome graph of the exon 1 huntingtin gene (\textit{HTTexon1}): \textbf{(a)} Excerpt of vital statistics of the \textit{HTTexon1} graph displayed by MultiQC's ODGI module.
+	\caption{Features of a 90 haplotypes human pangenome graph of the exon 1 huntingtin gene (\textit{HTTexon1}): \textbf{(a)} Excerpt of vital statistics of the \textit{HTTexon1} graph displayed by MultiQC's ODGI module.
 The very high GC content of 73.0\% compared to a human genomic mean GC content of 40.9\% \cite{Piovesan2019} is in accordance with the literature (see for example \cite{Sathasivam2013, Neueder2017}).
-\textbf{(b)} \texttt{odgi viz} visualization of the \textit{HTTexon1} graph.
+\textbf{(b)} \texttt{odgi viz} visualization of the 13 largest gene alleles, CHM13, and GRCH38 of the \textit{HTTexon1} graph.
 \textbf{(c)} Per nucleotide node depth distribution of CHM13 in the \textit{HTTexon1} graph.

### 2021-10-11 manuscript/main.tex
keep going on intro

@@
 These flexible models let us encode any kind of variation, allowing the generation of truly unbiased data systems on which to base analyses of genome evolution.
-Although these structures are of clear utility to researchers \citep{cpang2018}, and have been the focus of numerous applications in read mapping \cite{Garrison:2018,Baaijens_2019,Hickey:2020,Sibbesen_2021}, we observe that the community lacks a set of tools focused on graph manipulation and interrogation.
+Although these structures are of clear utility to researchers \citep{cpang2018}, and have been the focus of numerous applications in read mapping \cite{Garrison:2018,Baaijens_2019,Hickey:2020,Sibbesen_2021}, we observe that the community lacks a toolset specifically focused on graph manipulation and interrogation.
 
-%In node-labeled sequence graphs, nodes indicate DNA sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph.
+Projects like the Human Pangenome Reference Consortium (HPRC) and Telomere-to-Telomere (T2T) consortium \citep{Miga:2020, Logsdon_2021, Nurk_2021} have recently demonstrated that high-quality \textit{de novo} assemblies can be routinely generated from third-generation long read sequencing data.
+We anticipate that in the near future, studies will frequently begin from assemblies of similar quality, leading to demand for methods that allow us to understand pangenomes.
+As experience in the HPRC has shown, pangenome graphs are a natural model to support these kinds of analyses, but studies based on pangenome graphs are difficult due to a lack of downstream tools and standard operations.
 
-%A bidirected sequence graph represents both strands of DNA.
-%On this model, variation graphs add the concept of path to embed linear sequences (e.g., genomes or haplotypes) into the graph.
-%Paths provide a stable coordinate system, and allow the graph to support annotations and  different graphs.
+Here, we present the Optimized Dynamic Genome/Graph Implementation (ODGI) toolkit, a genome graph interrogation and transformation system specifically built to handle the data scales encountered when working with pangenomes built from hundreds of haplotype-resolved genomes.
+ODGI provides a set of standard operations on the pangenome variation graph data model, generalizing ``genome arithmetic'' concepts like those found in BEDTools \citep{Quinlan_2010} to work on genome graphs, and providing new operations, such as visualizations, sorting, and liftover projections critical to understanding the structure of pangenome graphs.
+
+%Tools in ODGI operate on an efficient dynamic HandleGraph model \citep{Eizenga_2020_BX}.
+%Algorithms written against this abstract API can be applied to the graph.% applies algorithms based on the HandleGraph API.
+%This common API lets us reuse and and extend algorithms shared with the VG toolkit (VG) \citep{Garrison:2018}.
+%We specifically develop new methodsfocused on problems encountered when building pangenome graphs at the scale of vertebrate populations.
+%To ease interactive use, the majority of the ODGI tools are implemented in an index-free manner, avoiding the need to create index structures at each step of complex graph processing pipelines.
+%Thanks to ODGI's efficient path representation, its tools can work with variation graphs with highly complex regions.
+%This eliminates one of the major bottlenecks when working with very large and deep variation graphs, and allows researchers to build and understand graphs of previously-inaccessible complexity and scale.
 
@@
 %\footnote{\url{https://sites.google.com/ucsc.edu/t2tworkinggroup/home} (accessed Oct 2021)}.
-
-Thanks to advances in sequencing technologies, new full-length genome assemblies are produced at a high rate, including efforts from the Human Pangenome Reference Consortium (HPRC) and Telomere-to-Telomere (T2T) projects \citep{Miga:2020, Logsdon_2021, Nurk_2021}, from the T2T consortium.
-This amount of data offers the opportunity to study genomic variation as never before, but it also brings challenges in how to represent and manipulate hundreds of genomes at a gigabase scale.
-Variation graphs compactly represent the full genetic variation across a population, but graph-based data structures present computational overheads: in addition to the sequences, they must also represent the graph topology.
-Several genome data structures were developed to address these challenges \citep{Eizenga_2020_BX}.
-Here, we build on these approaches, developing essential functions for graph interrogation and manipulation, and refining our implementation to scale to the complexity found in T2T genome assemblies produced with modern long read sequencing.
-
-%that we have found to be essential during our work in the HPRC.
-%Most memory-efficient representations do not allow dynamic updates to the graph, such as adding new genomes or simplifying the graph topology.
-%Furthermore, dynamic updates must perform efficiently on the complex graphs generated from whole-genome alignment of vertebrates.
-%Highly repetitive regions, such as centromeres, segmental duplications, and acrocentric chromosomes lead to graphs that present paths with very high-depth nodes.
-%The node depth is defined as the number of times in which the node is crossed by all the paths embedded in the graph.
-%High depth increases the complexity of the operations performed on the paths and the nodes they cross.
-
-%To overcome this limitation, we have developed an 
-We implement these functions in the Optimized Dynamic Genome/Graph Implementation (ODGI).
-Each tool in ODGI loads graphs into a succinct dynamic HandleGraph model \citep{Eizenga_2020_BX} and applies methods based on the HandleGraph API.
-In addition to reusing and extending algorithms shared with the VG toolkit (VG) \citep{Garrison:2018}, we develop new methods to focused specifically on problems encountered when building pangenome graphs at the scale of vertebrate populations.
-%However, although similar in scope to VG, ODGI importantly differs in its design to support highly complex pangenome graphs.
-To ease interactive use, the majority of the ODGI tools are implemented in an index-free manner, avoiding the need to create index structures at each step of complex graph processing pipelines.
-%Indeed, index-constrained implementations would require building indexes also for single or little operations on the graphs.
-Thanks to ODGI's efficient path representation, its tools can work with variation graphs with highly complex regions.
-This eliminates one of the major bottlenecks when working with very large and deep variation graphs, and allows researchers to build and interrogate graphs of previously-inaccessible complexity and scale.
-%efficiently scaling up to hundreds of large vertebrate genomes.
-
-%\begin{methods}
 

### 2021-10-12 manuscript/main.tex
odgi viz/draw

@@
 
-By visualizing pangenome graphs, we can gain insight into the relationship between genomes.
-odgi viz applies binning and direct rendering to a raster image to generate a 1D representation able to scale up to gigabase pangenome graphs (Figure 1.A).
-However, pangenome graphs may present complex structures to visualize: odgi sort finds the best node order to simplify 1D graph visualization by applying several cutting-edge sorting algorithms.
-odgi draw extends the visualization in 2D, taking in input the layouts built by odgi layout.
-
 Applying odgi stats, users can retrieve metrics describing the graph properties, such as the number of nodes, edges, paths, and graph length.
@@
 
-% viz / draw
-These provide scalable ways of generating human-viewable pictures of the high-level structure of the graph.
 % interesting that we don't go base-level at all... we use vg for that, or STM
+
+Pangenome graph visualization is one of the first steps to gain insight into the mutual relationship between the sequences and their variation.
+\texttt{odgi viz} and \texttt{odgi draw} provide scalable ways of generating human-viewable pictures of the high-level structure of the graph.
+
+\texttt{odgi viz} offers a binned, linearized rendering in 1 dimension (1D).
+
+% Explain odgi viz's visualization
+
+However, complex, nonlinear graph structures are difficult to display and interpret in a convenient low number of dimensions.
+To overcome such a limitation, \texttt{odgi viz} supports multiple visualization modes, making it easy to grasp the properties of the graph and its complexity.
+
+% Explain odgi viz's color by path position
+% Explain odgi viz's color by strandness
+% Explain odgi viz's color by coverage
+
+\texttt{odgi draw} extends the visualization in 2 dimensions (2D), taking in input the layouts built by \texttt{odgi layout}.
+
+% Explain odgi draw's visualization
 

### 2021-10-12 manuscript/fig_metrics.tex
Fixes: in general use input instead of include        put files with dots in braces, see https://tex.stackexchange.com/questions/110513/unknown-graphics-extension-1-png

@@
 		% include first image
-		\includegraphics[width=1.0\linewidth]{fig/metrics/chr4.pan.HTTex1.gfa.multiqc_odgi_stats.png}  
+		\includegraphics[width=1.0\linewidth]{fig/metrics/{chr4.pan.HTTex1.gfa.multiqc_odgi_stats}.png}  
 		\label{fig:metrics-multiqc}
@@
 		% include second image
-		\includegraphics[width=1.0\linewidth, trim=0 0 -1cm 0 ]{fig/metrics/chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.tiny.og.png.svg.pdf}  
+		\includegraphics[width=1.0\linewidth, trim=0 0 -1cm 0 ]{fig/metrics/{chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.tiny.og.png.svg}.pdf}  
 		\label{fig:metrics-viz}
@@
 		% include second image
-		\includegraphics[width=\linewidth,trim=+.225cm 0 +0.425cm +2cm]{fig/metrics/chr4.HTT.chm13.depth.w1.bed.pdf}  
+		\includegraphics[width=\linewidth,trim=+.225cm 0 +0.425cm +2cm]{fig/metrics/{chr4.HTT.chm13.depth.w1.bed}.pdf}  
 		\label{fig:metrics-depth}
@@
 		% include fourth image
-		\includegraphics[width=\linewidth,trim=+.225cm 0 +.425cm +2cm]{fig/metrics/chr4.HTT.chm13.degree.w1.bed.pdf}  
+		\includegraphics[width=\linewidth,trim=+.225cm 0 +.425cm +2cm]{fig/metrics/{chr4.HTT.chm13.degree.w1.bed}.pdf}  
 		\label{fig:metrics-degree}

### 2021-10-13 manuscript/main.tex
Fixes: in general use input instead of include        put files with dots in braces, see https://tex.stackexchange.com/questions/110513/unknown-graphics-extension-1-png
broken

@@
 
-A pangenome models the full set of genomic elements in a given species or clade \citep{Eizenga_2020}.
+A pangenome models the full set of genomic elements in a given species or clade \citep{cpang2018,Eizenga_2020}.
 In contrast to reference-based approaches which relate sequences to a single genome, these data structures encode the mutual relationships between all the genomes represented.
@@
 These flexible models let us encode any kind of variation, allowing the generation of truly unbiased data systems on which to base analyses of genome evolution.
-Although these structures are of clear utility to researchers \citep{cpang2018}, and have been the focus of numerous applications in read mapping \cite{Garrison:2018,Baaijens_2019,Hickey:2020,Sibbesen_2021}, we observe that the community lacks a toolset specifically focused on graph manipulation and interrogation.
+Although these structures are of clear utility to researchers \citep{cpang2018}, and have been the focus of numerous applications in read mapping \citep{Garrison:2018,Baaijens_2019,Hickey:2020,Sibbesen_2021}, we observe that the community lacks a toolset specifically focused on graph manipulation and interrogation.
 
@@
 Here, we present the Optimized Dynamic Genome/Graph Implementation (ODGI) toolkit, a genome graph interrogation and transformation system specifically built to handle the data scales encountered when working with pangenomes built from hundreds of haplotype-resolved genomes.
-ODGI provides a set of standard operations on the pangenome variation graph data model, generalizing ``genome arithmetic'' concepts like those found in BEDTools \citep{Quinlan_2010} to work on genome graphs, and providing new operations, such as visualizations, sorting, and liftover projections critical to understanding the structure of pangenome graphs.
+ODGI provides a set of standard operations on the pangenome variation graph data model, generalizing ``genome arithmetic'' concepts like those found in BEDTools \citep{Quinlan_2010} to work on genome graphs, and providing new operations, such as visualizations, sorting, and liftover projections critical to understanding and exploiting pangenome graphs.
 
@@
 %ODGI is designed to build and modify paths in parallel, keeping its representation in memory efficient.
-The ODGI toolkit builds on existing approaches to efficiently store and manipulate variation graphs \citep{Eizenga_2020_BX}. %, presenting a refined and well-tested implementation.
-We begin with several key properties which hold for most pangenome variation graphs that we have built: 1) The graphs are relatively sparse, with low node degree; 2) The graphs can be sorted, so that most edges go between nodes that are close together in the sort order (and identifier space); 3) There is high similarity between paths in the graph.
+The ODGI toolkit builds on existing approaches to efficiently store and manipulate variation graphs \citep{Garrison:2018,Eizenga_2020_BX}. %, presenting a refined and well-tested implementation.
+Its implementation rests on three key properties which hold for most pangenome variation graphs:
+
+\begin{enumerate}
+\item They are relatively sparse, with low average node degree.
+\item They can be sorted, so that most edges go between nodes that are close together in the sort order (and identifier space).
+\item Paths are similar to each other.
 %We incorporate concepts first introduced in the dynamic version of the Graph BWT \citep{Siren:2020} (GBWT), a scalable implementation of the graph extension of the positional Burrows-Wheeler transform \citep{Durbin_2014}.
-To exploit these properties, we follow a design first explored in the dynamic Graph BWT (GBWT) \citep{Siren:2020} and build a node-centric model of the graph, where the graph is stored in a vector of node structures, each of which presents a node-local view of the graph sequence, topology, and path layout.
+\end{enumerate}
+
+%These properties and this  encourage us to use a node-centric model for the graph, where we apply a local, compressed representation of the node neighborhood.
+
+%These structures are complex, but we can make sense of them so long as we can load and analyze   %In order to make sense of such structures
+%(Although we usually seek to remove as many such motifs as possible through graph normalization, to do so requires that we be able to load and traverse the graph.)
+%Our objective is to allow dynamic modification of graphs even when they contain nodes with very high path depth.
+%This requires an efficient
+
+A key design objective, motivated by experience working with graphs built from high quality eukaryotic genome assemblies, is to support graphs with very high path depth.
+Such motifs can occur in collapsed repeat structures generated by ambiguous sequence homology relationships in repeats found in the centromeres and other segmental duplications.
+To meet these needs, we follow the approach of the dynamic Graph BWT (GBWT) \citep{Siren:2020}, and built a dynamic, compressed model of the paths.
+
+%wherein we save space using a local encoding of paths 
+%We construct a node-centric model of the graph.
+We store the graph in a vector of node structures, each of which presents a node-local view of the graph sequence, topology, and path layout.
 %The identifier of each node is its offset in this vector plus 1.
@@
 
-\include{fig_metrics.tex}
+\input{fig_metrics}
 

### 2021-10-13 manuscript/main.tex
working into the body

@@
 \item They can be sorted, so that most edges go between nodes that are close together in the sort order (and identifier space).
-\item Paths are similar to each other.
+\item Paths are similar to each other---in general, sequences from a species pangenome are similar enough to be easily alignable.
 %We incorporate concepts first introduced in the dynamic version of the Graph BWT \citep{Siren:2020} (GBWT), a scalable implementation of the graph extension of the positional Burrows-Wheeler transform \citep{Durbin_2014}.
@@
 
-%These properties and this  encourage us to use a node-centric model for the graph, where we apply a local, compressed representation of the node neighborhood.
-
-%These structures are complex, but we can make sense of them so long as we can load and analyze   %In order to make sense of such structures
-%(Although we usually seek to remove as many such motifs as possible through graph normalization, to do so requires that we be able to load and traverse the graph.)
-%Our objective is to allow dynamic modification of graphs even when they contain nodes with very high path depth.
-%This requires an efficient
-
-A key design objective, motivated by experience working with graphs built from high quality eukaryotic genome assemblies, is to support graphs with very high path depth.
+These properties have already been used to build efficient dynamic variation graph data structures \citep{Eizenga_2020_BX}.
+%Here, we build on this work and focus on issues that have arisen during our recent work on the Human Pangenome.
+However, we take this work further, focusing on bottlenecks that arose during our work with high-quality assemblies that cover almost all parts of the genome \citep{Logsdon_2021,Nurk_2021}.
+%A key design objective, motivated by experience working with graphs built from high quality eukaryotic genome assemblies, is to support graphs with very high path depth.
+We find that it is necessary to support graphs that have regions of very high path depth.
 Such motifs can occur in collapsed repeat structures generated by ambiguous sequence homology relationships in repeats found in the centromeres and other segmental duplications.
-To meet these needs, we follow the approach of the dynamic Graph BWT (GBWT) \citep{Siren:2020}, and built a dynamic, compressed model of the paths.
+If we cannot process such regions, we have only two options---we can remove these regions, or we can leave them unaligned.
+Neither of these solutions allows us to investigate their biological features.
+To seamlessy represent such difficult regions, we follow the approach of the dynamic Graph BWT (GBWT) \citep{Siren:2020}, and built a node-centric, dynamic, compressed model of the paths.
 

### 2021-10-14 manuscript/fig_metrics.tex
remove { from figure paths
remove dots from figure file names

--- 2021-10-11 manuscript/fig_metrics.tex
@@
 		% include first image
-		\includegraphics[width=1.0\linewidth]{fig/metrics/chr4.pan.HTTex1.gfa.multiqc_odgi_stats.png}  
+		\includegraphics[width=1.0\linewidth]{fig/metrics/chr4_pan_HTTex1_gfa_multiqc_odgi_stats.png}  
 		\label{fig:metrics-multiqc}
@@
 		% include second image
-		\includegraphics[width=1.0\linewidth, trim=0 0 -1cm 0 ]{fig/metrics/chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.tiny.og.png.svg.pdf}  
+		\includegraphics[width=1.0\linewidth, trim=0 0 -1cm 0 ]{fig/metrics/chr4_pan_fa_a2fb268_e820cd3_9ea71d8_smooth_gfa_og_HTTex1_og_O_og_tiny_og_png_svg.pdf}  
 		\label{fig:metrics-viz}
@@
 		% include second image
-		\includegraphics[width=\linewidth,trim=+.225cm 0 +0.425cm +2cm]{fig/metrics/chr4.HTT.chm13.depth.w1.bed.pdf}  
+		\includegraphics[width=\linewidth,trim=+.225cm 0 +0.425cm +2cm]{fig/metrics/chr4_HTT_chm13_depth_w1_bed.pdf}  
 		\label{fig:metrics-depth}
@@
 		% include fourth image
-		\includegraphics[width=\linewidth,trim=+.225cm 0 +.425cm +2cm]{fig/metrics/chr4.HTT.chm13.degree.w1.bed.pdf}  
+		\includegraphics[width=\linewidth,trim=+.225cm 0 +.425cm +2cm]{fig/metrics/chr4_HTT_chm13_degree_w1_bed.pdf}  
 		\label{fig:metrics-degree}

### 2021-10-14 scripts/metrics.sh

@@ -1,7 +1,5 @@
 #!/bin/bash
 
-SVG=$1
-
 wget https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/scratch/2021_07_30_pggb/chroms/chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.gz
 gunzip chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.gz
 wget https://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/genes/hg38.ncbiRefSeq.gtf.gz
@@ -23,7 +21,7 @@ odgi depth -i chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og -
 odgi degree -i chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og -r chm13#chr4:3073405-3073983 -P -t 16 | bedtools makewindows -b /dev/stdin -w 1 > chr4.HTT.chm13.degree.bed
 odgi degree -i chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og -b chr4.HTT.chm13.degree.bed | bedtools sort > chr4.HTT.chm13.degree.w1.bed
 
-Rscript metrics.R chr4.HTT.chm13.depth.w1.bed chr4.HTT.chm13.depth.w1.bed.pdf chr4.HTT.chm13.degree.w1.bed chr4.HTT.chm13.degree.w1.bed.pdf
+Rscript metrics.R chr4.HTT.chm13.depth.w1.bed chr4_HTT_chm13_depth_w1_bed.pdf chr4.HTT.chm13.degree.w1.bed chr4_HTT_chm13_degree_w1_bed.pdf
 
 odgi stats -i chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og -m > chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.stats.yml
 
@@ -57,7 +55,7 @@ odgi viz -i chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.tin
 # check if the svg already exists, else boil out
[…]

### 2021-10-14 manuscript/main.tex
hard going

@@
 
+\section{Model}
+
+A pangenome graph is a sequence model that encodes the mutual alignment of many genomes \citep{Garrison_2019_thesis,Eizenga_2020}.
+In the variation graph, $V = (N, E, P)$, nodes $N = n_1\ldots n_{|N|}$ contain sequences of DNA.
+Each node $n_i$ has an identifier $i$ and an implicit reverse complement $\bar{n_i}$, and a node strand $s$ corresponds to one such orientation.
+Edges $E = e_1\ldots e_{|E|}$ represent ordered pairs of node strands: $e_i = ( s_a, s_b )$.
+Paths $P = p_1\ldots p_{|P|}$ describe walks over node strands: $p_i = s_1 \ldots s_{|p_i|}$.
+When used as a pangenome graph, $V$ expresses sequences, haplotypes, and annotations as paths.
+%The utility of the variation graph model lies in its lossless representation of genomes and their alignment.
+By containing both the sequences and information about their relative variations, the variation graph provides a powerful foundation for many bioinformatic applications.
+
+Such a graph can be constructed by a process like multiple sequence alignment \citep{Lee_2002,Grasso_2004}, but this partially-ordered graph cannot represent complex variation like inversions and cycles that occur at structural scales in genomes.
+General approaches to building pangenome graphs transitively reduce an alignment between sequences to an equivalent, labeled sequence graph \citep{Kehr_2014,Garrison_2019_thesis}.
+Current methods to build these graphs are still in development \citep{Li:2020,Armstrong:2020,pggb}, but they have largely settled on a common data model, represented in the Graphical Fragment Assembly (GFA) format \citep{GFA}).
+%The improvement of these methods depends on tools that allow us to investigate 
+This standardization supports the development of a reference set of tools that operate on the pangenome graph model, while the need to understand the graphs made by these methods motivates it.
+
+%The improvement of these approaches depends on scalable tools for understanding pangenome graphs, a need that our work responds to.
+
+%As a toolkit for understanding these graphs, ODGI
+
+%A more general approach is to build the graph from an alignment between sequences.
+%We first transitively collapse characters in the input sequences that align together into single character in the output genome
+%By projecting input sequences through this transformation, we obtain paths that losslessly encode the original sequences, but in the space of the graph \citep{Garrison_2019_thesis}.
+%Methods to build such graphs are in active development, and here we focus on graphs built with the PanGenome Graph Builder pipeline \citep{pggb}, a process that converts an input pangenome (in FASTA) into a combined graphical model (in the Graphical Fragment Assembly (GFA) format \citep{GFA}).
+
 \section{Implementation}
@@
 %ODGI is designed to build and modify paths in parallel, keeping its representation in memory efficient.
-The ODGI toolkit builds on existing approaches to efficiently store and manipulate variation graphs \citep{Garrison:2018,Eizenga_2020_BX}. %, presenting a refined and well-tested implementation.
+The ODGI toolkit builds on existing approaches to efficiently store and manipulate variation graphs \citep{Garrison:2018,Eizenga_2020_BX}.
+It implements the \textsc{HandleGraph} model \cite{Eizenga_2020_BX}.
 Its implementation rests on three key properties which hold for most pangenome variation graphs:
@@
 \item They can be sorted, so that most edges go between nodes that are close together in the sort order (and identifier space).
-\item Paths are similar to each other---in general, sequences from a species pangenome are similar enough to be easily alignable.
+\item Paths are locally similar to each other, 
 %We incorporate concepts first introduced in the dynamic version of the Graph BWT \citep{Siren:2020} (GBWT), a scalable implementation of the graph extension of the positional Burrows-Wheeler transform \citep{Durbin_2014}.
@@
 
-These properties have already been used to build efficient dynamic variation graph data structures \citep{Eizenga_2020_BX}.
+These properties have already been used to build efficient dynamic variation graph data structures \citep{Siren:2020,Eizenga_2020_BX}.
 %Here, we build on this work and focus on issues that have arisen during our recent work on the Human Pangenome.
-However, we take this work further, focusing on bottlenecks that arose during our work with high-quality assemblies that cover almost all parts of the genome \citep{Logsdon_2021,Nurk_2021}.
+Here, we improve on these prior efforts, focusing on issues that arose during our work with high-quality assemblies that cover almost all parts of the genome \citep{Logsdon_2021,Nurk_2021}.
 %A key design objective, motivated by experience working with graphs built from high quality eukaryotic genome assemblies, is to support graphs with very high path depth.
@@
 To seamlessy represent such difficult regions, we follow the approach of the dynamic Graph BWT (GBWT) \citep{Siren:2020}, and built a node-centric, dynamic, compressed model of the paths.
+This design supports node-local modification and update of the graph, which lets us build and transform the model in parallel.
 
-%wherein we save space using a local encoding of paths 
-%We construct a node-centric model of the graph.
 We store the graph in a vector of node structures, each of which presents a node-local view of the graph sequence, topology, and path layout.
-%The identifier of each node is its offset in this vector plus 1.
 Each structure contains the sequence of the node, its edges in both directions (forward and backward), and a vector of path steps that describes the previous and next steps in paths that walk across the node.
-%Each step records the path identifier and the previous and next path steps on the same path.
 As in the GBWT, to compress the encoding, we encode path steps using a local alphabet that maps the N neighbours of the node into the range $1\ldots N$.
@@
 
-%\end{methods}
-
 ODGI is entirely focused on “first-order” operations on the graph (Table 1), providing, unlike VG, no method to construct graphs from raw sequences or map reads.
@@
 
+\begin{comment}
 %\subsection{Core functionality}
@@
 odgi position finds, translates, and liftovers graph and path positions between different graphs by exploiting their shared path sequences (Figure 1.B).
+\end{comment}
 

### 2021-10-14 manuscript/main.tex
onward

@@
 %ODGI is designed to build and modify paths in parallel, keeping its representation in memory efficient.
-The ODGI toolkit builds on existing approaches to efficiently store and manipulate variation graphs \citep{Garrison:2018,Eizenga_2020_BX}.
-It implements the \textsc{HandleGraph} model \cite{Eizenga_2020_BX}.
-Its implementation rests on three key properties which hold for most pangenome variation graphs:
+The ODGI toolkit builds on existing approaches to efficiently store and manipulate variation graphs \citep{Garrison:2018}.
+%It implements the \textsc{HandleGraph} model \cite{Eizenga_2020_BX}.
+Like other efficient libaries presenting the \textsc{HandleGraph} model \citep{Eizenga_2020_BX}, its implementation rests on three key properties which hold for most pangenome variation graphs:
 
@@
 \item They are relatively sparse, with low average node degree.
-\item They can be sorted, so that most edges go between nodes that are close together in the sort order (and identifier space).
-\item Paths are locally similar to each other, 
+\item They can be sorted so that most edges go between nodes that are close together in the sort order.
+\item Paths are locally similar to each other.
 %We incorporate concepts first introduced in the dynamic version of the Graph BWT \citep{Siren:2020} (GBWT), a scalable implementation of the graph extension of the positional Burrows-Wheeler transform \citep{Durbin_2014}.
@@
 These properties have already been used to build efficient dynamic variation graph data structures \citep{Siren:2020,Eizenga_2020_BX}.
+Sparsity (1) allows us to encode edges $E$ using adjacency lists rather than matricies or hash tables.
+The local linear structure of the graph (2) lets us assign node identifiers that increase along the linear components of the graph, which supports a compact storage of edges and path steps as relativistic (usually small) differences rather than absolute (always large) integer identifiers.
+Path similarity (3) allows us to write local compressors that reduce the storage cost of collections of path steps.
+
 %Here, we build on this work and focus on issues that have arisen during our recent work on the Human Pangenome.
@@
 We store the graph in a vector of node structures, each of which presents a node-local view of the graph sequence, topology, and path layout.
+Expressed in terms of the variation graph $V$, the structure for node  the node sequence $seq(n_i)$
+
 Each structure contains the sequence of the node, its edges in both directions (forward and backward), and a vector of path steps that describes the previous and next steps in paths that walk across the node.

### 2021-10-14 manuscript/main.tex
going for node struct model
make the node struct

@@
 \usepackage{subcaption}
+\usepackage{algorithm2e}
+
+\SetAlgoLined
+\SetKwProg{MyStruct}{Struct}{ contains}{end}
 
@@
 We store the graph in a vector of node structures, each of which presents a node-local view of the graph sequence, topology, and path layout.
-Expressed in terms of the variation graph $V$, the structure for node  the node sequence $seq(n_i)$
+Expressed in terms of the variation graph $V$, ODGI's core $Node$ structure includes.
+
+\begin{algorithm}
+\MyStruct{Node}{
+    \textbf{id} $\in \mathbb{N}$ \tcp{an identifier}
+    \textbf{lock} \tcp{atomic locking primitive}
+    \textbf{sequence} $= [ATGC]*$ \tcp{DNA, RNA, protein}
+    \tcp{succinct dynamic bitvector of edges}
+    \textbf{edges} $= (s_{x},s_{y})* : x \land y \in$ decoding \\
+    \tcp{a vector of id deltas}
+    \textbf{decoding} $= (1 \ldots x \ldots N)*$ \\
+    \tcp{a vector of path steps}
+    \textbf{paths}
+}
+\caption{ODGI's relativistically-packed node structure.}
+\end{algorithm}
+
+\begin{comment}
+TODO:
+    struct step_t {
+        uint64_t path_id;
+        bool is_rev;
+        bool is_start;
+        bool is_end;
+        uint64_t prev_id;
+        uint64_t prev_rank;
+        uint64_t next_id;
+        uint64_t next_rank;
+    };
+\end{comment}
+
+%\begin{enumerate}
+%\item a node sequence $seq(n_i)$
+%\end{enumerate}
 

### 2021-10-14 manuscript/main.tex
sharing the magic

--- 2021-10-14 manuscript/main.tex
@@
 We store the graph in a vector of node structures, each of which presents a node-local view of the graph sequence, topology, and path layout.
-Expressed in terms of the variation graph $V$, the structure for node  the node sequence $seq(n_i)$
+Expressed in terms of the variation graph $V$, ODGI's core $Node$ structure includes a decoder that maps the neighbors of each node to a dense range of integers.
+For a given node $i$ and neighbor $j$, the decoder itself does not store the $id$s of $j$, but rather $Node_i.id - Node_j.id$.
+This keeps even the size of the encoding small, per common variation graph property (2).
+
+\begin{algorithm}
+\MyStruct{Node}{
+    \textbf{id} $\in \mathbb{N}$ \tcp{an identifier}
+    \textbf{lock} \tcp{atomic locking primitive}
+    \textbf{sequence} $= [$A$|$T$|$G$|$C$]*$ \tcp{DNA, RNA, protein}
+    \tcp{bit-packed vector of edges}
+    \textbf{edges} $= (s_{x},s_{y})* : x \land y \in [1\ldots N]$ \\
+    \tcp{bit-packed vector of id deltas}
+    \textbf{decoding} $x_i \ldots x_N \in \mathbb{N}^N$ \\
+    \tcp{bit-packed vector of path steps}
+    \textbf{paths} $[Step_1 \ldots Step_M]*$
+}
+\MyStruct{Step}{
+    \textbf{path\_id} $\in \mathbb{N}$ \tcp{the path's global id}
+    \textbf{is\_rev} $\in ( 0, 1 )$ \tcp{the step orientation}
+    \textbf{is\_start} $\in ( 0, 1 )$ \tcp{if first step in path}
+    \textbf{is\_end} $\in ( 0, 1 )$ \tcp{if last step in path}
+    \textbf{prev$_\delta$} $\in [1\ldots N]$ \tcp{$\delta$-encoded previous node}
+    \textbf{prev\_rank} $\in \mathbb{N}$ \tcp{step rank on previous node}
+    \textbf{next$_\delta$} $\in [1\ldots N]$ \tcp{$\delta$-encoded previous node}
+    \textbf{next\_rank} $\in \mathbb{N}$ \tcp{step rank on next node}
+}
+\caption{ODGI's relativistically-packed node structure and the step structure used to represent the paths as doubly-linked lists.}
+\end{algorithm}
+
+%\begin{enumerate}
+%\item a node sequence $seq(n_i)$
+%\end{enumerate}
 

### 2021-10-18 manuscript/main.tex
working through the results intro

@@
 \newcommand{\vocab}{\textbf}
+
+\def\labelitemi{--}
 
@@
 Expressed in terms of the variation graph $V$, ODGI's core $Node$ structure includes a decoder that maps the neighbors of each node to a dense range of integers.
-For a given node $i$ and neighbor $j$, the decoder itself does not store the $id$s of $j$, but rather $Node_i.id - Node_j.id$.
+For a given node $i$ and neighbor $j$, the decoder itself does not store the $id$s of $j$, but rather a compact representation of the relative difference between the node ids: $\delta = Node_i.id - Node_j.id$.
 This keeps even the size of the encoding small, per common variation graph property (2).
+We define the edges and path steps traversing the node in terms of this alphabet of $\delta$'s.
+The structures in algorithm \ref{alg:structs} describes our encoding.
 
@@
     \tcp{bit-packed vector of edges}
-    \textbf{edges} $= (s_{x},s_{y})* : x \land y \in [1\ldots N]$ \\
+    \textbf{edges} $= (x_i,x_j)* : (i, j) \in [1\ldots \Sigma]^2$ \\
     \tcp{bit-packed vector of id deltas}
-    \textbf{decoding} $x_i \ldots x_N \in \mathbb{N}^N$ \\
+    \textbf{decoding} $x_1 \ldots x_{\Sigma} \in \mathbb{N}^\Sigma$ \\
     \tcp{bit-packed vector of path steps}
-    \textbf{paths} $[Step_1 \ldots Step_M]*$
+    \textbf{path\_steps} $[Step_1 \ldots Step_\Omega]*$
 }
@@
     \textbf{is\_end} $\in ( 0, 1 )$ \tcp{if last step in path}
-    \textbf{prev$_\delta$} $\in [1\ldots N]$ \tcp{$\delta$-encoded previous node}
+    \textbf{prev\_$\delta$} $\in [1\ldots \Sigma]$ \tcp{$\delta$-encoded previous node}
     \textbf{prev\_rank} $\in \mathbb{N}$ \tcp{step rank on previous node}
-    \textbf{next$_\delta$} $\in [1\ldots N]$ \tcp{$\delta$-encoded previous node}
+    \textbf{next\_$\delta$} $\in [1\ldots \Sigma]$ \tcp{$\delta$-encoded previous node}
     \textbf{next\_rank} $\in \mathbb{N}$ \tcp{step rank on next node}
 }
-\caption{ODGI's relativistically-packed node structure and the step structure used to represent the paths as doubly-linked lists.}
+\caption{ODGI's relativistically-packed $Node$ structure and the $Step$ structure used to represent the paths as doubly-linked lists.}
+\label{alg:structs}
 \end{algorithm}
 
-%\begin{enumerate}
-%\item a node sequence $seq(n_i)$
-%\end{enumerate}
+Each structure contains the sequence of the node ($Node_i.sequence$), its edges in both directions ($Node_i.edges$), and a vector of path steps that describes the previous and next steps in paths that walk across the node ($Node_i.path\_steps$).
+%As in the GBWT, to compress the encoding, we encode path steps using a local alphabet that maps the N neighbours of the node into the range $1\ldots N$.
+%To further save space, node deltas, rather than IDs, are stored in this alphabet.
+%The delta between two nodes is defined as their distance in the graph vector (i.e., the difference between the node offsets).
+For efficiency $Node_i.sequence$ is stored as a plain string, while the $edges$ and $path\_steps$ are stored using a dynamic succinct integer vector that requires $O(2nw)$ bits for the edges and $O(5nw)$ bits for the path steps, where $w$ is $\approx log_2(n)$ \citep{prezza2017framework}.
 
-Each structure contains the sequence of the node, its edges in both directions (forward and backward), and a vector of path steps that describes the previous and next steps in paths that walk across the node.
-As in the GBWT, to compress the encoding, we encode path steps using a local alphabet that maps the N neighbours of the node into the range $1\ldots N$.
-To further save space, node deltas, rather than IDs, are stored in this alphabet.
-The delta between two nodes is defined as their distance in the graph vector (i.e., the difference between the node offsets).
-The path step vector, of length n, is stored using a dynamic succinct integer vector that requires O(5nw) bits, where w is ~log2(N).
-Finally, to allow operating on path steps in parallel, each node structure includes a byte-width mutex lock.
-
-ODGI is entirely focused on “first-order” operations on the graph (Table 1), providing, unlike VG, no method to construct graphs from raw sequences or map reads.
-It supports interrogation of the graph for basic dimensional properties, positional and path-related queries, and visualization in one-dimension (1D) and two-dimensions (2D).
-ODGI also supports operations on the graph such as subset, subdivide, break, combine, normalize, or order its components, nodes, and paths.
-Most of the tools are designed to be applied together, piping the output of one tool into the next, thereby preventing the creation of intermediate files, and reducing the number of IO operations.
+To allow edit operations in parallel, each node structure includes a byte-width mutex $lock$.
+All edit operations on the graph must touch at most two $Node$ structs at a time.
+(Both our edge and path step representations are doubly-linked.)
+To avoid deadlocks, we acquire the node locks in ascending $Node.id$ order and release them in desceding order.
+In addition to node-local features of the graph, we must maintain some global information.
+Specifically, we record the start and end of paths, as well as a name to path id mapping in lock-free hashtables.
+The use of lock-free hashtables lets us avoid a global lock when looking up path or graph metadata, which would quickly become a bottleneck during parallel operations on the graph.
+By avoiding global locks, we implement many of the operations in ODGI using maximum parallelism available.
+This ``atomic'' approach to edit operations on the graph is key to enabling our method to scale to the largest genome graphs that we can currently build.
 
@@
 
+ODGI provides a set of first-order interrogative and manipulative operations on pangenome graphs.
+We have established these tools to support our exploration of graphs built from hundreds of large eukaryotic genomes.
+They are practical and able to work with high levels of graph complexity, even with regions of very high path depth (10$^5$ to 10$^6$-fold coverage).
+To further our understanding of these graphs, we have developed a set of methods that let us inspect and manipulate them.
+
+Low-dimensional projections and related visualizations support a complete understanding of the graph.
+To provide a universal coordinate space for the graph, we project its nodes into a low-dimensional space (in practice, 1 or 2 dimensions) by learning the projection via a HOGWILD!, path-guided stochastic gradient descent (PG-SGD) model.
+This approach updates node relative positions to best-match their distance in the paths running through the graph.
+Based on this projection, we can trivially generate 1D and 2D visualizations.
+We can also detect regions whose layout is distorted, and which thus may represent structural variation or assembly error.
+%To our knowledge, these methods are the first to trivially scale to operate on large base-level graphs of this type.
+
+%To foster a high-level understanding of large graphs, we pursue a novel approach to visualization.
+%ODGI provides a linear-time 1D visualization that easily renders gigabase-scale pangenomes in a human-interpretable format suitable for understanding the topology and genome releationships in the pangenome graph.
+%A similar 2D visualization is more costly to compute, but similarly provides linear-time scaling, allowing us to apply it to large pangenomes.
+%This removes a major bottleneck in
+
+Subsetting operations allow us to zoom in on particular genes or regions of interest for more precise examination.
+
+We extract key information about the graph using its paths as a pangenome reference system.
+
+We find that, to understanding relationships between sequences and positions in pangenome graphs, we must use context mapping to ``untangle'' complex regions found in VNTRs, segmental duplications, and centromeric repeats.
+This approach lets use pangenome graphs as a foundation for universal liftover between any two sets of sequences in the graph.
+
+We begin with a set of HandleGraph-based algorithms first establish in the VG toolkit.
+These include algorithms for graph traversal, partition-finding, $k$-mer and character enumeration, sorting, and pruning.
+However, ODGI is more focused than VG.
+Unlike VG, we provide no methods to construct graphs, map reads to the graph, or derive variants.
+
+%It supports interrogation of the graph for basic dimensional properties, positional and path-related queries, and visualization in one-dimension (1D) and two-dimensions (2D).
+%ODGI also supports operations on the graph such as subset, subdivide, break, combine, normalize, or order its components, nodes, and paths.
+%Most of the tools are designed to be applied together, piping the output of one tool into the next, thereby preventing the creation of intermediate files, and reducing the number of IO operations.
+
 \subsection{Building the \textsc{ODGI} model}
@@
 Figure: .dot showing the relationship between all the tools and the data formats we use.
+
+\subsection{Extracting regions of interest}
+
+% andrea (?)
+
+% extract
+Pangenome graphs are very large, but we often only want to work with a small portion (e.g. a single gene).
+We can extract such regions using coordinates on the paths in the graph to guide us.
+
+Figure showing extraction process---either schematic or ``real''.
 
@@
 Figure here with 1D + 2D vizs.
+
+\subsection{Untangling and navigating the pangenome with path-based coordinates}
+
+% erik
+
+% position / tips
+% untangle (this follows on the jaccard graph mapping concepts in the previous section}
+The graph is a model of an alignment of many genomes.
+We can use coordinates in any genome to refer to it, but this requires a few basic operations ...
+Obtaining unambiguous mappings between different genomes requires the use of a new kind of graph based mapping (path graph jaccard).
+
+Figure showing the untangling of C4 ?
+Figure showing tips over a single human chromosome from HPRC.
 
@@
 These are commonly-needed basic operations on the topology of the graph.
-
-\subsection{Untangling and navigating the pangenome with path-based coordinates}
-
-% erik
-
-% position / tips
-% untangle (this follows on the jaccard graph mapping concepts in the previous section}
-The graph is a model of an alignment of many genomes.
-We can use coordinates in any genome to refer to it, but this requires a few basic operations ...
-Obtaining unambiguous mappings between different genomes requires the use of a new kind of graph based mapping (path graph jaccard).
-
-Figure showing the untangling of C4 ?
-Figure showing tips over a single human chromosome from HPRC.
-
-\subsection{Extracting regions of interest}
-
-% andrea (?)
-
-% extract
-Pangenome graphs are very large, but we often only want to work with a small portion (e.g. a single gene).
-We can extract such regions using coordinates on the paths in the graph to guide us.
-
-Figure showing extraction process---either schematic or ``real''.
 

### 2021-10-18 manuscript/main.tex
Added textual output types
integrate odgi tools figure

@@
 
-Low-dimensional projections and related visualizations support a complete understanding of the graph.
+A few design considerations shape these tools.
+To navigate the graph, we need a coordinate system.
+ODGI uses the embedded pangenome sequences to provide a set of coordinates in a standard, 1D format.
+These coordinates depend on the sequences, and are graph-independent.
+
 To provide a universal coordinate space for the graph, we project its nodes into a low-dimensional space (in practice, 1 or 2 dimensions) by learning the projection via a HOGWILD!, path-guided stochastic gradient descent (PG-SGD) model.
@@
 We can also detect regions whose layout is distorted, and which thus may represent structural variation or assembly error.
+ODGI can project vector and matrix representations of the graph relative to these learned coordinate spaces.
+
+In addition to these projections, we use the paths in the graph to provide a universal coordinate space.
+It also supports several kinds of ``lift-over'' of coordinates between different genomes.
+These allow for the direct translation of single coordinates and ranges, in an interactive way.
+And, they let us extract a pairwise 1:many alignment between a given set of ``query'' sequences and a given set of ``target'' or reference sequences.
+This lets us convert the graph to lift-over maps compatible with standard software for projecting annotations and alignments from one genome to another.
+
 %To our knowledge, these methods are the first to trivially scale to operate on large base-level graphs of this type.
@@
 %Most of the tools are designed to be applied together, piping the output of one tool into the next, thereby preventing the creation of intermediate files, and reducing the number of IO operations.
+
+\input{fig_operations}
 

### 2021-10-18 manuscript/main.tex
Remove dot output
another potential discussion point
Intro: minor edits and comments

@@
 \usepackage{algorithm2e}
+\usepackage[usenames,dvipsnames]{xcolor}
 
@@
 \newcommand{\vocab}{\textbf}
+\newcommand{\red}[1]{{\textcolor{Red}{#1}}}
+\newcommand{\FIXME}[1]{\red{[FIXME: #1]}}
 
@@
 In pangenome graphs \citep{Paten:2017}, homologous regions between genomes are compressed into a single representative model of all alleles present in the pangenome.
-These flexible models let us encode any kind of variation, allowing the generation of truly unbiased data systems on which to base analyses of genome evolution.
+These flexible models let us encode any kind of variation, allowing the generation of truly unbiased \FIXME{unbiased needs a little explanation} data systems on which to base analyses of genome evolution.
 Although these structures are of clear utility to researchers \citep{cpang2018}, and have been the focus of numerous applications in read mapping \citep{Garrison:2018,Baaijens_2019,Hickey:2020,Sibbesen_2021}, we observe that the community lacks a toolset specifically focused on graph manipulation and interrogation.
 
-Projects like the Human Pangenome Reference Consortium (HPRC) and Telomere-to-Telomere (T2T) consortium \citep{Miga:2020, Logsdon_2021, Nurk_2021} have recently demonstrated that high-quality \textit{de novo} assemblies can be routinely generated from third-generation long read sequencing data.
-We anticipate that in the near future, studies will frequently begin from assemblies of similar quality, leading to demand for methods that allow us to understand pangenomes.
-As experience in the HPRC has shown, pangenome graphs are a natural model to support these kinds of analyses, but studies based on pangenome graphs are difficult due to a lack of downstream tools and standard operations.
+The Human Pangenome Reference Consortium (HPRC) and Telomere-to-Telomere (T2T) consortium \citep{Miga:2020, Logsdon_2021, Nurk_2021} have recently demonstrated that high-quality \textit{de novo} assemblies can be routinely generated from third-generation long read sequencing data.
+We anticipate that in the near future, studies will frequently begin from assemblies of similar quality, leading to demand for methods that allow us to create and explore pangenomes.
+As experience \FIXME{this is a bit odd} in the HPRC has shown, pangenome graphs are a natural model to support these kinds of analyses, but studies based on pangenome graphs are difficult due to a lack of downstream tools and standard operations.
 
-Here, we present the Optimized Dynamic Genome/Graph Implementation (ODGI) toolkit, a genome graph interrogation and transformation system specifically built to handle the data scales encountered when working with pangenomes built from hundreds of haplotype-resolved genomes.
+Here, we present the Optimized Dynamic Genome/Graph Implementation (ODGI) toolkit, a pangenome graph interrogation and transformation system specifically built to handle the data scales encountered when working with pangenomes built from hundreds of haplotype-resolved genomes.
 ODGI provides a set of standard operations on the pangenome variation graph data model, generalizing ``genome arithmetic'' concepts like those found in BEDTools \citep{Quinlan_2010} to work on genome graphs, and providing new operations, such as visualizations, sorting, and liftover projections critical to understanding and exploiting pangenome graphs.
@@
 
+\texttt{odgi view} can output GFA and internal graph representations \\
+\texttt{odgi stats} outputs statistics in TSV or YAML. \\
+\texttt{odgi bin} can output JSON. \\
+\texttt{odgi degree/position/depth/untangle/tips} can output BED. \\
+\texttt{odgi untangle} outputs PAF.
+
 Figure: .dot showing the relationship between all the tools and the data formats we use.
@@
 
+is ODGI ready for scWGS data? potentially, each cell is a path in a graph!! very very high path depth
+
 some cancer stuff?!

### 2021-10-18 manuscript/main.tex
Model: minor edits and comments

@@
 
-Such a graph can be constructed by a process like multiple sequence alignment \citep{Lee_2002,Grasso_2004}, but this partially-ordered graph cannot represent complex variation like inversions and cycles that occur at structural scales in genomes.
+Such a graph can be constructed by multiple sequence alignment \citep{Lee_2002,Grasso_2004}, but the resulting partially-ordered graph cannot represent complex variation like inversions and cycles that occur at structural scales in genomes.
 General approaches to building pangenome graphs transitively reduce an alignment between sequences to an equivalent, labeled sequence graph \citep{Kehr_2014,Garrison_2019_thesis}.
@@
 This standardization supports the development of a reference set of tools that operate on the pangenome graph model, while the need to understand the graphs made by these methods motivates it.
+\FIXME{Sentence is a bit odd}.
 

### 2021-10-18 manuscript/main.tex
Results: minor edits and comments

@@
 %It implements the \textsc{HandleGraph} model \cite{Eizenga_2020_BX}.
-Like other efficient libaries presenting the \textsc{HandleGraph} model \citep{Eizenga_2020_BX}, its implementation rests on three key properties which hold for most pangenome variation graphs:
+Similar to other efficient libaries presenting the \textsc{HandleGraph} model \citep{Eizenga_2020_BX}, ODGI's implementation rests on three key properties which hold for most pangenome variation graphs:
 
@@
 \item They can be sorted so that most edges go between nodes that are close together in the sort order.
-\item Paths are locally similar to each other.
+\item They contain paths are locally similar to each other.
 %We incorporate concepts first introduced in the dynamic version of the Graph BWT \citep{Siren:2020} (GBWT), a scalable implementation of the graph extension of the positional Burrows-Wheeler transform \citep{Durbin_2014}.
@@
 
-These properties have already been used to build efficient dynamic variation graph data structures \citep{Siren:2020,Eizenga_2020_BX}.
-Sparsity (1) allows us to encode edges $E$ using adjacency lists rather than matricies or hash tables.
+These properties are used to build efficient dynamic variation graph data structures \citep{Siren:2020,Eizenga_2020_BX}.
+Sparsity (1) allows us to encode edges $E$ using adjacency lists rather than matrices or hash tables.
 The local linear structure of the graph (2) lets us assign node identifiers that increase along the linear components of the graph, which supports a compact storage of edges and path steps as relativistic (usually small) differences rather than absolute (always large) integer identifiers.
 Path similarity (3) allows us to write local compressors that reduce the storage cost of collections of path steps.
+\FIXME{without a figure we may lose the reader here a little}
 
 %Here, we build on this work and focus on issues that have arisen during our recent work on the Human Pangenome.
-Here, we improve on these prior efforts, focusing on issues that arose during our work with high-quality assemblies that cover almost all parts of the genome \citep{Logsdon_2021,Nurk_2021}.
+ODGI improves on prior efforts, based on issues that arose during our work with high-quality assemblies that cover almost all parts of the genome \citep{Logsdon_2021,Nurk_2021}.
 %A key design objective, motivated by experience working with graphs built from high quality eukaryotic genome assemblies, is to support graphs with very high path depth.
-We find that it is necessary to support graphs that have regions of very high path depth.
+For example, we find that it is necessary to support graphs that have regions of very high path depth.
 Such motifs can occur in collapsed repeat structures generated by ambiguous sequence homology relationships in repeats found in the centromeres and other segmental duplications.
@@
 Neither of these solutions allows us to investigate their biological features.
-To seamlessy represent such difficult regions, we follow the approach of the dynamic Graph BWT (GBWT) \citep{Siren:2020}, and built a node-centric, dynamic, compressed model of the paths.
+To seamlessy represent such difficult regions, we follow \FIXME{or introduce?} the approach of the dynamic Graph BWT (GBWT) \citep{Siren:2020}, and built a node-centric, dynamic, compressed model of the paths.
+\FIXME{So this has never been done before?}
 This design supports node-local modification and update of the graph, which lets us build and transform the model in parallel.
+\FIXME{Maybe plug ODGI for we?}
 
 We store the graph in a vector of node structures, each of which presents a node-local view of the graph sequence, topology, and path layout.
+\FIXME{Some repetition here?}
 Expressed in terms of the variation graph $V$, ODGI's core $Node$ structure includes a decoder that maps the neighbors of each node to a dense range of integers.
@@
 This ``atomic'' approach to edit operations on the graph is key to enabling our method to scale to the largest genome graphs that we can currently build.
+\FIXME{This is great!}
 
@@
 We have established these tools to support our exploration of graphs built from hundreds of large eukaryotic genomes.
+\FIXME{why not just say human, mouse, \ldots?}
 They are practical and able to work with high levels of graph complexity, even with regions of very high path depth (10$^5$ to 10$^6$-fold coverage).
@@
 To provide a universal coordinate space for the graph, we project its nodes into a low-dimensional space (in practice, 1 or 2 dimensions) by learning the projection via a HOGWILD!, path-guided stochastic gradient descent (PG-SGD) model.
+\FIXME{Cite or explain?!}
 This approach updates node relative positions to best-match their distance in the paths running through the graph.
@@
 Unlike VG, we provide no methods to construct graphs, map reads to the graph, or derive variants.
+\FIXME{Above switches from ODGI to WE --- why not hold on to the former?}
 

### 2021-10-18 manuscript/main.tex
Discussion: minor edits and comments
Note on pggb

@@
 
+\FIXME{Would it be an idea to put part of pggb in a box and explain steps? That way people can visualize using the toolkit and we can refer to it in the text}
+
 \subsection{Building the \textsc{ODGI} model}
@@
 (recent) interactive visualization tools?! CITE PANACHE, also indicate that no published tool for interactive exploration of such graphs exist
+\FIXME{mention gfaestus?}
 
@@
 some cancer stuff?!
+
+\FIXME{I think we can leave RDF and FAIR out as it does not really relate to performance and ODGI explorations.
+Cancer, yes, we could state that the tools are applicable to cancer genome exploration as wild variation can be represented :)}
 

### 2021-10-19 manuscript/main.tex
lots of stuff

@@
 In pangenome graphs \citep{Paten:2017}, homologous regions between genomes are compressed into a single representative model of all alleles present in the pangenome.
-These flexible models let us encode any kind of variation, allowing the generation of truly unbiased \FIXME{unbiased needs a little explanation} data systems on which to base analyses of genome evolution.
+These flexible models let us encode any kind of variation, allowing the generation of comprehensive data systems on which to base analyses of genome evolution.
 Although these structures are of clear utility to researchers \citep{cpang2018}, and have been the focus of numerous applications in read mapping \citep{Garrison:2018,Baaijens_2019,Hickey:2020,Sibbesen_2021}, we observe that the community lacks a toolset specifically focused on graph manipulation and interrogation.
@@
 We anticipate that in the near future, studies will frequently begin from assemblies of similar quality, leading to demand for methods that allow us to create and explore pangenomes.
-As experience \FIXME{this is a bit odd} in the HPRC has shown, pangenome graphs are a natural model to support these kinds of analyses, but studies based on pangenome graphs are difficult due to a lack of downstream tools and standard operations.
+Pangenome graphs are a natural model to support these kinds of analyses, but studies based on pangenome graphs are difficult due to a lack of downstream tools and standard operations.
 
@@
 %The improvement of these methods depends on tools that allow us to investigate
-This standardization supports the development of a reference set of tools that operate on the pangenome graph model, while the need to understand the graphs made by these methods motivates it.
-\FIXME{Sentence is a bit odd}.
+This standardization supports the development of a reference set of tools that operate on the pangenome graph model.
+%The need to understand the graphs made by these methods motivates it.
+We require such a toolkit to understand the graphs made by these methods.
 
@@
 \item They can be sorted so that most edges go between nodes that are close together in the sort order.
-\item They contain paths are locally similar to each other.
+\item Their embedded paths are locally similar to each other.
 %We incorporate concepts first introduced in the dynamic version of the Graph BWT \citep{Siren:2020} (GBWT), a scalable implementation of the graph extension of the positional Burrows-Wheeler transform \citep{Durbin_2014}.
@@
 Path similarity (3) allows us to write local compressors that reduce the storage cost of collections of path steps.
-\FIXME{without a figure we may lose the reader here a little}
+%\FIXME{without a figure we may lose the reader here a little} -- see algorithm 1
 
@@
 Neither of these solutions allows us to investigate their biological features.
-To seamlessy represent such difficult regions, we follow \FIXME{or introduce?} the approach of the dynamic Graph BWT (GBWT) \citep{Siren:2020}, and built a node-centric, dynamic, compressed model of the paths.
-\FIXME{So this has never been done before?}
+To seamlessy represent such difficult regions, we follow an approach implemented in the dynamic version of the Graph BWT (GBWT) \citep{Siren:2020}, and built a node-centric, dynamic, compressed model of the paths.
 This design supports node-local modification and update of the graph, which lets us build and transform the model in parallel.
-\FIXME{Maybe plug ODGI for we?}
 
 We store the graph in a vector of node structures, each of which presents a node-local view of the graph sequence, topology, and path layout.
-\FIXME{Some repetition here?}
 Expressed in terms of the variation graph $V$, ODGI's core $Node$ structure includes a decoder that maps the neighbors of each node to a dense range of integers.
@@
 This ``atomic'' approach to edit operations on the graph is key to enabling our method to scale to the largest genome graphs that we can currently build.
-\FIXME{This is great!}
 
@@
 
+\input{fig_operations}
+
 \section{Results}
@@
 We have established these tools to support our exploration of graphs built from hundreds of large eukaryotic genomes.
-\FIXME{why not just say human, mouse, \ldots?}
 They are practical and able to work with high levels of graph complexity, even with regions of very high path depth (10$^5$ to 10$^6$-fold coverage).
-To further our understanding of these graphs, we have developed a set of methods that let us inspect and manipulate them.
+ODGI's tools cover common operations that we have found to be essential when working with these complex biological models:
 
-A few design considerations shape these tools.
-To navigate the graph, we need a coordinate system.
-ODGI uses the embedded pangenome sequences to provide a set of coordinates in a standard, 1D format.
+\begin{itemize}
+\item \textit{odgi build} constructs the ODGI data model from GFA (\S\ref{sec:build}).
+\item \textit{odgi view} converts the graph into standard text formats (\S\ref{sec:text}).
+\item \textit{odgi stats} provides numerical properties of the graph (\S\ref{sec:text}).
+\item \textit{odgi matrix} derives the pangenome matrix (\S\ref{sec:text}).
+\item \textit{odgi extract} subsets the graph based on path ranges (\S\ref{sec:extract}).
+\item \textit{odgi chop} breaks long nodes into shorter ones (\S\ref{sec:edit}).
+\item \textit{odgi unchop} combines redundant nodes (\S\ref{sec:edit}).
+\item \textit{odgi explode} breaks the graph into connected components (\S\ref{sec:edit}).
+\item \textit{odgi squeeze} unifies disjoint graphs (\S\ref{sec:edit}).
+\item \textit{odgi prune} removes complex regions (\S\ref{sec:edit}).
+\item \textit{odgi viz} provides a scalable 1D rendering of the graph (\S\ref{sec:viz}).
+\item \textit{odgi untangle} deconvolutes paths relative to a reference (\S\ref{sec:untangle}).
+\item \textit{odgi sort} orders the graph with sorting pipelines (\S\ref{sec:sort}).
+\item \textit{odgi layout} establishes a 2D layout using SGD (\S\ref{sec:sort}).
+\item \textit{odgi depth} describes path depth over the graph (\S\ref{sec:metrics}).
+\item \textit{odgi tips} finds path end points relative to a reference (\S\ref{sec:metrics}).
+\end{itemize}
+
+%To further our understanding of these graphs, we have developed a set of methods that let us inspect and manipulate them.
+
+%We begin with a set of HandleGraph-based algorithms first establish in the VG toolkit \citep{Garrison:2018}.
+%These include algorithms for graph traversal, partition-finding, $k$-mer and character enumeration, sorting, and pruning.
+%Unlike VG, we provide no methods to construct graphs, map reads to the graph, or derive variants.
+
+%To our knowledge, these methods are the first to trivially scale to operate on large base-level graphs of this type.
+
+%To foster a high-level understanding of large graphs, we pursue a novel approach to visualization.
+%ODGI provides a linear-time 1D visualization that easily renders gigabase-scale pangenomes in a human-interpretable format suitable for understanding the topology and genome releationships in the pangenome graph.
+%A similar 2D visualization is more costly to compute, but similarly provides linear-time scaling, allowing us to apply it to large pangenomes.
+%This removes a major bottleneck in
+
+%Subsetting operations allow us to zoom in on particular genes or regions of interest for more precise examination.
+
+%We extract key information about the graph using its paths as a pangenome reference system.
+
+%We find that, to understanding relationships between sequences and positions in pangenome graphs, we must use context mapping to ``untangle'' complex regions found in VNTRs, segmental duplications, and centromeric repeats.
+%This approach lets use pangenome graphs as a foundation for universal liftover between any two sets of sequences in the graph.
+
+%It supports interrogation of the graph for basic dimensional properties, positional and path-related queries, and visualization in one-dimension (1D) and two-dimensions (2D).
+%ODGI also supports operations on the graph such as subset, subdivide, break, combine, normalize, or order its components, nodes, and paths.
+%Most of the tools are designed to be applied together, piping the output of one tool into the next, thereby preventing the creation of intermediate files, and reducing the number of IO operations.
+
+\subsection{Building the \textsc{ODGI} model}
+\label{sec:build}
+
+% erik
+
+% build
+Solves problem of working with big and complex graphs in limited memory.
+
+Figure: encoding model as a cartoon --  HTT exon 1 as tiny example graph (maybe)
+
+\subsection{Converting to standard text formats}
+\label{sec:text}
+
+% pjotr & andrea
+
+% view / paths
+Allows us to export basic data of the graph (genomes and graph structure).
+
+\textit{odgi view} can output GFA and internal graph representations \\
+\textit{odgi stats} outputs statistics in TSV or YAML. \\
+\textit{odgi bin} can output JSON. \\
+\textit{odgi degree/position/depth/untangle/tips} can output BED. \\
+\textit{odgi untangle} outputs PAF.
+
+Figure: .dot showing the relationship between all the tools and the data formats we use.
+
+\subsection{Extracting regions of interest}
+\label{sec:extract}
+
+% andrea (?)
+
+% extract
+Pangenome graphs are very large, but we often only want to work with a small portion (e.g. a single gene).
+We can extract such regions using coordinates on the paths in the graph to guide us.
+
+Figure showing extraction process---either schematic or ``real''.
+
+\subsection{Editing the graph structure}
+\label{sec:edit}
+
+% andrea (?)
+
+% chop / unchop / squeeze / break / prune / explode
+These are commonly-needed basic operations on the topology of the graph.
+
+\subsection{Visualizing pangenome graphs}
+\label{sec:viz}
+
+% andrea (viz) & erik (draw)
+
+% interesting that we don't go base-level at all... we use vg for that, or STM
+
+Pangenome graph visualization is one of the first steps to gain insight into the mutual relationship between the sequences and their variation.
+\textit{odgi viz} and \textit{odgi draw} provide scalable ways of generating human-viewable pictures of the high-level structure of the graph.
+
+\textit{odgi viz} offers a binned, linearized rendering in 1 dimension (1D).
+
+% Explain odgi viz's visualization
+
+However, complex, nonlinear graph structures are difficult to display and interpret in a convenient low number of dimensions.
+To overcome such a limitation, \textit{odgi viz} supports multiple visualization modes, making it easy to grasp the properties of the graph and its complexity.
+
+% Explain odgi viz's color by path position
+% Explain odgi viz's color by strandness
+% Explain odgi viz's color by coverage
+
+\textit{odgi draw} extends the visualization in 2 dimensions (2D), taking in input the layouts built by \textit{odgi layout}.
+
+% Explain odgi draw's visualization
+
+Figure here with 1D + 2D vizs.
+
+\subsection{Untangling and navigating the pangenome with path-based coordinates}
+\label{sec:untangle}
+
+% erik
+
+The key data in a pangenome graph is a representation of the alignment (or homology) relationships between sequences in the pangenome.
+Navigating and understanding the graph requires coordinate systems that we can use to link other data into the graph model, and thus to all genomes in the pangenome.
+Primarily, ODGI uses the embedded pangenome sequences to provide a set of coordinates in a standard, 1D format.e
 These coordinates depend on the sequences, and are graph-independent.
+Our for positional conversion (\textit{odgi position}), annotation liftover (\textit{odgi untangle}), subgraph extraction (\textit{odgi extract}), visualization (\textit{odgi viz}), path depth (\textit{odgi depth}), and graph complexity measurement (\textit{odgi degree}) can all operate on coordinates of paths in the graph.
 
-To provide a universal coordinate space for the graph, we project its nodes into a low-dimensional space (in practice, 1 or 2 dimensions) by learning the projection via a HOGWILD!, path-guided stochastic gradient descent (PG-SGD) model.
-\FIXME{Cite or explain?!}
+For understanding the large-scale organization of the graph, we benefit from an even simpler coordinate system.
+To provide a universal coordinate space for the graph, we project its nodes into a low-dimensional space (in practice, 1 or 2 dimensions) by learning the projection via a HOGWILD!
+\citep{niu2011hogwild}, path-guided stochastic gradient descent (PG-SGD) layout algorithm that adapts SGD-based drawing to pangenome graphs \citep{zheng2018graph}.
 This approach updates node relative positions to best-match their distance in the paths running through the graph.
@@
 
-%To our knowledge, these methods are the first to trivially scale to operate on large base-level graphs of this type.
-
-%To foster a high-level understanding of large graphs, we pursue a novel approach to visualization.
-%ODGI provides a linear-time 1D visualization that easily renders gigabase-scale pangenomes in a human-interpretable format suitable for understanding the topology and genome releationships in the pangenome graph.
-%A similar 2D visualization is more costly to compute, but similarly provides linear-time scaling, allowing us to apply it to large pangenomes.
-%This removes a major bottleneck in
-
-Subsetting operations allow us to zoom in on particular genes or regions of interest for more precise examination.
-
-We extract key information about the graph using its paths as a pangenome reference system.
-
-We find that, to understanding relationships between sequences and positions in pangenome graphs, we must use context mapping to ``untangle'' complex regions found in VNTRs, segmental duplications, and centromeric repeats.
-This approach lets use pangenome graphs as a foundation for universal liftover between any two sets of sequences in the graph.
-
-We begin with a set of HandleGraph-based algorithms first establish in the VG toolkit.
-These include algorithms for graph traversal, partition-finding, $k$-mer and character enumeration, sorting, and pruning.
-However, ODGI is more focused than VG.
-Unlike VG, we provide no methods to construct graphs, map reads to the graph, or derive variants.
-\FIXME{Above switches from ODGI to WE --- why not hold on to the former?}
-
-%It supports interrogation of the graph for basic dimensional properties, positional and path-related queries, and visualization in one-dimension (1D) and two-dimensions (2D).
-%ODGI also supports operations on the graph such as subset, subdivide, break, combine, normalize, or order its components, nodes, and paths.
-%Most of the tools are designed to be applied together, piping the output of one tool into the next, thereby preventing the creation of intermediate files, and reducing the number of IO operations.
-
-\input{fig_operations}
-
-\FIXME{Would it be an idea to put part of pggb in a box and explain steps? That way people can visualize using the toolkit and we can refer to it in the text}
-
-\subsection{Building the \textsc{ODGI} model}
-
-% erik
-
-% build
-Solves problem of working with big and complex graphs in limited memory.
-
-Figure: encoding model as a cartoon --  HTT exon 1 as tiny example graph (maybe)
-
-\subsection{Converting to standard text formats}
-
-% pjotr & andrea
-
-% view / paths
-Allows us to export basic data of the graph (genomes and graph structure).
-
-\texttt{odgi view} can output GFA and internal graph representations \\
-\texttt{odgi stats} outputs statistics in TSV or YAML. \\
-\texttt{odgi bin} can output JSON. \\
-\texttt{odgi degree/position/depth/untangle/tips} can output BED. \\
-\texttt{odgi untangle} outputs PAF.
-
-Figure: .dot showing the relationship between all the tools and the data formats we use.
-
-\subsection{Extracting regions of interest}
-
-% andrea (?)
-
-% extract
-Pangenome graphs are very large, but we often only want to work with a small portion (e.g. a single gene).
-We can extract such regions using coordinates on the paths in the graph to guide us.
-
-Figure showing extraction process---either schematic or ``real''.
-
-\subsection{Visualizing pangenome graphs}
-
-% andrea (viz) & erik (draw)
-
-% interesting that we don't go base-level at all... we use vg for that, or STM
-
-Pangenome graph visualization is one of the first steps to gain insight into the mutual relationship between the sequences and their variation.
-\texttt{odgi viz} and \texttt{odgi draw} provide scalable ways of generating human-viewable pictures of the high-level structure of the graph.
-
-\texttt{odgi viz} offers a binned, linearized rendering in 1 dimension (1D).
-
-% Explain odgi viz's visualization
-
-However, complex, nonlinear graph structures are difficult to display and interpret in a convenient low number of dimensions.
-To overcome such a limitation, \texttt{odgi viz} supports multiple visualization modes, making it easy to grasp the properties of the graph and its complexity.
-
-% Explain odgi viz's color by path position
-% Explain odgi viz's color by strandness
-% Explain odgi viz's color by coverage
-
-\texttt{odgi draw} extends the visualization in 2 dimensions (2D), taking in input the layouts built by \texttt{odgi layout}.
-
-% Explain odgi draw's visualization
-
-Figure here with 1D + 2D vizs.
-
-\subsection{Untangling and navigating the pangenome with path-based coordinates}
-
-% erik
-
 % position / tips
 % untangle (this follows on the jaccard graph mapping concepts in the previous section}
-The graph is a model of an alignment of many genomes.
-We can use coordinates in any genome to refer to it, but this requires a few basic operations ...
-Obtaining unambiguous mappings between different genomes requires the use of a new kind of graph based mapping (path graph jaccard).
+%The graph is a model of an alignment of many genomes.
+%We can use coordinates in any genome to refer to it, but this requires a few basic operations ...
+%Obtaining unambiguous mappings between different genomes requires the use of a new kind of graph based mapping (path graph jaccard).
 
@@
 
-\subsection{Finding latent structure in the pangenome graph topology}
+\subsection{Sorting the pangenome graph topology}
+\label{sec:sort}
 
@@
 
-Pangenome graphs built from raw sets of alignments may have complex structures which can introduce difficulty in downstream analyses.
+%Pangenome graphs built from raw sets of alignments may have complex structures which can introduce difficulty in downstream analyses.
 Uncovering their latent topology allows us to understand their typical sparse structures.
 
-\texttt{odgi sort} offers multiple sorting algorithms to find the best node order in one dimension.
+\textit{odgi sort} offers multiple sorting algorithms to find the best node order in one dimension.
 Most notably, nodes can be sorted topologically, randomly, by breaking cycles in the graph, by grooming, or by using a novel path-guided stochastic gradient descent (PG-SGD) algorithm.
@@
 Combining a series of sorts can mitigate the weaknesses of individually applied sorting algorithms.
-In Figure \textbf{\textit{TODO MHC wgg.85}}, a comparison of an unsorted human pangenome MHC graph and a sorted one are displayed.
+%In Figure \textbf{\textit{TODO MHC wgg.85}}, a comparison of an unsorted human pangenome MHC graph and a sorted one are displayed.
 
-For the establishment of a layout in two dimensions \texttt{odgi layout} applies a 2D adjusted PG-SGD algorithm.
-A calculated layout can be visualized statically with \texttt{odgi draw} (see Figure \textit{\textbf{TODO MHC wgg.85}}) or interactively explored with the in-development \texttt{gfaestus}\footnote{\url{https://github.com/chfi/gfaestus}}.
+%For the establishment of a layout in two dimensions \textit{odgi layout} applies a 2D adjusted PG-SGD algorithm.
+A calculated layout can be visualized statically with \textit{odgi draw} (see Figure \textit{\textbf{TODO MHC wgg.85}}) or interactively explored with the in-development \textit{gfaestus}\footnote{\url{https://github.com/chfi/gfaestus}}.
 
 \subsection{Obtaining metrics of the pangenome graph}
+\label{sec:metrics}
 
@@
 
-Applying \texttt{odgi stats}, users can retrieve metrics describing the graph properties, such as the number of nodes, edges, paths, and graph length.
+Applying \textit{odgi stats}, users can retrieve metrics describing the graph properties, such as the number of nodes, edges, paths, and graph length.
 MultiQC's \citep{Ewels_2016} ODGI module\footnote{\url{https://multiqc.info/docs/\#odgi} (accessed Oct 2021)} provides an interactive way to comparatively explore the statistics of an arbitrary number of graphs.
 ODGI also offers more advanced tools for the interrogation of the graphs.
-\texttt{odgi bin} summarizes the path information into bins of a specified size, enabling a summarized view of gigabase scale graphs.
+\textit{odgi bin} summarizes the path information into bins of a specified size, enabling a summarized view of gigabase scale graphs.
 This concept found its way into the pangenome graph ontology \cite{Yokoyama2020}.
-\texttt{odgi depth} returns the node depth as defined by query criteria, allowing users to assess the complexity of regions in the graph.
-Collapsed sequences with highly identical repeats indicate intricate \textit{loci}.
-Complex motifs can be also detected with \texttt{odgi degree}, which returns the node degree as defined by query criteria.
+\textit{odgi depth} returns the node depth as defined by query criteria, allowing users to assess the complexity of regions in the graph.
+Collapsed sequences with highly identical repeats indicate intricate loci.
+Complex motifs can be also detected with \textit{odgi degree}, which returns the node degree as defined by query criteria.
 High degree nodes can be the mirror of variation, misassemblies, or problems in the pangenome building, making the tool useful for validation.
@@
 % Figure: small graph with all the stats shown. A screenshot of MultiQC, visualize summary per bin somehow, R visualizations of degree and depth HTT exon 1 as tiny example graph
-
-\subsection{Editing the graph structure}
-
-% andrea (?)
-
-% chop / unchop / squeeze / break / prune / explode
-These are commonly-needed basic operations on the topology of the graph.
 

### 2021-10-19 manuscript/main.tex
ok now write the sections

@@
 \item \textit{odgi matrix} derives the pangenome matrix (\S\ref{sec:text}).
+\item \textit{odgi paths} lists and extracts paths in FASTA (\S\ref{sec:text}).
+\item \textit{odgi flatten} converts the graph to FASTA and BED (\S\ref{sec:text}).
 \item \textit{odgi extract} subsets the graph based on path ranges (\S\ref{sec:extract}).
@@
 \item \textit{odgi prune} removes complex regions (\S\ref{sec:edit}).
-\item \textit{odgi viz} provides a scalable 1D rendering of the graph (\S\ref{sec:viz}).
+\item \textit{odgi viz} provides a scalable linear visualization of the graph (\S\ref{sec:viz}).
+\item \textit{odgi draw} renders a 2D image of the graph (\S\ref{sec:viz}).
+\item \textit{odgi position} lift coordinates between paths and graph positions (\S\ref{sec:untangle}).
 \item \textit{odgi untangle} deconvolutes paths relative to a reference (\S\ref{sec:untangle}).
@@
 \item \textit{odgi layout} establishes a 2D layout using SGD (\S\ref{sec:sort}).
-\item \textit{odgi depth} describes path depth over the graph (\S\ref{sec:metrics}).
+\item \textit{odgi depth} describes path depth over graph and path positions (\S\ref{sec:metrics}).
+\item \textit{odgi degree} describes node degree over graph and path positions (\S\ref{sec:metrics}).
 \item \textit{odgi tips} finds path end points relative to a reference (\S\ref{sec:metrics}).
 \end{itemize}
+
+Each tool focuses on a small set of related operations.
+Most read or write the native ODGI format (.og) (Figure \ref{fig:operations}) and work with standard text based data formats common to bioinformatics.
+This supports the implementation of flexible, composable, graph processing pipelines based on a common data type.
+While the pangenome graph is a first-class entity in ODGI, we focus on the embedded paths to provide a universal coordinate system.
+By considering all paths in the graph as potential reference or query sequences, we make the graph invisible to downstream tools that operate on collections of sequences (e.g.
+bedtools, samtools, aligners).
+This approach benefits from the information in the graph without strongly embedding our methods in this difficult new research context.
+%ODGI provides methods that empower ``legacy'' toolsets.
+
+%we focus on common reference-based data formats like BED and PAF to input and output information from the graph.
 

### 2021-10-19 manuscript/main.tex
nada
build

@@
 We define the edges and path steps traversing the node in terms of this alphabet of $\delta$'s.
-The structures in algorithm \ref{alg:structs} describes our encoding.
+The structures in Algorithm \ref{alg:structs} describes our encoding.
 
@@
 % build
-Solves problem of working with big and complex graphs in limited memory.
+We begin by transforming the ``columnar'' storage model of GFA, in which nodes, edges, and paths are described independently, into the ODGI node-centric encoding with \textit{odgi build}.
+This construction step can present a significant bottleneck, in particular as the size of the path set of the graph increases.
+The ODGI data structure (Algorithm \ref{alg:structs}) allows algorithms to dynamically construct the graph and its path set to operate in parallel.
+In \textit{odgi build}, we initially construct the node vector in a serial operation that scans across the input GFA.
+Edges, and then paths, can be added in parallel by obtaining mutex $Node.lock$ for pairs of nodes and adding edges or steps to their corresponding vectors in both nodes.
+This parallelism speeds ODGI model construction by many-fold when testing against graphs built by the HPRC (TODO).
 
-Figure: encoding model as a cartoon --  HTT exon 1 as tiny example graph (maybe)
+%Figure: encoding model as a cartoon --  HTT exon 1 as tiny example graph (maybe)
 

### 2021-10-19 manuscript/main.tex
more

@@
 This construction step can present a significant bottleneck, in particular as the size of the path set of the graph increases.
-The ODGI data structure (Algorithm \ref{alg:structs}) allows algorithms to dynamically construct the graph and its path set to operate in parallel.
+The ODGI data structure (Algorithm \ref{alg:structs}) allows algorithms that build and modify the graph to operate in parallel, without any global locks.
 In \textit{odgi build}, we initially construct the node vector in a serial operation that scans across the input GFA.
-Edges, and then paths, can be added in parallel by obtaining mutex $Node.lock$ for pairs of nodes and adding edges or steps to their corresponding vectors in both nodes.
+Edges, and then paths, can be added in parallel by obtaining mutex $Node.lock$ for pairs of nodes and adding the edge or path step their $Node.edges$ and $Node.path\_steps$ vectors.
 This parallelism speeds ODGI model construction by many-fold when testing against graphs built by the HPRC (TODO).

### 2021-10-19 manuscript/main.tex
matthy pickyness
fix serial/parallel stuff; a bit of Italianity in the text

@@
 These models offer the opportunity to study the total genomic diversity of a population, including structurally complex regions.
-Analyzing hundreds of gigabase-scale genomes using this model is difficult and not well-supported by existing tools.
+However, analyzing hundreds of gigabase-scale genomes using pangenome graphs is difficult and not well-supported by existing tools.
 %In addition, vertebrate genomes present highly repetitive regions which increase the complexity of the analysis.
@@
 We anticipate that in the near future, studies will frequently begin from assemblies of similar quality, leading to demand for methods that allow us to create and explore pangenomes.
-Pangenome graphs are a natural model to support these kinds of analyses, but studies based on pangenome graphs are difficult due to a lack of downstream tools and standard operations.
+Pangenome graphs are a natural model to support these kinds of analyses, but studies based on pangenome graphs are difficult due to a lack of downstream tools and standard workflows.
 
@@
 Such motifs can occur in collapsed repeat structures generated by ambiguous sequence homology relationships in repeats found in the centromeres and other segmental duplications.
-If we cannot process such regions, we have only two options---we can remove these regions, or we can leave them unaligned.
-Neither of these solutions allows us to investigate their biological features.
+If we cannot process such regions, we have only two options: 1) we can remove these regions, or 2) we can leave them unaligned.
+However, neither of these solutions allows us to investigate their biological features.
 To seamlessy represent such difficult regions, we follow an approach implemented in the dynamic version of the Graph BWT (GBWT) \citep{Siren:2020}, and built a node-centric, dynamic, compressed model of the paths.
@@
     \textbf{lock} \tcp{atomic locking primitive}
-    \textbf{sequence} $= [$A$|$T$|$G$|$C$]*$ \tcp{DNA, RNA, protein}
+    \textbf{sequence} $= [$A$|$T$|$G$|$C$|$N$]*$ \tcp{DNA}
     \tcp{bit-packed vector of edges}
@@
     \tcp{bit-packed vector of path steps}
-    \textbf{path\_steps} $[Step_1 \ldots Step_\Omega]*$
+    \textbf{path\_steps} $[Step_1 \ldots Step_n]*$
 }
@@
 %The delta between two nodes is defined as their distance in the graph vector (i.e., the difference between the node offsets).
-For efficiency $Node_i.sequence$ is stored as a plain string, while the $edges$ and $path\_steps$ are stored using a dynamic succinct integer vector that requires $O(2nw)$ bits for the edges and $O(5nw)$ bits for the path steps, where $w$ is $\approx log_2(n)$ \citep{prezza2017framework}.
+For efficiency, $Node_i.sequence$ is stored as a plain string, while the $edges$ and $path\_steps$ are stored using a dynamic succinct integer vector that requires $O(2nw)$ bits for the edges and $O(5nw)$ bits for the path steps, where $n$ is the number of steps on the node, and $w$ is $\approx log_2(n)$ \citep{prezza2017framework}.
 
 To allow edit operations in parallel, each node structure includes a byte-width mutex $lock$.
-All edit operations on the graph must touch at most two $Node$ structs at a time.
-(Both our edge and path step representations are doubly-linked.)
+All edit operations on the graph must touch at most two $Node$ structs at a time (both edge and path step representations are doubly-linked).
 To avoid deadlocks, we acquire the node locks in ascending $Node.id$ order and release them in desceding order.
@@
 By avoiding global locks, we implement many of the operations in ODGI using maximum parallelism available.
-This ``atomic'' approach to edit operations on the graph is key to enabling our method to scale to the largest genome graphs that we can currently build.
+This ``atomic'' approach to edit operations on the graph is key to enable our methods to scale to the largest genome graphs that we can currently build.
 
@@
 Each tool focuses on a small set of related operations.
-Most read or write the native ODGI format (.og) (Figure \ref{fig:operations}) and work with standard text based data formats common to bioinformatics.
+Most read or write the native ODGI format (`og' extension) (Figure \ref{fig:operations}) and work with standard text based data formats common to bioinformatics.
 This supports the implementation of flexible, composable, graph processing pipelines based on a common data type.
@@
 % build
-We begin by transforming the ``columnar'' storage model of GFA, in which nodes, edges, and paths are described independently, into the ODGI node-centric encoding with \textit{odgi build}.
+We begin by transforming the storage model of the GFA format (in which nodes, edges, and paths are described independently) into the ODGI node-centric encoding with \textit{odgi build}.
 This construction step can present a significant bottleneck, in particular as the size of the path set of the graph increases.
@@
 In \textit{odgi build}, we initially construct the node vector in a serial operation that scans across the input GFA.
-Edges, and then paths, can be added in parallel by obtaining mutex $Node.lock$ for pairs of nodes and adding the edge or path step their $Node.edges$ and $Node.path\_steps$ vectors.
+Then, in serial, we add edges in the $Node.edges$ vectors of pairs of nodes.
+Finally, we create paths in serial, and extend them in parallel by obtaining mutex $Node.lock$ for pairs of nodes and by adding the path step in their $Node.path\_steps$ vectors.
 This parallelism speeds ODGI model construction by many-fold when testing against graphs built by the HPRC (TODO).

### 2021-10-19 manuscript/main.tex
explain viz/draw earlier

@@
 
-\title[short Title]{ODGI: understanding pangenome graphs}
+\title[ODGI: understanding pangenome graphs]{ODGI: understanding pangenome graphs}
 \author[Guarracino, Heumos \textit{et~al}.]{
@@
 
-Variation graphs are commonly represented by a subset of version 1 of the Graphical Fragment Assembly format (GFAv1): odgi build and odgi view allow to convert graphs from GFA to ODGI format and vice versa, respectively.
 In the variation graph model, paths have to respect the graph’s topology: this can be verified with odgi validate, to ensure no errors in the input or edited graphs.
-
-Applying odgi stats, users can retrieve metrics describing the graph properties, such as the number of nodes, edges, paths, and graph length.
-ODGI also offers more advanced tools for interrogating the graphs. odgi bin summarizes the path information into bins of specified size, enabling a summarized view of gigabase scale graphs.
-odgi depth returns the node depth as defined by query criteria, allowing users to detect the complex regions in the graph due to sequences with highly identical repeats.
-Complex motifs can be also detected with odgi degree, which returns the node degree as defined by query criteria.
-High degree nodes are the mirror of misassemblies or problems in the pangenome building, making the tool useful for debugging the process as well.
-
-ODGI allows edit operations on the graphs.
-Cycles in the graph complicate downstream analyses: odgi break removes the cycles, reducing the complexity of the graph topology.
-odgi groom removes spurious inverting links by exploring the graph from the orientation supported by most paths.
-To enable efficient sequence alignment against the graph, long nodes can be divided into shorter nodes at a maximum requested size using odgi chop.
-Partial order alignment, which consists of aligning sequences against a directed acyclic graph (DAG), is frequently used in pangenome building pipelines, but the current implementations return DAGs with 1-bp long nodes; odgi unchop allows joining nodes that can be merged without changing the graph topology, nor the embedded sequences, obtaining an equivalent, but more compact, representation of the graph.
-
-Pangenome graphs can embed multiple chromosomes as separated connected components (inter-chromosomal structural variants would join the components into bigger ones).
-odgi explode separates the connected components in different ODGI format files, while odgi squeeze allows merging multiple graphs into the same ODGI format file, preventing node ID collisions.
-odgi extract allows extracting specific subregions of the graph as defined by query criteria, thereby simplifying the downstream analyses and reducing the resources to work only with the extracted region.
-
 In variation graphs the coordinates are provided by the embedded path sequences.
@@
 \item \textit{odgi flatten} converts the graph to FASTA and BED (\S\ref{sec:text}).
+\item \textit{odgi viz} provides a scalable linear visualization of the graph (\S\ref{sec:viz}).
+\item \textit{odgi draw} renders a 2D image of the graph (\S\ref{sec:viz}).
 \item \textit{odgi extract} subsets the graph based on path ranges (\S\ref{sec:extract}).
@@
 \item \textit{odgi prune} removes complex regions (\S\ref{sec:edit}).
-\item \textit{odgi viz} provides a scalable linear visualization of the graph (\S\ref{sec:viz}).
-\item \textit{odgi draw} renders a 2D image of the graph (\S\ref{sec:viz}).
 \item \textit{odgi position} lift coordinates between paths and graph positions (\S\ref{sec:untangle}).
@@
 This approach benefits from the information in the graph without strongly embedding our methods in this difficult new research context.
-%ODGI provides methods that empower ``legacy'' toolsets.
-
-%we focus on common reference-based data formats like BED and PAF to input and output information from the graph.
-
-%To further our understanding of these graphs, we have developed a set of methods that let us inspect and manipulate them.
 
@@
 
-%To our knowledge, these methods are the first to trivially scale to operate on large base-level graphs of this type.
-
-%To foster a high-level understanding of large graphs, we pursue a novel approach to visualization.
-%ODGI provides a linear-time 1D visualization that easily renders gigabase-scale pangenomes in a human-interpretable format suitable for understanding the topology and genome releationships in the pangenome graph.
-%A similar 2D visualization is more costly to compute, but similarly provides linear-time scaling, allowing us to apply it to large pangenomes.
-%This removes a major bottleneck in
-
-%Subsetting operations allow us to zoom in on particular genes or regions of interest for more precise examination.
-
-%We extract key information about the graph using its paths as a pangenome reference system.
-
 %We find that, to understanding relationships between sequences and positions in pangenome graphs, we must use context mapping to ``untangle'' complex regions found in VNTRs, segmental duplications, and centromeric repeats.
@@
 
-%It supports interrogation of the graph for basic dimensional properties, positional and path-related queries, and visualization in one-dimension (1D) and two-dimensions (2D).
-%ODGI also supports operations on the graph such as subset, subdivide, break, combine, normalize, or order its components, nodes, and paths.
 %Most of the tools are designed to be applied together, piping the output of one tool into the next, thereby preventing the creation of intermediate files, and reducing the number of IO operations.
@@
 
-\subsection{Extracting regions of interest}
-\label{sec:extract}
-
-% andrea (?)
-
-% extract
-Pangenome graphs are very large, but we often only want to work with a small portion (e.g. a single gene).
-We can extract such regions using coordinates on the paths in the graph to guide us.
-
-Figure showing extraction process---either schematic or ``real''.
-
-\subsection{Editing the graph structure}
-\label{sec:edit}
-
-% andrea (?)
-
-% chop / unchop / squeeze / break / prune / explode
-These are commonly-needed basic operations on the topology of the graph.
-
 \subsection{Visualizing pangenome graphs}
@@
 % interesting that we don't go base-level at all... we use vg for that, or STM
+
+%To foster a high-level understanding of large graphs, we pursue a novel approach to visualization.
+%ODGI provides a linear-time 1D visualization that easily renders gigabase-scale pangenomes in a human-interpretable format suitable for understanding the topology and genome releationships in the pangenome graph.
+%A similar 2D visualization is more costly to compute, but similarly provides linear-time scaling, allowing us to apply it to large pangenomes.
 
@@
 Figure here with 1D + 2D vizs.
+
+\subsection{Extracting regions of interest}
+\label{sec:extract}
+
+% andrea (?)
+
+% extract
+Pangenome graphs are very large, but we often only want to work with a small portion (e.g. a single gene).
+We can extract such regions using coordinates on the paths in the graph to guide us.
+
+Figure showing extraction process---either schematic or ``real''.
+
+odgi extract allows extracting specific subregions of the graph as defined by query criteria, thereby simplifying the downstream analyses and reducing the resources to work only with the extracted region.
+
+%Subsetting operations allow us to zoom in on particular genes or regions of interest for more precise examination.
+
+%We extract key information about the graph using its paths as a pangenome reference system.
+
+\subsection{Editing the graph structure}
+\label{sec:edit}
+
+% andrea (?)
+
+% chop / unchop / squeeze / break / prune / explode
+
+To enable efficient sequence alignment against the graph, long nodes can be divided into shorter nodes at a maximum requested size using odgi chop.
+Partial order alignment, which consists of aligning sequences against a directed acyclic graph (DAG), is frequently used in pangenome building pipelines, but the current implementations return DAGs with 1-bp long nodes; odgi unchop allows joining nodes that can be merged without changing the graph topology, nor the embedded sequences, obtaining an equivalent, but more compact, representation of the graph.
+
+Pangenome graphs can embed multiple chromosomes as separated connected components (inter-chromosomal structural variants would join the components into bigger ones).
+odgi explode separates the connected components in different ODGI format files, while odgi squeeze allows merging multiple graphs into the same ODGI format file, preventing node ID collisions.
+
+Cycles in the graph complicate downstream analyses: odgi break removes the cycles, reducing the complexity of the graph topology.
+odgi groom removes spurious inverting links by exploring the graph from the orientation supported by most paths.
+
+These are commonly-needed basic operations on the topology of the graph.
 

### 2021-10-19 manuscript/main.tex
drafting odgi viz and its placeholder image

@@
 
-%ODGI is designed to build and modify paths in parallel, keeping its representation in memory efficient.
 The ODGI toolkit builds on existing approaches to efficiently store and manipulate variation graphs \citep{Garrison:2018}.
-%It implements the \textsc{HandleGraph} model \cite{Eizenga_2020_BX}.
 Similar to other efficient libaries presenting the \textsc{HandleGraph} model \citep{Eizenga_2020_BX}, ODGI's implementation rests on three key properties which hold for most pangenome variation graphs:
@@
 
-% erik
-
-% build
 We begin by transforming the storage model of the GFA format (in which nodes, edges, and paths are described independently) into the ODGI node-centric encoding with \textit{odgi build}.
@@
 
-% andrea (viz) & erik (draw)
-
 % interesting that we don't go base-level at all... we use vg for that, or STM
 
-%To foster a high-level understanding of large graphs, we pursue a novel approach to visualization.
-%ODGI provides a linear-time 1D visualization that easily renders gigabase-scale pangenomes in a human-interpretable format suitable for understanding the topology and genome releationships in the pangenome graph.
-%A similar 2D visualization is more costly to compute, but similarly provides linear-time scaling, allowing us to apply it to large pangenomes.
+Pangenome graph visualization is one of the first steps to gain insight into the mutual relationship between the sequences and their variation.
+We pursue a novel approach to visualization with \textit{odgi viz} and \textit{odgi draw}, two tools which provide scalable ways of generating pictures of the high-level structure of the graph.
 
-Pangenome graph visualization is one of the first steps to gain insight into the mutual relationship between the sequences and their variation.
-\textit{odgi viz} and \textit{odgi draw} provide scalable ways of generating human-viewable pictures of the high-level structure of the graph.
-
-\textit{odgi viz} offers a binned, linearized rendering in 1 dimension (1D).
-
-% Explain odgi viz's visualization
+\textit{odgi viz} supports a binned, linearized rendering in 1 dimension (1D).
+This visualization is computed in linear-time and offers a human-interpretable format suitable for understanding the topology and genome relationships in the pangenome graph (Fig.
+\ref{fig:odgi_viz}).
+% Explain here odgi viz
 
 However, complex, nonlinear graph structures are difficult to display and interpret in a convenient low number of dimensions.
-To overcome such a limitation, \textit{odgi viz} supports multiple visualization modes, making it easy to grasp the properties of the graph and its complexity.
+To overcome such a limitation, \textit{odgi viz} supports multiple visualization modes (Fig. \ref{fig:odgi_viz}\textbf{b-d}), making it easy to grasp the properties of the graph and its complexity.
+% Explain here 3 odgi viz modes
 
-% Explain odgi viz's color by path position
-% Explain odgi viz's color by strandness
-% Explain odgi viz's color by coverage
-
-\textit{odgi draw} extends the visualization in 2 dimensions (2D), taking in input the layouts built by \textit{odgi layout}.
-
-% Explain odgi draw's visualization
-
-Figure here with 1D + 2D vizs.
+\textit{odgi draw} extends the visualization in 2 dimensions (2D) by rendering the layout built by \textit{odgi layout} (\S\ref{sec:sort}).
+A 2D rendering is more costly to compute, but we similarly provide a linear-time scaling implementation, allowing us to apply it to large pangenome graphs.
 
@@
 
-% simon
 % HTT exon1
-% Only chr8 longs if I can't produce someting nice
 
 % stats / multiQC / bin / degree / depth
-
-% Solves the problem of understanding the overall structure, size, and features of the graph and its sequences.
 
@@
 \ref{fig:metrics}, the characteristics of a 90 haplotypes human pangenome graph of the exon 1 huntingtin gene \citep{Sathasivam2013,Neueder2017} (\textit{HTTexon1}), obtained with the ODGI metrics tools, are shown.
+
+\input{fig_odgi_viz}
 

### 2021-10-20 manuscript/fig_odgi_viz.tex
added odgi_viz placeholder

--- 2021-10-14 manuscript/fig_metrics.tex
@@
 	\end{subfigure}
-
 	\begin{subfigure}{\linewidth}
@@
 %	\includegraphics[width=\linewidth]{fig/metrics/chr4.pan.HTTex1.gfa.multiqc_odgi_stats.png}
-	\caption{Features of a 90 haplotypes human pangenome graph of the exon 1 huntingtin gene (\textit{HTTexon1}): \textbf{(a)} Excerpt of vital statistics of the \textit{HTTexon1} graph displayed by MultiQC's ODGI module.
-The very high GC content of 73.0\% compared to a human genomic mean GC content of 40.9\% \cite{Piovesan2019} is in accordance with the literature (see for example \cite{Sathasivam2013, Neueder2017}).
-\textbf{(b)} \texttt{odgi viz} visualization of the 13 largest gene alleles, CHM13, and GRCH38 of the \textit{HTTexon1} graph.
-\textbf{(c)} Per nucleotide node depth distribution of CHM13 in the \textit{HTTexon1} graph.
-The alternating depth around position 200 indicates polymorphic variation.
-\textbf{(d)} Per nucleotide node degree distribution of CHM13 in the \textit{HTTexon1} graph.
-The polymorphic region around position 200 complements the above node depth analysis.
-Figures \textbf{(b)-(d)} clearly highlight the variant region around position 200 showing the variable number of glutamine residues of the different individuals as reported by \cite{Nance1999}.}
-	\label{fig:metrics}
+	\caption{Visualizing the (\textit{C4 locus}) from a human pangenome graph of 90 haplotypes: \textbf{(a)} standard mode: explain the bio stuff.
+\textbf{(b)} odgi viz's color by path position: explain the bio stuff XXX.
+\textbf{(c)} odgi viz's color by strandness: explain the bio stuff XXX.
+`textbf{(d)}  odgi viz's color by coverage: explain the bio stuff XXX.}
+	\label{fig:odgi_viz}
 \end{figure*}

### 2021-10-20 manuscript/main.tex
Added vg in intro and closing discussion paragraph

@@
 Sven Nahnsen\,$^{\text{\sfb 2}}$,
-Pjotr Prins\,$^{\text{\sfb 3}}$,
-and Erik Garrison\,$^{\text{\sfb 3}*}$
+Pjotr~Prins\,$^{\text{\sfb 3}}$,
+and~Erik~Garrison\,$^{\text{\sfb 3}*}$
 }
@@
 Hence, fast and versatile software is required to ask intricate questions of such data in an efficient way. \\
-\textbf{Results:} Here we present ODGI, a new suite of tools that implements scalable algorithms and has an efficient in-memory representation of DNA variation graphs.
+\textbf{Results:} We wrote ODGI, a new suite of tools that implements scalable algorithms and has an efficient in-memory representation of DNA variation graphs.
 ODGI includes tools for detecting complex regions, extracting loci, removing artifacts, exploratory analysis, manipulation, validation, and visualization.
@@
 %The tools scale efficiently to large collections of eukaryotic genomes, allowing users to work easily with pangenome graphs of tens to thousands of large genomes while maintaining fast execution runtime and low memory overhead.
-\textbf{Availability:} ODGI is under the MIT license.
-Source code can be downloaded from \url{https://github.com/pangenome/odgi} while the documentation is available at \url{https://odgi.readthedocs.io}.
-ODGI can be installed via Bioconda \url{https://bioconda.github.io/recipes/odgi/README.html} or Guix \url{https://github.com/ekg/guix-genomics/blob/master/odgi.scm}. \\
+\textbf{Availability:} ODGI is published as free software under the MIT open source license.
+Source code can be downloaded from \url{https://github.com/pangenome/odgi} and documentation is available at \url{https://odgi.readthedocs.io}.
+ODGI can be installed via Bioconda \url{https://bioconda.github.io/recipes/odgi/README.html} or GNU Guix \url{https://github.com/ekg/guix-genomics/blob/master/odgi.scm}. \\
 \textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
@@
 Current methods to build these graphs are still in development \citep{Li:2020,Armstrong:2020,pggb}, but they have largely settled on a common data model, represented in the Graphical Fragment Assembly (GFA) format \citep{GFA}).
+Because ODGI make use of the GFA file standard its functionality can be mixed with other pangenome tools, such as the VG toolkit\citep{Garrison:2018}.
 %The improvement of these methods depends on tools that allow us to investigate
@@
 We require such a toolkit to understand the graphs made by these methods.
+\FIXME{last sentence not really clear}
 
@@
 With ODGI we implemented a state-of-the-art tool suite that can transform, analyse, simplify, validate, and visualise pangenome graphs at large scale.
+%
 Lifting over annotations and linearizing nested graph structures place the suite as the bridge between traditional linear reference genome analysis and pangenome graphs.
@@
 
-rdf sparql side, bring in annotation for great biological interpretation \cite{Yokoyama2020}, federated queries, sharing, FAIR
+ODGI provides common operation on on standard pangenomic file formats, such as GFA.
+ODGI aims to be a reference implementation for many pangenome related operations.
+In the near future we aim to improve RDF support for annotation in the in-memory graph\cite{Yokoyama2020}, for example, and allow for federated queries on the pangenome, and FAIR data sharing.
 
@@
 
-some cancer stuff?!
-
-\FIXME{I think we can leave RDF and FAIR out as it does not really relate to performance and ODGI explorations.
-Cancer, yes, we could state that the tools are applicable to cancer genome exploration as wild variation can be represented :)}
+With the increased adoption of long read sequencing we expect pangenome tools to become increasingly common in biomedical research.
+Particularly for targets that involve complex variation, such as cancer, plant genomics and metagenomics, ODGI should facilitate disentangling, describing and analysing a much larger set of variation than previously was possible with tools that depend on short reads and reference genomes.
 

### 2021-10-20 manuscript/main.tex
notes

@@
 Current methods to build these graphs are still in development \citep{Li:2020,Armstrong:2020,pggb}, but they have largely settled on a common data model, represented in the Graphical Fragment Assembly (GFA) format \citep{GFA}).
-Because ODGI make use of the GFA file standard its functionality can be mixed with other pangenome tools, such as the VG toolkit\citep{Garrison:2018}.
+Because ODGI make use of the GFA file standard its functionality can be mixed with other pangenome tools, such as the VG toolkit~\citep{Garrison:2018}.
+\FIXME{may want to move GFA standard+VG to intro section 1}
 %The improvement of these methods depends on tools that allow us to investigate
@@
 The ODGI toolkit builds on existing approaches to efficiently store and manipulate variation graphs \citep{Garrison:2018}.
-Similar to other efficient libaries presenting the \textsc{HandleGraph} model \citep{Eizenga_2020_BX}, ODGI's implementation rests on three key properties which hold for most pangenome variation graphs:
+Similar to other efficient libaries presenting the \red{so-called} \textsc{HandleGraph} model \citep{Eizenga_2020_BX}, ODGI's implementation rests on three key properties which hold for most pangenome variation graphs:
 

### 2021-10-20 manuscript/main.tex
add when the figure is ready
shorter

--- 2021-10-20 manuscript/main.tex
@@
 The Human Pangenome Reference Consortium (HPRC) and Telomere-to-Telomere (T2T) consortium \citep{Miga:2020, Logsdon_2021, Nurk_2021} have recently demonstrated that high-quality \textit{de novo} assemblies can be routinely generated from third-generation long read sequencing data.
-We anticipate that in the near future, studies will frequently begin from assemblies of similar quality, leading to demand for methods that allow us to create and explore pangenomes.
+We anticipate that assemblies of similar quality will become common, leading to demand for methods that allow us to create and explore pangenomes.
 Pangenome graphs are a natural model to support these kinds of analyses, but studies based on pangenome graphs are difficult due to a lack of downstream tools and standard workflows.
@@
 
-Such a graph can be constructed by multiple sequence alignment \citep{Lee_2002,Grasso_2004}, but the resulting partially-ordered graph cannot represent complex variation like inversions and cycles that occur at structural scales in genomes.
-General approaches to building pangenome graphs transitively reduce an alignment between sequences to an equivalent, labeled sequence graph \citep{Kehr_2014,Garrison_2019_thesis}.
+Such a graph can be constructed by multiple sequence alignment \citep{Lee_2002,Grasso_2004}, or by transitively reducing an alignment between sequences to an equivalent, labeled sequence graph \citep{Kehr_2014,Garrison_2019_thesis}.
 Current methods to build these graphs are still in development \citep{Li:2020,Armstrong:2020,pggb}, but they have largely settled on a common data model, represented in the Graphical Fragment Assembly (GFA) format \citep{GFA}).
-Because ODGI make use of the GFA file standard its functionality can be mixed with other pangenome tools, such as the VG toolkit\citep{Garrison:2018}.
+%\FIXME{may want to move GFA standard+VG to intro section 1}
 %The improvement of these methods depends on tools that allow us to investigate
 This standardization supports the development of a reference set of tools that operate on the pangenome graph model.
+Such an effort began with the VG toolkit toolkit~\citep{Garrison:2018}.
+Here we refocus it with with ODGI---a compatible, but independent set of algorithms focused on first-order interrogation and transformation of pangenome graphs.
+
+%Because ODGI uses this standard to  can be mixed with other pangenome tools, such as the VG toolkit
+%Here, we enhance ongoing efforts with the VG 
 %The need to understand the graphs made by these methods motivates it.
-We require such a toolkit to understand the graphs made by these methods.
-\FIXME{last sentence not really clear}
+%We require such a toolkit to understand the graphs made by these methods.
+%\FIXME{last sentence not really clear}
 
@@
 
-\input{fig_odgi_viz}
+%\input{fig_odgi_viz}
 

### 2021-10-21 manuscript/main.tex
File format section

@@
 
-\subsection{Converting to standard text formats}
+\subsection{Converting to standard file formats}
 \label{sec:text}
@@
 % pjotr & andrea
+ODGI maintains its own efficient binary format for storing graphs on disk.
+ODGI also supports reading and writing the textual GFA file format for interchange with other tools such as with the VG toolkit toolkit~\citep{Garrison:2018}.
 
-% view / paths
-Allows us to export basic data of the graph (genomes and graph structure).
+\textit{odgi view} can convert a graph in ODGI binary format to the standard GFAv1 format. It can reveal a graph’s internal structures for e.g. debugging processes.
 
-\textit{odgi view} can output GFA and internal graph representations \\
-\textit{odgi stats} outputs statistics in TSV or YAML. \\
-\textit{odgi bin} can output JSON. \\
-\textit{odgi degree/position/depth/untangle/tips} can output BED. \\
-\textit{odgi untangle} outputs PAF.
+\textit{odgi stats} outputs pangenome statistics in TSV or YAML textual file formats.
+Among other metrics, it can calculate the number of nodes, edges, paths and the total nucleotide length of the graph.
 
-Figure: .dot showing the relationship between all the tools and the data formats we use.
+\textit{odgi bin} generates a summarized view of gigabase scale graphs providing tab delimited or JSON `bins'. Also it is possible to output a FASTA file.
+
+\textit{odgi degree/position/depth/untangle/tips} all can output the BED format~\citep{Quinlan_2010}.
+
+\textit{odgi untangle} outputs the Pairwise mApping Format (PAF). PAF is a text format describing the approximate mapping positions between two set of sequences.
+The odgi untangle command projects paths into a reference-relative BEDPE file, decomposing paralogy relationships.
+During this process, it is capable of untangling loopy region resulting in linearized pairs of regions in the BEDPE file. A self dotplot assists in debugging and understanding the untangle process.
 

### 2021-10-21 manuscript/fig_odgi_viz.tex
C4 locus scripting

@@
         % include first image
-		\includegraphics[width=1.0\linewidth]{fig/metrics/chr4_pan_HTTex1_gfa_multiqc_odgi_stats.png}  
-		\label{fig:metrics-multiqc}
+        \includegraphics[width=1.0\linewidth, trim=0 +2cm 0 0]{fig/visualization_1D/chr6.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.grch38_chr6_31889244-32095493.0.sorted}
+        \label{fig:odgi_viz_default}
     \end{subfigure}
@@
         % include second image
-		\includegraphics[width=1.0\linewidth, trim=0 0 -1cm 0 ]{fig/metrics/chr4_pan_fa_a2fb268_e820cd3_9ea71d8_smooth_gfa_og_HTTex1_og_O_og_tiny_og_png_svg.pdf}  
-		\label{fig:metrics-viz}
+        \includegraphics[width=1.0\linewidth, trim=0 +2cm 0 0]{fig/visualization_1D/chr6.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.grch38_chr6_31889244-32095493.0.sorted.du}
+        \label{fig:odgi_viz_color_by_path_pos}
     \end{subfigure}
@@
         % include second image
-		\includegraphics[width=\linewidth,trim=+.225cm 0 +0.425cm +2cm]{fig/metrics/chr4_HTT_chm13_depth_w1_bed.pdf}  
-		\label{fig:metrics-depth}
+        \includegraphics[width=\linewidth, trim=0 +2cm 0 0]{fig/visualization_1D/chr6.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.grch38_chr6_31889244-32095493.0.sorted.z}
+        \label{fig:odgi_viz_color_by_inversion_rate}
     \end{subfigure}
@@
         % include fourth image
-		\includegraphics[width=\linewidth,trim=+.225cm 0 +.425cm +2cm]{fig/metrics/chr4_HTT_chm13_degree_w1_bed.pdf}  
-		\label{fig:metrics-degree}
+        \includegraphics[width=\linewidth, trim=0 0 0 0]{fig/visualization_1D/chr6.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.grch38_chr6_31889244-32095493.0.sorted.m}
+        \label{fig:odgi_viz_color_by_path_depth}
     \end{subfigure}
-%	\includegraphics[width=\linewidth]{fig/metrics/chr4.pan.HTTex1.gfa.multiqc_odgi_stats.png}
-	\caption{Visualizing the (\textit{C4 locus}) from a human pangenome graph of 90 haplotypes: \textbf{(a)} standard mode: explain the bio stuff.
+    \caption{Visualizing the complement component 4 (C4) \textit{locus} from a human pangenome graph of 90 haplotypes: \textbf{(a)} standard mode: explain the bio stuff.
 \textbf{(b)} odgi viz's color by path position: explain the bio stuff XXX.

### 2021-10-21 manuscript/main.tex
C4 locus scripting

@@
 ODGI maintains its own efficient binary format for storing graphs on disk.
-ODGI also supports reading and writing the textual GFA file format for interchange with other tools such as with the VG toolkit toolkit~\citep{Garrison:2018}.
+ODGI also supports reading and writing the textual GFA file format for interchange with other tools such as with the VG toolkit~\citep{Garrison:2018}.
 
@@
 
-However, complex, nonlinear graph structures are difficult to display and interpret in a convenient low number of dimensions.
+Nevertheless, complex, nonlinear graph structures are difficult to display and interpret in a convenient low number of dimensions.
 To overcome such a limitation, \textit{odgi viz} supports multiple visualization modes (Fig. \ref{fig:odgi_viz}\textbf{b-d}), making it easy to grasp the properties of the graph and its complexity.

### 2021-10-21 scripts/visualization.sh

@@ -1,0 +1,40 @@
+#!/bin/bash
+
+#======================================================================================================================
+# odgi viz with C4A/C4B locus graph
+#==================================
+
+filename_chr6_gfa=chr6.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa
+
+# Download and build the graph
+wget https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/scratch/2021_07_30_pggb/chroms/${filename_chr6_gfa}.gz
+gunzip ${filename_chr6_gfa}.gz
+odgi build -g ${filename_chr6_gfa} -o ${filename_chr6_gfa}.og -t 16 -P
+
+# TO_IGNORE
+## Find C4 coordinates
+#wget https://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/genes/hg38.ncbiRefSeq.gtf.gz
+#zgrep 'gene_id "C4A"\|gene_id "C4B"' hg38.ncbiRefSeq.gtf.gz | awk '$1 == "chr6"' | cut -f 1,4,5 | bedtools sort | bedtools merge -d 15000 | sed 's/chr6/chm13#chr6/g' > C4.coordinates.bed
+#odgi extract -i ${filename_chr6_gfa}.og -b C4.coordinates.bed -o - --full-range -L 130000 -t 16 -P | \
+#odgi sort -i - -o - --optimize -t 16 -P | \
+#odgi sort -i - -o ${filename_chr6_gfa}.og.C4.sorted.og -p Ygs -x 100 -t 16 -P
+#odgi viz -i ${filename_chr6_gfa}.og.C4.sorted.og -o ${filename_chr6_gfa}.og.C4.sorted.png -w 150  -m -B Spectral:4
+# TO_IGNORE
+
+# Extraction, optimization, and sorting
+odgi extract -i ${filename_chr6_gfa}.og -r grch38#chr6:31889244-32095493 -o - --full-range -t 16 -P | odgi explode -i - --biggest 1 --optimize -p ${filename_chr6_gfa}.og.grch38_chr6_31889244-32095493
+odgi sort -i ${filename_chr6_gfa}.og.grch38_chr6_31889244-32095493.0.og -o ${filename_chr6_gfa}.og.grch38_chr6_31889244-32095493.0.sorted.og -p Ygs -x 100 -t 16 -P
+
+# Default (binned) mode
[…]

### 2021-10-22 manuscript/main.tex
Add NIH funding support

@@
 
-S.H. acknowledges funding from the Central Innovation Programme (ZIM) for SMEs of the Federal Ministry for Economic Affairs and Energy of Germany.
+We gratefully acknowledge support from NIH/NIDA U01DA047638 (EG) and NIH/NIGMS R01GM123489 (EG and PP).
+SH acknowledges funding from the Central Innovation Programme (ZIM) for SMEs of the Federal Ministry for Economic Affairs and Energy of Germany.
 This work was supported by the BMBF-funded de.NBI Cloud within the German Network for Bioinformatics Infrastructure (de.NBI) (031A537B, 031A533A, 031A538A, 031A533B, 031A535A, 031A537C, 031A534A, 031A532B).

### 2021-10-22 manuscript/main.tex
Added missing output 4.2 commands listed in intro of 4.

@@
 
+\textit{odgi matrix} derives the pangenome matrix and writes the graph topology in sparse matrix formats.
+
+\textit{odgi paths} allows the investigation of paths of a given variation graph. It can calculate overlap statistics of groupings of paths and extracts FASTA.
+
+\textit{odgi flatten} projects the graph sequence and paths into FASTA and BED.
+
 \textit{odgi bin} generates a summarized view of gigabase scale graphs providing tab delimited or JSON `bins'. Also it is possible to output a FASTA file.
 
-\textit{odgi degree/position/depth/untangle/tips} all can output the BED format~\citep{Quinlan_2010}.
+\textit{odgi degree/position/depth/untangle/tips} all can output the BED format.
 
@@
 ODGI aims to be a reference implementation for many pangenome related operations.
-In the near future we aim to improve RDF support for annotation in the in-memory graph\cite{Yokoyama2020}, for example, and allow for federated queries on the pangenome, and FAIR data sharing.
+In the near future we aim to improve RDF support for annotation in the in-memory graph~\citep{Yokoyama2020}, for example, and allow for federated queries on the pangenome, and FAIR data sharing.
 

### 2021-10-22 manuscript/fig_metrics.tex
compress images putting subcaptions on the left

@@
 		% include first image
-		\includegraphics[width=1.0\linewidth]{fig/metrics/chr4_pan_HTTex1_gfa_multiqc_odgi_stats.png}  
+		\includegraphics[width=1.0\linewidth, trim=0 1cm 0 0]{fig/metrics/chr4_pan_HTTex1_gfa_multiqc_odgi_stats.png}
 		\label{fig:metrics-multiqc}
@@
 		% include second image
-		\includegraphics[width=1.0\linewidth, trim=0 0 -1cm 0 ]{fig/metrics/chr4_pan_fa_a2fb268_e820cd3_9ea71d8_smooth_gfa_og_HTTex1_og_O_og_tiny_og_png_svg.pdf}  
+		\includegraphics[width=1.0\linewidth, trim=0 0.7cm -1cm 0 ]{fig/metrics/chr4_pan_fa_a2fb268_e820cd3_9ea71d8_smooth_gfa_og_HTTex1_og_O_og_tiny_og_png_svg.pdf}
 		\label{fig:metrics-viz}
@@
 		% include second image
-		\includegraphics[width=\linewidth,trim=+.225cm 0 +0.425cm +2cm]{fig/metrics/chr4_HTT_chm13_depth_w1_bed.pdf}  
+		\includegraphics[width=\linewidth,trim=+.225cm 1.4cm +0.425cm +2cm]{fig/metrics/chr4_HTT_chm13_depth_w1_bed.pdf}
 		\label{fig:metrics-depth}
@@
 		% include fourth image
-		\includegraphics[width=\linewidth,trim=+.225cm 0 +.425cm +2cm]{fig/metrics/chr4_HTT_chm13_degree_w1_bed.pdf}  
+		\includegraphics[width=\linewidth,trim=+.225cm 1.5cm +.425cm +2cm]{fig/metrics/chr4_HTT_chm13_degree_w1_bed.pdf}
 		\label{fig:metrics-degree}

### 2021-10-22 manuscript/fig_odgi_viz.tex
drafted odgi viz part

@@
         % include first image
-        \includegraphics[width=1.0\linewidth, trim=0 +2cm 0 0]{fig/visualization_1D/chr6.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.grch38_chr6_31889244-32095493.0.sorted}
+        \includegraphics[width=1.0\linewidth, trim=0 +2cm 0 0]{fig/visualization_1D/chr6.pan.fa.c3d3224.7748b33.395c7f4.smooth.gfa.C4.sorted}
         \label{fig:odgi_viz_default}
@@
         % include second image
-        \includegraphics[width=1.0\linewidth, trim=0 +2cm 0 0]{fig/visualization_1D/chr6.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.grch38_chr6_31889244-32095493.0.sorted.du}
+        \includegraphics[width=1.0\linewidth, trim=0 +2cm 0 0]{fig/visualization_1D/chr6.pan.fa.c3d3224.7748b33.395c7f4.smooth.gfa.C4.sorted.du}
         \label{fig:odgi_viz_color_by_path_pos}
@@
         % include second image
-        \includegraphics[width=\linewidth, trim=0 +2cm 0 0]{fig/visualization_1D/chr6.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.grch38_chr6_31889244-32095493.0.sorted.z}
+        \includegraphics[width=\linewidth, trim=0 +2cm 0 0]{fig/visualization_1D/chr6.pan.fa.c3d3224.7748b33.395c7f4.smooth.gfa.C4.sorted.z}
         \label{fig:odgi_viz_color_by_inversion_rate}
@@
         % include fourth image
-        \includegraphics[width=\linewidth, trim=0 0 0 0]{fig/visualization_1D/chr6.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.grch38_chr6_31889244-32095493.0.sorted.m}
+        \includegraphics[width=\linewidth, trim=0 1.5cm 0 0]{fig/visualization_1D/chr6.pan.fa.c3d3224.7748b33.395c7f4.smooth.gfa.C4.sorted.m}
         \label{fig:odgi_viz_color_by_path_depth}
     \end{subfigure}
-    \caption{Visualizing the complement component 4 (C4) \textit{locus} from a human pangenome graph of 90 haplotypes: \textbf{(a)} standard mode: explain the bio stuff.
-\textbf{(b)} odgi viz's color by path position: explain the bio stuff XXX.
-\textbf{(c)} odgi viz's color by strandness: explain the bio stuff XXX.
-`textbf{(d)}  odgi viz's color by coverage: explain the bio stuff XXX.}
+    \caption{
+        Visualizing the complement component 4 (C4) pangenome graph extracted from a whole human pangenome graph of 90 haplotypes.
+        In all visualizations, 10 paths are displayed: 2 reference genomes (chm13 and grch38 on the top) and 8 haplotypes from 4 individuals.
+        \textbf{(a)} default modality: the image shows a quite linear graph.
+        The 2 longer links at the bottom indicate the presence of a structural variant (long link) with another structural variant nested inside it (short link).
+        Indeed, human C4 exists as 2 functionally distinct genes, C4A and C4B, which both vary in structure and copy number \citep{Sekar_2016}.
+The longer link indicates that the copy number status varies across the haplotypes represented in the pangenome.
+        Moreover, C4A and C4B genes segregate in both long and short genomic forms, distinguished by the presence or absence of a human endogenous retroviral (HERV) sequence, as also highlighted by the short nested link on the bottom.
+        \textbf{(b)} color by path position: the color gradients are smooths, highlighting that its node are well sorted in 1 dimension.
+        The top two reference genomes and 2 haplotypes (with names starting with HG01978\#1 and HG03492\#2) go from left to right, while 5 haplotypes go in the opposite direction, as indicated by the black color on their left.
+        \textbf{(c)} color by strandness: the red paths indicate the haplotypes that were assembled in reverse with respect to the 2 reference genomes.
+        \textbf{(d)} color by path depth: using the Spectra color palette with 4 level of path depths, white indicates no depth, while grey, red, and yellow indicate depth 1, 2, and 3, respectively.
+        Coloring by path depth, we can see that the two references present two different allele copies of the C4 genes, both of them including the HERV sequence.
+The entirely grey paths have one copy of these genes.
+        The path with name starting with HG01071\#2 presents 3 copies of the genes (indicated by the orange color), of which only 1 with the HERV sequence (gray color in the middle of the orange region).
+    }
     \label{fig:odgi_viz}

### 2021-10-22 manuscript/main.tex
compress images putting subcaptions on the left
drafted odgi viz part

@@
 \usepackage{comment}
-\usepackage{subcaption}
+
+% singlelinecheck=false puts subcaptions on the left
+\usepackage[singlelinecheck=false]{subcaption}
+
 \usepackage{algorithm2e}
@@
 
-\textit{odgi viz} supports a binned, linearized rendering in 1 dimension (1D).
-This visualization is computed in linear-time and offers a human-interpretable format suitable for understanding the topology and genome relationships in the pangenome graph (Fig.
-\ref{fig:odgi_viz}).
-% Explain here odgi viz
+\textit{odgi viz} supports a binned, linearized rendering in 1 dimension (that is, all graph nodes lie on the same axis).
+This visualization is computed in linear-time and offers a human-interpretable format suitable for understanding the topology and genome relationships in the pangenome graph (Fig. \ref{fig:odgi_viz}).
+Graph nodes are arranged on a single axis, from left to right, with the colored bar indicating the paths and the nodes they cross.
+White spaces indicate where paths do not traverse the nodes.
+The meaning of the colors depends on how \textit{odgi viz} is executed.
+By default, path colors has no particular meaning, but each path has its own color (Fig. \ref{fig:odgi_viz}\textbf{a}).
+Path names are displayed on the left of the paths.
+The black lines on the bottom indicate the edges connecting the nodes and, therefore, represent the graph topology.
 
-Nevertheless, complex, nonlinear graph structures are difficult to display and interpret in a convenient low number of dimensions.
-To overcome such a limitation, \textit{odgi viz} supports multiple visualization modes (Fig. \ref{fig:odgi_viz}\textbf{b-d}), making it easy to grasp the properties of the graph and its complexity.
-% Explain here 3 odgi viz modes
+Nevertheless, complex, nonlinear graph structures are difficult to display and interpret in a low number of dimensions.
+To overcome such a limitation, \textit{odgi viz} supports multiple visualization modalities (Fig.
+\ref{fig:odgi_viz}\textbf{b-d}), making it easy to grasp the properties of the graph and its complexity.
+Graph nodes order can affects downstream analyses on pangenome graphs.
+With \textit{odgi viz} we can color the paths by path position (Fig.
+\ref{fig:odgi_viz}\textbf{b}), with light grey indicating where paths begin and dark grey where they ends.
+This visualization is suitable for checking graph nodes' linear order, as smooth color gradients indicate that the graph reflects well the linear paths' coordinate system.
+Pangenome graphs represents both strand of DNA sequences.
+\textit{odgi viz} supports also coloring the paths by orientation, with paths colored in red and black where their sequence is reversed or in forward, respectively, with respect to the sequences represented by the graph nodes (Fig.
+\ref{fig:odgi_viz}\textbf{b}).
+Eukaryotic genomes experience gains and losses of genetic material, resulting in copy number variation (CNV) across the population.
+With \textit{odgi viz}, we can use multiple color palettes to color the paths by path depth, highlighting the different copy number statuses in the genomes represented in the pangenome graph (Fig.
+\ref{fig:odgi_viz}\textbf{d}).
 
@@
 
-%\input{fig_odgi_viz}
+\input{fig_odgi_viz}
 

### 2021-10-22 scripts/visualization.sh

@@ -4,37 +4,38 @@
 # odgi viz with C4A/C4B locus graph
 #==================================
 
-filename_chr6_gfa=chr6.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa
+filename_chr6_gfa=chr6.pan.fa.c3d3224.7748b33.395c7f4.smooth.gfa
 
 # Download and build the graph
-wget https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/scratch/2021_07_30_pggb/chroms/${filename_chr6_gfa}.gz
+wget https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/scratch/2021_06_27_pggb/chroms/${filename_chr6_gfa}.gz
 gunzip ${filename_chr6_gfa}.gz
 odgi build -g ${filename_chr6_gfa} -o ${filename_chr6_gfa}.og -t 16 -P
 
-# TO_IGNORE
-## Find C4 coordinates
-#wget https://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/genes/hg38.ncbiRefSeq.gtf.gz
-#zgrep 'gene_id "C4A"\|gene_id "C4B"' hg38.ncbiRefSeq.gtf.gz | awk '$1 == "chr6"' | cut -f 1,4,5 | bedtools sort | bedtools merge -d 15000 | sed 's/chr6/chm13#chr6/g' > C4.coordinates.bed
-#odgi extract -i ${filename_chr6_gfa}.og -b C4.coordinates.bed -o - --full-range -L 130000 -t 16 -P | \
-#odgi sort -i - -o - --optimize -t 16 -P | \
-#odgi sort -i - -o ${filename_chr6_gfa}.og.C4.sorted.og -p Ygs -x 100 -t 16 -P
-#odgi viz -i ${filename_chr6_gfa}.og.C4.sorted.og -o ${filename_chr6_gfa}.og.C4.sorted.png -w 150  -m -B Spectral:4
-# TO_IGNORE
+# Find C4 coordinates
+wget http://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/hg38.chrom.sizes
+wget https://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/genes/hg38.ncbiRefSeq.gtf.gz
[…]

### 2021-10-22 manuscript/fig_metrics.tex
compressss

--- 2021-10-14 manuscript/fig_metrics.tex
@@
 		% include first image
-		\includegraphics[width=1.0\linewidth]{fig/metrics/chr4_pan_HTTex1_gfa_multiqc_odgi_stats.png}  
+		\includegraphics[width=1.0\linewidth, trim=0 1cm 0 0.5cm]{fig/metrics/chr4_pan_HTTex1_gfa_multiqc_odgi_stats}
 		\label{fig:metrics-multiqc}
@@
 		% include second image
-		\includegraphics[width=1.0\linewidth, trim=0 0 -1cm 0 ]{fig/metrics/chr4_pan_fa_a2fb268_e820cd3_9ea71d8_smooth_gfa_og_HTTex1_og_O_og_tiny_og_png_svg.pdf}  
+		\includegraphics[width=1.0\linewidth, trim=0 0.7cm -1cm 0.2cm]{fig/metrics/chr4_pan_fa_a2fb268_e820cd3_9ea71d8_smooth_gfa_og_HTTex1_og_O_og_tiny_og_png_svg}
 		\label{fig:metrics-viz}
@@
 		% include second image
-		\includegraphics[width=\linewidth,trim=+.225cm 0 +0.425cm +2cm]{fig/metrics/chr4_HTT_chm13_depth_w1_bed.pdf}  
+		\includegraphics[width=\linewidth,trim=+.225cm 1.4cm +0.425cm +2.2cm]{fig/metrics/chr4_HTT_chm13_depth_w1_bed}
 		\label{fig:metrics-depth}
@@
 		% include fourth image
-		\includegraphics[width=\linewidth,trim=+.225cm 0 +.425cm +2cm]{fig/metrics/chr4_HTT_chm13_degree_w1_bed.pdf}  
+		\includegraphics[width=\linewidth,trim=+.225cm 1.5cm +.425cm +2.2cm]{fig/metrics/chr4_HTT_chm13_degree_w1_bed}
 		\label{fig:metrics-degree}

### 2021-10-22 manuscript/fig_odgi_viz.tex
compressss

@@
         % include first image
-        \includegraphics[width=1.0\linewidth, trim=0 +2cm 0 0]{fig/visualization_1D/chr6.pan.fa.c3d3224.7748b33.395c7f4.smooth.gfa.C4.sorted}
+        \includegraphics[width=1.0\linewidth, trim=0 +2cm 0 0.5cm]{fig/visualization_1D/chr6.pan.fa.c3d3224.7748b33.395c7f4.smooth.gfa.C4.sorted}
         \label{fig:odgi_viz_default}
@@
         % include second image
-        \includegraphics[width=1.0\linewidth, trim=0 +2cm 0 0]{fig/visualization_1D/chr6.pan.fa.c3d3224.7748b33.395c7f4.smooth.gfa.C4.sorted.du}
+        \includegraphics[width=1.0\linewidth, trim=0 +2cm 0 0.5cm]{fig/visualization_1D/chr6.pan.fa.c3d3224.7748b33.395c7f4.smooth.gfa.C4.sorted.du}
         \label{fig:odgi_viz_color_by_path_pos}
@@
         % include second image
-        \includegraphics[width=\linewidth, trim=0 +2cm 0 0]{fig/visualization_1D/chr6.pan.fa.c3d3224.7748b33.395c7f4.smooth.gfa.C4.sorted.z}
+        \includegraphics[width=\linewidth, trim=0 +2cm 0 0.5cm]{fig/visualization_1D/chr6.pan.fa.c3d3224.7748b33.395c7f4.smooth.gfa.C4.sorted.z}
         \label{fig:odgi_viz_color_by_inversion_rate}
@@
         % include fourth image
-        \includegraphics[width=\linewidth, trim=0 1.5cm 0 0]{fig/visualization_1D/chr6.pan.fa.c3d3224.7748b33.395c7f4.smooth.gfa.C4.sorted.m}
+        \includegraphics[width=\linewidth, trim=0 1.5cm 0 0.5cm]{fig/visualization_1D/chr6.pan.fa.c3d3224.7748b33.395c7f4.smooth.gfa.C4.sorted.m}
         \label{fig:odgi_viz_color_by_path_depth}

### 2021-10-22 manuscript/fig_odgi_viz.tex
smaller image, new graph

--- 2021-10-22 manuscript/fig_odgi_viz.tex
@@
         % include first image
-        \includegraphics[width=1.0\linewidth, trim=0 +2cm 0 0]{fig/visualization_1D/chr6.pan.fa.c3d3224.7748b33.395c7f4.smooth.gfa.C4.sorted}
+        \includegraphics[width=1.0\linewidth, trim=0 +2cm 0 0.5cm]{fig/visualization_1D/chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.C4.sorted}
         \label{fig:odgi_viz_default}
@@
         % include second image
-        \includegraphics[width=1.0\linewidth, trim=0 +2cm 0 0]{fig/visualization_1D/chr6.pan.fa.c3d3224.7748b33.395c7f4.smooth.gfa.C4.sorted.du}
+        \includegraphics[width=1.0\linewidth, trim=0 +2cm 0 0.5cm]{fig/visualization_1D/chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.C4.sorted.du}
         \label{fig:odgi_viz_color_by_path_pos}
@@
         % include second image
-        \includegraphics[width=\linewidth, trim=0 +2cm 0 0]{fig/visualization_1D/chr6.pan.fa.c3d3224.7748b33.395c7f4.smooth.gfa.C4.sorted.z}
+        \includegraphics[width=\linewidth, trim=0 +2cm 0 0.5cm]{fig/visualization_1D/chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.C4.sorted.z}
         \label{fig:odgi_viz_color_by_inversion_rate}
@@
         % include fourth image
-        \includegraphics[width=\linewidth, trim=0 1.5cm 0 0]{fig/visualization_1D/chr6.pan.fa.c3d3224.7748b33.395c7f4.smooth.gfa.C4.sorted.m}
+        \includegraphics[width=\linewidth, trim=0 1.5cm 0 0.5cm]{fig/visualization_1D/chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.C4.sorted.m}
         \label{fig:odgi_viz_color_by_path_depth}
@@
         Visualizing the complement component 4 (C4) pangenome graph extracted from a whole human pangenome graph of 90 haplotypes.
-        In all visualizations, 10 paths are displayed: 2 reference genomes (chm13 and grch38 on the top) and 8 haplotypes from 4 individuals.
+        In all visualizations, 8 paths are displayed: 2 reference genomes (chm13 and grch38 on the top) and 6 haplotypes of 3 individuals.
         \textbf{(a)} default modality: the image shows a quite linear graph.
-        The 2 longer links at the bottom indicate the presence of a structural variant (long link) with another structural variant nested inside it (short link).
+        The longer links at the bottom indicate the presence of a structural variant (long link) with another structural variant nested inside it (short link on the left).
         Indeed, human C4 exists as 2 functionally distinct genes, C4A and C4B, which both vary in structure and copy number \citep{Sekar_2016}.
 The longer link indicates that the copy number status varies across the haplotypes represented in the pangenome.
-        Moreover, C4A and C4B genes segregate in both long and short genomic forms, distinguished by the presence or absence of a human endogenous retroviral (HERV) sequence, as also highlighted by the short nested link on the bottom.
+        Moreover, C4A and C4B genes segregate in both long and short genomic forms, distinguished by the presence or absence of a human endogenous retroviral (HERV) sequence, as also highlighted by the short nested link on the left.
         \textbf{(b)} color by path position: the color gradients are smooths, highlighting that its node are well sorted in 1 dimension.
-        The top two reference genomes and 2 haplotypes (with names starting with HG01978\#1 and HG03492\#2) go from left to right, while 5 haplotypes go in the opposite direction, as indicated by the black color on their left.
+        The top two reference genomes and 2 haplotypes (with name starting with HG01952\#2) go from left to right, while 5 haplotypes go in the opposite direction, as indicated by the black color on their left.
         \textbf{(c)} color by strandness: the red paths indicate the haplotypes that were assembled in reverse with respect to the 2 reference genomes.
-        \textbf{(d)} color by path depth: using the Spectra color palette with 4 level of path depths, white indicates no depth, while grey, red, and yellow indicate depth 1, 2, and 3, respectively.
+        \textbf{(d)} color by path depth: using the Spectra color palette with 4 level of path depths, white indicates no depth, while grey, red, and yellow indicate depth 1, 2, and greater than or equal to 3, respectively.
         Coloring by path depth, we can see that the two references present two different allele copies of the C4 genes, both of them including the HERV sequence.
@@
         The path with name starting with HG01071\#2 presents 3 copies of the genes (indicated by the orange color), of which only 1 with the HERV sequence (gray color in the middle of the orange region).
+        In the haplotype with the name starting with HG01952\#1, the HERV sequence is absent.
     }

### 2021-10-22 scripts/visualization.sh

@@ -4,10 +4,10 @@
 # odgi viz with C4A/C4B locus graph
 #==================================
 
-filename_chr6_gfa=chr6.pan.fa.c3d3224.7748b33.395c7f4.smooth.gfa
+filename_chr6_gfa=chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa
 
 # Download and build the graph
-wget https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/scratch/2021_06_27_pggb/chroms/${filename_chr6_gfa}.gz
+wget http://hypervolu.me/~guarracino/${filename_chr6_gfa}.gz
 gunzip ${filename_chr6_gfa}.gz
 odgi build -g ${filename_chr6_gfa} -o ${filename_chr6_gfa}.og -t 16 -P
 
@@ -20,6 +20,8 @@ zgrep 'gene_id "C4A"\|gene_id "C4B"' hg38.ncbiRefSeq.gtf.gz | awk '$1 == "chr6"'
 odgi extract -i ${filename_chr6_gfa}.og -b C4.coordinates.bed -o - --full-range -t 16 -P |
   odgi explode -i - --biggest 1 --sorting-criteria P --optimize -p ${filename_chr6_gfa}.C4
 odgi sort -i ${filename_chr6_gfa}.C4.0.og -o ${filename_chr6_gfa}.C4.sorted.og -p Ygs -x 100 -t 16 -P
+
+odgi paths -i ${filename_chr6_gfa}.C4.sorted.og -L | grep 'chr6\|HG00438\|HG0107\|HG01952' > chr6.selected_paths.txt
 
 # Default (binned) mode
 odgi viz -i ${filename_chr6_gfa}.C4.sorted.og -o ${filename_chr6_gfa}.C4.sorted.png -w 50 -y 50 -p chr6.selected_paths.txt

### 2021-10-23 manuscript/fig_odgi_viz.tex
pjotrized image names

@@
         % include first image
-        \includegraphics[width=1.0\linewidth, trim=0 +2cm 0 0.5cm]{fig/visualization_1D/chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.C4.sorted}
+        \includegraphics[width=1.0\linewidth, trim=0 +2cm 0 0.5cm]{fig/visualization_1D/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted}
         \label{fig:odgi_viz_default}
@@
         % include second image
-        \includegraphics[width=1.0\linewidth, trim=0 +2cm 0 0.5cm]{fig/visualization_1D/chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.C4.sorted.du}
+        \includegraphics[width=1.0\linewidth, trim=0 +2cm 0 0.5cm]{fig/visualization_1D/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_du}
         \label{fig:odgi_viz_color_by_path_pos}
@@
         % include second image
-        \includegraphics[width=\linewidth, trim=0 +2cm 0 0.5cm]{fig/visualization_1D/chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.C4.sorted.z}
+        \includegraphics[width=\linewidth, trim=0 +2cm 0 0.5cm]{fig/visualization_1D/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_z}
         \label{fig:odgi_viz_color_by_inversion_rate}
@@
         % include fourth image
-        \includegraphics[width=\linewidth, trim=0 1.5cm 0 0.5cm]{fig/visualization_1D/chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.C4.sorted.m}
+        \includegraphics[width=\linewidth, trim=0 1.5cm 0 0.5cm]{fig/visualization_1D/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_m}
         \label{fig:odgi_viz_color_by_path_depth}

### 2021-10-23 scripts/visualization.sh

@@ -24,16 +24,16 @@ odgi sort -i ${filename_chr6_gfa}.C4.0.og -o ${filename_chr6_gfa}.C4.sorted.og -
 odgi paths -i ${filename_chr6_gfa}.C4.sorted.og -L | grep 'chr6\|HG00438\|HG0107\|HG01952' > chr6.selected_paths.txt
 
 # Default (binned) mode
-odgi viz -i ${filename_chr6_gfa}.C4.sorted.og -o ${filename_chr6_gfa}.C4.sorted.png -w 50 -y 50 -p chr6.selected_paths.txt
+odgi viz -i ${filename_chr6_gfa}.C4.sorted.og -o "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_C4_sorted.png -w 50 -y 50 -p chr6.selected_paths.txt
 
 # Color by strand
-odgi viz -i ${filename_chr6_gfa}.C4.sorted.og -o ${filename_chr6_gfa}.C4.sorted.z.png -w 50 -y 50 -p chr6.selected_paths.txt -z
+odgi viz -i ${filename_chr6_gfa}.C4.sorted.og -o "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_C4_sorted_z.png -w 50 -y 50 -p chr6.selected_paths.txt -z
 
 # Color by position
-odgi viz -i ${filename_chr6_gfa}.C4.sorted.og -o ${filename_chr6_gfa}.C4.sorted.du.png -w 50 -y 50 -p chr6.selected_paths.txt -du
+odgi viz -i ${filename_chr6_gfa}.C4.sorted.og -o "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_C4_sorted_du.png -w 50 -y 50 -p chr6.selected_paths.txt -du
 
 # Color by depth
-odgi viz -i ${filename_chr6_gfa}.C4.sorted.og -o ${filename_chr6_gfa}.C4.sorted.m.png -w 50 -y 50 -p chr6.selected_paths.txt -m -B Spectral:4
+odgi viz -i ${filename_chr6_gfa}.C4.sorted.og -o "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_C4_sorted_m.png -w 50 -y 50 -p chr6.selected_paths.txt -m -B Spectral:4
 
 #======================================================================================================================
 

### 2021-10-25 manuscript/main.tex
reordering of some subcommand sections

@@
 Expressed in terms of the variation graph $V$, ODGI's core $Node$ structure includes a decoder that maps the neighbors of each node to a dense range of integers.
-For a given node $i$ and neighbor $j$, the decoder itself does not store the $id$s of $j$, but rather a compact representation of the relative difference between the node ids: $\delta = Node_i.id - Node_j.id$.
+For a given node $i$ and neighbor $j$, the decoder itself does not store the $id$ of $j$, but rather a compact representation of the relative difference between the node ids: $\delta = Node_i.id - Node_j.id$.
 This keeps even the size of the encoding small, per common variation graph property (2).
@@
 \item \textit{odgi view} converts the graph into standard text formats (\S\ref{sec:text}).
-\item \textit{odgi stats} provides numerical properties of the graph (\S\ref{sec:text}).
 \item \textit{odgi matrix} derives the pangenome matrix (\S\ref{sec:text}).
@@
 \item \textit{odgi untangle} deconvolutes paths relative to a reference (\S\ref{sec:untangle}).
+\item \textit{odgi tips} finds path end points relative to a reference (\S\ref{sec:untangle}).
 \item \textit{odgi sort} orders the graph with sorting pipelines (\S\ref{sec:sort}).
 \item \textit{odgi layout} establishes a 2D layout using SGD (\S\ref{sec:sort}).
+\item \textit{odgi stats} provides numerical properties of the graph (\S\ref{sec:metrics}).
+\item \textit{odgi bin} generates a summarized view of the graph (\S\ref{sec:metrics}).
 \item \textit{odgi depth} describes path depth over graph and path positions (\S\ref{sec:metrics}).
 \item \textit{odgi degree} describes node degree over graph and path positions (\S\ref{sec:metrics}).
-\item \textit{odgi tips} finds path end points relative to a reference (\S\ref{sec:metrics}).
 \end{itemize}
@@
 
-\textit{odgi bin} generates a summarized view of gigabase scale graphs providing tab delimited or JSON `bins'. Also it is possible to output a FASTA file.
+\textit{odgi bin} generates a summarized view of gigabase scale graphs providing tab delimited or JSON `bins'.
 
@@
 Applying \textit{odgi stats}, users can retrieve metrics describing the graph properties, such as the number of nodes, edges, paths, and graph length.
+It outputs pangenome statistics in TSV or YAML textual file formats.
 MultiQC's \citep{Ewels_2016} ODGI module\footnote{\url{https://multiqc.info/docs/\#odgi} (accessed Oct 2021)} provides an interactive way to comparatively explore the statistics of an arbitrary number of graphs.
@@
 \textit{odgi bin} summarizes the path information into bins of a specified size, enabling a summarized view of gigabase scale graphs.
+It provides tab delimited or JSON `bins'.
 This concept found its way into the pangenome graph ontology \cite{Yokoyama2020}.
@@
 
-ODGI provides common operation on on standard pangenomic file formats, such as GFA.
+ODGI provides common operation on standard pangenomic file formats, such as GFA.
 ODGI aims to be a reference implementation for many pangenome related operations.

### 2021-10-26 manuscript/fig_metrics.tex
massively squeeze the metrics figure

@@
 	\end{subfigure}
-
-	\begin{subfigure}{\linewidth}
+	\begin{subfigure}{1\linewidth}
 	\caption{}
 	\centering
-		% include second image
-		\includegraphics[width=1.0\linewidth, trim=0 0.7cm -1cm 0.2cm]{fig/metrics/chr4_pan_fa_a2fb268_e820cd3_9ea71d8_smooth_gfa_og_HTTex1_og_O_og_tiny_og_png_svg}
-		\label{fig:metrics-viz}
+	% include fourth image
+	\includegraphics[width=\linewidth,trim=+.225cm 2.2cm +.425cm +3cm]{fig/metrics/chr4_HTT_chm13_degree_w1_bed}
+	\label{fig:metrics-degree}
 	\end{subfigure}
@@
 	% include second image
-		\includegraphics[width=\linewidth,trim=+.225cm 1.4cm +0.425cm +2.2cm]{fig/metrics/chr4_HTT_chm13_depth_w1_bed}
+	\includegraphics[width=\linewidth,trim=+.225cm 1.2cm +0.425cm +3cm]{fig/metrics/chr4_HTT_chm13_depth_w1_bed}
 	\label{fig:metrics-depth}
 	\end{subfigure}
-	\begin{subfigure}{1\linewidth}
+	\begin{subfigure}{\linewidth}
 		\caption{}
 		\centering
-		% include fourth image
-		\includegraphics[width=\linewidth,trim=+.225cm 1.5cm +.425cm +2.2cm]{fig/metrics/chr4_HTT_chm13_degree_w1_bed}
-		\label{fig:metrics-degree}
+		% include second image
+		\includegraphics[width=1.0\linewidth, trim=-1.5cm 0.75cm -0.8cm 0.5cm]{fig/metrics/chr4_pan_fa_a2fb268_e820cd3_9ea71d8_smooth_gfa_og_HTTex1_og_O_og_tiny_og_png_svg}
+		\label{fig:metrics-viz}
 	\end{subfigure}
@@
 The very high GC content of 73.0\% compared to a human genomic mean GC content of 40.9\% \cite{Piovesan2019} is in accordance with the literature (see for example \cite{Sathasivam2013, Neueder2017}).
-\textbf{(b)} \texttt{odgi viz} visualization of the 13 largest gene alleles, CHM13, and GRCH38 of the \textit{HTTexon1} graph.
+\textbf{(b)} Per nucleotide node degree distribution of CHM13 in the \textit{HTTexon1} graph.
+Around position 200 there is a huge variation in node degree.
 \textbf{(c)} Per nucleotide node depth distribution of CHM13 in the \textit{HTTexon1} graph.
-The alternating depth around position 200 indicates polymorphic variation.
-\textbf{(d)} Per nucleotide node degree distribution of CHM13 in the \textit{HTTexon1} graph.
-The polymorphic region around position 200 complements the above node depth analysis.
+The alternating depth around position 200 indicates polymorphic variation complementing the above node degree analysis.
+\textbf{(d)} \textit{odgi viz} visualization of the 13 largest gene alleles, CHM13, and GRCH38 of the \textit{HTTexon1} graph.
 Figures \textbf{(b)-(d)} clearly highlight the variant region around position 200 showing the variable number of glutamine residues of the different individuals as reported by \cite{Nance1999}.}

### 2021-10-26 manuscript/main.tex
funding SN
massively squeeze the metrics figure

@@
 \usepackage[usenames,dvipsnames]{xcolor}
+
+% we squeeze our figures even more together
+\captionsetup{belowskip=-2pt}
 
@@
 
+\textit{odgi position} translates graph and path positions between or within graphs emitting the liftovers in TSV format.
+
+\textit{odgi server} implements a basic HTTP server which can translate a 'path:position' to a 'pangenome:position' very efficiently.
+
 \textit{odgi degree/position/depth/untangle/tips} all can output the BED format.
@@
 
-To enable efficient sequence alignment against the graph, long nodes can be divided into shorter nodes at a maximum requested size using odgi chop.
-Partial order alignment, which consists of aligning sequences against a directed acyclic graph (DAG), is frequently used in pangenome building pipelines, but the current implementations return DAGs with 1-bp long nodes; odgi unchop allows joining nodes that can be merged without changing the graph topology, nor the embedded sequences, obtaining an equivalent, but more compact, representation of the graph.
+To enable efficient sequence alignment against the graph, long nodes can be divided into shorter nodes at a maximum requested size using \textit{odgi chop}.
+Partial order alignment, which consists of aligning sequences against a directed acyclic graph (DAG), is frequently used in pangenome building pipelines, but the current implementations return DAGs with 1-bp long nodes; \textit{odgi unchop} allows joining nodes that can be merged without changing the graph topology, nor the embedded sequences, obtaining an equivalent, but more compact, representation of the graph.
 
 Pangenome graphs can embed multiple chromosomes as separated connected components (inter-chromosomal structural variants would join the components into bigger ones).
-odgi explode separates the connected components in different ODGI format files, while odgi squeeze allows merging multiple graphs into the same ODGI format file, preventing node ID collisions.
+\textit{odgi explode} separates the connected components in different ODGI format files, while \textit{odgi squeeze} allows merging multiple graphs into the same ODGI format file, preventing node ID collisions.
 
-Cycles in the graph complicate downstream analyses: odgi break removes the cycles, reducing the complexity of the graph topology.
-odgi groom removes spurious inverting links by exploring the graph from the orientation supported by most paths.
+Cycles in the graph complicate downstream analyses: \textit{odgi break} removes the cycles, reducing the complexity of the graph topology.
+\textit{odgi groom} removes spurious inverting links by exploring the graph from the orientation supported by most paths.
 
@@
 Navigating and understanding the graph requires coordinate systems that we can use to link other data into the graph model, and thus to all genomes in the pangenome.
-Primarily, ODGI uses the embedded pangenome sequences to provide a set of coordinates in a standard, 1D format.e
+Primarily, ODGI uses the embedded pangenome sequences to provide a set of coordinates in a standard, 1D format.	
 These coordinates depend on the sequences, and are graph-independent.
-Our for positional conversion (\textit{odgi position}), annotation liftover (\textit{odgi untangle}), subgraph extraction (\textit{odgi extract}), visualization (\textit{odgi viz}), path depth (\textit{odgi depth}), and graph complexity measurement (\textit{odgi degree}) can all operate on coordinates of paths in the graph.
+Our for positional conversion (\textit{odgi position}), annotation liftover (\textit{odgi untangle}), breakpoint identification (\textit{odgi tips}), subgraph extraction (\textit{odgi extract}), visualization (\textit{odgi viz}), path depth (\textit{odgi depth}), and graph complexity measurement (\textit{odgi degree}) can all operate on coordinates of paths in the graph.
 
@@
 SH acknowledges funding from the Central Innovation Programme (ZIM) for SMEs of the Federal Ministry for Economic Affairs and Energy of Germany.
+SN acknowleges Germany’s Excellence Strategy (CMFI), EXC-2124 and (iFIT) - EXC 2180 – 390900677.
 This work was supported by the BMBF-funded de.NBI Cloud within the German Network for Bioinformatics Infrastructure (de.NBI) (031A537B, 031A533A, 031A538A, 031A533B, 031A535A, 031A537C, 031A534A, 031A532B).

### 2021-10-26 scripts/metrics.R

@@ -23,7 +23,7 @@ degree <- read.table(bed_degree, sep = "\t", quote = "", as.is = T,
 pdf(degree_pdf, height = 3, width = 14)
 plot(degree$V2, degree$V4, pch=20, 
 #     main = "odgi degree - chm13#chr4:3073405-3073983 - HTT exon1", 
-     xlab = "position of chm13 in HTT exon1",
+     xlab = "",
      ylab = "degree")
 lines(degree$V2, degree$V4, xlim=range(degree$V2), ylim=range(degree$V4), pch=1)
 dev.off()

### 2021-10-26 manuscript/main.tex
discussion: viz + scWGS

@@
 
-(recent) interactive visualization tools?! CITE PANACHE, also indicate that no published tool for interactive exploration of such graphs exist
-\FIXME{mention gfaestus?}
+ODGI's visualization tools produce static, large-scale snapshots.
+Recent interactive browsers are reference centric \citep{Beyer2019, Yokoyama2019, Durant2021} or focusing on one dimension \citep{Wick_2015}.
+However, an interactive visualization solution that combines the 1D and 2D layout of a graph with annotation and read mapping information across different zoom levels is still missing.
+ODGI's graph layout algorithms could become a backbone of such tools, as is already the case for \textit{gfaestus}\footnote{\url{https://github.com/chfi/gfaestus}}, which currently relies on the \textit{odgi layout} output for its interactive 2D visualization.
 
-is ODGI ready for scWGS data? potentially, each cell is a path in a graph!! very very high path depth
+So far, ODGI has not been applied to graphs build from scWGS \citep{Zhuo2021} data.
+Each cell would be represented as a path in the graph, the path depth would be very high.
+As ODGI was designed to deal with the path depth of human pangenome graphs, it remains to be seen, if it will work equally well with scWGS data.
 

### 2021-10-27 manuscript/fig_metrics.tex
one column-width metrics figure

@@
-\begin{figure*}[h!]
+\begin{figure}[h!]
 	\begin{subfigure}{\linewidth}
@@
 		% include first image
-		\includegraphics[width=1.0\linewidth, trim=0 1cm 0 0.5cm]{fig/metrics/chr4_pan_HTTex1_gfa_multiqc_odgi_stats}
+		\includegraphics[width=1.0\linewidth, trim=-1.25cm 3cm 0 1.75cm]{fig/metrics/chr4_pan_HTTex1_gfa_multiqc_odgi_stats}
 		\label{fig:metrics-multiqc}
@@
 	% include fourth image
-	\includegraphics[width=\linewidth,trim=+.225cm 2.2cm +.425cm +3cm]{fig/metrics/chr4_HTT_chm13_degree_w1_bed}
+	\includegraphics[width=\linewidth,trim=+.225cm 3cm +.425cm +3cm]{fig/metrics/chr4_HTT_chm13_degree_w1_bed}
 	\label{fig:metrics-degree}
@@
 	% include second image
-	\includegraphics[width=\linewidth,trim=+.225cm 1.2cm +0.425cm +3cm]{fig/metrics/chr4_HTT_chm13_depth_w1_bed}
+	\includegraphics[width=\linewidth,trim=+.225cm 3cm +0.425cm +3cm]{fig/metrics/chr4_HTT_chm13_depth_w1_bed}
 	\label{fig:metrics-depth}
@@
 The alternating depth around position 200 indicates polymorphic variation complementing the above node degree analysis.
-\textbf{(d)} \textit{odgi viz} visualization of the 13 largest gene alleles, CHM13, and GRCH38 of the \textit{HTTexon1} graph.
+\textbf{(d)} \textit{odgi viz} visualization of the 23 largest gene alleles, CHM13, and GRCH38 of the \textit{HTTexon1} graph.
 Figures \textbf{(b)-(d)} clearly highlight the variant region around position 200 showing the variable number of glutamine residues of the different individuals as reported by \cite{Nance1999}.}
 	\label{fig:metrics}
-\end{figure*}
+\end{figure}

### 2021-10-27 manuscript/main.tex
forgot citation
one column-width metrics figure

@@
 By considering all paths in the graph as potential reference or query sequences, we make the graph invisible to downstream tools that operate on collections of sequences (e.g.
-bedtools, samtools, aligners).
+BEDTools, SAMtools \citep{Li2009}, aligners).
 This approach benefits from the information in the graph without strongly embedding our methods in this difficult new research context.
@@
 
-\textit{odgi view} can convert a graph in ODGI binary format to the standard GFAv1 format. It can reveal a graph’s internal structures for e.g. debugging processes.
+\textit{odgi view} can convert a graph in ODGI binary format to the standard GFAv1\footnote{\url{https://github.com/GFA-spec/GFA-spec/blob/master/GFA1.md} (accessed Oct 2021)}  format.
+It can reveal a graph’s internal structures for e.g.
+debugging processes.
 
@@
 Among other metrics, it can calculate the number of nodes, edges, paths and the total nucleotide length of the graph.
+{\color{red} FIXME: All this information is also listed in the metrics section.
+How to split?
+No mentioning of file formats in the following sections?}
 
@@
 ODGI's visualization tools produce static, large-scale snapshots.
-Recent interactive browsers are reference centric \citep{Beyer2019, Yokoyama2019, Durant2021} or focusing on one dimension \citep{Wick_2015}.
+Recent interactive browsers are reference centric \citep{Beyer2019, Yokoyama2019, Durant2021, Liang2021} or focusing on one dimension \citep{Wick_2015, Gonnella2018}.
 However, an interactive visualization solution that combines the 1D and 2D layout of a graph with annotation and read mapping information across different zoom levels is still missing.

### 2021-10-27 scripts/metrics.sh

@@ -43,15 +43,17 @@ odgi build -g chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.g
 odgi paths -i chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.tiny.og -f > chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.tiny.og.fasta
 samtools faidx chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.tiny.og.fasta
 
-cat chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.tiny.og.fasta.fai | sort -k2 -r | head -n 14 | grep "chm13" -v | cut -f 1 > paths_to_viz
+cat chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.tiny.og.fasta.fai | sort -k2 -r | head -n 24 | grep "chm13" -v | cut -f 1 > paths_to_viz
 # take care of chm13 and grch13
 echo -e "chm13\ngrch38\n$(cat paths_to_viz)" > paths_to_viz
-odgi viz -i chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.tiny.og -o chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.tiny.og.png -a 8  -p paths_to_viz -c 100
+odgi viz -i chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.tiny.og -o chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.tiny.og.png -a 1  -p paths_to_viz
+
 #### MANUAL INTERVENTION ####
 # Go to https://drububu.com/tutorial/bitmap-to-vector.html. 
 # Drag and drop chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.tiny.og.png into the dashed field.
[…]

### 2021-10-27 manuscript/main.tex
remove or replace all footnotes

@@
 
-\textit{odgi view} can convert a graph in ODGI binary format to the standard GFAv1\footnote{\url{https://github.com/GFA-spec/GFA-spec/blob/master/GFA1.md} (accessed Oct 2021)}  format.
-It can reveal a graph’s internal structures for e.g.
-debugging processes.
+\textit{odgi view} can convert a graph in ODGI binary format to the standard GFAv1 \citep{GFA} format. It can reveal a graph’s internal structures for e.g. debugging processes.
 
@@
 %For the establishment of a layout in two dimensions \textit{odgi layout} applies a 2D adjusted PG-SGD algorithm.
-A calculated layout can be visualized statically with \textit{odgi draw} (see Figure \textit{\textbf{TODO MHC wgg.85}}) or interactively explored with the in-development \textit{gfaestus}\footnote{\url{https://github.com/chfi/gfaestus}}.
+A calculated layout can be visualized statically with \textit{odgi draw} (see Figure \textit{\textbf{TODO MHC wgg.85}}) or interactively explored with the in-development \textit{gfaestus}\{\url{https://github.com/chfi/gfaestus}}.
 
@@
 It outputs pangenome statistics in TSV or YAML textual file formats.
-MultiQC's \citep{Ewels_2016} ODGI module\footnote{\url{https://multiqc.info/docs/\#odgi} (accessed Oct 2021)} provides an interactive way to comparatively explore the statistics of an arbitrary number of graphs.
+MultiQC's \citep{Ewels_2016} ODGI module %\footnote{\url{https://multiqc.info/docs/\#odgi} (accessed Oct 2021)} 
+provides an interactive way to comparatively explore the statistics of an arbitrary number of graphs.
 ODGI also offers more advanced tools for the interrogation of the graphs.
@@
 It provides tab delimited or JSON `bins'.
-This concept found its way into the pangenome graph ontology \cite{Yokoyama2020}.
+%This concept found its way into the pangenome graph ontology \cite{Yokoyama2020}. %maybe we want to save this up for the discussion
 \textit{odgi depth} returns the node depth as defined by query criteria, allowing users to assess the complexity of regions in the graph.

### 2021-10-27 manuscript/main.tex
replace some more lost footnotes

@@
 ODGI is a unique set of tools that enables scientists to explore and discover the underlying biology of pangenome graphs.
-Already, the tools are the backbone of pipelines such as the Pangenome Graph Builder\footnote{\url{https://github.com/pangenome/pggb} (accessed Oct 2021)} (PGGB) or nf-core/pangenome\footnote{\url{https://github.com/nf-core/pangenome} (accessed Oct 2021)}.
+Already, the tools are the backbone of pipelines such as the Pangenome Graph Builder \citep{pggb} (PGGB) or nf-core/pangenome \citep{pangenome}.
+%\footnote{\url{https://github.com/nf-core/pangenome} (accessed Oct 2021)}.
 Future work will add support for RNA and protein sequences and expand on metadata capabilities of large pangenome graphs.
@@
 However, an interactive visualization solution that combines the 1D and 2D layout of a graph with annotation and read mapping information across different zoom levels is still missing.
-ODGI's graph layout algorithms could become a backbone of such tools, as is already the case for \textit{gfaestus}\footnote{\url{https://github.com/chfi/gfaestus}}, which currently relies on the \textit{odgi layout} output for its interactive 2D visualization.
+ODGI's graph layout algorithms could become a backbone of such tools, as is already the case for \textit{gfaestus} \citep{gfaestus}%\footnote{\url{https://github.com/chfi/gfaestus}}, which currently relies on the \textit{odgi layout} output for its interactive 2D visualization.
 

### 2021-10-27 scripts/visualization.sh
odgi layout and draw command lines; code cleaning and commenting

@@ -1,9 +1,5 @@
 #!/bin/bash
 
-#======================================================================================================================
-# odgi viz with C4A/C4B locus graph
-#==================================
-
 filename_chr6_gfa=chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa
 
 # Download and build the graph
@@ -11,33 +7,39 @@ wget http://hypervolu.me/~guarracino/${filename_chr6_gfa}.gz
 gunzip ${filename_chr6_gfa}.gz
 odgi build -g ${filename_chr6_gfa} -o ${filename_chr6_gfa}.og -t 16 -P
 
+
 # Find C4 coordinates
 wget http://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/hg38.chrom.sizes
 wget https://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/genes/hg38.ncbiRefSeq.gtf.gz
 zgrep 'gene_id "C4A"\|gene_id "C4B"' hg38.ncbiRefSeq.gtf.gz | awk '$1 == "chr6"' | cut -f 1,4,5 | bedtools sort | bedtools merge -d 15000 | bedtools slop -l 10000 -r 20000 -g hg38.chrom.sizes | sed 's/chr6/grch38#chr6/g' >C4.coordinates.bed
+
 
 # Extraction, explosion, optimization, and sorting
 odgi extract -i ${filename_chr6_gfa}.og -b C4.coordinates.bed -o - --full-range -t 16 -P |
   odgi explode -i - --biggest 1 --sorting-criteria P --optimize -p ${filename_chr6_gfa}.C4
 odgi sort -i ${filename_chr6_gfa}.C4.0.og -o ${filename_chr6_gfa}.C4.sorted.og -p Ygs -x 100 -t 16 -P
 
+
+# Select haplotypes
 odgi paths -i ${filename_chr6_gfa}.C4.sorted.og -L | grep 'chr6\|HG00438\|HG0107\|HG01952' > chr6.selected_paths.txt
 
-# Default (binned) mode
+
+# odgi viz: default (binned) mode
 odgi viz -i ${filename_chr6_gfa}.C4.sorted.og -o "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_C4_sorted.png -w 50 -y 50 -p chr6.selected_paths.txt
 
-# Color by strand
+# odgi viz: color by strand
[…]

### 2021-10-27 manuscript/fig_odgi_viz.tex
change figure 2 layout

@@
 \begin{figure*}[h!]
-    \begin{subfigure}{\linewidth}
+    \begin{subfigure}[b]{.50\linewidth}
         \caption{}
@@
         % include first image
-        \includegraphics[width=1.0\linewidth, trim=0 +2cm 0 0.5cm]{fig/visualization_1D/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted}
+        \includegraphics[width=\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/visualization_1D/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted}
         \label{fig:odgi_viz_default}
     \end{subfigure}
-    \begin{subfigure}{\linewidth}
+    \begin{subfigure}[b]{.50\linewidth}
+        \caption{}
+        \centering
+        % include first image
+        \includegraphics[width=0.5\linewidth, trim=0 +2cm 0 0.5cm]{fig/visualization_1D/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_layout}
+        \label{fig:odgi_viz_default2}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{.50\linewidth}
         \caption{}
@@
         % include second image
-        \includegraphics[width=1.0\linewidth, trim=0 +2cm 0 0.5cm]{fig/visualization_1D/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_du}
+        \includegraphics[width=\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/visualization_1D/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_du}
         \label{fig:odgi_viz_color_by_path_pos}
     \end{subfigure}
-    \begin{subfigure}{\linewidth}
+
+    \begin{subfigure}[b]{.50\linewidth}
         \caption{}
@@
         % include second image
-        \includegraphics[width=\linewidth, trim=0 +2cm 0 0.5cm]{fig/visualization_1D/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_z}
+        \includegraphics[width=\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/visualization_1D/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_z}
         \label{fig:odgi_viz_color_by_inversion_rate}
     \end{subfigure}
-    \begin{subfigure}{1\linewidth}
+    \begin{subfigure}[b]{.50\linewidth}
+        \caption{}
+        \centering
+        % include second image
+        %todo BANDAGE
+        \includegraphics[width=0.5\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/visualization_1D/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_layout}
+        \label{fig:odgi_viz_color_by_inversion_rate2}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{.50\linewidth}
         \caption{}
@@
         % include fourth image
-        \includegraphics[width=\linewidth, trim=0 1.5cm 0 0.5cm]{fig/visualization_1D/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_m}
+        \includegraphics[width=\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/visualization_1D/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_m}
         \label{fig:odgi_viz_color_by_path_depth}
     \end{subfigure}
+
     \caption{

### 2021-10-27 scripts/visualization.sh

@@ -25,21 +25,21 @@ odgi paths -i ${filename_chr6_gfa}.C4.sorted.og -L | grep 'chr6\|HG00438\|HG0107
 
 
 # odgi viz: default (binned) mode
-odgi viz -i ${filename_chr6_gfa}.C4.sorted.og -o "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_C4_sorted.png -w 50 -y 50 -p chr6.selected_paths.txt
+odgi viz -i ${filename_chr6_gfa}.C4.sorted.og -o "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_C4_sorted.png -c 12 -w 100 -y 50 -p chr6.selected_paths.txt
 
 # odgi viz: color by strand
-odgi viz -i ${filename_chr6_gfa}.C4.sorted.og -o "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_C4_sorted_z.png -w 50 -y 50 -p chr6.selected_paths.txt -z
+odgi viz -i ${filename_chr6_gfa}.C4.sorted.og -o "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_C4_sorted_z.png -c 12 -w 100 -y 50 -p chr6.selected_paths.txt -z
 
 # odgi viz: color by position
-odgi viz -i ${filename_chr6_gfa}.C4.sorted.og -o "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_C4_sorted_du.png -w 50 -y 50 -p chr6.selected_paths.txt -du
+odgi viz -i ${filename_chr6_gfa}.C4.sorted.og -o "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_C4_sorted_du.png -c 12 -w 100 -y 50 -p chr6.selected_paths.txt -du
 
 # odgi viz: color by depth
-odgi viz -i ${filename_chr6_gfa}.C4.sorted.og -o "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_C4_sorted_m.png -w 50 -y 50 -p chr6.selected_paths.txt -m -B Spectral:4
+odgi viz -i ${filename_chr6_gfa}.C4.sorted.og -o "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_C4_sorted_m.png -c 12 -w 100 -y 50 -p chr6.selected_paths.txt -m -B Spectral:4
 
 
[…]

### 2021-10-27 scripts/visualization.sh
rename fig2 directory; improved fig2 layout

@@ -38,8 +38,8 @@ odgi viz -i ${filename_chr6_gfa}.C4.sorted.og -o "$(echo ${filename_chr6_gfa} | 
 
 
 # Compute layout
-odgi layout -i ${filename_chr6_gfa}.C4.sorted.og -o ${filename_chr6_gfa}.C4.sorted.lay -T ${filename_chr6_gfa}.C4.sorted.tsv -x 30 -t 16 -P
+odgi layout -i ${filename_chr6_gfa}.C4.sorted.og -o ${filename_chr6_gfa}.C4.sorted.lay -T ${filename_chr6_gfa}.C4.sorted.tsv -x 100 -t 16 -P
 
 # odgi draw
-odgi draw -i ${filename_chr6_gfa}.C4.sorted.og -c ${filename_chr6_gfa}.C4.sorted.lay -p "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_C4_sorted_layout.png
+odgi draw -i ${filename_chr6_gfa}.C4.sorted.og -c ${filename_chr6_gfa}.C4.sorted.lay -p "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_C4_sorted_layout.png -H 1000 -w 1000 -B 500
 

### 2021-10-27 manuscript/fig_build_time.tex
odgi build performance figure

\begin{figure}[h!]
	\centering
	\includegraphics[width=\linewidth]{fig/build/mean_build_time.pdf}
	\caption{Performance evaluation of \textit{odgi build} when translating a 90 haplotypes human chromosome 6 graph into ODGI's native format. The error margins are the standard deviations computed from 10 runs for each of the different threading configurations.}
	\label{fig:build}
\end{figure}

### 2021-10-27 scripts/build_time.sh

@@ -1,0 +1,42 @@
+wget http://hypervolu.me/~guarracino/chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.gz
+gunzip chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.gz
+
+G=chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa
+
+# t1
+for i in {1..10}
+do
+    /usr/bin/time odgi build -g "$G" -o "$G".og -t 1 2> chr6_odgi_build_time_t1_"$i"
+done
+
+#t2
+for i in {1..10}
+do
+    /usr/bin/time odgi build -g "$G" -o "$G".og -t 2 2> chr6_odgi_build_time_t2_"$i"
+done
+
+#t4
+for i in {1..10}
+do
+    /usr/bin/time odgi build -g "$G" -o "$G".og -t 4 2> chr6_odgi_build_time_t4_"$i"
+done
+
+#t8
+for i in {1..10}
+do
+    /usr/bin/time odgi build -g "$G" -o "$G".og -t 8 2> chr6_odgi_build_time_t8_"$i"
+done
+
+#t16
+for i in {1..10}
+do
+    /usr/bin/time odgi build -g "$G" -o "$G".og -t 16 2> chr6_odgi_build_time_t16_"$i"
+done
+
+for T in 1 2 4 8 16
+do
+    grep elapsed chr6_odgi_build_time_t"$T"_* | tr ' ' '\n' | grep elapsed | sed 's/elapsed//g' | awk -F: '{ print ($1 * 60) + ($2) + $3 }' | sed -e "s/$/,$T/" >> mean_build_time.csv
+    
+done
+
+Rscript time.R mean_build_time.csv mean_build_time.pdf

### 2021-10-27 scripts/time.R

@@ -1,0 +1,53 @@
+#!/usr/bin/env Rscript
+
+library(ggplot2)
+
+args <- commandArgs(trailingOnly = T)
+
+times_csv <- args[1]
+times_pdf <- args[2]
+
+# times_csv <- "/home/heumos/Downloads/odgi_paper/MHC/mean_build_time.csv"
+
+times <- read.table(times_csv, sep = ",", quote = "", as.is = T, 
+                    check.names = F, header = F, comment.char = "")
+colnames(times) <- c("time", "threads")
+
+times$threads <- as.character(times$threads)
+times$threads <- factor(times$threads, levels=unique(times$threads))
+
+#+++++++++++++++++++++++++
+# Function to calculate the mean and the standard deviation
+# for each group
+#+++++++++++++++++++++++++
+# data : a data frame
+# varname : the name of a column containing the variable
+#to be summariezed
+# groupnames : vector of column names to be used as
+# grouping variables
+data_summary <- function(data, varname, groupnames){
+  require(plyr)
+  summary_func <- function(x, col){
+    c(mean = mean(x[[col]], na.rm=TRUE),
+      sd = sd(x[[col]], na.rm=TRUE))
+  }
+  data_sum<-ddply(data, groupnames, .fun=summary_func,
+                  varname)
+  data_sum <- rename(data_sum, c("mean" = varname))
+  return(data_sum)
+}
+
+df2 <- data_summary(times, varname="time", 
+                    groupnames=c("threads"))
+df2$threads=factor(df2$threads, levels=unique(times$threads))
+
+p <- ggplot(data=df2, aes(x=threads, y=time, color=threads, group=1)) +
+  geom_errorbar(aes(ymin=time-sd, ymax=time+sd), width=.2) +
+  geom_line(color = "black") +
+  geom_point(size = 1) + 
+  # geom_bar(stat="identity") +
+  # scale_fill_brewer(palette="Dark2") +
+  labs(x = "number of threads", y = "time in seconds")
+pdf(times_pdf, height = 2, width = 3)
+p
+dev.off()

### 2021-10-27 manuscript/fig_odgi_viz.tex

--- 2021-10-23 manuscript/fig_odgi_viz.tex
@@
-\begin{figure*}[h!]
-    \begin{subfigure}{\linewidth}
+\begin{figure*}[!htb]
+    \centering
+    \begin{tabular}[t]{cc}%{|c|c|}
+        %\hline
+        \begin{tabular}{c}% if you add [t], then sub images are pushed down
+            %\smallskip
+            \begin{subfigure}[t]{0.5\textwidth}
+                \centering
                 \caption{}
+                \includegraphics[width=\linewidth, trim=0 +1cm -0.5cm 0.5cm]{fig/fig2_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted}
+            \end{subfigure} \\
+            \begin{subfigure}[t]{0.5\textwidth}
                 \centering
-        % include first image
-        \includegraphics[width=1.0\linewidth, trim=0 +2cm 0 0.5cm]{fig/visualization_1D/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted}
-        \label{fig:odgi_viz_default}
+                \caption{}
+                \includegraphics[width=\linewidth, trim=0 +1cm -0.5cm 0.5cm]{fig/fig2_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_du}
+            \end{subfigure} \\
+            \begin{subfigure}[t]{0.5\textwidth}
+                \centering
+                \caption{}
+                \includegraphics[width=\linewidth, trim=0 +1cm -0.5cm 0.5cm]{fig/fig2_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_z}
+            \end{subfigure} \\
+            \begin{subfigure}[t]{0.5\textwidth}
+                \centering
+                \caption{}
+                \includegraphics[width=\linewidth, trim=0 +1cm -0.5cm 0.5cm]{fig/fig2_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_m}
             \end{subfigure}
-    \begin{subfigure}{\linewidth}
+        \end{tabular}
+        %&
+        \begin{tabular}{c}% if you add [t], than sub images are pushed down
+            %\smallskip
+            \begin{subfigure}[t]{0.5\textwidth}
+                \centering
+                \includegraphics[width=0.65\linewidth, trim=0 +2cm 0 0.5cm]{fig/fig2_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_layout}
                 \caption{}
+            \end{subfigure} \\
+            \\
+
+            \begin{subfigure}[t]{0.5\textwidth}
                 \centering
-        % include second image
-        \includegraphics[width=1.0\linewidth, trim=0 +2cm 0 0.5cm]{fig/visualization_1D/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_du}
-        \label{fig:odgi_viz_color_by_path_pos}
+                \includegraphics[width=0.65\linewidth, trim=0 +2cm 0 0.5cm]{fig/fig2_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_layout}
+                \caption{}
             \end{subfigure}
-    \begin{subfigure}{\linewidth}
-        \caption{}
-        \centering
-        % include second image
-        \includegraphics[width=\linewidth, trim=0 +2cm 0 0.5cm]{fig/visualization_1D/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_z}
-        \label{fig:odgi_viz_color_by_inversion_rate}
-    \end{subfigure}
-    \begin{subfigure}{1\linewidth}
-        \caption{}
-        \centering
-        % include fourth image
-        \includegraphics[width=\linewidth, trim=0 1.5cm 0 0.5cm]{fig/visualization_1D/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_m}
-        \label{fig:odgi_viz_color_by_path_depth}
-    \end{subfigure}
+        \end{tabular}\\
+        %\hline
+    \end{tabular}
     \caption{
@@
 \end{figure*}
+
+%\begin{figure*}[h!]
+%    \begin{subfigure}[t]{.50\linewidth}
+%        \caption{}
+%        \centering
+%        % include first image
+%        \includegraphics[width=\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/fig2_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted}
+%        \label{fig:odgi_viz_default}
+%    \end{subfigure}
+%
+%    \begin{subfigure}[t]{.50\linewidth}
+%        \caption{}
+%        \centering
+%        % include second image
+%        \includegraphics[width=\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/fig2_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_du}
+%        \label{fig:odgi_viz_color_by_path_pos}
+%    \end{subfigure}
+%
+%    \begin{subfigure}[t]{.50\linewidth}
+%        \caption{}
+%        \centering
+%        % include second image
+%        \includegraphics[width=\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/fig2_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_z}
+%        \label{fig:odgi_viz_color_by_inversion_rate}
+%    \end{subfigure}
+%
+%    \begin{subfigure}[t]{.50\linewidth}
+%        \caption{}
+%        \centering
+%        % include fourth image
+%        \includegraphics[width=\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/fig2_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_m}
+%        \label{fig:odgi_viz_color_by_path_depth}
+%    \end{subfigure}
+%    \begin{subfigure}[t]{.50\linewidth}
+%        \caption{}
+%        \centering
+%        % include first image
+%        \includegraphics[width=0.5\linewidth, trim=0 +2cm 0 0.5cm]{fig/fig2_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_layout}
+%        \label{fig:odgi_viz_default2}
+%    \end{subfigure}
+%    \begin{subfigure}[t]{.50\linewidth}
+%        \caption{}
+%        \centering
+%        % include second image
+%        %todo BANDAGE
+%        \includegraphics[width=0.5\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/fig2_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_layout}
+%        \label{fig:odgi_viz_color_by_inversion_rate2}
+%    \end{subfigure}
+%
+%    \caption{
+%        Visualizing the complement component 4 (C4) pangenome graph extracted from a whole human pangenome graph of 90 haplotypes.
+%        In all visualizations, 8 paths are displayed: 2 reference genomes (chm13 and grch38 on the top) and 6 haplotypes of 3 individuals.
+%        \textbf{(a)} default modality: the image shows a quite linear graph.
+%        The longer links at the bottom indicate the presence of a structural variant (long link) with another structural variant nested inside it (short link on the left).
+%        Indeed, human C4 exists as 2 functionally distinct genes, C4A and C4B, which both vary in structure and copy number \citep{Sekar_2016}.
+The longer link indicates that the copy number status varies across the haplotypes represented in the pangenome.
+%        Moreover, C4A and C4B genes segregate in both long and short genomic forms, distinguished by the presence or absence of a human endogenous retroviral (HERV) sequence, as also highlighted by the short nested link on the left.
+%        \textbf{(b)} color by path position: the color gradients are smooths, highlighting that its node are well sorted in 1 dimension.
+%        The top two reference genomes and 2 haplotypes (with name starting with HG01952\#2) go from left to right, while 5 haplotypes go in the opposite direction, as indicated by the black color on their left.
+%        \textbf{(c)} color by strandness: the red paths indicate the haplotypes that were assembled in reverse with respect to the 2 reference genomes.
+%        \textbf{(d)} color by path depth: using the Spectra color palette with 4 level of path depths, white indicates no depth, while grey, red, and yellow indicate depth 1, 2, and greater than or equal to 3, respectively.
+%        Coloring by path depth, we can see that the two references present two different allele copies of the C4 genes, both of them including the HERV sequence.
+%        The entirely grey paths have one copy of these genes.
+%        The path with name starting with HG01071\#2 presents 3 copies of the genes (indicated by the orange color), of which only 1 with the HERV sequence (gray color in the middle of the orange region).
+%        In the haplotype with the name starting with HG01952\#1, the HERV sequence is absent.
+%    }
+%    \label{fig:odgi_viz}
+%\end{figure*}

### 2021-10-27 manuscript/fig_odgi_viz.tex
added code for the bandage figure

@@
                 \caption{}
-                \includegraphics[width=\linewidth, trim=0 +1cm -0.5cm 0.5cm]{fig/fig2_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted}
+                \includegraphics[width=\linewidth, trim=0 +1cm -0.5cm 0.5cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted}
             \end{subfigure} \\
@@
                 \caption{}
-                \includegraphics[width=\linewidth, trim=0 +1cm -0.5cm 0.5cm]{fig/fig2_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_du}
+                \includegraphics[width=\linewidth, trim=0 +1cm -0.5cm 0.5cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_du}
             \end{subfigure} \\
@@
                 \caption{}
-                \includegraphics[width=\linewidth, trim=0 +1cm -0.5cm 0.5cm]{fig/fig2_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_z}
+                \includegraphics[width=\linewidth, trim=0 +1cm -0.5cm 0.5cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_z}
             \end{subfigure} \\
@@
                 \caption{}
-                \includegraphics[width=\linewidth, trim=0 +1cm -0.5cm 0.5cm]{fig/fig2_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_m}
+                \includegraphics[width=\linewidth, trim=0 +1cm -0.5cm 0.5cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_m}
             \end{subfigure}
@@
                 \centering
-                \includegraphics[width=0.65\linewidth, trim=0 +2cm 0 0.5cm]{fig/fig2_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_layout}
+                \includegraphics[width=0.65\linewidth, trim=0 +2cm 0 0.5cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_layout}
                 \caption{}
@@
 %        % include first image
-%        \includegraphics[width=\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/fig2_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted}
+%        \includegraphics[width=\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted}
 %        \label{fig:odgi_viz_default}
@@
 %        % include second image
-%        \includegraphics[width=\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/fig2_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_du}
+%        \includegraphics[width=\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_du}
 %        \label{fig:odgi_viz_color_by_path_pos}
@@
 %        % include second image
-%        \includegraphics[width=\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/fig2_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_z}
+%        \includegraphics[width=\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_z}
 %        \label{fig:odgi_viz_color_by_inversion_rate}
@@
 %        % include fourth image
-%        \includegraphics[width=\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/fig2_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_m}
+%        \includegraphics[width=\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_m}
 %        \label{fig:odgi_viz_color_by_path_depth}
@@
 %        % include first image
-%        \includegraphics[width=0.5\linewidth, trim=0 +2cm 0 0.5cm]{fig/fig2_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_layout}
+%        \includegraphics[width=0.5\linewidth, trim=0 +2cm 0 0.5cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_layout}
 %        \label{fig:odgi_viz_default2}
@@
 %        %todo BANDAGE
-%        \includegraphics[width=0.5\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/fig2_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_layout}
+%        \includegraphics[width=0.5\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_layout}
 %        \label{fig:odgi_viz_color_by_inversion_rate2}

### 2021-10-27 scripts/visualization.sh

@@ -11,7 +11,7 @@ odgi build -g ${filename_chr6_gfa} -o ${filename_chr6_gfa}.og -t 16 -P
 # Find C4 coordinates
 wget http://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/hg38.chrom.sizes
 wget https://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/genes/hg38.ncbiRefSeq.gtf.gz
-zgrep 'gene_id "C4A"\|gene_id "C4B"' hg38.ncbiRefSeq.gtf.gz | awk '$1 == "chr6"' | cut -f 1,4,5 | bedtools sort | bedtools merge -d 15000 | bedtools slop -l 10000 -r 20000 -g hg38.chrom.sizes | sed 's/chr6/grch38#chr6/g' >C4.coordinates.bed
+zgrep 'gene_id "C4A"\|gene_id "C4B"' hg38.ncbiRefSeq.gtf.gz | awk '$1 == "chr6"' | cut -f 1,4,5 | bedtools sort | bedtools merge -d 15000 | bedtools slop -l 10000 -r 20000 -g hg38.chrom.sizes | sed 's/chr6/grch38#chr6/g' > C4.coordinates.bed
 
 
 # Extraction, explosion, optimization, and sorting
@@ -43,3 +43,11 @@ odgi layout -i ${filename_chr6_gfa}.C4.sorted.og -o ${filename_chr6_gfa}.C4.sort
 # odgi draw
 odgi draw -i ${filename_chr6_gfa}.C4.sorted.og -c ${filename_chr6_gfa}.C4.sorted.lay -p "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_C4_sorted_layout.png -H 1000 -w 1000 -B 500
 
+
+# Get C4-GTF
+zgrep 'gene_id "C4A"\|gene_id "C4B"' hg38.ncbiRefSeq.gtf.gz | awk '$1 == "chr6" && $3 == "transcript"' | sed 's/chr6/grch38#chr6/g' | head -n 2 > chr6.C4.gtf
+#grch38#chr6	ncbiRefSeq.2021-09-09	transcript	32014795	32035418	.	+	.	gene_id "C4B"; transcript_id "NM_001002029.4_4";  gene_name "C4B";
[…]

### 2021-10-28 scripts/mhc.sh
add MHC 2D layout

@@ -1,0 +1,21 @@
+wget http://hypervolu.me/~guarracino/chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.gz
+gunzip chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.gz
+
+MHC=chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.og
+odgi build -g chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa -o "$MHC" -t 28 -P
+odgi extract -i "$MHC" -o "$MHC".mhc.og -E -P -t 28 -r grch38#chr6:29000000-34000000
+odgi sort -i "$MHC".mhc.og -o "$MHC".mhc.og.O.og -O
+odgi layout -i "$MHC".mhc.og.O.og -o "$MHC".mhc.og.O.og.lay -T "$MHC".mhc.og.O.og.lay.tsv -P -t 28 -G 20 -x 100
+odgi draw -i chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.og.mhc.og.O.og -c chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.og.mhc.og.O.og.lay -p chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.og.mhc.og.O.og.lay.H1000w10.png -H 1000 -C -w 10 -t 28 -P
+
+#### MANUAL INTERVENTION ####
+# Go to https://drububu.com/tutorial/bitmap-to-vector.html. 
+# Drag and drop chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.og.mhc.og.O.og.lay.H1000w100.png into the dashed field.
+# Click "save as svg" to chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.og.mhc.og.O.og.lay.H1000w100.png.svg. This might take some time, be patient!
+
+inkscape chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.og.mhc.og.O.og.lay.H1000w100.png.svg --export-pdf=chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_og_mhc_og_O_og_lay_H1000w100_png_svg.pdf
+
+#### MANUAL INTERVENTION ####
+# Add box around C4 using Inkscape
+
+

### 2021-10-28 manuscript/fig_odgi_viz.tex
add MHC 2D layout
update caption

@@
         %\hline
+            \begin{subfigure}[t]{1.0\textwidth}
+        	\centering
+    		\caption{}
+			\includegraphics[width=\linewidth, trim=0cm 0cm -10cm 0cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_og_mhc_og_O_og_lay_H1000w100_png_svg.pdf}
+        \end{subfigure} \\
         \begin{tabular}{c}% if you add [t], then sub images are pushed down
@@
     \caption{
-        Visualizing the complement component 4 (C4) pangenome graph extracted from a whole human pangenome graph of 90 haplotypes.
+        \textbf{(a)} 2 dimensional drawing of the major histocompatibility complex (MHC) pangenome graph extracted from a whole human pangenome graph of 90 haplotypes created with \textit{odgi draw}.
+The red box highlights the complement component 4 (C4) gene.
+\\
+        \textbf{(b-e)} Visualizing the C4 pangenome graph extracted from a whole human pangenome graph of 90 haplotypes.
         In all visualizations, 8 paths are displayed: 2 reference genomes (chm13 and grch38 on the top) and 6 haplotypes of 3 individuals.
-        \textbf{(a)} default modality: the image shows a quite linear graph.
+        \textbf{(b)} default modality: the image shows a quite linear graph.
         The longer links at the bottom indicate the presence of a structural variant (long link) with another structural variant nested inside it (short link on the left).
@@
         Moreover, C4A and C4B genes segregate in both long and short genomic forms, distinguished by the presence or absence of a human endogenous retroviral (HERV) sequence, as also highlighted by the short nested link on the left.
-        \textbf{(b)} color by path position: the color gradients are smooths, highlighting that its node are well sorted in 1 dimension.
+        \textbf{(c)} color by path position: the color gradients are smooths, highlighting that its node are well sorted in 1 dimension.
         The top two reference genomes and 2 haplotypes (with name starting with HG01952\#2) go from left to right, while 5 haplotypes go in the opposite direction, as indicated by the black color on their left.
-        \textbf{(c)} color by strandness: the red paths indicate the haplotypes that were assembled in reverse with respect to the 2 reference genomes.
-        \textbf{(d)} color by path depth: using the Spectra color palette with 4 level of path depths, white indicates no depth, while grey, red, and yellow indicate depth 1, 2, and greater than or equal to 3, respectively.
+        \textbf{(d)} color by strandness: the red paths indicate the haplotypes that were assembled in reverse with respect to the 2 reference genomes.
+        \textbf{(e)} color by path depth: using the Spectra color palette with 4 level of path depths, white indicates no depth, while grey, red, and yellow indicate depth 1, 2, and greater than or equal to 3, respectively.
         Coloring by path depth, we can see that the two references present two different allele copies of the C4 genes, both of them including the HERV sequence.

### 2021-10-28 manuscript/fig_odgi_viz.tex
placeholders

--- 2021-10-27 manuscript/fig_odgi_viz.tex
@@
         %\hline
+        \begin{subfigure}[t]{1.0\textwidth}
+            \centering
+            \caption{}
+            \includegraphics[width=\linewidth, trim=0cm 0cm -10cm 0cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_og_mhc_og_O_og_lay_H1000w100_png_svg}
+        \end{subfigure} \\
         \begin{tabular}{c}% if you add [t], then sub images are pushed down
@@
         \end{tabular}\\
+        \begin{subfigure}[t]{1.0\textwidth}
+            \centering
+            \caption{}
+            \includegraphics[width=\linewidth, trim=0cm 0cm -10cm 0cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_og_mhc_og_O_og_lay_H1000w100_png_svg}
+        \end{subfigure} \\
         %\hline
@@
     \caption{
-        Visualizing the complement component 4 (C4) pangenome graph extracted from a whole human pangenome graph of 90 haplotypes.
-        In all visualizations, 8 paths are displayed: 2 reference genomes (chm13 and grch38 on the top) and 6 haplotypes of 3 individuals.
-        \textbf{(a)} default modality: the image shows a quite linear graph.
+        Visualizing the MHC and C4 \textit{loci} pangenome graphs.
+\textbf{(a)} \textit{odgi draw} layout of the MHC pangenome graph extracted from a whole human pangenome graph of 90 haplotypes.
+The red box highlights the C4 region.
+        \textbf{(b-e)} \textit{odgi viz} visualizations of the C4 pangenome graph, where 8 paths are displayed: 2 reference genomes (chm13 and grch38 on the top) and 6 haplotypes of 3 individuals.
+        \textbf{(b)} \textit{odgi viz} default modality: the image shows a quite linear graph.
         The longer links at the bottom indicate the presence of a structural variant (long link) with another structural variant nested inside it (short link on the left).
@@
         Moreover, C4A and C4B genes segregate in both long and short genomic forms, distinguished by the presence or absence of a human endogenous retroviral (HERV) sequence, as also highlighted by the short nested link on the left.
-        \textbf{(b)} color by path position: the color gradients are smooths, highlighting that its node are well sorted in 1 dimension.
+        \textbf{(c)} Color by path position: the color gradients are smooth, highlighting that its node are well sorted in 1 dimension.
         The top two reference genomes and 2 haplotypes (with name starting with HG01952\#2) go from left to right, while 5 haplotypes go in the opposite direction, as indicated by the black color on their left.
-        \textbf{(c)} color by strandness: the red paths indicate the haplotypes that were assembled in reverse with respect to the 2 reference genomes.
-        \textbf{(d)} color by path depth: using the Spectra color palette with 4 level of path depths, white indicates no depth, while grey, red, and yellow indicate depth 1, 2, and greater than or equal to 3, respectively.
+        \textbf{(d)} \textit{odgi viz} color by strandness: the red paths indicate the haplotypes that were assembled in reverse with respect to the 2 reference genomes.
+        \textbf{(e)} \textit{odgi viz} color by path depth: using the Spectra color palette with 4 level of path depths, white indicates no depth, while grey, red, and yellow indicate depth 1, 2, and greater than or equal to 3, respectively.
         Coloring by path depth, we can see that the two references present two different allele copies of the C4 genes, both of them including the HERV sequence.
@@
         In the haplotype with the name starting with HG01952\#1, the HERV sequence is absent.
+        \textbf{(f)} \textit{Bandage} layout of the C4 pangenome graph. The red box highlighs XXX.
+        \textbf{(g)} \textit{Bandage} layout of the XXX.
+        \textbf{(h)} Visualization of \textit{odgi untangle} output: XXX.
     }

### 2021-10-28 manuscript/fig_odgi_viz.tex
added C4 bandage stuff, highlighting the SNPs and the corresponding encoded proteins

@@
                 \centering
-                \includegraphics[width=0.65\linewidth, trim=0 +2cm 0 0.5cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_layout}
+                \includegraphics[width=0.65\linewidth, trim=0 +2cm 0 0.5cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_bandage}
                 \caption{}
@@
                 \centering
-                \includegraphics[width=0.65\linewidth, trim=0 +2cm 0 0.5cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_layout}
+                \includegraphics[width=0.65\linewidth, trim=0 +2cm 0 0.5cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_bandage_zoom_in}
                 \caption{}
@@
 \textbf{(a)} \textit{odgi draw} layout of the MHC pangenome graph extracted from a whole human pangenome graph of 90 haplotypes.
-The red box highlights the C4 region.
+The red rectangle highlights the C4 region.
         \textbf{(b-e)} \textit{odgi viz} visualizations of the C4 pangenome graph, where 8 paths are displayed: 2 reference genomes (chm13 and grch38 on the top) and 6 haplotypes of 3 individuals.
@@
         The longer links at the bottom indicate the presence of a structural variant (long link) with another structural variant nested inside it (short link on the left).
-        Indeed, human C4 exists as 2 functionally distinct genes, C4A and C4B, which both vary in structure and copy number \citep{Sekar_2016}.
+        Indeed, human C4 exists as 2 functionally distinct genes, \textit{C4A} and \textit{C4B}, which both vary in structure and copy number \citep{Sekar_2016}.
 The longer link indicates that the copy number status varies across the haplotypes represented in the pangenome.
-        Moreover, C4A and C4B genes segregate in both long and short genomic forms, distinguished by the presence or absence of a human endogenous retroviral (HERV) sequence, as also highlighted by the short nested link on the left.
+        Moreover, \textit{C4A} and \textit{C4B} genes segregate in both long and short genomic forms, distinguished by the presence or absence of a human endogenous retroviral (HERV) sequence, as also highlighted by the short nested link on the left.
         \textbf{(c)} Color by path position: the color gradients are smooth, highlighting that its node are well sorted in 1 dimension.
@@
         In the haplotype with the name starting with HG01952\#1, the HERV sequence is absent.
-        \textbf{(f)} \textit{Bandage} layout of the C4 pangenome graph. The red box highlighs XXX.
-        \textbf{(g)} \textit{Bandage} layout of the XXX.
+        \textbf{(f)} \textit{Bandage} layout of the C4 pangenome graph. Green nodes indicate the C4 genes. The red rectangle highlights the regions where \textit{C4A} and \textit{C4B} genes differ.
+        \textbf{(g)} \textit{Bandage} layout of the region where \textit{C4A} and \textit{C4B} genes differ due to single nucleotide variants leading to changes in the encoded protein sequences.
         \textbf{(h)} Visualization of \textit{odgi untangle} output: XXX.

### 2021-10-28 scripts/visualization.sh

@@ -45,9 +45,14 @@ odgi draw -i ${filename_chr6_gfa}.C4.sorted.og -c ${filename_chr6_gfa}.C4.sorted
 
 
 # Get C4-GTF
-zgrep 'gene_id "C4A"\|gene_id "C4B"' hg38.ncbiRefSeq.gtf.gz | awk '$1 == "chr6" && $3 == "transcript"' | sed 's/chr6/grch38#chr6/g' | head -n 2 > chr6.C4.gtf
-#grch38#chr6	ncbiRefSeq.2021-09-09	transcript	32014795	32035418	.	+	.	gene_id "C4B"; transcript_id "NM_001002029.4_4";  gene_name "C4B";
-#grch38#chr6	ncbiRefSeq.2021-09-09	transcript	31982057	32002681	.	+	.	gene_id "C4A"; transcript_id "NM_001252204.2_3";  gene_name "C4A";
+zgrep 'gene_id "C4A"\|gene_id "C4B"' hg38.ncbiRefSeq.gtf.gz | awk '$1 == "chr6" && $3 == "transcript"' | cut -f 1 -d';' | sed 's/gene_id //g' | sed 's/"//g' | sed 's/chr6/grch38#chr6/g' | head -n 2 > chr6.C4.gtf
+#grch38#chr6	ncbiRefSeq.2021-09-09	transcript	32014795	32035418	.	+	.	C4B
+#grch38#chr6	ncbiRefSeq.2021-09-09	transcript	31982057	32002681	.	+	.	C4A
+
 
 # Compute Bandage annotations
 odgi position -i ${filename_chr6_gfa}.C4.sorted.og -E chr6.C4.gtf -t 16 > ${filename_chr6_gfa}.C4.sorted.anno.csv
+
+# Convert to GFA for Bandage
+odgi view -i ${filename_chr6_gfa}.C4.sorted.og -g > ${filename_chr6_gfa}.C4.sorted.gfa
+

### 2021-10-29 manuscript/fig_odgi_viz.tex
explicit Bandage annotation by `odgi position`
odgi untangle visualization

@@
             \caption{}
-            \includegraphics[width=\linewidth, trim=0cm 0cm -10cm 0cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_og_mhc_og_O_og_lay_H1000w100_png_svg}
+            \begin{center}
+                \includegraphics[width=\linewidth, trim=0cm 0cm -2.5cm 0.9cm]{fig/fig_extract_viz_draw_position_untangle/chr6.C4.sorted.untangle.bed.subset.tsv}
+            \end{center}
         \end{subfigure} \\
@@
         In the haplotype with the name starting with HG01952\#1, the HERV sequence is absent.
-        \textbf{(f)} \textit{Bandage} layout of the C4 pangenome graph. Green nodes indicate the C4 genes. The red rectangle highlights the regions where \textit{C4A} and \textit{C4B} genes differ.
+        \textbf{(f)} \textit{Bandage} layout of the C4 pangenome graph, annotated by using \textit{odgi position}.
+Green nodes indicate the C4 genes.
+The red rectangle highlights the regions where \textit{C4A} and \textit{C4B} genes differ.
         \textbf{(g)} \textit{Bandage} layout of the region where \textit{C4A} and \textit{C4B} genes differ due to single nucleotide variants leading to changes in the encoded protein sequences.
+Node labels were annoted by using \textit{odgi position}.
         \textbf{(h)} Visualization of \textit{odgi untangle} output: XXX.

### 2021-10-29 scripts/fig_extract_viz_draw_position_untangle.sh

@@ -56,3 +56,52 @@ odgi position -i ${filename_chr6_gfa}.C4.sorted.og -E chr6.C4.gtf -t 16 > ${file
 # Convert to GFA for Bandage
 odgi view -i ${filename_chr6_gfa}.C4.sorted.og -g > ${filename_chr6_gfa}.C4.sorted.gfa
 
+
+# Untangle
+
+f=${filename_chr6_gfa}.C4.sorted.og
+( echo query.name query.start query.end ref.name ref.start ref.end score inv self.cov x|
+  tr ' ' '\t'; odgi untangle -i $f -r $(odgi paths -i $f -L | grep chm13) -t 16 -m 256 |
+  bedtools sort -i - ) | awk '$8 == "-" { x=$6; $6=$5; $5=x; } { print }' |
+  tr ' ' '\t'   >$f.untangle.bed.tsv
+
+# cat $f.untangle.bed.tsv | grep '^chm13\|^grch38\|^HG00438\|^HG0107\|^HG01952'
+
+# R
+library(ggplot2)
+x <- read.table('/home/guarracino/Desktop/pangenomics/odgi_pub/chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.C4.sorted.og.untangle.bed.tsv', sep = '\t', header = T, comment.char="$")
+x$query.name <- gsub(":.*","",x$query.name)
+
+ggplot(
+  subset(x, query.name %in% c(
+    ##"chm13#chr6:31825251-31908851",
+    #"grch38#chr6:31972046-32055647",
+    ##"HG00438#1#JAHBCB010000040.1:24269348-24320210",
+    #"HG00438#2#JAHBCA010000042.1:24398231-24449090",
+    ##"HG01071#1#JAHBCF010000017.1:706180-783405",
+    #"HG01071#2#JAHBCE010000076.1:7794179-7897781",
+    #"HG01952#1#JAHAME010000044.1:28380191-28451052",
+    #"HG01952#2#JAHAMD010000016.1:31974838-32052065",
+
+    #"chm13#chr6",
+    "grch38#chr6",
+    #"HG00438#1#JAHBCB010000040.1",
+    "HG00438#2#JAHBCA010000042.1",
+    #"HG01071#1#JAHBCF010000017.1",
+    "HG01071#2#JAHBCE010000076.1",
+    "HG01952#1#JAHAME010000044.1",
+    "HG01952#2#JAHAMD010000016.1"
+    )
[…]

### 2021-10-29 manuscript/fig_metrics.tex
squeezed figure 3 and 4

@@
 		% include second image
-		\includegraphics[width=1.0\linewidth, trim=-1.5cm 0.75cm -0.8cm 0.5cm]{fig/metrics/chr4_pan_fa_a2fb268_e820cd3_9ea71d8_smooth_gfa_og_HTTex1_og_O_og_tiny_og_png_svg}
+		\includegraphics[width=1.0\linewidth, trim=-1.5cm 2cm -0.1cm 0.5cm]{fig/metrics/chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.tiny.og}
 		\label{fig:metrics-viz}

### 2021-10-29 scripts/metrics.sh

@@ -46,7 +46,7 @@ samtools faidx chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.
 cat chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.tiny.og.fasta.fai | sort -k2 -r | head -n 24 | grep "chm13" -v | cut -f 1 > paths_to_viz
 # take care of chm13 and grch13
 echo -e "chm13\ngrch38\n$(cat paths_to_viz)" > paths_to_viz
-odgi viz -i chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.tiny.og -o chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.tiny.og.png -a 1  -p paths_to_viz
+odgi viz -i chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.tiny.og -o chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.tiny.og.png -a 3  -p paths_to_viz
 
 #### MANUAL INTERVENTION ####
 # Go to https://drububu.com/tutorial/bitmap-to-vector.html. 

### 2021-10-29 manuscript/fig_metrics.tex
trimming stuff

@@
 		% include second image
-		\includegraphics[width=1.0\linewidth, trim=-1.5cm 2cm -0.1cm 0.5cm]{fig/metrics/chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.tiny.og}
+		\includegraphics[width=1.0\linewidth, trim=-1.95cm 2cm -1.0cm 0.5cm]{fig/metrics/chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.tiny.og}
 		\label{fig:metrics-viz}

### 2021-10-29 scripts/fig_extract_viz_draw_position_untangle.sh
change reference in the untangling

@@ -61,7 +61,7 @@ odgi view -i ${filename_chr6_gfa}.C4.sorted.og -g > ${filename_chr6_gfa}.C4.sort
 
 f=${filename_chr6_gfa}.C4.sorted.og
 ( echo query.name query.start query.end ref.name ref.start ref.end score inv self.cov x|
-  tr ' ' '\t'; odgi untangle -i $f -r $(odgi paths -i $f -L | grep chm13) -t 16 -m 256 |
+  tr ' ' '\t'; odgi untangle -i $f -r $(odgi paths -i $f -L | grep grch38) -t 16 -m 256 |
   bedtools sort -i - ) | awk '$8 == "-" { x=$6; $6=$5; $5=x; } { print }' |
   tr ' ' '\t'   >$f.untangle.bed.tsv
 

### 2021-10-29 manuscript/fig_odgi_viz.tex
squeezed figure 3 and 4
explain odgi untangle output visualization

@@
             \begin{center}
-                \includegraphics[width=\linewidth, trim=0cm 0cm -2.5cm 0.9cm]{fig/fig_extract_viz_draw_position_untangle/chr6.C4.sorted.untangle.bed.subset.tsv}
+                \includegraphics[width=\linewidth, trim=0cm 0.9cm -2.5cm 0.9cm]{fig/fig_extract_viz_draw_position_untangle/chr6.C4.sorted.untangle.bed.subset.tsv}
             \end{center}
@@
 Node labels were annoted by using \textit{odgi position}.
-        \textbf{(h)} Visualization of \textit{odgi untangle} output: XXX.
+        \textbf{(h)} Visualization of \textit{odgi untangle} output in the C4 pangenome graph: the plots show the copy number status of the sequences in the C4 region with respect to the grch38 reference sequence, making clear, for example, that in the haplotype with name starting with HG00438\#2, \textit{C4B} gene is missing.
     }

### 2021-10-29 manuscript/main.tex
odgi build performance figure
update fig letter in odgi viz/draw section

@@
 Finally, we create paths in serial, and extend them in parallel by obtaining mutex $Node.lock$ for pairs of nodes and by adding the path step in their $Node.path\_steps$ vectors.
-This parallelism speeds ODGI model construction by many-fold when testing against graphs built by the HPRC (TODO).
+This parallelism speeds ODGI model construction by many-fold when testing against graphs built by the HPRC (Figure \ref{fig:build}).
 
@@
 The meaning of the colors depends on how \textit{odgi viz} is executed.
-By default, path colors has no particular meaning, but each path has its own color (Fig. \ref{fig:odgi_viz}\textbf{a}).
+By default, path colors has no particular meaning, but each path has its own color (Fig. \ref{fig:odgi_viz}\textbf{b}).
 Path names are displayed on the left of the paths.
@@
 To overcome such a limitation, \textit{odgi viz} supports multiple visualization modalities (Fig.
-\ref{fig:odgi_viz}\textbf{b-d}), making it easy to grasp the properties of the graph and its complexity.
+\ref{fig:odgi_viz}\textbf{c-e}), making it easy to grasp the properties of the graph and its complexity.
 Graph nodes order can affects downstream analyses on pangenome graphs.
 With \textit{odgi viz} we can color the paths by path position (Fig.
-\ref{fig:odgi_viz}\textbf{b}), with light grey indicating where paths begin and dark grey where they ends.
+\ref{fig:odgi_viz}\textbf{c}), with light grey indicating where paths begin and dark grey where they ends.
 This visualization is suitable for checking graph nodes' linear order, as smooth color gradients indicate that the graph reflects well the linear paths' coordinate system.
@@
 \textit{odgi viz} supports also coloring the paths by orientation, with paths colored in red and black where their sequence is reversed or in forward, respectively, with respect to the sequences represented by the graph nodes (Fig.
-\ref{fig:odgi_viz}\textbf{b}).
+\ref{fig:odgi_viz}\textbf{d}).
 Eukaryotic genomes experience gains and losses of genetic material, resulting in copy number variation (CNV) across the population.
 With \textit{odgi viz}, we can use multiple color palettes to color the paths by path depth, highlighting the different copy number statuses in the genomes represented in the pangenome graph (Fig.
-\ref{fig:odgi_viz}\textbf{d}).
+\ref{fig:odgi_viz}\textbf{e}).
 
-\textit{odgi draw} extends the visualization in 2 dimensions (2D) by rendering the layout built by \textit{odgi layout} (\S\ref{sec:sort}).
+\textit{odgi draw} extends the visualization in 2 dimensions (2D)  (Fig.
+\ref{fig:odgi_viz}\textbf{a}) by rendering the layout built by \textit{odgi layout} (\S\ref{sec:sort}).
 A 2D rendering is more costly to compute, but we similarly provide a linear-time scaling implementation, allowing us to apply it to large pangenome graphs.
@@
 
+\input{fig_build_time}
+
 \input{fig_odgi_viz}

### 2021-10-29 manuscript/main.tex
specify where we use odgi position

@@
 In addition to these projections, we use the paths in the graph to provide a universal coordinate space.
-It also supports several kinds of ``lift-over'' of coordinates between different genomes.
+It also supports several kinds of ``lift-over'' of coordinates between different genomes (Fig. \ref{fig:odgi_viz}\textbf{f-g}).
 These allow for the direct translation of single coordinates and ranges, in an interactive way.
@@
 
-Figure showing the untangling of C4 ?
+Figure showing the untangling of C4 (Fig. \ref{fig:odgi_viz}\textbf{f-g})
 Figure showing tips over a single human chromosome from HPRC.

### 2021-10-29 manuscript/main.tex
typo
draft odgi untangle

@@
 
-Figure showing the untangling of C4 (Fig. \ref{fig:odgi_viz}\textbf{f-g})
+To obtain a more precise overview of collapsed \textit{loci}, we can apply \textit{odgi untangle} to segment paths into linear segments by breaking these segments where the paths loop back on themselves.
+In this way, we can obtain base-level information on the copy number status of the sequences in the locus.
+
+Figure showing the untangling of C4 (Fig. \ref{fig:odgi_viz}\textbf{h})
 Figure showing tips over a single human chromosome from HPRC.

### 2021-10-30 manuscript/main.tex
italianization: abstract

@@
 \author[Guarracino, Heumos \textit{et~al}.]{
-Andrea Guarracino\,$^{\text{\sfb 1} \dagger}$,
-Simon Heumos\,$^{\text{\sfb 2} \dagger}$,
-Sven Nahnsen\,$^{\text{\sfb 2}}$,
+Andrea~Guarracino\,$^{\text{\sfb 1} \dagger}$,
+Simon~Heumos\,$^{\text{\sfb 2} \dagger}$,
+Sven~Nahnsen\,$^{\text{\sfb 2}}$,
 Pjotr~Prins\,$^{\text{\sfb 3}}$,
@@
 \abstract{
-\textbf{Motivation:} Pangenome graphs provide a complete representation of the mutual alignment of collections of genomes.
-These models offer the opportunity to study the total genomic diversity of a population, including structurally complex regions.
-However, analyzing hundreds of gigabase-scale genomes using pangenome graphs is difficult and not well-supported by existing tools.
-%In addition, vertebrate genomes present highly repetitive regions which increase the complexity of the analysis.
-Hence, fast and versatile software is required to ask intricate questions of such data in an efficient way. \\
-\textbf{Results:} We wrote ODGI, a new suite of tools that implements scalable algorithms and has an efficient in-memory representation of DNA variation graphs.
-ODGI includes tools for detecting complex regions, extracting loci, removing artifacts, exploratory analysis, manipulation, validation, and visualization.
-Its fast parallel execution facilitates routine pangenomic tasks as well as pipelines that can quickly answer complex biological questions of gigabase-scale pangenome graphs.
-\\
-%ODGI supports graphs in the Graphical Fragment Assembly format version 1 (GFAv1).
-%The tools scale efficiently to large collections of eukaryotic genomes, allowing users to work easily with pangenome graphs of tens to thousands of large genomes while maintaining fast execution runtime and low memory overhead.
-\textbf{Availability:} ODGI is published as free software under the MIT open source license.
+\textbf{Motivation:}
+Pangenome graphs provide a complete representation of the mutual alignment of collections of genomes.
+These models offer the opportunity to study the entire genomic diversity of a population, including structurally complex regions.
+Nevertheless, analyzing hundreds of gigabase-scale genomes using pangenome graphs is difficult as it is not well-supported by existing tools.
+Hence, fast and versatile software is required to ask advanced questions to such data in an efficient way. \\
+\textbf{Results:}
+We wrote ODGI, a novel suite of tools that implements scalable algorithms and has an efficient in-memory representation of DNA variation graphs.
+ODGI includes tools for detecting complex regions, extracting \textit{loci}, removing artifacts, exploratory analysis, manipulation, validation, and visualization.
+Its fast parallel execution facilitates routine pangenomic tasks as well as pipelines that can quickly answer complex biological questions of gigabase-scale pangenome graphs. \\
+\textbf{Availability:}
+ODGI is published as free software under the MIT open source license.
 Source code can be downloaded from \url{https://github.com/pangenome/odgi} and documentation is available at \url{https://odgi.readthedocs.io}.
@@
 
-A pangenome models the full set of genomic elements in a given species or clade \citep{cpang2018,Eizenga_2020}.
+A pangenome models the full set of genomic elements in a given species or clade~\citep{cpang2018,Eizenga_2020}.
 In contrast to reference-based approaches which relate sequences to a single genome, these data structures encode the mutual relationships between all the genomes represented.
-In pangenome graphs \citep{Paten:2017}, homologous regions between genomes are compressed into a single representative model of all alleles present in the pangenome.
+In pangenome graphs~\citep{Paten:2017}, homologous regions between genomes are compressed into a single representative model of all alleles present in the pangenome.
 These flexible models let us encode any kind of variation, allowing the generation of comprehensive data systems on which to base analyses of genome evolution.
-Although these structures are of clear utility to researchers \citep{cpang2018}, and have been the focus of numerous applications in read mapping \citep{Garrison:2018,Baaijens_2019,Hickey:2020,Sibbesen_2021}, we observe that the community lacks a toolset specifically focused on graph manipulation and interrogation.
+Although these structures are of clear utility to researchers~\citep{cpang2018}, and have been the focus of numerous applications in read mapping~\citep{Garrison:2018,Baaijens_2019,Hickey:2020,Sibbesen_2021}, we observe that the community lacks a toolset specifically focused on graph manipulation and interrogation.
 
-The Human Pangenome Reference Consortium (HPRC) and Telomere-to-Telomere (T2T) consortium \citep{Miga:2020, Logsdon_2021, Nurk_2021} have recently demonstrated that high-quality \textit{de novo} assemblies can be routinely generated from third-generation long read sequencing data.
+The Human Pangenome Reference Consortium (HPRC) and Telomere-to-Telomere (T2T) consortium~\citep{Miga:2020, Logsdon_2021, Nurk_2021} have recently demonstrated that high-quality \textit{de novo} assemblies can be routinely generated from third-generation long read sequencing data.
 We anticipate that assemblies of similar quality will become common, leading to demand for methods that allow us to create and explore pangenomes.
@@
 Here, we present the Optimized Dynamic Genome/Graph Implementation (ODGI) toolkit, a pangenome graph interrogation and transformation system specifically built to handle the data scales encountered when working with pangenomes built from hundreds of haplotype-resolved genomes.
-ODGI provides a set of standard operations on the pangenome variation graph data model, generalizing ``genome arithmetic'' concepts like those found in BEDTools \citep{Quinlan_2010} to work on genome graphs, and providing new operations, such as visualizations, sorting, and liftover projections critical to understanding and exploiting pangenome graphs.
+ODGI provides a set of standard operations on the pangenome variation graph data model, generalizing ``genome arithmetic'' concepts like those found in BEDTools~\citep{Quinlan_2010} to work on genome graphs, and providing new operations, such as visualizations, sorting, and liftover projections critical to understanding and exploiting pangenome graphs.
 
@@
 
-\textit{odgi view} can convert a graph in ODGI binary format to the standard GFAv1 \citep{GFA} format. It can reveal a graph’s internal structures for e.g. debugging processes.
+\textit{odgi view} can convert a graph in ODGI binary format to the standard GFAv1~\citep{GFA} format. It can reveal a graph’s internal structures for e.g. debugging processes.
 

### 2021-10-30 manuscript/main.tex
italianization: introduction; avoid saying "genome graphs", we are alwasy pan

@@
 \section{Introduction}
-
 A pangenome models the full set of genomic elements in a given species or clade~\citep{cpang2018,Eizenga_2020}.
 In contrast to reference-based approaches which relate sequences to a single genome, these data structures encode the mutual relationships between all the genomes represented.
-In pangenome graphs~\citep{Paten:2017}, homologous regions between genomes are compressed into a single representative model of all alleles present in the pangenome.
+In pangenome graphs~\citep{Paten:2017}, homologous regions between genomes are compressed into a single representative of all alleles present in the pangenome.
 These flexible models let us encode any kind of variation, allowing the generation of comprehensive data systems on which to base analyses of genome evolution.
-Although these structures are of clear utility to researchers~\citep{cpang2018}, and have been the focus of numerous applications in read mapping~\citep{Garrison:2018,Baaijens_2019,Hickey:2020,Sibbesen_2021}, we observe that the community lacks a toolset specifically focused on graph manipulation and interrogation.
+Although these data structures are of clear utility to researchers~\citep{cpang2018}, and have been the focus of numerous applications in read mapping~\citep{Garrison:2018,Baaijens_2019,Hickey:2020,Sibbesen_2021}, the scientific community still lacks a toolset specifically focused on graph manipulation and interrogation.
 
 The Human Pangenome Reference Consortium (HPRC) and Telomere-to-Telomere (T2T) consortium~\citep{Miga:2020, Logsdon_2021, Nurk_2021} have recently demonstrated that high-quality \textit{de novo} assemblies can be routinely generated from third-generation long read sequencing data.
-We anticipate that assemblies of similar quality will become common, leading to demand for methods that allow us to create and explore pangenomes.
+We anticipate that \textit{de novo} assemblies of similar quality will become common, leading to demand for methods that allow us to create and explore pangenomes.
 Pangenome graphs are a natural model to support these kinds of analyses, but studies based on pangenome graphs are difficult due to a lack of downstream tools and standard workflows.
 
-Here, we present the Optimized Dynamic Genome/Graph Implementation (ODGI) toolkit, a pangenome graph interrogation and transformation system specifically built to handle the data scales encountered when working with pangenomes built from hundreds of haplotype-resolved genomes.
-ODGI provides a set of standard operations on the pangenome variation graph data model, generalizing ``genome arithmetic'' concepts like those found in BEDTools~\citep{Quinlan_2010} to work on genome graphs, and providing new operations, such as visualizations, sorting, and liftover projections critical to understanding and exploiting pangenome graphs.
+Here, we present the Optimized Dynamic Genome/Graph Implementation (ODGI) toolkit, a pangenome graph interrogation and transformation system specifically implemented to handle the data scales encountered when working with pangenomes built from hundreds of haplotype-resolved genomes.
+ODGI provides a set of standard operations on the variation graph data model, generalizing ``genome arithmetic'' concepts like those found in BEDTools~\citep{Quinlan_2010} to work on pangenome graphs, and providing a variety of operations, such as visualization, sorting, and liftover projections, all critical to understand and exploit pangenome graphs.
 
@@
 %This common API lets us reuse and and extend algorithms shared with the VG toolkit (VG) \citep{Garrison:2018}.
-%We specifically develop new methodsfocused on problems encountered when building pangenome graphs at the scale of vertebrate populations.
+%We specifically develop new methods focused on problems encountered when building pangenome graphs at the scale of vertebrate populations.
 %To ease interactive use, the majority of the ODGI tools are implemented in an index-free manner, avoiding the need to create index structures at each step of complex graph processing pipelines.
@@
 By avoiding global locks, we implement many of the operations in ODGI using maximum parallelism available.
-This ``atomic'' approach to edit operations on the graph is key to enable our methods to scale to the largest genome graphs that we can currently build.
+This ``atomic'' approach to edit operations on the graph is key to enable our methods to scale to the largest pangenome graphs that we can currently build.
 

### 2021-10-30 manuscript/main.tex
italianization: model; typos, little changes

@@
 \section{Model}
-
-A pangenome graph is a sequence model that encodes the mutual alignment of many genomes \citep{Garrison_2019_thesis,Eizenga_2020}.
+A pangenome graph is a sequence model that encodes the mutual alignment of many genomes~\citep{Garrison_2019_thesis,Eizenga_2020}.
 In the variation graph, $V = (N, E, P)$, nodes $N = n_1\ldots n_{|N|}$ contain sequences of DNA.
-Each node $n_i$ has an identifier $i$ and an implicit reverse complement $\bar{n_i}$, and a node strand $s$ corresponds to one such orientation.
+Each node $n_i$ has an identifier $i$ and an implicit reverse complement $\bar{n_i}$, and a node strand $s$ corresponds to one such orientations.
 Edges $E = e_1\ldots e_{|E|}$ represent ordered pairs of node strands: $e_i = ( s_a, s_b )$.
@@
 %The utility of the variation graph model lies in its lossless representation of genomes and their alignment.
-By containing both the sequences and information about their relative variations, the variation graph provides a powerful foundation for many bioinformatic applications.
+By containing both the sequences and information about their relative variations, the variation graph provides a complete and powerful foundation for many bioinformatic applications.
 
-Such a graph can be constructed by multiple sequence alignment \citep{Lee_2002,Grasso_2004}, or by transitively reducing an alignment between sequences to an equivalent, labeled sequence graph \citep{Kehr_2014,Garrison_2019_thesis}.
-Current methods to build these graphs are still in development \citep{Li:2020,Armstrong:2020,pggb}, but they have largely settled on a common data model, represented in the Graphical Fragment Assembly (GFA) format \citep{GFA}).
-%\FIXME{may want to move GFA standard+VG to intro section 1}
-%The improvement of these methods depends on tools that allow us to investigate
+Pangenome graphs can be constructed by multiple sequence alignment~\citep{Lee_2002,Grasso_2004} or by transitively reducing an alignment between sequences to an equivalent, labeled sequence graph~\citep{Kehr_2014,Garrison_2019_thesis}.
+Current methods to build these graphs are still under active development~\citep{Li:2020,Armstrong:2020,pggb}, but they have largely settled on a common data model, represented in the Graphical Fragment Assembly (GFA) format \citep{GFA}).
+%\FIXME{may want to move GFA standard+VG to intro section 1}. AG: it is borderline, but it seems fine to have it here too.
 This standardization supports the development of a reference set of tools that operate on the pangenome graph model.
-Such an effort began with the VG toolkit toolkit~\citep{Garrison:2018}.
-Here we refocus it with with ODGI---a compatible, but independent set of algorithms focused on first-order interrogation and transformation of pangenome graphs.
-
-%Because ODGI uses this standard to  can be mixed with other pangenome tools, such as the VG toolkit
-%Here, we enhance ongoing efforts with the VG
-%The need to understand the graphs made by these methods motivates it.
-%We require such a toolkit to understand the graphs made by these methods.
-%\FIXME{last sentence not really clear}
-
-%The improvement of these approaches depends on scalable tools for understanding pangenome graphs, a need that our work responds to.
-
-%As a toolkit for understanding these graphs, ODGI
+Such an effort began with the VG toolkit~\citep{Garrison:2018}.
+Here we refocus it with ODGI, a compatible, but independent set of algorithms focused on visualization, interrogation, and transformation of pangenome graphs.
 
@@
 %By projecting input sequences through this transformation, we obtain paths that losslessly encode the original sequences, but in the space of the graph \citep{Garrison_2019_thesis}.
-%Methods to build such graphs are in active development, and here we focus on graphs built with the PanGenome Graph Builder pipeline \citep{pggb}, a process that converts an input pangenome (in FASTA) into a combined graphical model (in the Graphical Fragment Assembly (GFA) format \citep{GFA}).
 
 \section{Implementation}
-
 The ODGI toolkit builds on existing approaches to efficiently store and manipulate variation graphs \citep{Garrison:2018}.

### 2021-10-30 manuscript/main.tex
italianization: implementation; explicit who is depth, use model only for the pangenomes, typos, little changes

@@
 Pangenome graphs can be constructed by multiple sequence alignment~\citep{Lee_2002,Grasso_2004} or by transitively reducing an alignment between sequences to an equivalent, labeled sequence graph~\citep{Kehr_2014,Garrison_2019_thesis}.
-Current methods to build these graphs are still under active development~\citep{Li:2020,Armstrong:2020,pggb}, but they have largely settled on a common data model, represented in the Graphical Fragment Assembly (GFA) format \citep{GFA}).
+Current methods to build these graphs are still under active development~\citep{Li:2020,Armstrong:2020,pggb}, but they have largely settled on a common data model, represented in the Graphical Fragment Assembly (GFA) format~\citep{GFA}).
 %\FIXME{may want to move GFA standard+VG to intro section 1}. AG: it is borderline, but it seems fine to have it here too.
@@
 \section{Implementation}
-The ODGI toolkit builds on existing approaches to efficiently store and manipulate variation graphs \citep{Garrison:2018}.
-Similar to other efficient libaries presenting the \textsc{HandleGraph} model \citep{Eizenga_2020_BX}, ODGI's implementation rests on three key properties which hold for most pangenome variation graphs:
+The ODGI toolkit builds on existing approaches to efficiently store and manipulate variation graphs~\citep{Garrison:2018}.
+Similar to other efficient libaries presenting the \textsc{HandleGraph} model~\citep{Eizenga_2020_BX}, the implementation of ODGI's tools rests on three key properties which hold for most pangenome variation graphs:
 
@@
 \item Their embedded paths are locally similar to each other.
-%We incorporate concepts first introduced in the dynamic version of the Graph BWT \citep{Siren:2020} (GBWT), a scalable implementation of the graph extension of the positional Burrows-Wheeler transform \citep{Durbin_2014}.
 \end{enumerate}
 
-These properties are used to build efficient dynamic variation graph data structures \citep{Siren:2020,Eizenga_2020_BX}.
+These properties are used to build efficient dynamic variation graph data structures~\citep{Siren:2020,Eizenga_2020_BX}.
 Sparsity (1) allows us to encode edges $E$ using adjacency lists rather than matrices or hash tables.
@@
 
-%Here, we build on this work and focus on issues that have arisen during our recent work on the Human Pangenome.
-ODGI improves on prior efforts, based on issues that arose during our work with high-quality assemblies that cover almost all parts of the genome \citep{Logsdon_2021,Nurk_2021}.
-%A key design objective, motivated by experience working with graphs built from high quality eukaryotic genome assemblies, is to support graphs with very high path depth.
-For example, we find that it is necessary to support graphs that have regions of very high path depth.
+ODGI improves on prior efforts, based on issues that arose during our work with high-quality \textit{de novo} assemblies that cover almost all parts of the human genome~\citep{Logsdon_2021,Nurk_2021}.
+For example, we find that it is necessary to support graphs that have regions with paths that present very high depth nodes.
 Such motifs can occur in collapsed repeat structures generated by ambiguous sequence homology relationships in repeats found in the centromeres and other segmental duplications.
@@
 However, neither of these solutions allows us to investigate their biological features.
-To seamlessy represent such difficult regions, we follow an approach implemented in the dynamic version of the Graph BWT (GBWT) \citep{Siren:2020}, and built a node-centric, dynamic, compressed model of the paths.
-This design supports node-local modification and update of the graph, which lets us build and transform the model in parallel.
+To seamlessly represent such difficult regions, we follow an approach implemented in the dynamic version of the Graph BWT (GBWT)~\citep{Siren:2020} and built a node-centric, dynamic, compressed model of the paths.
+This design supports node-local modification and update of the graph, which lets us operate on paths in parallel.
 
@@
 For a given node $i$ and neighbor $j$, the decoder itself does not store the $id$ of $j$, but rather a compact representation of the relative difference between the node ids: $\delta = Node_i.id - Node_j.id$.
-This keeps even the size of the encoding small, per common variation graph property (2).
+%The $\delta between two nodes is computed as their distance in the graph vector (i.e., the difference between the node offsets).
+This keeps the size of the encoding small, per common variation graph property (2).
 We define the edges and path steps traversing the node in terms of this alphabet of $\delta$'s.
-The structures in Algorithm \ref{alg:structs} describes our encoding.
+The structures in Algorithm~\ref{alg:structs} describes our encoding.
 
@@
 Each structure contains the sequence of the node ($Node_i.sequence$), its edges in both directions ($Node_i.edges$), and a vector of path steps that describes the previous and next steps in paths that walk across the node ($Node_i.path\_steps$).
-%As in the GBWT, to compress the encoding, we encode path steps using a local alphabet that maps the N neighbours of the node into the range $1\ldots N$.
-%To further save space, node deltas, rather than IDs, are stored in this alphabet.
-%The delta between two nodes is defined as their distance in the graph vector (i.e., the difference between the node offsets).
 For efficiency, $Node_i.sequence$ is stored as a plain string, while the $edges$ and $path\_steps$ are stored using a dynamic succinct integer vector that requires $O(2nw)$ bits for the edges and $O(5nw)$ bits for the path steps, where $n$ is the number of steps on the node, and $w$ is $\approx log_2(n)$ \citep{prezza2017framework}.
@@
 By avoiding global locks, we implement many of the operations in ODGI using maximum parallelism available.
-This ``atomic'' approach to edit operations on the graph is key to enable our methods to scale to the largest pangenome graphs that we can currently build.
+This approach is key to enable our methods to scale to the largest pangenome graphs that we can currently build.
 

### 2021-10-30 manuscript/main.tex
italianization: implementation; explicit who is depth, use model only for the pangenomes, typos, little changes
italianization: resultsn; avoid acronyms not introduced, simplify, typos

@@
 ODGI improves on prior efforts, based on issues that arose during our work with high-quality \textit{de novo} assemblies that cover almost all parts of the human genome~\citep{Logsdon_2021,Nurk_2021}.
-For example, we find that it is necessary to support graphs that have regions with paths that present very high depth nodes.
+For example, we find that it is necessary to support graphs that have regions where paths present very high depth nodes.
 Such motifs can occur in collapsed repeat structures generated by ambiguous sequence homology relationships in repeats found in the centromeres and other segmental duplications.
@@
 
-\begin{comment}
 %\subsection{Core functionality}
-
-In the variation graph model, paths have to respect the graph’s topology: this can be verified with odgi validate, to ensure no errors in the input or edited graphs.
-In variation graphs the coordinates are provided by the embedded path sequences.
-Indeed, the node IDs are not meant to be stable.
+%In the variation graph model, paths have to respect the graph’s topology: this can be verified with odgi validate, to ensure no errors in the input or edited graphs.
+%In variation graphs the coordinates are provided by the embedded path sequences.
+%Indeed, the node IDs are not meant to be stable.
 odgi position finds, translates, and liftovers graph and path positions between different graphs by exploiting their shared path sequences (Figure 1.B).
-\end{comment}
 
@@
 
-ODGI provides a set of first-order interrogative and manipulative operations on pangenome graphs.
+ODGI provides a set of interrogative and manipulative operations on pangenome graphs.
 We have established these tools to support our exploration of graphs built from hundreds of large eukaryotic genomes.
-They are practical and able to work with high levels of graph complexity, even with regions of very high path depth (10$^5$ to 10$^6$-fold coverage).
-ODGI's tools cover common operations that we have found to be essential when working with these complex biological models:
+ODGI's tools are practical and able to work with high levels of graph complexity, even with regions where paths present very high depth nodes (10$^5$ to 10$^6$-fold depth).
+
+ODGI covers common operations that we have found to be essential when working with complex pangenome graphs:
 
@@
 \item \textit{odgi build} constructs the ODGI data model from GFA (\S\ref{sec:build}).
-\item \textit{odgi view} converts the graph into standard text formats (\S\ref{sec:text}).
+\item \textit{odgi view} converts the ODGI data model into GFA (\S\ref{sec:text}).
 \item \textit{odgi matrix} derives the pangenome matrix (\S\ref{sec:text}).
@@
 \item \textit{odgi flatten} converts the graph to FASTA and BED (\S\ref{sec:text}).
-\item \textit{odgi viz} provides a scalable linear visualization of the graph (\S\ref{sec:viz}).
+\item \textit{odgi viz} provides a linear visualization of the graph (\S\ref{sec:viz}).
 \item \textit{odgi draw} renders a 2D image of the graph (\S\ref{sec:viz}).
-\item \textit{odgi extract} subsets the graph based on path ranges (\S\ref{sec:extract}).
+\item \textit{odgi extract} subsets of the graph based on path ranges (\S\ref{sec:extract}).
 \item \textit{odgi chop} breaks long nodes into shorter ones (\S\ref{sec:edit}).
@@
 \item \textit{odgi prune} removes complex regions (\S\ref{sec:edit}).
-\item \textit{odgi position} lift coordinates between paths and graph positions (\S\ref{sec:untangle}).
+\item \textit{odgi position} lifts coordinates between paths and graph positions (\S\ref{sec:untangle}).
 \item \textit{odgi untangle} deconvolutes paths relative to a reference (\S\ref{sec:untangle}).
 \item \textit{odgi tips} finds path end points relative to a reference (\S\ref{sec:untangle}).
-\item \textit{odgi sort} orders the graph with sorting pipelines (\S\ref{sec:sort}).
-\item \textit{odgi layout} establishes a 2D layout using SGD (\S\ref{sec:sort}).
+\item \textit{odgi sort} orders the graph nodes (\S\ref{sec:sort}).
+\item \textit{odgi layout} establishes a 2D layout (\S\ref{sec:sort}).
 \item \textit{odgi stats} provides numerical properties of the graph (\S\ref{sec:metrics}).
 \item \textit{odgi bin} generates a summarized view of the graph (\S\ref{sec:metrics}).
-\item \textit{odgi depth} describes path depth over graph and path positions (\S\ref{sec:metrics}).
+\item \textit{odgi depth} describes node depth over graph and path positions (\S\ref{sec:metrics}).
 \item \textit{odgi degree} describes node degree over graph and path positions (\S\ref{sec:metrics}).
@@
 Most read or write the native ODGI format (`og' extension) (Figure \ref{fig:operations}) and work with standard text based data formats common to bioinformatics.
-This supports the implementation of flexible, composable, graph processing pipelines based on a common data type.
+This supports the implementation of flexible and composable graph processing pipelines based on a common data type.
 While the pangenome graph is a first-class entity in ODGI, we focus on the embedded paths to provide a universal coordinate system.
-By considering all paths in the graph as potential reference or query sequences, we make the graph invisible to downstream tools that operate on collections of sequences (e.g.
-BEDTools, SAMtools \citep{Li2009}, aligners).
+By considering all paths in the graph as potential reference or query sequence, we make the graph invisible to downstream tools that operate on collections of sequences (\textit{e.g.} BEDTools, SAMtools~\citep{Li2009}, aligners).
 This approach benefits from the information in the graph without strongly embedding our methods in this difficult new research context.

### 2021-10-30 manuscript/main.tex
ODGI is a graph-wrapper

@@
 While the pangenome graph is a first-class entity in ODGI, we focus on the embedded paths to provide a universal coordinate system.
-By considering all paths in the graph as potential reference or query sequence, we make the graph invisible to downstream tools that operate on collections of sequences (\textit{e.g.} BEDTools, SAMtools~\citep{Li2009}, aligners).
+Despite being recent systems, pangenome graphs are not incompatible with current standard ways of working with bioinformatics data.
+ODGI provides a set of interfaces that let us approach these graphs from the perspective of standard reference- and sequence-based data models.
+Indeed, by considering all paths in the graph as potential reference or query sequence, we make graphs invisible to downstream tools that operate on collections of sequences or rely on a reference sequence (\textit{e.g.} SAMtools~\citep{Li2009}), enabling interoperability.
 This approach benefits from the information in the graph without strongly embedding our methods in this difficult new research context.

### 2021-10-30 manuscript/main.tex
current subsection 4.2 is tricky

@@
 \begin{itemize}
-\item \textit{odgi build} constructs the ODGI data model from GFA (\S\ref{sec:build}).
-\item \textit{odgi view} converts the ODGI data model into GFA (\S\ref{sec:text}).
+\item \textit{odgi build} constructs the ODGI data model from GFA file (\S\ref{sec:build}).
+\item \textit{odgi view} converts the ODGI data model into GFA file (\S\ref{sec:build}).
 \item \textit{odgi matrix} derives the pangenome matrix (\S\ref{sec:text}).
@@
 Each tool focuses on a small set of related operations.
-Most read or write the native ODGI format (`og' extension) (Figure \ref{fig:operations}) and work with standard text based data formats common to bioinformatics.
+Most read or write the native ODGI format (`og' extension) (Figure~\ref{fig:operations}) and work with standard text based data formats common to bioinformatics.
 This supports the implementation of flexible and composable graph processing pipelines based on a common data type.
@@
 
-%We find that, to understanding relationships between sequences and positions in pangenome graphs, we must use context mapping to ``untangle'' complex regions found in VNTRs, segmental duplications, and centromeric repeats.
-%This approach lets use pangenome graphs as a foundation for universal liftover between any two sets of sequences in the graph.
-
 %Most of the tools are designed to be applied together, piping the output of one tool into the next, thereby preventing the creation of intermediate files, and reducing the number of IO operations.
@@
 \label{sec:build}
+ODGI maintains its own efficient ODGI binary format for storing graphs on disk.
+We begin by transforming the storage model of the standard GFAv1~\citep{GFA} format (in which nodes, edges, and paths are described independently) into the ODGI node-centric encoding with \textit{odgi build}.
+This construction step can be a significant bottleneck, in particular as the size of the path set of the graph increases.
 
-We begin by transforming the storage model of the GFA format (in which nodes, edges, and paths are described independently) into the ODGI node-centric encoding with \textit{odgi build}.
-This construction step can present a significant bottleneck, in particular as the size of the path set of the graph increases.
-The ODGI data structure (Algorithm \ref{alg:structs}) allows algorithms that build and modify the graph to operate in parallel, without any global locks.
-In \textit{odgi build}, we initially construct the node vector in a serial operation that scans across the input GFA.
-Then, in serial, we add edges in the $Node.edges$ vectors of pairs of nodes.
-Finally, we create paths in serial, and extend them in parallel by obtaining mutex $Node.lock$ for pairs of nodes and by adding the path step in their $Node.path\_steps$ vectors.
-This parallelism speeds ODGI model construction by many-fold when testing against graphs built by the HPRC (Figure \ref{fig:build}).
+The ODGI data structure (Algorithm~\ref{alg:structs}) allows algorithms that build and modify the graph to operate in parallel, without any global locks.
+In \textit{odgi build}, we initially construct the node vector in a serial operation that scans across the input GFA file.
+Then, we serially add edges in the $Node.edges$ vectors of pairs of nodes.
+Finally, we create paths in serial, and extend them in parallel by obtaining the mutex $Node.lock$ for pairs of nodes and by adding the path step in their $Node.path\_steps$ vectors.
+This parallelism speeds ODGI model construction by many-fold when testing against graphs built by the HPRC (Figure~\ref{fig:build}).
 
-%Figure: encoding model as a cartoon --  HTT exon 1 as tiny example graph (maybe)
+For interchange with other tools, \textit{odgi view} can convert a graph in ODGI binary format to the standard GFAv1~\citep{GFA}.
 
-\subsection{Converting to standard file formats}
+\subsection{Converting to standard file formats{\color{red} FIXME: We need a different title/subsection}
+}
 \label{sec:text}
-
-% pjotr & andrea
-ODGI maintains its own efficient binary format for storing graphs on disk.
-ODGI also supports reading and writing the textual GFA file format for interchange with other tools such as with the VG toolkit~\citep{Garrison:2018}.
-
-\textit{odgi view} can convert a graph in ODGI binary format to the standard GFAv1~\citep{GFA} format. It can reveal a graph’s internal structures for e.g. debugging processes.
-
-\textit{odgi stats} outputs pangenome statistics in TSV or YAML textual file formats.
-Among other metrics, it can calculate the number of nodes, edges, paths and the total nucleotide length of the graph.
-{\color{red} FIXME: All this information is also listed in the metrics section.
-How to split?
-No mentioning of file formats in the following sections?}
 
@@
 
+{\color{red} FIXME: We should put the following information in the metric section, of move some commands here}
+
 \textit{odgi bin} generates a summarized view of gigabase scale graphs providing tab delimited or JSON `bins'.
@@
 \textit{odgi server} implements a basic HTTP server which can translate a 'path:position' to a 'pangenome:position' very efficiently.
+{\color{red} FIXME: this command is listed in the list in 4 Results}
 
@@
 During this process, it is capable of untangling loopy region resulting in linearized pairs of regions in the BEDPE file. A self dotplot assists in debugging and understanding the untangle process.
+
+\textit{odgi stats} outputs pangenome statistics in TSV or YAML textual file formats.
+Among other metrics, it can calculate the number of nodes, edges, paths and the total nucleotide length of the graph.
 
@@
 
-% erik
+%We find that, to understanding relationships between sequences and positions in pangenome graphs, we must use context mapping to ``untangle'' complex regions found in VNTRs, segmental duplications, and centromeric repeats.
+%This approach lets use pangenome graphs as a foundation for universal liftover between any two sets of sequences in the graph.
 
@@
 \label{sec:sort}
-
-% simon
 
@@
 \section{Discussion}
-
-% erik & pjotr & anyone
 

### 2021-10-31 manuscript/main.tex
added commads; moved view, matrix, paths, flatten, explode, squeezed, server, commands; multiple changes in results, extraction, editing, and metrics sections

@@
 The ODGI toolkit builds on existing approaches to efficiently store and manipulate variation graphs~\citep{Garrison:2018}.
-Similar to other efficient libaries presenting the \textsc{HandleGraph} model~\citep{Eizenga_2020_BX}, the implementation of ODGI's tools rests on three key properties which hold for most pangenome variation graphs:
+Similar to other efficient libraries presenting the \textsc{HandleGraph} model~\citep{Eizenga_2020_BX}, the implementation of ODGI's tools rests on three key properties which hold for most pangenome variation graphs:
 
@@
 \item \textit{odgi view} converts the ODGI data model into GFA file (\S\ref{sec:build}).
-\item \textit{odgi matrix} derives the pangenome matrix (\S\ref{sec:text}).
-\item \textit{odgi paths} lists and extracts paths in FASTA (\S\ref{sec:text}).
-\item \textit{odgi flatten} converts the graph to FASTA and BED (\S\ref{sec:text}).
 \item \textit{odgi viz} provides a linear visualization of the graph (\S\ref{sec:viz}).
@@
 \item \textit{odgi extract} subsets of the graph based on path ranges (\S\ref{sec:extract}).
+\item \textit{odgi explode} breaks the graph into connected components (\S\ref{sec:extract}).
+\item \textit{odgi squeeze} unifies disjoint graphs (\S\ref{sec:extract}).
 \item \textit{odgi chop} breaks long nodes into shorter ones (\S\ref{sec:edit}).
 \item \textit{odgi unchop} combines redundant nodes (\S\ref{sec:edit}).
-\item \textit{odgi explode} breaks the graph into connected components (\S\ref{sec:edit}).
-\item \textit{odgi squeeze} unifies disjoint graphs (\S\ref{sec:edit}).
+\item \textit{odgi break} removes cycles in the graph (\S\ref{sec:edit}).
 \item \textit{odgi prune} removes complex regions (\S\ref{sec:edit}).
-\item \textit{odgi position} lifts coordinates between paths and graph positions (\S\ref{sec:untangle}).
+\item \textit{odgi groom} resolves spurious inverting links (\S\ref{sec:edit}).
+\item \textit{odgi position} lifts coordinates between path and graph positions (\S\ref{sec:untangle}).
+\item \textit{odgi server} provides a HTTP server to lift coordinates (\S\ref{sec:untangle}).
 \item \textit{odgi untangle} deconvolutes paths relative to a reference (\S\ref{sec:untangle}).
@@
 \item \textit{odgi layout} establishes a 2D layout (\S\ref{sec:sort}).
+\item \textit{odgi matrix} derives the pangenome matrix (\S\ref{sec:metrics}).
+\item \textit{odgi paths} lists and extracts paths in FASTA (\S\ref{sec:metrics}).
+\item \textit{odgi flatten} converts the graph to FASTA and BED (\S\ref{sec:metrics}).
 \item \textit{odgi stats} provides numerical properties of the graph (\S\ref{sec:metrics}).
@@
 
-\subsection{Converting to standard file formats{\color{red} FIXME: We need a different title/subsection}
-}
-\label{sec:text}
-
-\textit{odgi matrix} derives the pangenome matrix and writes the graph topology in sparse matrix formats.
-
-\textit{odgi paths} allows the investigation of paths of a given variation graph. It can calculate overlap statistics of groupings of paths and extracts FASTA.
-
-\textit{odgi flatten} projects the graph sequence and paths into FASTA and BED.
-
-{\color{red} FIXME: We should put the following information in the metric section, of move some commands here}
-
-\textit{odgi bin} generates a summarized view of gigabase scale graphs providing tab delimited or JSON `bins'.
-
-\textit{odgi position} translates graph and path positions between or within graphs emitting the liftovers in TSV format.
-
-\textit{odgi server} implements a basic HTTP server which can translate a 'path:position' to a 'pangenome:position' very efficiently.
-{\color{red} FIXME: this command is listed in the list in 4 Results}
-
-\textit{odgi degree/position/depth/untangle/tips} all can output the BED format.
-
-\textit{odgi untangle} outputs the Pairwise mApping Format (PAF). PAF is a text format describing the approximate mapping positions between two set of sequences.
-The odgi untangle command projects paths into a reference-relative BEDPE file, decomposing paralogy relationships.
-During this process, it is capable of untangling loopy region resulting in linearized pairs of regions in the BEDPE file. A self dotplot assists in debugging and understanding the untangle process.
-
-\textit{odgi stats} outputs pangenome statistics in TSV or YAML textual file formats.
-Among other metrics, it can calculate the number of nodes, edges, paths and the total nucleotide length of the graph.
-
 \subsection{Visualizing pangenome graphs}
@@
 \textit{odgi viz} supports a binned, linearized rendering in 1 dimension (that is, all graph nodes lie on the same axis).
-This visualization is computed in linear-time and offers a human-interpretable format suitable for understanding the topology and genome relationships in the pangenome graph (Fig. \ref{fig:odgi_viz}).
+This visualization is computed in linear-time and offers a human-interpretable format suitable for understanding the topology and genome relationships in the pangenome graph (Fig.~\ref{fig:odgi_viz}).
 Graph nodes are arranged on a single axis, from left to right, with the colored bar indicating the paths and the nodes they cross.
@@
 The meaning of the colors depends on how \textit{odgi viz} is executed.
-By default, path colors has no particular meaning, but each path has its own color (Fig. \ref{fig:odgi_viz}\textbf{b}).
+By default, path colors has no particular meaning, but each path has its own color (Fig.~\ref{fig:odgi_viz}\textbf{b}).
 Path names are displayed on the left of the paths.
@@
 Nevertheless, complex, nonlinear graph structures are difficult to display and interpret in a low number of dimensions.
-To overcome such a limitation, \textit{odgi viz} supports multiple visualization modalities (Fig.
-\ref{fig:odgi_viz}\textbf{c-e}), making it easy to grasp the properties of the graph and its complexity.
+To overcome such a limitation, \textit{odgi viz} supports multiple visualization modalities (Fig.~\ref{fig:odgi_viz}\textbf{c-e}), making it easy to grasp the properties of the graph and its complexity.
 Graph nodes order can affects downstream analyses on pangenome graphs.
-With \textit{odgi viz} we can color the paths by path position (Fig.
-\ref{fig:odgi_viz}\textbf{c}), with light grey indicating where paths begin and dark grey where they ends.
+With \textit{odgi viz} we can color the paths by path position (Fig.~\ref{fig:odgi_viz}\textbf{c}), with light grey indicating where paths begin and dark grey where they ends.
 This visualization is suitable for checking graph nodes' linear order, as smooth color gradients indicate that the graph reflects well the linear paths' coordinate system.
 Pangenome graphs represents both strand of DNA sequences.
-\textit{odgi viz} supports also coloring the paths by orientation, with paths colored in red and black where their sequence is reversed or in forward, respectively, with respect to the sequences represented by the graph nodes (Fig.
-\ref{fig:odgi_viz}\textbf{d}).
+\textit{odgi viz} supports also coloring the paths by orientation, with paths colored in red and black where their sequence is reversed or in forward, respectively, with respect to the sequences represented by the graph nodes (Fig.~\ref{fig:odgi_viz}\textbf{d}).
 Eukaryotic genomes experience gains and losses of genetic material, resulting in copy number variation (CNV) across the population.
-With \textit{odgi viz}, we can use multiple color palettes to color the paths by path depth, highlighting the different copy number statuses in the genomes represented in the pangenome graph (Fig.
-\ref{fig:odgi_viz}\textbf{e}).
+With \textit{odgi viz}, we can use multiple color palettes to color the paths by path depth, highlighting the different copy number statuses in the genomes represented in the pangenome graph (Fig.~\ref{fig:odgi_viz}\textbf{e}).
 
-\textit{odgi draw} extends the visualization in 2 dimensions (2D)  (Fig.
-\ref{fig:odgi_viz}\textbf{a}) by rendering the layout built by \textit{odgi layout} (\S\ref{sec:sort}).
+\textit{odgi draw} extends the visualization in 2 dimensions (2D)  (Fig.~\ref{fig:odgi_viz}\textbf{a}) by rendering the layout built by \textit{odgi layout} (\S\ref{sec:sort}).
 A 2D rendering is more costly to compute, but we similarly provide a linear-time scaling implementation, allowing us to apply it to large pangenome graphs.
 
-\subsection{Extracting regions of interest}
+\subsection{Extracting or joining regions of interest}
 \label{sec:extract}
 
-% andrea (?)
+Pangenome graphs built from hundreds of haplotype-resolved \textit{de novo} genome assemblies are very large, but it is often necessary to work with only a small portion of the genomes represented,
+such as a specific \textit{locus} (Fig.~\ref{fig:odgi_viz}\textbf{a}) or a smaller region (Fig.~\ref{fig:odgi_viz}\textbf{b-g}), or even a single gene (Fig.~\ref{fig:metrics}),
+thereby simplifying the downstream analyses and reducing the resources to work only with the extracted graphs.
+%Subsetting operations allow us to zoom in on particular genes or regions of interest for more precise examination.
+Graph portions can be extracted by using the paths in the graph as coordinate systems to guide the process.
+For such operation, \textit{odgi extract} allows users to extract specific regions of the graph as defined by query criteria.
+Regions of interest can be specified by graph nodes or path range(s), also in BED format. Furthermore, it is possible to indicate a list of paths to be preserved completely in the extracted graph.
 
-% extract
-Pangenome graphs are very large, but we often only want to work with a small portion (e.g. a single gene).
-We can extract such regions using coordinates on the paths in the graph to guide us.
+In \textit{odgi extract}, we begin by collecting all graph nodes that fall within the ranges to extract (and the paths to preserve, if requested).
+Users can specify the number of steps or nucleotides to expand the selection and include neighboring nodes.
+Then, edges connecting all selected nodes are added in the subgraph under construction.
+Finally, the portions of the paths (\textit{i.e.}, the subpaths) walking through the selected nodes are extracted and added to the new subgraph.
+Subpaths are search in parallel, created serially, and extended in parallel again thanks to the parallelism enabled by the ODGI data structure (see \S\ref{sec:build}), making \textit{odgi extract} a scalable solution to extract also complex subregions presenting nodes with very high depth.
 
-Figure showing extraction process---either schematic or ``real''.
+% If needed, we can have a figure showing extraction process---either schematic or ``real''.
 
-odgi extract allows extracting specific subregions of the graph as defined by query criteria, thereby simplifying the downstream analyses and reducing the resources to work only with the extracted region.
-
-%Subsetting operations allow us to zoom in on particular genes or regions of interest for more precise examination.
-
-%We extract key information about the graph using its paths as a pangenome reference system.
+Pangenome graphs can embed multiple chromosomes as separated connected components (inter-chromosomal structural variants would join the components into bigger ones).
+\textit{odgi explode} separates the connected components in different ODGI format files, while \textit{odgi squeeze} allows merging multiple graphs into the same ODGI format file, preventing node ID collisions.
 
@@
 
-% andrea (?)
+Pangenome graphs can be used in a variety of applications, ranging from read mapping to variant identification and genotyping ~\citep{Eizenga_2020}.
+However, graphs presenting complex topology can increase the computational overhead of many downstream analyses.
+ODGI offers multiple commonly-needed basic operations on the topology of the graph and its nodes.
 
-% chop / unchop / squeeze / break / prune / explode
+With \textit{odgi prune} users can remove complex parts as defined by query criteria, while with \textit{odgi break} they can remove cycles in the graph, reducing the complexity of the graph topology.
+Furthermore, \textit{odgi groom} allows removing spurious inverting links by exploring the graph from the orientation supported by most paths.
 
 To enable efficient sequence alignment against the graph, long nodes can be divided into shorter nodes at a maximum requested size using \textit{odgi chop}.
-Partial order alignment, which consists of aligning sequences against a directed acyclic graph (DAG), is frequently used in pangenome building pipelines, but the current implementations return DAGs with 1-bp long nodes; \textit{odgi unchop} allows joining nodes that can be merged without changing the graph topology, nor the embedded sequences, obtaining an equivalent, but more compact, representation of the graph.
+Partial order alignment, which consists of aligning sequences against a directed acyclic graph (DAG), is frequently used in pangenome building pipelines~\citep{pggb}, but the current implementations return DAGs with 1-bp long nodes.
+\textit{odgi unchop} allows joining nodes that can be merged without changing the graph topology, nor the embedded sequences, obtaining an equivalent, but more compact, representation of the graph.
 
-Pangenome graphs can embed multiple chromosomes as separated connected components (inter-chromosomal structural variants would join the components into bigger ones).
-\textit{odgi explode} separates the connected components in different ODGI format files, while \textit{odgi squeeze} allows merging multiple graphs into the same ODGI format file, preventing node ID collisions.
-
-Cycles in the graph complicate downstream analyses: \textit{odgi break} removes the cycles, reducing the complexity of the graph topology.
-\textit{odgi groom} removes spurious inverting links by exploring the graph from the orientation supported by most paths.
-
-These are commonly-needed basic operations on the topology of the graph.
-
-\subsection{Untangling and navigating the pangenome with path-based coordinates}
+\subsection{Untangling and navigating the pangenome}
 \label{sec:untangle}
@@
 For understanding the large-scale organization of the graph, we benefit from an even simpler coordinate system.
-To provide a universal coordinate space for the graph, we project its nodes into a low-dimensional space (in practice, 1 or 2 dimensions) by learning the projection via a HOGWILD!
-\citep{niu2011hogwild}, path-guided stochastic gradient descent (PG-SGD) layout algorithm that adapts SGD-based drawing to pangenome graphs \citep{zheng2018graph}.
+To provide a universal coordinate space for the graph, we project its nodes into a low-dimensional space (in practice, 1 or 2 dimensions) by learning the projection via a HOGWILD!~\citep{niu2011hogwild}, path-guided stochastic gradient descent (PG-SGD) layout algorithm that adapts SGD-based drawing to pangenome graphs \citep{zheng2018graph}.
 This approach updates node relative positions to best-match their distance in the paths running through the graph.
@@
 In addition to these projections, we use the paths in the graph to provide a universal coordinate space.
-It also supports several kinds of ``lift-over'' of coordinates between different genomes (Fig. \ref{fig:odgi_viz}\textbf{f-g}).
+It also supports several kinds of ``lift-over'' of coordinates between different genomes (Fig.~\ref{fig:odgi_viz}\textbf{f-g}).
 These allow for the direct translation of single coordinates and ranges, in an interactive way.
@@
 
-Figure showing the untangling of C4 (Fig. \ref{fig:odgi_viz}\textbf{h})
+Figure showing the untangling of C4 (Fig.~\ref{fig:odgi_viz}\textbf{h})
 Figure showing tips over a single human chromosome from HPRC.
 
-\subsection{Sorting the pangenome graph topology}
+\FIXME{}
+\textit{odgi position} translates graph and path positions between or within graphs emitting the liftovers in TSV format.
+
+\textit{odgi server} implements a basic HTTP server which can translate a 'path:position' to a 'pangenome:position' very efficiently.
+
+\textit{odgi position/untangle/tips} all can output the BED format.
+
+\textit{odgi untangle} outputs the Pairwise mApping Format (PAF). PAF is a text format describing the approximate mapping positions between two set of sequences.
+The odgi untangle command projects paths into a reference-relative BEDPE file, decomposing paralogy relationships.
+During this process, it is capable of untangling loopy region resulting in linearized pairs of regions in the BEDPE file. A self dotplot assists in debugging and understanding the untangle process.
+\FIXME{}
+
+\subsection{Sorting the pangenome graph}
 \label{sec:sort}
@@
 
-% HTT exon1
-
-% stats / multiQC / bin / degree / depth
-
 Graphs statistics provide alternative ways to gain insight into pangenomes complexity revealing the overall structure, size, and features of a graph and its sequences.
 
+Pangenome graph topology can be derived by applying \textit{odgi matrix}, obtaining information on graph nodes connections in textual sparse matrix format.
+To investigate on the genomes encoded in the graph, \textit{odgi paths} allows users to calculate pairwise overlap statistics of groupings of paths and emit all path sequences in FASTA format.
+Furthermore, \textit{odgi flatten} generate a linearization of the graph by emitting the pangenome sequence (the concatenation of all graph node sequences) in FASTA format, and the projection of all paths on the linearized sequence in BED format.
+
 Applying \textit{odgi stats}, users can retrieve metrics describing the graph properties, such as the number of nodes, edges, paths, and graph length.
-It outputs pangenome statistics in TSV or YAML textual file formats.
-MultiQC's \citep{Ewels_2016} ODGI module %\footnote{\url{https://multiqc.info/docs/\#odgi} (accessed Oct 2021)}
-provides an interactive way to comparatively explore the statistics of an arbitrary number of graphs.
+It outputs pangenome statistics in tab-separated values (TSV) or YAML textual file formats.
+MultiQC's~\citep{Ewels_2016} ODGI module %\footnote{\url{https://multiqc.info/docs/\#odgi} (accessed Oct 2021)}
+provides an interactive way to comparatively explore such statistics of an arbitrary number of graphs.
+
 ODGI also offers more advanced tools for the interrogation of the graphs.
-\textit{odgi bin} summarizes the path information into bins of a specified size, enabling a summarized view of gigabase scale graphs.
-It provides tab delimited or JSON `bins'.
+To study very large pangenomes, users can use \textit{odgi bin} to summarize the path information into bins of a specified size, generating a summarized view of gigabase scale graphs in TSV or JSON file formats.
 %This concept found its way into the pangenome graph ontology \cite{Yokoyama2020}. %maybe we want to save this up for the discussion
-\textit{odgi depth} returns the node depth as defined by query criteria, allowing users to assess the complexity of regions in the graph.
-Collapsed sequences with highly identical repeats indicate intricate loci.
-Complex motifs can be also detected with \textit{odgi degree}, which returns the node degree as defined by query criteria.
-High degree nodes can be the mirror of variation, misassemblies, or problems in the pangenome building, making the tool useful for validation.
-In Fig.
-\ref{fig:metrics}, the characteristics of a 90 haplotypes human pangenome graph of the exon 1 huntingtin gene \citep{Sathasivam2013,Neueder2017} (\textit{HTTexon1}), obtained with the ODGI metrics tools, are shown.
+
+Genomes presenting sequences with highly identical repeats result in pangenome graphs with complex motifs that can be detected by \textit{odgi depth} and \textit{odgi degree}, which return the node depth and node degree, respectively, as defined by query criteria.
+Both tools emit the output in BED format, allowing users to assess the complexity of the graph and detect intricate regions.
+Indeed, high depth/degree nodes can be the mirror of genetic variation (Fig.~\ref{fig:metrics}), but also misassemblies or problems in the pangenome building, making the tools further useful for graph validation.
+%In Fig.~\ref{fig:metrics}, the characteristics of a 90 haplotypes human pangenome graph of the exon 1 huntingtin gene~\citep{Sathasivam2013,Neueder2017} (\textit{HTTexon1}) are shown.
 
@@
 
-Pangenome graphs stand to become a ubiquitous tool in genomics \citep{Eizenga_2020}.
+Pangenome graphs stand to become a ubiquitous tool in genomics~\citep{Eizenga_2020}.
 With ODGI we implemented a state-of-the-art tool suite that can transform, analyse, simplify, validate, and visualise pangenome graphs at large scale.
@@
 ODGI is a unique set of tools that enables scientists to explore and discover the underlying biology of pangenome graphs.
-Already, the tools are the backbone of pipelines such as the Pangenome Graph Builder \citep{pggb} (PGGB) or nf-core/pangenome \citep{pangenome}.
+Already, the tools are the backbone of pipelines such as the Pangenome Graph Builder~\citep{pggb} (PGGB) or nf-core/pangenome \citep{pangenome}.
 %\footnote{\url{https://github.com/nf-core/pangenome} (accessed Oct 2021)}.
@@
 ODGI's visualization tools produce static, large-scale snapshots.
-Recent interactive browsers are reference centric \citep{Beyer2019, Yokoyama2019, Durant2021, Liang2021} or focusing on one dimension \citep{Wick_2015, Gonnella2018}.
+Recent interactive browsers are reference centric~\citep{Beyer2019, Yokoyama2019, Durant2021, Liang2021} or focusing on one dimension~\citep{Wick_2015, Gonnella2018}.
 However, an interactive visualization solution that combines the 1D and 2D layout of a graph with annotation and read mapping information across different zoom levels is still missing.
@@
 
-So far, ODGI has not been applied to graphs build from scWGS \citep{Zhuo2021} data.
+So far, ODGI has not been applied to graphs build from scWGS~\citep{Zhuo2021} data.
 Each cell would be represented as a path in the graph, the path depth would be very high.

### 2021-11-01 manuscript/main.tex
Untangling and navigating the pangenome

@@
 %We specifically develop new methods focused on problems encountered when building pangenome graphs at the scale of vertebrate populations.
-%To ease interactive use, the majority of the ODGI tools are implemented in an index-free manner, avoiding the need to create index structures at each step of complex graph processing pipelines.
+%To ease interactive use, the majority of the ODGI's tools are implemented in an index-free manner, avoiding the need to create index structures at each step of complex graph processing pipelines.
 %Thanks to ODGI's efficient path representation, its tools can work with variation graphs with highly complex regions.
@@
 
-The key data in a pangenome graph is a representation of the alignment (or homology) relationships between sequences in the pangenome.
+The key data in a pangenome graph is a representation of the alignment (or homology) relationships between the sequences represented.
 Navigating and understanding the graph requires coordinate systems that we can use to link other data into the graph model, and thus to all genomes in the pangenome.
-Primarily, ODGI uses the embedded pangenome sequences to provide a set of coordinates in a standard, 1D format.
-These coordinates depend on the sequences, and are graph-independent.
-Our for positional conversion (\textit{odgi position}), annotation liftover (\textit{odgi untangle}), breakpoint identification (\textit{odgi tips}), subgraph extraction (\textit{odgi extract}), visualization (\textit{odgi viz}), path depth (\textit{odgi depth}), and graph complexity measurement (\textit{odgi degree}) can all operate on coordinates of paths in the graph.
+ODGI's tools use the embedded sequences to provide a universal coordinate space that is graph-independent, thereby remaining stable across different graphs built with the same genomes.
+%Our for positional conversion (\textit{odgi position}), annotation liftover (\textit{odgi untangle}), breakpoint identification (\textit{odgi tips}), subgraph extraction (\textit{odgi extract}), visualization (\textit{odgi viz}), path depth (\textit{odgi depth}), and graph complexity measurement (\textit{odgi degree}) can all operate on coordinates of paths in the graph.
 
-For understanding the large-scale organization of the graph, we benefit from an even simpler coordinate system.
-To provide a universal coordinate space for the graph, we project its nodes into a low-dimensional space (in practice, 1 or 2 dimensions) by learning the projection via a HOGWILD!~\citep{niu2011hogwild}, path-guided stochastic gradient descent (PG-SGD) layout algorithm that adapts SGD-based drawing to pangenome graphs \citep{zheng2018graph}.
-This approach updates node relative positions to best-match their distance in the paths running through the graph.
-Based on this projection, we can trivially generate 1D and 2D visualizations.
-We can also detect regions whose layout is distorted, and which thus may represent structural variation or assembly error.
-ODGI can project vector and matrix representations of the graph relative to these learned coordinate spaces.
+A universal coordinate system allows us to support several kinds of ``lift-over'' of coordinates between different genomes in the same or different graphs.
+For basic liftover operations, \textit{odgi server} provides a basic HTTP server that can translate very efficiently a path position to a pangenomic position within the same graph.
+For more complex coordinates conversions, \textit{odgi position} translates graph and path positions between or within graphs, emitting the liftovers in BED format.
+Coordinates can be specified in BED format, but users can even specify a GFF/GTF file to project the annotations into the pangenome graph (Fig.~\ref{fig:odgi_viz}\textbf{f-g}).
 
-In addition to these projections, we use the paths in the graph to provide a universal coordinate space.
-It also supports several kinds of ``lift-over'' of coordinates between different genomes (Fig.~\ref{fig:odgi_viz}\textbf{f-g}).
-These allow for the direct translation of single coordinates and ranges, in an interactive way.
-And, they let us extract a pairwise 1:many alignment between a given set of ``query'' sequences and a given set of ``target'' or reference sequences.
-This lets us convert the graph to lift-over maps compatible with standard software for projecting annotations and alignments from one genome to another.
-
-% position / tips
 % untangle (this follows on the jaccard graph mapping concepts in the previous section}
-%The graph is a model of an alignment of many genomes.
 %We can use coordinates in any genome to refer to it, but this requires a few basic operations ...
@@
 
-To obtain a more precise overview of collapsed \textit{loci}, we can apply \textit{odgi untangle} to segment paths into linear segments by breaking these segments where the paths loop back on themselves.
-In this way, we can obtain base-level information on the copy number status of the sequences in the locus.
+Pangenome graphs model alignments of many genomes.
+With \textit{odgi untangle}, users can extract pairwise 1:many alignment information between a given set of ``query'' sequences and a given set of ``target'' sequences (used as references)
+This lets us decompose both homology and paralogy relationships by also untangling loopy motifs in the graph, and obtaining linearized pairs of regions in BEDPE or PAF format.
+Being able to work with any sets of reference sequences lets us convert the graph to lift-over maps compatible with standard software for projecting annotations and alignments from one genome to another.
+As an example, by untangling the graph we can study the variation that lies in the collapsed regions due to sequence repeats (as shown in Fig.~\ref{fig:odgi_viz}\textbf{f}).
+Indeed, to obtain a more precise overview of the \textit{locus} in Fig.~\ref{fig:odgi_viz}\textbf{b-e},
+we can apply \textit{odgi untangle} to segment paths into linear segments by breaking these segments where the paths loop back on themselves.
+In this way, we obtain information on the position and copy number status of the sequences in the collapsed \textit{locus} (Fig.~\ref{fig:odgi_viz}\textbf{h}).
+Moreover, to obtain base-level precise information on the relationships between the repeated sequences, we can use any sequence aligner to align the pairs of regions that came from the untangling (as {cite wfmash}).
 
-Figure showing the untangling of C4 (Fig.~\ref{fig:odgi_viz}\textbf{h})
-Figure showing tips over a single human chromosome from HPRC.
-
-\FIXME{}
-\textit{odgi position} translates graph and path positions between or within graphs emitting the liftovers in TSV format.
-
-\textit{odgi server} implements a basic HTTP server which can translate a 'path:position' to a 'pangenome:position' very efficiently.
-
-\textit{odgi position/untangle/tips} all can output the BED format.
-
-\textit{odgi untangle} outputs the Pairwise mApping Format (PAF). PAF is a text format describing the approximate mapping positions between two set of sequences.
-The odgi untangle command projects paths into a reference-relative BEDPE file, decomposing paralogy relationships.
-During this process, it is capable of untangling loopy region resulting in linearized pairs of regions in the BEDPE file. A self dotplot assists in debugging and understanding the untangle process.
-\FIXME{}
+\FIXME{odgi tips, BED format, and a figure for it showing tips over a single human chromosome from HPRC?}
 

### 2021-11-01 manuscript/fig_extract_viz_draw_position_untangle.tex
Sorting the pangenome graph; added new image; partially updated to the new wgg.87 pggb run

--- 2021-10-29 manuscript/fig_odgi_viz.tex
@@
             \caption{}
-            \includegraphics[width=\linewidth, trim=0cm 0cm -10cm 0cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_og_mhc_og_O_og_lay_H1000w100_png_svg}
+            \includegraphics[width=\linewidth, trim=0cm 0cm -10cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_og_mhc_og_O_og_lay_H1000w100_png_svg}
         \end{subfigure} \\
@@
                 \caption{}
-                \includegraphics[width=\linewidth, trim=0 +1cm -0.5cm 0.5cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted}
+                \includegraphics[width=\linewidth, trim=0 +1cm -0.5cm 0.5cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted}
             \end{subfigure} \\
@@
                 \caption{}
-                \includegraphics[width=\linewidth, trim=0 +1cm -0.5cm 0.5cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_du}
+                \includegraphics[width=\linewidth, trim=0 +1cm -0.5cm 0.5cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_du}
             \end{subfigure} \\
@@
                 \caption{}
-                \includegraphics[width=\linewidth, trim=0 +1cm -0.5cm 0.5cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_z}
+                \includegraphics[width=\linewidth, trim=0 +1cm -0.5cm 0.5cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_z}
             \end{subfigure} \\
@@
                 \caption{}
-                \includegraphics[width=\linewidth, trim=0 +1cm -0.5cm 0.5cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_m}
+                \includegraphics[width=\linewidth, trim=0 +1cm -0.5cm 0.5cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_m}
             \end{subfigure}
@@
                 \centering
-                \includegraphics[width=0.65\linewidth, trim=0 +2cm 0 0.5cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_bandage}
+                \includegraphics[width=0.65\linewidth, trim=0 +2cm 0 0.5cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_bandage}
                 \caption{}
@@
                 \centering
-                \includegraphics[width=0.65\linewidth, trim=0 +2cm 0 0.5cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_bandage_zoom_in}
+                \includegraphics[width=0.65\linewidth, trim=0 +2cm 0 0.5cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_bandage_zoom_in}
                 \caption{}
@@
             \begin{center}
-                \includegraphics[width=\linewidth, trim=0cm 0.9cm -2.5cm 0.9cm]{fig/fig_extract_viz_draw_position_untangle/chr6.C4.sorted.untangle.bed.subset.tsv}
+                \includegraphics[width=\linewidth, trim=0cm 0.9cm -2.5cm 0.9cm]{fig/extract_viz_draw_position_untangle/chr6.C4.sorted.untangle.bed.subset.tsv}
             \end{center}
@@
     \caption{
-        Visualizing the MHC and C4 \textit{loci} pangenome graphs.
+        Visualizing the MHC and C4 pangenome graphs.
 \textbf{(a)} \textit{odgi draw} layout of the MHC pangenome graph extracted from a whole human pangenome graph of 90 haplotypes.
@@
         The longer links at the bottom indicate the presence of a structural variant (long link) with another structural variant nested inside it (short link on the left).
-        Indeed, human C4 exists as 2 functionally distinct genes, \textit{C4A} and \textit{C4B}, which both vary in structure and copy number \citep{Sekar_2016}.
+        Indeed, human C4 exists as 2 functionally distinct genes, \textit{C4A} and \textit{C4B}, which both vary in structure and copy number~\citep{Sekar_2016}.
 The longer link indicates that the copy number status varies across the haplotypes represented in the pangenome.
@@
 The red rectangle highlights the regions where \textit{C4A} and \textit{C4B} genes differ.
-        \textbf{(g)} \textit{Bandage} layout of the region where \textit{C4A} and \textit{C4B} genes differ due to single nucleotide variants leading to changes in the encoded protein sequences.
+        \textbf{(g)} Annotated \textit{bandage} layout of the C4 region where \textit{C4A} and \textit{C4B} genes differ due to single nucleotide variants leading to changes in the encoded protein sequences.
 Node labels were annoted by using \textit{odgi position}.
@@
 %        % include first image
-%        \includegraphics[width=\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted}
+%        \includegraphics[width=\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted}
 %        \label{fig:odgi_viz_default}
@@
 %        % include second image
-%        \includegraphics[width=\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_du}
+%        \includegraphics[width=\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_du}
 %        \label{fig:odgi_viz_color_by_path_pos}
@@
 %        % include second image
-%        \includegraphics[width=\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_z}
+%        \includegraphics[width=\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_z}
 %        \label{fig:odgi_viz_color_by_inversion_rate}
@@
 %        % include fourth image
-%        \includegraphics[width=\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_m}
+%        \includegraphics[width=\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_m}
 %        \label{fig:odgi_viz_color_by_path_depth}
@@
 %        % include first image
-%        \includegraphics[width=0.5\linewidth, trim=0 +2cm 0 0.5cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_layout}
+%        \includegraphics[width=0.5\linewidth, trim=0 +2cm 0 0.5cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_layout}
 %        \label{fig:odgi_viz_default2}
@@
 %        %todo BANDAGE
-%        \includegraphics[width=0.5\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/fig_extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_layout}
+%        \includegraphics[width=0.5\linewidth, trim=0 +1.5cm -0.5cm 0.5cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_layout}
 %        \label{fig:odgi_viz_color_by_inversion_rate2}

### 2021-11-01 manuscript/fig_sorting.tex
Sorting the pangenome graph; added new image; partially updated to the new wgg.87 pggb run

\begin{figure}[h!]
	\begin{subfigure}{\linewidth}
		%\caption{}
		\centering
		% include first image
		\includegraphics[width=1.0\linewidth, trim=-0cm 2cm 0 0cm]{fig/sorting/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_C4_bad_sorted}
		\label{fig:bad-sorting}
	\end{subfigure}
%	\includegraphics[width=\linewidth]{fig/metrics/chr4.pan.HTTex1.gfa.multiqc_odgi_stats.png}
	\caption{
		Visualization of a poorly sorted C4 pangenome graph.
		Same \textit{odgi viz} visualization as the one shown in Fig.~\ref{fig:odgi_viz}\textbf{e}, but with a graph where the bad node order hides the underlying copy number variation status present in the C4 region.
	}
	\label{fig:sorting}
\end{figure}

### 2021-11-01 manuscript/main.tex
Sorting the pangenome graph; added new image; partially updated to the new wgg.87 pggb run

@@
 
-Pangenome graphs can be used in a variety of applications, ranging from read mapping to variant identification and genotyping ~\citep{Eizenga_2020}.
+Pangenome graphs can be used in a variety of applications, ranging from read mapping to variant identification and genotyping~\citep{Eizenga_2020}.
 However, graphs presenting complex topology can increase the computational overhead of many downstream analyses.
@@
 
-With \textit{odgi prune} users can remove complex parts as defined by query criteria, while with \textit{odgi break} they can remove cycles in the graph, reducing the complexity of the graph topology.
+For simplifying the graph structure, users can use \textit{odgi prune} to take away complex parts as defined by query criteria,
+while with \textit{odgi break} they can remove cycles in the graph, reducing the complexity of the graph topology.
 Furthermore, \textit{odgi groom} allows removing spurious inverting links by exploring the graph from the orientation supported by most paths.
@@
 In this way, we obtain information on the position and copy number status of the sequences in the collapsed \textit{locus} (Fig.~\ref{fig:odgi_viz}\textbf{h}).
-Moreover, to obtain base-level precise information on the relationships between the repeated sequences, we can use any sequence aligner to align the pairs of regions that came from the untangling (as {cite wfmash}).
+Moreover, to obtain base-level precise information on the relationships between the repeated sequences, we can align them the pairs of regions that came from the untangling to guide the alignment~\citep{wfmash}.
 
@@
 
-% sort / layout / NO tension! We keep that for the sorting paper
+Pangenome graphs can hide their underlying latent structures, introducing difficulties in the analysis and interpretation (Fig.~\ref{fig:sorting}\textbf{a}).
+Among the causes of this is the correct ordering of the graph nodes in a convenient number of dimensions.
+ODGI provides a variety of sorting algorithms to find the best graph node order in 1 or 2 dimensions (1D and 2D respectively), allowing us
+to understand the sparse structures typically found in pangenome graphs and the genetic variation they represent.
 
-% Allows us to understand the sparse structures typically found in pangenome graphs.
-
-%Pangenome graphs built from raw sets of alignments may have complex structures which can introduce difficulty in downstream analyses.
-Uncovering their latent topology allows us to understand their typical sparse structures.
-
-\textit{odgi sort} offers multiple sorting algorithms to find the best node order in one dimension.
-Most notably, nodes can be sorted topologically, randomly, by breaking cycles in the graph, by grooming, or by using a novel path-guided stochastic gradient descent (PG-SGD) algorithm.
+To find the best order of graph nodes in 1D, \textit{odgi sort} provides multiple sorting algorithms which can be combined into a series of sorts to take advantage of the strength of each.
+Most notably, nodes can be sorted topologically, randomly, by breaking cycles in the graph (\S\ref{sec:edit}), by grooming (\S\ref{sec:edit}), and/or by using a novel path-guided (PG) stochastic gradient descent (SGD) algorithm, also called PG-SGD.
 The latter one exploits the biological information in the paths for sorting the nodes.
-Combining a series of sorts can mitigate the weaknesses of individually applied sorting algorithms.
-%In Figure \textbf{\textit{TODO MHC wgg.85}}, a comparison of an unsorted human pangenome MHC graph and a sorted one are displayed.
-
-%For the establishment of a layout in two dimensions \textit{odgi layout} applies a 2D adjusted PG-SGD algorithm.
-A calculated layout can be visualized statically with \textit{odgi draw} (see Figure \textit{\textbf{TODO MHC wgg.85}}) or interactively explored with the in-development \textit{gfaestus}\{\url{https://github.com/chfi/gfaestus}}.
+%For understanding the large-scale organization of the graph, we benefit from an even simpler coordinate system.
+In particular, to provide a universal coordinate space for the graph, we project its nodes into a low-dimensional space
+by learning the projection via a HOGWILD! approach~\citep{niu2011hogwild} implemented in the PG-SGD algorithm.
+The PG-SGD adapts a SGD-based drawing to pangenome graphs~\citep{zheng2018graph}.
+This approach updates node relative positions to best-match their distance in the paths running through the graph.
+ODGI can project vector (in 1D) and matrix (2D) representations of the graph relative to these learned coordinate spaces.
+Based on this projection, we can trivially sort graph nodes in 1D. Moreover, we support the same concept in 2D in \textit{odgi layout}
+by providing a 2D implementation of the PG-SGD algorithm.
+As the PG-SGD sorts graph nodes with respect to the paths which walk through them, we can also detect regions whose layout is distorted, which may represent structural variation or assembly error.
 
@@
 
-\input{fig_odgi_viz}
+\input{fig_extract_viz_draw_position_untangle}
+
+\input{fig_sorting}
 

### 2021-11-01 scripts/fig_extract_viz_draw_position_untangle.sh

@@ -1,21 +1,24 @@
 #!/bin/bash
 
-filename_chr6_gfa=chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa
-
-# Download and build the graph
-wget http://hypervolu.me/~guarracino/${filename_chr6_gfa}.gz
-gunzip ${filename_chr6_gfa}.gz
-odgi build -g ${filename_chr6_gfa} -o ${filename_chr6_gfa}.og -t 16 -P
-
-
 # Find C4 coordinates
 wget http://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/hg38.chrom.sizes
 wget https://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/genes/hg38.ncbiRefSeq.gtf.gz
-zgrep 'gene_id "C4A"\|gene_id "C4B"' hg38.ncbiRefSeq.gtf.gz | awk '$1 == "chr6"' | cut -f 1,4,5 | bedtools sort | bedtools merge -d 15000 | bedtools slop -l 10000 -r 20000 -g hg38.chrom.sizes | sed 's/chr6/grch38#chr6/g' > C4.coordinates.bed
+zgrep 'gene_id "C4A"\|gene_id "C4B"' hg38.ncbiRefSeq.gtf.gz |
+  awk '$1 == "chr6"' | cut -f 1,4,5 |
+  bedtools sort | bedtools merge -d 15000 | bedtools slop -l 10000 -r 20000 -g hg38.chrom.sizes |
+  sed 's/chr6/grch38#chr6/g' > hg38.ncbiRefSeq.C4.coordinates.bed
+
+
+filename_chr6_gfa=chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth
+
+# Download and build the graph
+wget http://hypervolu.me/~guarracino/${filename_chr6_gfa}.gfa.gz
+gunzip ${filename_chr6_gfa}.gfa.gz
+odgi build -g ${filename_chr6_gfa}.gfa -o ${filename_chr6_gfa}.og -t 16 -P
 
 
 # Extraction, explosion, optimization, and sorting
-odgi extract -i ${filename_chr6_gfa}.og -b C4.coordinates.bed -o - --full-range -t 16 -P |
[…]

### 2021-11-01 scripts/fig_sorting.sh

@@ -1,0 +1,7 @@
+#!/bin/bash
+
+filename_chr6_gfa=chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth
+
+# Take the ${filename_chr6_gfa}.C4.0.og from `fig_extract_viz_draw_position_untangle.sh`
+
+odgi sort -i chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.C4.sorted.og -o - -x 1 -Y | odgi viz -i - -o "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_C4_bad_sorted.png -c 12 -w 100 -y 50 -p chr6.selected_paths.txt -m -B Spectral:4

### 2021-11-01 manuscript/main.tex
thinner bandage layout; formatting

@@
 In this way, we obtain information on the position and copy number status of the sequences in the collapsed \textit{locus} (Fig.~\ref{fig:odgi_viz}\textbf{h}).
-Moreover, to obtain base-level precise information on the relationships between the repeated sequences, we can align them the pairs of regions that came from the untangling to guide the alignment~\citep{wfmash}.
+Moreover, to obtain base-level precise information on the relationships between the repeated sequences, we can align them by using the pairs of regions that came from the untangling to guide the alignment~\citep{wfmash}.
 
@@
 With ODGI we implemented a state-of-the-art tool suite that can transform, analyse, simplify, validate, and visualise pangenome graphs at large scale.
-%
+
 Lifting over annotations and linearizing nested graph structures place the suite as the bridge between traditional linear reference genome analysis and pangenome graphs.
 ODGI is a unique set of tools that enables scientists to explore and discover the underlying biology of pangenome graphs.
-Already, the tools are the backbone of pipelines such as the Pangenome Graph Builder~\citep{pggb} (PGGB) or nf-core/pangenome \citep{pangenome}.
+Already, the tools are the backbone of pipelines such as the Pangenome Graph Builder~\citep{pggb} (PGGB) or nf-core/pangenome~\citep{pangenome}.
 %\footnote{\url{https://github.com/nf-core/pangenome} (accessed Oct 2021)}.
@@
 
-So far, ODGI has not been applied to graphs build from scWGS~\citep{Zhuo2021} data.
-Each cell would be represented as a path in the graph, the path depth would be very high.
+So far, ODGI has not been applied to graphs build from scWGS~\citep{Zhuo2021} data. Each cell would be represented as a path in the graph, the path depth would be very high.
 As ODGI was designed to deal with the path depth of human pangenome graphs, it remains to be seen, if it will work equally well with scWGS data.

### 2021-11-01 manuscript/main.tex
spoiler

@@
 \address{
-$^{\text{\sf 1}}$University of Tor Vergata, Rome, Italy \\
+$^{\text{\sf 1}}$Human Technopole, Milan, Italy \\
 $^{\text{\sf 2}}$Quantitative Biology Center (QBiC), University of T\"ubingen, T\"ubingen, Germany, 72076 \\

### 2021-11-01 manuscript/main.tex
fix the spoiler

@@
 \address{
-$^{\text{\sf 1}}$Human Technopole, Milan, Italy \\
+$^{\text{\sf 1}}$Genomics Research Centre, Human Technopole, Milan, Italy \\
 $^{\text{\sf 2}}$Quantitative Biology Center (QBiC), University of T\"ubingen, T\"ubingen, Germany, 72076 \\

### 2021-11-01 manuscript/main.tex
the discussion is hard

@@
 \textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
-\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
+%\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
 }
@@
 These flexible models let us encode any kind of variation, allowing the generation of comprehensive data systems on which to base analyses of genome evolution.
-Although these data structures are of clear utility to researchers~\citep{cpang2018}, and have been the focus of numerous applications in read mapping~\citep{Garrison:2018,Baaijens_2019,Hickey:2020,Sibbesen_2021}, the scientific community still lacks a toolset specifically focused on graph manipulation and interrogation.
+Although these data structures are of clear utility to researchers~\citep{cpang2018}, and have been the focus of numerous applications in read mapping~\citep{Garrison:2018,Baaijens_2019,Hickey:2020,Sibbesen_2021},
+the scientific community still lacks a toolset specifically focused on graph manipulation and interrogation.
 
@@
 Here, we present the Optimized Dynamic Genome/Graph Implementation (ODGI) toolkit, a pangenome graph interrogation and transformation system specifically implemented to handle the data scales encountered when working with pangenomes built from hundreds of haplotype-resolved genomes.
-ODGI provides a set of standard operations on the variation graph data model, generalizing ``genome arithmetic'' concepts like those found in BEDTools~\citep{Quinlan_2010} to work on pangenome graphs, and providing a variety of operations, such as visualization, sorting, and liftover projections, all critical to understand and exploit pangenome graphs.
+ODGI provides a set of standard operations on the variation graph data model, generalizing ``genome arithmetic'' concepts like those found in BEDTools~\citep{Quinlan_2010} to work on pangenome graphs,
+and providing a variety of operations, such as visualization, sorting, and liftover projections, all critical to understand and exploit pangenome graphs.
 
@@
 \section{Discussion}
+% Short recap
+% Confessions
+% Future plans
+Pangenome graphs stand to become a ubiquitous model in genomics~\citep{Eizenga_2020}.
+With ODGI we implemented a state-of-the-art tool suite that enables users to explore and discover the underlying biology in pangenomes graphs, filling the gap that made pangenomic analyzes difficult.
+Indeed, with ODGI's tools users can easily transform, analyse, simplify, validate, and visualise pangenome graphs at large scale.
 
-Pangenome graphs stand to become a ubiquitous tool in genomics~\citep{Eizenga_2020}.
-With ODGI we implemented a state-of-the-art tool suite that can transform, analyse, simplify, validate, and visualise pangenome graphs at large scale.
+ODGI's visualization tools produce static, large-scale snapshots.
+Recent interactive browsers are reference centric~\citep{Beyer2019, Yokoyama2019, Durant2021, Liang2021} or focusing on 1D~\citep{Wick_2015, Gonnella2018}.
+However, an interactive visualization solution that combines the 1D and 2D layout of a graph with annotation and read mapping information across different zoom levels is still missing.
+Nevertheless, ODGI's graph layout algorithms could become a backbone of such tools, as is already the case for \textit{gfaestus}~\citep{gfaestus} %\footnote{\url{https://github.com/chfi/gfaestus}},
+which currently relies on the \textit{odgi layout} output for its interactive 2D visualization.
 
 Lifting over annotations and linearizing nested graph structures place the suite as the bridge between traditional linear reference genome analysis and pangenome graphs.
-ODGI is a unique set of tools that enables scientists to explore and discover the underlying biology of pangenome graphs.
+
 Already, the tools are the backbone of pipelines such as the Pangenome Graph Builder~\citep{pggb} (PGGB) or nf-core/pangenome~\citep{pangenome}.
@@
 
-ODGI provides common operation on standard pangenomic file formats, such as GFA.
 ODGI aims to be a reference implementation for many pangenome related operations.
 In the near future we aim to improve RDF support for annotation in the in-memory graph~\citep{Yokoyama2020}, for example, and allow for federated queries on the pangenome, and FAIR data sharing.
-
-ODGI's visualization tools produce static, large-scale snapshots.
-Recent interactive browsers are reference centric~\citep{Beyer2019, Yokoyama2019, Durant2021, Liang2021} or focusing on one dimension~\citep{Wick_2015, Gonnella2018}.
-However, an interactive visualization solution that combines the 1D and 2D layout of a graph with annotation and read mapping information across different zoom levels is still missing.
-ODGI's graph layout algorithms could become a backbone of such tools, as is already the case for \textit{gfaestus} \citep{gfaestus}%\footnote{\url{https://github.com/chfi/gfaestus}}, which currently relies on the \textit{odgi layout} output for its interactive 2D visualization.
 
@@
 With the increased adoption of long read sequencing we expect pangenome tools to become increasingly common in biomedical research.
-Particularly for targets that involve complex variation, such as cancer, plant genomics and metagenomics, ODGI should facilitate disentangling, describing and analysing a much larger set of variation than previously was possible with tools that depend on short reads and reference genomes.
+Particularly for targets that involve complex variation, such as cancer, plant genomics and metagenomics, ODGI will facilitate disentangling,
+describing and analysing a much larger set of variation than previously was possible with tools that depend on short reads and reference genomes.
 
@@
 
-Data used to build Human pangenome graphs is available at \url{https://github.com/human-pangenomics/HPP_Year1_Data_Freeze_v1.0}.
+\FIXME{Data used to build Human pangenome graphs is available at \url{https://github.com/human-pangenomics/HPP_Year1_Data_Freeze_v1.0}.}
+Code used to build images in this manuscript is available at \url{https://github.com/pangenome/odgi-paper}.
 TODO!

### 2021-11-02 manuscript/main.tex
drafted the discussion

@@
 \section{Discussion}
-% Short recap
-% Confessions
-% Future plans
-Pangenome graphs stand to become a ubiquitous model in genomics~\citep{Eizenga_2020}.
-With ODGI we implemented a state-of-the-art tool suite that enables users to explore and discover the underlying biology in pangenomes graphs, filling the gap that made pangenomic analyzes difficult.
-Indeed, with ODGI's tools users can easily transform, analyse, simplify, validate, and visualise pangenome graphs at large scale.
+Pangenome graphs stand to become a ubiquitous model in genomics thanks to their capability to represent any genetic variant without being affected by the reference-bias~\citep{Eizenga_2020}.
+However, despite this great potential, their spread is impeded by the lack of tools capable of managing and analyzing pangenome graphs easily and efficiently.
 
-ODGI's visualization tools produce static, large-scale snapshots.
+With ODGI we implemented a state-of-the-art tool suite that enables users to explore and discover the underlying biology in pangenomes graphs, filling the gap that made pangenomic analyses difficult.
+As a result, we provides tools to easily transform, analyse, simplify, validate, and visualise pangenome graphs at large scale.
+In particular, lifting over annotations and linearizing nested graph structures place the suite as the bridge between traditional linear reference genome analysis and pangenome graphs.
+With the increased adoption of long read sequencing we expect pangenomic tools to become increasingly common in biomedical research.
+Particularly for targets that involve complex variation, such as cancer, plant genomics and metagenomics, ODGI will facilitate disentangling,
+describing and analysing a much larger set of variation than previously was possible with tools that depend on short reads and reference genomes,.
+Futhermore, users can even consider ODGI as a framework, taking advantage of its algorithms to develop new and more advanced tools that work on pangenome graphs,
+thus expanding the type of possible pangenomic analyses available to the scientific community.
+
+ODGI is already the backbone of the pipelines such as the Pangenome Graph Builder~\citep{pggb} or nf-core/pangenome~\citep{pangenome}, where it is also applied for debugging purposes.
+%\footnote{\url{https://github.com/nf-core/pangenome} (accessed Oct 2021)}.
+Indeed, its static, large-scale 1D and 2D visualizations of the pangenome graphs allow grasping the underlying structure and any problems in the graph construction.
+However, an interactive solution that combines the 1D and 2D layout of a graph with annotation and read mapping information across different zoom levels is still missing.
 Recent interactive browsers are reference centric~\citep{Beyer2019, Yokoyama2019, Durant2021, Liang2021} or focusing on 1D~\citep{Wick_2015, Gonnella2018}.
-However, an interactive visualization solution that combines the 1D and 2D layout of a graph with annotation and read mapping information across different zoom levels is still missing.
-Nevertheless, ODGI's graph layout algorithms could become a backbone of such tools, as is already the case for \textit{gfaestus}~\citep{gfaestus} %\footnote{\url{https://github.com/chfi/gfaestus}},
-which currently relies on the \textit{odgi layout} output for its interactive 2D visualization.
+In ODGI, we provide a graph layout algorithm that can become a backbone of such tools, as is already the case for \textit{gfaestus}~\citep{gfaestus} %\footnote{\url{https://github.com/chfi/gfaestus}},
+which currently relies on the \textit{odgi layout} output for its interactive 2D graph visualization.
 
-Lifting over annotations and linearizing nested graph structures place the suite as the bridge between traditional linear reference genome analysis and pangenome graphs.
+ODGI aims to be a reference implementation for many pangenome related operations, thus future work will be focused on multiple directions.
+First, we will expand its metadata capabilities to annotate large pangenome graphs.
+Futhermore, we will improve RDF support for annotation in the in-memory graph~\citep{Yokoyama2020} and allow for federated queries on pangenome graphs.
+Finally, we will explore the addition of support for RNA sequences and proteins to extend the application of pangenomic approaches to other levels in the genetic information flow from DNA to phenotype.
 
-Already, the tools are the backbone of pipelines such as the Pangenome Graph Builder~\citep{pggb} (PGGB) or nf-core/pangenome~\citep{pangenome}.
-%\footnote{\url{https://github.com/nf-core/pangenome} (accessed Oct 2021)}.
-Future work will add support for RNA and protein sequences and expand on metadata capabilities of large pangenome graphs.
-
-ODGI aims to be a reference implementation for many pangenome related operations.
-In the near future we aim to improve RDF support for annotation in the in-memory graph~\citep{Yokoyama2020}, for example, and allow for federated queries on the pangenome, and FAIR data sharing.
-
-So far, ODGI has not been applied to graphs build from scWGS~\citep{Zhuo2021} data. Each cell would be represented as a path in the graph, the path depth would be very high.
-As ODGI was designed to deal with the path depth of human pangenome graphs, it remains to be seen, if it will work equally well with scWGS data.
-
-With the increased adoption of long read sequencing we expect pangenome tools to become increasingly common in biomedical research.
-Particularly for targets that involve complex variation, such as cancer, plant genomics and metagenomics, ODGI will facilitate disentangling,
-describing and analysing a much larger set of variation than previously was possible with tools that depend on short reads and reference genomes.
+%So far, ODGI has not been applied to graphs build from scWGS~\citep{Zhuo2021} data. Each cell would be represented as a path in the graph, the path depth would be very high.
+%As ODGI was designed to deal with the path depth of human pangenome graphs, it remains to be seen, if it will work equally well with scWGS data.
 

### 2021-11-02 manuscript/main.tex
the path jaccard concept and odgi tips

@@
 Paths $P = p_1\ldots p_{|P|}$ describe walks over node strands: $p_i = s_1 \ldots s_{|p_i|}$.
-When used as a pangenome graph, $V$ expresses sequences, haplotypes, and annotations as paths.
+When used as a pangenome graph, $V$ expresses sequences, haplotypes, contigs, and annotations as paths.
 %The utility of the variation graph model lies in its lossless representation of genomes and their alignment.
@@
 \item \textit{odgi draw} renders a 2D image of the graph (\S\ref{sec:viz}).
-\item \textit{odgi extract} subsets of the graph based on path ranges (\S\ref{sec:extract}).
+\item \textit{odgi extract} excerpts subsets of the graph based on path ranges (\S\ref{sec:extract}).
 \item \textit{odgi explode} breaks the graph into connected components (\S\ref{sec:extract}).
@@
 \item \textit{odgi chop} breaks long nodes into shorter ones (\S\ref{sec:edit}).
-\item \textit{odgi unchop} combines redundant nodes (\S\ref{sec:edit}).
+\item \textit{odgi unchop} combines unitig nodes (\S\ref{sec:edit}).
 \item \textit{odgi break} removes cycles in the graph (\S\ref{sec:edit}).
@@
 
-For interchange with other tools, \textit{odgi view} can convert a graph in ODGI binary format to the standard GFAv1~\citep{GFA}.
+For interchange with other tools, \textit{odgi view} can convert a graph in ODGI binary format to the standard GFAv1.
 
@@
 
-\textit{odgi viz} supports a binned, linearized rendering in 1 dimension (that is, all graph nodes lie on the same axis).
+\textit{odgi viz} supports a binned, linearized rendering in 1 dimension (1D) (that is, all graph nodes lie on the same axis).
 This visualization is computed in linear-time and offers a human-interpretable format suitable for understanding the topology and genome relationships in the pangenome graph (Fig.~\ref{fig:odgi_viz}).
@@
 Pangenome graphs built from hundreds of haplotype-resolved \textit{de novo} genome assemblies are very large, but it is often necessary to work with only a small portion of the genomes represented,
-such as a specific \textit{locus} (Fig.~\ref{fig:odgi_viz}\textbf{a}) or a smaller region (Fig.~\ref{fig:odgi_viz}\textbf{b-g}), or even a single gene (Fig.~\ref{fig:metrics}),
-thereby simplifying the downstream analyses and reducing the resources to work only with the extracted graphs.
+such as a specific \textit{locus} (Fig.~\ref{fig:odgi_viz}\textbf{a}) or a smaller region (Fig.~\ref{fig:odgi_viz}\textbf{b-g}), or even a single gene (Fig.~\ref{fig:metrics}).
+This simplifies the downstream analyses and reduces the resources to work only with the extracted graphs.
 %Subsetting operations allow us to zoom in on particular genes or regions of interest for more precise examination.
@@
 Finally, the portions of the paths (\textit{i.e.}, the subpaths) walking through the selected nodes are extracted and added to the new subgraph.
-Subpaths are search in parallel, created serially, and extended in parallel again thanks to the parallelism enabled by the ODGI data structure (see \S\ref{sec:build}), making \textit{odgi extract} a scalable solution to extract also complex subregions presenting nodes with very high depth.
+Subpaths are searched in parallel, created serially, and extended in parallel again thanks to the parallelism enabled by the ODGI data structure (see \S\ref{sec:build}), making \textit{odgi extract} a scalable solution to extract also complex subregions presenting nodes with very high depth.
 
@@
 
+For a precise translation process when conversing a query position to a reference position in a repeat region the \textit{path jaccard} concept is applied: It could be that the found reference node is visited several times by the reference.
+To ensure a precise translation, the best jaccard for the query step and each reference candidate step is calculated.
+
 % untangle (this follows on the jaccard graph mapping concepts in the previous section}
@@
 Pangenome graphs model alignments of many genomes.
-With \textit{odgi untangle}, users can extract pairwise 1:many alignment information between a given set of ``query'' sequences and a given set of ``target'' sequences (used as references)
+With \textit{odgi untangle}, users can extract pairwise \textit{1}:\textit{many} alignment information between a given set of ``query'' sequences and a given set of ``target'' sequences (used as references).
 This lets us decompose both homology and paralogy relationships by also untangling loopy motifs in the graph, and obtaining linearized pairs of regions in BEDPE or PAF format.
@@
 we can apply \textit{odgi untangle} to segment paths into linear segments by breaking these segments where the paths loop back on themselves.
+The \textit{path jaccard} concept ensures that the breaking is nucleotide-precise.
 In this way, we obtain information on the position and copy number status of the sequences in the collapsed \textit{locus} (Fig.~\ref{fig:odgi_viz}\textbf{h}).
@@
 
-\FIXME{odgi tips, BED format, and a figure for it showing tips over a single human chromosome from HPRC?}
+Having build a pangenome graph from assemblies in contig state, we can ask questions like: Where do our contigs' ends match the reference(s)?
+An answer to this question can highlight regions that are tricky to assemble~\ref{fig:odgi_tips}.
+\textit{odgi tips} can identify the break point positions of the contigs relative to the reference(s) in the graph by walking from the ends of each contig until a reference node is found. 
+It could be that the reference visits the node several times.
+Therefore, for each contig range (a tip) \textit{odgi tips} takes a look at each possible reference window and finds the most similar one using the \textit{path jaccard} concept.
+The output is a BED file with the best reference hit and position for each of the contigs' ends.
+
+\FIXME{add odgi tips figure: only focus on the beta-defensin cluster [p23.3-p22] or more?}
 
@@
 Among the causes of this is the correct ordering of the graph nodes in a convenient number of dimensions.
-ODGI provides a variety of sorting algorithms to find the best graph node order in 1 or 2 dimensions (1D and 2D respectively), allowing us
+ODGI provides a variety of sorting algorithms to find the best graph node order in 1 or 2 dimensions, allowing us
 to understand the sparse structures typically found in pangenome graphs and the genetic variation they represent.
@@
 The PG-SGD adapts a SGD-based drawing to pangenome graphs~\citep{zheng2018graph}.
-This approach updates node relative positions to best-match their distance in the paths running through the graph.
+This approach updates node relative positions to best-match their nucleotide distance in the paths running through the graph.
 ODGI can project vector (in 1D) and matrix (2D) representations of the graph relative to these learned coordinate spaces.

### 2021-11-02 manuscript/fig_metrics.tex
remove '.' from figure names

--- 2021-10-29 manuscript/fig_metrics.tex
@@
 		% include second image
-		\includegraphics[width=1.0\linewidth, trim=-1.5cm 2cm -0.1cm 0.5cm]{fig/metrics/chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.HTTex1.og.O.og.tiny.og}
+		\includegraphics[width=1.0\linewidth, trim=-1.95cm 2cm -1.0cm 0.5cm]{fig/metrics/chr4_pan_fa_a2fb268_e820cd3_9ea71d8_smooth_gfa_og_HTTex1_og_O_og_tiny_og}
 		\label{fig:metrics-viz}

### 2021-11-03 scripts/metrics.sh
hide odgi server and its index

@@ -1,7 +1,7 @@
 #!/bin/bash
 
-wget https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/scratch/2021_07_30_pggb/chroms/chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.gz
-gunzip chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.gz
+wget http://hypervolu.me/~guarracino/chr4.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
+gunzip chr4.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
 wget https://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/genes/hg38.ncbiRefSeq.gtf.gz
 gunzip hg38.ncbiRefSeq.gtf.gz
 
@@ -9,21 +9,21 @@ grep HTT hg38.ncbiRefSeq.gtf | grep "exon_number \"1\"" | grep "gene_name \"HTT\
 #chr4	ncbiRefSeq.2021-09-09	exon	3074681	3075088	.	+	.	gene_id "HTT"; transcript_id "NM_001388492.1"; exon_number "1"; exon_id "NM_001388492.1.1"; gene_name "HTT";
 #chr4	ncbiRefSeq.2021-09-09	exon	3074681	3075088	.	+	.	gene_id "HTT"; transcript_id "NM_002111.8"; exon_number "1"; exon_id "NM_002111.8.1"; gene_name "HTT";
 
-odgi build -g chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa -o chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og -t 16 -P
-odgi paths -L -i chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og > chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.paths
+odgi build -g chr4.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa -o chr4.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.og -t 16 -P
[…]

### 2021-11-03 manuscript/main.tex
reduce the warnings and the citation error (I'm skipping whatever remains of this entry)

@@
 \item \textit{odgi groom} resolves spurious inverting links (\S\ref{sec:edit}).
-\item \textit{odgi position} lifts coordinates between path and graph positions (\S\ref{sec:untangle}).
-\item \textit{odgi server} provides a HTTP server to lift coordinates (\S\ref{sec:untangle}).
+\item \textit{odgi position} lifts coordinates between path and graph positions (\S\ref{sec:untangle}).%
+%\item \textit{odgi server} provides a HTTP server to lift coordinates (\S\ref{sec:untangle}).
 \item \textit{odgi untangle} deconvolutes paths relative to a reference (\S\ref{sec:untangle}).
@@
 A universal coordinate system allows us to support several kinds of ``lift-over'' of coordinates between different genomes in the same or different graphs.
-For basic liftover operations, \textit{odgi server} provides a basic HTTP server that can translate very efficiently a path position to a pangenomic position within the same graph.
-For more complex coordinates conversions, \textit{odgi position} translates graph and path positions between or within graphs, emitting the liftovers in BED format.
+%For basic liftover operations, \textit{odgi server} provides a basic HTTP server that can translate very efficiently a path position to a pangenomic position within the same graph.
+\textit{odgi position} translates graph and path positions between or within graphs, emitting the liftovers in BED format.
 Coordinates can be specified in BED format, but users can even specify a GFF/GTF file to project the annotations into the pangenome graph (Fig.~\ref{fig:odgi_viz}\textbf{f-g}).

### 2021-11-03 manuscript/fig_metrics.tex
reduce the warnings and the citation error (I'm skipping whatever remains of this entry)
metrics: add nucleotide-level vg viz

@@
 	\end{subfigure}
+	\begin{subfigure}{\linewidth}
+	\caption{}
+	\centering
+	% include second image
+	\includegraphics[width=1.0\linewidth, trim=1.25cm 0.5cm 1.5cm 0cm]{fig/metrics/chr4_pan_HTTex1_STR_xg_svg.pdf}
+	\label{fig:metrics-str}
+\end{subfigure}
 %	\includegraphics[width=\linewidth]{fig/metrics/chr4.pan.HTTex1.gfa.multiqc_odgi_stats.png}
@@
 \textbf{(d)} \textit{odgi viz} visualization of the 23 largest gene alleles, CHM13, and GRCH38 of the \textit{HTTexon1} graph.
-Figures \textbf{(b)-(d)} clearly highlight the variant region around position 200 showing the variable number of glutamine residues of the different individuals as reported by \cite{Nance1999}.}
+\textbf{(e)} \textit{vg viz} nucleotide-level visualization of 10 gene alleles, CHM13, GRCH38 of the \textit{HTTexon1} graph focusing on the CAG variable repeat region.
+Figures \textbf{(b)-(e)} clearly highlight the variant region around position 200 of CHM13 showing the variable number of glutamine residues of the different individuals as reported by \cite{Nance1999}.}
 	\label{fig:metrics}

### 2021-11-03 scripts/metrics.sh
update metrics script

@@ -1,7 +1,7 @@
 #!/bin/bash
 
-wget http://hypervolu.me/~guarracino/chr4.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
-gunzip chr4.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
+wget https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/scratch/2021_07_30_pggb/chroms/chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.gz
+gunzip chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.gz
 wget https://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/genes/hg38.ncbiRefSeq.gtf.gz
 gunzip hg38.ncbiRefSeq.gtf.gz
 
@@ -9,21 +9,21 @@ grep HTT hg38.ncbiRefSeq.gtf | grep "exon_number \"1\"" | grep "gene_name \"HTT\
 #chr4	ncbiRefSeq.2021-09-09	exon	3074681	3075088	.	+	.	gene_id "HTT"; transcript_id "NM_001388492.1"; exon_number "1"; exon_id "NM_001388492.1.1"; gene_name "HTT";
 #chr4	ncbiRefSeq.2021-09-09	exon	3074681	3075088	.	+	.	gene_id "HTT"; transcript_id "NM_002111.8"; exon_number "1"; exon_id "NM_002111.8.1"; gene_name "HTT";
 
-odgi build -g chr4.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa -o chr4.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.og -t 16 -P
-odgi paths -L -i chr4.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.og > chr4.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.og.paths
+odgi build -g chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa -o chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og -t 16 -P
[…]

### 2021-11-03 manuscript/fig_tips.tex
first half of tips figure

\begin{figure}[ht!]
	\begin{subfigure}{\linewidth}
		\caption{}
		\centering
		% include second image
		\includegraphics[width=1.0\linewidth, trim=0cm 2.75cm 0cm 0cm]{fig/tips/chr8_chm13_beta_defensin_locus_odgi_tips_w50000_karyoploteR}
		\label{fig:tips-karyo}
	\end{subfigure}
	%	\includegraphics[width=\linewidth]{fig/metrics/chr4.pan.HTTex1.gfa.multiqc_odgi_stats.png}
	\caption{Visualizing the contigs of a beta-defensin locus pangenome graph. (\textbf{a}) Breakpoint ranges of the contigs in the beta-defensin locus pangenome graph of 90 haplotypes relative to CHM13. The \textit{odgi tips} results are visualized as the density of BED ranges across the whole \textit{p23.1} cytoband. The high peak at the beginning of \textit{p23.1} indicates that 39 contigs of the graph broke there relative to CHM13. (\textbf{b}) \FIXME{ADD VIZ AS SOON AS THE SORTING IS FINISHED.}}
	\label{fig:tips}
\end{figure}

### 2021-11-04 scripts/chr8.chm13.beta_defensin.tips.R
add odgi viz of p23.1 with caption

@@ -1,0 +1,73 @@
+#!/usr/bin/env Rscript
+
+library(karyoploteR)
+library(BSgenome)
+library(data.table)
+
+args <- commandArgs(trailingOnly = T)
+
+chroms.bed <- args[1]
+chm13.cyto <- args[2]
+beta_in <- args[3]
+pdf_out <- args[4]
+
+chr.start.end <- read.table(chroms.bed, as.is = T, header = F, sep = "\t", check.names = F, quote = "", comment.char = "")
+colnames(chr.start.end) <- c("chr", "start", "end")
+chr.start.end$start <- chr.start.end$start + 1
+
+chr.start.end <- chr.start.end[chr.start.end$chr == "chm13#chr8",]
+
+# chm13.cyto="chm13.CytoBandIdeo.v2.p23.1.txt"
+CYTO_COLORS = getCytobandColors()
+CYTOtmp = toGRanges(fread(chm13.cyto, col.names = c("chr","start","end", "name","gieStain")))
+
+p23.1_start <- start(CYTOtmp)
+p23.1_end <- end(CYTOtmp)
+
+chr.start.end$start <- p23.1_start
+chr.start.end$end <- p23.1_end
+
+CYTOtmp <- GenomicRanges::shift(CYTOtmp, -p23.1_start)
+CYTO_FULL <- c(CYTOtmp)
+
+chroms_split <- strsplit(chr.start.end$chr, "#")
+chroms <- c()
+for (chrom in chroms_split) {
+  chroms <- c(chroms, chrom[[2]])
+}
+chroms <- unique(chroms)
+
+for (chrom in chroms) {
+
+beta <- GRanges(chr.start.end)
+beta <- GenomicRanges::shift(beta, -p23.1_start)
+pp <- getDefaultPlotParams(plot.type=1)
+pp$leftmargin <- 0.15
+pp$data1height <- 500
+pp$data1inmargin <- 25
+pp$data2inmargin <- 0
+pp$ideogramheight <- 100
+beta.bed <- read.table(beta_in, as.is = T, header = F, sep = "\t", check.names = F, quote = "", comment.char = "")
+colnames(beta.bed) <- c("chr", "start", "end", "median_range", "path", "path_pos", "walking_from_front", "jaccard")
+beta.bed$start <- beta.bed$start + 1
+beta.bed <- beta.bed[beta.bed$start >= p23.1_start & beta.bed$end <= p23.1_end,]
+
[…]

### 2021-11-04 manuscript/fig_build_time.tex
passing through

@@
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width=\linewidth]{fig/build/mean_build_time.pdf}
-	\caption{Performance evaluation of \textit{odgi build} when translating a 90 haplotypes human chromosome 6 graph into ODGI's native format.
-The error margins are the standard deviations computed from 10 runs for each of the different threading configurations.}
-	\label{fig:build}
+\begin{figure}[ht!]
+%  \centering
+\floatbox[{\capbeside\thisfloatsetup{capbesideposition={left,top},capbesidewidth=0.45\linewidth}}]{figure}[\FBwidth]
+         {\caption{Performance evaluation of \textit{odgi build} when translating a 90-haplotype graph of human chromosome 6 into ODGI's native format.
+             Build time decreases as parallelism increases, but with diminishing per-thread processing rates.
+           }\label{fig:build}}
+         {\includegraphics[width=\linewidth]{fig/build/mean_build_time.pdf}} %\includegraphics[width=5cm]{name}}
+	     %\includegraphics[width=\linewidth]{fig/build/mean_build_time.pdf}
+	     %\caption{Performance evaluation of \textit{odgi build} when translating a 90-haplotype graph of human chromosome 6 into ODGI's native format.}
+
 \end{figure}

### 2021-11-04 manuscript/fig_extract_viz_draw_position_untangle.tex
remove '.' from figure names
passing through

@@
             \begin{center}
-                \includegraphics[width=\linewidth, trim=0cm 0.9cm -2.5cm 0.9cm]{fig/extract_viz_draw_position_untangle/chr6.C4.sorted.untangle.bed.subset.tsv}
+                \includegraphics[width=\linewidth, trim=0cm 0.9cm -2.5cm 0.9cm]{fig/extract_viz_draw_position_untangle/chr6_C4_sorted_untangle_bed_subset_tsv.pdf}
             \end{center}
@@
         \textbf{(c)} Color by path position: the color gradients are smooth, highlighting that its node are well sorted in 1 dimension.
-        The top two reference genomes and 2 haplotypes (with name starting with HG01952\#2) go from left to right, while 5 haplotypes go in the opposite direction, as indicated by the black color on their left.
+        The top two reference genomes and 2 haplotypes (HG01952\#2) go from left to right, while 5 haplotypes go in the opposite direction, as indicated by the black color on their left.
         \textbf{(d)} \textit{odgi viz} color by strandness: the red paths indicate the haplotypes that were assembled in reverse with respect to the 2 reference genomes.
@@
         The entirely grey paths have one copy of these genes.
-        The path with name starting with HG01071\#2 presents 3 copies of the genes (indicated by the orange color), of which only 1 with the HERV sequence (gray color in the middle of the orange region).
-        In the haplotype with the name starting with HG01952\#1, the HERV sequence is absent.
+        HG01071\#2 presents 3 copies of locus (orange), of which one contains the HERV sequence (gray in the middle of the orange).
+        In HG01952\#1, the HERV sequence is absent.
         \textbf{(f)} \textit{Bandage} layout of the C4 pangenome graph, annotated by using \textit{odgi position}.
@@
 Node labels were annoted by using \textit{odgi position}.
-        \textbf{(h)} Visualization of \textit{odgi untangle} output in the C4 pangenome graph: the plots show the copy number status of the sequences in the C4 region with respect to the grch38 reference sequence, making clear, for example, that in the haplotype with name starting with HG00438\#2, \textit{C4B} gene is missing.
+        \textbf{(h)} Visualization of \textit{odgi untangle} output in the C4 pangenome graph: the plots show the copy number status of the sequences in the C4 region with respect to the grch38 reference sequence, making clear, for example, that in HG00438\#2, \textit{C4B} gene is missing.
     }

### 2021-11-04 manuscript/main.tex
first half of tips figure
passing through

@@
 \usepackage{comment}
+
+\usepackage{floatrow}
 
@@
 ODGI improves on prior efforts, based on issues that arose during our work with high-quality \textit{de novo} assemblies that cover almost all parts of the human genome~\citep{Logsdon_2021,Nurk_2021}.
-For example, we find that it is necessary to support graphs that have regions where paths present very high depth nodes.
+For example, we find that it is necessary to support graphs with regions with very high numbers of path traversals (high path depth).
 Such motifs can occur in collapsed repeat structures generated by ambiguous sequence homology relationships in repeats found in the centromeres and other segmental duplications.
@@
 Most read or write the native ODGI format (`og' extension) (Figure~\ref{fig:operations}) and work with standard text based data formats common to bioinformatics.
-This supports the implementation of flexible and composable graph processing pipelines based on a common data type.
-While the pangenome graph is a first-class entity in ODGI, we focus on the embedded paths to provide a universal coordinate system.
-Despite being recent systems, pangenome graphs are not incompatible with current standard ways of working with bioinformatics data.
-ODGI provides a set of interfaces that let us approach these graphs from the perspective of standard reference- and sequence-based data models.
+This supports the implementation of flexible and composable graph processing pipelines based on graphs (GFA/ODGI) and standard bioinformatic data types representing positions, genomic ranges (BED), and pairwise mappings (PAF).
+%While the pangenome graph is a first-class entity in ODGI,
+% develop a universal, stable, coordinate system, rendering these data types relative to reference and query sequences embedded in the graph.
+%To develop a stable and compatible coordinate system, 
+We use variation graph paths to provide a universal coordinate system, representing annotations and pairwise sequence relationships using the paths as reference and query sequences.
+%This avoids the complexity of establishing a stable coordinate system over the topology of the graph, which may change when it is modified or subset.
+%Although they present new complexities, we find that pangenome graphs are compatible with current standard ways of working with bioinformatics data.
+ODGI thus provides a set of interfaces that let us approach these graphs from the perspective of standard reference- and sequence-based data models.
 Indeed, by considering all paths in the graph as potential reference or query sequence, we make graphs invisible to downstream tools that operate on collections of sequences or rely on a reference sequence (\textit{e.g.} SAMtools~\citep{Li2009}), enabling interoperability.
@@
 \label{sec:build}
-ODGI maintains its own efficient ODGI binary format for storing graphs on disk.
+ODGI maintains its own efficient binary format for storing graphs on disk.
 We begin by transforming the storage model of the standard GFAv1~\citep{GFA} format (in which nodes, edges, and paths are described independently) into the ODGI node-centric encoding with \textit{odgi build}.
@@
 
-For interchange with other tools, \textit{odgi view} can convert a graph in ODGI binary format to the standard GFAv1.
+To support interchange with other pangenome tools or text-based processing, \textit{odgi view} converts a graph in ODGI binary format to GFAv1.
+%ODGI uses the ``S'' and ``L'' records to describe the graph topology, and ``P'' records to describe paths.
+
+\input{fig_build_time}
 
@@
 
-Pangenome graph visualization is one of the first steps to gain insight into the mutual relationship between the sequences and their variation.
-We pursue a novel approach to visualization with \textit{odgi viz} and \textit{odgi draw}, two tools which provide scalable ways of generating pictures of the high-level structure of the graph.
+Pangenome graph visualization is one of the first steps to gain insight into the mutual relationship between the sequences in the graph and their variation.
+We pursue a novel approach to visualization with \textit{odgi viz} and \textit{odgi draw}, two tools which provide scalable ways of generating pictures of the high-level structure of large pangenome graphs.
 
@@
 The meaning of the colors depends on how \textit{odgi viz} is executed.
-By default, path colors has no particular meaning, but each path has its own color (Fig.~\ref{fig:odgi_viz}\textbf{b}).
+By default, path colors are derived from a hash of the path name (Fig.~\ref{fig:odgi_viz}\textbf{b}).
 Path names are displayed on the left of the paths.
@@
 Nevertheless, complex, nonlinear graph structures are difficult to display and interpret in a low number of dimensions.
-To overcome such a limitation, \textit{odgi viz} supports multiple visualization modalities (Fig.~\ref{fig:odgi_viz}\textbf{c-e}), making it easy to grasp the properties of the graph and its complexity.
-Graph nodes order can affects downstream analyses on pangenome graphs.
-With \textit{odgi viz} we can color the paths by path position (Fig.~\ref{fig:odgi_viz}\textbf{c}), with light grey indicating where paths begin and dark grey where they ends.
-This visualization is suitable for checking graph nodes' linear order, as smooth color gradients indicate that the graph reflects well the linear paths' coordinate system.
-Pangenome graphs represents both strand of DNA sequences.
-\textit{odgi viz} supports also coloring the paths by orientation, with paths colored in red and black where their sequence is reversed or in forward, respectively, with respect to the sequences represented by the graph nodes (Fig.~\ref{fig:odgi_viz}\textbf{d}).
+To overcome such a limitation, \textit{odgi viz} supports multiple visualization modalities (Fig.~\ref{fig:odgi_viz}\textbf{c-e}), making it easy to grasp the properties and shape of the graph.
+Graph node order can affect downstream analyses on pangenome graphs.
+With \textit{odgi viz} we can color the paths by path position (Fig.~\ref{fig:odgi_viz}\textbf{c}), with light grey indicating where paths begin and dark grey where they end.
+This visualization is suitable for understanding graph node order, as smooth color gradients indicate that the graph order respects the linear paths' coordinate systems.
+Pangenome graphs represent both strands of DNA sequences.
+\textit{odgi viz} supports also coloring the paths by orientation, with paths colored where their sequence is reverse-complemented (red) or in direct orientation (black) with respect to the sequences of the graph nodes (Fig.~\ref{fig:odgi_viz}\textbf{d}).
 Eukaryotic genomes experience gains and losses of genetic material, resulting in copy number variation (CNV) across the population.
@@
 \textit{odgi draw} extends the visualization in 2 dimensions (2D)  (Fig.~\ref{fig:odgi_viz}\textbf{a}) by rendering the layout built by \textit{odgi layout} (\S\ref{sec:sort}).
-A 2D rendering is more costly to compute, but we similarly provide a linear-time scaling implementation, allowing us to apply it to large pangenome graphs.
+A 2D rendering is more costly to compute, but we similarly provide an implementation that scales linearly with pangenome sequence size, allowing us to apply it to large pangenome graphs.
 
@@
 Finally, the portions of the paths (\textit{i.e.}, the subpaths) walking through the selected nodes are extracted and added to the new subgraph.
-Subpaths are searched in parallel, created serially, and extended in parallel again thanks to the parallelism enabled by the ODGI data structure (see \S\ref{sec:build}), making \textit{odgi extract} a scalable solution to extract also complex subregions presenting nodes with very high depth.
+Subpaths are searched in parallel, created serially, and extended in parallel again thanks to the parallelism enabled by the ODGI data structure (see \S\ref{sec:build}), making \textit{odgi extract} a scalable solution to extract also complex subregions presenting nodes with very high path depth.
 
@@
 
-For a precise translation process when conversing a query position to a reference position in a repeat region the \textit{path jaccard} concept is applied: It could be that the found reference node is visited several times by the reference.
-To ensure a precise translation, the best jaccard for the query step and each reference candidate step is calculated.
+For a precise translation process when conversing a query position to a reference position in a repeat region, we apply the \textit{path jaccard} context mapping concept.
+It could be that the found reference node is visited several times by the reference.
+To ensure a precise translation, we select the reference position whose context (the multiset of $Node.id$s reached within a distance of e.g.
+10kbp) has the best jaccard metric when compared to the query context.
+%the best jaccard for the query step and each reference candidate step is calculated by looking at a context range (e.g. 10kbp) around the locus.
 
@@
 Pangenome graphs model alignments of many genomes.
-With \textit{odgi untangle}, users can extract pairwise \textit{1}:\textit{many} alignment information between a given set of ``query'' sequences and a given set of ``target'' sequences (used as references).
-This lets us decompose both homology and paralogy relationships by also untangling loopy motifs in the graph, and obtaining linearized pairs of regions in BEDPE or PAF format.
+With \textit{odgi untangle}, users can extract pairwise \textit{many}:\textit{1} alignment information between a given set of ``query'' sequences and a given set of ``target'' sequences (used as references).
+This lets us decompose both homology and paralogy relationships by also untangling looping motifs in the graph, and obtaining linearized pairs of regions in BEDPE or PAF format.
 Being able to work with any sets of reference sequences lets us convert the graph to lift-over maps compatible with standard software for projecting annotations and alignments from one genome to another.
-As an example, by untangling the graph we can study the variation that lies in the collapsed regions due to sequence repeats (as shown in Fig.~\ref{fig:odgi_viz}\textbf{f}).
+As an example, by untangling the graph we can study the variation that lies in regions collapsed due to ambiguous alignments over sequence repeats (as shown in Fig.~\ref{fig:odgi_viz}\textbf{f}).
 Indeed, to obtain a more precise overview of the \textit{locus} in Fig.~\ref{fig:odgi_viz}\textbf{b-e},
 we can apply \textit{odgi untangle} to segment paths into linear segments by breaking these segments where the paths loop back on themselves.
-The \textit{path jaccard} concept ensures that the breaking is nucleotide-precise.
+We ``untangle'' by finding the target segment that best match each query segment using the \textit{path jaccard} context mapping model.
+%The \textit{path jaccard} concept ensures that the breaking is nucleotide-precise.
 In this way, we obtain information on the position and copy number status of the sequences in the collapsed \textit{locus} (Fig.~\ref{fig:odgi_viz}\textbf{h}).
@@
 
-Having build a pangenome graph from assemblies in contig state, we can ask questions like: Where do our contigs' ends match the reference(s)?
-An answer to this question can highlight regions that are tricky to assemble~\ref{fig:odgi_tips}.
+%Having built a pangenome graph from assembly contigs, we can ask questions like: Where do our contigs' ends match the reference(s)?
+An answer to this question can highlight regions that are tricky to assemble (Fig.
+~\ref{fig:tips}).
 \textit{odgi tips} can identify the break point positions of the contigs relative to the reference(s) in the graph by walking from the ends of each contig until a reference node is found. 
@@
 
-\FIXME{add odgi tips figure: only focus on the beta-defensin cluster [p23.3-p22] or more?}
-
 \subsection{Sorting the pangenome graph}
@@
 
-Pangenome graphs can hide their underlying latent structures, introducing difficulties in the analysis and interpretation (Fig.~\ref{fig:sorting}\textbf{a}).
+Pangenome graphs can hide their underlying latent structures, introducing difficulties in the analysis and interpretation. % (Fig.~\ref{fig:sorting}\textbf{a}).
 Among the causes of this is the correct ordering of the graph nodes in a convenient number of dimensions.
@@
 
-To find the best order of graph nodes in 1D, \textit{odgi sort} provides multiple sorting algorithms which can be combined into a series of sorts to take advantage of the strength of each.
-Most notably, nodes can be sorted topologically, randomly, by breaking cycles in the graph (\S\ref{sec:edit}), by grooming (\S\ref{sec:edit}), and/or by using a novel path-guided (PG) stochastic gradient descent (SGD) algorithm, also called PG-SGD.
-The latter one exploits the biological information in the paths for sorting the nodes.
+To find the best order of graph nodes in 1D, \textit{odgi sort} provides multiple sorting algorithms which can be combined into a sorting pipeline to take advantage of the strength of each.
+Most notably, nodes can be sorted topologically, randomly, by breaking cycles in the graph (\S\ref{sec:edit}), by grooming (\S\ref{sec:edit}), and/or by using a novel path-guided (PG) stochastic gradient descent (SGD) algorithm: PG-SGD.
+This exploits path information to order the graph nodes.
 %For understanding the large-scale organization of the graph, we benefit from an even simpler coordinate system.
-In particular, to provide a universal coordinate space for the graph, we project its nodes into a low-dimensional space
-by learning the projection via a HOGWILD! approach~\citep{niu2011hogwild} implemented in the PG-SGD algorithm.
-The PG-SGD adapts a SGD-based drawing to pangenome graphs~\citep{zheng2018graph}.
-This approach updates node relative positions to best-match their nucleotide distance in the paths running through the graph.
+PG-SGD learns a 1D or 2D organization of the graph nodes that matches distances in graph paths.
+To scale to large graphs, we learn this projection in parallel via a HOGWILD! approach~\citep{niu2011hogwild}.
+Our approach can be seen as an adaptotion of SGD-based drawing to pangenome graphs~\citep{zheng2018graph}.
+In parallel, each HOGWILD! thread updates node relative positions to best-match their nucleotide distance in the paths running through the graph.
+Following standard SGD approaches, a learning rate is reduced as the algorithm progresses, and execution continues until the adjustments to the model fall below a target threshold $\epsilon$.
 ODGI can project vector (in 1D) and matrix (2D) representations of the graph relative to these learned coordinate spaces.
-Based on this projection, we can trivially sort graph nodes in 1D. Moreover, we support the same concept in 2D in \textit{odgi layout}
-by providing a 2D implementation of the PG-SGD algorithm.
-As the PG-SGD sorts graph nodes with respect to the paths which walk through them, we can also detect regions whose layout is distorted, which may represent structural variation or assembly error.
+Based on this projection, we can trivially sort graph nodes in 1D.
+Moreover, we support the same concept in 2D in \textit{odgi layout} by providing a 2D implementation of the PG-SGD algorithm.
+%As the PG-SGD sorts graph nodes with respect to the paths which walk through them, we can also detect regions whose layout is distorted, which may represent structural variation or assembly error.
+
+\input{fig_metrics}
 
@@
 Pangenome graph topology can be derived by applying \textit{odgi matrix}, obtaining information on graph nodes connections in textual sparse matrix format.
-To investigate on the genomes encoded in the graph, \textit{odgi paths} allows users to calculate pairwise overlap statistics of groupings of paths and emit all path sequences in FASTA format.
-Furthermore, \textit{odgi flatten} generate a linearization of the graph by emitting the pangenome sequence (the concatenation of all graph node sequences) in FASTA format, and the projection of all paths on the linearized sequence in BED format.
+To investigate on the genomes encoded in the graph, \textit{odgi paths} allows users to calculate pairwise overlap statistics of groupings of paths and emit all path sequences in FASTA format, and it also allows the generation of a ``pangenome matrix'' that reports the copy number (presence/absence) of each path over each node.
+\textit{odgi flatten} generate a linearization of the graph by emitting the pangenome sequence (the concatenation of all graph node sequences) in FASTA format, and the projection of all paths on the linearized sequence in BED format.
 
@@
 
-\input{fig_build_time}
-
 \input{fig_extract_viz_draw_position_untangle}
 
-\input{fig_sorting}
+%\input{fig_tips}
 
-\input{fig_metrics}
+%\input{fig_sorting}
 
@@
 \section{Discussion}
-Pangenome graphs stand to become a ubiquitous model in genomics thanks to their capability to represent any genetic variant without being affected by the reference-bias~\citep{Eizenga_2020}.
+Pangenome graphs stand to become a ubiquitous model in genomics thanks to their capability to represent any genetic variant without being affected by reference bias~\citep{Eizenga_2020}.
 However, despite this great potential, their spread is impeded by the lack of tools capable of managing and analyzing pangenome graphs easily and efficiently.
 
-With ODGI we implemented a state-of-the-art tool suite that enables users to explore and discover the underlying biology in pangenomes graphs, filling the gap that made pangenomic analyses difficult.
-As a result, we provides tools to easily transform, analyse, simplify, validate, and visualise pangenome graphs at large scale.
+ODGI is a state-of-the-art tool suite that enables users to explore and discover the underlying biology in pangenomes graphs, filling the gap that made pangenomic analyses difficult.
+It provides tools to easily transform, analyse, simplify, validate, and visualise pangenome graphs at large scale.
 In particular, lifting over annotations and linearizing nested graph structures place the suite as the bridge between traditional linear reference genome analysis and pangenome graphs.
@@
 Particularly for targets that involve complex variation, such as cancer, plant genomics and metagenomics, ODGI will facilitate disentangling,
-describing and analysing a much larger set of variation than previously was possible with tools that depend on short reads and reference genomes,.
+describing and analysing a much larger set of variation than previously was possible with tools that depend on short reads and reference genomes.
 Futhermore, users can even consider ODGI as a framework, taking advantage of its algorithms to develop new and more advanced tools that work on pangenome graphs,
@@
 
-ODGI is already the backbone of the pipelines such as the Pangenome Graph Builder~\citep{pggb} or nf-core/pangenome~\citep{pangenome}, where it is also applied for debugging purposes.
+ODGI is already the backbone of the pipelines such as the Pangenome Graph Builder~\citep{pggb} or nf-core/pangenome~\citep{pangenome}.
+%, where it is also applied for debugging purposes.
 %\footnote{\url{https://github.com/nf-core/pangenome} (accessed Oct 2021)}.
-Indeed, its static, large-scale 1D and 2D visualizations of the pangenome graphs allow grasping the underlying structure and any problems in the graph construction.
+Its static, large-scale 1D and 2D visualizations of the pangenome graphs allow an unprecedented high-level perspective on variation in pangenomes, and have also been critical in the development of pangenome graph building methods.
+%the underlying structure and any problems in graph construction.
 However, an interactive solution that combines the 1D and 2D layout of a graph with annotation and read mapping information across different zoom levels is still missing.
-Recent interactive browsers are reference centric~\citep{Beyer2019, Yokoyama2019, Durant2021, Liang2021} or focusing on 1D~\citep{Wick_2015, Gonnella2018}.
+Recent interactive browsers are reference centric~\citep{Beyer2019, Yokoyama2019, Durant2021, Liang2021} or focusing on 2D~\citep{Wick_2015, Gonnella2018}.
 In ODGI, we provide a graph layout algorithm that can become a backbone of such tools, as is already the case for \textit{gfaestus}~\citep{gfaestus} %\footnote{\url{https://github.com/chfi/gfaestus}},

### 2021-11-04 scripts/time.R

@@ -1,6 +1,8 @@
 #!/usr/bin/env Rscript
 
-library(ggplot2)
+library(tidyverse)
+library(ggbeeswarm)
+library(knitr)
 
 args <- commandArgs(trailingOnly = T)
 
@@ -9,45 +11,17 @@ times_pdf <- args[2]
 
 # times_csv <- "/home/heumos/Downloads/odgi_paper/MHC/mean_build_time.csv"
 
-times <- read.table(times_csv, sep = ",", quote = "", as.is = T, 
-                    check.names = F, header = F, comment.char = "")
-colnames(times) <- c("time", "threads")
+times <- read.delim(times_csv)
 
-times$threads <- as.character(times$threads)
+summary(times)
+
 times$threads <- factor(times$threads, levels=unique(times$threads))
 
-#+++++++++++++++++++++++++
-# Function to calculate the mean and the standard deviation
-# for each group
-#+++++++++++++++++++++++++
-# data : a data frame
-# varname : the name of a column containing the variable
-#to be summariezed
-# groupnames : vector of column names to be used as
-# grouping variables
-data_summary <- function(data, varname, groupnames){
-  require(plyr)
-  summary_func <- function(x, col){
-    c(mean = mean(x[[col]], na.rm=TRUE),
-      sd = sd(x[[col]], na.rm=TRUE))
-  }
-  data_sum<-ddply(data, groupnames, .fun=summary_func,
-                  varname)
-  data_sum <- rename(data_sum, c("mean" = varname))
-  return(data_sum)
-}
+p <- ggplot(times, aes(x=threads, y=time, group=threads)) +
+    geom_violin() +
+    #geom_beeswarm(size=0.5) +
+    labs(x = "number of threads", y = "time in seconds")
+#pdf(times_pdf, height = 2.5, width = 3)
+ggsave(times_pdf, device = cairo_pdf, height = 2.5, width = 3)
+knitr::plot_crop(times_pdf)
 
-df2 <- data_summary(times, varname="time", 
-                    groupnames=c("threads"))
-df2$threads=factor(df2$threads, levels=unique(times$threads))
-
-p <- ggplot(data=df2, aes(x=threads, y=time, color=threads, group=1)) +
-  geom_errorbar(aes(ymin=time-sd, ymax=time+sd), width=.2) +
-  geom_line(color = "black") +
-  geom_point(size = 1) + 
-  # geom_bar(stat="identity") +
[…]

### 2021-11-04 manuscript/fig_tips.tex

@@
 		% include second image
-		\includegraphics[width=1.0\linewidth, trim=0cm 2.75cm 0cm 0cm]{fig/tips/chr8_chm13_beta_defensin_locus_odgi_tips_w50000_karyoploteR}
+		\includegraphics[width=1.0\linewidth, trim=0cm 2cm 0cm 2cm]{fig/tips/chr8_chm13_beta_defensin_locus_odgi_tips_w50000_karyoploteR}
 		\label{fig:tips-karyo}
+	\end{subfigure}
+	\begin{subfigure}{\linewidth}
+	\caption{}
+	\centering
+	% include second image
+	\includegraphics[width=1.0\linewidth, trim=-10cm 6cm -3.4cm 3cm]{fig/tips/chr8_pan_fa_c3d3224_7748b33.395c7f4_smooth_og_Y_og_p23_1_og_90paths_m}
+	\label{fig:tips-viz}
 \end{subfigure}
@@
 The \textit{odgi tips} results are visualized as the density of BED ranges across the whole \textit{p23.1} cytoband.
-The high peak at the beginning of \textit{p23.1} indicates that 39 contigs of the graph broke there relative to CHM13.
-(\textbf{b}) \FIXME{ADD VIZ AS SOON AS THE SORTING IS FINISHED.}}
+The high peak at the beginning of \textit{p23.1} indicates that 39 contigs of the graph broke at this specific position relative to CHM13.
+(\textbf{b}) \textit{odgi viz} color by path depth visualization of the 90 longest paths in the beta-defensin locus pangenome graph including CHM13 and GRCH38 on the top: using the Spectra color palette with 4 levels of path depths.
+White, grey, red, and yellow indicate a path depth of 0, 1, 2, and greater than or equal 3, respectively.
+Coloring by path depth we can see a very high path depth region on the right part of the  \textit{p23.1}.
+In this complex region, lot's of contigs break as can be seen in (\textbf{a}).
+There is no high path depth in the left part of the \textit{p23.1} where a huge number of contigs break at one specific reference position.
+This could be due to the fact that during graph construction, very small repeats are resolved instead of collapsed, and are therefore not highlighted in the visualization.}
 	\label{fig:tips}

### 2021-11-04 manuscript/main.tex
data

@@
 
-\FIXME{Data used to build Human pangenome graphs is available at \url{https://github.com/human-pangenomics/HPP_Year1_Data_Freeze_v1.0}.}
-Code used to build images in this manuscript is available at \url{https://github.com/pangenome/odgi-paper}.
-TODO!
+Data used to build human pangenome graphs: \url{https://github.com/human-pangenomics/HPP_Year1_Data_Freeze_v1.0}.
+Code used to build this manuscript and its figures: \url{https://github.com/pangenome/odgi-paper}.
 

### 2021-11-04 manuscript/main.tex
tweak the discussion

@@
 
-ODGI is already the backbone of the pipelines such as the Pangenome Graph Builder~\citep{pggb} or nf-core/pangenome~\citep{pangenome}.
+ODGI is already the backbone of the pipelines such as the Pangenome Graph Builder~\citep{pggb}.
+% or nf-core/pangenome~\citep{pangenome}.
 %, where it is also applied for debugging purposes.
@@
 However, an interactive solution that combines the 1D and 2D layout of a graph with annotation and read mapping information across different zoom levels is still missing.
-Recent interactive browsers are reference centric~\citep{Beyer2019, Yokoyama2019, Durant2021, Liang2021} or focusing on 2D~\citep{Wick_2015, Gonnella2018}.
-In ODGI, we provide a graph layout algorithm that can become a backbone of such tools, as is already the case for \textit{gfaestus}~\citep{gfaestus} %\footnote{\url{https://github.com/chfi/gfaestus}},
-which currently relies on the \textit{odgi layout} output for its interactive 2D graph visualization.
+Recent interactive browsers are reference centric~\citep{Beyer2019, Yokoyama2019, Durant2021, Liang2021} or focus primarily on 2D~\citep{Wick_2015, Gonnella2018}.
+Our graph sorting and layout algorithms can provide the foundation for future tools of this type.
+We plan to focus on using these learned models to detect structural variation and assembly errors.
+%In ODGI, we provide a graph layout algorithm that can become a backbone of such tools, as is already the case for \textit{gfaestus}~\citep{gfaestus} %\footnote{\url{https://github.com/chfi/gfaestus}},
+%which currently relies on the \textit{odgi layout} output for its interactive 2D graph visualization.
 
-ODGI aims to be a reference implementation for many pangenome related operations, thus future work will be focused on multiple directions.
-First, we will expand its metadata capabilities to annotate large pangenome graphs.
-Futhermore, we will improve RDF support for annotation in the in-memory graph~\citep{Yokoyama2020} and allow for federated queries on pangenome graphs.
-Finally, we will explore the addition of support for RNA sequences and proteins to extend the application of pangenomic approaches to other levels in the genetic information flow from DNA to phenotype.
+ODGI has allowed us to explore \textit{context mapping} via the path jaccard metric.
+This resolves a major conceptual issue that has strongly guided existing algorithms to construct pangenome graphs \citep{Li:2020}.
+Previously, great efforts have been made to prevent ``collapse'' of non-orthologous sequences in the graph topology itself.
+While our presentation is primarily qualitative, our work demonstrates that we can completely mitigate this issue by exploiting the pangenome graph not as a static reference, but as a dynamic model of the mutual alignment of many genomes.
+%Our future work will focus on further extensions of this topic.
+
+%Here, we present only qualitative evaluation of this result, but our future work will focus on exploiting this mechanism to simplify and aid the interpretation of large pangenomes.
+
+%ODGI aims to be a reference implementation for many pangenome related operations, thus future work will be focused on multiple directions.
+%First, we will expand its metadata capabilities to annotate large pangenome graphs.
+%Futhermore, we will improve RDF support for annotation in the in-memory graph~\citep{Yokoyama2020} and allow for federated queries on pangenome graphs.
+%Finally, we will explore the addition of support for RNA sequences and proteins to extend the application of pangenomic approaches to other levels in the genetic information flow from DNA to phenotype.
 

### 2021-11-04 manuscript/fig_metrics.tex
first half of tips figure
redo metrics figures with wgg.87

@@
-\begin{figure}[h!]
+\begin{figure}[ht!]
 	\begin{subfigure}{\linewidth}
@@
 		% include first image
-		\includegraphics[width=1.0\linewidth, trim=-1.25cm 3cm 0 1.75cm]{fig/metrics/chr4_pan_HTTex1_gfa_multiqc_odgi_stats}
+		\includegraphics[width=1.0\linewidth, trim=-2.25cm 1cm 0cm 2.75cm]{fig/metrics/chr4_pan_HTTex1_gfa_multiqc_odgi_stats}
 		\label{fig:metrics-multiqc}
@@
 	% include second image
-	\includegraphics[width=1.0\linewidth, trim=1.25cm 0.5cm 1.5cm 0cm]{fig/metrics/chr4_pan_HTTex1_STR_xg_svg.pdf}
+	\includegraphics[width=1.0\linewidth, trim=-0.4cm 0.1cm 3.15cm 0cm, clip]{fig/metrics/chr4_pan_HTTex1_STR_xg_svg.pdf}
 	\label{fig:metrics-str}

### 2021-11-04 scripts/build_time.sh

@@ -1,7 +1,7 @@
-wget http://hypervolu.me/~guarracino/chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.gz
-gunzip chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.gz
+wget https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/scratch/2021_11_04_pggb_wgg.87/chroms/chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
+gunzip chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
 
-G=chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa
+G=chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa
 
 # t1
 for i in {1..10}

### 2021-11-04 scripts/metrics.sh

@@ -1,29 +1,31 @@
 #!/bin/bash
 
-wget https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/scratch/2021_07_30_pggb/chroms/chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.gz
-gunzip chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.gz
+wget https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/scratch/2021_11_04_pggb_wgg.87/chroms/chr4.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
+gunzip chr4.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
 wget https://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/genes/hg38.ncbiRefSeq.gtf.gz
 gunzip hg38.ncbiRefSeq.gtf.gz
+
+G=chr4.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa
 
 grep HTT hg38.ncbiRefSeq.gtf | grep "exon_number \"1\"" | grep "gene_name \"HTT\"" | grep -P "\texon"
 #chr4	ncbiRefSeq.2021-09-09	exon	3074681	3075088	.	+	.	gene_id "HTT"; transcript_id "NM_001388492.1"; exon_number "1"; exon_id "NM_001388492.1.1"; gene_name "HTT";
 #chr4	ncbiRefSeq.2021-09-09	exon	3074681	3075088	.	+	.	gene_id "HTT"; transcript_id "NM_002111.8"; exon_number "1"; exon_id "NM_002111.8.1"; gene_name "HTT";
 
-odgi build -g chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa -o chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og -t 16 -P
-odgi paths -L -i chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og > chr4.pan.fa.a2fb268.e820cd3.9ea71d8.smooth.gfa.og.paths
+odgi build -g "$G" -o "$G".og -t 16 -P
+odgi paths -L -i "$G".og > "$G".og.paths
 
[…]

### 2021-11-04 scripts/metrics.sh
pdferize all metrics figures

@@ -33,6 +33,7 @@ multiqc .
 sed -i 's/"smooth">smooth/"chr4.pan.HTTex1.gfa">chr4.pan.HTTex1.gfa/g' multiqc_report.html
 #### MANUAL INTERVENTION ####
 # A manual screenshot only from the Detailed ODGI stats table. Zoom in Firefox: 200%.
+inkscape chr4_pan_HTTex1_gfa_multiqc_odgi_stats.png.svg --export-pdf=chr4_pan_HTTex1_gfa_multiqc_odgi_stats_svg.pdf
 
 odgi paths -i "$G".og.HTTex1.og.O.og -L > "$G".og.HTTex1.og.O.og.paths
 # we accidentally cut off stuff from chm13 and grch13 

### 2021-11-04 scripts/mhc.sh

@@ -1,11 +1,11 @@
-wget http://hypervolu.me/~guarracino/chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.gz
-gunzip chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.gz
+wget https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/scratch/2021_11_04_pggb_wgg.87/chroms/chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
+gunzip chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
 
-MHC=chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.og
-odgi build -g chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa -o "$MHC" -t 28 -P
+MHC=chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.og
+odgi build -g chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa -o "$MHC" -t 28 -P
 odgi extract -i "$MHC" -o "$MHC".mhc.og -E -P -t 28 -r grch38#chr6:29000000-34000000
 odgi sort -i "$MHC".mhc.og -o "$MHC".mhc.og.O.og -O
-odgi layout -i "$MHC".mhc.og.O.og -o "$MHC".mhc.og.O.og.lay -T "$MHC".mhc.og.O.og.lay.tsv -P -t 28 -G 20 -x 100
+odgi layout -i "$MHC".mhc.og.O.og -o "$MHC".mhc.og.O.og.lay -T "$MHC".mhc.og.O.og.lay.tsv -P -t 28 -x 100
 odgi draw -i chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.og.mhc.og.O.og -c chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.og.mhc.og.O.og.lay -p chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.og.mhc.og.O.og.lay.H1000w10.png -H 1000 -C -w 10 -t 28 -P
 
 #### MANUAL INTERVENTION ####

### 2021-11-04 manuscript/fig_extract_viz_draw_position_untangle.tex
simpler
there is no place to hide

@@
                 \caption{}
-                \includegraphics[width=\linewidth, trim=0 +1cm -0.5cm 0.5cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted}
+                \includegraphics[width=\linewidth, trim=0 2cm -0.5cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted}
             \end{subfigure} \\
@@
                 \caption{}
-                \includegraphics[width=\linewidth, trim=0 +1cm -0.5cm 0.5cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_du}
+                \includegraphics[width=\linewidth, trim=0 +2cm -0.5cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_du}
             \end{subfigure} \\
@@
                 \caption{}
-                \includegraphics[width=\linewidth, trim=0 +1cm -0.5cm 0.5cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_z}
+                \includegraphics[width=\linewidth, trim=0 +2cm -0.5cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_z}
             \end{subfigure} \\
@@
                 \caption{}
-                \includegraphics[width=\linewidth, trim=0 +1cm -0.5cm 0.5cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_m}
+                \includegraphics[width=\linewidth, trim=0 +2cm -0.5cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_m}
             \end{subfigure}
@@
             \begin{center}
-                \includegraphics[width=\linewidth, trim=0cm 0.9cm -2.5cm 0.9cm]{fig/extract_viz_draw_position_untangle/chr6_C4_sorted_untangle_bed_subset_tsv.pdf}
+                \includegraphics[width=\linewidth, trim=0cm 0.9cm -2.5cm 0.5cm]{fig/extract_viz_draw_position_untangle/chr6_C4_sorted_untangle_bed_subset_tsv.pdf}
             \end{center}
@@
         Moreover, \textit{C4A} and \textit{C4B} genes segregate in both long and short genomic forms, distinguished by the presence or absence of a human endogenous retroviral (HERV) sequence, as also highlighted by the short nested link on the left.
-        \textbf{(c)} Color by path position: the color gradients are smooth, highlighting that its node are well sorted in 1 dimension.
+        \textbf{(c)} Color by path position. %: the color gradients are smooth, highlighting that its node are well sorted in 1 dimension.
         The top two reference genomes and 2 haplotypes (HG01952\#2) go from left to right, while 5 haplotypes go in the opposite direction, as indicated by the black color on their left.

### 2021-11-04 manuscript/fig_extract_viz_draw_position_untangle.tex
mhc in fig4, correct fig3d

@@
             \caption{}
-            \includegraphics[width=\linewidth, trim=0cm 0cm -10cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_og_mhc_og_O_og_lay_H1000w100_png_svg}
+            \includegraphics[width=\linewidth, trim=0cm 0cm -10cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_gfa_og_mhc_og_O_og_lay_H1000w100_png_svg}
         \end{subfigure} \\

### 2021-11-04 manuscript/fig_extract_viz_draw_position_untangle.tex
update and squeeze

@@
         %\hline
-        \begin{subfigure}[t]{1.0\textwidth}
+        \begin{subfigure}[t]{\textwidth}
             \centering
             \caption{}
-            \includegraphics[width=\linewidth, trim=0cm 0cm -10cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_gfa_og_mhc_og_O_og_lay_H1000w100_png_svg}
+            \includegraphics[width=\linewidth, trim=0cm 10cm -18cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_og_mhc_og_O_og_lay_H1000w100_png_svg}
         \end{subfigure} \\
@@
                 \caption{}
-                \includegraphics[width=\linewidth, trim=0 2cm -0.5cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted}
+                \includegraphics[width=\linewidth, trim=0 1.7cm 0cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_C4_sorted}
             \end{subfigure} \\
@@
                 \caption{}
-                \includegraphics[width=\linewidth, trim=0 +2cm -0.5cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_du}
+                \includegraphics[width=\linewidth, trim=0 +1.7cm 0cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_C4_sorted_du}
             \end{subfigure} \\
@@
                 \caption{}
-                \includegraphics[width=\linewidth, trim=0 +2cm -0.5cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_z}
+                \includegraphics[width=\linewidth, trim=0 +1.7cm 0cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_C4_sorted_z}
             \end{subfigure} \\
@@
                 \caption{}
-                \includegraphics[width=\linewidth, trim=0 +2cm -0.5cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_m}
+                \includegraphics[width=\linewidth, trim=0 +1.7cm 0cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_C4_sorted_m}
             \end{subfigure}
@@
             %\smallskip
+            \\\\
             \begin{subfigure}[t]{0.5\textwidth}
@@
                 \caption{}
-            \end{subfigure} \\
-            \\
-
+            \end{subfigure}
+            \\\\
             \begin{subfigure}[t]{0.5\textwidth}
                 \centering
-                \includegraphics[width=0.65\linewidth, trim=0 +2cm 0 0.5cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_bandage_zoom_in}
+                \includegraphics[width=0.65\linewidth, trim=0 +8cm 0 0.5cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_C4_sorted_bandage_zoom_in}
                 \caption{}
@@
             \begin{center}
-                \includegraphics[width=\linewidth, trim=0cm 0.9cm -2.5cm 0.5cm]{fig/extract_viz_draw_position_untangle/chr6_C4_sorted_untangle_bed_subset_tsv.pdf}
+                \includegraphics[width=\linewidth, trim=0cm 0.9cm -2.5cm 0.5cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_C4_sorted_untangle_bed}
             \end{center}

### 2021-11-04 scripts/fig_extract_viz_draw_position_untangle.sh

@@ -66,12 +66,14 @@ odgi view -i ${filename_chr6_gfa}.C4.sorted.og -g > ${filename_chr6_gfa}.C4.sort
   bedtools sort -i - ) | awk '$8 == "-" { x=$6; $6=$5; $5=x; } { print }' |
   tr ' ' '\t'   >${filename_chr6_gfa}.C4.sorted.untangle.bed
 
-# cat $f.untangle.bed | grep '^chm13\|^grch38\|^HG00438\|^HG0107\|^HG01952'
+# cat ${filename_chr6_gfa}.C4.sorted.untangle.bed | grep '^chm13\|^grch38\|^HG00438\|^HG0107\|^HG01952'
 
 # R
 library(ggplot2)
-x <- read.table('/home/guarracino/Desktop/pangenomics/odgi_pub/chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.C4.sorted.og.untangle.bed.tsv', sep = '\t', header = T, comment.char="$")
+#library(knitr)
+x <- read.table('/home/guarracino/Desktop/pangenomics/odgi_pub/chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.C4.sorted.untangle.bed', sep = '\t', header = T, comment.char="$")
 x$query.name <- gsub(":.*","",x$query.name)
+x$query.name <- gsub("#J.*","",x$query.name)
 
 ggplot(
   subset(x, query.name %in% c(
@@ -86,23 +88,29 @@ ggplot(
 
     #"chm13#chr6",
     "grch38#chr6",
-    #"HG00438#1#JAHBCB010000040.1",
-    "HG00438#2#JAHBCA010000042.1",
-    #"HG01071#1#JAHBCF010000017.1",
-    "HG01071#2#JAHBCE010000076.1",
-    "HG01952#1#JAHAME010000044.1",
-    "HG01952#2#JAHAMD010000016.1"
+    #"HG00438#1",
+    "HG00438#2",
+    #"HG01071#1",
+    "HG01071#2",
+    "HG01952#1",
+    "HG01952#2"
     )
   ), aes(x=query.start, xend=query.end, y=ref.start, yend=ref.end)) +
-    geom_segment() +
+    geom_segment(size=0.3) +
     facet_grid(. ~ query.name) +
     coord_fixed() +
+    #theme_bw() +
     theme(
       text = element_text(size = 12.6),
       axis.text.x = element_text(size = 12, angle = 90),
[…]

### 2021-11-04 manuscript/fig_extract_viz_draw_position_untangle.tex

@@
         %\hline
-        \begin{subfigure}[t]{\textwidth}
+        \begin{subfigure}[t]{1.0\textwidth}
             \centering
             \caption{}
-            \includegraphics[width=\linewidth, trim=0cm 10cm -18cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_og_mhc_og_O_og_lay_H1000w100_png_svg}
+            \includegraphics[width=\linewidth, trim=0cm 0cm -10cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_gfa_og_mhc_og_O_og_lay_H1000w100_png_svg}
         \end{subfigure} \\

### 2021-11-04 manuscript/fig_metrics.tex
pdferize all metrics figures

@@
 		% include first image
-		\includegraphics[width=1.0\linewidth, trim=-2.25cm 1cm 0cm 2.75cm]{fig/metrics/chr4_pan_HTTex1_gfa_multiqc_odgi_stats}
+		\includegraphics[width=1.0\linewidth, trim=-2.25cm 1cm 0cm 2.75cm]{fig/metrics/chr4_pan_HTTex1_gfa_multiqc_odgi_stats_svg}
 		\label{fig:metrics-multiqc}
@@
 		% include second image
-		\includegraphics[width=1.0\linewidth, trim=-1.95cm 2cm -1.0cm 0.5cm]{fig/metrics/chr4_pan_fa_a2fb268_e820cd3_9ea71d8_smooth_gfa_og_HTTex1_og_O_og_tiny_og}
+		\includegraphics[width=1.0\linewidth, trim=-1.5cm 2cm -1.0cm 0.5cm]{fig/metrics/chr4_pan_fa_a2fb268_e820cd3_9ea71d8_smooth_gfa_og_HTTex1_og_O_og_tiny_og_png_svg}
 		\label{fig:metrics-viz}

### 2021-11-04 manuscript/fig_extract_viz_draw_position_untangle.tex
update and squeeze

@@
             \caption{}
-            \includegraphics[width=\linewidth, trim=0cm 0cm -10cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_gfa_og_mhc_og_O_og_lay_H1000w100_png_svg}
+            \includegraphics[width=\linewidth, trim=0cm 10cm -19cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_gfa_og_mhc_og_O_og_lay_H1000w100_png_svg}
         \end{subfigure} \\

### 2021-11-04 manuscript/fig_extract_viz_draw_position_untangle.tex
squeeze!

--- 2021-11-04 manuscript/fig_extract_viz_draw_position_untangle.tex
@@
             \caption{}
-            \includegraphics[width=\linewidth, trim=0cm 0cm -10cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_gfa_og_mhc_og_O_og_lay_H1000w100_png_svg}
+            \includegraphics[width=\linewidth, trim=0cm 20cm -20cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_gfa_og_mhc_og_O_og_lay_H1000w100_png_svg}
         \end{subfigure} \\
@@
                 \caption{}
-                \includegraphics[width=\linewidth, trim=0 1.7cm 0cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_C4_sorted}
+                \includegraphics[width=\linewidth, trim=0 1.9cm 0cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_C4_sorted}
             \end{subfigure} \\
@@
                 \caption{}
-                \includegraphics[width=\linewidth, trim=0 +1.7cm 0cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_C4_sorted_du}
+                \includegraphics[width=\linewidth, trim=0 +1.9cm 0cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_C4_sorted_du}
             \end{subfigure} \\
@@
                 \caption{}
-                \includegraphics[width=\linewidth, trim=0 +1.7cm 0cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_C4_sorted_z}
+                \includegraphics[width=\linewidth, trim=0 +1.9cm 0cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_C4_sorted_z}
             \end{subfigure} \\
@@
                 \caption{}
-                \includegraphics[width=\linewidth, trim=0 +1.7cm 0cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_C4_sorted_m}
+                \includegraphics[width=\linewidth, trim=0 +1.9cm 0cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_C4_sorted_m}
             \end{subfigure}
@@
         \textbf{(c)} Color by path position. %: the color gradients are smooth, highlighting that its node are well sorted in 1 dimension.
-        The top two reference genomes and 2 haplotypes (HG01952\#2) go from left to right, while 5 haplotypes go in the opposite direction, as indicated by the black color on their left.
+        The top two reference genomes and 2 haplotypes (HG01952\#2) xgo from left to right, while 5 haplotypes go in the opposite direction, as indicated by the black color on their left.
         \textbf{(d)} \textit{odgi viz} color by strandness: the red paths indicate the haplotypes that were assembled in reverse with respect to the 2 reference genomes.

### 2021-11-04 scripts/fig_extract_viz_draw_position_untangle.sh

@@ -28,16 +28,16 @@ odgi paths -i ${filename_chr6_gfa}.C4.sorted.og -L | grep 'chr6\|HG00438\|HG0107
 
 
 # odgi viz: default (binned) mode
-odgi viz -i ${filename_chr6_gfa}.C4.sorted.og -o "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_C4_sorted.png -c 12 -w 100 -y 50 -p chr6.selected_paths.txt
+odgi viz -i ${filename_chr6_gfa}.C4.sorted.og -o "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_C4_sorted.png -a 9 -c 12 -w 100 -y 50 -p chr6.selected_paths.txt
 
 # odgi viz: color by strand
-odgi viz -i ${filename_chr6_gfa}.C4.sorted.og -o "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_C4_sorted_z.png -c 12 -w 100 -y 50 -p chr6.selected_paths.txt -z
+odgi viz -i ${filename_chr6_gfa}.C4.sorted.og -o "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_C4_sorted_z.png -a 9 -c 12 -w 100 -y 50 -p chr6.selected_paths.txt -z
 
 # odgi viz: color by position
-odgi viz -i ${filename_chr6_gfa}.C4.sorted.og -o "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_C4_sorted_du.png -c 12 -w 100 -y 50 -p chr6.selected_paths.txt -du
+odgi viz -i ${filename_chr6_gfa}.C4.sorted.og -o "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_C4_sorted_du.png -a 9 -c 12 -w 100 -y 50 -p chr6.selected_paths.txt -du
 
 # odgi viz: color by depth
-odgi viz -i ${filename_chr6_gfa}.C4.sorted.og -o "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_C4_sorted_m.png -c 12 -w 100 -y 50 -p chr6.selected_paths.txt -m -B Spectral:4
[…]

### 2021-11-04 manuscript/fig_metrics.tex
pass through

@@
 \textbf{(e)} \textit{vg viz} nucleotide-level visualization of 10 gene alleles, CHM13, GRCH38 of the \textit{HTTexon1} graph focusing on the CAG variable repeat region.
-Figures \textbf{(b)-(e)} clearly highlight the variant region around position 200 of CHM13 showing the variable number of glutamine residues of the different individuals as reported by \cite{Nance1999}.}
+Figures \textbf{(b)-(e)} highlight the variant region around position 200 of CHM13 showing the variable number of glutamine residues of the different individuals as reported by \cite{Nance1999}.}
 	\label{fig:metrics}

### 2021-11-04 manuscript/main.tex
redo metrics figures with wgg.87
pass through

@@
 To investigate on the genomes encoded in the graph, \textit{odgi paths} allows users to calculate pairwise overlap statistics of groupings of paths and emit all path sequences in FASTA format, and it also allows the generation of a ``pangenome matrix'' that reports the copy number (presence/absence) of each path over each node.
-\textit{odgi flatten} generate a linearization of the graph by emitting the pangenome sequence (the concatenation of all graph node sequences) in FASTA format, and the projection of all paths on the linearized sequence in BED format.
+\textit{odgi flatten} generates a linearization of the graph by emitting the pangenome sequence (the concatenation of all node sequences) in FASTA format, and the projection of all paths on the linearized sequence in BED format.
 
@@
 
-ODGI is already the backbone of the pipelines such as the Pangenome Graph Builder~\citep{pggb}.
-% or nf-core/pangenome~\citep{pangenome}.
+ODGI is already the backbone of the pipelines such as the Pangenome Graph Builder~\citep{pggb} and its Nextflow child nf-core/pangenome~\citep{pangenome}.
 %, where it is also applied for debugging purposes.
@@
 
-Data used to build human pangenome graphs: \url{https://github.com/human-pangenomics/HPP_Year1_Data_Freeze_v1.0}.
+%Data used to build human pangenome graphs: \url{https://github.com/human-pangenomics/HPP_Year1_Data_Freeze_v1.0}. 
+The graphs we generated the figures from: \url{https://s3-us-west-2.amazonaws.com/human-pangenomics/index.html?prefix=pangenomes/scratch/2021_11_04_pggb_wgg.87/}.
 Code used to build this manuscript and its figures: \url{https://github.com/pangenome/odgi-paper}.

### 2021-11-04 manuscript/fig_extract_viz_draw_position_untangle.tex
a bit more appealing odgi viz

@@
     \centering
-    \begin{tabular}[t]{cc}%{|c|c|}
+    \begin{tabular}[t]{c}%{|c|c|}
         %\hline
@@
             \caption{}
-            \includegraphics[width=\linewidth, trim=0cm 20cm -20cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_gfa_og_mhc_og_O_og_lay_H1000w100_png_svg}
+            \includegraphics[width=\linewidth, trim=0cm 18cm -20cm 0.5cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_gfa_og_mhc_og_O_og_lay_H1000w100_png_svg}
         \end{subfigure} \\

### 2021-11-04 manuscript/fig_extract_viz_draw_position_untangle.tex
pass through

--- 2021-11-04 manuscript/fig_extract_viz_draw_position_untangle.tex
@@
         \textbf{(c)} Color by path position. %: the color gradients are smooth, highlighting that its node are well sorted in 1 dimension.
-        The top two reference genomes and 2 haplotypes (HG01952\#2) xgo from left to right, while 5 haplotypes go in the opposite direction, as indicated by the black color on their left.
+        The top two reference genomes and 2 haplotypes (HG01952\#2) go from left to right, while 5 haplotypes go in the opposite direction, as indicated by the black color on their left.
         \textbf{(d)} \textit{odgi viz} color by strandness: the red paths indicate the haplotypes that were assembled in reverse with respect to the 2 reference genomes.
@@
 Node labels were annoted by using \textit{odgi position}.
-        \textbf{(h)} Visualization of \textit{odgi untangle} output in the C4 pangenome graph: the plots show the copy number status of the sequences in the C4 region with respect to the grch38 reference sequence, making clear, for example, that in HG00438\#2, \textit{C4B} gene is missing.
+        \textbf{(h)} Visualization of \textit{odgi untangle} output in the C4 pangenome graph: the plots show the copy number status of the sequences in the C4 region with respect to the grch38 reference sequence, making clear, for example, that in HG00438\#2, the \textit{C4B} gene is missing.
     }

### 2021-11-04 manuscript/fig_metrics.tex
fix and squeeze metrics

--- 2021-11-04 manuscript/fig_metrics.tex
@@
         % include first image
-		\includegraphics[width=1.0\linewidth, trim=-2.25cm 1cm 0cm 2.75cm]{fig/metrics/chr4_pan_HTTex1_gfa_multiqc_odgi_stats}
+        \includegraphics[width=1.0\linewidth, trim=-2.25cm 3.4cm 0cm 2.75cm]{fig/metrics/chr4_pan_HTTex1_gfa_multiqc_odgi_stats_svg}
         \label{fig:metrics-multiqc}
@@
         % include fourth image
-	\includegraphics[width=\linewidth,trim=+.225cm 3cm +.425cm +3cm]{fig/metrics/chr4_HTT_chm13_degree_w1_bed}
+        \includegraphics[width=\linewidth,trim=-.225cm 4cm +.425cm +3cm]{fig/metrics/chr4_HTT_chm13_degree_w1_bed}
         \label{fig:metrics-degree}
@@
         % include second image
-	\includegraphics[width=\linewidth,trim=+.225cm 3cm +0.425cm +3cm]{fig/metrics/chr4_HTT_chm13_depth_w1_bed}
+        \includegraphics[width=\linewidth,trim=-.225cm 3.3cm +0.425cm +3cm]{fig/metrics/chr4_HTT_chm13_depth_w1_bed}
         \label{fig:metrics-depth}
@@
         % include second image
-		\includegraphics[width=1.0\linewidth, trim=-1.95cm 2cm -1.0cm 0.5cm]{fig/metrics/chr4_pan_fa_a2fb268_e820cd3_9ea71d8_smooth_gfa_og_HTTex1_og_O_og_tiny_og}
+        \includegraphics[width=1.0\linewidth, trim=-1.75cm 2.2cm -1.0cm 0.5cm]{fig/metrics/chr4_pan_fa_a2fb268_e820cd3_9ea71d8_smooth_gfa_og_HTTex1_og_O_og_tiny_og_png_svg}
         \label{fig:metrics-viz}
@@
         % include second image
-	\includegraphics[width=1.0\linewidth, trim=-0.4cm 0.1cm 3.15cm 0cm, clip]{fig/metrics/chr4_pan_HTTex1_STR_xg_svg.pdf}
+        \includegraphics[width=1.0\linewidth, trim=-0.4cm 0.4cm 3.15cm 0cm]{fig/metrics/chr4_pan_HTTex1_STR_xg_svg}
         \label{fig:metrics-str}

### 2021-11-05 manuscript/main.tex
implementation tuning

@@
 ODGI improves on prior efforts, based on issues that arose during our work with high-quality \textit{de novo} assemblies that cover almost all parts of the human genome~\citep{Logsdon_2021,Nurk_2021}.
-For example, we find that it is necessary to support graphs with regions with very high numbers of path traversals (high path depth).
-Such motifs can occur in collapsed repeat structures generated by ambiguous sequence homology relationships in repeats found in the centromeres and other segmental duplications.
-If we cannot process such regions, we have only two options: 1) we can remove these regions, or 2) we can leave them unaligned.
+In more detail, we find that it is necessary to support graphs with regions with very high numbers of path traversals (high path depth).
+Such motifs can occur in collapsed structures generated by ambiguous sequence homology relationships in repeats found in the centromeres and other segmental duplications.
+If we cannot process such regions, there are only two options: 1) we can remove such regions, or 2) we can leave them unaligned.
 However, neither of these solutions allows us to investigate their biological features.
-To seamlessly represent such difficult regions, we follow an approach implemented in the dynamic version of the Graph BWT (GBWT)~\citep{Siren:2020} and built a node-centric, dynamic, compressed model of the paths.
+To seamlessly represent such difficult regions, we followed an approach implemented in the dynamic version of the Graph BWT (GBWT)~\citep{Siren:2020} and built a node-centric, dynamic, compressed model of the paths.
 This design supports node-local modification and update of the graph, which lets us operate on paths in parallel.
@@
 Expressed in terms of the variation graph $V$, ODGI's core $Node$ structure includes a decoder that maps the neighbors of each node to a dense range of integers.
-For a given node $i$ and neighbor $j$, the decoder itself does not store the $id$ of $j$, but rather a compact representation of the relative difference between the node ids: $\delta = Node_i.id - Node_j.id$.
+For a given $Node_i$ and neighbor $Node_j$, the decoder itself does not store the $id$ of $Node_j$, but rather a compact representation of the relative difference between the node ids: $\delta = Node_i.id - Node_j.id$.
 %The $\delta between two nodes is computed as their distance in the graph vector (i.e., the difference between the node offsets).
@@
 Each structure contains the sequence of the node ($Node_i.sequence$), its edges in both directions ($Node_i.edges$), and a vector of path steps that describes the previous and next steps in paths that walk across the node ($Node_i.path\_steps$).
-For efficiency, $Node_i.sequence$ is stored as a plain string, while the $edges$ and $path\_steps$ are stored using a dynamic succinct integer vector that requires $O(2nw)$ bits for the edges and $O(5nw)$ bits for the path steps, where $n$ is the number of steps on the node, and $w$ is $\approx log_2(n)$ \citep{prezza2017framework}.
+For efficiency, $Node_i.sequence$ is stored as a plain string, while the $edges$ and $path\_steps$ are stored using a dynamic succinct integer vector that requires $O(2nw)$ bits for the edges and $O(5nw)$ bits for the path steps, where $n$ is the number of steps on the node and $w$ is $\approx log_2(n)$ \citep{prezza2017framework}.
 
 To allow edit operations in parallel, each node structure includes a byte-width mutex $lock$.
-All edit operations on the graph must touch at most two $Node$ structs at a time (both edge and path step representations are doubly-linked).
+All changes on the graph can involve at most two $Node$ structs at a time (both edge and path step representations are doubly-linked).
 To avoid deadlocks, we acquire the node locks in ascending $Node.id$ order and release them in desceding order.
@@
 By avoiding global locks, we implement many of the operations in ODGI using maximum parallelism available.
-This approach is key to enable our methods to scale to the largest pangenome graphs that we can currently build.
+This approach is key to enable our methods to scale to the largest pangenome graphs that we can currently build (with hundreds of vertebrate genomes).
 
@@
 
-ODGI is already the backbone of the pipelines such as the Pangenome Graph Builder~\citep{pggb} and its Nextflow child nf-core/pangenome~\citep{pangenome}.
-%, where it is also applied for debugging purposes.
-%\footnote{\url{https://github.com/nf-core/pangenome} (accessed Oct 2021)}.
+ODGI is already the backbone of the Pangenome Graph Builder pipeline~\citep{pggb}.
 Its static, large-scale 1D and 2D visualizations of the pangenome graphs allow an unprecedented high-level perspective on variation in pangenomes, and have also been critical in the development of pangenome graph building methods.

### 2021-11-05 manuscript/main.tex
feedback Sven Nahnsen

@@
 Simon~Heumos\,$^{\text{\sfb 2} \dagger}$,
-Sven~Nahnsen\,$^{\text{\sfb 2}}$,
-Pjotr~Prins\,$^{\text{\sfb 3}}$,
-and~Erik~Garrison\,$^{\text{\sfb 3}*}$
+Sven~Nahnsen\,$^{\text{\sfb 2,3}}$,
+Pjotr~Prins\,$^{\text{\sfb 4}}$,
+and~Erik~Garrison\,$^{\text{\sfb 4}*}$
 }
@@
 $^{\text{\sf 2}}$Quantitative Biology Center (QBiC), University of T\"ubingen, T\"ubingen, Germany, 72076 \\
-$^{\text{\sf 3}}$University of Tennessee Health Science Center, Memphis, TN, USA
+$^{\text{\sf 3}}$Biomedical Data Science, Dept. of Computer Science, University of T\"ubingen, T\"ubingen, Germany, 72076 \\
+$^{\text{\sf 4}}$University of Tennessee Health Science Center, Memphis, TN, USA
 }
@@
 ODGI includes tools for detecting complex regions, extracting \textit{loci}, removing artifacts, exploratory analysis, manipulation, validation, and visualization.
-Its fast parallel execution facilitates routine pangenomic tasks as well as pipelines that can quickly answer complex biological questions of gigabase-scale pangenome graphs. \\
+Its fast parallel execution facilitates routine pangenomic tasks, as well as pipelines that can quickly answer complex biological questions of gigabase-scale pangenome graphs. \\
 \textbf{Availability:}
@@
 In pangenome graphs~\citep{Paten:2017}, homologous regions between genomes are compressed into a single representative of all alleles present in the pangenome.
-These flexible models let us encode any kind of variation, allowing the generation of comprehensive data systems on which to base analyses of genome evolution.
+These flexible models let us encode any kind of variation, allowing the generation of comprehensive data systems that builds the basis for the analyses of genome evolution.
 Although these data structures are of clear utility to researchers~\citep{cpang2018}, and have been the focus of numerous applications in read mapping~\citep{Garrison:2018,Baaijens_2019,Hickey:2020,Sibbesen_2021},
@@
 ODGI improves on prior efforts, based on issues that arose during our work with high-quality \textit{de novo} assemblies that cover almost all parts of the human genome~\citep{Logsdon_2021,Nurk_2021}.
-In more detail, we find that it is necessary to support graphs with regions with very high numbers of path traversals (high path depth).
+In more detail, we find that it is necessary to support graphs with regions of very high numbers of path traversals (high path depth).
 Such motifs can occur in collapsed structures generated by ambiguous sequence homology relationships in repeats found in the centromeres and other segmental duplications.
-If we cannot process such regions, there are only two options: 1) we can remove such regions, or 2) we can leave them unaligned.
+If we cannot process such regions, there are only two options: 1) remove such regions, or 2) leave them unaligned.
 However, neither of these solutions allows us to investigate their biological features.
@@
 %Although they present new complexities, we find that pangenome graphs are compatible with current standard ways of working with bioinformatics data.
-ODGI thus provides a set of interfaces that let us approach these graphs from the perspective of standard reference- and sequence-based data models.
+Thus, ODGI provides a set of interfaces that let us approach these graphs from the perspective of standard reference- and sequence-based data models.
 Indeed, by considering all paths in the graph as potential reference or query sequence, we make graphs invisible to downstream tools that operate on collections of sequences or rely on a reference sequence (\textit{e.g.} SAMtools~\citep{Li2009}), enabling interoperability.

### 2021-11-05 manuscript/fig_metrics.tex
adjust fig3d

@@
         % include second image
-        \includegraphics[width=1.0\linewidth, trim=-1.75cm 2.2cm -1.0cm 0.5cm]{fig/metrics/chr4_pan_fa_a2fb268_e820cd3_9ea71d8_smooth_gfa_og_HTTex1_og_O_og_tiny_og_png_svg}
+        \includegraphics[width=1.0\linewidth, trim=-1.75cm 2.2cm -1cm 0.5cm]{fig/metrics/chr4_pan_fa_a2fb268_e820cd3_9ea71d8_smooth_gfa_og_HTTex1_og_O_og_tiny_og_png_svg}
         \label{fig:metrics-viz}

### 2021-11-05 manuscript/fig_metrics.tex
tweaks

@@
 %	\includegraphics[width=\linewidth]{fig/metrics/chr4.pan.HTTex1.gfa.multiqc_odgi_stats.png}
-	\caption{Features of a 90 haplotypes human pangenome graph of the exon 1 huntingtin gene (\textit{HTTexon1}): \textbf{(a)} Excerpt of vital statistics of the \textit{HTTexon1} graph displayed by MultiQC's ODGI module.
+	\caption{Features of a 90-haplotype human pangenome graph of the exon 1 huntingtin gene (\textit{HTTexon1}): \textbf{(a)} Excerpt of vital statistics of the \textit{HTTexon1} graph displayed by MultiQC's ODGI module.
 The very high GC content of 73.0\% compared to a human genomic mean GC content of 40.9\% \cite{Piovesan2019} is in accordance with the literature (see for example \cite{Sathasivam2013, Neueder2017}).

### 2021-11-05 manuscript/main.tex
tweaks

@@
 These flexible models let us encode any kind of variation, allowing the generation of comprehensive data systems that builds the basis for the analyses of genome evolution.
-Although these data structures are of clear utility to researchers~\citep{cpang2018}, and have been the focus of numerous applications in read mapping~\citep{Garrison:2018,Baaijens_2019,Hickey:2020,Sibbesen_2021},
-the scientific community still lacks a toolset specifically focused on graph manipulation and interrogation.
+Although these data structures are of utility to researchers~\citep{cpang2018,Garrison:2018,Baaijens_2019,Hickey:2020,Sibbesen_2021}, the scientific community still lacks a toolset specifically focused on graph manipulation and interrogation.
 
@@
 We anticipate that \textit{de novo} assemblies of similar quality will become common, leading to demand for methods that allow us to create and explore pangenomes.
-Pangenome graphs are a natural model to support these kinds of analyses, but studies based on pangenome graphs are difficult due to a lack of downstream tools and standard workflows.
+%Pangenome graphs are a natural model to support these kinds of analyses, but studies based on pangenome graphs are difficult due to a lack of downstream tools and standard workflows.
 
@@
 ODGI improves on prior efforts, based on issues that arose during our work with high-quality \textit{de novo} assemblies that cover almost all parts of the human genome~\citep{Logsdon_2021,Nurk_2021}.
-In more detail, we find that it is necessary to support graphs with regions of very high numbers of path traversals (high path depth).
+In particular, we find that it is necessary to support graphs with regions of very high numbers of path traversals (high path depth).
 Such motifs can occur in collapsed structures generated by ambiguous sequence homology relationships in repeats found in the centromeres and other segmental duplications.
@@
 Previously, great efforts have been made to prevent ``collapse'' of non-orthologous sequences in the graph topology itself.
-While our presentation is primarily qualitative, our work demonstrates that we can completely mitigate this issue by exploiting the pangenome graph not as a static reference, but as a dynamic model of the mutual alignment of many genomes.
+While our presentation is primarily qualitative, our work demonstrates that we can mitigate this issue by exploiting the pangenome graph not as a static reference, but as a dynamic model of the mutual alignment of many genomes.
+
 %Our future work will focus on further extensions of this topic.
@@
 %Data used to build human pangenome graphs: \url{https://github.com/human-pangenomics/HPP_Year1_Data_Freeze_v1.0}. 
-The graphs we generated the figures from: \url{https://s3-us-west-2.amazonaws.com/human-pangenomics/index.html?prefix=pangenomes/scratch/2021_11_04_pggb_wgg.87/}.
-Code used to build this manuscript and its figures: \url{https://github.com/pangenome/odgi-paper}.
+%The graphs we generated the figures from: \url{https://s3-us-west-2.amazonaws.com/human-pangenomics/index.html?prefix=pangenomes/scratch/2021_11_04_pggb_wgg.87/}.
+Code and links to data resources used to build this manuscript and its figures, can be found in the paper's public repository: \url{https://github.com/pangenome/odgi-paper}.
 

### 2021-11-06 manuscript/main.tex
our most famous coauthor

@@
 Sven~Nahnsen\,$^{\text{\sfb 2,3}}$,
+Human Pangenome Reference Consortium,
 Pjotr~Prins\,$^{\text{\sfb 4}}$,
@@
 $^{\text{\sf 3}}$Biomedical Data Science, Dept. of Computer Science, University of T\"ubingen, T\"ubingen, Germany, 72076 \\
+%$^{\text{\sf 4}}$\url{https://humanpangenome.org/} \\
 $^{\text{\sf 4}}$University of Tennessee Health Science Center, Memphis, TN, USA
@@
 Finally, we create paths in serial, and extend them in parallel by obtaining the mutex $Node.lock$ for pairs of nodes and by adding the path step in their $Node.path\_steps$ vectors.
-This parallelism speeds ODGI model construction by many-fold when testing against graphs built by the HPRC (Figure~\ref{fig:build}).
+This parallelism speeds ODGI model construction by many-fold when testing against graphs made from assemblies produced by the HPRC (Figure~\ref{fig:build}).
 
@@
 Pangenome graphs model alignments of many genomes.
-With \textit{odgi untangle}, users can extract pairwise \textit{many}:\textit{1} alignment information between a given set of ``query'' sequences and a given set of ``target'' sequences (used as references).
-This lets us decompose both homology and paralogy relationships by also untangling looping motifs in the graph, and obtaining linearized pairs of regions in BEDPE or PAF format.
+With \textit{odgi untangle}, users can extract pairwise alignment information between a given set of ``query'' sequences and a given set of ``target'' sequences (used as references).
+While pangenome graphs may contain looping structures that imply many-to-many alignments between the pangenome sequences, these untangled alignments map each segment of the queries to a single segment in the set of targets.
+%This lets us decompose both homology and paralogy relationships by also untangling looping motifs in the graph, and obtaining linearized pairs of regions in BEDPE or PAF format.
 Being able to work with any sets of reference sequences lets us convert the graph to lift-over maps compatible with standard software for projecting annotations and alignments from one genome to another.
@@
 we can apply \textit{odgi untangle} to segment paths into linear segments by breaking these segments where the paths loop back on themselves.
-We ``untangle'' by finding the target segment that best match each query segment using the \textit{path jaccard} context mapping model.
+We first discover segment boundaries using standard approaches for detecting repeats in sequence graphs \citep{Pevzner_2004}.
+We finally ``untangle'' by finding the target segment that best match each query segment using the \textit{path jaccard} context mapping model.
 %The \textit{path jaccard} concept ensures that the breaking is nucleotide-precise.
@@
 
-ODGI has allowed us to explore \textit{context mapping} via the path jaccard metric.
-This resolves a major conceptual issue that has strongly guided existing algorithms to construct pangenome graphs \citep{Li:2020}.
-Previously, great efforts have been made to prevent ``collapse'' of non-orthologous sequences in the graph topology itself.
+ODGI has allowed us to explore \textit{context mapping} deconvolution of pangenome graph structures via the path jaccard metric.
+This resolves a major conceptual issue that has strongly guided existing algorithms to construct pangenome graphs.
+Previously, great efforts have been made to prevent the ``collapse'' of non-orthologous sequences in the graph topology itself  \citep{Li:2020}.
+This has been seen as essential to making these new bioinformatic models interpretable.
 While our presentation is primarily qualitative, our work demonstrates that we can mitigate this issue by exploiting the pangenome graph not as a static reference, but as a dynamic model of the mutual alignment of many genomes.
+Because pangenome graphs can contain complete genomes, we are able to query them to polarize the information they contain in easily-interpretable and reusable pairwise formats that are widely supported in bioinformatics.
+We expect that ODGI will provide a reference interface between pangenomic and genomic approaches to understanding genome variation.
+
+%To understand this model, we query it using any reference that it contains.
 

### 2021-11-07 manuscript/main.tex
maybe the famous author isn't needed

@@
 Sven~Nahnsen\,$^{\text{\sfb 2,3}}$,
-Human Pangenome Reference Consortium,
 Pjotr~Prins\,$^{\text{\sfb 4}}$,
@@
 $^{\text{\sf 3}}$Biomedical Data Science, Dept. of Computer Science, University of T\"ubingen, T\"ubingen, Germany, 72076 \\
-%$^{\text{\sf 4}}$\url{https://humanpangenome.org/} \\
 $^{\text{\sf 4}}$University of Tennessee Health Science Center, Memphis, TN, USA

### 2021-11-08 manuscript/fig_extract_viz_draw_position_untangle.tex
spacy
update Fig4 with the 'final' WGG, with a bit more zoomed MHC locus; simplified the scripts

@@
             \caption{}
-            \includegraphics[width=\linewidth, trim=0cm 18cm -20cm 0.5cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_gfa_og_mhc_og_O_og_lay_H1000w100_png_svg}
+            \includegraphics[width=\linewidth, trim=0cm 18cm -20cm 0.5cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_mhc_H1000w120}
         \end{subfigure} \\
@@
         \textbf{(h)} Visualization of \textit{odgi untangle} output in the C4 pangenome graph: the plots show the copy number status of the sequences in the C4 region with respect to the grch38 reference sequence, making clear, for example, that in HG00438\#2, the \textit{C4B} gene is missing.
+\vspace{-1em}
     }

### 2021-11-08 scripts/fig_extract_viz_draw_position_untangle.sh

@@ -1,5 +1,23 @@
 #!/bin/bash
 
+filename_chr6_gfa=chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth
+
+# Download and build the graph
+wget https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/scratch/2021_11_04_pggb_wgg.87/chroms/${filename_chr6_gfa}.gfa.gz
+gunzip ${filename_chr6_gfa}.gfa.gz
+odgi build -g ${filename_chr6_gfa}.gfa -o ${filename_chr6_gfa}.og -t 16 -P
+
+
+# Extraction and optimization of the MHC locus
+odgi extract -i ${filename_chr6_gfa}.og -r grch38#chr6:29000000-34000000 -o - --full-range -t 16 -P | odgi sort -i - -o ${filename_chr6_gfa}.mhc.og --optimize
+
+
+# MHC locus layout
+odgi layout -i ${filename_chr6_gfa}.mhc.og -o ${filename_chr6_gfa}.mhc.lay -T ${filename_chr6_gfa}.mhc.tsv -x 100 -t 16 -P
+odgi draw -i ${filename_chr6_gfa}.mhc.og -c ${filename_chr6_gfa}.mhc.lay -p "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_mhc_H1000w10.png -H 1000 -w 120 -C -t 16 -P
+# Add red box (30px) around C4 locus using GIMP
+
+
 # Find C4 coordinates
 wget http://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/hg38.chrom.sizes
 wget https://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/genes/hg38.ncbiRefSeq.gtf.gz
@@ -7,14 +25,6 @@ zgrep 'gene_id "C4A"\|gene_id "C4B"' hg38.ncbiRefSeq.gtf.gz |
   awk '$1 == "chr6"' | cut -f 1,4,5 |
   bedtools sort | bedtools merge -d 15000 | bedtools slop -l 10000 -r 20000 -g hg38.chrom.sizes |
   sed 's/chr6/grch38#chr6/g' > hg38.ncbiRefSeq.C4.coordinates.bed
-
-
-filename_chr6_gfa=chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth
-
-# Download and build the graph
[…]

### 2021-11-08 scripts/metrics.sh

@@ -3,11 +3,10 @@
 wget https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/scratch/2021_11_04_pggb_wgg.87/chroms/chr4.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
 gunzip chr4.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
 wget https://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/genes/hg38.ncbiRefSeq.gtf.gz
-gunzip hg38.ncbiRefSeq.gtf.gz
 
 G=chr4.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa
 
-grep HTT hg38.ncbiRefSeq.gtf | grep "exon_number \"1\"" | grep "gene_name \"HTT\"" | grep -P "\texon"
+zgrep HTT hg38.ncbiRefSeq.gtf | grep "exon_number \"1\"" | grep "gene_name \"HTT\"" | grep -P "\texon"
 #chr4	ncbiRefSeq.2021-09-09	exon	3074681	3075088	.	+	.	gene_id "HTT"; transcript_id "NM_001388492.1"; exon_number "1"; exon_id "NM_001388492.1.1"; gene_name "HTT";
 #chr4	ncbiRefSeq.2021-09-09	exon	3074681	3075088	.	+	.	gene_id "HTT"; transcript_id "NM_002111.8"; exon_number "1"; exon_id "NM_002111.8.1"; gene_name "HTT";
 

### 2021-11-08 scripts/mhc.sh

@@ -1,21 +1,0 @@
-wget https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/scratch/2021_11_04_pggb_wgg.87/chroms/chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
-gunzip chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
-
-MHC=chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.og
-odgi build -g chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa -o "$MHC" -t 28 -P
-odgi extract -i "$MHC" -o "$MHC".mhc.og -E -P -t 28 -r grch38#chr6:29000000-34000000
-odgi sort -i "$MHC".mhc.og -o "$MHC".mhc.og.O.og -O
-odgi layout -i "$MHC".mhc.og.O.og -o "$MHC".mhc.og.O.og.lay -T "$MHC".mhc.og.O.og.lay.tsv -P -t 28 -x 100
-odgi draw -i chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.og.mhc.og.O.og -c chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.og.mhc.og.O.og.lay -p chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.og.mhc.og.O.og.lay.H1000w10.png -H 1000 -C -w 10 -t 28 -P
-
-#### MANUAL INTERVENTION ####
-# Go to https://drububu.com/tutorial/bitmap-to-vector.html. 
-# Drag and drop chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.og.mhc.og.O.og.lay.H1000w100.png into the dashed field.
-# Click "save as svg" to chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.og.mhc.og.O.og.lay.H1000w100.png.svg. This might take some time, be patient!
-
-inkscape chr6.pan.fa.a2fb268.4030258.d9f1245.smooth.gfa.og.mhc.og.O.og.lay.H1000w100.png.svg --export-pdf=chr6_pan_fa_a2fb268_4030258_d9f1245_smooth_gfa_og_mhc_og_O_og_lay_H1000w100_png_svg.pdf
-
-#### MANUAL INTERVENTION ####
-# Add box around C4 using Inkscape
-
-

### 2021-11-08 manuscript/fig_metrics.tex
redo metrics figures with wgg.88

@@
         % include second image
-        \includegraphics[width=1.0\linewidth, trim=-1.75cm 2.2cm -1cm 0.5cm]{fig/metrics/chr4_pan_fa_a2fb268_e820cd3_9ea71d8_smooth_gfa_og_HTTex1_og_O_og_tiny_og_png_svg}
+        \includegraphics[width=1.0\linewidth, trim=-1.75cm 2.2cm -0.75cm 0.5cm]{fig/metrics/chr4_pan_fa_a2fb268_4030258_6a1ecc2_smooth_gfa_og_HTTex1_og_O_og_tiny_og_png_svg.pdf}
         \label{fig:metrics-viz}

### 2021-11-08 scripts/metrics.sh

@@ -1,10 +1,12 @@
 #!/bin/bash
 
-wget https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/scratch/2021_11_04_pggb_wgg.87/chroms/chr4.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
-gunzip chr4.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
+#### FIXME once wgg.88 is in the HPRC repository ####
+wget wget http://hypervolu.me/~guarracino/chr4.pan.fa.a2fb268.4030258.6a1ecc2.smooth.og.gz
+gunzip chr4.pan.fa.a2fb268.4030258.6a1ecc2.smooth.og.gz
 wget https://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/genes/hg38.ncbiRefSeq.gtf.gz
 
-G=chr4.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa
+# G=chr4.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa
+G=chr4.pan.fa.a2fb268.4030258.6a1ecc2.smooth
 
 zgrep HTT hg38.ncbiRefSeq.gtf | grep "exon_number \"1\"" | grep "gene_name \"HTT\"" | grep -P "\texon"
 #chr4	ncbiRefSeq.2021-09-09	exon	3074681	3075088	.	+	.	gene_id "HTT"; transcript_id "NM_001388492.1"; exon_number "1"; exon_id "NM_001388492.1.1"; gene_name "HTT";
@@ -59,7 +61,7 @@ odgi viz -i "$G".og.HTTex1.og.O.og.tiny.og -o "$G".og.HTTex1.og.O.og.tiny.og.png
 # check if the svg already exists, else boil out
 if [[ -f "$G".og.HTTex1.og.O.og.tiny.og.png.svg ]]
 then
-    inkscape "$G".og.HTTex1.og.O.og.tiny.og.png.svg --export-pdf=chr4_pan_fa_a2fb268_e820cd3_9ea71d8_smooth_gfa_og_HTTex1_og_O_og_tiny_og_png_svg.pdf
+    inkscape "$G".og.HTTex1.og.O.og.tiny.og.png.svg --export-pdf=chr4_pan_fa_a2fb268_4030258_6a1ecc2_smooth_gfa_og_HTTex1_og_O_og_tiny_og_png_svg.pdf
 else
[…]

### 2021-11-08 scripts/fig_extract_viz_draw_position_untangle.sh
put the temporary link for chr6 too

@@ -3,7 +3,8 @@
 filename_chr6_gfa=chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth
 
 # Download and build the graph
-wget https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/scratch/2021_11_04_pggb_wgg.87/chroms/${filename_chr6_gfa}.gfa.gz
+# FIXME with the final link
+wget wget http://hypervolu.me/~guarracino/chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.og.gz
 gunzip ${filename_chr6_gfa}.gfa.gz
 odgi build -g ${filename_chr6_gfa}.gfa -o ${filename_chr6_gfa}.og -t 16 -P
 

### 2021-11-08 scripts/fig_extract_viz_draw_position_untangle.sh
share GFA files

@@ -1,21 +1,21 @@
 #!/bin/bash
 
-filename_chr6_gfa=chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth
+prefix_chr6_smooth=chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth
 
 # Download and build the graph
 # FIXME with the final link
-wget wget http://hypervolu.me/~guarracino/chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.og.gz
-gunzip ${filename_chr6_gfa}.gfa.gz
-odgi build -g ${filename_chr6_gfa}.gfa -o ${filename_chr6_gfa}.og -t 16 -P
+wget wget /lizardfs/erikg/HPRC/year1v2genbank/wgg.88/chr6.pan/${prefix_chr6_smooth}.gfa.gz
+gunzip ${prefix_chr6_smooth}.gfa.gz
+odgi build -g ${prefix_chr6_smooth}.gfa -o ${prefix_chr6_smooth}.og -t 16 -P
 
 
 # Extraction and optimization of the MHC locus
-odgi extract -i ${filename_chr6_gfa}.og -r grch38#chr6:29000000-34000000 -o - --full-range -t 16 -P | odgi sort -i - -o ${filename_chr6_gfa}.mhc.og --optimize
+odgi extract -i ${prefix_chr6_smooth}.og -r grch38#chr6:29000000-34000000 -o - --full-range -t 16 -P | odgi sort -i - -o ${prefix_chr6_smooth}.mhc.og --optimize
 
 
 # MHC locus layout
-odgi layout -i ${filename_chr6_gfa}.mhc.og -o ${filename_chr6_gfa}.mhc.lay -T ${filename_chr6_gfa}.mhc.tsv -x 100 -t 16 -P
-odgi draw -i ${filename_chr6_gfa}.mhc.og -c ${filename_chr6_gfa}.mhc.lay -p "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_mhc_H1000w10.png -H 1000 -w 120 -C -t 16 -P
+odgi layout -i ${prefix_chr6_smooth}.mhc.og -o ${prefix_chr6_smooth}.mhc.lay -T ${prefix_chr6_smooth}.mhc.tsv -x 100 -t 16 -P
[…]

### 2021-11-08 scripts/metrics.sh

@@ -1,18 +1,17 @@
 #!/bin/bash
 
 #### FIXME once wgg.88 is in the HPRC repository ####
-wget wget http://hypervolu.me/~guarracino/chr4.pan.fa.a2fb268.4030258.6a1ecc2.smooth.og.gz
-gunzip chr4.pan.fa.a2fb268.4030258.6a1ecc2.smooth.og.gz
+wget wget http://hypervolu.me/~guarracino/chr4.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
+gunzip chr4.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
 wget https://hgdownload.soe.ucsc.edu/goldenPath/hg38/bigZips/genes/hg38.ncbiRefSeq.gtf.gz
 
-# G=chr4.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa
 G=chr4.pan.fa.a2fb268.4030258.6a1ecc2.smooth
 
 zgrep HTT hg38.ncbiRefSeq.gtf | grep "exon_number \"1\"" | grep "gene_name \"HTT\"" | grep -P "\texon"
 #chr4	ncbiRefSeq.2021-09-09	exon	3074681	3075088	.	+	.	gene_id "HTT"; transcript_id "NM_001388492.1"; exon_number "1"; exon_id "NM_001388492.1.1"; gene_name "HTT";
 #chr4	ncbiRefSeq.2021-09-09	exon	3074681	3075088	.	+	.	gene_id "HTT"; transcript_id "NM_002111.8"; exon_number "1"; exon_id "NM_002111.8.1"; gene_name "HTT";
 
-odgi build -g "$G" -o "$G".og -t 16 -P
+odgi build -g "$G".gfa -o "$G".og -t 16 -P
 odgi paths -L -i "$G".og > "$G".og.paths
 
 odgi extract -i "$G".og -r grch38#chr4:3074681-3075088 -o "$G".og.HTTex1.og -P -t 16 -E

### 2021-11-09 manuscript/fig_metrics.tex
go back to 7 pages again
give thanks

@@
         % include first image
-        \includegraphics[width=1.0\linewidth, trim=-2.25cm 3.4cm 0cm 2.75cm]{fig/metrics/chr4_pan_HTTex1_gfa_multiqc_odgi_stats_svg}
+        \includegraphics[width=1.0\linewidth, trim=-2.25cm 1.0cm 0cm 3.75cm]{fig/metrics/chr4_pan_HTTex1_gfa_multiqc_odgi_stats_svg}
         \label{fig:metrics-multiqc}
+        	\vspace{-2em}
     \end{subfigure}
@@
         \label{fig:metrics-str}
+        \vspace{-0.5em}
     \end{subfigure}
@@
 	\caption{Features of a 90-haplotype human pangenome graph of the exon 1 huntingtin gene (\textit{HTTexon1}): \textbf{(a)} Excerpt of vital statistics of the \textit{HTTexon1} graph displayed by MultiQC's ODGI module.
-The very high GC content of 73.0\% compared to a human genomic mean GC content of 40.9\% \cite{Piovesan2019} is in accordance with the literature (see for example \cite{Sathasivam2013, Neueder2017}).
+The very high GC content of 73.0\% compared to a human genomic mean GC content of 40.9\% \citep{Piovesan2019} is in accordance with the literature \citep{Neueder2017}.
 \textbf{(b)} Per nucleotide node degree distribution of CHM13 in the \textit{HTTexon1} graph.
@@
 \textbf{(e)} \textit{vg viz} nucleotide-level visualization of 10 gene alleles, CHM13, GRCH38 of the \textit{HTTexon1} graph focusing on the CAG variable repeat region.
-Figures \textbf{(b)-(e)} highlight the variant region around position 200 of CHM13 showing the variable number of glutamine residues of the different individuals as reported by \cite{Nance1999}.}
+Figures \textbf{(b)-(e)} highlight the variant region around position 200 of CHM13 showing the variable number of glutamine residues of the different individuals as reported by \citep{Nance1999}.}
 	\label{fig:metrics}

### 2021-11-09 manuscript/main.tex
write a sentence about stats/ML implied todo
give thanks

@@
 Because pangenome graphs can contain complete genomes, we are able to query them to polarize the information they contain in easily-interpretable and reusable pairwise formats that are widely supported in bioinformatics.
+ODGI also projects variation graphs into vector and matrix representations that allow the direct application of machine learning and statistical models to the pangenome.
+%By providing convenient linearizations and matrix transformations, ODGI also provides an interface for machine learning on pangenomes.
+%can be linearized into feature vector or matrix form, and ODGI thus provides a natural interface for machine learning and statistical methods that operate directly on data of this shape.
 We expect that ODGI will provide a reference interface between pangenomic and genomic approaches to understanding genome variation.
@@
 %As ODGI was designed to deal with the path depth of human pangenome graphs, it remains to be seen, if it will work equally well with scWGS data.
+
+\section*{Acknowledgements}
+
+We thank members of the HRPC ``Pangenomes Working Group'' for their insightful discussion and feedback, and members of the HPRC production teams for their development of resources used in our exposition.
 

### 2021-11-09 manuscript/main.tex
lingering parentheses

@@
 Pangenome graphs can be constructed by multiple sequence alignment~\citep{Lee_2002,Grasso_2004} or by transitively reducing an alignment between sequences to an equivalent, labeled sequence graph~\citep{Kehr_2014,Garrison_2019_thesis}.
-Current methods to build these graphs are still under active development~\citep{Li:2020,Armstrong:2020,pggb}, but they have largely settled on a common data model, represented in the Graphical Fragment Assembly (GFA) format~\citep{GFA}).
+Current methods to build these graphs are still under active development~\citep{Li:2020,Armstrong:2020,pggb}, but they have largely settled on a common data model, represented in the Graphical Fragment Assembly (GFA) format~\citep{GFA}.
 %\FIXME{may want to move GFA standard+VG to intro section 1}. AG: it is borderline, but it seems fine to have it here too.

### 2021-11-09 manuscript/main.tex
coca cola and typos

--- 2021-11-09 manuscript/main.tex
@@
 All changes on the graph can involve at most two $Node$ structs at a time (both edge and path step representations are doubly-linked).
-To avoid deadlocks, we acquire the node locks in ascending $Node.id$ order and release them in desceding order.
+To avoid deadlocks, we acquire the node locks in ascending $Node.id$ order and release them in descending order.
 In addition to node-local features of the graph, we must maintain some global information.
-Specifically, we record the start and end of paths, as well as a name to path id mapping in lock-free hashtables.
-The use of lock-free hashtables lets us avoid a global lock when looking up path or graph metadata, which would quickly become a bottleneck during parallel operations on the graph.
+Specifically, we record the start and end of paths, as well as a name to path id mapping in lock-free hash tables.
+The use of lock-free hash tables lets us avoid a global lock when looking up path or graph metadata, which would quickly become a bottleneck during parallel operations on the graph.
 By avoiding global locks, we implement many of the operations in ODGI using maximum parallelism available.
@@
 To scale to large graphs, we learn this projection in parallel via a HOGWILD! approach~\citep{niu2011hogwild}.
-Our approach can be seen as an adaptotion of SGD-based drawing to pangenome graphs~\citep{zheng2018graph}.
+Our approach can be seen as an adaptation of SGD-based drawing to pangenome graphs~\citep{zheng2018graph}.
 In parallel, each HOGWILD! thread updates node relative positions to best-match their nucleotide distance in the paths running through the graph.
@@
 ODGI is a state-of-the-art tool suite that enables users to explore and discover the underlying biology in pangenomes graphs, filling the gap that made pangenomic analyses difficult.
-It provides tools to easily transform, analyse, simplify, validate, and visualise pangenome graphs at large scale.
+It provides tools to easily transform, analyze, simplify, validate, and visualize pangenome graphs at large scale.
 In particular, lifting over annotations and linearizing nested graph structures place the suite as the bridge between traditional linear reference genome analysis and pangenome graphs.
@@
 Particularly for targets that involve complex variation, such as cancer, plant genomics and metagenomics, ODGI will facilitate disentangling,
-describing and analysing a much larger set of variation than previously was possible with tools that depend on short reads and reference genomes.
-Futhermore, users can even consider ODGI as a framework, taking advantage of its algorithms to develop new and more advanced tools that work on pangenome graphs,
+describing and analyzing a much larger set of variation than previously was possible with tools that depend on short reads and reference genomes.
+Furthermore, users can even consider ODGI as a framework, taking advantage of its algorithms to develop new and more advanced tools that work on pangenome graphs,
 thus expanding the type of possible pangenomic analyses available to the scientific community.
@@
 However, an interactive solution that combines the 1D and 2D layout of a graph with annotation and read mapping information across different zoom levels is still missing.
-Recent interactive browsers are reference centric~\citep{Beyer2019, Yokoyama2019, Durant2021, Liang2021} or focus primarily on 2D~\citep{Wick_2015, Gonnella2018}.
+Recent interactive browsers are reference-centric~\citep{Beyer2019, Yokoyama2019, Durant2021, Liang2021} or focus primarily on 2D~\citep{Wick_2015, Gonnella2018}.
 Our graph sorting and layout algorithms can provide the foundation for future tools of this type.
@@
 
-We thank members of the HRPC ``Pangenomes Working Group'' for their insightful discussion and feedback, and members of the HPRC production teams for their development of resources used in our exposition.
+We thank members of the HPRC ``Pangenomes Working Group'' for their insightful discussion and feedback, and members of the HPRC production teams for their development of resources used in our exposition.
 

### 2021-11-09 manuscript/submission/letter.tex
add letter

\documentclass[12pt,hidelinks,letterpaper]{article}
\usepackage[margin=0.75in,top=0.75in,footskip=0.5in]{geometry}
\usepackage{comment}

\usepackage{helvet}
\renewcommand{\familydefault}{\sfdefault}

% use Unicode characters - try changing the option if you run into troubles with special characters (e.g. umlauts)
\usepackage[utf8]{inputenc}

% clean citations
\usepackage{cite}

% hyperref makes references clicky. use \url{www.example.com} or \href{www.example.com}{description} to add a clicky url
\usepackage{nameref,hyperref}

% line numbers
%\usepackage[right]{lineno}

% improves typesetting in LaTeX
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% Remove % for double line spacing
%\usepackage{setspace} 
%\doublespacing

% use adjustwidth environment to exceed text width (see examples in text)
\usepackage{changepage}

% use \textcolor{color}{text} for colored text (e.g. highlight to-do areas)
\usepackage{color}

% this is required to include graphics
\usepackage{graphicx}

% use for have text wrap around figures
\usepackage{wrapfig}

%\usepackage{setspace}
%\doublespacing
\usepackage{multicol}

% document begins here
\begin{document}

\begin{flushright}
  Erik Garrison \\
  egarris5@uthsc.edu \\
  %+1 502 382 6005 \\
  %+39 320 244 2758 \\
  Department of Genetics, Genomics, and Informatics \\
  University of Tennessee Health Science Center \\
  Memphis, TN 38163, USA
\end{flushright}

%\hfill \break

\begin{flushleft}
  % madlib: replace with address
  Bioinformatics \\
  Oxford University Press \\
  Great Clarendon Street \\
  Oxford OX2 6DP, UK
\end{flushleft}

%\hfill \break
%\hfill \break

Dear editors,
%\hfill \break
%\hfill \break

In the included manuscript, we present the Optimized Dynamic Genome/Graph Implementation (ODGI) toolkit.
This system of interoperable methods provides researchers with a coherent set of approaches to understand the structure and significance of genome alignments encoded in pangenome graphs.
In addition to exposing the features of pangenome graphs to standard bioinformatic approaches based on pairwise comparisons, ranges, positions, feature vector and matrix models, ODGI provides unique methods to visualize and label large pangenome graphs like those currently being constructed in the Human Pangenome Project (HPP/HPRC).

Our description of ODGI is intended to describe a practical modality for working with pangenomes that ODGI implements.
As such, it is largely qualitative and theoretical.
As our focus is primarily conceptual, we have omitted performance based comparisons with the vg toolkit, which is the only similar existing system to ODGI.
However, we do demonstrate that ODGI is capable of efficiently working with large, complex pangenomes.
This high-level focus on the toolkit necessitates the omission of deep algorithmic details that might be of formal interest to the community.
Likewise, there are enough small pieces to the toolkit that a full description of all parts may not be helpful.
We hope that through review we can strike a suitable balance of focus on such detail.

To exposit this system, we take a handful of examples from the ongoing work in the HPP/HPRC.
As members of this project, we have followed the data usage agreement and contacted the steering committee to indicate this use, and obtained authorization to submit this manuscript to you and as a public preprint.
The discussion of how methods papers published before the main project analysis should reflect the HPP/HPRC is ongoing, and some changes, such as the potential inclusion of an HPRC ``banner author'' may eventually be required should \textit{Bioinformatics} choose to publish a version of our manuscript.

ODGI has provided us with a critical resource during the initial phase of the HPP/HPRC.
It is a fundamentally practical system that we have developed to answer basic questions about the pangenome graphs that the community is still learning how to build.
We trust that this toolkit will be of wide interest to the ``pangenomic'' community that is now building up methods to understand variation between hundreds of complete genome assemblies in many species.

%\hfill \break
\hfill \break

%\begin{flushright}
%\hspace*{100mm}
\indent Sincerely,\\
\hfill \break
\includegraphics[width=0.38\textwidth]{signature_Erik_Garrison.pdf}
\hfill \break
\indent Erik Garrison
%\end{flushright}

\begin{comment}
\hfill \break
\begingroup
\let\oldthebibliography\thebibliography
\let\endoldthebibliography\endthebibliography
\renewenvironment{thebibliography}[1]{
  \begin{oldthebibliography}{#1}
    \setlength{\itemsep}{0em}
    \setlength{\parskip}{0em}
}
{
  \end{oldthebibliography}
}
\renewcommand{\section}[2]{}%
\bibliographystyle{unsrt}
%\subsection*{References}
{\small \bibliography{references}}
\endgroup
\end{comment}

\end{document}

### 2021-11-09 manuscript/submission/letter.tex
tweaks
make all happy

@@
 Dear editors,
-%\hfill \break
+\hfill \break
 %\hfill \break
@@
 
-Our description of ODGI is intended to describe a practical modality for working with pangenomes that ODGI implements.
+Our description of ODGI is intended to present a practical modality for working with pangenomes that ODGI implements.
 As such, it is largely qualitative and theoretical.
-As our focus is primarily conceptual, we have omitted performance based comparisons with the vg toolkit, which is the only similar existing system to ODGI.
+%As our focus is primarily conceptual, we have omitted performance based comparisons with the vg toolkit, which is the only similar existing system to ODGI.
 However, we do demonstrate that ODGI is capable of efficiently working with large, complex pangenomes.
@@
 %\hfill \break
+
 \hfill \break
@@
 \indent Sincerely,\\
-\hfill \break
-\includegraphics[width=0.38\textwidth]{signature_Erik_Garrison.pdf}
+%\hfill \break
+\includegraphics[width=0.34\textwidth]{signature_Erik_Garrison.pdf}
 \hfill \break

### 2021-11-09 manuscript/main.tex
pwg name

@@
 
-We thank members of the HPRC ``Pangenomes Working Group'' for their insightful discussion and feedback, and members of the HPRC production teams for their development of resources used in our exposition.
+We thank members of the HPRC Pangenome Working Group for their insightful discussion and feedback, and members of the HPRC production teams for their development of resources used in our exposition.
 

### 2021-11-10 manuscript/fig_extract_viz_draw_position_untangle.tex
be consistent with the `odgi untangle` output. C4A is before C4B in the reference

@@
 Node labels were annoted by using \textit{odgi position}.
-        \textbf{(h)} Visualization of \textit{odgi untangle} output in the C4 pangenome graph: the plots show the copy number status of the sequences in the C4 region with respect to the grch38 reference sequence, making clear, for example, that in HG00438\#2, the \textit{C4B} gene is missing.
+        \textbf{(h)} Visualization of \textit{odgi untangle} output in the C4 pangenome graph: the plots show the copy number status of the sequences in the C4 region with respect to the grch38 reference sequence, making clear, for example, that in HG00438\#2, the \textit{C4A} gene is missing.
 \vspace{-1em}

### 2021-11-10 manuscript/fig_extract_viz_draw_position_untangle.tex
missing `s`

@@
         \textbf{(d)} \textit{odgi viz} color by strandness: the red paths indicate the haplotypes that were assembled in reverse with respect to the 2 reference genomes.
-        \textbf{(e)} \textit{odgi viz} color by path depth: using the Spectra color palette with 4 level of path depths, white indicates no depth, while grey, red, and yellow indicate depth 1, 2, and greater than or equal to 3, respectively.
+        \textbf{(e)} \textit{odgi viz} color by path depth: using the Spectra color palette with 4 levels of path depths, white indicates no depth, while grey, red, and yellow indicate depth 1, 2, and greater than or equal to 3, respectively.
         Coloring by path depth, we can see that the two references present two different allele copies of the C4 genes, both of them including the HERV sequence.

### 2021-11-10 scripts/fig_extract_viz_draw_position_untangle.sh
put a temporary, but working link for chr6

@@ -4,7 +4,7 @@ prefix_chr6_smooth=chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth
 
 # Download and build the graph
 # FIXME with the final link
-wget wget /lizardfs/erikg/HPRC/year1v2genbank/wgg.88/chr6.pan/${prefix_chr6_smooth}.gfa.gz
+wget http://hypervolu.me/~guarracino/${prefix_chr6_smooth}.gfa.gz
 gunzip ${prefix_chr6_smooth}.gfa.gz
 odgi build -g ${prefix_chr6_smooth}.gfa -o ${prefix_chr6_smooth}.og -t 16 -P
 

### 2022-01-12 scripts/fig_extract_viz_draw_position_untangle.sh
clean untangle plotting code; add C4A/C4B annotation in the untangle scatterplots

@@ -78,49 +78,4 @@ odgi view -i ${prefix_chr6_smooth}.C4.sorted.og -g > ${prefix_chr6_smooth}.C4.so
 
 # cat ${prefix_chr6_smooth}.C4.sorted.untangle.bed | grep '^chm13\|^grch38\|^HG00438\|^HG0107\|^HG01952'
 
-# R
-library(ggplot2)
-#library(knitr)
-x <- read.table('/home/guarracino/Desktop/pangenomics/odgi_pub/chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.C4.sorted.untangle.bed', sep = '\t', header = T, comment.char="$")
-x$query.name <- gsub(":.*","",x$query.name)
-x$query.name <- gsub("#J.*","",x$query.name)
-
-ggplot(
-  subset(x, query.name %in% c(
-    ##"chm13#chr6:31825251-31908851",
-    #"grch38#chr6:31972046-32055647",
-    ##"HG00438#1#JAHBCB010000040.1:24269348-24320210",
-    #"HG00438#2#JAHBCA010000042.1:24398231-24449090",
-    ##"HG01071#1#JAHBCF010000017.1:706180-783405",
-    #"HG01071#2#JAHBCE010000076.1:7794179-7897781",
-    #"HG01952#1#JAHAME010000044.1:28380191-28451052",
-    #"HG01952#2#JAHAMD010000016.1:31974838-32052065",
-
-    #"chm13#chr6",
-    "grch38#chr6",
-    #"HG00438#1",
-    "HG00438#2",
-    #"HG01071#1",
-    "HG01071#2",
-    "HG01952#1",
-    "HG01952#2"
-    )
-  ), aes(x=query.start, xend=query.end, y=ref.start, yend=ref.end)) +
-    geom_segment(size=0.3) +
-    facet_grid(. ~ query.name) +
-    coord_fixed() +
-    #theme_bw() +
-    theme(
-      text = element_text(size = 12.6),
-      axis.text.x = element_text(size = 12, angle = 90),
-      axis.text.y = element_text(size = 12),
-
-      #panel.grid.minor = element_line(size = 0.125),
-      #panel.grid.major = element_line(size = 0.25)
-    )  +
-      xlab("Query start") +
-      ylab("Reference start")
-filename <- 'chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_C4_sorted_untangle_bed.pdf'
[…]

### 2022-01-12 scripts/plot_untangle.R

@@ -1,0 +1,71 @@
+args <- commandArgs()
+path_untangle_bed <- args[6]
+
+library(ggplot2)
+#library(knitr)
+x <- read.table(path_untangle_bed, sep = '\t', header = T, comment.char = "$")
+x$query.name <- gsub(":.*", "", x$query.name)
+x$query.name <- gsub("#J.*", "", x$query.name)
+
+x_subset <- subset(x, query.name %in% c(
+  ##"chm13#chr6:31825251-31908851",
+  #"grch38#chr6:31972046-32055647",
+  ##"HG00438#1#JAHBCB010000040.1:24269348-24320210",
+  #"HG00438#2#JAHBCA010000042.1:24398231-24449090",
+  ##"HG01071#1#JAHBCF010000017.1:706180-783405",
+  #"HG01071#2#JAHBCE010000076.1:7794179-7897781",
+  #"HG01952#1#JAHAME010000044.1:28380191-28451052",
+  #"HG01952#2#JAHAMD010000016.1:31974838-32052065",
+
+  #"chm13#chr6",
+  "grch38#chr6",
+  #"HG00438#1",
+  "HG00438#2",
+  #"HG01071#1",
+  "HG01071#2",
+  "HG01952#1",
+  "HG01952#2"
+)
+)
+
+p <- ggplot(
+  x_subset, aes(x = query.start, xend = query.end, y = ref.start, yend = ref.end)) +
+  geom_segment(size = 0.3) +
+  facet_grid(. ~ query.name) +
+  coord_fixed() +
+  #theme_bw() +
+  theme(
+    text = element_text(size = 12.6),
+    axis.text.x = element_text(size = 12, angle = 90),
+    axis.text.y = element_text(size = 12),
+
+    #panel.grid.minor = element_line(size = 0.125),
+    #panel.grid.major = element_line(size = 0.25)
+  ) +
+  xlab("Query start") +
+  ylab("Reference start")
+
+
+# C4A annotation
+# https://stackoverflow.com/questions/65013846/why-geom-rect-looking-different-across-facets
+p <- p +
+  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 31982057 - 31972046, ymax = 32002681 - 31972046, fill = "#EE2222", alpha = .2, color = "#444444", size = 0.1)
+#https://stackoverflow.com/questions/11889625/annotating-text-on-individual-facet-in-ggplot2
+C4A_ann_text <- data.frame(
[…]

### 2022-01-12 manuscript/fig_extract_viz_draw_position_untangle.tex
missing the and locus in italic
fix bandage reference and add it in the main

@@
         The entirely grey paths have one copy of these genes.
-        HG01071\#2 presents 3 copies of locus (orange), of which one contains the HERV sequence (gray in the middle of the orange).
+        HG01071\#2 presents 3 copies of the \textit{locus} (orange), of which one contains the HERV sequence (gray in the middle of the orange).
         In HG01952\#1, the HERV sequence is absent.
-        \textbf{(f)} \textit{Bandage} layout of the C4 pangenome graph, annotated by using \textit{odgi position}.
+        \textbf{(f)} \textit{Bandage}~\citep{Wick_2015} layout of the C4 pangenome graph, annotated by using \textit{odgi position}.
 Green nodes indicate the C4 genes.

### 2022-01-13 manuscript/main.tex
reformatting footer for bioRxiv
explain variation graphs and reduce pangenome/variation graph confusion

@@
 \newcommand{\FIXME}[1]{\red{[FIXME: #1]}}
+\newcommand{\REVIEWED}[1]{{\textcolor{Red}{#1}}}
 
@@
 
-\copyrightyear{XXXX} \pubyear{XXXX}
+\copyrightyear{2021} \pubyear{XXXX}
 
@@
 \textbf{Results:}
-We wrote ODGI, a novel suite of tools that implements scalable algorithms and has an efficient in-memory representation of DNA variation graphs.
+We wrote ODGI, a novel suite of tools that implements scalable algorithms and has an efficient in-memory representation of DNA \REVIEWED{pangenome graphs in the form of variation graphs}.
 ODGI includes tools for detecting complex regions, extracting \textit{loci}, removing artifacts, exploratory analysis, manipulation, validation, and visualization.
@@
 In contrast to reference-based approaches which relate sequences to a single genome, these data structures encode the mutual relationships between all the genomes represented.
-In pangenome graphs~\citep{Paten:2017}, homologous regions between genomes are compressed into a single representative of all alleles present in the pangenome.
+%In pangenome graphs~\citep{Paten:2017}, homologous regions between genomes are compressed into a single representative of all alleles present in the pangenome.
+\REVIEWED{A class of methods to represent pangenomes involves the sequence graphs~\citep{Hein1989, Paten:2017}, where homologous regions between genomes are compressed into a single representative of all alleles present in the pangenome.
+In node-labeled sequence graphs, nodes indicate DNA sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph.
+A bidirected sequence graph represents both strands of DNA.
+On this model, variation graphs add the concept of path to embed linear sequences (e.g., genomes or haplotypes) into the graph~\citep{Garrison:2018}.
+}
 These flexible models let us encode any kind of variation, allowing the generation of comprehensive data systems that builds the basis for the analyses of genome evolution.
@@
 \section{Implementation}
-The ODGI toolkit builds on existing approaches to efficiently store and manipulate variation graphs~\citep{Garrison:2018}.
-Similar to other efficient libraries presenting the \textsc{HandleGraph} model~\citep{Eizenga_2020_BX}, the implementation of ODGI's tools rests on three key properties which hold for most pangenome variation graphs:
+The ODGI toolkit builds on existing approaches to efficiently store and manipulate \REVIEWED{pangenome graphs in the form of} variation graphs~\citep{Garrison:2018}.
+Similar to other efficient libraries presenting the \textsc{HandleGraph} model~\citep{Eizenga_2020_BX}, the implementation of ODGI's tools rests on three key properties which hold for most pangenome graphs:
 
@@
 %The $\delta between two nodes is computed as their distance in the graph vector (i.e., the difference between the node offsets).
-This keeps the size of the encoding small, per common variation graph property (2).
+This keeps the size of the encoding small, per common pangenome graph property (2).
 We define the edges and path steps traversing the node in terms of this alphabet of $\delta$'s.

### 2022-01-13 manuscript/main.tex
state that odgi viz/draw produce static pictures

@@
 Pangenome graph visualization is one of the first steps to gain insight into the mutual relationship between the sequences in the graph and their variation.
-We pursue a novel approach to visualization with \textit{odgi viz} and \textit{odgi draw}, two tools which provide scalable ways of generating pictures of the high-level structure of large pangenome graphs.
+We pursue a novel approach to visualization with \textit{odgi viz} and \textit{odgi draw}, two tools which provide scalable ways of generating \REVIEWED{static} pictures of the high-level structure of large pangenome graphs.
 

### 2022-01-14 scripts/build_time.sh
update performance measurement scripts

@@ -1,42 +1,80 @@
-wget https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/scratch/2021_11_04_pggb_wgg.87/chroms/chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
+wget https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/scratch/2021_11_16_pggb_wgg.88/chroms/chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
 gunzip chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
 
 G=chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa
 
-# t1
-for i in {1..10}
+for T in 1 2 4 8 16
 do
-    /usr/bin/time odgi build -g "$G" -o "$G".og -t 1 2> chr6_odgi_build_time_t1_"$i"
+    for i in {1..10}
+    do
+        /usr/bin/time --verbose odgi build -g "$G" -o "$G".og -t "$T" 2> chr6_odgi_build_time_"$T"_"$i"
+        /usr/bin/time --verbose vg convert -g -x -t "$T" "$G" 2> chr6_vg_convert_time_"$T"_"$i" 1> "$G".xg
+    done    
 done
 
-#t2
-for i in {1..10}
-do
-    /usr/bin/time odgi build -g "$G" -o "$G".og -t 2 2> chr6_odgi_build_time_t2_"$i"
-done
-
-#t4
-for i in {1..10}
-do
-    /usr/bin/time odgi build -g "$G" -o "$G".og -t 4 2> chr6_odgi_build_time_t4_"$i"
-done
-
-#t8
-for i in {1..10}
-do
-    /usr/bin/time odgi build -g "$G" -o "$G".og -t 8 2> chr6_odgi_build_time_t8_"$i"
-done
-
-#t16
-for i in {1..10}
-do
-    /usr/bin/time odgi build -g "$G" -o "$G".og -t 16 2> chr6_odgi_build_time_t16_"$i"
-done
+if [ -f odgi_build_threads_time.csv ] ; then
+    rm odgi_build_threads_time.csv
+fi
+if [ -f vg_convert_threads_time.csv ] ; then
+    rm vg_convert_threads_time.csv
+fi
 
 for T in 1 2 4 8 16
 do
[…]

### 2022-01-14 scripts/extract_time.sh

@@ -1,0 +1,64 @@
+wget https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/scratch/2021_11_16_pggb_wgg.88/chroms/chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
+gunzip chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
+
+G=chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa
+
+odgi paths -i "$G" -L -l -t 16 > "$G".paths
+
+# build our graphs by number of haplotypes
+for H in 1 2 4 8 16 32 64
+do
+    if [[ H == 1 ]]; then
+        grep grch38 "$G".paths | sed 's/\t1\t/\t0\t/g' > "$G"."$H"haps.bed
+        cut -f 1 "$G"."$H"haps.bed > "$G"."$H"haps
+    elif [[ H == 2 ]]; then
+        grep -E 'grch38|chm13' "$G".paths | sed 's/\t1\t/\t0\t/g' > "$G"."$H"haps.bed
+        cut -f 1 "$G"."$H"haps.bed > "$G"."$H"haps
+    else
+        grep -E $(echo $(head -n $(expr "$H" - 2) "$G".non_ref.haps | sed -z 's/\n/\|/g')"grch38|chm13") "$G".paths | sed 's/\t1\t/\t0\t/g' > "$G"."$H"haps.bed
+        cut -f 1 "$G"."$H"haps.bed > "$G"."$H"haps
+    fi
+    odgi extract -i "$G" -o - -b "G$"."$H"haps.bed -p "$G"."$H"haps -t 16 -P | odgi sort -i - -O -o "$G"."$H"haps.og
+    odgi view -i "$G"."$H"haps.og -g > "$G"."$H"haps.og.gfa
+    for i in {1..10}
+    do
+        /usr/bin/time --verbose odgi build -g "$G"."$H"haps.og.gfa -o "$G"."$H"haps.og.gfa.og -t 16
+        /usr/bin/time --verbose vg convert -g -x -t 16 "$G"."$H"haps.og.gfa > "$G"."$H"haps.og.gfa.xg
+    done
+done
+
+for H in 1 2 4 8 16 32 64
+do
+    for T in 1 2 4 8 16
+    do
+        for i in {1..10}
+        do
[…]

### 2022-01-14 scripts/pos_time.sh

@@ -1,0 +1,64 @@
+wget https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/scratch/2021_11_16_pggb_wgg.88/chroms/chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
+gunzip chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
+
+G=chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa
+
+odgi paths -i "$G" -L -l -t 16 > "$G".paths
+
+# build our graphs by number of haplotypes
+for H in 1 2 4 8 16 32 64
+do
+    if [[ H == 1 ]]; then
+        grep grch38 "$G".paths | sed 's/\t1\t/\t0\t/g' > "$G"."$H"haps.bed
+        cut -f 1 "$G"."$H"haps.bed > "$G"."$H"haps
+    elif [[ H == 2 ]]; then
+        grep -E 'grch38|chm13' "$G".paths | sed 's/\t1\t/\t0\t/g' > "$G"."$H"haps.bed
+        cut -f 1 "$G"."$H"haps.bed > "$G"."$H"haps
+    else
+        grep -E $(echo $(head -n $(expr "$H" - 2) "$G".non_ref.haps | sed -z 's/\n/\|/g')"grch38|chm13") "$G".paths | sed 's/\t1\t/\t0\t/g' > "$G"."$H"haps.bed
+        cut -f 1 "$G"."$H"haps.bed > "$G"."$H"haps
+    fi
+    odgi extract -i "$G" -o - -b "G$"."$H"haps.bed -p "$G"."$H"haps -t 16 -P | odgi sort -i - -O -o "$G"."$H"haps.og
+    odgi view -i "$G"."$H"haps.og -g > "$G"."$H"haps.og.gfa
+    for i in {1..10}
+    do
+        /usr/bin/time --verbose odgi build -g "$G"."$H"haps.og.gfa -o "$G"."$H"haps.og.gfa.og -t 16
+        /usr/bin/time --verbose vg convert -g -x -t 16 "$G"."$H"haps.og.gfa > "$G"."$H"haps.og.gfa.xg
+    done
+done
+
+for H in 1 2 4 8 16 32 64
+do
+    for T in 1 2 4 8 16
+    do
+        for i in {1..10}
+        do
+            # 4701798 nodes, so we take the one in the middle
[…]

### 2022-01-14 scripts/tips.sh

@@ -1,24 +1,0 @@
-#!/bin/bash
-
-wget https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/scratch/2021_06_27_pggb/chroms/chr8.pan.fa.c3d3224.7748b33.395c7f4.smooth.og.gz
-gunzip chr8.pan.fa.c3d3224.7748b33.395c7f4.smooth.og.gz
-
-# paths
-odgi paths -i chr8.pan.fa.c3d3224.7748b33.395c7f4.smooth.og -L > chr8.pan.fa.c3d3224.7748b33.395c7f4.smooth.og.paths
-# ref paths
-grep "chm13\|grch38" chr8.pan.fa.c3d3224.7748b33.395c7f4.smooth.og.paths > chr8.pan.fa.c3d3224.7748b33.395c7f4.smooth.og.ref_paths
-# tips
-odgi tips -i chr8.pan.fa.c3d3224.7748b33.395c7f4.smooth.og -t 28 -P -R chr8.pan.fa.c3d3224.7748b33.395c7f4.smooth.og.ref_paths -v chr8.pan.fa.c3d3224.7748b33.395c7f4.smooth.og.tips.no_ref_hit.tsv -w 50000 -j > chr8.pan.fa.c3d3224.7748b33.395c7f4.smooth.og.tips.bed
-bedtools sort -i chr8.pan.fa.c3d3224.7748b33.395c7f4.smooth.og.tips.bed > chr8.pan.fa.c3d3224.7748b33.395c7f4.smooth.og.tips.bed.sorted.bed
-
-#### TODO ####
-# we got this cytobands annoation after personal communication with Mitchell R. Vollger
-grep -P "^chm13#chr8\t" chm13.CytoBandIdeo.v2.txt | grep p23.1
-
-odgi sort -i chr8.pan.fa.c3d3224.7748b33.395c7f4.smooth.og -t 28 -G 10 -x 100 -o chr8.pan.fa.c3d3224.7748b33.395c7f4.smooth.og.Y.og -Y -P
-odgi extract -i chr8.pan.fa.c3d3224.7748b33.395c7f4.smooth.og.Y.og -E -r chm13#chr8:6054497-13066158 -t 28 -P -o chr8.pan.fa.c3d3224.7748b33.395c7f4.smooth.og.Y.og.p23.1.og
[…]

### 2022-01-14 scripts/viz_time.sh

@@ -1,0 +1,58 @@
+wget https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/scratch/2021_11_16_pggb_wgg.88/chroms/chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
+gunzip chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
+
+G=chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa
+
+odgi paths -i "$G" -L -l -t 16 > "$G".paths
+
+# build our graphs by number of haplotypes
+for H in 1 2 4 8 16 32 64
+do
+    if [[ H == 1 ]]; then
+        grep grch38 "$G".paths | sed 's/\t1\t/\t0\t/g' > "$G"."$H"haps.bed
+        cut -f 1 "$G"."$H"haps.bed > "$G"."$H"haps
+    elif [[ H == 2 ]]; then
+        grep -E 'grch38|chm13' "$G".paths | sed 's/\t1\t/\t0\t/g' > "$G"."$H"haps.bed
+        cut -f 1 "$G"."$H"haps.bed > "$G"."$H"haps
+    else
+        grep -E $(echo $(head -n $(expr "$H" - 2) "$G".non_ref.haps | sed -z 's/\n/\|/g')"grch38|chm13") "$G".paths | sed 's/\t1\t/\t0\t/g' > "$G"."$H"haps.bed
+        cut -f 1 "$G"."$H"haps.bed > "$G"."$H"haps
+    fi
+    odgi extract -i "$G" -o - -b "G$"."$H"haps.bed -p "$G"."$H"haps -t 16 -P | odgi sort -i - -O -o "$G"."$H"haps.og
+    odgi view -i "$G"."$H"haps.og -g > "$G"."$H"haps.og.gfa
+    for i in {1..10}
+    do
+        /usr/bin/time --verbose odgi build -g "$G"."$H"haps.og.gfa -o "$G"."$H"haps.og.gfa.og -t 16
+        /usr/bin/time --verbose vg convert -g -x -t 16 "$G"."$H"haps.og.gfa > "$G"."$H"haps.og.gfa.xg
+    done
+done
+
+for H in 1 2 4 8 16 32 64
+do
+    for i in {1..10}
+    do
[…]

### 2022-01-15 scripts/build_time.sh
1..2, bug fixes, cheating for vg find

@@ -7,6 +7,7 @@ for T in 1 2 4 8 16
 do
     for i in {1..10}
     do
+        echo "T: ""$T"" i: ""$i"
         /usr/bin/time --verbose odgi build -g "$G" -o "$G".og -t "$T" 2> chr6_odgi_build_time_"$T"_"$i"
         /usr/bin/time --verbose vg convert -g -x -t "$T" "$G" 2> chr6_vg_convert_time_"$T"_"$i" 1> "$G".xg
     done    
@@ -31,7 +32,11 @@ done
 # TODO
 #Rscript time.R mean_build_time.csv mean_build_time.pdf
 
-odgi paths -i "$G" -L -l -t 16 > "$G".paths
+odgi build -g "$G" -o "$G".og -t 16 -P
+odgi paths -i "$G".og -L -l -t 16 > "$G".paths
+grep -E 'grch38|chm13' "$G".paths -v > "$G".non_ref.paths
+cut -f 1,2 "$G".non_ref.paths -d '#' | sort| uniq > "$G".non_ref.haps
+
 
 if [ -f odgi_build_haps_time.csv ] ; then
     rm odgi_build_haps_time.csv
@@ -41,23 +46,27 @@ if [ -f vg_convert_haps_time.csv ] ; then
 fi
 
 for H in 1 2 4 8 16 32 64
-do
-    if [[ H == 1 ]]; then
+do  
+    echo "hap: ""$H" 
+    if [[ "$H" == 1 ]]; then
+        echo "H == 1"
         grep grch38 "$G".paths | sed 's/\t1\t/\t0\t/g' > "$G"."$H"haps.bed
         cut -f 1 "$G"."$H"haps.bed > "$G"."$H"haps
-    elif [[ H == 2 ]]; then
+    elif [[ "$H" == 2 ]]; then
+        echo "H == 2"
         grep -E 'grch38|chm13' "$G".paths | sed 's/\t1\t/\t0\t/g' > "$G"."$H"haps.bed
         cut -f 1 "$G"."$H"haps.bed > "$G"."$H"haps
     else
         grep -E $(echo $(head -n $(expr "$H" - 2) "$G".non_ref.haps | sed -z 's/\n/\|/g')"grch38|chm13") "$G".paths | sed 's/\t1\t/\t0\t/g' > "$G"."$H"haps.bed
[…]

### 2022-01-15 scripts/extract_time.sh

@@ -3,36 +3,55 @@ gunzip chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
 
 G=chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa
 
-odgi paths -i "$G" -L -l -t 16 > "$G".paths
+odgi build -g "$G" -o "$G".og -t 16 -P
+odgi paths -i "$G".og -L -l -t 16 > "$G".paths
+grep -E 'grch38|chm13' "$G".paths -v > "$G".non_ref.paths
+cut -f 1,2 "$G".non_ref.paths -d '#' | sort| uniq > "$G".non_ref.haps
 
-# build our graphs by number of haplotypes
 for H in 1 2 4 8 16 32 64
-do
-    if [[ H == 1 ]]; then
+do  
+    echo "hap: ""$H" 
+    if [[ "$H" == 1 ]]; then
+        echo "H == 1"
         grep grch38 "$G".paths | sed 's/\t1\t/\t0\t/g' > "$G"."$H"haps.bed
         cut -f 1 "$G"."$H"haps.bed > "$G"."$H"haps
-    elif [[ H == 2 ]]; then
+    elif [[ "$H" == 2 ]]; then
+        echo "H == 2"
         grep -E 'grch38|chm13' "$G".paths | sed 's/\t1\t/\t0\t/g' > "$G"."$H"haps.bed
         cut -f 1 "$G"."$H"haps.bed > "$G"."$H"haps
     else
         grep -E $(echo $(head -n $(expr "$H" - 2) "$G".non_ref.haps | sed -z 's/\n/\|/g')"grch38|chm13") "$G".paths | sed 's/\t1\t/\t0\t/g' > "$G"."$H"haps.bed
         cut -f 1 "$G"."$H"haps.bed > "$G"."$H"haps
     fi
-    odgi extract -i "$G" -o - -b "G$"."$H"haps.bed -p "$G"."$H"haps -t 16 -P | odgi sort -i - -O -o "$G"."$H"haps.og
+    odgi extract -i "$G".og -o - -b "$G"."$H"haps.bed -p "$G"."$H"haps -t 16 -P | odgi sort -i - -O -o "$G"."$H"haps.og
     odgi view -i "$G"."$H"haps.og -g > "$G"."$H"haps.og.gfa
-    for i in {1..10}
+    # replace the output path names with the original ones, because we always take the full paths
[…]

### 2022-01-15 scripts/pos_time.sh

@@ -3,39 +3,59 @@ gunzip chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
 
 G=chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa
 
-odgi paths -i "$G" -L -l -t 16 > "$G".paths
+odgi build -g "$G" -o "$G".og -t 16 -P
+odgi paths -i "$G".og -L -l -t 16 > "$G".paths
+grep -E 'grch38|chm13' "$G".paths -v > "$G".non_ref.paths
+cut -f 1,2 "$G".non_ref.paths -d '#' | sort| uniq > "$G".non_ref.haps
 
-# build our graphs by number of haplotypes
 for H in 1 2 4 8 16 32 64
-do
-    if [[ H == 1 ]]; then
+do  
+    echo "hap: ""$H" 
+    if [[ "$H" == 1 ]]; then
+        echo "H == 1"
         grep grch38 "$G".paths | sed 's/\t1\t/\t0\t/g' > "$G"."$H"haps.bed
         cut -f 1 "$G"."$H"haps.bed > "$G"."$H"haps
-    elif [[ H == 2 ]]; then
+    elif [[ "$H" == 2 ]]; then
+        echo "H == 2"
         grep -E 'grch38|chm13' "$G".paths | sed 's/\t1\t/\t0\t/g' > "$G"."$H"haps.bed
         cut -f 1 "$G"."$H"haps.bed > "$G"."$H"haps
     else
         grep -E $(echo $(head -n $(expr "$H" - 2) "$G".non_ref.haps | sed -z 's/\n/\|/g')"grch38|chm13") "$G".paths | sed 's/\t1\t/\t0\t/g' > "$G"."$H"haps.bed
         cut -f 1 "$G"."$H"haps.bed > "$G"."$H"haps
     fi
-    odgi extract -i "$G" -o - -b "G$"."$H"haps.bed -p "$G"."$H"haps -t 16 -P | odgi sort -i - -O -o "$G"."$H"haps.og
+    odgi extract -i "$G".og -o - -b "$G"."$H"haps.bed -p "$G"."$H"haps -t 16 -P | odgi sort -i - -O -o "$G"."$H"haps.og
     odgi view -i "$G"."$H"haps.og -g > "$G"."$H"haps.og.gfa
-    for i in {1..10}
+    # replace the output path names with the original ones, because we always take the full paths
[…]

### 2022-01-15 scripts/viz_time.sh

@@ -3,33 +3,36 @@ gunzip chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
 
 G=chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa
 
-odgi paths -i "$G" -L -l -t 16 > "$G".paths
+odgi build -g "$G" -o "$G".og -t 16 -P
+odgi paths -i "$G".og -L -l -t 16 > "$G".paths
+grep -E 'grch38|chm13' "$G".paths -v > "$G".non_ref.paths
+cut -f 1,2 "$G".non_ref.paths -d '#' | sort| uniq > "$G".non_ref.haps
 
-# build our graphs by number of haplotypes
 for H in 1 2 4 8 16 32 64
-do
-    if [[ H == 1 ]]; then
+do  
+    echo "hap: ""$H" 
+    if [[ "$H" == 1 ]]; then
+        echo "H == 1"
         grep grch38 "$G".paths | sed 's/\t1\t/\t0\t/g' > "$G"."$H"haps.bed
         cut -f 1 "$G"."$H"haps.bed > "$G"."$H"haps
-    elif [[ H == 2 ]]; then
+    elif [[ "$H" == 2 ]]; then
+        echo "H == 2"
         grep -E 'grch38|chm13' "$G".paths | sed 's/\t1\t/\t0\t/g' > "$G"."$H"haps.bed
         cut -f 1 "$G"."$H"haps.bed > "$G"."$H"haps
     else
         grep -E $(echo $(head -n $(expr "$H" - 2) "$G".non_ref.haps | sed -z 's/\n/\|/g')"grch38|chm13") "$G".paths | sed 's/\t1\t/\t0\t/g' > "$G"."$H"haps.bed
         cut -f 1 "$G"."$H"haps.bed > "$G"."$H"haps
     fi
-    odgi extract -i "$G" -o - -b "G$"."$H"haps.bed -p "$G"."$H"haps -t 16 -P | odgi sort -i - -O -o "$G"."$H"haps.og
+    odgi extract -i "$G".og -o - -b "$G"."$H"haps.bed -p "$G"."$H"haps -t 16 -P | odgi sort -i - -O -o "$G"."$H"haps.og
     odgi view -i "$G"."$H"haps.og -g > "$G"."$H"haps.og.gfa
-    for i in {1..10}
-    do
[…]

### 2022-01-15 scripts/extract_time.sh
extracty

@@ -41,8 +41,8 @@ do
       fi
     done < "$G"."$H"haps.og.gfa
 
-    odgi build -g "$G"."$H"haps.og.gfa -o "$G"."$H"haps.og.gfa.og -t 16 -P
-    vg convert -g -x -t 16 "$G"."$H"haps.og.gfa > "$G"."$H"haps.og.gfa.xg
+    odgi build -g "$G"."$H"haps.og.gfa.orig.gfa -o "$G"."$H"haps.og.gfa.og -t 16 -P
+    vg convert -g -x -t 16 "$G"."$H"haps.og.gfa.orig.gfa > "$G"."$H"haps.og.gfa.xg
 done
 
 for H in 1 2 4 8 16 32 64
@@ -53,7 +53,6 @@ do
         do
             echo "H: ""$H"" T: ""$T"" i: ""$i"
             /usr/bin/time --verbose odgi extract -i "$G"."$H"haps.og.gfa.og -t "$T" -c 10 -r grch38#chr6:58553888-59829934 -o "$G"."$H"haps.og.gfa.og.centromeres.og 2> chr6_odgi_extract_time_"$T"_"$H"_"$i"
-            # TODO this never finished with 28 threads so far, maybe we can't run it!
             /usr/bin/time --verbose vg chunk -x "$G"."$H"haps.og.gfa.xg -t "$T" -p grch38#chr6:58553888-59829934 -c 10 2> chr6_vg_chunk_time_"$T"_"$H"_"$i" 1> "$G"."$H"haps.og.gfa.xg.centromeres.vg
         done
     done

### 2022-01-19 manuscript/main.tex
do not explicit that path colors are derived from a hash of the path names
Conflict of Interest
ORCIDs, enable hyperref in black

@@
 
-\usepackage[draft]{hyperref}
+\usepackage[]{hyperref}
+\hypersetup{colorlinks=true,urlcolor=black,linkcolor=black,citecolor=black}
 \usepackage{makecell}
@@
 \usepackage[usenames,dvipsnames]{xcolor}
+
+\usepackage{orcidlink}
 
@@
 \author[Guarracino, Heumos \textit{et~al}.]{
-Andrea~Guarracino\,$^{\text{\sfb 1} \dagger}$,
-Simon~Heumos\,$^{\text{\sfb 2} \dagger}$,
-Sven~Nahnsen\,$^{\text{\sfb 2,3}}$,
-Pjotr~Prins\,$^{\text{\sfb 4}}$,
-and~Erik~Garrison\,$^{\text{\sfb 4}*}$
+Andrea~Guarracino\orcidlink{0000-0001-9744-131X}\,$^{\text{\sfb 1} \dagger}$,
+Simon~Heumos\orcidlink{0000-0003-3326-817X},$^{\text{\sfb 2} \dagger}$
+Sven~Nahnsen\orcidlink{https://orcid.org/0000-0002-4375-0691}$^{\text{\sfb 2,3}}$,
+Pjotr \\Prins\orcidlink{0000-0002-8021-9162}$^{\text{\sfb 4}}$,
+and~Erik~Garrison\orcidlink{0000-0003-3821-631X}$^{\text{\sfb 4}*}$
 }
@@
 The meaning of the colors depends on how \textit{odgi viz} is executed.
-By default, path colors are derived from a hash of the path name (Fig.~\ref{fig:odgi_viz}\textbf{b}).
+By default, path colors are derived from the path names (Fig.~\ref{fig:odgi_viz}\textbf{b}).
 Path names are displayed on the left of the paths.
@@
 
+\section*{Conflict of Interest}
+
+The authors declare that they have no competing interests.
+
 \section*{Data availability}

### 2022-01-20 manuscript/main.tex
keywords

@@
 
-\usepackage[]{hyperref}
+\usepackage[draft]{hyperref}
 \hypersetup{colorlinks=true,urlcolor=black,linkcolor=black,citecolor=black}
@@
 %\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
+\\
+\REVIEWED{\textbf{Keywords:} Navigating Variation Graphs. Pangenomic Liftover. Gigabase-Scale Pangenome Visualization. T2T Pangenome Graph Exploration.}
 }
@@
 \section{Introduction}
-A pangenome models the full set of genomic elements in a given species or clade~\citep{cpang2018,Eizenga_2020}.
+A pangenome models the full set of genomic elements in a given species or clade~\REVIEWED{\citep{cpang2018,Eizenga_2020}}.
 In contrast to reference-based approaches which relate sequences to a single genome, these data structures encode the mutual relationships between all the genomes represented.
 %In pangenome graphs~\citep{Paten:2017}, homologous regions between genomes are compressed into a single representative of all alleles present in the pangenome.
-\REVIEWED{A class of methods to represent pangenomes involves the sequence graphs~\citep{Hein1989, Paten:2017}, where homologous regions between genomes are compressed into a single representative of all alleles present in the pangenome.
+\REVIEWED{[2,5] A class of methods to represent pangenomes involves the sequence graphs~\citep{Hein1989, Paten:2017}, where homologous regions between genomes are compressed into a single representative of all alleles present in the pangenome.
 In node-labeled sequence graphs, nodes indicate DNA sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph.
 A bidirected sequence graph represents both strands of DNA.
-On this model, variation graphs add the concept of path to embed linear sequences (e.g., genomes or haplotypes) into the graph~\citep{Garrison:2018}.
+On this model, variation graphs add the concept of paths representing linear sequences as traversals through the nodes of the graph~\citep{Garrison:2018}.
+A path can be a genome, haplotype, contig, or read.
+[2,5]
 }
@@
 
-\section*{Conflict of Interest}
+\REVIEWED{\section*{Conflict of Interest}}
 
-The authors declare that they have no competing interests.
+\REVIEWED{The authors declare that they have no competing interests.}
 

### 2022-01-20 manuscript/main.tex
for Andrea

@@
 \section{Implementation}
-The ODGI toolkit builds on existing approaches to efficiently store and manipulate \REVIEWED{pangenome graphs in the form of} variation graphs~\citep{Garrison:2018}.
+The ODGI toolkit builds on existing approaches to efficiently store and manipulate \REVIEWED{[2,5] pangenome graphs in the form of [2,5]} variation graphs~\citep{Garrison:2018}.
 Similar to other efficient libraries presenting the \textsc{HandleGraph} model~\citep{Eizenga_2020_BX}, the implementation of ODGI's tools rests on three key properties which hold for most pangenome graphs:

### 2022-01-20 manuscript/main.tex
heavily-sh reworked introduction

--- 2022-01-20 manuscript/main.tex
@@
 
-\copyrightyear{2021} \pubyear{XXXX}
+\copyrightyear{2022} \pubyear{XXXX}
 
@@
 In contrast to reference-based approaches which relate sequences to a single genome, these data structures encode the mutual relationships between all the genomes represented.
-%In pangenome graphs~\citep{Paten:2017}, homologous regions between genomes are compressed into a single representative of all alleles present in the pangenome.
 \REVIEWED{[2,5] A class of methods to represent pangenomes involves the sequence graphs~\citep{Hein1989, Paten:2017}, where homologous regions between genomes are compressed into a single representative of all alleles present in the pangenome.
@@
 A path can be a genome, haplotype, contig, or read.
-[2,5]
-}
-These flexible models let us encode any kind of variation, allowing the generation of comprehensive data systems that builds the basis for the analyses of genome evolution.
-Although these data structures are of utility to researchers~\citep{cpang2018,Garrison:2018,Baaijens_2019,Hickey:2020,Sibbesen_2021}, the scientific community still lacks a toolset specifically focused on graph manipulation and interrogation.
+[2,5]}
 
+\REVIEWED{Pangenome graphs can be constructed by multiple sequence alignment~\citep{Lee_2002,Grasso_2004} or by transitively reducing an alignment between sequences to an equivalent, labeled sequence graph~\citep{Kehr_2014,Garrison_2019_thesis}.
+Current methods to build these graphs are still under active development~\citep{Li:2020,Armstrong:2020,pggb}, but they have largely settled on a common data model, represented in the Graphical Fragment Assembly (GFA) format~\citep{GFA}.
+This standardization supports the development of a reference set of tools that operate on the pangenome graph model.}
+
+Pangenome graphs let us encode any kind of variation, allowing the generation of comprehensive data systems that builds the basis for the analyses of genome evolution.
 The Human Pangenome Reference Consortium (HPRC) and Telomere-to-Telomere (T2T) consortium~\citep{Miga:2020, Logsdon_2021, Nurk_2021} have recently demonstrated that high-quality \textit{de novo} assemblies can be routinely generated from third-generation long read sequencing data.
-We anticipate that \textit{de novo} assemblies of similar quality will become common, leading to demand for methods that allow us to create and explore pangenomes.
-%Pangenome graphs are a natural model to support these kinds of analyses, but studies based on pangenome graphs are difficult due to a lack of downstream tools and standard workflows.
+We anticipate that \textit{de novo} assemblies of similar quality will become common, leading to demand for methods to \REVIEWED{analyze} pangenomes.
 
-Here, we present the Optimized Dynamic Genome/Graph Implementation (ODGI) toolkit, a pangenome graph interrogation and transformation system specifically implemented to handle the data scales encountered when working with pangenomes built from hundreds of haplotype-resolved genomes.
+Although \REVIEWED{pangenome graphs} are data structures of utility to researchers~\citep{cpang2018,Garrison:2018,Baaijens_2019,Hickey:2020,Sibbesen_2021}, the scientific community still lacks a toolset \REVIEWED{capable of operating on gigabase-scale pangenomes built from whole-genome assemblies}.
+Such an effort began with the VG toolkit~\citep{Garrison:2018}\REVIEWED{, but its tools do not handle efficiently pangenome graphs presenting complex motifs that result from repetitive sequences like centromeres.}
+Here we refocus it with the Optimized Dynamic Genome/Graph Implementation (ODGI) toolkit, a compatible, but independent pangenome graph interrogation and transformation system specifically implemented to handle the data scales encountered when working with pangenomes built from hundreds of haplotype-resolved genomes.
 ODGI provides a set of standard operations on the variation graph data model, generalizing ``genome arithmetic'' concepts like those found in BEDTools~\citep{Quinlan_2010} to work on pangenome graphs,
@@
 By containing both the sequences and information about their relative variations, the variation graph provides a complete and powerful foundation for many bioinformatic applications.
-
-Pangenome graphs can be constructed by multiple sequence alignment~\citep{Lee_2002,Grasso_2004} or by transitively reducing an alignment between sequences to an equivalent, labeled sequence graph~\citep{Kehr_2014,Garrison_2019_thesis}.
-Current methods to build these graphs are still under active development~\citep{Li:2020,Armstrong:2020,pggb}, but they have largely settled on a common data model, represented in the Graphical Fragment Assembly (GFA) format~\citep{GFA}.
-%\FIXME{may want to move GFA standard+VG to intro section 1}. AG: it is borderline, but it seems fine to have it here too.
-This standardization supports the development of a reference set of tools that operate on the pangenome graph model.
-Such an effort began with the VG toolkit~\citep{Garrison:2018}.
-Here we refocus it with ODGI, a compatible, but independent set of algorithms focused on visualization, interrogation, and transformation of pangenome graphs.
 
@@
 This work was supported by the BMBF-funded de.NBI Cloud within the German Network for Bioinformatics Infrastructure (de.NBI) (031A537B, 031A533A, 031A538A, 031A533B, 031A535A, 031A537C, 031A534A, 031A532B).
-
-\REVIEWED{\section*{Conflict of Interest}}
-
-\REVIEWED{The authors declare that they have no competing interests.}
+\linebreak
+\linebreak
+\REVIEWED{\textit{Conflict of Interest}: the authors declare no competing interests.}
 

### 2022-02-04 manuscript/main.tex
ODGI's build process is lossless
genomic sequences of the DNA

@@
 \REVIEWED{[2,5] A class of methods to represent pangenomes involves the sequence graphs~\citep{Hein1989, Paten:2017}, where homologous regions between genomes are compressed into a single representative of all alleles present in the pangenome.
-In node-labeled sequence graphs, nodes indicate DNA sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph.
-A bidirected sequence graph represents both strands of DNA.
+In node-labeled sequence graphs, nodes indicate \REVIEWED{[1,2] genomic sequences [1,2]}, with edges connecting the nodes that are concatenated in the sequences represented in the graph.
+A bidirected sequence graph can \REVIEWED{[1,2] represent both strands of the genomic sequences of the DNA. [1,2]}
 On this model, variation graphs add the concept of paths representing linear sequences as traversals through the nodes of the graph~\citep{Garrison:2018}.
@@
 A pangenome graph is a sequence model that encodes the mutual alignment of many genomes~\citep{Garrison_2019_thesis,Eizenga_2020}.
-In the variation graph, $V = (N, E, P)$, nodes $N = n_1\ldots n_{|N|}$ contain sequences of DNA.
+In the variation graph, $V = (N, E, P)$, nodes $N = n_1\ldots n_{|N|}$ contain \REVIEWED{[1,2] genomic sequences. [1,2]}
 Each node $n_i$ has an identifier $i$ and an implicit reverse complement $\bar{n_i}$, and a node strand $s$ corresponds to one such orientations.
@@
     \textbf{lock} \tcp{atomic locking primitive}
-    \textbf{sequence} $= [$A$|$T$|$G$|$C$|$N$]*$ \tcp{DNA}
+    \textbf{sequence} $= [$A$|$T$|$G$|$C$|$N$]*$ \REVIEWED{[1,2] \tcp{genomic} [1,2]}
     \tcp{bit-packed vector of edges}
@@
 This construction step can be a significant bottleneck, in particular as the size of the path set of the graph increases.
+\REVIEWED{[1,5] The process itself is lossless.
+A graph in ODGI format represents everything that is in the input GFAv1 graph, without any loss of information.
+[1,5]}
 
@@
 This visualization is suitable for understanding graph node order, as smooth color gradients indicate that the graph order respects the linear paths' coordinate systems.
-Pangenome graphs represent both strands of DNA sequences.
+Pangenome graphs represent both strands \REVIEWED{[1,2] of the genomic sequences of the DNA. [1,2]}
 \textit{odgi viz} supports also coloring the paths by orientation, with paths colored where their sequence is reverse-complemented (red) or in direct orientation (black) with respect to the sequences of the graph nodes (Fig.~\ref{fig:odgi_viz}\textbf{d}).

### 2022-02-04 manuscript/main.tex
more genomic sequences

@@
 A pangenome models the full set of genomic elements in a given species or clade~\REVIEWED{\citep{cpang2018,Eizenga_2020}}.
-In contrast to reference-based approaches which relate sequences to a single genome, these data structures encode the mutual relationships between all the genomes represented.
+In contrast to reference-based approaches which relate sequences to a single genome, these data structures encode the mutual relationships between all the genomes \REVIEWED{[1,2]and genomic sequences[1,2]} represented.
 \REVIEWED{[2,5] A class of methods to represent pangenomes involves the sequence graphs~\citep{Hein1989, Paten:2017}, where homologous regions between genomes are compressed into a single representative of all alleles present in the pangenome.
@@
 A bidirected sequence graph can \REVIEWED{[1,2] represent both strands of the genomic sequences of the DNA. [1,2]}
-On this model, variation graphs add the concept of paths representing linear sequences as traversals through the nodes of the graph~\citep{Garrison:2018}.
-A path can be a genome, haplotype, contig, or read.
+On this model, variation graphs add the concept of paths representing linear \REVIEWED{[1,2]genomic[1,2]} sequences as traversals through the nodes of the graph~\citep{Garrison:2018}.
+\REVIEWED{[1,2]For example, a[1,2]} path can be a genome, haplotype, contig, or read.
 [2,5]}
@@
 If we cannot process such regions, there are only two options: 1) remove such regions, or 2) leave them unaligned.
-However, neither of these solutions allows us to investigate their biological features.
+However, neither of these solutions allows us  their biological features.
 To seamlessly represent such difficult regions, we followed an approach implemented in the dynamic version of the Graph BWT (GBWT)~\citep{Siren:2020} and built a node-centric, dynamic, compressed model of the paths.
@@
     \textbf{lock} \tcp{atomic locking primitive}
-    \textbf{sequence} $= [$A$|$T$|$G$|$C$|$N$]*$ \REVIEWED{[1,2] \tcp{genomic} [1,2]}
+    \textbf{sequence} $= [$A$|$T$|$G$|$C$|$N$]*$ \REVIEWED{[1,2] \tcp{genomic}}
     \tcp{bit-packed vector of edges}
@@
 Eukaryotic genomes experience gains and losses of genetic material, resulting in copy number variation (CNV) across the population.
-With \textit{odgi viz}, we can use multiple color palettes to color the paths by path depth, highlighting the different copy number statuses in the genomes represented in the pangenome graph (Fig.~\ref{fig:odgi_viz}\textbf{e}).
+With \textit{odgi viz}, we can use multiple color palettes to color the paths by path depth, highlighting the different copy number statuses in the \REVIEWED{[1,2]genomic sequences[1,2]} represented in the pangenome graph (Fig.~\ref{fig:odgi_viz}\textbf{e}).
 
@@
 The key data in a pangenome graph is a representation of the alignment (or homology) relationships between the sequences represented.
-Navigating and understanding the graph requires coordinate systems that we can use to link other data into the graph model, and thus to all genomes in the pangenome.
+Navigating and understanding the graph requires coordinate systems that we can use to link other data into the graph model, and thus to all genomes \REVIEWED{[1,2]and genomic sequences[1,2]} in the pangenome.
 ODGI's tools use the embedded sequences to provide a universal coordinate space that is graph-independent, thereby remaining stable across different graphs built with the same genomes.
@@
 
-Pangenome graphs model alignments of many genomes.
+Pangenome graphs \REVIEWED{can} model alignments of many genomes.
 With \textit{odgi untangle}, users can extract pairwise alignment information between a given set of ``query'' sequences and a given set of ``target'' sequences (used as references).
@@
 Pangenome graph topology can be derived by applying \textit{odgi matrix}, obtaining information on graph nodes connections in textual sparse matrix format.
-To investigate on the genomes encoded in the graph, \textit{odgi paths} allows users to calculate pairwise overlap statistics of groupings of paths and emit all path sequences in FASTA format, and it also allows the generation of a ``pangenome matrix'' that reports the copy number (presence/absence) of each path over each node.
+To investigate on the \REVIEWED{[1,2] genomic sequences [1,2]} encoded in the graph, \textit{odgi paths} allows users to calculate pairwise overlap statistics of groupings of paths and emit all path sequences in FASTA format, and it also allows the generation of a ``pangenome matrix'' that reports the copy number (presence/absence) of each path over each node.
 \textit{odgi flatten} generates a linearization of the graph by emitting the pangenome sequence (the concatenation of all node sequences) in FASTA format, and the projection of all paths on the linearized sequence in BED format.
@@
 This has been seen as essential to making these new bioinformatic models interpretable.
-While our presentation is primarily qualitative, our work demonstrates that we can mitigate this issue by exploiting the pangenome graph not as a static reference, but as a dynamic model of the mutual alignment of many genomes.
+While our presentation is primarily qualitative, our work demonstrates that we can mitigate this issue by exploiting the pangenome graph not as a static reference, but as a dynamic model of the mutual alignment of many \REVIEWED{[1.2] genomic sequences [1.2]}.
 Because pangenome graphs can contain complete genomes, we are able to query them to polarize the information they contain in easily-interpretable and reusable pairwise formats that are widely supported in bioinformatics.

### 2022-02-08 manuscript/main.tex
GFAv2 and rGFA are conditionally acceptable

@@
 
-Although \REVIEWED{pangenome graphs} are data structures of utility to researchers~\citep{cpang2018,Garrison:2018,Baaijens_2019,Hickey:2020,Sibbesen_2021}, the scientific community still lacks a toolset \REVIEWED{capable of operating on gigabase-scale pangenomes built from whole-genome assemblies}.
+Although \REVIEWED{pangenome graphs} are data structures of utility to researchers~\citep{cpang2018,Garrison:2018,Baaijens_2019,Hickey:2020,Sibbesen_2021}, the scientific community still lacks a toolset \REVIEWED{capable of operating on gigabase-scale pangenome graphs constructed from whole-genome assemblies}.
 Such an effort began with the VG toolkit~\citep{Garrison:2018}\REVIEWED{, but its tools do not handle efficiently pangenome graphs presenting complex motifs that result from repetitive sequences like centromeres.}
-Here we refocus it with the Optimized Dynamic Genome/Graph Implementation (ODGI) toolkit, a compatible, but independent pangenome graph interrogation and transformation system specifically implemented to handle the data scales encountered when working with pangenomes built from hundreds of haplotype-resolved genomes.
-ODGI provides a set of standard operations on the variation graph data model, generalizing ``genome arithmetic'' concepts like those found in BEDTools~\citep{Quinlan_2010} to work on pangenome graphs,
+Here we refocus it with the Optimized Dynamic Genome/Graph Implementation (ODGI) toolkit, a compatible, but independent pangenome graph interrogation and transformation system specifically implemented to handle the data scales encountered when working \REVIEWED{[1,10]with previously constructed pangenomes comprising hundreds of haplotype-resolved genomes.[1,10]}
+ODGI offers a set of standard operations on the variation graph data model, generalizing ``genome arithmetic'' concepts like those found in BEDTools~\citep{Quinlan_2010} to work on pangenome graphs,
 and providing a variety of operations, such as visualization, sorting, and liftover projections, all critical to understand and exploit pangenome graphs.
@@
 This construction step can be a significant bottleneck, in particular as the size of the path set of the graph increases.
-\REVIEWED{[1,5] The process itself is lossless.
-A graph in ODGI format represents everything that is in the input GFAv1 graph, without any loss of information.
-[1,5]}
+\REVIEWED{[1,5]The process itself is lossless.
+A graph in ODGI format represents everything that is in the input GFAv1 graph, without any loss of information.[1,5]} \REVIEWED{[2,1]ODGI does not support GFAv2 nor rGFA.
+The less popular GFAv2 format was specifically designed for assembly graphs.
+rGFA requires a genomic sequence to be the reference sequence, but in GFAv1 we don't have that limitation which is fundamental to implement reference-free approaches.[2,1]}
 
@@
 
-ODGI is already the backbone of the Pangenome Graph Builder pipeline~\citep{pggb}.
+\REVIEWED{[1,10]While ODGI does not construct graphs from scratch nor is is capable of extending them, ODGI[1,10]} is already the backbone of the Pangenome Graph Builder pipeline~\citep{pggb}.
 Its static, large-scale 1D and 2D visualizations of the pangenome graphs allow an unprecedented high-level perspective on variation in pangenomes, and have also been critical in the development of pangenome graph building methods.

### 2022-02-09 scripts/build_time.sh
update performance evaluation scripts

@@ -1,89 +1,0 @@
-wget https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/scratch/2021_11_16_pggb_wgg.88/chroms/chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
-gunzip chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
-
-G=chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa
-
-for T in 1 2 4 8 16
-do
-    for i in {1..10}
-    do
-        echo "T: ""$T"" i: ""$i"
-        /usr/bin/time --verbose odgi build -g "$G" -o "$G".og -t "$T" 2> chr6_odgi_build_time_"$T"_"$i"
-        /usr/bin/time --verbose vg convert -g -x -t "$T" "$G" 2> chr6_vg_convert_time_"$T"_"$i" 1> "$G".xg
-    done    
-done
-
-if [ -f odgi_build_threads_time.csv ] ; then
-    rm odgi_build_threads_time.csv
-fi
-if [ -f vg_convert_threads_time.csv ] ; then
-    rm vg_convert_threads_time.csv
-fi
-
-for T in 1 2 4 8 16
-do
-    for i in {1..10}
-    do
-      echo "$T","$i",$(grep Elapsed chr6_odgi_build_time_"$T"_"$i" | cut -f 8 -d ' ' | awk -F: '{ print ($1 * 60) + ($2) + $3 }'),$(grep "Maximum" chr6_odgi_build_time_"$T"_"$i" | cut -f 6 -d ' ') >> odgi_build_threads_time.csv
-      echo "$T","$i",$(grep Elapsed chr6_vg_convert_time_"$T"_"$i" | cut -f 8 -d ' ' | awk -F: '{ print ($1 * 60) + ($2) + $3 }'),$(grep "Maximum" chr6_vg_convert_time_"$T"_"$i" | cut -f 6 -d ' ') >> vg_convert_threads_time.csv
-    done
-done
-
-# TODO
-#Rscript time.R mean_build_time.csv mean_build_time.pdf
-
-odgi build -g "$G" -o "$G".og -t 16 -P
-odgi paths -i "$G".og -L -l -t 16 > "$G".paths
-grep -E 'grch38|chm13' "$G".paths -v > "$G".non_ref.paths
[…]

### 2022-02-09 scripts/build_time_haps.sh

@@ -1,3 +1,5 @@
+#!/bin/bash
+
 wget https://s3-us-west-2.amazonaws.com/human-pangenomics/pangenomes/scratch/2021_11_16_pggb_wgg.88/chroms/chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
 gunzip chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.gfa.gz
 
@@ -7,6 +9,13 @@ odgi build -g "$G" -o "$G".og -t 16 -P
 odgi paths -i "$G".og -L -l -t 16 > "$G".paths
 grep -E 'grch38|chm13' "$G".paths -v > "$G".non_ref.paths
 cut -f 1,2 "$G".non_ref.paths -d '#' | sort| uniq > "$G".non_ref.haps
+
+if [ -f odgi_build_threads_haps_time.csv ] ; then
+    rm odgi_build_threads_haps_time.csv
+fi
+if [ -f vg_convert_threads_haps_time.csv ] ; then
+    rm vg_convert_threads_haps_time.csv
+fi
 
 for H in 1 2 4 8 16 32 64
 do  
@@ -25,35 +34,26 @@ do
     fi
     odgi extract -i "$G".og -o - -b "$G"."$H"haps.bed -p "$G"."$H"haps -t 16 -P | odgi sort -i - -O -o "$G"."$H"haps.og
     odgi view -i "$G"."$H"haps.og -g > "$G"."$H"haps.og.gfa
-
-    odgi build -g "$G"."$H"haps.og.gfa -o "$G"."$H"haps.og.gfa.og -t 16 -P
-    vg convert -g -x -t 16 "$G"."$H"haps.og.gfa > "$G"."$H"haps.og.gfa.xg
-done
-
-for H in 1 2 4 8 16 32 64
-do
-    for i in 1 2 3 4 5 6 7 8 9 10
+    for T in 1 2 4 8 16
     do
-        /usr/bin/time --verbose odgi viz -i "$G"."$H"haps.og.gfa.og -o "$G"."$H"haps.og.gfa.og.png 2> chr6_odgi_viz_time_"$H"_"$i"
-        #TODO if we can actually produce something...
-        /usr/bin/time --verbose vg viz -D -x "$G"."$H"haps.og.gfa.xg -o "$G"."$H"haps.og.gfa.xg.svg 2> chr6_vg_viz_time_"$H"_"$i"
[…]

### 2022-02-09 scripts/calc_max_edges.R

@@ -1,0 +1,16 @@
+calc_max_edges <- function(num_nodes) {
+  if (num_nodes == 0) {
+    return(0)
+  }
+  value <- 0
+  for (i in seq(1, num_nodes)) {
+    value <- value + i - 1  
+  }
+  
+  return(2*value + num_nodes)
+}
+options(digits = 22)
+# print(calc_max_edges(3))
+print(calc_max_edges(35)) #1225
+print(calc_max_edges(377123)) #142221757129
+print(calc_max_edges(2000000)) #4e+10

### 2022-02-09 scripts/extract_time.sh

@@ -49,7 +49,7 @@ for H in 1 2 4 8 16 32 64
 do
     for T in 1 2 4 8 16
     do
-        for i in {1..2}
+        for i in 1 2 3 4 5 6 7 8 9 10
         do
             echo "H: ""$H"" T: ""$T"" i: ""$i"
             /usr/bin/time --verbose odgi extract -i "$G"."$H"haps.og.gfa.og -t "$T" -c 10 -r grch38#chr6:58553888-59829934 -o "$G"."$H"haps.og.gfa.og.centromeres.og 2> chr6_odgi_extract_time_"$T"_"$H"_"$i"
@@ -69,7 +69,7 @@ for H in 1 2 4 8 16 32 64
 do
     for T in 1 2 4 8 16
     do
-        for i in {1..2}
+        for i in 1 2 3 4 5 6 7 8 9 10
         do
             echo "$T","$H","$i",$(grep Elapsed chr6_odgi_extract_time_"$T"_"$H"_"$i" | cut -f 8 -d ' ' | awk -F: '{ print ($1 * 60) + ($2) + $3 }'),$(grep "Maximum" chr6_odgi_extract_time_"$T"_"$H"_"$i" | cut -f 6 -d ' ') >> odgi_extract_time.csv
             echo "$T","$H","$i",$(grep Elapsed chr6_vg_chunk_time_"$T"_"$H"_"$i" | cut -f 8 -d ' ' | awk -F: '{ print ($1 * 60) + ($2) + $3 }'),$(grep "Maximum" chr6_vg_chunk_time_"$T"_"$H"_"$i" | cut -f 6 -d ' ') >> vg_chunk_time.csv 

### 2022-02-09 scripts/pos_time.sh

@@ -8,6 +8,27 @@ odgi paths -i "$G".og -L -l -t 16 > "$G".paths
 grep -E 'grch38|chm13' "$G".paths -v > "$G".non_ref.paths
 cut -f 1,2 "$G".non_ref.paths -d '#' | sort| uniq > "$G".non_ref.haps
 
+vg_find_helper="#!/bin/bash
+
+if [ -f \"\$G\".\"\$H\"haps.og.gfa.xg.find ] ; then
+    rm \"\$G\".\"\$H\"haps.og.gfa.xg.find
+fi
+
+G=\"\$1\"
+H=\"\$2\"
+T=\"\$3\"
+n=\"\$4\"
+
+while IFS= read -r pos
+    do
+    if [[ \"\$pos\" != *\"strand.vs.ref\" ]]; then
+        vg find -x \"\$G\".\"\$H\"haps.og.gfa.xg -P \$(echo \"\$pos\" | cut -f 2 | grep \"target.path.pos\" -v | cut -f 1 -d ',') -n \"\$n\" >> \"\$G\".\"\$H\"haps.og.gfa.xg.find
+    fi
+done < \"\$G\".\"\$H\"haps.og.gfa.og.pos"
+
+echo "$vg_find_helper" > vg_find_helper.sh
+chmod +x vg_find_helper.sh
+
 for H in 1 2 4 8 16 32 64
 do  
     echo "hap: ""$H" 
@@ -47,21 +68,21 @@ done
 
 for H in 1 2 4 8 16 32 64
 do
+    # we take a position relative to the middle of graphs with more than 1 haplotype
+    odgi position -i "$G"."$H"haps.og.gfa.og -p grch38#chr6,147658481,+ -t 1 -v > "$G"."$H"haps.og.gfa.og.graphpos
     for T in 1 2 4 8 16
     do
-        for i in {1..2}
+        for i in 1 2 3 4 5 6 7 8 9 10
         do
-            # 4701798 nodes, so we take the one in the middle
             echo "H: ""$H"" T: ""$T"" i: ""$i"
-            /usr/bin/time --verbose odgi position -i "$G"."$H"haps.og.gfa.og -g 2350000 -I -t "$T" 2> chr6_odgi_position_time_"$T"_"$H"_"$i" 1> "$G"."$H"haps.og.gfa.og.pos
-            # TODO this won't work with one haplotype
[…]

### 2022-02-09 manuscript/main.tex
genome vs genomic sequence

@@
 A pangenome models the full set of genomic elements in a given species or clade~\REVIEWED{\citep{cpang2018,Eizenga_2020}}.
-In contrast to reference-based approaches which relate sequences to a single genome, these data structures encode the mutual relationships between all the genomes \REVIEWED{[1,2]and genomic sequences[1,2]} represented.
+In contrast to reference-based approaches which relate \REVIEWED{samples} to a single genome, these data structures encode the mutual relationships between all the genomes represented.
 \REVIEWED{[2,5] A class of methods to represent pangenomes involves the sequence graphs~\citep{Hein1989, Paten:2017}, where homologous regions between genomes are compressed into a single representative of all alleles present in the pangenome.
 In node-labeled sequence graphs, nodes indicate \REVIEWED{[1,2] genomic sequences [1,2]}, with edges connecting the nodes that are concatenated in the sequences represented in the graph.
-A bidirected sequence graph can \REVIEWED{[1,2] represent both strands of the genomic sequences of the DNA. [1,2]}
-On this model, variation graphs add the concept of paths representing linear \REVIEWED{[1,2]genomic[1,2]} sequences as traversals through the nodes of the graph~\citep{Garrison:2018}.
-\REVIEWED{[1,2]For example, a[1,2]} path can be a genome, haplotype, contig, or read.
+A bidirected sequence graph can \REVIEWED{[1,2] represent both strands of the genomic sequences [1,2]}.
+On this model, variation graphs add the concept of paths representing linear \REVIEWED{[1,2] genomic [1,2]} sequences as traversals through the nodes of the graph~\citep{Garrison:2018}.
+\REVIEWED{[1,2] For example, a [1,2]} path can be a genome, haplotype, contig, or read.
 [2,5]}
@@
     \textbf{lock} \tcp{atomic locking primitive}
-    \textbf{sequence} $= [$A$|$T$|$G$|$C$|$N$]*$ \REVIEWED{[1,2] \tcp{genomic}}
+    \textbf{sequence} $= [$A$|$T$|$G$|$C$|$N$]*$ \\
     \tcp{bit-packed vector of edges}
@@
 This visualization is suitable for understanding graph node order, as smooth color gradients indicate that the graph order respects the linear paths' coordinate systems.
-Pangenome graphs represent both strands \REVIEWED{[1,2] of the genomic sequences of the DNA. [1,2]}
+Pangenome graphs represent both strands \REVIEWED{[1,2] of the genomic sequences [1,2]}.
 \textit{odgi viz} supports also coloring the paths by orientation, with paths colored where their sequence is reverse-complemented (red) or in direct orientation (black) with respect to the sequences of the graph nodes (Fig.~\ref{fig:odgi_viz}\textbf{d}).
 Eukaryotic genomes experience gains and losses of genetic material, resulting in copy number variation (CNV) across the population.
-With \textit{odgi viz}, we can use multiple color palettes to color the paths by path depth, highlighting the different copy number statuses in the \REVIEWED{[1,2]genomic sequences[1,2]} represented in the pangenome graph (Fig.~\ref{fig:odgi_viz}\textbf{e}).
+With \textit{odgi viz}, we can use multiple color palettes to color the paths by path depth, highlighting the different copy number statuses in the \REVIEWED{[1,2] genomic sequences [1,2]} represented in the pangenome graph (Fig.~\ref{fig:odgi_viz}\textbf{e}).
 
@@
 The key data in a pangenome graph is a representation of the alignment (or homology) relationships between the sequences represented.
-Navigating and understanding the graph requires coordinate systems that we can use to link other data into the graph model, and thus to all genomes \REVIEWED{[1,2]and genomic sequences[1,2]} in the pangenome.
+Navigating and understanding the graph requires coordinate systems that we can use to link other data into the graph model, and thus to the \REVIEWED{[1,2] genomic sequences [1,2]} in the pangenome.
 ODGI's tools use the embedded sequences to provide a universal coordinate space that is graph-independent, thereby remaining stable across different graphs built with the same genomes.

### 2022-02-09 manuscript/main.tex
broad relevance of ODGI plus more citations

@@
 In particular, lifting over annotations and linearizing nested graph structures place the suite as the bridge between traditional linear reference genome analysis and pangenome graphs.
-With the increased adoption of long read sequencing we expect pangenomic tools to become increasingly common in biomedical research.
-Particularly for targets that involve complex variation, such as cancer, plant genomics and metagenomics, ODGI will facilitate disentangling,
+With the increased adoption of long read sequencing we expect pangenomic tools to become increasingly common in \REVIEWED{the genomic studies at different taxonomic levels and} in biomedical research.
+Particularly for targets that involve complex variation, such as cancer\REVIEWED{~\citep{CompPan2016}}, plant genomics\REVIEWED{~\citep{Bayer2020}} and metagenomics\REVIEWED{~\citep{Zhong2021}}, ODGI will facilitate disentangling,
 describing and analyzing a much larger set of variation than previously was possible with tools that depend on short reads and reference genomes.

### 2022-02-09 manuscript/main.tex
explicit in the abstract that ODGI takes pre-built graphs

@@
 We wrote ODGI, a novel suite of tools that implements scalable algorithms and has an efficient in-memory representation of DNA \REVIEWED{pangenome graphs in the form of variation graphs}.
+\REVIEWED{ODGI supports pre-built graphs in the Graphical Fragment Assembly format version 1.}
 ODGI includes tools for detecting complex regions, extracting \textit{loci}, removing artifacts, exploratory analysis, manipulation, validation, and visualization.

### 2022-02-10 scripts/build.R
construction timings incoming

@@ -1,0 +1,100 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(ggbeeswarm)
+library(knitr)
+library(gridExtra)
+library(R.devices)
+
+args <- commandArgs(trailingOnly = T)
+
+odgi_build_csv <- args[1]
+vg_convert_csv <- args[2]
+build_pdf <- args[3]
+build_csv <- args[4]
+
+#odgi_build_csv <- "/home/heumos/git/odgi-paper/data/odgi_build_threads_haps_time.csv"
+odgi_build <- read.delim(odgi_build_csv, sep = ",", as.is = T, header = F, stringsAsFactors = F, strip.white = T)
+colnames(odgi_build) <- c("threads", "haps", "run", "time", "memory")
+# time is in seconds
+# memory is in kilobytes
+odgi_build$tool <- "odgi build"
+summary(odgi_build)
+
+#vg_convert_csv <- "/home/heumos/git/odgi-paper/data/vg_convert_threads_haps_time.csv"
+vg_convert <- read.delim(vg_convert_csv, sep = ",", as.is = T, header = F, stringsAsFactors = F, strip.white = T)
+colnames(vg_convert) <- c("threads", "haps", "run", "time", "memory")
+vg_convert$tool <- "vg convert"
+summary(vg_convert)
+
+construct <- rbind(odgi_build, vg_convert)
+construct$threads <- factor(construct$threads, levels=unique(construct$threads))
+construct$haps <- factor(construct$haps, levels=unique(construct$haps))
+# construct$memory <- factor(construct$memory, levels=unique(construct$memory))
+construct$memory <- as.numeric(construct$memory)/1000000
+
+construct_64haps <- construct[construct$haps == 64,]
+construct_16threads <- construct[construct$threads == 16,]
+
+dodge <- position_dodge(width = 0.0)
+
+ct_p_t <- ggplot(construct_64haps, aes(x=threads, y=time, fill=tool, color = tool)) +
+  geom_violin(position = dodge) +
+  #geom_beeswarm(size=0.5) +
+  labs(x = "number of threads", y = "time in seconds") +
+  expand_limits(x = 0, y = 0) + 
+  theme(legend.position = "none")
+#ct_p_t
+
+ch_p_t <- ggplot(construct_16threads, aes(x=haps, y=time, fill=tool, color = tool)) +
[…]

### 2022-02-10 scripts/time.R

@@ -1,28 +1,0 @@
-#!/usr/bin/env Rscript
-
-library(tidyverse)
-library(ggbeeswarm)
-library(knitr)
-
-args <- commandArgs(trailingOnly = T)
-
-times_csv <- args[1]
-times_pdf <- args[2]
-
-# times_csv <- "/home/heumos/git/odgi-paper/data/mean_build_time.csv"
-
-times <- read.delim(times_csv, sep = ",", as.is = T, header = T)
-
-summary(times)
-
-times$threads <- factor(times$threads, levels=unique(times$threads))
-
-p <- ggplot(times, aes(x=threads, y=time, group=threads)) +
-    geom_violin() +
-    #geom_beeswarm(size=0.5) +
-    labs(x = "number of threads", y = "time in seconds") +
-    expand_limits(x = 0, y = 0)
-#pdf(times_pdf, height = 2.5, width = 3)
-ggsave(times_pdf, device = cairo_pdf, height = 2.5, width = 3)
-knitr::plot_crop(times_pdf)
-

### 2022-02-10 scripts/viz.R
viz evaluation files ready

@@ -1,0 +1,97 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(ggbeeswarm)
+library(knitr)
+library(gridExtra)
+
+args <- commandArgs(trailingOnly = T)
+
+odgi_viz_csv <- args[1]
+vg_viz_csv <- args[2]
+viz_pdf <- args[3]
+viz_csv <- args[4]
+
+#odgi_viz_csv <- "/home/heumos/git/odgi-paper/data/odgi_viz_time.csv"
+odgi_viz <- read.delim(odgi_viz_csv, sep = ",", as.is = T, header = F, stringsAsFactors = F, strip.white = T)
+colnames(odgi_viz) <- c("haps", "run", "time", "memory")
+# time is in seconds
+# memory is in kilobytes
+odgi_viz$tool <- "odgi viz"
+summary(odgi_viz)
+
+#vg_viz_csv <- "/home/heumos/git/odgi-paper/data/vg_viz_time.csv"
+vg_viz <- read.delim(vg_viz_csv, sep = ",", as.is = T, header = F, stringsAsFactors = F, strip.white = T)
+colnames(vg_viz) <- c("haps", "run", "time", "memory")
+vg_viz$tool <- "vg viz"
+summary(vg_viz)
+
+viz <- rbind(odgi_viz, vg_viz)
+viz$haps <- factor(viz$haps, levels=unique(viz$haps))
+viz$memory <- as.numeric(viz$memory)/1000000
+
+dodge <- position_dodge(width = 0.0)
+
+first_column <- c(1,2,4,8,16,32,64)
+second_column <- c("*", "**", "**", "**", "**", "**", "**")
+df <- data.frame(first_column, second_column)
+colnames(df) <- c("haps", "label")
+
+
+vh_p_t <- ggplot(viz, aes(x=haps, y=time, fill=tool, color = tool)) +
+  geom_violin(position = dodge) +
+  #geom_beeswarm(size=0.5) +
+  labs(x = "number of haplotypes", y = "time in seconds") +
+  expand_limits(x = 0, y = 0) + theme(legend.direction = "horizontal") + 
+  labs(color = "tool   ") + 
+  labs(fill = "tool   ") + 
+  theme(legend.title = element_text(size=15)) +
+  theme(legend.key.size = unit(0.4, 'cm')) +
[…]

### 2022-02-10 scripts/viz_time.sh

@@ -35,7 +35,7 @@ do
     for i in 1 2 3 4 5 6 7 8 9 10
     do
         /usr/bin/time --verbose odgi viz -i "$G"."$H"haps.og.gfa.og -o "$G"."$H"haps.og.gfa.og.png 2> chr6_odgi_viz_time_"$H"_"$i"
-        #TODO if we can actually produce something...
+        # if we can actually produce something...
         /usr/bin/time --verbose vg viz -D -x "$G"."$H"haps.og.gfa.xg -o "$G"."$H"haps.og.gfa.xg.svg 2> chr6_vg_viz_time_"$H"_"$i"
     done
 done
@@ -52,10 +52,10 @@ do
     for i in 1 2 3 4 5 6 7 8 9 10
     do
         echo "$H","$i",$(grep Elapsed chr6_odgi_viz_time_"$H"_"$i" | cut -f 8 -d ' ' | awk -F: '{ print ($1 * 60) + ($2) + $3 }'),$(grep "Maximum" chr6_odgi_viz_time_"$H"_"$i" | cut -f 6 -d ' ') >> odgi_viz_time.csv
-        #TODO if we can actually produce something...
+        # if we can actually produce something...
         echo "$H","$i",$(grep Elapsed chr6_vg_viz_time_"$H"_"$i" | cut -f 8 -d ' ' | awk -F: '{ print ($1 * 60) + ($2) + $3 }'),$(grep "Maximum" chr6_vg_viz_time_"$H"_"$i" | cut -f 6 -d ' ') >> vg_viz_time.csv 
     done
 done
 
-#TODO
 #Rscript ...
+Rscript viz.R odgi_viz_time.csv vg_viz_time.csv viz_eval.pdf viz_supp.csv

### 2022-02-10 manuscript/main.tex
explain better odgi viz output

@@
 White spaces indicate where paths do not traverse the nodes.
+\REVIEWED{Directly consecutive nodes are displayed with no white space between the two.}
 The meaning of the colors depends on how \textit{odgi viz} is executed.
@@
 The black lines on the bottom indicate the edges connecting the nodes and, therefore, represent the graph topology.
+\REVIEWED{Given a path X traversing two nodes A and B, the corresponding edge is represented by a black line starting from the left or right of node A if this node is traversed in reverse or forward, respectively, by path X;
+the black line ends to the left or right of node B if this node is traversed in forward or reverse, respectively, by path X.
+Consequently, if two consecutive nodes are linked both in forward, no edge is shown (it would be 0 pixels long, as it would start from the right of the first node and ends to the left of the second one).}
 

### 2022-02-10 manuscript/fig_extract_viz_draw_position_untangle.tex
explain what are MHC and C4, specifying the acronyms

@@
     \caption{
-        Visualizing the MHC and C4 pangenome graphs.
+        Visualizing the \REVIEWED{major histocompatibility complex (MHC) and complement component 4 (C4)} pangenome graphs.
 \textbf{(a)} \textit{odgi draw} layout of the MHC pangenome graph extracted from a whole human pangenome graph of 90 haplotypes.
-The red rectangle highlights the C4 region.
+        \REVIEWED{The MHC locus is located on the human chromosome 6 and it encodes molecules involved in antigen presentation, inflammation regulation, the complement system, and the innate and adaptive immune responses~\citep{Shiina2009}.}
+        The red rectangle highlights the C4 region\REVIEWED{, which encodes proteins involved in the complement system.}
         \textbf{(b-e)} \textit{odgi viz} visualizations of the C4 pangenome graph, where 8 paths are displayed: 2 reference genomes (chm13 and grch38 on the top) and 6 haplotypes of 3 individuals.

### 2022-02-10 manuscript/fig_extract_viz_draw_position_untangle.tex
Bandage is not in ODGI

@@
         In HG01952\#1, the HERV sequence is absent.
-        \textbf{(f)} \textit{Bandage}~\citep{Wick_2015} layout of the C4 pangenome graph, annotated by using \textit{odgi position}.
+        \textbf{(f)} \REVIEWED{Layout of the C4 pangenome graph made with the \textit{Bandage} tool~\citep{Wick_2015} and} annotated by using \textit{odgi position}.
 Green nodes indicate the C4 genes.
 The red rectangle highlights the regions where \textit{C4A} and \textit{C4B} genes differ.
-        \textbf{(g)} Annotated \textit{bandage} layout of the C4 region where \textit{C4A} and \textit{C4B} genes differ due to single nucleotide variants leading to changes in the encoded protein sequences.
+        \textbf{(g)} Annotated \textit{Bandage} layout of the C4 region where \textit{C4A} and \textit{C4B} genes differ due to single nucleotide variants leading to changes in the encoded protein sequences.
 Node labels were annoted by using \textit{odgi position}.

### 2022-02-11 manuscript/fig_evaluation.tex
row row row your boat, row your boat

\begin{figure}[]
	\begin{subfigure}{\linewidth}
		\caption{}
		\centering
		% include first image
		\includegraphics[width=1.0\linewidth, trim=0cm 1.25cm 0cm 0cm]{fig/performance/build_eval.pdf}
		\label{fig:eval-build}
	\end{subfigure}
	\begin{subfigure}{1\linewidth}
		\caption{}
		\centering
		% include fourth image
		\includegraphics[width=\linewidth, trim=0cm 1.25cm 0cm 0cm]{fig/performance/viz_eval.pdf}
		\label{fig:eval-viz}
	\end{subfigure}
	\begin{subfigure}{\linewidth}
		\caption{}
		\centering
		% include second image
		\includegraphics[width=\linewidth, trim=0cm 1.25cm 0cm 0cm]{fig/performance/extract_eval.pdf}
		\label{fig:eval-extract}
	\end{subfigure}
	\caption{TODO}
	\label{fig:evaluation}
\end{figure}

### 2022-02-11 manuscript/main.tex
row row row your boat, row your boat

--- 2022-02-09 manuscript/main.tex
@@
 \usepackage{orcidlink}
+\usepackage{paralist}
 
@@
 Finally, we create paths in serial, and extend them in parallel by obtaining the mutex $Node.lock$ for pairs of nodes and by adding the path step in their $Node.path\_steps$ vectors.
-This parallelism speeds ODGI model construction by many-fold when testing against graphs made from assemblies produced by the HPRC (Figure~\ref{fig:build}).
+This parallelism speeds ODGI model construction by many-fold when testing against graphs made from assemblies produced by the HPRC (\autoref{Evaluation}).
 
@@
 %ODGI uses the ``S'' and ``L'' records to describe the graph topology, and ``P'' records to describe paths.
-
-\input{fig_build_time}
 
@@
 
+\subsection{Evaluation}
+\label{Evaluation}
+
+We measured the efficiency of ODGI (VERSION!) and VG (VERSION) for routine pangenome tasks. In particular, we measured the execution time and memory usage 
+\begin{inparaenum}[(i)]
+	\item of transforming a GFAv1 into the tool's native format,
+	\item the visualization of a pangenome graph, 
+	\item the extraction of a subgraph, and
+	\item the finding of path positions in a pangenome graph.
+\end{inparaenum}
+key graph operations, why
+deNBI
+data
+threads (64haps), haps (16 threads)
+general findings: VG sucks, hehehe
+
 \input{fig_extract_viz_draw_position_untangle}
+
+\input{fig_evaluation}
 

### 2022-02-11 scripts/build.R

@@ -97,4 +97,17 @@ construct_supp$memory <- construct_supp$memory/1000000
 construct_supp <- aggregate(cbind(memory,time) ~threads+haps+tool, data=construct_supp, FUN=mean)
 construct_supp <- construct_supp[order(construct_supp$threads, construct_supp$haps),]
 
-write.table(construct_supp, file = build_csv, sep = ",", row.names = F, quote = F)
+construct_latex <- setNames(data.frame(matrix(ncol = 6, nrow = 0)), 
+                            c("threads", "haps", "time_odgi_build", "time_vg_convert", "memory_odgi_build", "memory_vg_convert"))
+i <- 1
+j <- i + 1
+
+while(i < dim(construct_supp)[1]) {
+  row_row_row_your_boat <- c(construct_supp[i,1], construct_supp[i,2], construct_supp[i,5], construct_supp[j,5], construct_supp[i,4], construct_supp[j,4])
+  construct_latex <- rbind(construct_latex, row_row_row_your_boat)
+  i <- i + 2
+  j <- i + 1
+}
+colnames(construct_latex) <-  c("threads", "haps", "time_odgi_build", "time_vg_convert", "memory_odgi_build", "memory_vg_convert")
+
+write.table(construct_latex, file = build_csv, sep = ",", row.names = F, quote = F)

### 2022-02-11 scripts/extract.R

@@ -1,0 +1,112 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(ggbeeswarm)
+library(knitr)
+library(gridExtra)
+
+args <- commandArgs(trailingOnly = T)
+
+odgi_extract_csv <- args[1]
+vg_chunk_csv <- args[2]
+extract_pdf <- args[3]
+extract_csv <- args[4]
+
+#odgi_extract_csv <- "/home/heumos/git/odgi-paper/data/odgi_extract_time.csv"
+odgi_extract <- read.delim(odgi_extract_csv, sep = ",", as.is = T, header = F, stringsAsFactors = F, strip.white = T)
+colnames(odgi_extract) <- c("threads", "haps", "run", "time", "memory")
+# time is in seconds
+# memory is in kilobytes
+odgi_extract$tool <- "odgi extract"
+summary(odgi_extract)
+
+#vg_chunk_csv <- "/home/heumos/git/odgi-paper/data/vg_chunk_time.csv"
+vg_chunk <- read.delim(vg_chunk_csv, sep = ",", as.is = T, header = F, stringsAsFactors = F, strip.white = T)
+colnames(vg_chunk) <- c("threads", "haps", "run", "time", "memory")
+vg_chunk$tool <- "vg chunk"
+summary(vg_chunk)
+
+extract <- rbind(odgi_extract, vg_chunk)
+extract$threads <- factor(extract$threads, levels=unique(extract$threads))
+extract$haps <- factor(extract$haps, levels=unique(extract$haps))
+# extract$memory <- factor(extract$memory, levels=unique(extract$memory))
+extract$memory <- as.numeric(extract$memory)/1000000
+
+extract_64haps <- extract[extract$haps == 64,]
+extract_16threads <- extract[extract$threads == 16,]
+
+dodge <- position_dodge(width = 0.0)
+
+et_p_t <- ggplot(extract_64haps, aes(x=threads, y=time, fill=tool, color = tool)) +
+  geom_violin(position = dodge) +
+  #geom_beeswarm(size=0.5) +
+  labs(x = "number of threads", y = "time in seconds") +
+  expand_limits(x = 0, y = 0) + 
+  theme(legend.position = "none")
+et_p_t
+
+eh_p_t <- ggplot(extract_16threads, aes(x=haps, y=time, fill=tool, color = tool)) +
+  geom_violin(position = dodge) +
[…]

### 2022-02-11 scripts/position.R

@@ -1,0 +1,46 @@
+#!/usr/bin/env Rscript
+
+library(tidyverse)
+library(ggbeeswarm)
+library(knitr)
+library(gridExtra)
+
+args <- commandArgs(trailingOnly = T)
+
+odgi_position_csv <- args[1]
+vg_find_csv <- args[2]
+position_csv <- args[3]
+
+#odgi_position_csv <- "/home/heumos/git/odgi-paper/data/odgi_position_time.csv"
+odgi_position <- read.delim(odgi_position_csv, sep = ",", as.is = T, header = F, stringsAsFactors = F, strip.white = T)
+colnames(odgi_position) <- c("threads", "haps", "run", "time", "memory")
+# time is in seconds
+# memory is in kilobytes
+odgi_position$tool <- "odgi position"
+summary(odgi_position)
+
+#vg_find_csv <- "/home/heumos/git/odgi-paper/data/vg_find_time.csv"
+vg_find <- read.delim(vg_find_csv, sep = ",", as.is = T, header = F, stringsAsFactors = F, strip.white = T)
+colnames(vg_find) <- c("threads", "haps", "run", "time", "memory")
+vg_find$tool <- "vg find"
+summary(vg_find)
+
+position_supp <- rbind(odgi_position, vg_find)
+position_supp$memory <- position_supp$memory/1000000
+position_supp <- aggregate(cbind(memory,time) ~threads+haps+tool, data=position_supp, FUN=mean)
+position_supp <- position_supp[order(position_supp$threads, position_supp$haps),]
+
+position_latex <- setNames(data.frame(matrix(ncol = 6, nrow = 0)), 
+                           c("threads", "haps", "time_odgi_position", "time_vg_find", "memory_odgi_position", "memory_vg_find"))
+i <- 1
+j <- i + 1
+
+while(i < dim(position_supp)[1]) {
+  row_row_row_your_boat <- c(position_supp[i,1], position_supp[i,2], position_supp[i,5], position_supp[j,5], position_supp[i,4], position_supp[j,4])
+  position_latex <- rbind(position_latex, row_row_row_your_boat)
+  i <- i + 2
+  j <- i + 1
+}
+colnames(position_latex) <-  c("threads", "haps", "time_odgi_position", "time_vg_find", "memory_odgi_position", "memory_vg_find")
+
+write.table(position_latex, file = position_csv, sep = ",", row.names = F, quote = F)

### 2022-02-11 scripts/viz.R

@@ -86,7 +86,7 @@ vh_p_m
 grid.arrange(arrangeGrob(vh_p_t, vh_p_m, nrow = 1), legend, nrow = 2, heights = c(1,0.05))
 
 viz <- arrangeGrob(arrangeGrob(vh_p_t, vh_p_m, nrow = 1), legend, nrow = 2, heights = c(1,0.05))
-ggsave(file=viz_pdf, viz, width = 7, height = 5)
+ggsave(file=viz_pdf, viz, width = 7, height = 3.5)
 file.remove("Rplots.pdf")
 
 viz_supp <- rbind(odgi_viz, vg_viz)
@@ -94,4 +94,17 @@ viz_supp$memory <- viz_supp$memory/1000000
 viz_supp <- aggregate(cbind(memory,time) ~haps+tool, data=viz_supp, FUN=mean)
 viz_supp <- viz_supp[order(viz_supp$haps),]
 
-write.table(viz_supp, file = viz_csv, sep = ",", row.names = F, quote = F)
+viz_latex <- setNames(data.frame(matrix(ncol = 5, nrow = 0)), 
+                      c("haps", "time_odgi_viz", "time_vg_viz", "memory_odgi_viz", "memory_vg_viz"))
+i <- 1
+j <- i + 1
+
+while(i < dim(viz_supp)[1]) {
+  row_row_row_your_boat <- c(viz_supp[i,1], viz_supp[i,4], viz_supp[j,4], viz_supp[i,3], viz_supp[j,3])
+  viz_latex <- rbind(viz_latex, row_row_row_your_boat)
+  i <- i + 2
+  j <- i + 1
+}
+colnames(viz_latex) <-  c("haps", "time_odgi_viz", "time_vg_viz", "memory_odgi_viz", "memory_vg_viz")
+
+write.table(viz_latex, file = viz_csv, sep = ",", row.names = F, quote = F)

### 2022-02-14 manuscript/fig_evaluation.tex
performance evaluation

@@
 	\end{subfigure}
+	\begin{subfigure}{\linewidth}
+	\caption{}
+	\centering
+	% include second image
+	\includegraphics[width=\linewidth, trim=0cm 1.25cm 0cm 0cm]{fig/performance/extract_eval.pdf}
+	\label{fig:eval-extract}
+	\end{subfigure}
 	\begin{subfigure}{1\linewidth}
@@
 	\end{subfigure}
-	\begin{subfigure}{\linewidth}
-		\caption{}
-		\centering
-		% include second image
-		\includegraphics[width=\linewidth, trim=0cm 1.25cm 0cm 0cm]{fig/performance/extract_eval.pdf}
-		\label{fig:eval-extract}
-	\end{subfigure}
-	\caption{TODO}
+	\caption{\REVIEWED{Performance on a graph of human chromosome 6 from the HPRC.
+Clearly, ODGI compares favorably to VG across all routine pangenomic tasks.
+Evaluations across threads were done using a 64 human haplotype graph.
+Evaluations across haplotypes were done using 16 threads.
+\textbf{(a)} Performance evaluation when translating a graph into the tools' respective native formats.
+\textit{odgi build}: Transformation time decreases as parallelism increases, but with diminishing per-thread processing rates.
+Transformation time increases as haplotypes increase, but in a sub-linear relationship.
+Surprisingly, less memory consumption with increasing number of threads.
+Sub-linear relationship of transformation memory and number of haplotypes.
+\textit{vg convert}: Transformation time barely decreases as parallelism increases.
+Basically linear relationship of transformation time and number of haplotypes.
+The number of threads do not influence the memory consumption.
+Basically linear relationship of transformation memory and number of haplotypes.
+\textbf{(b)} Performance evaluation when extracting the centromeric region from the HPRC graph.
+\textit{odgi extract}: Extraction time decreases as parallelism increases.
+Extraction time increases as haplotypes increase, but in a sub-linear relationship.
+The memory consumption is stable across the number of threads.
+Sub-linear relationship of extraction memory and number of haplotypes.
+\textit{vg chunk}: Extraction time barely decreases as parallelism increases.
+Basically linear relationship of extraction time and number of haplotypes.
+The number of threads do not influence the memory consumption.
+Basically linear relationship of extraction memory and number of haplotypes.
+\textbf{(c)} Performance evaluation when visualizing a graph.
+Both tools were run with only one thread.
+\textit{odgi viz}: Visualization time increases as haplotypes increase, but in a sub-linear relationship.
+The number of haplotypes has very little influence on the memory consumption.
+\textit{vg viz}: \textbf{*}A 816MB SVG was produced which can't be opened by any program.
+\textbf{**}All produced SVGs only contain an XML header, nothing else.}}
 	\label{fig:evaluation}

### 2022-02-14 manuscript/main.tex
performance evaluation

@@
 \newcommand{\REVIEWED}[1]{{\textcolor{Red}{#1}}}
+\newcommand{\beginsupplement}{%
+	\setcounter{table}{0}
+	\renewcommand{\thetable}{S\arabic{table}}%
+	\setcounter{figure}{0}
+	\renewcommand{\thefigure}{S\arabic{figure}}%
+}
 
@@
 \textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
-%\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
+%\textbf{ information:} Supplementary data are available at \textit{Bioinformatics} online.
 \\
@@
 
-We measured the efficiency of ODGI (VERSION!) and VG (VERSION) for routine pangenome tasks. In particular, we measured the execution time and memory usage 
+We compared the efficiency of ODGI (v0.6.3-56-gebc493f "Pulizia") and VG (v1.37.0 "Monchio") for routine pangenome tasks. In particular, we measured the execution time and memory usage
 \begin{inparaenum}[(i)]
-	\item of transforming a GFAv1 into the tool's native format,
-	\item the visualization of a pangenome graph, 
-	\item the extraction of a subgraph, and
+	\item of transforming a GFAv1 into a tool's native format, 
+	\item the extraction of a subgraph,
+	\item the visualization of a pangenome graph, and
 	\item the finding of path positions in a pangenome graph.
 \end{inparaenum}
-key graph operations, why
-deNBI
-data
-threads (64haps), haps (16 threads)
-general findings: VG sucks, hehehe
+These graph operations are key when it comes to the understanding of pangenome graphs.
+We ran these operations for a varying number of threads and haplotypes in the graph for a scaling analysis.
+We ran each evaluation configuration 10 times and report the mean of each run.
+All evaluations were performed on a VM in the German Network for Bioinformatics Infrastructure (deNBI) cloud. It came with 28 cores and 256GB of RAM.
+The presented results are from a 90 haplotypes chromosome 6 human pangenome graph from the HPRC.
+Specifically, the graph contains the human references GRCH38, CHM13, and the contigs of 44 individuals encoding all possible variation including telomeres and centromeres.
+
+When transforming a GFAv1 with VG, we went for the XG file format, because all downstream tools involved in the evaluation require this static graph representation model in VG.
+
+Our general findings are that ODGI makes comparatively better use of multithreading and requires much less memory (Fig.~\ref{fig:evaluation}, \REVIEWED{TODO: Link to the results of finding path positions in the supplementary.}) across all operations.
+With raising number of haplotypes in the graph, ODGI scales much butter than VG.
+For example, extracting a subgraph (Fig.~\ref{fig:eval-extract}), ODGI is up to 40 times faster and requires 8 times less memory than VG.
+Surprisingly, \textit{odgi build} occupies less memory with more threads.
+We don't have an explanation for that.
+
+Both visualization tools can only make use of a single thread.
+For a 1 haplotype graph \textit{vg viz} produces a 816MB SVG which can't be opened by any program.
+For example, GIMP2 fails with "cannot load more than 1000000 XML elements".
+For larger graphs, \textit{vg viz} runs through and produces SVGs with only the XML header.
+This makes it unusable for larger graphs.
 
@@
 thus expanding the type of possible pangenomic analyses available to the scientific community.
+
+\REVIEWED{The performance analysis showed that ODGI clearly outperforms VG when handling large, complex pangenome graphs.
+This is one major achievement, because VG's XG data structure is actually static, compared to ODGI's dynamic graph implementation.
+Across the evaluation of key graph operations, ODGI's memory peak was 10GB.
+This makes it perfectly suited to be run interactively on a recent laptop.
+VG has limitations here.
+We expect that ODGI will be able to handle the next phase of the HPRC, constructing a pangenome graph from 300 individuals, without any problems.}
 
@@
 % \end{thebibliography}
+\clearpage
+\beginsupplement
+
+\section{Supplementary}
+
+\subsection{Evaluation}
+\input{table_viz}
+
 \end{document}

### 2022-02-14 manuscript/table_viz.tex
performance evaluation

\begin{table}[!ht]
	\centering
	\caption{TODO: PLACE THE CAPTION ON THE TOP. CONDITIONALLY DIFFICULT LATEX.}
	\begin{tabular}{|l|l|l|l|l|}
		\hline
		\textbf{haps} & \textbf{time\_odgi\_viz} & \textbf{time\_vg\_viz} & \textbf{memory\_odgi\_viz} & \textbf{memory\_vg\_viz} \\ \hline \hline
		1 & \textbf{4.143} & 43.401* & \textbf{1.0401276} & 7.0868568* \\ \hline
		2 & \textbf{5.303} & 57.94** & \textbf{1.1399148} & 9.6889932** \\ \hline
		4 & \textbf{7.852} & 76.151** & \textbf{1.2726568} & 14.047634** \\ \hline
		8 & \textbf{12.951} & 109.16** & \textbf{1.5252528} & 21.839728** \\ \hline
		16 & \textbf{22.334} & 170.72** & \textbf{1.8844984} & 36.9709848** \\ \hline
		32 & \textbf{49.236} & 303.397** & \textbf{2.8689692} & 69.8805692** \\ \hline
		64 & \textbf{102.816} & 543.5** & \textbf{4.7780848} & 127.3594708** \\ \hline
	\end{tabular}
\end{table}

### 2022-02-14 scripts/build.R

@@ -34,20 +34,20 @@ construct$haps <- factor(construct$haps, levels=unique(construct$haps))
 construct$memory <- as.numeric(construct$memory)/1000000
 
 construct_64haps <- construct[construct$haps == 64,]
+construct_64haps_m <- aggregate(cbind(memory,time) ~threads+haps+tool, data=construct_64haps, FUN=mean)
 construct_16threads <- construct[construct$threads == 16,]
+construct_16threads_m <- aggregate(cbind(memory,time) ~threads+haps+tool, data=construct_16threads, FUN=mean)
 
-dodge <- position_dodge(width = 0.0)
-
-ct_p_t <- ggplot(construct_64haps, aes(x=threads, y=time, fill=tool, color = tool)) +
-  geom_violin(position = dodge) +
+ct_p_t <- ggplot(construct_64haps_m, aes(x=threads, y=time, fill=tool, color = tool, group=tool)) +
+  geom_point() + geom_line() +
   #geom_beeswarm(size=0.5) +
   labs(x = "number of threads", y = "time in seconds") +
   expand_limits(x = 0, y = 0) + 
   theme(legend.position = "none")
-#ct_p_t
+ct_p_t
 
-ch_p_t <- ggplot(construct_16threads, aes(x=haps, y=time, fill=tool, color = tool)) +
-    geom_violin(position = dodge) +
+ch_p_t <- ggplot(construct_16threads_m,aes(x=haps, y=time, fill=tool, color = tool, group=tool)) +
+  geom_point() + geom_line() +
     #geom_beeswarm(size=0.5) +
     labs(x = "number of haplotypes", y = "time in seconds") +
     expand_limits(x = 0, y = 0) + theme(legend.direction = "horizontal") + 
@@ -55,7 +55,7 @@ ch_p_t <- ggplot(construct_16threads, aes(x=haps, y=time, fill=tool, color = too
     labs(fill = "tool   ") + 
     theme(legend.title = element_text(size=15)) +
     theme(legend.key.size = unit(0.4, 'cm'))
-#ch_p_t
+ch_p_t
 
 get_legend <- function(a.gplot) {
   tmp <- ggplot_gtable(ggplot_build(a.gplot))
@@ -68,25 +68,23 @@ legend <- get_legend(ch_p_t)
 ch_p_t <- ch_p_t + theme(legend.position = "none")
 
[…]

### 2022-02-14 scripts/extract.R

@@ -33,20 +33,20 @@ extract$haps <- factor(extract$haps, levels=unique(extract$haps))
 extract$memory <- as.numeric(extract$memory)/1000000
 
 extract_64haps <- extract[extract$haps == 64,]
+extract_64haps_m <- aggregate(cbind(memory,time) ~threads+haps+tool, data=extract_64haps, FUN=mean)
 extract_16threads <- extract[extract$threads == 16,]
+extract_16threads_m <- aggregate(cbind(memory,time) ~threads+haps+tool, data=extract_16threads, FUN=mean)
 
-dodge <- position_dodge(width = 0.0)
-
-et_p_t <- ggplot(extract_64haps, aes(x=threads, y=time, fill=tool, color = tool)) +
-  geom_violin(position = dodge) +
+et_p_t <- ggplot(extract_64haps_m, aes(x=threads, y=time, fill=tool, color = tool, group=tool)) +
+  geom_point() + geom_line() +
   #geom_beeswarm(size=0.5) +
   labs(x = "number of threads", y = "time in seconds") +
   expand_limits(x = 0, y = 0) + 
   theme(legend.position = "none")
 et_p_t
 
-eh_p_t <- ggplot(extract_16threads, aes(x=haps, y=time, fill=tool, color = tool)) +
-  geom_violin(position = dodge) +
+eh_p_t <- ggplot(extract_16threads_m, aes(x=haps, y=time, fill=tool, color = tool, group=tool)) +
+  geom_point() + geom_line() +
   #geom_beeswarm(size=0.5) +
   labs(x = "number of haplotypes", y = "time in seconds") +
   expand_limits(x = 0, y = 0) + theme(legend.direction = "horizontal") + 
@@ -67,7 +67,6 @@ legend <- get_legend(eh_p_t)
 eh_p_t <- eh_p_t + theme(legend.position = "none")
 
 # collapse memory by run and take the mean
-extract_64haps_m <- aggregate(cbind(memory,time) ~threads+haps+tool, data=extract_64haps, FUN=mean)
 et_p_m <- ggplot(extract_64haps_m, aes(x=threads, y=memory, fill=tool, color = tool, group = tool)) +
   geom_point() + geom_line() +
   #geom_beeswarm(size=0.5) +
@@ -76,7 +75,6 @@ et_p_m <- ggplot(extract_64haps_m, aes(x=threads, y=memory, fill=tool, color = t
   theme(legend.position = "none")
[…]

### 2022-02-14 scripts/viz.R

@@ -37,9 +37,10 @@ second_column <- c("*", "**", "**", "**", "**", "**", "**")
 df <- data.frame(first_column, second_column)
 colnames(df) <- c("haps", "label")
 
+viz_m <- aggregate(cbind(memory,time) ~haps+tool, data=viz, FUN=mean)
 
-vh_p_t <- ggplot(viz, aes(x=haps, y=time, fill=tool, color = tool)) +
-  geom_violin(position = dodge) +
+vh_p_t <- ggplot(viz_m, aes(x=haps, y=time, fill=tool, color = tool, group=tool)) +
+  geom_point() + geom_line() +
   #geom_beeswarm(size=0.5) +
   labs(x = "number of haplotypes", y = "time in seconds") +
   expand_limits(x = 0, y = 0) + theme(legend.direction = "horizontal") + 
@@ -67,11 +68,10 @@ legend <- get_legend(vh_p_t)
 vh_p_t <- vh_p_t + theme(legend.position = "none")
 
 # collapse memory by run and take the mean
-viz_m <- aggregate(cbind(memory,time) ~haps+tool, data=viz, FUN=mean)
 vh_p_m <- ggplot(viz_m, aes(x=haps, y=memory, fill=tool, color = tool, group = tool)) +
   geom_point() + geom_line() +
   #geom_beeswarm(size=0.5) +
-  labs(x = "number of threads", y = "memory in gigabytes") +
+  labs(x = "number of haplotypes", y = "memory in gigabytes") +
   expand_limits(x = 0, y = 0) +
   theme(legend.position = "none")+
   annotate("text", label = "*", x =1, y = (5 + max(viz[viz$tool == "vg viz" & viz$haps == 1,]$memory)), size = 7.5, colour = "black") +
@@ -107,4 +107,4 @@ while(i < dim(viz_supp)[1]) {
 }
 colnames(viz_latex) <-  c("haps", "time_odgi_viz", "time_vg_viz", "memory_odgi_viz", "memory_vg_viz")
 
-write.table(viz_latex, file = viz_csv, sep = ",", row.names = F, quote = F)
+write.table(format(viz_latex, digits=2, scientific=T), file = viz_csv, sep = ",", row.names = F, quote = F)

### 2022-02-15 manuscript/table_viz.tex
viz table

@@
 	\centering
-	\caption{TODO: PLACE THE CAPTION ON THE TOP. CONDITIONALLY DIFFICULT LATEX.}
-	\begin{tabular}{|l|l|l|l|l|}
-		\hline
-		\textbf{haps} & \textbf{time\_odgi\_viz} & \textbf{time\_vg\_viz} & \textbf{memory\_odgi\_viz} & \textbf{memory\_vg\_viz} \\ \hline \hline
-		1 & \textbf{4.143} & 43.401* & \textbf{1.0401276} & 7.0868568* \\ \hline
-		2 & \textbf{5.303} & 57.94** & \textbf{1.1399148} & 9.6889932** \\ \hline
-		4 & \textbf{7.852} & 76.151** & \textbf{1.2726568} & 14.047634** \\ \hline
-		8 & \textbf{12.951} & 109.16** & \textbf{1.5252528} & 21.839728** \\ \hline
-		16 & \textbf{22.334} & 170.72** & \textbf{1.8844984} & 36.9709848** \\ \hline
-		32 & \textbf{49.236} & 303.397** & \textbf{2.8689692} & 69.8805692** \\ \hline
-		64 & \textbf{102.816} & 543.5** & \textbf{4.7780848} & 127.3594708** \\ \hline
+	\ra{1.3}
+	\caption{\label{tab:viz} Performance measurements when visualizing a graph of human chromosome 6 from the HPRC.
+\textbf{haps} is the number of haplotypes in the graph.
+Both \textit{odgi viz} and \textit{vg viz} were run with 1 thread.
+Displayed are the mean results after 10 runs.
+\textbf{*}A 816MB SVG was produced which can't be opened by any program.
+\textbf{**}All produced SVGs were empty except for an XML header.}
+	\begin{tabular}{@{}SSSSS@{}}
+
+		& \multicolumn{2}{c}{$\mathbf{time\ in\ seconds}$} & \multicolumn{2}{c}{$\mathbf{memory\ in\ gigabytes}$} \\ \cmidrule(lr){2-3} \cmidrule(lr){4-5}
+		{$\mathbf{haps}$} & {$\mathbf{odgi\ viz}$} & {$\mathbf{vg\ viz}$} & {$\mathbf{odgi\ viz}$} & {$\mathbf{vg\ viz}$} \\ \midrule
+		1 & \textbf{4.14} & 43.40\ \ * & \textbf{1.04} & 7.09\ \ * \\ 
+		2 & \textbf{5.30} & 57.94\ \ ** & \textbf{1.14} & 9.69\ \ ** \\ 
+		4 & \textbf{7.85} & 76.15\ \ ** & \textbf{1.27} & 14.05\ \ ** \\ 
+		8 & \textbf{12.95} & 109.16** & \textbf{1.53} & 21.84\ \ ** \\ 
+		16 & \textbf{22.33} & 170.72** & \textbf{1.88} & 36.97\ \ ** \\ 
+		32 & \textbf{49.24} & 303.40** & \textbf{2.87} & 69.88\ \ ** \\ 
+		64 & \textbf{102.82} & 543.50** & \textbf{4.78} & 127.36** \\ \bottomrule
 	\end{tabular}

### 2022-02-15 scripts/build.R

@@ -108,4 +108,4 @@ while(i < dim(construct_supp)[1]) {
 }
 colnames(construct_latex) <-  c("threads", "haps", "time_odgi_build", "time_vg_convert", "memory_odgi_build", "memory_vg_convert")
 
-write.table(construct_latex, file = build_csv, sep = ",", row.names = F, quote = F)
+write.table(round(construct_latex,2), file = build_csv, sep = ",", row.names = F, quote = F)

### 2022-02-15 scripts/extract.R

@@ -107,4 +107,4 @@ while(i < dim(extract_supp)[1]) {
 }
 colnames(extract_latex) <-  c("threads", "haps", "time_odgi_extract", "time_vg_chunk", "memory_odgi_extract", "memory_vg_chunk")
 
-write.table(extract_latex, file = extract_csv, sep = ",", row.names = F, quote = F)
+write.table(round(extract_latex,2), file = extract_csv, sep = ",", row.names = F, quote = F)

### 2022-02-15 scripts/position.R

@@ -43,4 +43,4 @@ while(i < dim(position_supp)[1]) {
 }
 colnames(position_latex) <-  c("threads", "haps", "time_odgi_position", "time_vg_find", "memory_odgi_position", "memory_vg_find")
 
-write.table(position_latex, file = position_csv, sep = ",", row.names = F, quote = F)
+write.table(round(position_latex, 2), file = position_csv, sep = ",", row.names = F, quote = F)

### 2022-02-15 scripts/viz.R

@@ -107,4 +107,4 @@ while(i < dim(viz_supp)[1]) {
 }
 colnames(viz_latex) <-  c("haps", "time_odgi_viz", "time_vg_viz", "memory_odgi_viz", "memory_vg_viz")
 
-write.table(format(viz_latex, digits=2, scientific=T), file = viz_csv, sep = ",", row.names = F, quote = F)
+write.table(round(viz_latex, 2), file = viz_csv, sep = ",", row.names = F, quote = F)

### 2022-02-16 manuscript/main.tex
spelling
viz table
bring in all the tables
disk space table

@@
 \usepackage{paralist}
+
+% table stuff
+\usepackage{amsfonts}
+\usepackage{booktabs}
+\usepackage{siunitx}
+\newcommand{\ra}[1]{\renewcommand{\arraystretch}{#1}}
+\usepackage{float}
+\floatstyle{plaintop}
+\restylefloat{table}
 
@@
 
-Our general findings are that ODGI makes comparatively better use of multithreading and requires much less memory (Fig.~\ref{fig:evaluation}, \REVIEWED{TODO: Link to the results of finding path positions in the supplementary.}) across all operations.
+Our general findings are that ODGI makes comparatively better use of multithreading and requires much less memory (Fig.~\ref{fig:evaluation}, Tab.~\ref{tab:position}}) across all operations.
 With raising number of haplotypes in the graph, ODGI scales much butter than VG.
@@
 This makes it unusable for larger graphs.
+
+We also measured the disk space usage of GFAv1, ODGI's, and VG's binary format (Tab.
+\ref{tab:disk_space}).
+While VG's XG occupies less disk space for smaller graphs, ODGI requires less space for graphs having 32 haplotypes or more.
+With increasing number of haplotypes the centromeric sequence in the pangenome graph also increases which can be better compressed by ODGI's id delta encoding schema.
 
@@
 \subsection{Evaluation}
+\input{table_build}
+\clearpage
+\input{table_extract}
+\clearpage
 \input{table_viz}
+\clearpage
+\input{table_position}
+\clearpage
+\input{table_disk_space}
 

### 2022-02-16 manuscript/main.tex
minor restructure

@@
 
-\section{Results}
+\section{Overview}
 
@@
 
+\section{Results}
+
+\REVIEWED{Understanding pangenome graphs is not trivial. In the following, we highlight how ODGI's tools can assist in exploring the biological features of any pangenome graph.}
+
 \subsection{Visualizing pangenome graphs}
@@
 
-Pangenome graph visualization is one of the first steps to gain insight into the mutual relationship between the sequences in the graph and their variation.
+\REVIEWED{Pangenome graph visualization is one of the first steps to gain insight into the mutual relationship between the sequences in the graph and their variation, understanding the higher-level biology of a graph.}
 We pursue a novel approach to visualization with \textit{odgi viz} and \textit{odgi draw}, two tools which provide scalable ways of generating \REVIEWED{static} pictures of the high-level structure of large pangenome graphs.
@@
 
-We compared the efficiency of ODGI (v0.6.3-56-gebc493f "Pulizia") and VG (v1.37.0 "Monchio") for routine pangenome tasks. In particular, we measured the execution time and memory usage
+\REVIEWED{We compared the efficiency of ODGI (v0.6.3-56-gebc493f "Pulizia") and VG (v1.37.0 "Monchio") for routine pangenome tasks. In particular, we measured the execution time and memory usage}
 \begin{inparaenum}[(i)]
-	\item of transforming a GFAv1 into a tool's native format, 
-	\item the extraction of a subgraph,
-	\item the visualization of a pangenome graph, and
-	\item the finding of path positions in a pangenome graph.
+	\item \REVIEWED{of transforming a GFAv1 into a tool's native format, }
+	\item \REVIEWED{the extraction of a subgraph,}
+	\item \REVIEWED{the visualization of a pangenome graph, and}
+	\item \REVIEWED{the finding of path positions in a pangenome graph.}
 \end{inparaenum}
-These graph operations are key when it comes to the understanding of pangenome graphs.
+\REVIEWED{These graph operations are key when it comes to the understanding of pangenome graphs.
 We ran these operations for a varying number of threads and haplotypes in the graph for a scaling analysis.
@@
 Specifically, the graph contains the human references GRCH38, CHM13, and the contigs of 44 individuals encoding all possible variation including telomeres and centromeres.
+}
 
-When transforming a GFAv1 with VG, we went for the XG file format, because all downstream tools involved in the evaluation require this static graph representation model in VG.
+\REVIEWED{When transforming a GFAv1 with VG, we went for the XG file format, because all downstream tools involved in the evaluation require this static graph representation model in VG.}
 
-Our general findings are that ODGI makes comparatively better use of multithreading and requires much less memory (Fig.~\ref{fig:evaluation}, Tab.~\ref{tab:position}}) across all operations.
+\REVIEWED{Our general findings are that ODGI makes comparatively better use of multithreading and requires much less memory (Fig.~\ref{fig:evaluation}, Tab.~\ref{tab:position}) across all operations.
 With raising number of haplotypes in the graph, ODGI scales much butter than VG.
@@
 We don't have an explanation for that.
-
-Both visualization tools can only make use of a single thread.
+}
+\REVIEWED{Both visualization tools can only make use of a single thread.
 For a 1 haplotype graph \textit{vg viz} produces a 816MB SVG which can't be opened by any program.
@@
 For larger graphs, \textit{vg viz} runs through and produces SVGs with only the XML header.
-This makes it unusable for larger graphs.
+This makes it unusable for larger graphs.}
 
-We also measured the disk space usage of GFAv1, ODGI's, and VG's binary format (Tab.
+\REVIEWED{We also measured the disk space usage of GFAv1, ODGI's, and VG's binary format (Tab.
 \ref{tab:disk_space}).
@@
 With increasing number of haplotypes the centromeric sequence in the pangenome graph also increases which can be better compressed by ODGI's id delta encoding schema.
+}
 

### 2022-02-16 manuscript/main.tex

@@
 
-\section*{Acknowledgements}
+\section*{Acknowledgments}
 

### 2022-02-16 manuscript/table_disk_space.tex

\begin{table}[!ht]
	\centering
	\ra{1.2}
	\caption{\label{tab:disk_space} Disk space measurements of GFAv1 and the respective tools' binary formats. \textbf{haps} is the number of haplotypes in the graph. For VG, the file size of the XG format was measured.}
	\begin{tabular}{@{}rr|rr@{}}
		& \multicolumn{3}{c}{$\mathbf{disk\ space\ in\ megabytes}$} \\ \cmidrule(lr){2-4}
		{$\mathbf{haps}$} & {$\mathbf{GFAv1}$} & {$\mathbf{ODGI}$} & {$\mathbf{VG}$} \\ \hline
		1 & 283 & 604 & \textbf{189} \\ 
		2 & 333 & 660 & \textbf{237} \\ 
		4 & 432 & 745 & \textbf{394} \\ 
		8 & 622 & 968 & \textbf{706} \\ 
		16 & 947 & 1277 & \textbf{1200} \\ 
		32 & 1657 & \textbf{2131} & 2388 \\ 
		64 & 2913 & \textbf{3745} & 4426 \\ \bottomrule
	\end{tabular}
\end{table}

### 2022-02-16 manuscript/main.tex
the grooming is loseless

--- 2022-02-16 manuscript/main.tex
@@
 while with \textit{odgi break} they can remove cycles in the graph, reducing the complexity of the graph topology.
-Furthermore, \textit{odgi groom} allows removing spurious inverting links by exploring the graph from the orientation supported by most paths.
+Furthermore, \textit{odgi groom} allows removing spurious inverting links by exploring the graph from the orientation supported by most paths\REVIEWED{;
+the process does not remove any genetic information, but only edits how the sequences are represented in the graph.}
 

### 2022-02-17 manuscript/main.tex
more comments answered

--- 2022-02-16 manuscript/main.tex
@@
 A pangenome models the full set of genomic elements in a given species or clade~\REVIEWED{\citep{cpang2018,Eizenga_2020}}.
-In contrast to reference-based approaches which relate \REVIEWED{samples} to a single genome, these data structures encode the mutual relationships between all the genomes represented.
+In contrast to reference-based approaches which relate \REVIEWED{samples} to a single genome, these data structures encode the mutual relationships between all the genomes represented\REVIEWED{~\citep{Ballouz2019}}.
 \REVIEWED{[2,5] A class of methods to represent pangenomes involves the sequence graphs~\citep{Hein1989, Paten:2017}, where homologous regions between genomes are compressed into a single representative of all alleles present in the pangenome.
@@
 With the increased adoption of long read sequencing we expect pangenomic tools to become increasingly common in \REVIEWED{the genomic studies at different taxonomic levels and} in biomedical research.
-Particularly for targets that involve complex variation, such as cancer\REVIEWED{~\citep{CompPan2016}}, plant genomics\REVIEWED{~\citep{Bayer2020}} and metagenomics\REVIEWED{~\citep{Zhong2021}}, ODGI will facilitate disentangling,
+Particularly for targets that involve complex variation, such as cancer\REVIEWED{~\citep{CompPan2016}}, plant pangenomics\REVIEWED{~\citep{Bayer2020, Liu2020, Qin2021, Li2022}}, and metagenomics\REVIEWED{~\citep{Zhong2021}}.
+\REVIEWED{Also, when studying animals like bovine~\citep{Leonard2021, BPC}.
+Currently, bacterial pangenomes are best handled by specialized tools like PPanGGolin~\citep{Gautreau2020} or PanX~\citep{Wei2017}.
+However, the latter one doesn't build a graphical representation of  a pangenome.
+But, it already has a very developed eco-system, which allows a detailed analysis of bacterial pangenomes using an interactive GUI.
+ODGI does not come with such sophisticated tools, but it can serve as a strong basis to explore viral and bacterial pangenomes on a very large scale.}
+
+ODGI will facilitate disentangling,
 describing and analyzing a much larger set of variation than previously was possible with tools that depend on short reads and reference genomes.
@@
 
-\REVIEWED{The performance analysis showed that ODGI clearly outperforms VG when handling large, complex pangenome graphs.
+\REVIEWED{The performance analysis shows that ODGI clearly outperforms VG when handling large, complex pangenome graphs.
 This is one major achievement, because VG's XG data structure is actually static, compared to ODGI's dynamic graph implementation.
@@
 However, an interactive solution that combines the 1D and 2D layout of a graph with annotation and read mapping information across different zoom levels is still missing.
-Recent interactive browsers are reference-centric~\citep{Beyer2019, Yokoyama2019, Durant2021, Liang2021} or focus primarily on 2D~\citep{Wick_2015, Gonnella2018}.
+\REVIEWED{Recent interactive pangenome graph browsers are reference-centric}~\citep{Beyer2019, Yokoyama2019, Durant2021} or focus primarily on 2D~\citep{Wick_2015, Gonnella2018}.
 Our graph sorting and layout algorithms can provide the foundation for future tools of this type.

### 2022-02-17 manuscript/fig_extract_viz_draw_position_untangle.tex
node depth, degree; MHC drawing timings

@@
             \centering
-            \caption{}
+            \caption{\label{fig:odgi_layout}}
             \includegraphics[width=\linewidth, trim=0cm 18cm -20cm 0.5cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_mhc_H1000w120}
@@
         \textbf{(d)} \textit{odgi viz} color by strandness: the red paths indicate the haplotypes that were assembled in reverse with respect to the 2 reference genomes.
-        \textbf{(e)} \textit{odgi viz} color by path depth: using the Spectra color palette with 4 levels of path depths, white indicates no depth, while grey, red, and yellow indicate depth 1, 2, and greater than or equal to 3, respectively.
-        Coloring by path depth, we can see that the two references present two different allele copies of the C4 genes, both of them including the HERV sequence.
+        \textbf{(e)} \textit{odgi viz} color by node depth: using the Spectra color palette with 4 levels of node depths, white indicates no depth, while grey, red, and yellow indicate depth 1, 2, and greater than or equal to 3, respectively.
+        Coloring by node depth, we can see that the two references present two different allele copies of the C4 genes, both of them including the HERV sequence.
         The entirely grey paths have one copy of these genes.

### 2022-02-17 manuscript/main.tex

@@
 ODGI improves on prior efforts, based on issues that arose during our work with high-quality \textit{de novo} assemblies that cover almost all parts of the human genome~\citep{Logsdon_2021,Nurk_2021}.
-In particular, we find that it is necessary to support graphs with regions of very high numbers of path traversals (high path depth).
+In particular, we find that it is necessary to support graphs with regions of very high numbers of path traversals \REVIEWED{(high node depth)}.
 Such motifs can occur in collapsed structures generated by ambiguous sequence homology relationships in repeats found in the centromeres and other segmental duplications.
@@
 \begin{itemize}
-\item \textit{odgi build} constructs the ODGI data model from GFA file (\S\ref{sec:build}).
-\item \textit{odgi view} converts the ODGI data model into GFA file (\S\ref{sec:build}).
-\item \textit{odgi viz} provides a linear visualization of the graph (\S\ref{sec:viz}).
-\item \textit{odgi draw} renders a 2D image of the graph (\S\ref{sec:viz}).
-\item \textit{odgi extract} excerpts subsets of the graph based on path ranges (\S\ref{sec:extract}).
-\item \textit{odgi explode} breaks the graph into connected components (\S\ref{sec:extract}).
-\item \textit{odgi squeeze} unifies disjoint graphs (\S\ref{sec:extract}).
-\item \textit{odgi chop} breaks long nodes into shorter ones (\S\ref{sec:edit}).
-\item \textit{odgi unchop} combines unitig nodes (\S\ref{sec:edit}).
-\item \textit{odgi break} removes cycles in the graph (\S\ref{sec:edit}).
-\item \textit{odgi prune} removes complex regions (\S\ref{sec:edit}).
-\item \textit{odgi groom} resolves spurious inverting links (\S\ref{sec:edit}).
-\item \textit{odgi position} lifts coordinates between path and graph positions (\S\ref{sec:untangle}).%
+\item~\textit{odgi build} constructs the ODGI data model from GFA file (\S\ref{sec:build}).
+\item~\textit{odgi view} converts the ODGI data model into GFA file (\S\ref{sec:build}).
+\item~\textit{odgi viz} provides a linear visualization of the graph (\S\ref{sec:viz}).
+\item~\textit{odgi draw} renders a 2D image of the graph (\S\ref{sec:viz}).
+\item~\textit{odgi extract} excerpts subsets of the graph based on path ranges (\S\ref{sec:extract}).
+\item~\textit{odgi explode} breaks the graph into connected components (\S\ref{sec:extract}).
+\item~\textit{odgi squeeze} unifies disjoint graphs (\S\ref{sec:extract}).
+\item~\textit{odgi chop} breaks long nodes into shorter ones (\S\ref{sec:edit}).
+\item~\textit{odgi unchop} combines unitig nodes (\S\ref{sec:edit}).
+\item~\textit{odgi break} removes cycles in the graph (\S\ref{sec:edit}).
+\item~\textit{odgi prune} removes complex regions (\S\ref{sec:edit}).
+\item~\textit{odgi groom} resolves spurious inverting links (\S\ref{sec:edit}).
+\item~\textit{odgi position} lifts coordinates between path and graph positions (\S\ref{sec:untangle}).%
 %\item \textit{odgi server} provides a HTTP server to lift coordinates (\S\ref{sec:untangle}).
-\item \textit{odgi untangle} deconvolutes paths relative to a reference (\S\ref{sec:untangle}).
-\item \textit{odgi tips} finds path end points relative to a reference (\S\ref{sec:untangle}).
-\item \textit{odgi sort} orders the graph nodes (\S\ref{sec:sort}).
-\item \textit{odgi layout} establishes a 2D layout (\S\ref{sec:sort}).
-\item \textit{odgi matrix} derives the pangenome matrix (\S\ref{sec:metrics}).
-\item \textit{odgi paths} lists and extracts paths in FASTA (\S\ref{sec:metrics}).
-\item \textit{odgi flatten} converts the graph to FASTA and BED (\S\ref{sec:metrics}).
-\item \textit{odgi stats} provides numerical properties of the graph (\S\ref{sec:metrics}).
-\item \textit{odgi bin} generates a summarized view of the graph (\S\ref{sec:metrics}).
-\item \textit{odgi depth} describes node depth over graph and path positions (\S\ref{sec:metrics}).
-\item \textit{odgi degree} describes node degree over graph and path positions (\S\ref{sec:metrics}).
+\item~\textit{odgi untangle} deconvolutes paths relative to a reference (\S\ref{sec:untangle}).
+\item~\textit{odgi tips} finds path end points relative to a reference (\S\ref{sec:untangle}).
+\item~\textit{odgi sort} orders the graph nodes (\S\ref{sec:sort}).
+\item~\textit{odgi layout} establishes a 2D layout (\S\ref{sec:sort}).
+\item~\textit{odgi matrix} derives the pangenome matrix (\S\ref{sec:metrics}).
+\item~\textit{odgi paths} lists and extracts paths in FASTA (\S\ref{sec:metrics}).
+\item~\textit{odgi flatten} converts the graph to FASTA and BED (\S\ref{sec:metrics}).
+\item~\textit{odgi stats} provides numerical properties of the graph (\S\ref{sec:metrics}).
+\item~\textit{odgi bin} generates a summarized view of the graph (\S\ref{sec:metrics}).
+\item~\textit{odgi depth} describes node depth over graph and path positions (\S\ref{sec:metrics}).
+\item~\textit{odgi degree} describes node degree over graph and path positions (\S\ref{sec:metrics}).
 \end{itemize}
@@
 Eukaryotic genomes experience gains and losses of genetic material, resulting in copy number variation (CNV) across the population.
-With \textit{odgi viz}, we can use multiple color palettes to color the paths by path depth, highlighting the different copy number statuses in the \REVIEWED{[1,2] genomic sequences [1,2]} represented in the pangenome graph (Fig.~\ref{fig:odgi_viz}\textbf{e}).
+With \textit{odgi viz}, we can use multiple color palettes to color the paths \REVIEWED{by how many times they traverse a node, the so-called node depth,} highlighting the different copy number statuses in the \REVIEWED{[1,2] genomic sequences [1,2]} represented in the pangenome graph (Fig.~\ref{fig:odgi_viz}\textbf{e}).
 
 \textit{odgi draw} extends the visualization in 2 dimensions (2D)  (Fig.~\ref{fig:odgi_viz}\textbf{a}) by rendering the layout built by \textit{odgi layout} (\S\ref{sec:sort}).
+\REVIEWED{It took \textit{odgi layout} and \textit{odgi draw} 6 minutes and 21 seconds in total to produce the 2D layout of the MHC locus (Fig.~\ref{fig:odgi_layout}).}
 A 2D rendering is more costly to compute, but we similarly provide an implementation that scales linearly with pangenome sequence size, allowing us to apply it to large pangenome graphs.
@@
 Finally, the portions of the paths (\textit{i.e.}, the subpaths) walking through the selected nodes are extracted and added to the new subgraph.
-Subpaths are searched in parallel, created serially, and extended in parallel again thanks to the parallelism enabled by the ODGI data structure (see \S\ref{sec:build}), making \textit{odgi extract} a scalable solution to extract also complex subregions presenting nodes with very high path depth.
+Subpaths are searched in parallel, created serially, and extended in parallel again thanks to the parallelism enabled by the ODGI data structure (see \S\ref{sec:build}), making \textit{odgi extract} a scalable solution to extract also complex subregions presenting nodes with very high node depth.
 
@@
 Particularly for targets that involve complex variation, such as cancer\REVIEWED{~\citep{CompPan2016}}, plant pangenomics\REVIEWED{~\citep{Bayer2020, Liu2020, Qin2021, Li2022}}, and metagenomics\REVIEWED{~\citep{Zhong2021}}.
-\REVIEWED{Also, when studying animals like bovine~\citep{Leonard2021, BPC}.
-Currently, bacterial pangenomes are best handled by specialized tools like PPanGGolin~\citep{Gautreau2020} or PanX~\citep{Wei2017}.
-However, the latter one doesn't build a graphical representation of  a pangenome.
+\REVIEWED{Also, when studying animals like bovine~\citep{Leonard2021, BPC}.}
+
+\REVIEWED{Currently, bacterial pangenomes are best handled by specialized tools like PPanGGolin~\citep{Gautreau2020} or PanX~\citep{Wei2017}.
+The latter one doesn't build a graphical representation of  a pangenome.
 But, it already has a very developed eco-system, which allows a detailed analysis of bacterial pangenomes using an interactive GUI.

### 2022-02-17 manuscript/fig_sorting.tex
reduce the warnings and the citation error (I'm skipping whatever remains of this entry)
sorting supplementary

@@
-\begin{figure}[h!]
+\begin{figure*}[!htb]
     \begin{subfigure}{\linewidth}
-		%\caption{}
+        \caption{}
         \centering
         % include first image
-		\includegraphics[width=1.0\linewidth, trim=-0cm 2cm 0 0cm]{fig/sorting/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_C4_bad_sorted}
-		\label{fig:bad-sorting}
+        \includegraphics[width=1.0\linewidth, trim=-0cm 0cm 0 0cm]{fig/sorting/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_mhc_random_sorted}
+        \label{fig:mhc-random-node-order}
+        \vspace{-2em}
     \end{subfigure}
-%	\includegraphics[width=\linewidth]{fig/metrics/chr4.pan.HTTex1.gfa.multiqc_odgi_stats.png}
+    \begin{subfigure}{\linewidth}
+        \caption{}
+        \centering
+        % include first image
+        \includegraphics[width=1.0\linewidth, trim=-0cm 0cm 0 0cm]{fig/sorting/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_mhc_PGSGD_sorted}
+        \label{fig:mhc-pgsdg-sorting}
+        \vspace{-2em}
+    \end{subfigure}
+    \begin{subfigure}{1\linewidth}
+        \caption{}
+        \centering
+        % include fourth image
+        \includegraphics[width=1.0\linewidth, trim=-0cm 0.5cm 0 0cm]{fig/sorting/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_C4_PGSGD_sorted}
+        \label{fig:c4-pgsdg-sorting}
+    \end{subfigure}
+    \begin{subfigure}{1\linewidth}
+        \caption{}
+        \centering
+        % include fourth image
+        \includegraphics[width=1.0\linewidth, trim=-0cm 0.5cm 0 0cm]{fig/sorting/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_C4_topologically_sorted}
+        \label{fig:c4-topological-sorting}
+    \end{subfigure}
     \caption{
-		Visualization of a poorly sorted C4 pangenome graph.
-		Same \textit{odgi viz} visualization as the one shown in Fig.~\ref{fig:odgi_viz}\textbf{e}, but with a graph where the bad node order hides the underlying copy number variation status present in the C4 region.
+        Visualizing and visualizing the major histocompatibility complex (MHC) and complement component 4 (C4) pangenome graphs.
+        All visualization are made with \textit{odgi viz} by coloring the paths by node depth:
+        using the Spectra color palette with 4 levels of node depths, white indicates no depth, while grey, red, and yellow indicate depth 1, 2, and greater than or equal to 3, respectively.
+        \textbf{(a)} Visualization of a MHC pangenome graph with nodes randomly sorted from left to right.
+        The image shows a region 5 Mbp-long.
+        The bad node order completely hides the linear graph structure.
+        \textbf{(b)} Visualization of the same MHC pangenome graph but sorted by applying the path-guided stochastic gradient descent algorithm (PG-SGD).
+        The graph globally shows a linear structure, without long range links (at the bottom of the image).
+        Furthermore, the C4 region is visible as a region with red and orange paths.
+        \textbf{(c)} Visualization of the PG-SGD sorted C4 subgraph.
+        The image shows a region 83361 bp-long.
+        The C4 region is still not well sorted locally, with the node order that still hides the underlying copy number variation status present in such a region.
+        \textbf{(d)} Visualization of the same C4 pangenome graph but sorted by applying a topological sorting.
+        The graph shows its structure: the two references present two different allele copies of the C4 genes (in red), both of them including the HERV sequence.
+        HG01071\#2 presents 3 copies of the \textit{locus} (orange), of which one contains the HERV sequence (gray in the middle of the orange).
+        In HG01952\#1, the HERV sequence is absent (white in the middle of the red).
     }
     \label{fig:sorting}
-\end{figure}
+\end{figure*}

### 2022-02-17 manuscript/main.tex
sorting supplementary

--- 2022-02-17 manuscript/main.tex
@@
 A pangenome models the full set of genomic elements in a given species or clade~\REVIEWED{\citep{cpang2018,Eizenga_2020}}.
-In contrast to reference-based approaches which relate \REVIEWED{samples} to a single genome, these data structures encode the mutual relationships between all the genomes represented\REVIEWED{~\citep{Ballouz2019}}.
+In contrast to reference-based approaches which relate \REVIEWED{samples} to a single genome, these data structures encode the mutual relationships between all the genomes represented.
 \REVIEWED{[2,5] A class of methods to represent pangenomes involves the sequence graphs~\citep{Hein1989, Paten:2017}, where homologous regions between genomes are compressed into a single representative of all alleles present in the pangenome.
@@
 
-%\input{fig_sorting}
-
 % Figure: small graph with all the stats shown. A screenshot of MultiQC, visualize summary per bin somehow, R visualizations of degree and depth HTT exon 1 as tiny example graph
@@
 With the increased adoption of long read sequencing we expect pangenomic tools to become increasingly common in \REVIEWED{the genomic studies at different taxonomic levels and} in biomedical research.
-Particularly for targets that involve complex variation, such as cancer\REVIEWED{~\citep{CompPan2016}}, plant pangenomics\REVIEWED{~\citep{Bayer2020, Liu2020, Qin2021, Li2022}}, and metagenomics\REVIEWED{~\citep{Zhong2021}}.
-\REVIEWED{Also, when studying animals like bovine~\citep{Leonard2021, BPC}.
-Currently, bacterial pangenomes are best handled by specialized tools like PPanGGolin~\citep{Gautreau2020} or PanX~\citep{Wei2017}.
-However, the latter one doesn't build a graphical representation of  a pangenome.
-But, it already has a very developed eco-system, which allows a detailed analysis of bacterial pangenomes using an interactive GUI.
-ODGI does not come with such sophisticated tools, but it can serve as a strong basis to explore viral and bacterial pangenomes on a very large scale.}
-
-ODGI will facilitate disentangling,
+Particularly for targets that involve complex variation, such as cancer\REVIEWED{~\citep{CompPan2016}}, plant genomics\REVIEWED{~\citep{Bayer2020}} and metagenomics\REVIEWED{~\citep{Zhong2021}}, ODGI will facilitate disentangling,
 describing and analyzing a much larger set of variation than previously was possible with tools that depend on short reads and reference genomes.
@@
 
-\REVIEWED{The performance analysis shows that ODGI clearly outperforms VG when handling large, complex pangenome graphs.
+\REVIEWED{The performance analysis showed that ODGI clearly outperforms VG when handling large, complex pangenome graphs.
 This is one major achievement, because VG's XG data structure is actually static, compared to ODGI's dynamic graph implementation.
@@
 However, an interactive solution that combines the 1D and 2D layout of a graph with annotation and read mapping information across different zoom levels is still missing.
-\REVIEWED{Recent interactive pangenome graph browsers are reference-centric}~\citep{Beyer2019, Yokoyama2019, Durant2021} or focus primarily on 2D~\citep{Wick_2015, Gonnella2018}.
+Recent interactive browsers are reference-centric~\citep{Beyer2019, Yokoyama2019, Durant2021, Liang2021} or focus primarily on 2D~\citep{Wick_2015, Gonnella2018}.
 Our graph sorting and layout algorithms can provide the foundation for future tools of this type.
@@
 
+%\subsection{Sorting}
+\input{fig_sorting}
+
 \end{document}

### 2022-02-17 scripts/fig_sorting.sh

@@ -4,4 +4,12 @@ filename_chr6_gfa=chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth
 
 # Take the ${filename_chr6_gfa}.C4.0.og from `fig_extract_viz_draw_position_untangle.sh`
 
-odgi sort -i chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.C4.sorted.og -o - -x 1 -Y | odgi viz -i - -o "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_C4_bad_sorted.png -c 12 -w 100 -y 50 -p chr6.selected_paths.txt -m -B Spectral:4
+odgi paths -i chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.mhc.og -L | grep 'chr6\|HG00438\|HG0107\|HG01952' > chr6.mhc.selected_paths.txt
+odgi sort -i chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.mhc.og -o - -r | odgi viz -i - -o "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_mhc_random_sorted.png -c 12 -w 5000 -y 50 -p chr6.mhc.selected_paths.txt -m -B Spectral:4 -P
+odgi sort -i chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.mhc.og -o - -Y -P | odgi viz -i - -o "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_mhc_PGSGD_sorted.png -c 12 -w 000 -y 50 -p chr6.mhc.selected_paths.txt -m -B Spectral:4 -P
+
+odgi paths -i chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.C4.sorted.og -L | grep 'chr6\|HG00438\|HG0107\|HG01952' > chr6.selected_paths.txt
+odgi sort -i chr6.pan.fa.a2fb268.4030258.6a1ecc2.smooth.C4.sorted.og -o - -Y | odgi viz -i - -o "$(echo ${filename_chr6_gfa} | tr '.' '_' )"_C4_PGSGD_sorted.png -c 12 -w 100 -y 50 -p chr6.selected_paths.txt -m -B Spectral:4 -P
[…]

### 2022-02-17 manuscript/fig_extract_viz_draw_position_untangle.tex
explicit the color

@@
         \textbf{(f)} \REVIEWED{Layout of the C4 pangenome graph made with the \textit{Bandage} tool~\citep{Wick_2015} and} annotated by using \textit{odgi position}.
-Green nodes indicate the C4 genes.
+Green nodes indicate the C4 genes (in red).
 The red rectangle highlights the regions where \textit{C4A} and \textit{C4B} genes differ.
@@
 Node labels were annoted by using \textit{odgi position}.
-        \textbf{(h)} Visualization of \textit{odgi untangle} output in the C4 pangenome graph: the plots show the copy number status of the sequences in the C4 region with respect to the grch38 reference sequence, making clear, for example, that in HG00438\#2, the \textit{C4A} gene is missing.
+        \textbf{(h)} Visualization of \textit{odgi untangle} output in the C4 pangenome graph: the plots show the copy number status of the sequences in the C4 region with respect to the grch38 reference sequence, making clear, for example, that in HG00438\#2, the \textit{C4A} gene is missing (white in the middle of the red).
 \vspace{-1em}

### 2022-02-17 manuscript/main.tex
mention the supp sorting figure in the text

--- 2022-02-17 manuscript/main.tex
@@
 ODGI provides a variety of sorting algorithms to find the best graph node order in 1 or 2 dimensions, allowing us
-to understand the sparse structures typically found in pangenome graphs and the genetic variation they represent.
+to understand the sparse structures typically found in pangenome graphs and the genetic variation they represent \REVIEWED{(Figure~\ref{fig:sorting})}.
 
@@
 %\input{fig_tips}
-
-%\input{fig_sorting}
 

### 2022-02-17 manuscript/main.tex
to jacc or not to jacc

@@
 Particularly for targets that involve complex variation, such as cancer\REVIEWED{~\citep{CompPan2016}}, plant pangenomics\REVIEWED{~\citep{Bayer2020, Liu2020, Qin2021, Li2022}}, and metagenomics\REVIEWED{~\citep{Zhong2021}}.
-\REVIEWED{Also, when studying animals like bovine~\citep{Leonard2021, BPC}.}
+\REVIEWED{Also, when studying animals like bovine~\citep{Leonard2021, Talenti2022, BPC}.}
 
@@
 But, it already has a very developed eco-system, which allows a detailed analysis of bacterial pangenomes using an interactive GUI.
-ODGI does not come with such sophisticated tools, but it can serve as a strong basis to explore viral and bacterial pangenomes on a very large scale.}
+ODGI does not come with such sophisticated tooling, but it can serve as a strong basis to explore viral and bacterial pangenomes on a very large scale.}
 
@@
 
+\subsection{Path Jaccard Index}
+\textit{odgi tips}, \textit{odgi untangle}, and \textit{odgi position} all need to translate a query path position to a target path position.
+In a repeat region, it could be that the node of the query path position is visited several times by the target path.
+We use the graph \textit{path jaccard} concept to infer the best target position for a query path position.
+For each query position we look at each possible target window and find the most-similar one.
+ The \textit{path jaccard} context mapping ensures a precise position translation process. It explores the nucleotide context
+
 %\subsection{Sorting}

### 2022-02-18 manuscript/main.tex

--- 2022-02-17 manuscript/main.tex
@@
 \usepackage{paralist}
+
+\usepackage{amsmath}
 
@@
 
+\subsection{Path Jaccard Concept}
+\textit{odgi tips}, \textit{odgi untangle}, and \textit{odgi position} all translate a query path position to a target path position.
+In a repeat region, it could be that the node of the query path position ($N_q$) is visited several times by the target path.
+We use the graph \textit{path jaccard} concept to infer the best target position for a given query path position.
+For the query position we look at each possible target window and find the most-similar one:
+\begin{enumerate}
+	\item~Starting from $N_q$ and we follow the steps of one target path position a certain nucleotide distance to the left and a certain nucleotide distance to the right, also called context.
+The ODGI default distance is 10kbp for each direction.
+	\item~We collect all node identifiers that are visited this way in a multi-set.
+	\item~Repeat the same process when following the query path position steps.
+\end{enumerate}
+Now we have two multi-sets of node identifiers, the target path position $T$ one, and the query path position $Q$ one.
+We apply the Jaccard measure to estimate the similarity between these two multi-sets (Equ.~\ref{eq:jaccard}).
+\begin{equation}
+	J(T,Q)=\frac{|T\cap Q|}{|T\cup Q|}
+	\label{eq:jaccard}
+\end{equation}
+The "$|$" indicate that, after we joined our sets together, we actually calculate the total length in nucleotides represented by the intersection node set in the nominator, and by the union node set in the denominator, respectively.
+The division of the two nucleotide lengths then gives us the jaccard measure for the two path positions.
+We collect these jaccard measurements for all possible target path positions. The largest \textit{path jaccard} determines the actual target path position for the translation.
+
+Dependent on how repeat-streaked the graph is and how it was constructed, one might want to adjust context size, to get an even more precise positional translation.
+If we can't follow the full distance to one direction, because we are at an end of a sequence, we also adjust this for all other context evaluating windows.
+
+In the following an example which is based on the graph in Figure~\ref{fig:jaccard}.
+\input{fig_jaccard}
+Let's assume, we want to translate the query path position \textit{query:3} to a path position in path \textit{target} with a context distance of 2 nucleotides per direction.
+Node \textit{2} is where \textit{query:3} is located.
+The node is traversed two times by \textit{target}.
+Therefore, we apply the \textit{path jaccard} concept to do a precise position translation.
+Starting from node \textit{2}, we collect nodes $\{1,2,4\}$ for path \textit{query}.
+For path \textit{target} we have two possible steps to start our collection from: The first or the second step in node \textit{2}.
+We obtain the node sets $T_1=\{1,2,2\}$ and $T_2=\{2,2,3\}$.
+\begin{multline}
+	J_1(T_1,Q)=\frac{|T_1\cap Q|}{|T_1\cup Q|}=\frac{|\{1,2,2\}\cap \{1,2,4\}|}{|\{1,2,2\}\cup \{1,2,4\}|}\\=\frac{|\{1,2\}|}{|\{1,2,2,4\}|}=\frac{3+4}{3+4+4+1}\approx0.58
+	\label{eq:jaccard_t1}
+\end{multline}
+\begin{multline}
+	J_2(T_2,Q)=\frac{|T_2\cap Q|}{|T_2\cup Q|}=\frac{|\{2,2,3\}\cap \{1,2,4\}|}{|\{2,2,3\}\cup \{1,2,4\}|}\\=\frac{|\{2\}|}{|\{1,2,2,3,4\}|}=\frac{4}{3+4+4+2+1}\approx0.29
+	\label{eq:jaccard_t2}
+\end{multline}
+We observe a \textit{path jaccard} of $0.58$ for $T_1/Q$ (Equ.~\ref{eq:jaccard_t1}) and $0.28$ for $T_2Q$ (Equ.~\ref{eq:jaccard_t2}).
+Therefore, we translate query path position \textit{query:3} to the target path position \textit{target:3}.
+
 %\subsection{Sorting}

### 2022-02-18 manuscript/main.tex
typo
explain graph sorting and node order optimization

@@
 \end{multline}
-We observe a \textit{path jaccard} of $0.58$ for $T_1/Q$ (Equ.~\ref{eq:jaccard_t1}) and $0.28$ for $T_2Q$ (Equ.~\ref{eq:jaccard_t2}).
+We observe a \textit{path jaccard} of $0.58$ for $T_1/Q$ (Equ.~\ref{eq:jaccard_t1}) and $0.28$ for $T_2/Q$ (Equ.~\ref{eq:jaccard_t2}).
 Therefore, we translate query path position \textit{query:3} to the target path position \textit{target:3}.
 
-%\subsection{Sorting}
+\subsection{Graph Sorting}
+Most subcommands in ODGI require and verify that the input graph’s node identifiers (IDs) are optimized.
+If this assumption is violated, \textit{odgi sort} provides functionality to optimize the graph.
+This means that the first node identifier (ID) starts at 1 and the last node ID is the number of nodes.
+The graph is then updated in place: First, the node identifiers are normalized (from 1 to the number of nodes) including the adjustment of the edges.
+Second, path information, specifically each step of a path, is updated, too.
+
+When we sort a graph, we switch the node IDs of the nodes according to the result of the sorting algorithm.
+For example, if a random sort was applied, all existing node IDs would be replaced with new, random ones (the largest node ID would still correspond to the number of nodes on the graph).
+We would update the edge and path information as described in the paragraph above.
+The reordering of nodes has a great influence on how the pangenome looks like (Fig.~\ref{fig:sorting}).
 \input{fig_sorting}

### 2022-02-21 manuscript/main.tex
spacing
ORCID identical treatment Don't break name in author list.

@@
 Simon~Heumos\orcidlink{0000-0003-3326-817X},$^{\text{\sfb 2} \dagger}$
-Sven~Nahnsen\orcidlink{https://orcid.org/0000-0002-4375-0691}$^{\text{\sfb 2,3}}$,
-Pjotr \\Prins\orcidlink{0000-0002-8021-9162}$^{\text{\sfb 4}}$,
+Sven~Nahnsen\orcidlink{0000-0002-4375-0691}$^{\text{\sfb 2,3}}$, \\
+Pjotr~Prins\orcidlink{0000-0002-8021-9162}$^{\text{\sfb 4}}$,
 and~Erik~Garrison\orcidlink{0000-0003-3821-631X}$^{\text{\sfb 4}*}$

### 2022-02-21 manuscript/main.tex
Minor comments/fixes

@@
 In contrast to reference-based approaches which relate \REVIEWED{samples} to a single genome, these data structures encode the mutual relationships between all the genomes represented\REVIEWED{~\citep{Ballouz2019}}.
-\REVIEWED{[2,5] A class of methods to represent pangenomes involves the sequence graphs~\citep{Hein1989, Paten:2017}, where homologous regions between genomes are compressed into a single representative of all alleles present in the pangenome.
-In node-labeled sequence graphs, nodes indicate \REVIEWED{[1,2] genomic sequences [1,2]}, with edges connecting the nodes that are concatenated in the sequences represented in the graph.
-A bidirected sequence graph can \REVIEWED{[1,2] represent both strands of the genomic sequences [1,2]}.
-On this model, variation graphs add the concept of paths representing linear \REVIEWED{[1,2] genomic [1,2]} sequences as traversals through the nodes of the graph~\citep{Garrison:2018}.
+\REVIEWED{[2,5] A class of methods to represent pangenomes involves sequence graphs~\citep{Hein1989, Paten:2017} where homologous regions between genomes are compressed into a single representative \FIXME{representation?} of all alleles present in the pangenome.
+In node-labeled \FIXME{?} sequence graphs, nodes indicate \REVIEWED{[1,2] genomic sequences [1,2]} with edges connecting those nodes.
+A bidirected sequence graph can \REVIEWED{[1,2] represent both strands of DNA [1,2]}.
+On this model, variation graphs add the concept of paths representing linear \REVIEWED{[1,2] DNA [1,2]} sequences as traversals through the nodes of the graph~\citep{Garrison:2018}.
 \REVIEWED{[1,2] For example, a [1,2]} path can be a genome, haplotype, contig, or read.
@@
 Although \REVIEWED{pangenome graphs} are data structures of utility to researchers~\citep{cpang2018,Garrison:2018,Baaijens_2019,Hickey:2020,Sibbesen_2021}, the scientific community still lacks a toolset \REVIEWED{capable of operating on gigabase-scale pangenome graphs constructed from whole-genome assemblies}.
-Such an effort began with the VG toolkit~\citep{Garrison:2018}\REVIEWED{, but its tools do not handle efficiently pangenome graphs presenting complex motifs that result from repetitive sequences like centromeres.}
-Here we refocus it with the Optimized Dynamic Genome/Graph Implementation (ODGI) toolkit, a compatible, but independent pangenome graph interrogation and transformation system specifically implemented to handle the data scales encountered when working \REVIEWED{[1,10]with previously constructed pangenomes comprising hundreds of haplotype-resolved genomes.[1,10]}
+Such an effort began with the VG toolkit~\citep{Garrison:2018}\REVIEWED{, but its tools do not efficiently handle pangenome graphs presenting complex motifs that result from repetitive sequences (see Fig.~\ref{fig:1}).}
+Here we refocus it \FIXME{the effort?} with the Optimized Dynamic Genome/Graph Implementation (ODGI) toolkit, a compatible, but independent pangenome graph interrogation and transformation system specifically implemented to handle the data scales encountered when working \REVIEWED{[1,10]with previously constructed pangenomes comprising hundreds of haplotype-resolved genomes.[1,10]}
 ODGI offers a set of standard operations on the variation graph data model, generalizing ``genome arithmetic'' concepts like those found in BEDTools~\citep{Quinlan_2010} to work on pangenome graphs,
@@
 This visualization is suitable for understanding graph node order, as smooth color gradients indicate that the graph order respects the linear paths' coordinate systems.
-Pangenome graphs represent both strands \REVIEWED{[1,2] of the genomic sequences [1,2]}.
+Pangenome graphs represent both strands \REVIEWED{[1,2] of genomic sequences \FIXME{DNA?}[1,2]}.
 \textit{odgi viz} supports also coloring the paths by orientation, with paths colored where their sequence is reverse-complemented (red) or in direct orientation (black) with respect to the sequences of the graph nodes (Fig.~\ref{fig:odgi_viz}\textbf{d}).
@@
 
-\REVIEWED{When transforming a GFAv1 with VG, we went for the XG file format, because all downstream tools involved in the evaluation require this static graph representation model in VG.}
+\REVIEWED{When transforming a GFAv1 with VG, we used the XG file format because all downstream tools involved in the evaluation required this static graph representation model of VG.} \FIXME{Funny grammar, can be shorter}.
 
@@
 We don't have an explanation for that.
-}
+} \FIXME{I would leave that statement out}
 \REVIEWED{Both visualization tools can only make use of a single thread.
@@
 \REVIEWED{The performance analysis shows that ODGI clearly outperforms VG when handling large, complex pangenome graphs.
-This is one major achievement, because VG's XG data structure is actually static, compared to ODGI's dynamic graph implementation.
+This is a major achievement, because VG's XG data structure is static, compared to ODGI's dynamic graph implementation.
 Across the evaluation of key graph operations, ODGI's memory peak was 10GB.

### 2022-02-28 manuscript/main.tex
table fix
Princetown

@@
 In contrast to reference-based approaches which relate \REVIEWED{samples} to a single genome, these data structures encode the mutual relationships between all the genomes represented\REVIEWED{~\citep{Ballouz2019}}.
-\REVIEWED{[2,5] A class of methods to represent pangenomes involves sequence graphs~\citep{Hein1989, Paten:2017} where homologous regions between genomes are compressed into a single representative \FIXME{representation?} of all alleles present in the pangenome.
-In node-labeled \FIXME{?} sequence graphs, nodes indicate \REVIEWED{[1,2] genomic sequences [1,2]} with edges connecting those nodes.
+\REVIEWED{[2,5] A class of methods to represent pangenomes involves sequence graphs~\citep{Hein1989, Paten:2017} where homologous regions between genomes are compressed into a single representation of all alleles present in the pangenome.
+In sequence graphs, nodes' labels are \REVIEWED{[1,2] genomic sequences [1,2]} with edges connecting those nodes.
 A bidirected sequence graph can \REVIEWED{[1,2] represent both strands of DNA [1,2]}.
@@
 Such an effort began with the VG toolkit~\citep{Garrison:2018}\REVIEWED{, but its tools do not efficiently handle pangenome graphs presenting complex motifs that result from repetitive sequences (see Fig.~\ref{fig:1}).}
-Here we refocus it \FIXME{the effort?} with the Optimized Dynamic Genome/Graph Implementation (ODGI) toolkit, a compatible, but independent pangenome graph interrogation and transformation system specifically implemented to handle the data scales encountered when working \REVIEWED{[1,10]with previously constructed pangenomes comprising hundreds of haplotype-resolved genomes.[1,10]}
+Here we refocus the effort with the Optimized Dynamic Genome/Graph Implementation (ODGI) toolkit, a compatible, but independent pangenome graph interrogation and transformation system specifically implemented to handle the data scales encountered when working \REVIEWED{[1,10]with previously constructed pangenomes comprising hundreds of haplotype-resolved genomes.[1,10]}
 ODGI offers a set of standard operations on the variation graph data model, generalizing ``genome arithmetic'' concepts like those found in BEDTools~\citep{Quinlan_2010} to work on pangenome graphs,
@@
 This visualization is suitable for understanding graph node order, as smooth color gradients indicate that the graph order respects the linear paths' coordinate systems.
-Pangenome graphs represent both strands \REVIEWED{[1,2] of genomic sequences \FIXME{DNA?}[1,2]}.
+Pangenome graphs can represent both strands \REVIEWED{[1,2] of the genomic sequences of the DNA[1,2]}.~
 \textit{odgi viz} supports also coloring the paths by orientation, with paths colored where their sequence is reverse-complemented (red) or in direct orientation (black) with respect to the sequences of the graph nodes (Fig.~\ref{fig:odgi_viz}\textbf{d}).
@@
 
-\REVIEWED{When transforming a GFAv1 with VG, we used the XG file format because all downstream tools involved in the evaluation required this static graph representation model of VG.} \FIXME{Funny grammar, can be shorter}.
+\REVIEWED{When transforming a GFAv1 with VG, the static XG file format was used. All tools involved in the evaluation process require the XG format.}
 
@@
 For example, extracting a subgraph (Fig.~\ref{fig:eval-extract}), ODGI is up to 40 times faster and requires 8 times less memory than VG.
-Surprisingly, \textit{odgi build} occupies less memory with more threads.
-We don't have an explanation for that.
-} \FIXME{I would leave that statement out}
+}
 \REVIEWED{Both visualization tools can only make use of a single thread.
@@
 The reordering of nodes has a great influence on how the pangenome looks like (Fig.~\ref{fig:sorting}).
-\input{fig_sorting}
 

### 2022-02-28 manuscript/main.tex
PanGraph, Wheat Panache

@@
 With the increased adoption of long read sequencing we expect pangenomic tools to become increasingly common in \REVIEWED{the genomic studies at different taxonomic levels and} in biomedical research.
-Particularly for targets that involve complex variation, such as cancer\REVIEWED{~\citep{CompPan2016}}, plant pangenomics\REVIEWED{~\citep{Bayer2020, Liu2020, Qin2021, Li2022}}, and metagenomics\REVIEWED{~\citep{Zhong2021}}.
+Particularly for targets that involve complex variation, such as cancer\REVIEWED{~\citep{CompPan2016}}, plant pangenomics\REVIEWED{~\citep{Bayer2020, Liu2020, Qin2021, Li2022, Bayer2022}}, and metagenomics\REVIEWED{~\citep{Zhong2021}}.
 \REVIEWED{Also, when studying animals like bovine~\citep{Leonard2021, Talenti2022, BPC}.}
 
-\REVIEWED{Currently, bacterial pangenomes are best handled by specialized tools like PPanGGolin~\citep{Gautreau2020} or PanX~\citep{Wei2017}.
+\REVIEWED{Currently, bacterial pangenomes are best handled by specialized tools like PPanGGolin~\citep{Gautreau2020}, PanGraph~\citep{Noll2022} or PanX~\citep{Wei2017}.
 The latter one doesn't build a graphical representation of  a pangenome.
@@
 However, an interactive solution that combines the 1D and 2D layout of a graph with annotation and read mapping information across different zoom levels is still missing.
-\REVIEWED{Recent interactive pangenome graph browsers are reference-centric}~\citep{Beyer2019, Yokoyama2019, Durant2021} or focus primarily on 2D~\citep{Wick_2015, Gonnella2018}.
+\REVIEWED{Recent interactive pangenome graph browsers are reference-centric~\citep{Beyer2019, Yokoyama2019}, have a limited predefined coordinate system \citep{Durant2021}}, or focus primarily on 2D~\citep{Wick_2015, Gonnella2018}.
 Our graph sorting and layout algorithms can provide the foundation for future tools of this type.

### 2022-02-28 manuscript/fig_evaluation.tex
unreadme metrics section

--- 2022-02-11 manuscript/fig_evaluation.tex
@@
 	\end{subfigure}
-	\begin{subfigure}{\linewidth}
-		\caption{}
-		\centering
-		% include second image
-		\includegraphics[width=\linewidth, trim=0cm 1.25cm 0cm 0cm]{fig/performance/extract_eval.pdf}
-		\label{fig:eval-extract}
-	\end{subfigure}
-	\caption{TODO}
+	\caption{\REVIEWED{Performance on a graph of human chromosome 6 from the HPRC.
+ODGI compares favorably to VG across all routine pangenomic tasks.
+Evaluations across threads were done using a 64 human haplotype graph.
+Evaluations across haplotypes were done using 16 threads.
+\textbf{(a)} Performance evaluation when translating a graph into the tools' respective native formats.
+\textbf{(b)} Performance evaluation when extracting the centromeric region from the HPRC graph.
+\textbf{(c)} Performance evaluation when visualizing a graph.
+Both tools were run with only one thread.
+\textit{vg viz}: \textbf{*}A 816MB SVG was produced which can't be opened by any program.
+\textbf{**}All produced SVGs only contain an XML header, nothing else.}}
 	\label{fig:evaluation}

### 2022-02-28 manuscript/fig_metrics.tex
unreadme metrics section

@@
-\begin{figure}[ht!]
+\begin{figure}[]
     \begin{subfigure}{\linewidth}
@@
 	\caption{Features of a 90-haplotype human pangenome graph of the exon 1 huntingtin gene (\textit{HTTexon1}): \textbf{(a)} Excerpt of vital statistics of the \textit{HTTexon1} graph displayed by MultiQC's ODGI module.
-The very high GC content of 73.0\% compared to a human genomic mean GC content of 40.9\% \citep{Piovesan2019} is in accordance with the literature \citep{Neueder2017}.
 \textbf{(b)} Per nucleotide node degree distribution of CHM13 in the \textit{HTTexon1} graph.
@@
 \textbf{(d)} \textit{odgi viz} visualization of the 23 largest gene alleles, CHM13, and GRCH38 of the \textit{HTTexon1} graph.
-\textbf{(e)} \textit{vg viz} nucleotide-level visualization of 10 gene alleles, CHM13, GRCH38 of the \textit{HTTexon1} graph focusing on the CAG variable repeat region.
-Figures \textbf{(b)-(e)} highlight the variant region around position 200 of CHM13 showing the variable number of glutamine residues of the different individuals as reported by \citep{Nance1999}.}
+\textbf{(e)} \textit{vg viz} nucleotide-level visualization of 10 gene alleles, CHM13, GRCH38 of the \textit{HTTexon1} graph focusing on the CAG variable repeat region.}
 	\label{fig:metrics}

### 2022-02-28 manuscript/main.tex
minimal performance figure caption
unreadme metrics section

@@
 
+\FIXME{Update the references to the sections.}
 \begin{itemize}
@@
 \item~\textit{odgi layout} establishes a 2D layout (\S\ref{sec:sort}).
-\item~\textit{odgi matrix} derives the pangenome matrix (\S\ref{sec:metrics}).
-\item~\textit{odgi paths} lists and extracts paths in FASTA (\S\ref{sec:metrics}).
-\item~\textit{odgi flatten} converts the graph to FASTA and BED (\S\ref{sec:metrics}).
+\item~\textit{odgi matrix} derives the pangenome matrix (\S\ref{sec:supp_metrics}).
+\item~\textit{odgi paths} lists and extracts paths in FASTA (\S\ref{sec:supp_metrics}).
+\item~\textit{odgi flatten} converts the graph to FASTA and BED (\S\ref{sec:supp_metrics}).
 \item~\textit{odgi stats} provides numerical properties of the graph (\S\ref{sec:metrics}).
-\item~\textit{odgi bin} generates a summarized view of the graph (\S\ref{sec:metrics}).
+\item~\textit{odgi bin} generates a summarized view of the graph (\S\ref{sec:supp_metrics}).
 \item~\textit{odgi depth} describes node depth over graph and path positions (\S\ref{sec:metrics}).
@@
 Finally, we create paths in serial, and extend them in parallel by obtaining the mutex $Node.lock$ for pairs of nodes and by adding the path step in their $Node.path\_steps$ vectors.
-This parallelism speeds ODGI model construction by many-fold when testing against graphs made from assemblies produced by the HPRC (\autoref{Evaluation}).
+This parallelism speeds ODGI model construction by many-fold when testing against graphs made from assemblies produced by the HPRC (Fig.~\ref{fig:evaluation}).
 
@@
 
-Pangenome graph topology can be derived by applying \textit{odgi matrix}, obtaining information on graph nodes connections in textual sparse matrix format.
-To investigate on the \REVIEWED{[1,2] genomic sequences [1,2]} encoded in the graph, \textit{odgi paths} allows users to calculate pairwise overlap statistics of groupings of paths and emit all path sequences in FASTA format, and it also allows the generation of a ``pangenome matrix'' that reports the copy number (presence/absence) of each path over each node.
-\textit{odgi flatten} generates a linearization of the graph by emitting the pangenome sequence (the concatenation of all node sequences) in FASTA format, and the projection of all paths on the linearized sequence in BED format.
+As a use case study (Fig.~\ref{fig:metrics}), we took a look at the metrics of a 90-haplotype human pangenome graph of the exon 1 huntingtin gene (\textit{HTTexon1}).
+In particular, we obtained the number of nodes, edges, paths, components, bases, the graph length, and the GC content with \textit{odgi stats}.
+The output pangenome statistics in YAML textual file format was given to
+MultiQC's~\citep{Ewels_2016} newly added ODGI module.
+As can be seen in Figure~\ref{fig:metrics-multiqc}, we observe a very high GC content of 73.0\% in the \textit{HTTexon1} graph compared to the human genomic mean GC content of 40.9\% \citep{Piovesan2019}.
+This is in accordance with the literature~\citep{Neueder2017}.
+%\footnote{\url{https://multiqc.info/docs/\#odgi} (accessed Oct 2021)}
+Despite this discovery, the MultiQC module provides an interactive way to comparatively explore statistics of an arbitrary number of graphs.
 
-Applying \textit{odgi stats}, users can retrieve metrics describing the graph properties, such as the number of nodes, edges, paths, and graph length.
-It outputs pangenome statistics in tab-separated values (TSV) or YAML textual file formats.
-MultiQC's~\citep{Ewels_2016} ODGI module %\footnote{\url{https://multiqc.info/docs/\#odgi} (accessed Oct 2021)}
-provides an interactive way to comparatively explore such statistics of an arbitrary number of graphs.
-
-ODGI also offers more advanced tools for the interrogation of the graphs.
-To study very large pangenomes, users can use \textit{odgi bin} to summarize the path information into bins of a specified size, generating a summarized view of gigabase scale graphs in TSV or JSON file formats.
-%This concept found its way into the pangenome graph ontology \cite{Yokoyama2020}. %maybe we want to save this up for the discussion
-
-Genomes presenting sequences with highly identical repeats result in pangenome graphs with complex motifs that can be detected by \textit{odgi depth} and \textit{odgi degree}, which return the node depth and node degree, respectively, as defined by query criteria.
-Both tools emit the output in BED format, allowing users to assess the complexity of the graph and detect intricate regions.
-Indeed, high depth/degree nodes can be the mirror of genetic variation (Fig.~\ref{fig:metrics}), but also misassemblies or problems in the pangenome building, making the tools further useful for graph validation.
+Investigating in detail which intricate regions in the \textit{HTTexon1} graph are responsible for the genetic variation and high GC content, we took a look at the per nucleotide node degree distribution (Fig.~\ref{fig:metrics-degree}), and node depth distribution (Fig.~\ref{fig:metrics-depth}) of CHM13 using \textit{odgi depth}'s, and \textit{odgi degree}'s BED output, respectively.
+The results indicate a highly polymorphic region around position 200 in the graph.
+Figure~\ref{fig:metrics-viz} supports this analysis.
+Zooming in on this region with \textit{vg viz}, we can clearly identify the typical \textit{HTTexon1} CAG variable repeat region (Fig.~\ref{fig:metrics-str}).
+Figures~\ref{fig:metrics-degree}-\ref{fig:metrics-viz} highlight the variant region around position 200 of CHM13 showing the variable number of glutamine residues of the different individuals as reported by \citep{Nance1999}.
+In general, such complex motives can be the mirror of genetic variation, but also misassemblies or problems in the pangenome building, making the tools further useful for graph validation.
 %In Fig.~\ref{fig:metrics}, the characteristics of a 90 haplotypes human pangenome graph of the exon 1 huntingtin gene~\citep{Sathasivam2013,Neueder2017} (\textit{HTTexon1}) are shown.
@@
 \end{multline}
-We observe a \textit{path jaccard} of $0.58$ for $T_1/Q$ (Equ.~\ref{eq:jaccard_t1}) and $0.28$ for $T_2/Q$ (Equ.~\ref{eq:jaccard_t2}).
+We observe a \textit{path jaccard} of $0.58$ for $T_1/Q$ (Equ.~\ref{eq:jaccard_t1}) and $0.29$ for $T_2/Q$ (Equ.~\ref{eq:jaccard_t2}).
 Therefore, we translate query path position \textit{query:3} to the target path position \textit{target:3}.
@@
 The reordering of nodes has a great influence on how the pangenome looks like (Fig.~\ref{fig:sorting}).
+\input{fig_sorting}
+
+\subsection{Graph Metrics}
+\label{sec:supp_metrics}
+Pangenome graph topology can be derived by applying \textit{odgi matrix}, obtaining information on graph nodes connections in textual sparse matrix format.
+To investigate on the \REVIEWED{[1,2] genomic sequences [1,2]} encoded in the graph, \textit{odgi paths} allows users to calculate pairwise overlap statistics of groupings of paths and emit all path sequences in FASTA format, and it also allows the generation of a ``pangenome matrix'' that reports the copy number (presence/absence) of each path over each node.
+\textit{odgi flatten} generates a linearization of the graph by emitting the pangenome sequence (the concatenation of all node sequences) in FASTA format, and the projection of all paths on the linearized sequence in BED format.
+ODGI also offers more advanced tools for the interrogation of the graphs.
+To study very large pangenomes, users can use \textit{odgi bin} to summarize the path information into bins of a specified size, generating a summarized view of gigabase scale graphs in TSV or JSON file formats.
+This concept found its way into the pangenome graph ontology \citep{Yokoyama2020}. %maybe we want to save this up for the discussion
 

### 2022-03-01 manuscript/main.tex
moved edit section into supp, reworked sorting section

@@
 Although \REVIEWED{pangenome graphs} are data structures of utility to researchers~\citep{cpang2018,Garrison:2018,Baaijens_2019,Hickey:2020,Sibbesen_2021}, the scientific community still lacks a toolset \REVIEWED{capable of operating on gigabase-scale pangenome graphs constructed from whole-genome assemblies}.
-Such an effort began with the VG toolkit~\citep{Garrison:2018}\REVIEWED{, but its tools do not efficiently handle pangenome graphs presenting complex motifs that result from repetitive sequences (see Fig.~\ref{fig:1}).}
+Such an effort began with the VG toolkit~\citep{Garrison:2018}\REVIEWED{, but its tools do not efficiently handle pangenome graphs presenting complex motifs that result from repetitive sequences (see Fig.~\ref{fig:operations}).}
 Here we refocus the effort with the Optimized Dynamic Genome/Graph Implementation (ODGI) toolkit, a compatible, but independent pangenome graph interrogation and transformation system specifically implemented to handle the data scales encountered when working \REVIEWED{[1,10]with previously constructed pangenomes comprising hundreds of haplotype-resolved genomes.[1,10]}
@@
 \item~\textit{odgi extract} excerpts subsets of the graph based on path ranges (\S\ref{sec:extract}).
-\item~\textit{odgi explode} breaks the graph into connected components (\S\ref{sec:extract}).
-\item~\textit{odgi squeeze} unifies disjoint graphs (\S\ref{sec:extract}).
-\item~\textit{odgi chop} breaks long nodes into shorter ones (\S\ref{sec:edit}).
-\item~\textit{odgi unchop} combines unitig nodes (\S\ref{sec:edit}).
-\item~\textit{odgi break} removes cycles in the graph (\S\ref{sec:edit}).
-\item~\textit{odgi prune} removes complex regions (\S\ref{sec:edit}).
-\item~\textit{odgi groom} resolves spurious inverting links (\S\ref{sec:edit}).
+\item~\textit{odgi explode} breaks the graph into connected components (\S\ref{sec:supp_edit}).
+\item~\textit{odgi squeeze} unifies disjoint graphs (\S\ref{sec:supp_edit}).
+\item~\textit{odgi chop} breaks long nodes into shorter ones (\S\ref{sec:supp_edit}).
+\item~\textit{odgi unchop} combines unitig nodes (\S\ref{sec:supp_edit}).
+\item~\textit{odgi break} removes cycles in the graph (\S\ref{sec:supp_edit}).
+\item~\textit{odgi prune} removes complex regions (\S\ref{sec:supp_edit}).
+\item~\textit{odgi groom} resolves spurious inverting links (\S\ref{sec:supp_edit}).
 \item~\textit{odgi position} lifts coordinates between path and graph positions (\S\ref{sec:untangle}).%
@@
 
+The results of other useful graph editing operations are not shown here, but described in detail in $\S$\ref{sec:supp_edit}.
+
 % If needed, we can have a figure showing extraction process---either schematic or ``real''.
-
-Pangenome graphs can embed multiple chromosomes as separated connected components (inter-chromosomal structural variants would join the components into bigger ones).
-\textit{odgi explode} separates the connected components in different ODGI format files, while \textit{odgi squeeze} allows merging multiple graphs into the same ODGI format file, preventing node ID collisions.
-
-\subsection{Editing the graph structure}
-\label{sec:edit}
-
-Pangenome graphs can be used in a variety of applications, ranging from read mapping to variant identification and genotyping~\citep{Eizenga_2020}.
-However, graphs presenting complex topology can increase the computational overhead of many downstream analyses.
-ODGI offers multiple commonly-needed basic operations on the topology of the graph and its nodes.
-
-For simplifying the graph structure, users can use \textit{odgi prune} to take away complex parts as defined by query criteria,
-while with \textit{odgi break} they can remove cycles in the graph, reducing the complexity of the graph topology.
-Furthermore, \textit{odgi groom} allows removing spurious inverting links by exploring the graph from the orientation supported by most paths\REVIEWED{;
-the process does not remove any genetic information, but only edits how the sequences are represented in the graph.}
-
-To enable efficient sequence alignment against the graph, long nodes can be divided into shorter nodes at a maximum requested size using \textit{odgi chop}.
-Partial order alignment, which consists of aligning sequences against a directed acyclic graph (DAG), is frequently used in pangenome building pipelines~\citep{pggb}, but the current implementations return DAGs with 1-bp long nodes.
-\textit{odgi unchop} allows joining nodes that can be merged without changing the graph topology, nor the embedded sequences, obtaining an equivalent, but more compact, representation of the graph.
 
@@
 
-\subsection{Sorting the pangenome graph}
+\subsection{Latent graph structure reveals underlying biology}
 \label{sec:sort}
@@
 ODGI provides a variety of sorting algorithms to find the best graph node order in 1 or 2 dimensions, allowing us
-to understand the sparse structures typically found in pangenome graphs and the genetic variation they represent \REVIEWED{(Figure~\ref{fig:sorting})}.
+to understand the sparse structures typically found in pangenome graphs and the genetic variation they represent.
 
-To find the best order of graph nodes in 1D, \textit{odgi sort} provides multiple sorting algorithms which can be combined into a sorting pipeline to take advantage of the strength of each.
-Most notably, nodes can be sorted topologically, randomly, by breaking cycles in the graph (\S\ref{sec:edit}), by grooming (\S\ref{sec:edit}), and/or by using a novel path-guided (PG) stochastic gradient descent (SGD) algorithm: PG-SGD.
+\REVIEWED{We applied a variety of \textit{odgi sort}'s 1D algorithms to a 90-haplotype human pangenome MHC and a C4 graph(Fig.~\ref{fig:sorting}). 
+The randomly sorted MHC graph (Fig.~\ref{fig:mhc-random-node-order}) hides its linear graph structure, whereas our novel path-guided (PG) stochastic gradient descent (SGD) algorithm, PG-SGD, is able to produce a globally linear ordered graph revealing the C4 region (Fig.~\ref{fig:mhc-pgsdg-sorting}).
 This exploits path information to order the graph nodes.
@@
 In parallel, each HOGWILD! thread updates node relative positions to best-match their nucleotide distance in the paths running through the graph.
-Following standard SGD approaches, a learning rate is reduced as the algorithm progresses, and execution continues until the adjustments to the model fall below a target threshold $\epsilon$.
+Following standard SGD approaches, a learning rate is reduced as the algorithm progresses, and execution continues until the adjustments to the model fall below a target threshold $\epsilon$.}
+
+\REVIEWED{A PG-SGD sorting of the C4 does not work well locally.
+It hides the underlying copy number variation status in the region (Fig.~\ref{fig:c4-pgsdg-sorting}).
+A topological sort order finally gives rise to the graph's latent structure, revealing the copy number characteristics of this C4 region (Fig.~\ref{fig:c4-topological-sorting}).
+}
+%
+\REVIEWED{To find the best order of graph nodes in 1D, \textit{odgi sort}'s multiple sorting algorithms can be combined into a sorting pipeline to take advantage of the strength of each (results not shown here).
 ODGI can project vector (in 1D) and matrix (2D) representations of the graph relative to these learned coordinate spaces.
 Based on this projection, we can trivially sort graph nodes in 1D.
-Moreover, we support the same concept in 2D in \textit{odgi layout} by providing a 2D implementation of the PG-SGD algorithm.
+Moreover, we support the same concept in 2D in \textit{odgi layout} by providing a 2D implementation of the PG-SGD algorithm (Figure~\ref{fig:odgi_layout}).
+A detailed description of the node ordering process can be found at $\S$\ref{sec:supp_sorting}.
+As we have shown above, the node order is crucial to understand the biological features of a pangenome graph.}
 %As the PG-SGD sorts graph nodes with respect to the paths which walk through them, we can also detect regions whose layout is distorted, which may represent structural variation or assembly error.
@@
 
-\subsection{Obtaining metrics of the pangenome graph}
+\subsection{Graph features highlight complexity}
 \label{sec:metrics}
 
-Graphs statistics provide alternative ways to gain insight into pangenomes complexity revealing the overall structure, size, and features of a graph and its sequences.
+\REVIEWED{Graphs statistics provide alternative ways to gain insight into pangenomes complexity revealing the overall structure, size, and features of a graph and its sequences.}
 
-As a use case study (Fig.~\ref{fig:metrics}), we took a look at the metrics of a 90-haplotype human pangenome graph of the exon 1 huntingtin gene (\textit{HTTexon1}).
+\REVIEWED{As a use case study (Fig.~\ref{fig:metrics}), we took a look at the metrics of a 90-haplotype human pangenome graph of the exon 1 huntingtin gene (\textit{HTTexon1}).
 In particular, we obtained the number of nodes, edges, paths, components, bases, the graph length, and the GC content with \textit{odgi stats}.
@@
 %\footnote{\url{https://multiqc.info/docs/\#odgi} (accessed Oct 2021)}
-Despite this discovery, the MultiQC module provides an interactive way to comparatively explore statistics of an arbitrary number of graphs.
+Despite this discovery, the MultiQC module provides an interactive way to comparatively explore statistics of an arbitrary number of graphs.}
 
-Investigating in detail which intricate regions in the \textit{HTTexon1} graph are responsible for the genetic variation and high GC content, we took a look at the per nucleotide node degree distribution (Fig.~\ref{fig:metrics-degree}), and node depth distribution (Fig.~\ref{fig:metrics-depth}) of CHM13 using \textit{odgi depth}'s, and \textit{odgi degree}'s BED output, respectively.
+\REVIEWED{Investigating in detail which intricate regions in the \textit{HTTexon1} graph are responsible for the genetic variation and high GC content, we took a look at the per nucleotide node degree distribution (Fig.~\ref{fig:metrics-degree}), and node depth distribution (Fig.~\ref{fig:metrics-depth}) of CHM13 using \textit{odgi depth}'s, and \textit{odgi degree}'s BED output, respectively.
 The results indicate a highly polymorphic region around position 200 in the graph.
@@
 In general, such complex motives can be the mirror of genetic variation, but also misassemblies or problems in the pangenome building, making the tools further useful for graph validation.
-%In Fig.~\ref{fig:metrics}, the characteristics of a 90 haplotypes human pangenome graph of the exon 1 huntingtin gene~\citep{Sathasivam2013,Neueder2017} (\textit{HTTexon1}) are shown.
+Additional graph metrics interrogation tools are detailed in $\S$\ref{sec:supp_metrics}.}
 
@@
 \begin{multline}
-	J_1(T_1,Q)=\frac{|T_1\cap Q|}{|T_1\cup Q|}=\frac{|\{1,2,2\}\cap \{1,2,4\}|}{|\{1,2,2\}\cup \{1,2,4\}|}\\=\frac{|\{1,2\}|}{|\{1,2,2,4\}|}=\frac{3+4}{3+4+4+1}\approx0.58
+	J_1(T_1,Q)=\frac{|T_1\cap Q|}{|T_1\cup Q|}=\frac{|\{1,2,2\}\cap \{1,2,4\}|}{|\{1,2,2\}\cup \{1,2,4\}|}\\=\frac{|\{1,2\}|}{|\{1,2,2,4\}|}\widehat{=}\frac{3+4}{3+4+4+1}\approx0.58
 	\label{eq:jaccard_t1}
@@
 \begin{multline}
-	J_2(T_2,Q)=\frac{|T_2\cap Q|}{|T_2\cup Q|}=\frac{|\{2,2,3\}\cap \{1,2,4\}|}{|\{2,2,3\}\cup \{1,2,4\}|}\\=\frac{|\{2\}|}{|\{1,2,2,3,4\}|}=\frac{4}{3+4+4+2+1}\approx0.29
+	J_2(T_2,Q)=\frac{|T_2\cap Q|}{|T_2\cup Q|}=\frac{|\{2,2,3\}\cap \{1,2,4\}|}{|\{2,2,3\}\cup \{1,2,4\}|}\\=\frac{|\{2\}|}{|\{1,2,2,3,4\}|}\widehat{=}\frac{4}{3+4+4+2+1}\approx0.29
 	\label{eq:jaccard_t2}
@@
 
+\subsection{Graph Editing}
+\label{sec:supp_edit}
+Pangenome graphs can embed multiple chromosomes as separated connected components (inter-chromosomal structural variants would join the components into bigger ones).
+\textit{odgi explode} separates the connected components in different ODGI format files, while \textit{odgi squeeze} allows merging multiple graphs into the same ODGI format file, preventing node ID collisions.
+
+Pangenome graphs can be used in a variety of applications, ranging from read mapping to variant identification and genotyping~\citep{Eizenga_2020}.
+However, graphs presenting complex topology can increase the computational overhead of many downstream analyses.
+ODGI offers multiple commonly-needed basic operations on the topology of the graph and its nodes.
+
+For simplifying the graph structure, users can use \textit{odgi prune} to take away complex parts as defined by query criteria,
+while with \textit{odgi break} they can remove cycles in the graph, reducing the complexity of the graph topology.
+Furthermore, \textit{odgi groom} allows removing spurious inverting links by exploring the graph from the orientation supported by most paths\REVIEWED{;
+	the process does not remove any genetic information, but only edits how the sequences are represented in the graph.}
+
+To enable efficient sequence alignment against the graph, long nodes can be divided into shorter nodes at a maximum requested size using \textit{odgi chop}.
+Partial order alignment, which consists of aligning sequences against a directed acyclic graph (DAG), is frequently used in pangenome building pipelines~\citep{pggb}, but the current implementations return DAGs with 1-bp long nodes.
+\textit{odgi unchop} allows joining nodes that can be merged without changing the graph topology, nor the embedded sequences, obtaining an equivalent, but more compact, representation of the graph.
+
 \subsection{Graph Sorting}
+\label{sec:supp_sorting}
 Most subcommands in ODGI require and verify that the input graph’s node identifiers (IDs) are optimized.

### 2022-03-01 manuscript/main.tex
cite Figure 1 in a better place; short; switch figure 2 and 3; little tweaks

@@
 Although \REVIEWED{pangenome graphs} are data structures of utility to researchers~\citep{cpang2018,Garrison:2018,Baaijens_2019,Hickey:2020,Sibbesen_2021}, the scientific community still lacks a toolset \REVIEWED{capable of operating on gigabase-scale pangenome graphs constructed from whole-genome assemblies}.
-Such an effort began with the VG toolkit~\citep{Garrison:2018}\REVIEWED{, but its tools do not efficiently handle pangenome graphs presenting complex motifs that result from repetitive sequences (see Fig.~\ref{fig:operations}).}
+Such an effort began with the VG toolkit~\citep{Garrison:2018}\REVIEWED{, but its tools do not efficiently handle pangenome graphs presenting complex motifs that result from repetitive sequences.}
 Here we refocus the effort with the Optimized Dynamic Genome/Graph Implementation (ODGI) toolkit, a compatible, but independent pangenome graph interrogation and transformation system specifically implemented to handle the data scales encountered when working \REVIEWED{[1,10]with previously constructed pangenomes comprising hundreds of haplotype-resolved genomes.[1,10]}
-ODGI offers a set of standard operations on the variation graph data model, generalizing ``genome arithmetic'' concepts like those found in BEDTools~\citep{Quinlan_2010} to work on pangenome graphs,
-and providing a variety of operations, such as visualization, sorting, and liftover projections, all critical to understand and exploit pangenome graphs.
+ODGI offers a set of standard operations on the variation graph data model (Fig.~\ref{fig:operations}) , generalizing ``genome arithmetic'' concepts, like those found in BEDTools~\citep{Quinlan_2010}, to work on pangenome graphs.
+Furthermore, it provides a variety of tools for graph visualization, sorting, and liftover projections, all critical to understand and exploit pangenome graphs.
 
@@
 
-The results of other useful graph editing operations are not shown here, but described in detail in $\S$\ref{sec:supp_edit}.
+The results of other useful graph editing operations are described in $\S$\ref{sec:supp_edit}.
 
@@
 
+\input{fig_extract_viz_draw_position_untangle}
+
 \subsection{Latent graph structure reveals underlying biology}
@@
 
-\REVIEWED{We applied a variety of \textit{odgi sort}'s 1D algorithms to a 90-haplotype human pangenome MHC and a C4 graph(Fig.~\ref{fig:sorting}). 
+\REVIEWED{We applied a variety of \textit{odgi sort}'s 1D algorithms to a 90-haplotype human pangenome MHC and a C4 graph (Fig.~\ref{fig:sorting}).
 The randomly sorted MHC graph (Fig.~\ref{fig:mhc-random-node-order}) hides its linear graph structure, whereas our novel path-guided (PG) stochastic gradient descent (SGD) algorithm, PG-SGD, is able to produce a globally linear ordered graph revealing the C4 region (Fig.~\ref{fig:mhc-pgsdg-sorting}).
@@
 %For understanding the large-scale organization of the graph, we benefit from an even simpler coordinate system.
-PG-SGD learns a 1D or 2D organization of the graph nodes that matches distances in graph paths.
+PG-SGD learns a 1D or 2D organization of the graph nodes that matches nucleotide distances in graph paths (i.e., the sequences embedded in the graph).
 To scale to large graphs, we learn this projection in parallel via a HOGWILD! approach~\citep{niu2011hogwild}.
@@
 %
-\REVIEWED{To find the best order of graph nodes in 1D, \textit{odgi sort}'s multiple sorting algorithms can be combined into a sorting pipeline to take advantage of the strength of each (results not shown here).
+\REVIEWED{To find the best order of graph nodes in 1D, \textit{odgi sort}'s multiple sorting algorithms can be combined into a sorting pipeline to take advantage of the strength of each (results not shown).
 ODGI can project vector (in 1D) and matrix (2D) representations of the graph relative to these learned coordinate spaces.
@@
 }
-
-\input{fig_extract_viz_draw_position_untangle}
 

### 2022-03-01 manuscript/main.tex
better figure's labels management; tweak title 5.5

@@
 \usepackage{amsmath}
+
+\usepackage{cleveref}
 
@@
 \REVIEWED{Pangenome graph visualization is one of the first steps to gain insight into the mutual relationship between the sequences in the graph and their variation, understanding the higher-level biology of a graph.}
-We pursue a novel approach to visualization with \textit{odgi viz} and \textit{odgi draw}, two tools which provide scalable ways of generating \REVIEWED{static} pictures of the high-level structure of large pangenome graphs.
+We pursue a novel approach to visualization with \textit{odgi viz} and \textit{odgi draw}, two tools that provide scalable ways of generating \REVIEWED{static} pictures of the high-level structure of even large pangenome graphs.
 
-\textit{odgi viz} supports a binned, linearized rendering in 1 dimension (1D) (that is, all graph nodes lie on the same axis).
+\textit{odgi viz} supports a binned, linearized rendering in 1 dimension (1D) (i.e., all graph nodes lie on the same axis).
 This visualization is computed in linear-time and offers a human-interpretable format suitable for understanding the topology and genome relationships in the pangenome graph (Fig.~\ref{fig:odgi_viz}).
@@
 The meaning of the colors depends on how \textit{odgi viz} is executed.
-By default, path colors are derived from the path names (Fig.~\ref{fig:odgi_viz}\textbf{b}).
+By default, path colors are derived from the path names (Fig.~\ref{fig:viz_extract_untangle-C4-default}).
 Path names are displayed on the left of the paths.
@@
 Nevertheless, complex, nonlinear graph structures are difficult to display and interpret in a low number of dimensions.
-To overcome such a limitation, \textit{odgi viz} supports multiple visualization modalities (Fig.~\ref{fig:odgi_viz}\textbf{c-e}), making it easy to grasp the properties and shape of the graph.
+To overcome such a limitation, \textit{odgi viz} supports multiple visualization modalities (Fig.~\labelcref{fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage}), making it easy to grasp the properties and shape of the graph.
 Graph node order can affect downstream analyses on pangenome graphs.
-With \textit{odgi viz} we can color the paths by path position (Fig.~\ref{fig:odgi_viz}\textbf{c}), with light grey indicating where paths begin and dark grey where they end.
+With \textit{odgi viz} we can color the paths by path position (Fig.~\ref{fig:viz_extract_untangle-C4-gradient}), with light grey indicating where paths begin and dark grey where they end.
 This visualization is suitable for understanding graph node order, as smooth color gradients indicate that the graph order respects the linear paths' coordinate systems.
 Pangenome graphs can represent both strands \REVIEWED{[1,2] of the genomic sequences of the DNA[1,2]}.~
-\textit{odgi viz} supports also coloring the paths by orientation, with paths colored where their sequence is reverse-complemented (red) or in direct orientation (black) with respect to the sequences of the graph nodes (Fig.~\ref{fig:odgi_viz}\textbf{d}).
+\textit{odgi viz} supports also coloring the paths by orientation, with paths colored where their sequence is reverse-complemented (red) or in direct orientation (black) with respect to the sequences of the graph nodes (Fig.~\ref{fig:viz_extract_untangle-C4-strandness}).
 Eukaryotic genomes experience gains and losses of genetic material, resulting in copy number variation (CNV) across the population.
-With \textit{odgi viz}, we can use multiple color palettes to color the paths \REVIEWED{by how many times they traverse a node, the so-called node depth,} highlighting the different copy number statuses in the \REVIEWED{[1,2] genomic sequences [1,2]} represented in the pangenome graph (Fig.~\ref{fig:odgi_viz}\textbf{e}).
+With \textit{odgi viz}, we can use multiple color palettes to color the paths \REVIEWED{by how many times they traverse a node, the so-called node depth,} highlighting the different copy number statuses in the \REVIEWED{[1,2] genomic sequences [1,2]} represented in the pangenome graph (Fig.~\ref{fig:viz_extract_untangle-C4-coverage}).
 
-\textit{odgi draw} extends the visualization in 2 dimensions (2D)  (Fig.~\ref{fig:odgi_viz}\textbf{a}) by rendering the layout built by \textit{odgi layout} (\S\ref{sec:sort}).
+\textit{odgi draw} extends the visualization in 2 dimensions (2D)  (Fig.~\ref{fig:viz_extract_untangle-MHC}) by rendering the layout built by \textit{odgi layout} (\S\ref{sec:sort}).
 \REVIEWED{It took \textit{odgi layout} and \textit{odgi draw} 6 minutes and 21 seconds in total to produce the 2D layout of the MHC locus (Fig.~\ref{fig:odgi_layout}).}
@@
 Pangenome graphs built from hundreds of haplotype-resolved \textit{de novo} genome assemblies are very large, but it is often necessary to work with only a small portion of the genomes represented,
-such as a specific \textit{locus} (Fig.~\ref{fig:odgi_viz}\textbf{a}) or a smaller region (Fig.~\ref{fig:odgi_viz}\textbf{b-g}), or even a single gene (Fig.~\ref{fig:metrics}).
+such as a specific \textit{locus} (Fig.~\ref{fig:viz_extract_untangle-MHC}) or a smaller region (Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage,fig:viz_extract_untangle-C4-bandage,fig:viz_extract_untangle-C4-bandage-zoom}), or even a single gene (Fig.~\ref{fig:metrics}).
 This simplifies the downstream analyses and reduces the resources to work only with the extracted graphs.
@@
 \textit{odgi position} translates graph and path positions between or within graphs, emitting the liftovers in BED format.
-Coordinates can be specified in BED format, but users can even specify a GFF/GTF file to project the annotations into the pangenome graph (Fig.~\ref{fig:odgi_viz}\textbf{f-g}).
+Coordinates can be specified in BED format, but users can even specify a GFF/GTF file to project the annotations into the pangenome graph (Fig.~\labelcref{fig:viz_extract_untangle-C4-bandage,fig:viz_extract_untangle-C4-bandage-zoom}).
 
@@
 Being able to work with any sets of reference sequences lets us convert the graph to lift-over maps compatible with standard software for projecting annotations and alignments from one genome to another.
-As an example, by untangling the graph we can study the variation that lies in regions collapsed due to ambiguous alignments over sequence repeats (as shown in Fig.~\ref{fig:odgi_viz}\textbf{f}).
-Indeed, to obtain a more precise overview of the \textit{locus} in Fig.~\ref{fig:odgi_viz}\textbf{b-e},
+As an example, by untangling the graph we can study the variation that lies in regions collapsed due to ambiguous alignments over sequence repeats (as shown in Fig.~\ref{fig:viz_extract_untangle-C4-bandage}).
+Indeed, to obtain a more precise overview of the \textit{locus} in Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage},
 we can apply \textit{odgi untangle} to segment paths into linear segments by breaking these segments where the paths loop back on themselves.
@@
 %The \textit{path jaccard} concept ensures that the breaking is nucleotide-precise.
-In this way, we obtain information on the position and copy number status of the sequences in the collapsed \textit{locus} (Fig.~\ref{fig:odgi_viz}\textbf{h}).
+In this way, we obtain information on the position and copy number status of the sequences in the collapsed \textit{locus} (Fig.~\ref{fig:viz_extract_untangle-C4-untangle}).
 Moreover, to obtain base-level precise information on the relationships between the repeated sequences, we can align them by using the pairs of regions that came from the untangling to guide the alignment~\citep{wfmash}.
@@
 
-\subsection{Graph features highlight complexity}
+\subsection{Graph features highlight variation}
 \label{sec:metrics}

### 2022-03-01 manuscript/fig_extract_viz_draw_position_untangle.tex
better figure's labels management; tweak title 5.5
move a bit of biology from the figure's description to the main text

@@
             \includegraphics[width=\linewidth, trim=0cm 18cm -20cm 0.5cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_mhc_H1000w120}
+            \label{fig:viz_extract_untangle-MHC}
         \end{subfigure} \\
@@
                 \includegraphics[width=\linewidth, trim=0 1.9cm 0cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_C4_sorted}
+                \label{fig:viz_extract_untangle-C4-default}
             \end{subfigure} \\
@@
                 \includegraphics[width=\linewidth, trim=0 +1.9cm 0cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_C4_sorted_du}
+                \label{fig:viz_extract_untangle-C4-gradient}
             \end{subfigure} \\
@@
                 \includegraphics[width=\linewidth, trim=0 +1.9cm 0cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_C4_sorted_z}
+                \label{fig:viz_extract_untangle-C4-strandness}
             \end{subfigure} \\
@@
                 \includegraphics[width=\linewidth, trim=0 +1.9cm 0cm 0cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_C4_sorted_m}
+                \label{fig:viz_extract_untangle-C4-coverage}
             \end{subfigure}
@@
                 \caption{}
+                \label{fig:viz_extract_untangle-C4-bandage}
             \end{subfigure}
@@
                 \caption{}
+                \label{fig:viz_extract_untangle-C4-bandage-zoom}
             \end{subfigure}
@@
                 \includegraphics[width=\linewidth, trim=0cm 0.9cm -2.5cm 0.5cm]{fig/extract_viz_draw_position_untangle/chr6_pan_fa_a2fb268_4030258_6a1ecc2_smooth_C4_sorted_untangle_bed}
+                \label{fig:viz_extract_untangle-C4-untangle}
             \end{center}
@@
 \textbf{(a)} \textit{odgi draw} layout of the MHC pangenome graph extracted from a whole human pangenome graph of 90 haplotypes.
-        \REVIEWED{The MHC locus is located on the human chromosome 6 and it encodes molecules involved in antigen presentation, inflammation regulation, the complement system, and the innate and adaptive immune responses~\citep{Shiina2009}.}
-        The red rectangle highlights the C4 region\REVIEWED{, which encodes proteins involved in the complement system.}
-        \textbf{(b-e)} \textit{odgi viz} visualizations of the C4 pangenome graph, where 8 paths are displayed: 2 reference genomes (chm13 and grch38 on the top) and 6 haplotypes of 3 individuals.
+        The red rectangle highlights the C4 region.
+        \textbf{(b-e)} \textit{odgi viz} visualizations of the C4 pangenome graph, where 8 paths are displayed: 2 reference genomes (chm13 and grch38 on the top) and 6 haplotypes of 3 diploid individuals.
         \textbf{(b)} \textit{odgi viz} default modality: the image shows a quite linear graph.
-        The longer links at the bottom indicate the presence of a structural variant (long link) with another structural variant nested inside it (short link on the left).
-        Indeed, human C4 exists as 2 functionally distinct genes, \textit{C4A} and \textit{C4B}, which both vary in structure and copy number~\citep{Sekar_2016}.
-The longer link indicates that the copy number status varies across the haplotypes represented in the pangenome.
-        Moreover, \textit{C4A} and \textit{C4B} genes segregate in both long and short genomic forms, distinguished by the presence or absence of a human endogenous retroviral (HERV) sequence, as also highlighted by the short nested link on the left.
+        The links at the bottom indicate the presence of a structural variant (long link) with another structural variant nested inside it (short link on the left).
         \textbf{(c)} Color by path position. %: the color gradients are smooth, highlighting that its node are well sorted in 1 dimension.

### 2022-03-01 manuscript/main.tex
move a bit of biology from the figure's description to the main text

@@
 \REVIEWED{Pangenome graph visualization is one of the first steps to gain insight into the mutual relationship between the sequences in the graph and their variation, understanding the higher-level biology of a graph.}
-We pursue a novel approach to visualization with \textit{odgi viz} and \textit{odgi draw}, two tools that provide scalable ways of generating \REVIEWED{static} pictures of the high-level structure of even large pangenome graphs.
+We pursue a novel approach to visualization with \textit{odgi viz} and \textit{odgi draw}, two tools that provide scalable ways of generating \REVIEWED{static} pictures of the high-level structure of even large pangenome graphs (Fig.~\ref{fig:odgi_viz}).
 
-\textit{odgi viz} supports a binned, linearized rendering in 1 dimension (1D) (i.e., all graph nodes lie on the same axis).
-This visualization is computed in linear-time and offers a human-interpretable format suitable for understanding the topology and genome relationships in the pangenome graph (Fig.~\ref{fig:odgi_viz}).
-Graph nodes are arranged on a single axis, from left to right, with the colored bar indicating the paths and the nodes they cross.
+\textit{odgi viz} supports a binned, linearized rendering in 1 dimension (1D) (i.e., \REVIEWED{each path lies} on a single axis) (Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage}).
+For each path, graph nodes are arranged from left to right, with the colored bars indicating the paths and the nodes they cross.
 White spaces indicate where paths do not traverse the nodes.
@@
 The meaning of the colors depends on how \textit{odgi viz} is executed.
-By default, path colors are derived from the path names (Fig.~\ref{fig:viz_extract_untangle-C4-default}).
-Path names are displayed on the left of the paths.
-The black lines on the bottom indicate the edges connecting the nodes and, therefore, represent the graph topology.
-\REVIEWED{Given a path X traversing two nodes A and B, the corresponding edge is represented by a black line starting from the left or right of node A if this node is traversed in reverse or forward, respectively, by path X;
-the black line ends to the left or right of node B if this node is traversed in forward or reverse, respectively, by path X.
-Consequently, if two consecutive nodes are linked both in forward, no edge is shown (it would be 0 pixels long, as it would start from the right of the first node and ends to the left of the second one).}
+By default, path colors are derived from the path names (Fig.~\ref{fig:viz_extract_untangle-C4-default}), which are displayed on the left of the paths.
+The black lines on the bottom indicate the edges connecting the nodes and, therefore, represent the graph topology (see \S\ref{sec:supp_visualization} for a more detailed explanation).
+This visualization is computed in linear-time and offers a human-interpretable format suitable for understanding the topology and genome relationships in the pangenome graph\REVIEWED{, as shown when applied to the complement component 4 (C4) region.}
+\REVIEWED{Such region encodes proteins involved in the complement system.
+In humans, it exists as 2 functionally distinct genes, \textit{C4A} and \textit{C4B}, which both vary in structure and copy number~\citep{Sekar_2016}.
+Indeed, the longer link at the bottom of Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage} indicates that the copy number status of these genes varies across the haplotypes in the represented pangenome.
+Moreover, \textit{C4A} and \textit{C4B} genes segregate in both long and short genomic forms, distinguished by the presence or absence of a human endogenous retroviral (HERV) sequence, as also highlighted by the short nested link on the left.}
 
@@
 This visualization is suitable for understanding graph node order, as smooth color gradients indicate that the graph order respects the linear paths' coordinate systems.
-Pangenome graphs can represent both strands \REVIEWED{[1,2] of the genomic sequences of the DNA[1,2]}.~
+Pangenome graphs can represent both strands \REVIEWED{[1,2] of the genomic sequences of the DNA[1,2]}.
 \textit{odgi viz} supports also coloring the paths by orientation, with paths colored where their sequence is reverse-complemented (red) or in direct orientation (black) with respect to the sequences of the graph nodes (Fig.~\ref{fig:viz_extract_untangle-C4-strandness}).
@@
 
-\textit{odgi draw} extends the visualization in 2 dimensions (2D)  (Fig.~\ref{fig:viz_extract_untangle-MHC}) by rendering the layout built by \textit{odgi layout} (\S\ref{sec:sort}).
-\REVIEWED{It took \textit{odgi layout} and \textit{odgi draw} 6 minutes and 21 seconds in total to produce the 2D layout of the MHC locus (Fig.~\ref{fig:odgi_layout}).}
-A 2D rendering is more costly to compute, but we similarly provide an implementation that scales linearly with pangenome sequence size, allowing us to apply it to large pangenome graphs.
+\textit{odgi draw} extends the visualization in 2 dimensions (2D)  (Fig.~\ref{fig:viz_extract_untangle-MHC}) by rendering the graph layout built by \textit{odgi layout} (\S\ref{sec:sort})\REVIEWED{, as shown when applied to the major histocompatibility complex (MHC) locus.}
+\REVIEWED{The MHC locus is located on the human chromosome 6 and it encodes molecules involved in antigen presentation, inflammation regulation, the complement system, and the innate and adaptive immune responses~\citep{Shiina2009}.}
+\REVIEWED{Fig.~\ref{fig:viz_extract_untangle-MHC} shows the topology of the MHC pangenome graph and how the paths traverse the graph, with the regions of variations that appear as bubbles in the layout.}
+Of note, a 2D rendering is more costly to compute, but we provide an implementation that scales linearly with pangenome sequence size, allowing us to apply it to large pangenome graphs.
+%\REVIEWED{For example, \textit{odgi draw} took 6 minutes and 21 seconds in total to produce the 2D layout of the MHC locus (Fig.~\ref{fig:odgi_layout}).}
 
@@
 %Subsetting operations allow us to zoom in on particular genes or regions of interest for more precise examination.
+
 Graph portions can be extracted by using the paths in the graph as coordinate systems to guide the process.
@@
 Regions of interest can be specified by graph nodes or path range(s), also in BED format. Furthermore, it is possible to indicate a list of paths to be preserved completely in the extracted graph.
-
-In \textit{odgi extract}, we begin by collecting all graph nodes that fall within the ranges to extract (and the paths to preserve, if requested).
+We begin by collecting all graph nodes that fall within the ranges to extract (and the paths to preserve, if requested).
 Users can specify the number of steps or nucleotides to expand the selection and include neighboring nodes.
 Then, edges connecting all selected nodes are added in the subgraph under construction.
-Finally, the portions of the paths (\textit{i.e.}, the subpaths) walking through the selected nodes are extracted and added to the new subgraph.
+Finally, the portions of the paths (i.e., the subpaths) walking through the selected nodes are extracted and added to the new subgraph.
 Subpaths are searched in parallel, created serially, and extended in parallel again thanks to the parallelism enabled by the ODGI data structure (see \S\ref{sec:build}), making \textit{odgi extract} a scalable solution to extract also complex subregions presenting nodes with very high node depth.
 
-The results of other useful graph editing operations are described in $\S$\ref{sec:supp_edit}.
-
-% If needed, we can have a figure showing extraction process---either schematic or ``real''.
+Other graph editing operations are described in $\S$\ref{sec:supp_edit}.
 
@@
 Coordinates can be specified in BED format, but users can even specify a GFF/GTF file to project the annotations into the pangenome graph (Fig.~\labelcref{fig:viz_extract_untangle-C4-bandage,fig:viz_extract_untangle-C4-bandage-zoom}).
-
 For a precise translation process when conversing a query position to a reference position in a repeat region, we apply the \textit{path jaccard} context mapping concept.
@@
 
-\input{fig_extract_viz_draw_position_untangle}
-
 \subsection{Latent graph structure reveals underlying biology}
@@
 
+\input{fig_extract_viz_draw_position_untangle}
 \input{fig_metrics}
@@
 
+\subsection{Graph topology visualization}
+\label{sec:supp_visualization}
+\REVIEWED{In \textit{odgi viz} visualization, given a path X traversing two nodes A and B, the corresponding edge is represented by a black line starting from the left or right of node A if this node is traversed in reverse or forward, respectively, by path X;
+the black line ends to the left or right of node B if this node is traversed in forward or reverse, respectively, by path X.
+Consequently, if two consecutive nodes are linked both in forward, no edge is shown (it would be 0 pixels long, as it would start from the right of the first node and ends to the left of the second one).}
+
 \subsection{Evaluation}

### 2022-03-02 manuscript/main.tex
fix section order in Supplementary

@@
 
-The key data in a pangenome graph is a representation of the alignment (or homology) relationships between the sequences represented.
+The key data in a pangenome graph is a representation of the alignment or the homology relationships between the sequences represented.
 Navigating and understanding the graph requires coordinate systems that we can use to link other data into the graph model, and thus to the \REVIEWED{[1,2] genomic sequences [1,2]} in the pangenome.
 ODGI's tools use the embedded sequences to provide a universal coordinate space that is graph-independent, thereby remaining stable across different graphs built with the same genomes.
+Such a a universal coordinate system allows us to support several kinds of ``lift-over'' of coordinates between different genomes in the same or different graphs.
 %Our for positional conversion (\textit{odgi position}), annotation liftover (\textit{odgi untangle}), breakpoint identification (\textit{odgi tips}), subgraph extraction (\textit{odgi extract}), visualization (\textit{odgi viz}), path depth (\textit{odgi depth}), and graph complexity measurement (\textit{odgi degree}) can all operate on coordinates of paths in the graph.
 
-A universal coordinate system allows us to support several kinds of ``lift-over'' of coordinates between different genomes in the same or different graphs.
-%For basic liftover operations, \textit{odgi server} provides a basic HTTP server that can translate very efficiently a path position to a pangenomic position within the same graph.
+As a demonstration of this universality, we
+
 \textit{odgi position} translates graph and path positions between or within graphs, emitting the liftovers in BED format.
@@
 ~\ref{fig:tips}).
-\textit{odgi tips} can identify the break point positions of the contigs relative to the reference(s) in the graph by walking from the ends of each contig until a reference node is found.
-It could be that the reference visits the node several times.
-Therefore, for each contig range (a tip) \textit{odgi tips} takes a look at each possible reference window and finds the most similar one using the \textit{path jaccard} concept.
-The output is a BED file with the best reference hit and position for each of the contigs' ends.
 
@@
 \label{sec:supp_visualization}
-\REVIEWED{In \textit{odgi viz} visualization, given a path X traversing two nodes A and B, the corresponding edge is represented by a black line starting from the left or right of node A if this node is traversed in reverse or forward, respectively, by path X;
+\REVIEWED{In a \textit{odgi viz} visualization, given a path X traversing two nodes A and B, the corresponding edge is represented by a black line starting from the left or right of node A if this node is traversed in reverse or forward, respectively, by path X;
 the black line ends to the left or right of node B if this node is traversed in forward or reverse, respectively, by path X.
@@
 
-\subsection{Evaluation}
-\input{table_build}
-\clearpage
-\input{table_extract}
-\clearpage
-\input{table_viz}
-\clearpage
-\input{table_position}
-\clearpage
-\input{table_disk_space}
+\subsection{Graph navigation}
+\textit{odgi tips} can identify the break point positions of the contigs relative to the reference(s) in the graph by walking from the ends of each contig until a reference node is found.
+It could be that the reference visits the node several times.
+Therefore, for each contig range (a tip) \textit{odgi tips} takes a look at each possible reference window and finds the most similar one using the \textit{path jaccard} concept.
+The output is a BED file with the best reference hit and position for each of the contigs' ends.
 
-\subsection{Path Jaccard Concept}
+\subsubsection{Path Jaccard concept}
 \textit{odgi tips}, \textit{odgi untangle}, and \textit{odgi position} all translate a query path position to a target path position.
@@
 
+\clearpage
+\subsection{Evaluation}
+\input{table_build}
+\clearpage
+\input{table_extract}
+\clearpage
+\input{table_viz}
+\clearpage
+\input{table_position}
+\clearpage
+\input{table_disk_space}
+
 \end{document}

### 2022-03-02 manuscript/fig_extract_viz_draw_position_untangle.tex
unreadme untangling and navigating

@@
         The red rectangle highlights the C4 region.
-        \textbf{(b-e)} \textit{odgi viz} visualizations of the C4 pangenome graph, where 8 paths are displayed: 2 reference genomes (chm13 and grch38 on the top) and 6 haplotypes of 3 diploid individuals.
+        \textbf{(b-e)} \textit{odgi viz} visualizations of the C4 pangenome graph, where 8 paths are displayed: 2 reference genomes (\REVIEWED{CHM13} and \REVIEWED{GRCH38} on the top) and 6 haplotypes of 3 diploid individuals.
         \textbf{(b)} \textit{odgi viz} default modality: the image shows a quite linear graph.
@@
 Node labels were annoted by using \textit{odgi position}.
-        \textbf{(h)} Visualization of \textit{odgi untangle} output in the C4 pangenome graph: the plots show the copy number status of the sequences in the C4 region with respect to the grch38 reference sequence, making clear, for example, that in HG00438\#2, the \textit{C4A} gene is missing (white in the middle of the red).
+        \textbf{(h)} Visualization of \textit{odgi untangle} output in the C4 pangenome graph: the plots show the copy number status of the sequences in the C4 region with respect to the \REVIEWED{GRCH38} reference sequence, making clear, for example, that in HG00438\#2, the \textit{C4A} gene is missing (white in the middle of the red).
 \vspace{-1em}

### 2022-03-02 manuscript/main.tex
unreadme untangling and navigating

@@
 \item~\textit{odgi untangle} deconvolutes paths relative to a reference (\S\ref{sec:untangle}).
-\item~\textit{odgi tips} finds path end points relative to a reference (\S\ref{sec:untangle}).
+\item~\textit{odgi tips} finds path end points relative to a reference (\S\ref{sec:supp_navigation}).
 \item~\textit{odgi sort} orders the graph nodes (\S\ref{sec:sort}).
@@
 This visualization is suitable for understanding graph node order, as smooth color gradients indicate that the graph order respects the linear paths' coordinate systems.
-Pangenome graphs can represent both strands \REVIEWED{[1,2] of the genomic sequences of the DNA[1,2]}.
+Pangenome graphs can represent both strands \REVIEWED{[1,2] of the genomic sequences of the DNA[1,2]}.~
 \textit{odgi viz} supports also coloring the paths by orientation, with paths colored where their sequence is reverse-complemented (red) or in direct orientation (black) with respect to the sequences of the graph nodes (Fig.~\ref{fig:viz_extract_untangle-C4-strandness}).
@@
 
-As a demonstration of this universality, we
+\REVIEWED{As a demonstration of this universality, we took the C4 graph and added gene annotation from GRCH38 to its nodes from a GFF file using \textit{odgi position} ($\S$\ref{sec:supp_gff}).
+The resulting TSV contains pairs of node and color.
+Taking the graph and the TSV into Bandage~\citep{Wick_2015}, the actual C4 gene region is highlighted (Fig.~\ref{fig:viz_extract_untangle-C4-bandage}).
+Zooming to the nucleotide level, the annotation shows the single nucleotide differences of the \textit{C4A} and the \textit{C4B} genes (Fig.~\ref{fig:viz_extract_untangle-C4-bandage-zoom}).}
 
-\textit{odgi position} translates graph and path positions between or within graphs, emitting the liftovers in BED format.
-Coordinates can be specified in BED format, but users can even specify a GFF/GTF file to project the annotations into the pangenome graph (Fig.~\labelcref{fig:viz_extract_untangle-C4-bandage,fig:viz_extract_untangle-C4-bandage-zoom}).
+\textit{odgi position} can also translate graph and path positions between or within graphs, emitting the liftovers in BED format.
 For a precise translation process when conversing a query position to a reference position in a repeat region, we apply the \textit{path jaccard} context mapping concept.
@@
 10kbp) has the best jaccard metric when compared to the query context.
+\REVIEWED{For a more detailed explanation of the \textit{path jaccard} concept please see $\S$\ref{sec:supp_path_jaccard}.}
 %the best jaccard for the query step and each reference candidate step is calculated by looking at a context range (e.g. 10kbp) around the locus.
@@
 
-Pangenome graphs \REVIEWED{can} model alignments of many genomes.
-With \textit{odgi untangle}, users can extract pairwise alignment information between a given set of ``query'' sequences and a given set of ``target'' sequences (used as references).
-While pangenome graphs may contain looping structures that imply many-to-many alignments between the pangenome sequences, these untangled alignments map each segment of the queries to a single segment in the set of targets.
+\REVIEWED{To obtain a more precise overview of the \textit{locus} in Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage},
+we applied \textit{odgi untangle} with GRCH38 as a reference. \textit{odgi untangle} segments paths into linear segments by breaking these segments where the paths loop back on themselves.
+In this way, we obtain information on the position and copy number status of the sequences in the collapsed \textit{locus} (Fig.~\ref{fig:viz_extract_untangle-C4-untangle}).
+The shape of the line indicates if the copy number is in forward or in reverse orientation compared to GRCH38.
+\textit{odgi untangle} demonstrates that it is able to work with any sets of reference sequences converting the graph to lift-over maps compatible with standard software for projecting annotations and alignments from one genome to another.
+An explanation of the untangling process is given in $\S$\ref{sec:supp_navigation}.}
 %This lets us decompose both homology and paralogy relationships by also untangling looping motifs in the graph, and obtaining linearized pairs of regions in BEDPE or PAF format.
-Being able to work with any sets of reference sequences lets us convert the graph to lift-over maps compatible with standard software for projecting annotations and alignments from one genome to another.
-As an example, by untangling the graph we can study the variation that lies in regions collapsed due to ambiguous alignments over sequence repeats (as shown in Fig.~\ref{fig:viz_extract_untangle-C4-bandage}).
-Indeed, to obtain a more precise overview of the \textit{locus} in Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage},
-we can apply \textit{odgi untangle} to segment paths into linear segments by breaking these segments where the paths loop back on themselves.
-We first discover segment boundaries using standard approaches for detecting repeats in sequence graphs \citep{Pevzner_2004}.
-We finally ``untangle'' by finding the target segment that best match each query segment using the \textit{path jaccard} context mapping model.
-%The \textit{path jaccard} concept ensures that the breaking is nucleotide-precise.
-In this way, we obtain information on the position and copy number status of the sequences in the collapsed \textit{locus} (Fig.~\ref{fig:viz_extract_untangle-C4-untangle}).
-Moreover, to obtain base-level precise information on the relationships between the repeated sequences, we can align them by using the pairs of regions that came from the untangling to guide the alignment~\citep{wfmash}.
+%As an example, by untangling the graph we can study the variation that lies in regions collapsed due to ambiguous alignments over sequence repeats (as shown in Fig.~\ref{fig:viz_extract_untangle-C4-bandage}).
 
@@
 
-\subsection{Graph navigation}
+\subsection{Graph navigation and untangling}
+\label{sec:supp_navigation}
+Pangenome graphs \REVIEWED{can} model alignments of many genomes.
+With \textit{odgi untangle}, users can extract pairwise alignment information between a given set of ``query'' sequences and a given set of ``target'' sequences (used as references).
+While pangenome graphs may contain looping structures that imply many-to-many alignments between the pangenome sequences, these untangled alignments map each segment of the queries to a single segment in the set of targets.
+\textit{odgi untangle} first discovers segment boundaries using standard approaches for detecting repeats in sequence graphs \citep{Pevzner_2004}.
+We finally ``untangle'' by finding the target segment that best match each query segment using the \textit{path jaccard} context mapping model.
+%The \textit{path jaccard} concept ensures that the breaking is nucleotide-precise.
+Moreover, to obtain base-level precise information on the relationships between the repeated sequences, we can align them by using the pairs of regions that came from the untangling to guide the alignment~\citep{wfmash}.
+
 \textit{odgi tips} can identify the break point positions of the contigs relative to the reference(s) in the graph by walking from the ends of each contig until a reference node is found.
@@
 
+\subsubsection{The GFF liftover}
+\label{sec:supp_gff}
+A GFF file contains annotations for one or more paths in the graph.
+For each annotation, we know the start and end within that path.
+So we can annotate all nodes that are visited by such a path range with the information from the attribute field.
+If there are overlapping features, we append the annotation for each node.
+Using the same coloring schema as in \textit{odgi viz} we generate a color for each annotated node by its collected annotation.
+
+If a subgraph was as a result from e.g.
+\textit{odgi extract}, the path names are usually in the form of name:start-end.
+\textit{odgi position} is able to automatically detect this and adjust the positions given in the GFF on the fly to the new positions given in the subgraph.
+For each GFF entry, it just subtracts the “missing” number of nucleotides from the start and end field.
+That’s how we adjust for the subgraph annotation.
+
 \subsubsection{Path Jaccard concept}
+\label{sec:supp_path_jaccard}
 \textit{odgi tips}, \textit{odgi untangle}, and \textit{odgi position} all translate a query path position to a target path position.

### 2022-03-02 manuscript/main.tex
smoother `visualizing pangenome graphs` paragraph

@@
 \REVIEWED{Pangenome graph visualization is one of the first steps to gain insight into the mutual relationship between the sequences in the graph and their variation, understanding the higher-level biology of a graph.}
-We pursue a novel approach to visualization with \textit{odgi viz} and \textit{odgi draw}, two tools that provide scalable ways of generating \REVIEWED{static} pictures of the high-level structure of even large pangenome graphs (Fig.~\ref{fig:odgi_viz}).
+We pursue a novel approach to visualization with \textit{odgi draw} and \textit{odgi viz}, two tools that provide scalable ways of generating \REVIEWED{static} pictures of the high-level structure of even large pangenome graphs (Fig.~\ref{fig:odgi_viz}).
 
-\textit{odgi viz} supports a binned, linearized rendering in 1 dimension (1D) (i.e., \REVIEWED{each path lies} on a single axis) (Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage}).
+\REVIEWED{The the major histocompatibility complex (MHC) locus is located on the human chromosome 6 and it encodes molecules involved in antigen presentation, inflammation regulation, the complement system, and the innate and adaptive immune responses~\citep{Shiina2009}.
+MHC genes are highly polymorphic, i.e., there are multiple different alleles across individuals in a population.
+Such variability becomes evident when we apply \textit{odgi draw} to visualize the graph layout of a human MHC pangenome graph (Fig.~\ref{fig:viz_extract_untangle-MHC}).
+The visualization displays the graph topology in 2 dimensions (2D) and how the paths traverse the graph, with the regions of variations that appear as bubbles in the layout.
+The layout was built by \textit{odgi layout} (see \S\ref{sec:sort}).} Of note, a 2D rendering can be costly to compute, but we provide an implementation that scales linearly with pangenome sequence size, allowing us to apply it to large pangenome graphs.
+%\REVIEWED{For example, \textit{odgi draw} took 6 minutes and 21 seconds in total to produce the 2D layout of the MHC locus (Fig.~\ref{fig:odgi_layout}).}
+
+\REVIEWED{The MHC locus includes the complement component 4 (C4) region, which encodes proteins involved in the complement system.
+In Fig.~\ref{fig:viz_extract_untangle-MHC}, such a region corresponds to the little bubble highlighted by the red rectangle.
+To take a closer look, we obtained a C4 pangenome graph by extracting the C4 region from the full MHC pangenome graph (see \S\ref{sec:extract}).
+Then, we visualized such a subgraph by applying} \textit{odgi viz}, which produces binned, linearized renderings in 1 dimension (1D) (i.e., \REVIEWED{each path lies} on a single axis) (Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage}).
 For each path, graph nodes are arranged from left to right, with the colored bars indicating the paths and the nodes they cross.
@@
 The black lines on the bottom indicate the edges connecting the nodes and, therefore, represent the graph topology (see \S\ref{sec:supp_visualization} for a more detailed explanation).
-This visualization is computed in linear-time and offers a human-interpretable format suitable for understanding the topology and genome relationships in the pangenome graph\REVIEWED{, as shown when applied to the complement component 4 (C4) region.}
-\REVIEWED{Such region encodes proteins involved in the complement system.
-In humans, it exists as 2 functionally distinct genes, \textit{C4A} and \textit{C4B}, which both vary in structure and copy number~\citep{Sekar_2016}.
-Indeed, the longer link at the bottom of Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage} indicates that the copy number status of these genes varies across the haplotypes in the represented pangenome.
-Moreover, \textit{C4A} and \textit{C4B} genes segregate in both long and short genomic forms, distinguished by the presence or absence of a human endogenous retroviral (HERV) sequence, as also highlighted by the short nested link on the left.}
+This visualization is computed in linear-time and offers a human-interpretable format suitable for understanding the topology and genome relationships in the pangenome graph.
+\REVIEWED{In humans, it exists as 2 functionally distinct genes, \textit{C4A} and \textit{C4B}, which both vary in structure and copy number~\citep{Sekar_2016}.
+Indeed, the longer link at the bottom of Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage} indicates that the copy number status of these genes varies across the haplotypes represented in the pangenome.
+Moreover, the short nested link on the left highlights that \textit{C4A} and \textit{C4B} genes segregate in both long and short genomic forms, distinguished by the presence or absence of a human endogenous retroviral (HERV) sequence.}
 
-Nevertheless, complex, nonlinear graph structures are difficult to display and interpret in a low number of dimensions.
+Nevertheless, complex, nonlinear graph structures are difficult to interpret in a low number of dimensions.
 To overcome such a limitation, \textit{odgi viz} supports multiple visualization modalities (Fig.~\labelcref{fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage}), making it easy to grasp the properties and shape of the graph.
-Graph node order can affect downstream analyses on pangenome graphs.
-With \textit{odgi viz} we can color the paths by path position (Fig.~\ref{fig:viz_extract_untangle-C4-gradient}), with light grey indicating where paths begin and dark grey where they end.
-This visualization is suitable for understanding graph node order, as smooth color gradients indicate that the graph order respects the linear paths' coordinate systems.
-Pangenome graphs can represent both strands \REVIEWED{[1,2] of the genomic sequences of the DNA[1,2]}.~
-\textit{odgi viz} supports also coloring the paths by orientation, with paths colored where their sequence is reverse-complemented (red) or in direct orientation (black) with respect to the sequences of the graph nodes (Fig.~\ref{fig:viz_extract_untangle-C4-strandness}).
-Eukaryotic genomes experience gains and losses of genetic material, resulting in copy number variation (CNV) across the population.
-With \textit{odgi viz}, we can use multiple color palettes to color the paths \REVIEWED{by how many times they traverse a node, the so-called node depth,} highlighting the different copy number statuses in the \REVIEWED{[1,2] genomic sequences [1,2]} represented in the pangenome graph (Fig.~\ref{fig:viz_extract_untangle-C4-coverage}).
+%Graph node order can affect downstream analyses on pangenome graphs.
+For example, we can color the paths by path position (Fig.~\ref{fig:viz_extract_untangle-C4-gradient}), with light grey indicating where paths begin and dark grey where they end.
+This visualization is suitable for understanding graph node order, as smooth color gradients indicate that the node order respects the linear paths' coordinate systems.
+Pangenome graphs can represent both strands \REVIEWED{[1,2] of the genomic sequences of the DNA[1,2]}.
+We can display such information by coloring the paths by orientation, with paths colored where their sequence is reverse-complemented (red) or in direct orientation (black) with respect to the sequences of the graph nodes (Fig.~\ref{fig:viz_extract_untangle-C4-strandness}).
+%Eukaryotic genomes experience gains and losses of genetic material, resulting in copy number variation (CNV) across the population.
+Furthermore, we can use multiple color palettes to color the paths \REVIEWED{by how many times they traverse a node, the so-called node depth.}.
+\REVIEWED{This highlights that in the C4 pangenome graph, the haplotypes present different number of copies of the C4 genes (Fig.~\ref{fig:viz_extract_untangle-C4-coverage}).}
 
-\textit{odgi draw} extends the visualization in 2 dimensions (2D)  (Fig.~\ref{fig:viz_extract_untangle-MHC}) by rendering the graph layout built by \textit{odgi layout} (\S\ref{sec:sort})\REVIEWED{, as shown when applied to the major histocompatibility complex (MHC) locus.}
-\REVIEWED{The MHC locus is located on the human chromosome 6 and it encodes molecules involved in antigen presentation, inflammation regulation, the complement system, and the innate and adaptive immune responses~\citep{Shiina2009}.}
-\REVIEWED{Fig.~\ref{fig:viz_extract_untangle-MHC} shows the topology of the MHC pangenome graph and how the paths traverse the graph, with the regions of variations that appear as bubbles in the layout.}
-Of note, a 2D rendering is more costly to compute, but we provide an implementation that scales linearly with pangenome sequence size, allowing us to apply it to large pangenome graphs.
-%\REVIEWED{For example, \textit{odgi draw} took 6 minutes and 21 seconds in total to produce the 2D layout of the MHC locus (Fig.~\ref{fig:odgi_layout}).}
+\input{fig_extract_viz_draw_position_untangle}
 
@@
 
-\input{fig_extract_viz_draw_position_untangle}
 \input{fig_metrics}

### 2022-03-02 manuscript/main.tex
squeeze and rework the layout

@@
 
-The key data in a pangenome graph is a representation of the alignment or the homology relationships between the sequences represented.
-Navigating and understanding the graph requires coordinate systems that we can use to link other data into the graph model, and thus to the \REVIEWED{[1,2] genomic sequences [1,2]} in the pangenome.
-ODGI's tools use the embedded sequences to provide a universal coordinate space that is graph-independent, thereby remaining stable across different graphs built with the same genomes.
-Such a a universal coordinate system allows us to support several kinds of ``lift-over'' of coordinates between different genomes in the same or different graphs.
+The key data in a pangenome graph is a representation of the alignment (i.e., the homology relationships) between \REVIEWED{[1,2] genomic sequences [1,2]}.
+Navigating and understanding the graph requires coordinate systems to link other data to the sequences represented in the graph model.
+ODGI's tools use the embedded sequences to provide a universal coordinate space that is graph-independent, thereby remaining stable across different graphs built with the same sequences.
+Such a a universal coordinate system allows us to support several kinds of ``lift-over'' of coordinates between different sequences in the same or different graphs.
 %Our for positional conversion (\textit{odgi position}), annotation liftover (\textit{odgi untangle}), breakpoint identification (\textit{odgi tips}), subgraph extraction (\textit{odgi extract}), visualization (\textit{odgi viz}), path depth (\textit{odgi depth}), and graph complexity measurement (\textit{odgi degree}) can all operate on coordinates of paths in the graph.
-
-\REVIEWED{As a demonstration of this universality, we took the C4 graph and added gene annotation from GRCH38 to its nodes from a GFF file using \textit{odgi position} ($\S$\ref{sec:supp_gff}).
+\REVIEWED{As a demonstration, we took the C4 pangenome graph and added to its nodes gene annotation from GRCH38 (in GFF format file) using \textit{odgi position} ($\S$\ref{sec:supp_gff}).
 The resulting TSV contains pairs of node and color.
-Taking the graph and the TSV into Bandage~\citep{Wick_2015}, the actual C4 gene region is highlighted (Fig.~\ref{fig:viz_extract_untangle-C4-bandage}).
+Taking the graph and the TSV into Bandage~\citep{Wick_2015}, the actual C4 genes are highlighted (Fig.~\ref{fig:viz_extract_untangle-C4-bandage}).
 Zooming to the nucleotide level, the annotation shows the single nucleotide differences of the \textit{C4A} and the \textit{C4B} genes (Fig.~\ref{fig:viz_extract_untangle-C4-bandage-zoom}).}
@@
 10kbp) has the best jaccard metric when compared to the query context.
-\REVIEWED{For a more detailed explanation of the \textit{path jaccard} concept please see $\S$\ref{sec:supp_path_jaccard}.}
+\REVIEWED{For a more detailed explanation of the \textit{path jaccard} concept see $\S$\ref{sec:supp_path_jaccard}.}
 %the best jaccard for the query step and each reference candidate step is calculated by looking at a context range (e.g. 10kbp) around the locus.

### 2022-03-03 manuscript/main.tex
on heterozygosity and polyploidy
extract the extract

@@
 To support interchange with other pangenome tools or text-based processing, \textit{odgi view} converts a graph in ODGI binary format to GFAv1.
+
+As the GFA format itself does not store metadata information, ODGI follows the PanSN-spec~\citep{pansn-spec} convention for dealing with heterozygosity and polyploidy in the sequence name.
+This harmonizes the biosample information present in FASTA, VCF, BED, and GFF/GTF formats.
 %ODGI uses the ``S'' and ``L'' records to describe the graph topology, and ``P'' records to describe paths.
@@
 \REVIEWED{Understanding pangenome graphs is not trivial. In the following, we highlight how ODGI's tools can assist in exploring the biological features of any pangenome graph.}
+
+\REVIEWED{Pangenome graphs built from hundreds of haplotype-resolved \textit{de novo} genome assemblies are very large, but it is often necessary to work with only a small portion of the genomes represented,
+such as a specific \textit{locus} (Fig.~\ref{fig:viz_extract_untangle-MHC}) or a smaller region (Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage,fig:viz_extract_untangle-C4-bandage,fig:viz_extract_untangle-C4-bandage-zoom}), or even a single gene (Fig.~\ref{fig:metrics}).
+This simplifies the downstream analyses and reduces the resources to work only with the extracted graphs. More on graph extraction and edit operations can be found at $\S$\ref{sec:supp_edit}.}
+%Subsetting operations allow us to zoom in on particular genes or regions of interest for more precise examination.
 
@@
 
-\REVIEWED{The the major histocompatibility complex (MHC) locus is located on the human chromosome 6 and it encodes molecules involved in antigen presentation, inflammation regulation, the complement system, and the innate and adaptive immune responses~\citep{Shiina2009}.
+\REVIEWED{Using \textit{odgi extract}, we extracted the the major histocompatibility complex (MHC) locus from a 90 haplotypes human pangenome chromosome 6 graph form the HPRC.
+Specifically the graph contains the human references GRCH38, CHM13, and the contigs of 44 individuals that encode all possible variation including telomeres and centromeres.
+The MHC genes are involved in antigen presentation, inflammation regulation, the complement system, and the innate and adaptive immune responses~\citep{Shiina2009}.
 MHC genes are highly polymorphic, i.e., there are multiple different alleles across individuals in a population.
@@
 In Fig.~\ref{fig:viz_extract_untangle-MHC}, such a region corresponds to the little bubble highlighted by the red rectangle.
-To take a closer look, we obtained a C4 pangenome graph by extracting the C4 region from the full MHC pangenome graph (see \S\ref{sec:extract}).
+As an example use case, we took a closer look, at the C4 region of the MHC by extracting it from the full MHC pangenome graph with \textit{odgi extract}.
 Then, we visualized such a subgraph by applying} \textit{odgi viz}, which produces binned, linearized renderings in 1 dimension (1D) (i.e., \REVIEWED{each path lies} on a single axis) (Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage}).
@@
 This visualization is computed in linear-time and offers a human-interpretable format suitable for understanding the topology and genome relationships in the pangenome graph.
-\REVIEWED{In humans, it exists as 2 functionally distinct genes, \textit{C4A} and \textit{C4B}, which both vary in structure and copy number~\citep{Sekar_2016}.
+\REVIEWED{In humans, the C4 gene exists as 2 functionally distinct genes, \textit{C4A} and \textit{C4B}, which both vary in structure and copy number~\citep{Sekar_2016}.
 Indeed, the longer link at the bottom of Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage} indicates that the copy number status of these genes varies across the haplotypes represented in the pangenome.
@@
 %Eukaryotic genomes experience gains and losses of genetic material, resulting in copy number variation (CNV) across the population.
-Furthermore, we can use multiple color palettes to color the paths \REVIEWED{by how many times they traverse a node, the so-called node depth.}.
+Furthermore, we can use multiple color palettes to color the paths \REVIEWED{by how many times they traverse a node, the so-called node depth.}
 \REVIEWED{This highlights that in the C4 pangenome graph, the haplotypes present different number of copies of the C4 genes (Fig.~\ref{fig:viz_extract_untangle-C4-coverage}).}
@@
 \input{fig_extract_viz_draw_position_untangle}
-
-\subsection{Extracting or joining regions of interest}
-\label{sec:extract}
-
-Pangenome graphs built from hundreds of haplotype-resolved \textit{de novo} genome assemblies are very large, but it is often necessary to work with only a small portion of the genomes represented,
-such as a specific \textit{locus} (Fig.~\ref{fig:viz_extract_untangle-MHC}) or a smaller region (Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage,fig:viz_extract_untangle-C4-bandage,fig:viz_extract_untangle-C4-bandage-zoom}), or even a single gene (Fig.~\ref{fig:metrics}).
-This simplifies the downstream analyses and reduces the resources to work only with the extracted graphs.
-%Subsetting operations allow us to zoom in on particular genes or regions of interest for more precise examination.
-
-Graph portions can be extracted by using the paths in the graph as coordinate systems to guide the process.
-For such operation, \textit{odgi extract} allows users to extract specific regions of the graph as defined by query criteria.
-Regions of interest can be specified by graph nodes or path range(s), also in BED format. Furthermore, it is possible to indicate a list of paths to be preserved completely in the extracted graph.
-We begin by collecting all graph nodes that fall within the ranges to extract (and the paths to preserve, if requested).
-Users can specify the number of steps or nucleotides to expand the selection and include neighboring nodes.
-Then, edges connecting all selected nodes are added in the subgraph under construction.
-Finally, the portions of the paths (i.e., the subpaths) walking through the selected nodes are extracted and added to the new subgraph.
-Subpaths are searched in parallel, created serially, and extended in parallel again thanks to the parallelism enabled by the ODGI data structure (see \S\ref{sec:build}), making \textit{odgi extract} a scalable solution to extract also complex subregions presenting nodes with very high node depth.
-
-Other graph editing operations are described in $\S$\ref{sec:supp_edit}.
 
@@
 \label{sec:supp_edit}
+Graph portions can be extracted by using the paths in the graph as coordinate systems to guide the process.
+For such operation, \textit{odgi extract} allows users to extract specific regions of the graph as defined by query criteria.
+Regions of interest can be specified by graph nodes or path range(s), also in BED format. Furthermore, it is possible to indicate a list of paths to be preserved completely in the extracted graph.
+We begin by collecting all graph nodes that fall within the ranges to extract (and the paths to preserve, if requested).
+Users can specify the number of steps or nucleotides to expand the selection and include neighboring nodes.
+Then, edges connecting all selected nodes are added in the subgraph under construction.
+Finally, the portions of the paths (i.e., the subpaths) walking through the selected nodes are extracted and added to the new subgraph.
+Subpaths are searched in parallel, created serially, and extended in parallel again thanks to the parallelism enabled by the ODGI data structure (see \S\ref{sec:build}), making \textit{odgi extract} a scalable solution to extract also complex subregions presenting nodes with very high node depth.
+
 Pangenome graphs can embed multiple chromosomes as separated connected components (inter-chromosomal structural variants would join the components into bigger ones).

### 2022-03-07 manuscript/main.tex
fix layout distortions
fix wrong ref
pantools reference

@@
 \item~\textit{odgi draw} renders a 2D image of the graph (\S\ref{sec:viz}).
-\item~\textit{odgi extract} excerpts subsets of the graph based on path ranges (\S\ref{sec:extract}).
+\item~\textit{odgi extract} excerpts subsets of the graph based on path ranges (\S\ref{sec:supp_edit}).
 \item~\textit{odgi explode} breaks the graph into connected components (\S\ref{sec:supp_edit}).
@@
 
+\vspace{0.22cm}
+
 \subsection{Evaluation}
@@
 Particularly for targets that involve complex variation, such as cancer\REVIEWED{~\citep{CompPan2016}}, plant pangenomics\REVIEWED{~\citep{Bayer2020, Liu2020, Qin2021, Li2022, Bayer2022}}, and metagenomics\REVIEWED{~\citep{Zhong2021}}.
-\REVIEWED{Also, when studying animals like bovine~\citep{Leonard2021, Talenti2022, BPC}.}
+\REVIEWED{Also, when studying animals like bovines~\citep{Leonard2021, Talenti2022, BPC}.}
 
@@
 But, it already has a very developed eco-system, which allows a detailed analysis of bacterial pangenomes using an interactive GUI.
-ODGI does not come with such sophisticated tooling, but it can serve as a strong basis to explore viral and bacterial pangenomes on a very large scale.}
+ODGI does not come with such sophisticated tooling, but it can serve as a strong basis to explore viral and bacterial pangenomes on a very large scale.
+Other pangenome analysis platforms, like PanTools~\citep{Sheikhizadeh_2016} provide acces to pangenome analyses at the scales we demonstrate with ODGI, but use specialized de Bruijn graph models to achieve this---in contrast ODGI supports the highly generic variation graph model, which has greater representational power than de Bruijn graphs.}
 

### 2022-03-09 manuscript/main.tex
tweaks

@@
 We wrote ODGI, a novel suite of tools that implements scalable algorithms and has an efficient in-memory representation of DNA \REVIEWED{pangenome graphs in the form of variation graphs}.
-\REVIEWED{ODGI supports pre-built graphs in the Graphical Fragment Assembly format version 1.}
-ODGI includes tools for detecting complex regions, extracting \textit{loci}, removing artifacts, exploratory analysis, manipulation, validation, and visualization.
+\REVIEWED{ODGI supports pre-built graphs in the Graphical Fragment Assembly format.}
+ODGI includes tools for detecting complex regions, extracting pangenomic loci, removing artifacts, exploratory analysis, manipulation, validation, and visualization.
 Its fast parallel execution facilitates routine pangenomic tasks, as well as pipelines that can quickly answer complex biological questions of gigabase-scale pangenome graphs. \\
@@
 \section{Introduction}
-A pangenome models the full set of genomic elements in a given species or clade~\REVIEWED{\citep{cpang2018,Eizenga_2020}}.
+A pangenome models the full set of genomic elements in a given species or clade~\REVIEWED{\citep{Tettelin_2008,cpang2018,Eizenga_2020}}.
 In contrast to reference-based approaches which relate \REVIEWED{samples} to a single genome, these data structures encode the mutual relationships between all the genomes represented\REVIEWED{~\citep{Ballouz2019}}.
@@
 Such an effort began with the VG toolkit~\citep{Garrison:2018}\REVIEWED{, but its tools do not efficiently handle pangenome graphs presenting complex motifs that result from repetitive sequences.}
-Here we refocus the effort with the Optimized Dynamic Genome/Graph Implementation (ODGI) toolkit, a compatible, but independent pangenome graph interrogation and transformation system specifically implemented to handle the data scales encountered when working \REVIEWED{[1,10]with previously constructed pangenomes comprising hundreds of haplotype-resolved genomes.[1,10]}
+Here we refocus the effort with the Optimized Dynamic Genome/Graph Implementation (ODGI) toolkit, a compatible, but independent pangenome graph interrogation and transformation system specifically implemented to handle the data scales encountered when working \REVIEWED{[1,10]with pre-built constructed pangenomes comprising hundreds of haplotype-resolved genomes.[1,10]}
 ODGI offers a set of standard operations on the variation graph data model (Fig.~\ref{fig:operations}) , generalizing ``genome arithmetic'' concepts, like those found in BEDTools~\citep{Quinlan_2010}, to work on pangenome graphs.
@@
 Such motifs can occur in collapsed structures generated by ambiguous sequence homology relationships in repeats found in the centromeres and other segmental duplications.
-If we cannot process such regions, there are only two options: 1) remove such regions, or 2) leave them unaligned.
-However, neither of these solutions allows us  their biological features.
+If we cannot process such regions, \REVIEWED{we cannot understand them}, and there are only two options: 1) remove such regions, or 2) leave them unaligned.
+Neither of these solutions allows us \REVIEWED{to evaluate} their biological \REVIEWED{or technical} features.
+\REVIEWED{Such approaches presuppose that we are unable to understand variation in these regions.}
 To seamlessly represent such difficult regions, we followed an approach implemented in the dynamic version of the Graph BWT (GBWT)~\citep{Siren:2020} and built a node-centric, dynamic, compressed model of the paths.
@@
     \textbf{lock} \tcp{atomic locking primitive}
-    \textbf{sequence} $= [$A$|$T$|$G$|$C$|$N$]*$ \\
+    \textbf{sequence} $= [$A$|$T$|$G$|$C$|$N$]+$ \\
     \tcp{bit-packed vector of edges}

### 2022-03-10 manuscript/main.tex
onward

@@
 ODGI improves on prior efforts, based on issues that arose during our work with high-quality \textit{de novo} assemblies that cover almost all parts of the human genome~\citep{Logsdon_2021,Nurk_2021}.
-In particular, we find that it is necessary to support graphs with regions of very high numbers of path traversals \REVIEWED{(high node depth)}.
+In particular, we find that it is necessary to support graphs with regions of very high numbers of path traversals \REVIEWED{(high depth of path coverage of some nodes)}.
 Such motifs can occur in collapsed structures generated by ambiguous sequence homology relationships in repeats found in the centromeres and other segmental duplications.
-If we cannot process such regions, \REVIEWED{we cannot understand them}, and there are only two options: 1) remove such regions, or 2) leave them unaligned.
-Neither of these solutions allows us \REVIEWED{to evaluate} their biological \REVIEWED{or technical} features.
-\REVIEWED{Such approaches presuppose that we are unable to understand variation in these regions.}
+If we cannot process such regions, \REVIEWED{we cannot understand them}, and \REVIEWED{our only option is to build graphs that do not include them.}
+\REVIEWED{Our goal is to build tools that allow for a wide range of uses of pangenome graphs, including cases with potentially high path depth.}
+%Neither of these solutions allows us \REVIEWED{to evaluate} their biological \REVIEWED{or technical} features.
+%\REVIEWED{This approach  presuppose that we are unable to understand variation in these regions.}
 To seamlessly represent such difficult regions, we followed an approach implemented in the dynamic version of the Graph BWT (GBWT)~\citep{Siren:2020} and built a node-centric, dynamic, compressed model of the paths.
-This design supports node-local modification and update of the graph, which lets us operate on paths in parallel.
+This design supports node-local modification and update of the graph, which lets us \REVIEWED{build and modify the graph and its paths} in parallel.
 
-We store the graph in a vector of node structures, each of which presents a node-local view of the graph sequence, topology, and path layout.
+We store the graph in a vector of node structures, each of which presents a node-local view of the graph sequence, topology, and path layout (Algorithm~\ref{alg:structs}).
 Expressed in terms of the variation graph $V$, ODGI's core $Node$ structure includes a decoder that maps the neighbors of each node to a dense range of integers.
@@
 We define the edges and path steps traversing the node in terms of this alphabet of $\delta$'s.
-The structures in Algorithm~\ref{alg:structs} describes our encoding.
+%The structures in Algorithm~\ref{alg:structs} describes our encoding.
 
@@
 ODGI's tools are practical and able to work with high levels of graph complexity, even with regions where paths present very high depth nodes (10$^5$ to 10$^6$-fold depth).
-
 ODGI covers common operations that we have found to be essential when working with complex pangenome graphs:
 
-\FIXME{Update the references to the sections.}
+%\FIXME{Update the references to the sections.}
 \begin{itemize}
@@
 Indeed, by considering all paths in the graph as potential reference or query sequence, we make graphs invisible to downstream tools that operate on collections of sequences or rely on a reference sequence (\textit{e.g.} SAMtools~\citep{Li2009}), enabling interoperability.
-This approach benefits from the information in the graph without strongly embedding our methods in this difficult new research context.
+This approach benefits from the information in the graph without \REVIEWED{requiring that we build an entirely new set of bioinformatic methods to work} in this difficult new \REVIEWED{pangenomic} research context.
 
@@
 \REVIEWED{[1,5]The process itself is lossless.
-A graph in ODGI format represents everything that is in the input GFAv1 graph, without any loss of information.[1,5]} \REVIEWED{[2,1]ODGI does not support GFAv2 nor rGFA.
-The less popular GFAv2 format was specifically designed for assembly graphs.
-rGFA requires a genomic sequence to be the reference sequence, but in GFAv1 we don't have that limitation which is fundamental to implement reference-free approaches.[2,1]}
+A graph in ODGI format represents everything that is in the input GFAv1 graph, without any loss of information.[1,5]} \REVIEWED{[2,1]ODGI does not natively support GFAv2 or rGFA.
+GFAv2 is similar to GFAv1, but includes process-related annotations of assembly graphs not relevant for pangenome analyses.
+rGFA embeds a single coordinate hierarchy over the graph that links all sequences into a single base reference genome.
+This positional model depends on a particular graph induction algorithm \cite{Li:2020}.
+In contrast, ODGI implements coordinate translation dynamically (e.g. \textit{odgi position} and \textit{odgi untangle}), allowing use of any embedded genome as a reference.
+Its input graphs can represent any kind of alignment between the genomes.
+GFAv1 is fully capable of representing many reference genome coordinate systems simultaneously, which supports a reference-agnostic approach that uses the entire pangenome sequence space as a reference system.
+In doing so, our approach has the advantage of maintaining backward compatibility with existing tools based on genome sequences.
+ODGI's methods help us to use the pangenome graph model as a coordinating system for analyses that span the entire sequence space of the pangenome.[2,1]}
 
@@
 To support interchange with other pangenome tools or text-based processing, \textit{odgi view} converts a graph in ODGI binary format to GFAv1.
-
-As the GFA format itself does not store metadata information, ODGI follows the PanSN-spec~\citep{pansn-spec} convention for dealing with heterozygosity and polyploidy in the sequence name.
-This harmonizes the biosample information present in FASTA, VCF, BED, and GFF/GTF formats.
+\REVIEWED{ODGI utilizes the PanSN~\citep{pansn-spec} specification to embed sample and haplotype information in the sequence name.
+This harmonizes the biosample information present in FASTA, GFA, PAF, VCF, BED, and GFF/GTF formats related to the graph and its embedded genome sequences.
+%Although this convention may surprise
+%we believe that there is vastly greater value to harmonizing our work with existing systems than 
+By embedding all sequences into a single hierarchical namespace related to fundamental biological groupings in the input (e.g.
+biosample, individual, pooled group), PanSN allows us to utilize all assemblies in the pangenome as a combined reference coordinate model.
+%This pattern is directly backward\cite{Garrison:2018}.
+%Just as rGFA introduces a coordinate hierarchy, PanSN applies a naming hierarchy to all sequences in the pangenome.
+%Although many This simple practice is sufficient for all lossless variation graph models.
+}
 %ODGI uses the ``S'' and ``L'' records to describe the graph topology, and ``P'' records to describe paths.

### 2022-03-14 manuscript/main.tex
defusing the bomb

@@
 
-\REVIEWED{Understanding pangenome graphs is not trivial. In the following, we highlight how ODGI's tools can assist in exploring the biological features of any pangenome graph.}
+\REVIEWED{%Understanding pangenome graphs is not trivial.
+Here, we apply our methods to a series of analyses, highlighting how ODGI can assist in exploring the biological features of pangenome graphs.
+We follow typical analyses that we have found critical to interpreting whole genome alignments represented in the variation graph model.}
 
-\REVIEWED{Pangenome graphs built from hundreds of haplotype-resolved \textit{de novo} genome assemblies are very large, but it is often necessary to work with only a small portion of the genomes represented,
-such as a specific \textit{locus} (Fig.~\ref{fig:viz_extract_untangle-MHC}) or a smaller region (Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage,fig:viz_extract_untangle-C4-bandage,fig:viz_extract_untangle-C4-bandage-zoom}), or even a single gene (Fig.~\ref{fig:metrics}).
+\REVIEWED{To simplify our exposition, we will extract small graph regions that are easy to interpret and describe.
+Pangenome graphs built from hundreds of haplotype-resolved \textit{de novo} genome assemblies are very large, but it is often necessary to work with only a small portion of the genomes represented,
+such as a specific locus (Fig.~\ref{fig:viz_extract_untangle-MHC}) or a smaller region (Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage,fig:viz_extract_untangle-C4-bandage,fig:viz_extract_untangle-C4-bandage-zoom}), or even a single gene (Fig.~\ref{fig:metrics}).
 This simplifies the downstream analyses and reduces the resources to work only with the extracted graphs. More on graph extraction and edit operations can be found at $\S$\ref{sec:supp_edit}.}
@@
 
-\REVIEWED{Pangenome graph visualization is one of the first steps to gain insight into the mutual relationship between the sequences in the graph and their variation, understanding the higher-level biology of a graph.}
+\REVIEWED{Visualization methods help us quickly gain insight into otherwise opaque biological data.
+We find visualization essential to understanding pangenome variation graphs.}
 We pursue a novel approach to visualization with \textit{odgi draw} and \textit{odgi viz}, two tools that provide scalable ways of generating \REVIEWED{static} pictures of the high-level structure of even large pangenome graphs (Fig.~\ref{fig:odgi_viz}).
@@
 MHC genes are highly polymorphic, i.e., there are multiple different alleles across individuals in a population.
-Such variability becomes evident when we apply \textit{odgi draw} to visualize the graph layout of a human MHC pangenome graph (Fig.~\ref{fig:viz_extract_untangle-MHC}).
-The visualization displays the graph topology in 2 dimensions (2D) and how the paths traverse the graph, with the regions of variations that appear as bubbles in the layout.
-The layout was built by \textit{odgi layout} (see \S\ref{sec:sort}).} Of note, a 2D rendering can be costly to compute, but we provide an implementation that scales linearly with pangenome sequence size, allowing us to apply it to large pangenome graphs.
+Such variability becomes evident when we apply \textit{odgi draw} to visualize the graph layout of a human MHC pangenome graph (Fig.~\ref{fig:viz_extract_untangle-MHC}), using \textit{odgi layout} to generate the 2D projection (see \S\ref{sec:sort}).
+The visualization displays the graph topology in 2 dimensions (2D), with structural variation that appears as bubbles in the layout.}
+A 2D rendering can be costly to compute, but we provide an implementation that scales linearly with pangenome sequence size, allowing us to apply it to large pangenome graphs.
 %\REVIEWED{For example, \textit{odgi draw} took 6 minutes and 21 seconds in total to produce the 2D layout of the MHC locus (Fig.~\ref{fig:odgi_layout}).}
@@
 
-\REVIEWED{To obtain a more precise overview of the \textit{locus} in Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage},
+\REVIEWED{To obtain a more precise overview of the locus in Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage},
 we applied \textit{odgi untangle} with GRCH38 as a reference. \textit{odgi untangle} segments paths into linear segments by breaking these segments where the paths loop back on themselves.
-In this way, we obtain information on the position and copy number status of the sequences in the collapsed \textit{locus} (Fig.~\ref{fig:viz_extract_untangle-C4-untangle}).
+In this way, we obtain information on the position and copy number status of the sequences in the collapsed locus (Fig.~\ref{fig:viz_extract_untangle-C4-untangle}).
 The shape of the line indicates if the copy number is in forward or in reverse orientation compared to GRCH38.
@@
 ODGI does not come with such sophisticated tooling, but it can serve as a strong basis to explore viral and bacterial pangenomes on a very large scale.
-Other pangenome analysis platforms, like PanTools~\citep{Sheikhizadeh_2016} provide acces to pangenome analyses at the scales we demonstrate with ODGI, but use specialized de Bruijn graph models to achieve this---in contrast ODGI supports the highly generic variation graph model, which has greater representational power than de Bruijn graphs.}
+Other pangenome analysis platforms, like PanTools~\citep{Sheikhizadeh_2016} provide access to pangenome analyses at the scales we demonstrate with ODGI, but use specialized de Bruijn graph models to achieve this---in contrast ODGI supports the highly generic variation graph model, which has greater representational power than de Bruijn graphs.}
 

### 2022-03-14 manuscript/main.tex
slightly more defused

@@
 
-\REVIEWED{Using \textit{odgi extract}, we extracted the the major histocompatibility complex (MHC) locus from a 90 haplotypes human pangenome chromosome 6 graph form the HPRC.
+\REVIEWED{Using \textit{odgi extract}, we extracted the the major histocompatibility complex (MHC) locus from a 90-haplotype human pangenome chromosome 6 graph form the HPRC.
 Specifically the graph contains the human references GRCH38, CHM13, and the contigs of 44 individuals that encode all possible variation including telomeres and centromeres.
@@
 MHC genes are highly polymorphic, i.e., there are multiple different alleles across individuals in a population.
-Such variability becomes evident when we apply \textit{odgi draw} to visualize the graph layout of a human MHC pangenome graph (Fig.~\ref{fig:viz_extract_untangle-MHC}), using \textit{odgi layout} to generate the 2D projection (see \S\ref{sec:sort}).
+Such variability becomes evident when we apply \textit{odgi draw} to visualize the graph layout of a human MHC pangenome graph (Fig.~\ref{fig:viz_extract_untangle-MHC}) (\textit{odgi layout} first generates the drawn projection, see \S\ref{sec:sort}).
 The visualization displays the graph topology in 2 dimensions (2D), with structural variation that appears as bubbles in the layout.}
@@
 \REVIEWED{The MHC locus includes the complement component 4 (C4) region, which encodes proteins involved in the complement system.
-In Fig.~\ref{fig:viz_extract_untangle-MHC}, such a region corresponds to the little bubble highlighted by the red rectangle.
-As an example use case, we took a closer look, at the C4 region of the MHC by extracting it from the full MHC pangenome graph with \textit{odgi extract}.
-Then, we visualized such a subgraph by applying} \textit{odgi viz}, which produces binned, linearized renderings in 1 dimension (1D) (i.e., \REVIEWED{each path lies} on a single axis) (Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage}).
+In Fig.~\ref{fig:viz_extract_untangle-MHC}, C4 corresponds to the small bubble highlighted by the red rectangle.
+As an example use case, we took a closer look at the C4 region of the MHC by extracting it from the full MHC pangenome graph with \textit{odgi extract}.
+Then, we visualized this subgraph by applying} \textit{odgi viz}, which produces binned, linearized renderings in 1 dimension (1D) (\REVIEWED{the graph is ordered in 1D across the horizontal axis, with each path represented by a row of the vertical axis}) (Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage}).
 For each path, graph nodes are arranged from left to right, with the colored bars indicating the paths and the nodes they cross.
@@
 All evaluations were performed on a VM in the German Network for Bioinformatics Infrastructure (deNBI) cloud. It came with 28 cores and 256GB of RAM.
-The presented results are from a 90 haplotypes chromosome 6 human pangenome graph from the HPRC.
+The presented results are from a 90-haplotype chromosome 6 human pangenome graph from the HPRC.
 Specifically, the graph contains the human references GRCH38, CHM13, and the contigs of 44 individuals encoding all possible variation including telomeres and centromeres.

### 2022-03-14 manuscript/main.tex
more defused

@@
 \REVIEWED{To simplify our exposition, we will extract small graph regions that are easy to interpret and describe.
-Pangenome graphs built from hundreds of haplotype-resolved \textit{de novo} genome assemblies are very large, but it is often necessary to work with only a small portion of the genomes represented,
+We focus on a handful of difficult loci from the human pangenome, extracting them from a prototype human pangenome graph built with the Pangenome Graph Builder pipeline~\citep{pggb}.
+Pangenome graphs built from hundreds of haplotype-resolved \textit{de novo} genome assemblies are very large, but it is often only necessary to work with only a small portion of the genomes represented,
 such as a specific locus (Fig.~\ref{fig:viz_extract_untangle-MHC}) or a smaller region (Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage,fig:viz_extract_untangle-C4-bandage,fig:viz_extract_untangle-C4-bandage-zoom}), or even a single gene (Fig.~\ref{fig:metrics}).
@@
 We find visualization essential to understanding pangenome variation graphs.}
-We pursue a novel approach to visualization with \textit{odgi draw} and \textit{odgi viz}, two tools that provide scalable ways of generating \REVIEWED{static} pictures of the high-level structure of even large pangenome graphs (Fig.~\ref{fig:odgi_viz}).
+We pursue a novel approach to visualization with \textit{odgi draw} and \textit{odgi viz}, two tools that provide scalable ways of generating \REVIEWED{raster images showing} the high-level structure of even large pangenome graphs (Fig.~\ref{fig:odgi_viz}).
 
@@
 \REVIEWED{In humans, the C4 gene exists as 2 functionally distinct genes, \textit{C4A} and \textit{C4B}, which both vary in structure and copy number~\citep{Sekar_2016}.
-Indeed, the longer link at the bottom of Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage} indicates that the copy number status of these genes varies across the haplotypes represented in the pangenome.
-Moreover, the short nested link on the left highlights that \textit{C4A} and \textit{C4B} genes segregate in both long and short genomic forms, distinguished by the presence or absence of a human endogenous retroviral (HERV) sequence.}
+In combination with the observed changes in path self-coverage, which represents copy number of a given path relative to the graph (Fig.~\labelcref{fig:viz_extract_untangle-C4-coverage}), the longer link at the bottom of Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage} indicates that the copy number status of these genes varies across the haplotypes represented in the pangenome.
+Moreover, the short nested variation on the left of the locus highlights that \textit{C4A} and \textit{C4B} genes segregate in both long and short genomic forms, distinguished by the presence or absence of a human endogenous retroviral (HERV) sequence.}
 
 Nevertheless, complex, nonlinear graph structures are difficult to interpret in a low number of dimensions.
-To overcome such a limitation, \textit{odgi viz} supports multiple visualization modalities (Fig.~\labelcref{fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage}), making it easy to grasp the properties and shape of the graph.
+To overcome this limitation, \textit{odgi viz} supports multiple visualization modalities (Fig.~\labelcref{fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage}), making it easy to grasp the properties and shape of the graph.
 %Graph node order can affect downstream analyses on pangenome graphs.
@@
 %Eukaryotic genomes experience gains and losses of genetic material, resulting in copy number variation (CNV) across the population.
-Furthermore, we can use multiple color palettes to color the paths \REVIEWED{by how many times they traverse a node, the so-called node depth.}
+Furthermore, we can use multiple color palettes to color the paths \REVIEWED{by how many times they traverse a node, which can be referred to as the path's ``coverage'' of the node.}
 \REVIEWED{This highlights that in the C4 pangenome graph, the haplotypes present different number of copies of the C4 genes (Fig.~\ref{fig:viz_extract_untangle-C4-coverage}).}
@@
 In this way, we obtain information on the position and copy number status of the sequences in the collapsed locus (Fig.~\ref{fig:viz_extract_untangle-C4-untangle}).
-The shape of the line indicates if the copy number is in forward or in reverse orientation compared to GRCH38.
+The orientation of the line indicates if the copy number is in forward or in reverse orientation compared to GRCH38.
 \textit{odgi untangle} demonstrates that it is able to work with any sets of reference sequences converting the graph to lift-over maps compatible with standard software for projecting annotations and alignments from one genome to another.
@@
 to understand the sparse structures typically found in pangenome graphs and the genetic variation they represent.
+\REVIEWED{\textit{odgi sort} allows for the chaining of these sorting algorithms.
+As many of the algorithms are affected by the initial node order, this allows us to generate sorting pipelines that progressively refine the graph ordering.}
 
-\REVIEWED{We applied a variety of \textit{odgi sort}'s 1D algorithms to a 90-haplotype human pangenome MHC and a C4 graph (Fig.~\ref{fig:sorting}).
+\REVIEWED{We applied several of \textit{odgi sort}'s 1D algorithms to a 90-haplotype human pangenome MHC and a C4 subgraph (Fig.~\ref{fig:sorting}).
 The randomly sorted MHC graph (Fig.~\ref{fig:mhc-random-node-order}) hides its linear graph structure, whereas our novel path-guided (PG) stochastic gradient descent (SGD) algorithm, PG-SGD, is able to produce a globally linear ordered graph revealing the C4 region (Fig.~\ref{fig:mhc-pgsdg-sorting}).
@@
 To scale to large graphs, we learn this projection in parallel via a HOGWILD! approach~\citep{niu2011hogwild}.
-Our approach can be seen as an adaptation of SGD-based drawing to pangenome graphs~\citep{zheng2018graph}.
-In parallel, each HOGWILD! thread updates node relative positions to best-match their nucleotide distance in the paths running through the graph.
+PG-SGD can be seen as an adaptation of SGD-based drawing~\citep{zheng2018graph} to pangenome graphs.
+In parallel, each HOGWILD!
+thread updates the relative position of pairs of nodes so that their distance in the layout or order better-matches  their nucleotide distance in the paths running through the graph.
 Following standard SGD approaches, a learning rate is reduced as the algorithm progresses, and execution continues until the adjustments to the model fall below a target threshold $\epsilon$.}
 
-\REVIEWED{A PG-SGD sorting of the C4 does not work well locally.
-It hides the underlying copy number variation status in the region (Fig.~\ref{fig:c4-pgsdg-sorting}).
-A topological sort order finally gives rise to the graph's latent structure, revealing the copy number characteristics of this C4 region (Fig.~\ref{fig:c4-topological-sorting}).
-}
+\REVIEWED{A PG-SGD sort of C4 compresses both sides of the variant bubble into one dimension, leading to an interrupted pattern of nodes across the copy-number variable region (Fig.~\ref{fig:c4-pgsdg-sorting}).
+Subsequently applying a topological sort clarifies the graph's latent structure, simplifying interpretation (Fig.~\ref{fig:c4-topological-sorting}). }
 %
@@
 Zooming in on this region with \textit{vg viz}, we can clearly identify the typical \textit{HTTexon1} CAG variable repeat region (Fig.~\ref{fig:metrics-str}).
-Figures~\ref{fig:metrics-degree}-\ref{fig:metrics-viz} highlight the variant region around position 200 of CHM13 showing the variable number of glutamine residues of the different individuals as reported by \citep{Nance1999}.
-In general, such complex motives can be the mirror of genetic variation, but also misassemblies or problems in the pangenome building, making the tools further useful for graph validation.
+Figures~\ref{fig:metrics-degree}-\ref{fig:metrics-viz} highlight the variant region around position 200 of CHM13 showing the variable number of glutamine residues of the different individuals as reported by \citep{Nance1999}.}
+%In general, such complex motives can be the mirror of genetic variation, but also misassemblies or problems in the pangenome building, making the tools further useful for graph validation.
 Additional graph metrics interrogation tools are detailed in $\S$\ref{sec:supp_metrics}.}
@@
 
-\subsection{Evaluation}
+\subsection{Performance evaluation}
 \label{Evaluation}
 
-\REVIEWED{We compared the efficiency of ODGI (v0.6.3-56-gebc493f "Pulizia") and VG (v1.37.0 "Monchio") for routine pangenome tasks. In particular, we measured the execution time and memory usage}
+\REVIEWED{
+%Although many of the operations that ODGI provides are unique, in some we are able to build on a similar system, VG, and here we demonstrate the advantage of 
+We compared the efficiency of ODGI (v0.6.3-56-gebc493f "Pulizia") and VG (v1.37.0 "Monchio") for routine pangenome tasks. In particular, we measured the execution time and memory usage}
 \begin{inparaenum}[(i)]
@@
 \REVIEWED{These graph operations are key when it comes to the understanding of pangenome graphs.
-We ran these operations for a varying number of threads and haplotypes in the graph for a scaling analysis.
-We ran each evaluation configuration 10 times and report the mean of each run.
-All evaluations were performed on a VM in the German Network for Bioinformatics Infrastructure (deNBI) cloud. It came with 28 cores and 256GB of RAM.
-The presented results are from a 90-haplotype chromosome 6 human pangenome graph from the HPRC.
-Specifically, the graph contains the human references GRCH38, CHM13, and the contigs of 44 individuals encoding all possible variation including telomeres and centromeres.
-}
+They are also a set of functions implemented in both toolkits.
+We ran these operations for a varying number of threads and haplotypes in the graph for a scaling analysis. We ran each evaluation configuration 10 times and report the mean of each run.
+All evaluations were performed on a VM in the German Network for Bioinformatics Infrastructure (deNBI) cloud with 28 cores and 256GB of RAM.
+The presented results are from a 90-haplotype chromosome 6 human pangenome graph built with data from the HPRC.
+Specifically, the graph contains the human references GRCH38, CHM13, and the contigs of 44 individuals encoding all possible variation including telomeres and centromeres. }
 
-\REVIEWED{When transforming a GFAv1 with VG, the static XG file format was used. All tools involved in the evaluation process require the XG format.}
+\REVIEWED{When transforming a GFAv1 with VG, the static XG file format was used. The tools involved in the evaluation process require the XG format.}
 

### 2022-03-14 manuscript/main.tex
defused

@@
 \REVIEWED{
-%Although many of the operations that ODGI provides are unique, in some we are able to build on a similar system, VG, and here we demonstrate the advantage of 
-We compared the efficiency of ODGI (v0.6.3-56-gebc493f "Pulizia") and VG (v1.37.0 "Monchio") for routine pangenome tasks. In particular, we measured the execution time and memory usage}
+Although many of the operations that ODGI provides are unique, some are common with the existing VG toolkit.
+We compare with these to highlight the practical performance implications of our graph data structure design.
+Our results highlight the efficient parallel algorithm implementations enabled by this design.}
+
+\REVIEWED{We compared the efficiency of ODGI (v0.6.3-56-gebc493f "Pulizia") and VG (v1.37.0 "Monchio") for routine pangenome tasks. In particular, we measured the execution time and memory usage}
 \begin{inparaenum}[(i)]
@@
 The presented results are from a 90-haplotype chromosome 6 human pangenome graph built with data from the HPRC.
-Specifically, the graph contains the human references GRCH38, CHM13, and the contigs of 44 individuals encoding all possible variation including telomeres and centromeres. }
+Specifically, the graph contains the human references GRCH38, CHM13, and the contigs of 44 individuals encoding all possible variation including telomeres and centromeres.
+When transforming a GFAv1 with VG, the static XG file format was used.
+The tools involved in the evaluation process require the XG format.}
 
-\REVIEWED{When transforming a GFAv1 with VG, the static XG file format was used. The tools involved in the evaluation process require the XG format.}
+\REVIEWED{In general, we find that ODGI makes comparatively better use of multithreading and requires much less memory (Fig.~\ref{fig:evaluation}, Tab.~\ref{tab:position}) across all operations.
+ODGI scales much butter than VG when working with complex regions of the graph.
+For example, extracting a difficult centromeric subgraph (Fig.~\ref{fig:eval-extract}), ODGI is up to 40 times faster and requires 8 times less memory than VG.}
 
-\REVIEWED{Our general findings are that ODGI makes comparatively better use of multithreading and requires much less memory (Fig.~\ref{fig:evaluation}, Tab.~\ref{tab:position}) across all operations.
-With raising number of haplotypes in the graph, ODGI scales much butter than VG.
-For example, extracting a subgraph (Fig.~\ref{fig:eval-extract}), ODGI is up to 40 times faster and requires 8 times less memory than VG.
-}
 \REVIEWED{Both visualization tools can only make use of a single thread.
@@
 
-\REVIEWED{We also measured the disk space usage of GFAv1, ODGI's, and VG's binary format (Tab.
-\ref{tab:disk_space}).
+\REVIEWED{We also measured the disk space usage of GFAv1, ODGI's, and VG's binary format (Tab. \ref{tab:disk_space}).
 While VG's XG occupies less disk space for smaller graphs, ODGI requires less space for graphs having 32 haplotypes or more.
-With increasing number of haplotypes the centromeric sequence in the pangenome graph also increases which can be better compressed by ODGI's id delta encoding schema.
-}
+We hypothesize that this represents lower marginal cost for additional haplotypes when using ODGI's id delta encoding scheme. }
+%We note that the With increasing number of haplotypes the centromeric sequence in the pangenome graph also increases which can be better compressed by
 
@@
 
-ODGI is a state-of-the-art tool suite that enables users to explore and discover the underlying biology in pangenomes graphs, filling the gap that made pangenomic analyses difficult.
+\REVIEWED{By providing a set of standard analysis ``verbs'' to interact with pangenome graphs, ODGI enables users to explore and discover important biological features captured in this flexible, inclusive model.}
 It provides tools to easily transform, analyze, simplify, validate, and visualize pangenome graphs at large scale.
@@
 With the increased adoption of long read sequencing we expect pangenomic tools to become increasingly common in \REVIEWED{the genomic studies at different taxonomic levels and} in biomedical research.
-Particularly for targets that involve complex variation, such as cancer\REVIEWED{~\citep{CompPan2016}}, plant pangenomics\REVIEWED{~\citep{Bayer2020, Liu2020, Qin2021, Li2022, Bayer2022}}, and metagenomics\REVIEWED{~\citep{Zhong2021}}.
+This progression is already afoot, particularly for targets that involve complex variation, such as cancer\REVIEWED{~\citep{CompPan2016}}, plant pangenomics\REVIEWED{~\citep{Bayer2020, Liu2020, Qin2021, Li2022, Bayer2022}}, and metagenomics\REVIEWED{~\citep{Zhong2021}}.
 \REVIEWED{Also, when studying animals like bovines~\citep{Leonard2021, Talenti2022, BPC}.}
@@
 But, it already has a very developed eco-system, which allows a detailed analysis of bacterial pangenomes using an interactive GUI.
-ODGI does not come with such sophisticated tooling, but it can serve as a strong basis to explore viral and bacterial pangenomes on a very large scale.
+Unlike these approaches, which provide a monolithic, integrated solution to understanding pangenomes, ODGI is designed as a low-level toolkit that can work on a generic pangenome graph model frequently used by other existing methods.
+We hope that this design renders it useful to pangenome analysis pipeline authors.
+%but it can serve as a strong basis to explore viral and bacterial pangenomes on a very large scale.
 Other pangenome analysis platforms, like PanTools~\citep{Sheikhizadeh_2016} provide access to pangenome analyses at the scales we demonstrate with ODGI, but use specialized de Bruijn graph models to achieve this---in contrast ODGI supports the highly generic variation graph model, which has greater representational power than de Bruijn graphs.}
@@
 
-\REVIEWED{The performance analysis shows that ODGI clearly outperforms VG when handling large, complex pangenome graphs.
-This is a major achievement, because VG's XG data structure is static, compared to ODGI's dynamic graph implementation.
-Across the evaluation of key graph operations, ODGI's memory peak was 10GB.
-This makes it perfectly suited to be run interactively on a recent laptop.
-VG has limitations here.
+\REVIEWED{The performance analysis shows that ODGI outperforms VG when handling large, complex pangenome graphs.
+%This is a major achievement, because VG's XG data structure is static, compared to ODGI's dynamic graph implementation.
+Across the evaluation of key graph operations, ODGI's memory peak was 10GB. This makes it perfectly suited to be run interactively on a recent laptop.
+%VG has limitations here.
 We expect that ODGI will be able to handle the next phase of the HPRC, constructing a pangenome graph from 300 individuals, without any problems.}
 
-\REVIEWED{[1,10]While ODGI does not construct graphs from scratch nor is is capable of extending them, ODGI[1,10]} is already the backbone of the Pangenome Graph Builder pipeline~\citep{pggb}.
+\REVIEWED{[1,10]While ODGI does not construct graphs from scratch nor is is capable of extending them, it[1,10]} is already the backbone of the Pangenome Graph Builder pipeline~\citep{pggb}.
 Its static, large-scale 1D and 2D visualizations of the pangenome graphs allow an unprecedented high-level perspective on variation in pangenomes, and have also been critical in the development of pangenome graph building methods.
@@
 \label{sec:supp_visualization}
-\REVIEWED{In a \textit{odgi viz} visualization, given a path X traversing two nodes A and B, the corresponding edge is represented by a black line starting from the left or right of node A if this node is traversed in reverse or forward, respectively, by path X;
+\REVIEWED{In an \textit{odgi viz} visualization, given a path X traversing two nodes A and B, the corresponding edge is represented by a black line starting from the left or right of node A if this node is traversed in reverse or forward, respectively, by path X;
 the black line ends to the left or right of node B if this node is traversed in forward or reverse, respectively, by path X.

### 2022-03-15 manuscript/main.tex
ODGI does not construct a graph
refer to guix file in the odgi repo

@@
 Source code can be downloaded from \url{https://github.com/pangenome/odgi} and documentation is available at \url{https://odgi.readthedocs.io}.
-ODGI can be installed via Bioconda \url{https://bioconda.github.io/recipes/odgi/README.html} or GNU Guix \url{https://github.com/ekg/guix-genomics/blob/master/odgi.scm}. \\
+ODGI can be installed via Bioconda \url{https://bioconda.github.io/recipes/odgi/README.html} or GNU Guix \url{https://github.com/pangenome/odgi/blob/master/guix.scm}. \\
 \textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
@@
 %VG has limitations here.
-We expect that ODGI will be able to handle the next phase of the HPRC, constructing a pangenome graph from 300 individuals, without any problems.}
+We expect that ODGI will be able to handle the next phase of the HPRC, a pangenome graph constructed from 300 individuals, without any problems.}
 

### 2022-03-15 manuscript/fig_extract_viz_draw_position_untangle.tex
italianization

--- 2022-03-01 manuscript/fig_extract_viz_draw_position_untangle.tex
@@
         The red rectangle highlights the C4 region.
-        \textbf{(b-e)} \textit{odgi viz} visualizations of the C4 pangenome graph, where 8 paths are displayed: 2 reference genomes (chm13 and grch38 on the top) and 6 haplotypes of 3 diploid individuals.
+        \textbf{(b-e)} \textit{odgi viz} visualizations of the C4 pangenome graph, where 8 paths are displayed: 2 reference genomes (\REVIEWED{CHM13} and \REVIEWED{GRCh38} on the top) and 6 haplotypes of 3 diploid individuals.
         \textbf{(b)} \textit{odgi viz} default modality: the image shows a quite linear graph.
@@
         \textbf{(c)} Color by path position. %: the color gradients are smooth, highlighting that its node are well sorted in 1 dimension.
-        The top two reference genomes and 2 haplotypes (HG01952\#2) go from left to right, while 5 haplotypes go in the opposite direction, as indicated by the black color on their left.
+        The top two reference genomes and one haplotypes (HG01952\#2) go from left to right, while 5 haplotypes go in the opposite direction, as indicated by the black color on their left.
         \textbf{(d)} \textit{odgi viz} color by strandness: the red paths indicate the haplotypes that were assembled in reverse with respect to the 2 reference genomes.
@@
 Node labels were annoted by using \textit{odgi position}.
-        \textbf{(h)} Visualization of \textit{odgi untangle} output in the C4 pangenome graph: the plots show the copy number status of the sequences in the C4 region with respect to the grch38 reference sequence, making clear, for example, that in HG00438\#2, the \textit{C4A} gene is missing (white in the middle of the red).
+        \textbf{(h)} Visualization of \textit{odgi untangle} output in the C4 pangenome graph: the plots show the copy number status of the sequences in the C4 region with respect to the \REVIEWED{GRCh38} reference sequence, making clear, for example, that in HG00438\#2, the \textit{C4A} gene is missing (no black lines in the region annotated in red).
 \vspace{-1em}

### 2022-03-15 manuscript/fig_tips.tex
italianization

@@
 The high peak at the beginning of \textit{p23.1} indicates that 39 contigs of the graph broke at this specific position relative to CHM13.
-(\textbf{b}) \textit{odgi viz} color by path depth visualization of the 90 longest paths in the beta-defensin locus pangenome graph including CHM13 and GRCH38 on the top: using the Spectra color palette with 4 levels of path depths.
+(\textbf{b}) \textit{odgi viz} color by path depth visualization of the 90 longest paths in the beta-defensin locus pangenome graph including CHM13 and GRCh38 on the top: using the Spectra color palette with 4 levels of path depths.
 White, grey, red, and yellow indicate a path depth of 0, 1, 2, and greater than or equal 3, respectively.

### 2022-03-15 manuscript/main.tex
italianization

@@
 \\
-\REVIEWED{\textbf{Keywords:} Navigating Variation Graphs. Pangenomic Liftover. Gigabase-Scale Pangenome Visualization. T2T Pangenome Graph Exploration.}
+\REVIEWED{\textbf{Keywords:} Navigating Variation Graphs. Pangenomic Liftover. Gigabase-Scale Pangenome Visualization.}
 }
@@
 In contrast to reference-based approaches which relate \REVIEWED{samples} to a single genome, these data structures encode the mutual relationships between all the genomes represented\REVIEWED{~\citep{Ballouz2019}}.
-\REVIEWED{[2,5] A class of methods to represent pangenomes involves sequence graphs~\citep{Hein1989, Paten:2017} where homologous regions between genomes are compressed into a single representation of all alleles present in the pangenome.
-In sequence graphs, nodes' labels are \REVIEWED{[1,2] genomic sequences [1,2]} with edges connecting those nodes.
-A bidirected sequence graph can \REVIEWED{[1,2] represent both strands of DNA [1,2]}.
-On this model, variation graphs add the concept of paths representing linear \REVIEWED{[1,2] DNA [1,2]} sequences as traversals through the nodes of the graph~\citep{Garrison:2018}.
-\REVIEWED{[1,2] For example, a [1,2]} path can be a genome, haplotype, contig, or read.
-[2,5]}
+\REVIEWED{A class of methods to represent pangenomes involves sequence graphs~\citep{Hein1989, Paten:2017} where homologous regions between genomes are compressed into single representations of all alleles present in the pangenome.
+In sequence graphs, node labels are \REVIEWED{genomic sequences} with edges connecting those nodes.
+A bidirected sequence graph can \REVIEWED{represent both strands of DNA}.
+On this model, variation graphs add the concept of paths representing linear \REVIEWED{DNA} sequences as traversals through the nodes of the graph~\citep{Garrison:2018}.
+\REVIEWED{For example, a} path can be a genome, haplotype, contig, or read.}
 
@@
 Pangenome graphs let us encode any kind of variation, allowing the generation of comprehensive data systems that builds the basis for the analyses of genome evolution.
-The Human Pangenome Reference Consortium (HPRC) and Telomere-to-Telomere (T2T) consortium~\citep{Miga:2020, Logsdon_2021, Nurk_2021} have recently demonstrated that high-quality \textit{de novo} assemblies can be routinely generated from third-generation long read sequencing data.
+The Human Pangenome Reference Consortium (HPRC) and Telomere-to-Telomere (T2T) consortium~\citep{Miga:2020, Logsdon_2021, Nurk_2021, Jarvis2022} have recently demonstrated that high-quality \REVIEWED{haploid and diploid} \textit{de novo} assemblies can be routinely generated from third-generation long read sequencing data.
 We anticipate that \textit{de novo} assemblies of similar quality will become common, leading to demand for methods to \REVIEWED{analyze} pangenomes.
@@
 Such an effort began with the VG toolkit~\citep{Garrison:2018}\REVIEWED{, but its tools do not efficiently handle pangenome graphs presenting complex motifs that result from repetitive sequences.}
-Here we refocus the effort with the Optimized Dynamic Genome/Graph Implementation (ODGI) toolkit, a compatible, but independent pangenome graph interrogation and transformation system specifically implemented to handle the data scales encountered when working \REVIEWED{[1,10]with pre-built constructed pangenomes comprising hundreds of haplotype-resolved genomes.[1,10]}
+Here we refocus the effort with the Optimized Dynamic Genome/Graph Implementation (ODGI) toolkit, a compatible, but independent pangenome graph interrogation and transformation system specifically implemented to handle the data scales encountered when working \REVIEWED{with pre-built constructed pangenomes comprising hundreds of haplotype-resolved genomes.}
 ODGI offers a set of standard operations on the variation graph data model (Fig.~\ref{fig:operations}) , generalizing ``genome arithmetic'' concepts, like those found in BEDTools~\citep{Quinlan_2010}, to work on pangenome graphs.
@@
 A pangenome graph is a sequence model that encodes the mutual alignment of many genomes~\citep{Garrison_2019_thesis,Eizenga_2020}.
-In the variation graph, $V = (N, E, P)$, nodes $N = n_1\ldots n_{|N|}$ contain \REVIEWED{[1,2] genomic sequences. [1,2]}
-Each node $n_i$ has an identifier $i$ and an implicit reverse complement $\bar{n_i}$, and a node strand $s$ corresponds to one such orientations.
+In the variation graph, $V = (N, E, P)$, nodes $N = n_1\ldots n_{|N|}$ contain \REVIEWED{genomic sequences.}
+Each node $n_i$ has an identifier $i$ and an implicit reverse complement $\bar{n_i}$, and a node strand $s$ corresponds to one of such orientations.
 Edges $E = e_1\ldots e_{|E|}$ represent ordered pairs of node strands: $e_i = ( s_a, s_b )$.
@@
 \section{Implementation}
-The ODGI toolkit builds on existing approaches to efficiently store and manipulate \REVIEWED{[2,5] pangenome graphs in the form of [2,5]} variation graphs~\citep{Garrison:2018}.
+The ODGI toolkit builds on existing approaches to efficiently store and manipulate \REVIEWED{pangenome graphs in the form of} variation graphs~\citep{Garrison:2018}.
 Similar to other efficient libraries presenting the \textsc{HandleGraph} model~\citep{Eizenga_2020_BX}, the implementation of ODGI's tools rests on three key properties which hold for most pangenome graphs:
@@
 This construction step can be a significant bottleneck, in particular as the size of the path set of the graph increases.
-\REVIEWED{[1,5]The process itself is lossless.
-A graph in ODGI format represents everything that is in the input GFAv1 graph, without any loss of information.[1,5]} \REVIEWED{[2,1]ODGI does not natively support GFAv2 or rGFA.
+\REVIEWED{The process itself is lossless.
+A graph in ODGI format represents everything that is in the input GFAv1 graph, without any loss of information.} \REVIEWED{ODGI does not natively support GFAv2 or rGFA.
 GFAv2 is similar to GFAv1, but includes process-related annotations of assembly graphs not relevant for pangenome analyses.
 rGFA embeds a single coordinate hierarchy over the graph that links all sequences into a single base reference genome.
-This positional model depends on a particular graph induction algorithm \cite{Li:2020}.
+This positional model depends on a particular graph induction algorithm~\cite{Li:2020}.
 In contrast, ODGI implements coordinate translation dynamically (e.g. \textit{odgi position} and \textit{odgi untangle}), allowing use of any embedded genome as a reference.
@@
 In doing so, our approach has the advantage of maintaining backward compatibility with existing tools based on genome sequences.
-ODGI's methods help us to use the pangenome graph model as a coordinating system for analyses that span the entire sequence space of the pangenome.[2,1]}
+%ODGI's methods help us to use the pangenome graph model as a coordinating system for analyses that span the entire sequence space of the pangenome.
+}
 
@@
 Finally, we create paths in serial, and extend them in parallel by obtaining the mutex $Node.lock$ for pairs of nodes and by adding the path step in their $Node.path\_steps$ vectors.
-This parallelism speeds ODGI model construction by many-fold when testing against graphs made from assemblies produced by the HPRC (Fig.~\ref{fig:evaluation}).
+This parallelism speeds ODGI model construction by many-fold when testing against graphs made from assemblies produced by the HPRC (\S\ref{sec:evaluation}).
 
@@
 \REVIEWED{Visualization methods help us quickly gain insight into otherwise opaque biological data.
-We find visualization essential to understanding pangenome variation graphs.}
+We find visualization essential for understanding pangenome graphs.}
 We pursue a novel approach to visualization with \textit{odgi draw} and \textit{odgi viz}, two tools that provide scalable ways of generating \REVIEWED{raster images showing} the high-level structure of even large pangenome graphs (Fig.~\ref{fig:odgi_viz}).
 
-\REVIEWED{Using \textit{odgi extract}, we extracted the the major histocompatibility complex (MHC) locus from a 90-haplotype human pangenome chromosome 6 graph form the HPRC.
-Specifically the graph contains the human references GRCH38, CHM13, and the contigs of 44 individuals that encode all possible variation including telomeres and centromeres.
+\REVIEWED{Using \textit{odgi extract}, we extracted the major histocompatibility complex (MHC) locus from a 90-haplotype human chromosome 6 pangenome graph form the HPRC.
+Specifically, the graph contains the human references GRCh38, CHM13, and the contigs of 44 diploid individuals that encode all possible variations including those in telomeres and centromeres.
 The MHC genes are involved in antigen presentation, inflammation regulation, the complement system, and the innate and adaptive immune responses~\citep{Shiina2009}.
 MHC genes are highly polymorphic, i.e., there are multiple different alleles across individuals in a population.
-Such variability becomes evident when we apply \textit{odgi draw} to visualize the graph layout of a human MHC pangenome graph (Fig.~\ref{fig:viz_extract_untangle-MHC}) (\textit{odgi layout} first generates the drawn projection, see \S\ref{sec:sort}).
+Such variability becomes evident when we apply \textit{odgi draw} to visualize the graph layout of a human MHC pangenome graph (Fig.~\ref{fig:viz_extract_untangle-MHC}) (of note, \textit{odgi layout} first generates the drawn projection, see \S\ref{sec:sort}).
 The visualization displays the graph topology in 2 dimensions (2D), with structural variation that appears as bubbles in the layout.}
@@
 As an example use case, we took a closer look at the C4 region of the MHC by extracting it from the full MHC pangenome graph with \textit{odgi extract}.
-Then, we visualized this subgraph by applying} \textit{odgi viz}, which produces binned, linearized renderings in 1 dimension (1D) (\REVIEWED{the graph is ordered in 1D across the horizontal axis, with each path represented by a row of the vertical axis}) (Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage}).
+Then, we visualized this subgraph by applying} \textit{odgi viz}, which produces binned, linearized renderings in 1 dimension (1D)\REVIEWED{, where the graph is ordered in 1D across the horizontal axis, with each path represented by a row of the vertical axis} (Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage}).
 For each path, graph nodes are arranged from left to right, with the colored bars indicating the paths and the nodes they cross.
@@
 This visualization is suitable for understanding graph node order, as smooth color gradients indicate that the node order respects the linear paths' coordinate systems.
-Pangenome graphs can represent both strands \REVIEWED{[1,2] of the genomic sequences of the DNA[1,2]}.
+Pangenome graphs can represent both strands \REVIEWED{of the genomic sequences of the DNA}.
 We can display such information by coloring the paths by orientation, with paths colored where their sequence is reverse-complemented (red) or in direct orientation (black) with respect to the sequences of the graph nodes (Fig.~\ref{fig:viz_extract_untangle-C4-strandness}).
 %Eukaryotic genomes experience gains and losses of genetic material, resulting in copy number variation (CNV) across the population.
-Furthermore, we can use multiple color palettes to color the paths \REVIEWED{by how many times they traverse a node, which can be referred to as the path's ``coverage'' of the node.}
+Furthermore, we can use multiple color palettes to color the paths \REVIEWED{by how many times they traverse a node, which can be referred to as the path's depth or coverage of the node.}
 \REVIEWED{This highlights that in the C4 pangenome graph, the haplotypes present different number of copies of the C4 genes (Fig.~\ref{fig:viz_extract_untangle-C4-coverage}).}
@@
 
-The key data in a pangenome graph is a representation of the alignment (i.e., the homology relationships) between \REVIEWED{[1,2] genomic sequences [1,2]}.
+The key data in a pangenome graph is a representation of the alignment (i.e., the homology relationships) between \REVIEWED{genomic sequences}.
 Navigating and understanding the graph requires coordinate systems to link other data to the sequences represented in the graph model.
@@
 %Our for positional conversion (\textit{odgi position}), annotation liftover (\textit{odgi untangle}), breakpoint identification (\textit{odgi tips}), subgraph extraction (\textit{odgi extract}), visualization (\textit{odgi viz}), path depth (\textit{odgi depth}), and graph complexity measurement (\textit{odgi degree}) can all operate on coordinates of paths in the graph.
-\REVIEWED{As a demonstration, we took the C4 pangenome graph and added to its nodes gene annotation from GRCH38 (in GFF format file) using \textit{odgi position} ($\S$\ref{sec:supp_gff}).
-The resulting TSV contains pairs of node and color.
+\REVIEWED{As a demonstration, we took the C4 pangenome graph and added to its nodes gene annotation from GRCh38 (in GFF format file) using \textit{odgi position} ($\S$\ref{sec:supp_gff}).
+The resulting TSV contains pairs of node and colors.
 Taking the graph and the TSV into Bandage~\citep{Wick_2015}, the actual C4 genes are highlighted (Fig.~\ref{fig:viz_extract_untangle-C4-bandage}).
@@
 \REVIEWED{To obtain a more precise overview of the locus in Fig.~\labelcref{fig:viz_extract_untangle-C4-default,fig:viz_extract_untangle-C4-gradient,fig:viz_extract_untangle-C4-strandness,fig:viz_extract_untangle-C4-coverage},
-we applied \textit{odgi untangle} with GRCH38 as a reference. \textit{odgi untangle} segments paths into linear segments by breaking these segments where the paths loop back on themselves.
-In this way, we obtain information on the position and copy number status of the sequences in the collapsed locus (Fig.~\ref{fig:viz_extract_untangle-C4-untangle}).
-The orientation of the line indicates if the copy number is in forward or in reverse orientation compared to GRCH38.
-\textit{odgi untangle} demonstrates that it is able to work with any sets of reference sequences converting the graph to lift-over maps compatible with standard software for projecting annotations and alignments from one genome to another.
+we applied \textit{odgi untangle} with GRCh38 as a reference.
+\textit{odgi untangle} segments paths into linear segments by breaking these segments where the paths loop back on themselves.
+In this way, we obtain information on the position and copy number status of the sequences in the collapsed locus, in BEDPE or PAF format.
+In the representation in Fig.~\ref{fig:viz_extract_untangle-C4-untangle}, the orientation of the line indicates if the copy number is in forward or in reverse orientation compared to GRCh38.
+\textit{odgi untangle} is able to work with any sets of reference sequences, converting the graph to lift-over maps compatible with standard software for projecting annotations and alignments from one genome to another.
 An explanation of the untangling process is given in $\S$\ref{sec:supp_navigation}.}
@@
 to understand the sparse structures typically found in pangenome graphs and the genetic variation they represent.
-\REVIEWED{\textit{odgi sort} allows for the chaining of these sorting algorithms.
+\REVIEWED{\textit{odgi sort} allows the chaining of these sorting algorithms.
 As many of the algorithms are affected by the initial node order, this allows us to generate sorting pipelines that progressively refine the graph ordering.}
 
-\REVIEWED{We applied several of \textit{odgi sort}'s 1D algorithms to a 90-haplotype human pangenome MHC and a C4 subgraph (Fig.~\ref{fig:sorting}).
+\REVIEWED{We applied several of \textit{odgi sort}'s 1D algorithms to a 90-haplotype human MHC pangenome and a C4 subgraph (Fig.~\ref{fig:sorting}).
 The randomly sorted MHC graph (Fig.~\ref{fig:mhc-random-node-order}) hides its linear graph structure, whereas our novel path-guided (PG) stochastic gradient descent (SGD) algorithm, PG-SGD, is able to produce a globally linear ordered graph revealing the C4 region (Fig.~\ref{fig:mhc-pgsdg-sorting}).
@@
 In parallel, each HOGWILD!
-thread updates the relative position of pairs of nodes so that their distance in the layout or order better-matches  their nucleotide distance in the paths running through the graph.
-Following standard SGD approaches, a learning rate is reduced as the algorithm progresses, and execution continues until the adjustments to the model fall below a target threshold $\epsilon$.}
+thread updates the relative position of pairs of nodes so that their distance in the layout, or their order, better-matches  their nucleotide distance in the paths running through the graph.
+Following standard SGD approaches, the learning rate is reduced as the algorithm progresses, and execution continues until the adjustments to the model fall below a target threshold $\epsilon$.}
 
-\REVIEWED{A PG-SGD sort of C4 compresses both sides of the variant bubble into one dimension, leading to an interrupted pattern of nodes across the copy-number variable region (Fig.~\ref{fig:c4-pgsdg-sorting}).
+\REVIEWED{A PG-SGD sorting of C4 compresses both sides of the variant bubble into one dimension, leading to an interrupted pattern of nodes across the copy-number variable region (Fig.~\ref{fig:c4-pgsdg-sorting}).
 Subsequently applying a topological sort clarifies the graph's latent structure, simplifying interpretation (Fig.~\ref{fig:c4-topological-sorting}). }
@@
 
-\REVIEWED{Investigating in detail which intricate regions in the \textit{HTTexon1} graph are responsible for the genetic variation and high GC content, we took a look at the per nucleotide node degree distribution (Fig.~\ref{fig:metrics-degree}), and node depth distribution (Fig.~\ref{fig:metrics-depth}) of CHM13 using \textit{odgi depth}'s, and \textit{odgi degree}'s BED output, respectively.
+\REVIEWED{To investigate in detail which intricate regions in the \textit{HTTexon1} graph are responsible for its genetic variation and high GC content, we took a look at the per nucleotide node degree (Fig.~\ref{fig:metrics-degree}) and node depth (Fig.~\ref{fig:metrics-depth}) distributions of CHM13 by using \textit{odgi depth}'s and \textit{odgi degree}'s BED output, respectively.
 The results indicate a highly polymorphic region around position 200 in the graph.
@@
 Zooming in on this region with \textit{vg viz}, we can clearly identify the typical \textit{HTTexon1} CAG variable repeat region (Fig.~\ref{fig:metrics-str}).
-Figures~\ref{fig:metrics-degree}-\ref{fig:metrics-viz} highlight the variant region around position 200 of CHM13 showing the variable number of glutamine residues of the different individuals as reported by \citep{Nance1999}.}
+Figures~\ref{fig:metrics-degree}-\ref{fig:metrics-viz} highlight the variant region around position 200 of CHM13, showing the variable number of glutamine residues of the different individuals as reported by~\citep{Nance1999}.}
 %In general, such complex motives can be the mirror of genetic variation, but also misassemblies or problems in the pangenome building, making the tools further useful for graph validation.
@@
 \subsection{Performance evaluation}
-\label{Evaluation}
+\label{sec:evaluation}
 
@@
 \begin{inparaenum}[(i)]
-	\item \REVIEWED{of transforming a GFAv1 into a tool's native format, }
+	\item \REVIEWED{of transforming a GFAv1 file into a tool's native format, }
 	\item \REVIEWED{the extraction of a subgraph,}
@@
 The presented results are from a 90-haplotype chromosome 6 human pangenome graph built with data from the HPRC.
-Specifically, the graph contains the human references GRCH38, CHM13, and the contigs of 44 individuals encoding all possible variation including telomeres and centromeres.
-When transforming a GFAv1 with VG, the static XG file format was used.
+Specifically, the graph contains the human references GRCh38, CHM13, and the contigs of 44 diploid individuals that encode all possible variations including those in telomeres and centromeres.
+When transforming a GFAv1 file with VG, the static XG file format was used.
 The tools involved in the evaluation process require the XG format.}
@@
 \REVIEWED{Both visualization tools can only make use of a single thread.
-For a 1 haplotype graph \textit{vg viz} produces a 816MB SVG which can't be opened by any program.
-For example, GIMP2 fails with "cannot load more than 1000000 XML elements".
-For larger graphs, \textit{vg viz} runs through and produces SVGs with only the XML header.
-This makes it unusable for larger graphs.}
+For a 1 haplotype, graph \textit{vg viz} produces a 816MB SVG which can't be opened by any program.
+%For example, GIMP2 fails with "cannot load more than 1000000 XML elements".
+For larger graphs, \textit{vg viz} runs through and produces SVGs with only the XML header. This makes it unusable for large graphs.}
 
-\REVIEWED{We also measured the disk space usage of GFAv1, ODGI's, and VG's binary format (Tab. \ref{tab:disk_space}).
+\REVIEWED{We also measured the disk space usage of GFAv1, ODGI's, and VG's binary formats (Tab.~\ref{tab:disk_space}).
 While VG's XG occupies less disk space for smaller graphs, ODGI requires less space for graphs having 32 haplotypes or more.
-We hypothesize that this represents lower marginal cost for additional haplotypes when using ODGI's id delta encoding scheme. }
+We hypothesize that this indicates the lower marginal cost for additional haplotypes when using ODGI's id delta encoding scheme. }
 %We note that the With increasing number of haplotypes the centromeric sequence in the pangenome graph also increases which can be better compressed by
@@
 %but it can serve as a strong basis to explore viral and bacterial pangenomes on a very large scale.
-Other pangenome analysis platforms, like PanTools~\citep{Sheikhizadeh_2016} provide access to pangenome analyses at the scales we demonstrate with ODGI, but use specialized de Bruijn graph models to achieve this---in contrast ODGI supports the highly generic variation graph model, which has greater representational power than de Bruijn graphs.}
+Other pangenome analysis platforms, like PanTools~\citep{Sheikhizadeh_2016} provide access to pangenome analyses at the scales we demonstrate with ODGI, but use specialized de Bruijn graph models to achieve this.
+In contrast ODGI supports the highly generic variation graph model, which has greater representational power than de Bruijn graphs.}
 
@@
 
-\REVIEWED{[1,10]While ODGI does not construct graphs from scratch nor is is capable of extending them, it[1,10]} is already the backbone of the Pangenome Graph Builder pipeline~\citep{pggb}.
+\REVIEWED{While ODGI does not construct graphs from scratch nor is is capable of extending them, it} is already the backbone of the Pangenome Graph Builder pipeline~\citep{pggb}.
 Its static, large-scale 1D and 2D visualizations of the pangenome graphs allow an unprecedented high-level perspective on variation in pangenomes, and have also been critical in the development of pangenome graph building methods.
@@
 However, an interactive solution that combines the 1D and 2D layout of a graph with annotation and read mapping information across different zoom levels is still missing.
-\REVIEWED{Recent interactive pangenome graph browsers are reference-centric~\citep{Beyer2019, Yokoyama2019}, have a limited predefined coordinate system \citep{Durant2021}}, or focus primarily on 2D~\citep{Wick_2015, Gonnella2018}.
+\REVIEWED{Recent interactive pangenome graph browsers are reference-centric~\citep{Beyer2019, Yokoyama2019}, have a limited predefined coordinate system~\citep{Durant2021}}, or focus primarily on 2D representations~\citep{Wick_2015, Gonnella2018}.
 Our graph sorting and layout algorithms can provide the foundation for future tools of this type.
@@
 This resolves a major conceptual issue that has strongly guided existing algorithms to construct pangenome graphs.
-Previously, great efforts have been made to prevent the ``collapse'' of non-orthologous sequences in the graph topology itself  \citep{Li:2020}.
+Previously, great efforts have been made to prevent the ``collapse'' of non-orthologous sequences in the graph topology itself~\citep{Li:2020}.
 This has been seen as essential to making these new bioinformatic models interpretable.
-While our presentation is primarily qualitative, our work demonstrates that we can mitigate this issue by exploiting the pangenome graph not as a static reference, but as a dynamic model of the mutual alignment of many \REVIEWED{[1.2] genomic sequences [1.2]}.
+While our presentation is primarily qualitative, our work demonstrates that we can mitigate this issue by exploiting the pangenome graph not as a static reference, but as a dynamic model of the mutual alignment of many \REVIEWED{genomic sequences}.
 Because pangenome graphs can contain complete genomes, we are able to query them to polarize the information they contain in easily-interpretable and reusable pairwise formats that are widely supported in bioinformatics.
@@
 %can be linearized into feature vector or matrix form, and ODGI thus provides a natural interface for machine learning and statistical methods that operate directly on data of this shape.
-We expect that ODGI will provide a reference interface between pangenomic and genomic approaches to understanding genome variation.
+We expect that ODGI will provide a reference interface between pangenomic and genomic approaches for understanding genome variation.
 
@@
 \label{sec:supp_visualization}
-\REVIEWED{In an \textit{odgi viz} visualization, given a path X traversing two nodes A and B, the corresponding edge is represented by a black line starting from the left or right of node A if this node is traversed in reverse or forward, respectively, by path X;
+In an \textit{odgi viz} visualization, given a path X traversing two nodes A and B, the corresponding edge is represented by a black line starting from the left or right of node A if this node is traversed in reverse or forward, respectively, by path X;
 the black line ends to the left or right of node B if this node is traversed in forward or reverse, respectively, by path X.
-Consequently, if two consecutive nodes are linked both in forward, no edge is shown (it would be 0 pixels long, as it would start from the right of the first node and ends to the left of the second one).}
+Consequently, if two consecutive nodes are linked both in forward, no edge is shown (it would be 0 pixels long, as it would start from the right of the first node and ends to the left of the second one).
 
@@
 \label{sec:supp_navigation}
-Pangenome graphs \REVIEWED{can} model alignments of many genomes.
+Pangenome graphs can model alignments of many genomes.
 With \textit{odgi untangle}, users can extract pairwise alignment information between a given set of ``query'' sequences and a given set of ``target'' sequences (used as references).
 While pangenome graphs may contain looping structures that imply many-to-many alignments between the pangenome sequences, these untangled alignments map each segment of the queries to a single segment in the set of targets.
-\textit{odgi untangle} first discovers segment boundaries using standard approaches for detecting repeats in sequence graphs \citep{Pevzner_2004}.
+\textit{odgi untangle} first discovers segment boundaries using standard approaches for detecting repeats in sequence graphs~\citep{Pevzner_2004}.
 We finally ``untangle'' by finding the target segment that best match each query segment using the \textit{path jaccard} context mapping model.
@@
 while with \textit{odgi break} they can remove cycles in the graph, reducing the complexity of the graph topology.
-Furthermore, \textit{odgi groom} allows removing spurious inverting links by exploring the graph from the orientation supported by most paths\REVIEWED{;
-	the process does not remove any genetic information, but only edits how the sequences are represented in the graph.}
+Furthermore, \textit{odgi groom} allows removing spurious inverting links by exploring the graph from the orientation supported by most paths;
+the process does not remove any genetic information, but only edits how the sequences are represented in the graph.
 
@@
 This means that the first node identifier (ID) starts at 1 and the last node ID is the number of nodes.
-The graph is then updated in place: First, the node identifiers are normalized (from 1 to the number of nodes) including the adjustment of the edges.
+The graph is then updated in place. First, the node identifiers are normalized (from 1 to the number of nodes) including the adjustment of the edges.
 Second, path information, specifically each step of a path, is updated, too.
@@
 The reordering of nodes has a great influence on how the pangenome looks like (Fig.~\ref{fig:sorting}).
+
 \input{fig_sorting}
@@
 Pangenome graph topology can be derived by applying \textit{odgi matrix}, obtaining information on graph nodes connections in textual sparse matrix format.
-To investigate on the \REVIEWED{[1,2] genomic sequences [1,2]} encoded in the graph, \textit{odgi paths} allows users to calculate pairwise overlap statistics of groupings of paths and emit all path sequences in FASTA format, and it also allows the generation of a ``pangenome matrix'' that reports the copy number (presence/absence) of each path over each node.
+To investigate on the genomic sequences encoded in the graph, \textit{odgi paths} allows users to calculate pairwise overlap statistics of groupings of paths and emit all path sequences in FASTA format, and it also allows the generation of a ``pangenome matrix'' that reports the copy number (presence/absence) of each path over each node.
 \textit{odgi flatten} generates a linearization of the graph by emitting the pangenome sequence (the concatenation of all node sequences) in FASTA format, and the projection of all paths on the linearized sequence in BED format.
@@
 To study very large pangenomes, users can use \textit{odgi bin} to summarize the path information into bins of a specified size, generating a summarized view of gigabase scale graphs in TSV or JSON file formats.
-This concept found its way into the pangenome graph ontology \citep{Yokoyama2020}. %maybe we want to save this up for the discussion
+This concept found its way into the pangenome graph ontology~\citep{Yokoyama2020}. %maybe we want to save this up for the discussion
 

### 2022-03-15 manuscript/main.tex
typos

@@
 ODGI improves on prior efforts, based on issues that arose during our work with high-quality \textit{de novo} assemblies that cover almost all parts of the human genome~\citep{Logsdon_2021,Nurk_2021}.
-In particular, we find that it is necessary to support graphs with regions of very high numbers of path traversals \REVIEWED{(high depth of path coverage of some nodes)}.
+In particular, we find that it is necessary to support graphs with regions of very high numbers of path traversals \REVIEWED{(high depth of path coverage of some nodes, the so-called node depth)}.
 Such motifs can occur in collapsed structures generated by ambiguous sequence homology relationships in repeats found in the centromeres and other segmental duplications.

### 2022-03-15 manuscript/main.tex
ich habe fertig

@@
 \REVIEWED{ODGI utilizes the PanSN~\citep{pansn-spec} specification to embed sample and haplotype information in the sequence name.
-This harmonizes the biosample information present in FASTA, GFA, PAF, VCF, BED, and GFF/GTF formats related to the graph and its embedded genome sequences.
+This harmonizes the biosample information present in FASTA, GFA, PAF, VCF, BED, BEDPE, SAM/BAM, and GFF/GTF formats related to the graph and its embedded genome sequences.
 %Although this convention may surprise
@@
 
-\REVIEWED{Using \textit{odgi extract}, we extracted the major histocompatibility complex (MHC) locus from a 90-haplotype human chromosome 6 pangenome graph form the HPRC.
+\REVIEWED{Using \textit{odgi extract}, we extracted the major histocompatibility complex (MHC) locus from a 90-haplotype human chromosome 6 pangenome graph from the HPRC.
 Specifically, the graph contains the human references GRCh38, CHM13, and the contigs of 44 diploid individuals that encode all possible variations including those in telomeres and centromeres.
@@
 %Eukaryotic genomes experience gains and losses of genetic material, resulting in copy number variation (CNV) across the population.
-Furthermore, we can use multiple color palettes to color the paths \REVIEWED{by how many times they traverse a node, which can be referred to as the path's depth or coverage of the node.}
+Furthermore, we can use multiple color palettes to color the paths \REVIEWED{by how many times they traverse a node, which can be referred to as the path's depth or coverage of the node, the node depth.}
 \REVIEWED{This highlights that in the C4 pangenome graph, the haplotypes present different number of copies of the C4 genes (Fig.~\ref{fig:viz_extract_untangle-C4-coverage}).}
@@
 \begin{enumerate}
-	\item~Starting from $N_q$ and we follow the steps of one target path position a certain nucleotide distance to the left and a certain nucleotide distance to the right, also called context.
+	\item~Starting from $N_q$ we follow the steps of one target path position a certain nucleotide distance to the left and a certain nucleotide distance to the right, also called context.
 The ODGI default distance is 10kbp for each direction.

### 2022-03-15 manuscript/main.tex
rammstein

@@
 \textit{odgi untangle} first discovers segment boundaries using standard approaches for detecting repeats in sequence graphs~\citep{Pevzner_2004}.
-We finally ``untangle'' by finding the target segment that best match each query segment using the \textit{path jaccard} context mapping model.
+We finally ``untangle'' by finding the target segment that best matches each query segment using the \textit{path jaccard} context mapping model.
 %The \textit{path jaccard} concept ensures that the breaking is nucleotide-precise.
@@
 \label{sec:supp_edit}
-Graph portions can be extracted by using the paths in the graph as coordinate systems to guide the process.
+Subgraphs can be extracted by using the paths in the graph as coordinate systems to guide the process.
 For such operation, \textit{odgi extract} allows users to extract specific regions of the graph as defined by query criteria.
@@
 \label{sec:supp_sorting}
-Most subcommands in ODGI require and verify that the input graph’s node identifiers (IDs) are optimized.
+Most subcommands in ODGI require and verify that the input graph’s node identifiers (IDs) are optimized, that is compacted from 1 to N where N is the number of nodes in the graph. 
 If this assumption is violated, \textit{odgi sort} provides functionality to optimize the graph. 
 This means that the first node identifier (ID) starts at 1 and the last node ID is the number of nodes. 
-The graph is then updated in place. First, the node identifiers are normalized (from 1 to the number of nodes) including the adjustment of the edges.
-Second, path information, specifically each step of a path, is updated, too.
+All sorting operations update the graph in place with an efficient ID rewriting algorithm.
+The graph is then updated in place. 
+First, the node identifiers are normalized (from 1 to number of nodes) including the adjustment of the edges. 
+Second, path information, including both path metadata that points into the start and end steps of the path, plus each step of every path, is updated, too. 
+We point out that changing the node order does not change our coordinate systems based on paths. 
+These will now refer to a new node ordering.
 

### 2022-03-15 manuscript/main.tex
feedback Sven

@@
 \REVIEWED{As a demonstration, we took the C4 pangenome graph and added to its nodes gene annotation from GRCh38 (in GFF format file) using \textit{odgi position} ($\S$\ref{sec:supp_gff}).
-The resulting TSV contains pairs of node and colors.
+The resulting TSV contains pairs of nodes and a colors.
 Taking the graph and the TSV into Bandage~\citep{Wick_2015}, the actual C4 genes are highlighted (Fig.~\ref{fig:viz_extract_untangle-C4-bandage}).
@@
 
-\REVIEWED{In general, we find that ODGI makes comparatively better use of multithreading and requires much less memory (Fig.~\ref{fig:evaluation}, Tab.~\ref{tab:position}) across all operations.
+\REVIEWED{In general, ODGI makes comparatively better use of multithreading and requires much less memory (Fig.~\ref{fig:evaluation}, Tab.~\ref{tab:position}) across all operations.
 ODGI scales much butter than VG when working with complex regions of the graph.
@@
 \REVIEWED{Both visualization tools can only make use of a single thread.
-For a 1 haplotype, graph \textit{vg viz} produces a 816MB SVG which can't be opened by any program.
+For a 1 haplotype, graph \textit{vg viz} produces a 816MB SVG which can't be opened by the standard programs to date.
 %For example, GIMP2 fails with "cannot load more than 1000000 XML elements".

### 2022-03-18 manuscript/main.tex
a froggy color
keywords
mention odgi pav

@@
 \\
-\REVIEWED{\textbf{Keywords:} Navigating Variation Graphs. Pangenomic Liftover. Gigabase-Scale Pangenome Visualization.}
+\REVIEWED{\textbf{Keywords:} pangenomics, variation graphs, liftover, pangenome visualization, comparative genomics}
 }
@@
 \REVIEWED{As a demonstration, we took the C4 pangenome graph and added to its nodes gene annotation from GRCh38 (in GFF format file) using \textit{odgi position} ($\S$\ref{sec:supp_gff}).
-The resulting TSV contains pairs of nodes and a colors.
+The resulting TSV contains pairs of nodes and colors.
 Taking the graph and the TSV into Bandage~\citep{Wick_2015}, the actual C4 genes are highlighted (Fig.~\ref{fig:viz_extract_untangle-C4-bandage}).
@@
 \textit{odgi untangle} is able to work with any sets of reference sequences, converting the graph to lift-over maps compatible with standard software for projecting annotations and alignments from one genome to another.
-An explanation of the untangling process is given in $\S$\ref{sec:supp_navigation}.}
+An explanation of the untangling process is given in $\S$\ref{sec:supp_navigation}.
+Further investigation of the C4 locus can be performed with \textit{odgi pav} to identify presence/absence variants (PAVs) with respect to one or multiple paths in the pangenome graph (not shown).}
 %This lets us decompose both homology and paralogy relationships by also untangling looping motifs in the graph, and obtaining linearized pairs of regions in BEDPE or PAF format.

### 2022-03-18 manuscript/main.tex

@@
 \item~\textit{odgi flatten} converts the graph to FASTA and BED (\S\ref{sec:supp_metrics}).
+\REVIEWED{\item~\textit{odgi pav} generates a presence-abscence variation table (\S\ref{sec:supp_pav}).}
 \item~\textit{odgi stats} provides numerical properties of the graph (\S\ref{sec:metrics}).
@@
 
-\section{Supplementary}
+\section{Supplement}
 
@@
 
-\subsection{Graph Editing}
+\subsection{Editing}
 \label{sec:supp_edit}
@@
 
-\subsection{Graph Sorting}
+\subsection{Sorting and node identifier compaction}
 \label{sec:supp_sorting}
@@
 
-\subsection{Graph Metrics}
+\subsection{Linear projections}
 \label{sec:supp_metrics}
@@
 To investigate on the genomic sequences encoded in the graph, \textit{odgi paths} allows users to calculate pairwise overlap statistics of groupings of paths and emit all path sequences in FASTA format, and it also allows the generation of a ``pangenome matrix'' that reports the copy number (presence/absence) of each path over each node.
-\textit{odgi flatten} generates a linearization of the graph by emitting the pangenome sequence (the concatenation of all node sequences) in FASTA format, and the projection of all paths on the linearized sequence in BED format.
-ODGI also offers more advanced tools for the interrogation of the graphs.
-To study very large pangenomes, users can use \textit{odgi bin} to summarize the path information into bins of a specified size, generating a summarized view of gigabase scale graphs in TSV or JSON file formats.
-This concept found its way into the pangenome graph ontology~\citep{Yokoyama2020}. %maybe we want to save this up for the discussion
+
+In standard practice, pangenome analysis examines presence-absence variations (PAVs), which correspond to loci that are present in some samples but not others.
+\textit{odgi pav} uses a set of genomic intervals (in BED expressed in the coordinate space of paths in the graph) to demarcate PAVs.
+It then describes the coverage of all paths relative to these PAVs, yielding matrix or tabular representation.
+The precise determination of PAVs based on graph topology remains an open problem, but practically any method capable of generating a BED file can be used here.
+This lets us define PAVs using repeat or gene annotations of the genomes represented as paths in the graph.
+A simple technique is to take the output of \textit{odgi flatten}, which generates a linearization of the graph by emitting the pangenome sequence (the concatenation of all node sequences) in FASTA format, and the projection of all paths on the linearized sequence in BED format relative to the graph's paths.
+
+ODGI also supports an experimental binned representation of graphs designed to support the study and visualization of pangenomes at different scales of resolution.
+\textit{odgi bin} summarizes graph path information into bins of a specified size, generating a summarized view of gigabase scale graphs in TSV or JSON file formats.
+We have further supported this binning approach in pangenome graph ontology model~\citep{Yokoyama2020}. %maybe we want to save this up for the discussion
 

### 2022-03-18 manuscript/main.tex
Revert "mention odgi pav"
This reverts commit 39a6684df83aaacad07bbb337509c6a236f7e191.

@@
 \textit{odgi untangle} is able to work with any sets of reference sequences, converting the graph to lift-over maps compatible with standard software for projecting annotations and alignments from one genome to another.
-An explanation of the untangling process is given in $\S$\ref{sec:supp_navigation}.
-Further investigation of the C4 locus can be performed with \textit{odgi pav} to identify presence/absence variants (PAVs) with respect to one or multiple paths in the pangenome graph (not shown).}
+An explanation of the untangling process is given in $\S$\ref{sec:supp_navigation}.}
 %This lets us decompose both homology and paralogy relationships by also untangling looping motifs in the graph, and obtaining linearized pairs of regions in BEDPE or PAF format.

### 2022-04-01 manuscript/main.tex
correct subsection numbering
nice to PAV you
split out supplement
remove red
fix affiliations

@@
 \newcommand{\beginsupplement}{%
+    \renewcommand{\thesection}{S}
 	\setcounter{table}{0}
@@
 \address{
-$^{\text{\sf 1}}$Genomics Research Centre, Human Technopole, Milan, Italy \\
-$^{\text{\sf 2}}$Quantitative Biology Center (QBiC), University of T\"ubingen, T\"ubingen, Germany, 72076 \\
+$^{\text{\sf 1}}$Genomics Research Centre, Human Technopole, Viale Rita Levi‑Montalcini 1, Milan, 20157, Italy \\
+$^{\text{\sf 2}}$Quantitative Biology Center (QBiC), University of T\"ubingen, T\"ubingen, 72076, Germany \\
 $^{\text{\sf 3}}$Biomedical Data Science, Dept. of Computer Science, University of T\"ubingen, T\"ubingen, Germany, 72076 \\
-$^{\text{\sf 4}}$University of Tennessee Health Science Center, Memphis, TN, USA
+$^{\text{\sf 4}}$Department of Genetics, Genomics and Informatics, University of Tennessee Health Science Center, Memphis, 38163, Tennessee, USA
 }
@@
 \item~\textit{odgi flatten} converts the graph to FASTA and BED (\S\ref{sec:supp_metrics}).
-\REVIEWED{\item~\textit{odgi pav} generates a presence-abscence variation table (\S\ref{sec:supp_pav}).}
+\REVIEWED{\item~\textit{odgi pav} computes presence-abscence variations (\S\ref{sec:supp_metrics}).}
 \item~\textit{odgi stats} provides numerical properties of the graph (\S\ref{sec:metrics}).
@@
 \clearpage
+\setcounter{page}{1}
+
 \beginsupplement

### 2022-04-27 manuscript/main.tex
nsf
update Simon's affiliations

@@
 \newcommand{\FIXME}[1]{\red{[FIXME: #1]}}
-\newcommand{\REVIEWED}[1]{{\textcolor{Red}{#1}}}
+\newcommand{\REVIEWED}[1]{{\textcolor{Black}{#1}}}
 \newcommand{\beginsupplement}{%
@@
 Andrea~Guarracino\orcidlink{0000-0001-9744-131X}\,$^{\text{\sfb 1} \dagger}$,
-Simon~Heumos\orcidlink{0000-0003-3326-817X},$^{\text{\sfb 2} \dagger}$
+Simon~Heumos\orcidlink{0000-0003-3326-817X},$^{\text{\sfb 2,3} \dagger}$
 Sven~Nahnsen\orcidlink{0000-0002-4375-0691}$^{\text{\sfb 2,3}}$, \\
@@
 
-We gratefully acknowledge support from NIH/NIDA U01DA047638 (EG) and NIH/NIGMS R01GM123489 (EG and PP).
+We gratefully acknowledge support from NIH/NIDA U01DA047638 (EG), NIH/NIGMS R01GM123489 (EG and PP), and NSF PPoSS Award \#2118709 (EG and PP).
 SH acknowledges funding from the Central Innovation Programme (ZIM) for SMEs of the Federal Ministry for Economic Affairs and Energy of Germany.
